### 2018-06-11 main.tex
first commit

\documentclass{article}

\usepackage{mystyle}
\usepackage{mycommands}

\title{Continuous Reinforcement Learning}

\author{
  Corentin Tallec, Léonard Blier
}

\begin{document}
\maketitle

\begin{abstract}
Abstract
\end{abstract}

\section{Introduction}
\label{sec:introduction}

\section{Definitions and Notations}
\label{sec:definitions}

\begin{definition}[Continuous Markov Reward Process]
  A continuous markov decision process $(X_{t}, R_{t})_{t\in\BR_{+}}$ is a continuous random process defined by 
  \begin{itemize}
  \item $\mathcal{X}$ is the state set 
  \item 
  \end{itemize}
  process$\mathcal{X}$ is the set of states, $T$ a transition function, $r$ a reward function
\end{definition}

\section{Ordre de grandeur Corentin}
\label{sec:ordre-de-grandeur}

\section{MCTS continu}
\label{sec:mcts-continu}

\section{Conclusion}
\label{sec:conclusion}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2018-06-12 magnitudes.tex
Adding separate file magnitudes.tex

%!TEX root = main.tex

\section{Intuitions}
\label{sec:intuitions}

### 2018-06-12 magnitudes.tex
[ADD] Magnitudes.tex

@@
 \label{sec:intuitions}
+Reinforcement learning aims at tackling a wide variety of real world problems.
+Among them obviously stand physical problems, where an agent interacts with a
+physical environment, e.g.\ the real world.
+
+Resistance to a change of time discretization is a desirable property for
+algorithms aiming at attacking the reinforcement learning field in physical
+setups: such algorithms should only improve as the time discretization is
+refined, as refining the time discretization can only increase the amount of 
+eploitable information.

### 2018-06-12 magnitudes.tex
[ADD] Intuitions introduction.

@@
 eploitable information.
+
+However, some reinforcement learning algorithms such as \emph{Q-Learning} are
+ill-behaved in the regime of small time discretization. In what follow, we
+analyze in an handwavy manner the reason for this failure, and give an alternative
+algorithm that remains viable in the limit of small discretizations.

### 2018-06-12 magnitudes.tex
[ADD] Adding macros.
[ADD] Adds...

@@
 algorithm that remains viable in the limit of small discretizations.
+
+\subsection{Notations}
+In what follows, we loosely consider the notion of a $\emph{Markov Decision Process}$
+$\mathcal{M}$ with discretization step $\deltat$. Such an MDP is the discretization
+of a continuous MDP with discretization step $\deltat$. Notably, this means that
+contiguous states are at distance $\bigO{\deltat}$ and that instantaneous rewards
+have magnitude $\bigO{\deltat}$ for any policy, i.e.
+\begin{align}
+	\|s_{t + \deltat} - s_t\| &= \bigO{\deltat}\\
+	r_t = \reward_t \deltat, &\quad \reward_t = \bigO{1}.
+\end{align}
+Policies are functions from the state space $\statespace$ to distributions on the
+$\actionspace$
+\begin{equation}
+	\pi\colon \statespace \mapsto \mathcal{D}\left(\actionspace\right).
+\end{equation}
+The probability of an action $a$ in a given state $s$ under policy $\pi$ is
+denoted by $\pi\left(a\mid s\right)$.
+
+We consider discount factors of the form
+\begin{equation}
+	\gamma' = \left[1 - (1 - \gamma) \deltat\right],
+\end{equation}
+so has to have an effective discount factor close to $\gamma$.
+
+The state value function and action-state value function of a policy
+$\pi$ are defined as
+\begin{align}
+	V^\pi(s) &= \BE_\pi\left[
+		\sum\limits_{t=s}^\infty
+		{\gamma'}^{s - t} \reward_{t + (s-t)\deltat} \deltat
+		\mid
+		s_t = s
+	\right]\\
+	Q^\pi(s, a) &= \BE_\pi\left[
+		\sum\limits_{t=s}^\infty
+		{\gamma'}^{s - t} \reward_s \deltat
+		\mid
+		s_t = s, a_t = a
+	\right].
+\end{align}
+Both functions satisfy recurrent \emph{Bellman equations}
+\begin{align}
+	V^\pi(s) &= \BE_\pi\left[
+		\reward_t \deltat + \gamma' V^\pi(s_{t + \deltat})
+		\mid
+	s_t = s\right]\\
+	Q^\pi(s, a) &= \BE_\pi\left[
+		\reward_t \deltat + \gamma' V^\pi(s_{t + \deltat})
+		\mid
+	s_t = s, a_t = a\right].
+\end{align}
+
+\subsection{$Q^\pi = V^\pi$ in continuous time}
+\emph{Q-learning} uses the state-action value function to determine which
+action optimizes the future return. However, in continuous time, the state-action
+value function no longer carries any information on which action carries the best
+return. Intuitively, as the discretization step goes to zero, the difference
+in resulting state between two actions is of magnitude $\bigO{\deltat}$. Consequently,
+the value functions of the corresponding states, and thus the state-action value functions
+only differ by a factor of magnitude $\bigO{\deltat}$.

### 2018-06-12 magnitudes.tex
.

--- 2018-06-12 magnitudes.tex
@@
 algorithm that remains viable in the limit of small discretizations.
+
+\subsection{Notations}
+In what follows, we loosely consider the notion of a $\emph{Markov Decision Process}$
+$\mathcal{M}$ with discretization step $\deltat$. Such an MDP is the discretization
+of a continuous MDP with discretization step $\deltat$. Notably, this means that
+contiguous states are at distance $\bigO{\deltat}$ and that instantaneous rewards
+have magnitude $\bigO{\deltat}$ for any policy, i.e.
+\begin{align}
+	\|s_{t + \deltat} - s_t\| &= \bigO{\deltat}\\
+	r_t = \reward_t \deltat, &\quad \reward_t = \bigO{1}.
+\end{align}
+Policies are functions from the state space $\statespace$ to distributions on the
+$\actionspace$
+\begin{equation}
+	\pi\colon \statespace \mapsto \mathcal{D}\left(\actionspace\right).
+\end{equation}
+The probability of an action $a$ in a given state $s$ under policy $\pi$ is
+denoted by $\pi\left(a\mid s\right)$.

### 2018-06-15 main.tex
Accent and tiny tex formatting.
[ADD] Many docs compilation
[ADD] Adding macros.
[ADD] Adds...
first version of the convergence proof for discretization of MRP

@@
 \usepackage{mycommands}
+\usepackage{lmodern}
+
+\usepackage{todonotes}
+\input{macros.tex}
 
@@
 
-\author{
-  Corentin Tallec, Léonard Blier
-}
+\author{L\'eonard Blier, Corentin Tallec}
 
@@
 \begin{definition}[Continuous Markov Reward Process]
-  A continuous markov decision process $(X_{t}, R_{t})_{t\in\BR_{+}}$ is a continuous random process defined by 
+  A continuous markov reward process $(X_{t}, R_{t})_{t\in\BR_{+}}$ is a continuous random process defined by 
   \begin{itemize}
   \item $\mathcal{X}$ is the state set 
-  \item 
+  \item $T(x)$ the dynamics and $\Sigma (x)$ the noise
+  \item $\mathcal{R}(r|x)$ the reward law
+  \item $\gamma$ the discount factor
   \end{itemize}
-  process$\mathcal{X}$ is the set of states, $T$ a transition function, $r$ a reward function
+  The law for the process $X_{t}$ is given by
+  \begin{equation}
+    \label{eq:eqdiffstoch}
+    dX_{t} = T(X_{t}) + \Sigma(X_{t})dB_{t}
+  \end{equation}
 \end{definition}
 
-\section{Ordre de grandeur Corentin}
-\label{sec:ordre-de-grandeur}
+\begin{definition}[Discretization of the space]
+  A discretization of the space $(\mathcal{X}, \BR_{+})$ is a $(X, \delta t)$ where :
+  \begin{itemize}
+  \item $X$ is a partition of $X$
+  \item $\delta t > 0$
+  \end{itemize}
+  We define the precision $\delta(X)$ of the partition $X$ to be :
+  \begin{equation}
+    \label{eq:2}
+    \delta(X) = \sup_{s\in X}\sup_{x,y \in s}\|x-y\|
+  \end{equation}
+\end{definition}
+
+\begin{definition}[Convergence of discretizations]
+  We say that a sequence $(X_{k}, \delta t_{k})$ of discretizations of $\mathcal{X}$ converges if $(\delta(X_{k}), \delta t_{k}) \rightarrow 0$. 
+\end{definition}
+
+\begin{definition}[Discretization of a Markov Reward Process]
+  Let $(\mathcal{S}, T, R, \gamma)$ be a continuous MRP, and $(S, \delta t)$ a discretization of $\mathcal{X}$.
+We define the approximation of the continuous MRP associated to this discretization as the discrete MRP $(\tilde S, \tilde{T}, \tilde{R}, \tilde{\gamma})$ where :
+  \begin{itemize}
+  \item $\tilde{T}(s,s') = \BP(X_{t+\delta t}\in s' | X_{t}\in s)$. \todo{Formalize that $X_{t}$ uniform in $s$}
+  \item $R(s) \sim \int_{t}^{t+\delta t}R_{t'}dt'$\todo{idem}
+  \item $\tilde{\gamma} = \gamma^{\delta t}$ 
+  \end{itemize}
+\end{definition}
+
+\begin{theorem}
+  Let $(X_{t}, R_{t})$ be a continuous MRP, and $(X_{k}, \delta t_{k})$ a sequence of approximations which converges to $\mathcal{X}$.
+Let $(\tilde{X}^{k}_{t}, \tilde{R}^{k}_{t})$ be the associated discretization.
+Then, for all $T > 0$,
+  \begin{equation}
+    (X^{k}_{t}, R^{k}_{t})_{0 \leq t \leq T} \rightarrow (X_{t}, R_{t})_{0 \leq t \leq T}
+  \end{equation}
+\end{theorem}
+
+We define the operators on $\mathcal{M}$ :
+\begin{itemize}
+\item $U_{S}\mu = \sum_{s \in S}\frac{\mu(s)}{\ell(s)}1_{s}$
+\item $P_{S}\mu = \sum_{s \in S}\frac{\mu(s)}{\ell(s)}\delta_{x_{s}}$
+\item Family $(T_{t})_{t\in \BR_{+}}$, $T_{t}\mu$ is the law of $X_{t}$ if $X_{0}\sim \mu$.
+\item For a given $\delta t$, $\tilde{T}_{t}\mu$ is the law of $\tilde{\tilde X}_{t}$ if $X_{0}\sim \mu$. We have :
+  \begin{equation}
+    \tilde{T}_{\delta t} = P_{S}T_{\delta t}U_{S}
+  \end{equation}
+  and
+  \begin{equation}
+    \tilde{T}_{t} = \tilde{T}_{\delta t}^{\lfloor t/\delta t\rfloor}
+  \end{equation}
+\end{itemize}
+In the following, $W = W_{2}$ is the Wasserstein distance.
+\begin{lemma}
+  \begin{equation}
+    W(U_{S}\mu, \mu) \leq \delta(S)
+  \end{equation}
+  \begin{equation}
+    W(P_{S}\mu, \mu) \leq \delta(S)
+  \end{equation}
+  \begin{equation}
+    U_{S}P_{S} = U_{S}
+  \end{equation}
+  \begin{equation}
+    P_{S}U_{S} = P_{S}
+  \end{equation}
+\end{lemma}
+
+\begin{lemma}
+  Assume that $T$ is Lipschitz and $\mathcal{X}$ is compact, 
+  \begin{equation}
+    W(T_{t}\mu, T_{t}\nu) \leq (1+Ct)W(\mu, \nu)
+  \end{equation}
+\end{lemma}
+
+\begin{proof}
+  \todo{todo} IDEA : stability of $T$ and $\Sigma$
+\end{proof}
+
+\begin{lemma}
+   Let $(S_{k}, \delta t_{k})$ be a sequence of approximations which converges to $\mathcal{X}$ such that $\varepsilon_{k} := \frac{\delta(S_{k})}{\delta t_{k}} \rightarrow 0$.
+Let $(\tilde{X}^{k}_{t}, \tilde{R}^{k}_{t})$ be the associated discretization.
+Then, for all $t > 0$,
+  \begin{equation}
+    W(\tilde T^{k}_{t}\mu, T_{t}\nu) \leq W(\mu,\nu)e^{tC} + o_{k\rightarrow \infty}(1)
+  \end{equation}
+\end{lemma}
+
+\begin{proof}
+  Let $(S,\delta t)$ be a discretization.
+
+  \begin{align}
+    W(\tilde T_{\delta t}\mu, T_{t}\nu) &= \\
+  W(P_{S}(T_{\delta t}U_{S})^{\lfloor t/\delta t\rfloor}\mu , T_{t}\nu) &\leq \\ W(T_{\lfloor t/\delta t\rfloor \delta t}\mu, T_{t}\mu) + \delta(S) + W((T_{\delta t}U_{S})^{\lfloor t/\delta t\rfloor}\mu , T_{\delta t}^{\lfloor t/\delta t\rfloor}\nu)
+  \end{align}
+
+  But
+  \begin{equation}
+  W(T_{\lfloor t/\delta t\rfloor \delta t}\mu, T_{t}\mu) = W(T_{\delta t\{t/\delta t\}}\nu , \nu) \leq C_{1}\delta t
+\end{equation}
+Moreover :
+\begin{align}
+  W((T_{\delta t}U_{S})^{n}\mu , T_{\delta t}^{n}\nu) &= W(T_{\delta t}\left(U_{S}(T_{\delta t}U_{S})^{n-1}\mu\right) , T_{\delta t}\left(T_{\delta t}^{n-1}\nu\right)) \\
+                                                      &\leq (1+C\delta t)  W(U_{S}(T_{\delta t}U_{S})^{n-1}\mu , T_{\delta t}^{n-1}\nu)\\
+  & \leq (1+C\delta t)  (\delta(S) + W((T_{\delta t}U_{S})^{n-1}\mu , T_{\delta t}^{n-1}\nu))
+\end{align}
+Then, by induction :
+\begin{align}
+  W((T_{\delta t}U_{S})^{n}\mu , T_{\delta t}^{n}\mu) &\leq (1+C\delta t)\delta(S)\frac{(1+C\delta t)^{n} - 1}{C\delta t} + W(\mu, \nu)(1+C\delta t)^{n}
+\end{align}
+Therefore :
+\begin{align}
+  W((T_{\delta t}U_{S})^{\lfloor t/\delta t\rfloor}\mu , T_{\delta t}^{\lfloor t/\delta t\rfloor}\mu) &\leq (1+C\delta t)\delta(S)\frac{(1+C\delta t)^{\lfloor t/\delta t\rfloor} - 1}{C\delta t} + W(\mu, \nu)(1+C\delta t)^{\lfloor t/\delta t\rfloor} \\
+                                                                                                      &\leq (1+C\delta t)\delta(S)\frac{e^{tC} - 1}{C\delta t} + W(\mu, \nu)e^{tC}\\
+                                                                                                        &= W(\mu, \nu)e^{tC} + o(1)
+\end{align}
+Therefore, 
+\begin{align}
+  W(P_{S}(T_{\delta t_{k}}U_{S_{k}})^{\lfloor t/\delta t_{k}\rfloor}\mu, T_{t}\mu) & \leq W(\mu, \nu)e^{tC} + o(1) + \delta(S_{k}) + C'\delta t_{k}  \\
+  &= W(\mu, \nu)e^{tC} + o(1)
+\end{align}
+\end{proof}
+
+\begin{lemma}
+  Let $0 < t_{1} < ... < t_{l}<T$.
+  \begin{equation}
+    W((\tilde X_{t_{1}}, ..., \tilde X_{t_{l}}), (X_{t_{1}}, ..., X_{t_{l}})) \leq W(\mu,\nu)(e^{t_{1}C}+...+e^{t_{l}C}) + o(1)
+  \end{equation}
+\end{lemma}
+
+\begin{proof}
+  Let $P$ be the law of $(X_{t_{1}}, ..., X_{t_{l}})$ with $X_{0} \sim \nu$, and $\tilde P$ be the law of $(\tilde X_{t_{1}}, ..., \tilde X_{t_{l}})$ with $X_{0} \sim P_{s}\mu$.
+  For all $\mu, \nu$, we use the notation $\pi_{\mu,\nu}$ for a measure achieving the optimal transport distance up to $\epsilon$.
+
+  We define :
+  \begin{align}
+    \pi((\tilde x_{1},..., \tilde x_{l}), (x_{1}, ..., x_{l})) = \\ = \pi_{\tilde T_{t_{1}}\mu T_{t_{1}\nu}}(\tilde x_{1}, x_{1})\prod_{i=2}^{l}\pi_{\tilde T_{t_{i}-t_{i-1}}\delta_{\tilde x_{i-1}}, T_{t_{i}-t_{i-1}}\delta_{x_{i-1}}}(\tilde x_{i}, x_{i})
+  \end{align}
+  Then,
+  \begin{align}
+    & W((\tilde X_{t_{1}}, ..., \tilde X_{t_{l}}), (X_{t_{1}}, ..., X_{t_{l}}))
+    \\ & \leq \int_{(\tilde x_{1},..., \tilde x_{l}), (x_{1},..., x_{l})}\sum_{i=1}^{l}\|\tilde x_{i} - x_{i}\|^{2}\pi((\tilde x_{1},..., \tilde x_{l}), (x_{1}, ..., x_{l}))
+         \\ &\leq  W((\tilde X_{t_{1}}, ..., \tilde X_{t_{l-1}}), (X_{t_{1}}, ..., X_{t_{l-1}})) + \int_{(\tilde x_{1},..., \tilde x_{l}), (x_{1},..., x_{l})}\|\tilde x_{l} - x_{l}\|^{2}\pi((\tilde x_{1},..., \tilde x_{l}), (x_{1}, ..., x_{l}))
+  \end{align}
+
+  But we have :
+  \begin{align}
+    \int_{(\tilde x_{1},..., \tilde x_{l}), (x_{1},..., x_{l})}\|\tilde x_{l} - x_{l}\|^{2}\pi((\tilde x_{1},..., \tilde x_{l}), (x_{1}, ..., x_{l}))  \\
+    \leq \int_{(\tilde x_{1},..., \tilde x_{l-1}), (x_{1},..., x_{l-1})}W(\tilde T_{t_{l}-t_{l-1}}\delta_{\tilde x_{l-1}}, T_{t_{l}-t_{l-1}}\delta_{x_{l-1}})^{2}\pi((\tilde x_{1},..., \tilde x_{l-1}), (x_{1}, ..., x_{l-1}))\\
+    \leq \int_{(\tilde x_{1},..., \tilde x_{l-1}), (x_{1},..., x_{l-1})}(W(\delta_{\tilde x_{l-1}}, \delta_{x_{l-1}})^{2}e^{(t_{l}-t_{l-1})C}+\varepsilon)\pi((\tilde x_{1},..., \tilde x_{l-1}), (x_{1}, ..., x_{l-1}))\\
+    \leq \int_{(\tilde x_{1},..., \tilde x_{l-1}), (x_{1},..., x_{l-1})}(\|\tilde x_{l-1} - x_{l-1}\|^{2}e^{(t_{l}-t_{l-1})C}+\varepsilon)\pi((\tilde x_{1},..., \tilde x_{l-1}), (x_{1}, ..., x_{l-1}))
+  \end{align}
+
+  By induction :
+  \begin{align}
+    & W((\tilde X_{t_{1}}, ..., \tilde X_{t_{l}}), (X_{t_{1}}, ..., X_{t_{l}}))
+    \\ &\leq  W((\tilde X_{t_{1}}, ..., \tilde X_{t_{l-1}}), (X_{t_{1}}, ..., X_{t_{l-1}})) + W(\mu,\nu)e^{t_{l}C}
+    \\ &\leq W(\mu,\nu)(e^{t_{1}C}+...+e^{t_{l}C}) + o(1)
+  \end{align}
+\end{proof}
+
+\begin{proof}[Proof of the theorem]
+  We use theorem 13.5 from Billingsley \todo{cite}. We have to check two hypothesis :
+  \begin{enumerate}
+  \item For all $t_{1}<t_{2}<...<t_{l}<T$, $(X_{t_{1}}^{k}, ..., X_{t_{l}}^{k}) \rightarrow (X_{t_{1}}, ..., X_{t_{l}})$ weakly.
+  \item For all $r \leq s \leq t$, $\BE\left[|X_{s}-X_{r}|^{2}|X_{t}-X_{s}|^{2}\right] \leq (F(t) - F(r))^{2}$ where F is a continuous non-decreasing function.
+  \end{enumerate}
+  For the first point, done in Lemma
+
+  For the second point \todo{To formalize}:
+  \begin{align}
+    \BE\left[|X_{s}-X_{r}|^{2}|X_{t}-X_{s}|^{2}\right] &\leq C(r-s)(t-s)\\
+    &\leq C\frac{(t-r)^{2}}{4}
+  \end{align}
+  Then $F(t) = t\sqrt{C}/2$ solves the problem.
+\end{proof}
+
+\input{magnitudes.tex}
 

### 2018-06-15 magnitudes.tex
[FIX] HEAD remaining.

--- 2018-06-12 magnitudes.tex
@@
 \emph{Q-learning} uses the state-action value function to determine which
-action optimizes the future return. However, in continuous time, the state-action
-value function no longer carries any information on which action carries the best
-return. Intuitively, as the discretization step goes to zero, the difference
-in resulting state between two actions is of magnitude $\bigO{\deltat}$. Consequently,
-the value functions of the corresponding states, and thus the state-action value functions
-only differ by a factor of magnitude $\bigO{\deltat}$.
+action optimizes the future return.
+
+However, in continuous time, the state-action value function no longer carries
+any information on which action yields the best return. Intuitively, as the
+discretization step goes to zero, the difference in resulting state between two
+actions is of magnitude $\bigO{\deltat}$. Consequently, the value functions of
+the corresponding states, and thus the state-action value functions only differ
+by a factor of magnitude $\bigO{\deltat}$.
+
+Formally,
+\begin{align}
+	V^\pi(s) - Q^\pi(s, a) &= 
+	(\BE_\pi\left[\reward_t \deltat \mid s_t=s\right] -
+	\BE_\pi\left[\reward_t \deltat \mid s_t=s\right]) \deltat +\\
+	&\gamma'(\BE_\pi\left[
+		V^\pi(s_{t + \deltat})
+		\mid s_t = s
+	\right] - \BE_\pi\left[
+		V^\pi(s_{t + \deltat})
+		\mid s_t = s, a_t=a
+	\right]).\nonumber
+\end{align}
+Since $s_{t+\deltat} = s_t + \bigO{\deltat}$ and $V^\pi$ is differentiable,
+$V^\pi(s_{t+\deltat}) = V^\pi(s_t) + \bigO{\deltat}$. Consequently,
+\begin{equation}
+	V^\pi(s) - Q^\pi(s, a) = \bigO{\deltat}.
+\end{equation}
+
+Notably, decision based on the values or relative values of an approximation of
+the $Q^\pi(s, a)$'s for different $a$ are only meaningful when the
+approximation is tighter than $\bigO{\deltat}$, at least in term of relative
+values, i.e.  
+\begin{equation*}
+	|\tilde{Q}^\pi(s, a) - \tilde{Q}^\pi(s, a') - Q^\pi(s, a) +
+	Q^\pi(s, a')| = \bigO{\deltat}.
+\end{equation*}.
+
+Approximating with this degree of precision becomes harder as $\bigO{\deltat}$
+decreases. \NDC{This is not convincing at the moment: as $\deltat$ decreases,
+	the number of observations made increases, and this may help get a better
+	estimate $\tilde{Q}^\pi(s, a)$. My guess is that the number of observations
+	required to obtain an approximation of order $\varepsilon$ is superlinear in
+	$\frac{1}{\varepsilon}$ and thus the previous argument holds. We either need
+a simple theoretical setup where this works or an experimental one.}

### 2018-06-22 magnitudes.tex

@@
 	Q^\pi(s, a')| = \bigO{\deltat}.
-\end{equation*}.
+\end{equation*}
 
@@
 a simple theoretical setup where this works or an experimental one.}
+
+\paragraph{Directly learning the advantage function.}
+Let the system be defined by a state- and action-driven stochastic
+differential equation
+\begin{equation}
+s_{t+\deltat}=s_t+\deltat F(s_t,a_t)+\sqrt{\deltat}\mathcal{N}(0,C(s_t))
+\end{equation}
+where we assume that the covariance noise does not depend on $a_t$. (If
+$a_t$ itself is noisy, this will produce noise on $F$, but this noise
+would typically be of order $\deltat$ unless the noise on $a_t$ is of
+order $1/\sqrt{\deltat}$.)
+
+Then with $\gamma=1$ the Bellman equation is
+\begin{align}
+Q(s_t,a_t)=&\E[ \reward_t(s_t,a_t) \deltat + V(s_{t+\deltat}) \mid a_t]
+\\&=V(s_t)+\deltat \,\E\left[\reward_t(s_t,a_t)+\frac{\partial V(s_t)}{\partial
+s}F(s_t,a_t)+\frac12 \Tr \left(C(s_t)\frac{\partial^2 V(s_t)}{\partial
+s^2}\right)\mid a_t\right]+o(\deltat)
+\end{align}
+
+We only need to compute the advantage function $A$ defined by
+$Q(s_t,a_t)\eqd V(s_t)+\deltat A(s_t,a_t)$; or just the advantage function
+up to any term that depends only $s_t$ but not on $a_t$. By isolating the
+part that
+depends on $a_t$ in the above, the advantage function is
+\begin{equation}
+A(s_t,a_t)=\E\left[\reward_t(s_t,a_t)+\frac{\partial
+V(s_t)}{\partial
+s}F(s_t,a_t)\mid a_t\right]+\cdots(s_t)+o(1)
+\end{equation}
+
+Thus it is enough to learn a model of the instantaneous reward, a model
+of the value function $V$ (sufficiently regularized to allow for
+estimation of $\partial V(s)/\partial s$), and a model of the
+instantaneous dynamics $F(s,a)$.

### 2018-06-22 magnitudes.tex

@@
 \paragraph{Directly learning the advantage function.}
+(See Kenji Doya, Reinforcement Learning in Continuous Space and Time,
+Neural Comp.\ 2000.)
 Let the system be defined by a state- and action-driven stochastic

### 2018-06-22 magnitudes.tex

@@
 instantaneous dynamics $F(s,a)$.
+
+This extends to situations with continuous time and continuous or
+discrete space, where the state jumps from $s$ to $s'$ with probability $\deltat
+L(s,a,s')$ at each time. (Here $L$ is signed, with $L(s,a,s)$ usually
+negative to represent the jump from $s$; $L$ is the generator of the
+Markov process.) Writing $p_J(s,a,s')$ the law of
+$s'$ knowing that a jump has occurred, and $\deltat J(s,a)$ the
+probability of jumping, we have $L(s,a,s')=J(s,a)(p_J(s,a,s')-\delta(s))$
+and

### 2018-06-22 magnitudes.tex

@@
 and
+\begin{equation}
+A(s_t,a_t)=\E\left[\reward_t(s_t,a_t)+J(s_t,a_t)\,\E_{s'\sim
+p_J(s_t,a_t,s')} (V(s')-V(s_t))
+\mid a_t\right]+\cdots(s_t)
+\end{equation}
+
+This also extends to any combination of continuous trajectories and
+probabilistic jumps with probability $O(\deltat)$. However, this does not
+cover some kinds of jumps, such as jumps occurring with probability $1$
+if some part of the state reaches some threshold. Indeed, this relies on
+$s_{t+\deltat}$ being somehow close to $s_t$, either in distance or in
+probability.

### 2018-09-19 magnitudes.tex
Commiting some details both for Q-learning and advantage learning.

@@
 \subsection{$Q^\pi = V^\pi$ in continuous time}
+\label{sec:QContinuous}
+% {{{
 \emph{Q-learning} uses the state-action value function to determine which
@@
 Approximating with this degree of precision becomes harder as $\bigO{\deltat}$
-decreases. \NDC{This is not convincing at the moment: as $\deltat$ decreases,
+decreases. Typically, if the Q-function is learned only from \emph{Monte Carlo} samples
+of the return, the observed variance of the quantity to learn is of order
+$\frac{1}{{(1 - \gamma)} ^ 2 \deltat}$. An optimist estimate of the number of
+samples required to learn this quantity up to order $\bigO{\deltat}$ is 
+$\bigO{\frac{1}{{(1 - \gamma)}^2 \deltat^3}}$. If a sampling rate of $\deltat$ is
+assumed, the time it takes to learn $Q$ up to the correct degree of precission,
+increases quadratically with $\frac{1}{\deltat}$. Indeed, in a single real time step, with a
+sampling rate of $\deltat$, $\frac{1}{\deltat}$ samples are aquired. Consequently,
+$\frac{1}{{(1 - \gamma)}^2 \deltat^2}$ timesteps are required to obtain an estimate up
+to the desired precision.
+\NDC{This is not convincing at the moment: as $\deltat$ decreases,
 	the number of observations made increases, and this may help get a better
@@
 a simple theoretical setup where this works or an experimental one.}
+% }}}
 
 \paragraph{Directly learning the advantage function.}
+% {{{
 (See Kenji Doya, Reinforcement Learning in Continuous Space and Time,
@@
 probability.
+
+(Should also check ``Advantage updating'' Baird 1993, Harmon et al 1996.)
+% }}}
+
+\paragraph{Questions left unanswered with advantage learning.}
+As mentionned, advantage learning requires learning the gradient of $V$
+with respect to the states, or, in the discrete case, the difference of
+$V$'s between two neighboring rescaled by $\frac{1}{\deltat}$. Now, as
+mentionned in~\ref{sec:QContinuous}, $V$ can only be learned up to $\bigO{1}$
+in reasonable time. Consequently, $V(s) - V(s') = \bigO{1}$ in the general case,
+and $\frac{V(s) - V(s')}{\deltat} = \bigO{\frac{1}{\deltat}}$. This means that
+the learning procedure of $A$ can endure a bias of order $\bigO{\frac{1}{\deltat}}$
+which goes to $+\infty$ as $\deltat \to 0$.
+
+One could think that adding a gradient penalization to the learning criterion of
+$V$, or similarily, restricting the class of candidate function to a subset of 
+the set of $K$-lipschitz functions would enforce convergence of the gradient of the
+approximate $V$ function to the gradient of the true $V$ function as the approximate
+$V$ function gets closer to the true $V$ function. This is not true
+in the general case, as any constant function can be uniformly approximated by a
+$K$-lipschitz function whose derivative has unit infinity norm. A $K$-lipschitz
+approximation of the true $V$ can get arbitrarily close, while its derivative remains
+far from the true derivative.

### 2018-09-21 magnitudes.tex
Minor changes

@@
 $\mathcal{M}$ with discretization step $\deltat$. Such an MDP is the discretization
-of a continuous MDP with discretization step $\deltat$. Notably, this means that
-contiguous states are at distance $\bigO{\deltat}$ and that instantaneous rewards
-have magnitude $\bigO{\deltat}$ for any policy, i.e.
+of a continuous MDP with discretization step $\deltat$.
+Formally,
 \begin{align}
-	\|s_{t + \deltat} - s_t\| &= \bigO{\deltat}\\
-	r_t = \reward_t \deltat, &\quad \reward_t = \bigO{1}.
+	s_{t + \deltat} &= s_t + F(s_t, a_t) \deltat + \gauss(0, C(s_t, a_t)) \sqrt{\deltat}\\
+	r(s, a) &= \tilde{r}(s, a) \deltat + \gauss(0, \sigma_r(s, a)) \sqrt{\deltat}.
 \end{align}
@@
 
-We consider discount factors of the form
+Durations of the order of $\deltat$ are referred to as infinitesimal
+timesteps, while durations of the order of $1$ are referred to as timesteps.
+If the discount factor in timesteps is $\gamma$, the corresponding
+discount in infinitesimal timesteps is
 \begin{equation}
-	\gamma' = \left[1 - (1 - \gamma) \deltat\right],
+	\gamma^\deltat \approx (1 - (1 - \gamma \deltat)).
 \end{equation}
-so has to have an effective discount factor close to $\gamma$.
 
-The state value function and action-state value function of a policy
+The state value function and action-state value function under policy
 $\pi$ are defined as
 \begin{align}
-	V^\pi(s) &= \BE_\pi\left[
-		\sum\limits_{t=s}^\infty
-		{\gamma'}^{s - t} \reward_{t + (s-t)\deltat} \deltat
+	Q^\pi(s, a) &= \BE_\pi\left[
+		\sum\limits_{k=0}^\infty
+	{\gamma}^{k \deltat} r(s_k, a_k)
 		\mid
-		s_t = s
+		s_0 = s, a_0 = a
 	\right]\\
+	&= \BE_\pi\left[
+		\sum\limits_{k=0}^\infty
+		{\gamma}^{k \deltat} \tilde{r}(s_k, a_k) \deltat
+		\mid
+		s_0 = s, a_0 = a
+	\right]\\
+	V^\pi(s) &= \BE_\pi\left[Q^\pi(s_t, a_t) | s_t=s\right].
+\end{align}
+Both functions satisfy recurrent \emph{Bellman equations}, e.g.
+\begin{align}
 	Q^\pi(s, a) &= \BE_\pi\left[
-		\sum\limits_{t=s}^\infty
-		{\gamma'}^{s - t} \reward_s \deltat
+		\tilde{r}(s_t, a_t) \deltat + \gamma^\deltat V^\pi(s_{t + \deltat})
 		\mid
-		s_t = s, a_t = a
-	\right].
+	s_t = s, a_t = a\right]
 \end{align}
-Both functions satisfy recurrent \emph{Bellman equations}
+as well as $K$ steps \emph{Bellman equations}, for any $K$, and notably for
+$K = \frac{1}{\deltat}$, e.g.
 \begin{align}
-	V^\pi(s) &= \BE_\pi\left[
-		\reward_t \deltat + \gamma' V^\pi(s_{t + \deltat})
+	V^\pi(s) = \BE_\pi\left[
+		\sum\limits_{k=0}^{\frac{1}{\deltat} - 1} \gamma^{k\deltat} \tilde{r}(s_k, a_k) \deltat + \gamma V^\pi(s_{t + 1})
 		\mid
-	s_t = s\right]\\
-	Q^\pi(s, a) &= \BE_\pi\left[
-		\reward_t \deltat + \gamma' V^\pi(s_{t + \deltat})
+	s_0 = s\right].
+\end{align}
+With a fixed policy $\pi$, the trajectory of states in rollouts is a
+\emph{Markov Chain}. Under mild conditions, there exists a stationnary
+distribution $\rho$ for this markov chain. With those notations, the goal to
+maximize, the average return, is defined as
+$\BE_{s\sim\rho}\left[V^{\pi}(s)\right]$.
+Optimal $V$ and $Q$ functions, i.e. $V^*=V^{\pi^*}$ and $Q^*=Q^{\pi^*}$ with
+$\pi^*$ the policy that maximizes the average return, also satisfy Bellman equations
+equations, i.e.
+\begin{align}
+	Q^*(s, a) &= \BE\left[
+		\tilde{r}(s, a) \right] \deltat + \gamma^\deltat \max\limits_{a'} \BE\left[Q^*(s_{t + \deltat}, a')
 		\mid
 	s_t = s, a_t = a\right].
+	\label{eq:optq}
 \end{align}
+Popular approaches to optimizing
+the average return mainly fall in two categories:
+\begin{itemize}
+	\item $Q$-learning approaches, which use~\eqref{eq:optq} as a fixed point equation
+		to derive iterative approximations of the optimal $Q$ function.
+	\item Policy gradient approaches, which search through a parameterized-sub family of
+		all policies in order to optimize the average return through gradient descent. 
+		It is possible to estimate the gradient of the average return
+		using the policy gradient theorem, which yields
+		\begin{equation}
+			\nabla_\theta\BE_{s\sim\rho_\theta}\left[V^{\pi_\theta}(s)\right]
+			=
+			\BE_{s\sim\rho_\theta, a\sim\pi(\cdot\mid s)}\left[
+				Q^{\pi_\theta}(s, a)
+				\nabla_\theta \log \pi_\theta(a \mid s)
+			\right].
+		\end{equation}
+\end{itemize}
+Both approaches require estimating $Q^{\pi_\theta}$, and we will show that these
+estimations need to be precise up to $\bigO{\deltat}$.
 
-\subsection{$Q^\pi = V^\pi$ in continuous time}
+\paragraph{\emph{Monte Carlo} policy gradient in continuous time}~\\
+
+\paragraph{$Q^\pi = V^\pi$ in continuous time}~\\
 \label{sec:QContinuous}

### 2018-09-21 magnitudes.tex
Changes.

@@
 \label{sec:intuitions}
+% {{{
 Reinforcement learning aims at tackling a wide variety of real world problems.
@@
 
+% }}}
+
 \subsection{Notations}
+% {{{
 In what follows, we loosely consider the notion of a $\emph{Markov Decision Process}$
@@
 
-\paragraph{\emph{Monte Carlo} policy gradient in continuous time}~\\
+% }}}
 
-\paragraph{$Q^\pi = V^\pi$ in continuous time}~\\
+\paragraph{$Q$-functions and $V$-functions only differ by $\bigO{\deltat}$}~\\
 \label{sec:QContinuous}
@@
 any information on which action yields the best return. Intuitively, as the
-discretization step goes to zero, the difference in resulting state between two
-actions is of magnitude $\bigO{\deltat}$. Consequently, the value functions of
-the corresponding states, and thus the state-action value functions only differ
-by a factor of magnitude $\bigO{\deltat}$.
+discretization step goes to zero, the expected difference in resulting state
+values between two actions is of magnitude $\bigO{\deltat}$. Consequently, the
+state-action value functions only differ by a factor of magnitude
+$\bigO{\deltat}$.
 
@@
 \begin{align}
-	V^\pi(s) - Q^\pi(s, a) &= 
-	(\BE_\pi\left[\reward_t \deltat \mid s_t=s\right] -
-	\BE_\pi\left[\reward_t \deltat \mid s_t=s\right]) \deltat +\\
-	&\gamma'(\BE_\pi\left[
-		V^\pi(s_{t + \deltat})
-		\mid s_t = s
-	\right] - \BE_\pi\left[
-		V^\pi(s_{t + \deltat})
-		\mid s_t = s, a_t=a
-	\right]).\nonumber
+	&V^\pi(s) - Q^\pi(s, a)\\ 
+	= &\BE_{a'\sim \pi(\cdot\mid s)}
+	\left[\tilde{r}(s, a') - \tilde{r}(s, a)\right] \deltat
+	+
+	\gamma^\deltat 
+	(\BE\left[V^\pi(s_{t+\deltat}) \mid s_t=s\right]
+	-
+	\BE\left[V^\pi(s_{t+\deltat}) \mid s_t=s, a_t=a\right]).\nonumber
 \end{align}
-Since $s_{t+\deltat} = s_t + \bigO{\deltat}$ and $V^\pi$ is differentiable,
-$V^\pi(s_{t+\deltat}) = V^\pi(s_t) + \bigO{\deltat}$. Consequently,
+Now $s_{t + \deltat} = s_t + \F(s_t, a_t) \deltat +
+\gauss(0, C(s_t, a_t)) \sqrt{\deltat}$, thus 
 \begin{equation}
-	V^\pi(s) - Q^\pi(s, a) = \bigO{\deltat}.
+	V(s_{t+\deltat}) = V(s_t) + \partial_s V(s_t) \delta s_t
+	+ \frac{1}{2}\mathrm{tr}(C(s_t, a_t) H_V(s_t))\deltat.
 \end{equation}
+In expectation, irrespective of wether the expectation is conditionned on $a$,
+$V(s_{t+\deltat})$ and $V(s_t)$ are at distance $\bigO{\deltat}$. Consequently,
+$V^\pi(s) - Q^\pi(s, a) = \bigO{\deltat}$.
 
@@
 to the desired precision.
-\NDC{This is not convincing at the moment: as $\deltat$ decreases,
-	the number of observations made increases, and this may help get a better
-	estimate $\tilde{Q}^\pi(s, a)$. My guess is that the number of observations
-	required to obtain an approximation of order $\varepsilon$ is superlinear in
-	$\frac{1}{\varepsilon}$ and thus the previous argument holds. We either need
-a simple theoretical setup where this works or an experimental one.}
+% \NDC{This is not convincing at the moment: as $\deltat$ decreases,
+% 	the number of observations made increases, and this may help get a better
+% 	estimate $\tilde{Q}^\pi(s, a)$. My guess is that the number of observations
+% 	required to obtain an approximation of order $\varepsilon$ is superlinear in
+% 	$\frac{1}{\varepsilon}$ and thus the previous argument holds. We either need
+% a simple theoretical setup where this works or an experimental one.}
+% }}}
+
+\paragraph{\emph{Monte Carlo} policy gradient in continuous time}~\\
+% {{{
 % }}}
@@
 \paragraph{Questions left unanswered with advantage learning.}
+% {{{
 As mentionned, advantage learning requires learning the gradient of $V$
@@
 far from the true derivative.
+% }}}

### 2018-09-21 magnitudes.tex
Commit.

@@
 
-\paragraph{$Q$-functions and $V$-functions only differ by $\bigO{\deltat}$}~\\
+\paragraph{$Q$-functions and $V$-functions only differ by $\bigO{\deltat}$}
 \label{sec:QContinuous}
@@
 Approximating with this degree of precision becomes harder as $\bigO{\deltat}$
-decreases. Typically, if the Q-function is learned only from \emph{Monte Carlo} samples
+decreases. Typically, if the Q-function is learnt only from \emph{Monte Carlo} samples
 of the return, the observed variance of the quantity to learn is of order
-$\frac{1}{{(1 - \gamma)} ^ 2 \deltat}$. An optimist estimate of the number of
+$\frac{1}{1 - \gamma}$. An optimist estimate of the number of
 samples required to learn this quantity up to order $\bigO{\deltat}$ is 
-$\bigO{\frac{1}{{(1 - \gamma)}^2 \deltat^3}}$. If a sampling rate of $\deltat$ is
+$\bigO{\frac{1}{(1 - \gamma) \deltat^2}}$. If a sampling rate of $\deltat$ is
 assumed, the time it takes to learn $Q$ up to the correct degree of precission,
-increases quadratically with $\frac{1}{\deltat}$. Indeed, in a single real time step, with a
+increases linearily with $\frac{1}{\deltat}$. Indeed, in a single real time step, with a
 sampling rate of $\deltat$, $\frac{1}{\deltat}$ samples are aquired. Consequently,
-$\frac{1}{{(1 - \gamma)}^2 \deltat^2}$ timesteps are required to obtain an estimate up
-to the desired precision.
+$\frac{1}{(1 - \gamma) \deltat}$ timesteps are required to obtain an estimate up
+to the desired precision. As $\deltat \to 0$, it takes an infinite number of samples
+to obtain a viable estimate of $Q$, rendering $Q$-learning methods theoretically
+unsound.
 % \NDC{This is not convincing at the moment: as $\deltat$ decreases,

### 2018-09-21 magnitudes.tex
Changes.
Introducing A function.

@@
 		\begin{equation}
+			\label{eq:policygradient}
 			\nabla_\theta\BE_{s\sim\rho_\theta}\left[V^{\pi_\theta}(s)\right]
@@
 unsound.
+
+In what follows, we will refer to the advantage function $A^\pi(s, a)$ as
+$\frac{Q^\pi(s,a) - V^\pi(s)}{\deltat}$. Since $Q^\pi$ and $V^\pi$ only differ
+by a factor of scale $\bigO{\deltat}$, this quantity remains finite as $\deltat \to 0$.
 % \NDC{This is not convincing at the moment: as $\deltat$ decreases,
@@
 
-\paragraph{\emph{Monte Carlo} policy gradient in continuous time}~\\
+\paragraph{Policy gradient methods in continuous time} Policy gradient methods aim at
+maximizing the average return using gradient ascent on estimations of~\eqref{eq:policygradient}.
+One can observe that subtracting an action independent baseline to $Q^{\pi_\theta}$ in~\eqref{eq:policygradient} still yields an unbiased estimte of the gradient. Notably,
+\begin{equation}
+\end{equation}
+
 % {{{
@@
 $V$'s between two neighboring rescaled by $\frac{1}{\deltat}$. Now, as
-mentionned in~\ref{sec:QContinuous}, $V$ can only be learned up to $\bigO{1}$
+mentionned in~\ref{sec:QContinuous}, $V$ can only be learnt up to $\bigO{1}$
 in reasonable time. Consequently, $V(s) - V(s') = \bigO{1}$ in the general case,

### 2018-09-28 magnitudes.tex
Commit

@@
 \end{align}
-Now $s_{t + \deltat} = s_t + \F(s_t, a_t) \deltat +
+Now $s_{t + \deltat} = s_t + F(s_t, a_t) \deltat +
 \gauss(0, C(s_t, a_t)) \sqrt{\deltat}$, thus 
@@
 % }}}
+
+\paragraph{An experimental validation of \tdz convergence.}
+% {{{
+As mentionned in the previous paragraph, the learning algorithms described
+in~\ref{} are only expected to scale well if the value function approximated by
+\tdz converges in some kind of $\ho$ norm, or at least a norm that involves
+difference of function evaluations instead of function evaluations.
+
+We evaluate the validity of this statement on a toy 2D problem, where the
+value function can be efficiently nearly exactly approximated using finite
+element methods. To simplify things, we work with fixed policy, and only
+consider the state dynamic.
+
+Consider a stochastic continuous dynamical system evolving as
+\begin{equation}
+	ds_t = (\omega J -\theta) (s_t - \mu) dt + \sigma dB_t
+\end{equation}
+where $s$ is a $2$-d vector, $J$ is $\begin{pmatrix} 0 & -1 \\ 1 & 0\end{pmatrix}$,
+$\omega$ is an angular velocity parameter, $\mu$ is a mean parameter, $\theta$ is a
+recall parameter and $\sigma$ is a noise parameter. A typical trajectory of this process
+is given in Fig.~\ref{fig:trajectory}.
+\begin{figure}
+	\includegraphics[width=10cm]{imgs/trajectory.pdf}
+	\caption{Example trajectory, $\mu=0$, $\sigma=0.3$, $\omega=0.5$,
+	$\theta=0.03$}
+	\label{fig:trajectory}
+\end{figure}
+
+Given a reward function $r\colon \statespace \to \R$, Bellman equation yields
+\begin{equation}
+	V(s_{t}) = \BE\left[\tilde{r}(s_t)dt\right] + (1 - (1 - \gamma)dt)\BE\left[V(s_{t+dt})\right]
+\end{equation}
+Using Ito's lemma,
+\begin{align}
+	V(s_{t}) &= \BE\left[\tilde{r}(s_t)dt\right] + (1 - (1 - \gamma)dt)\left(V(s_t) + \frac{\partial V}{\partial s}(s_t) (\omega J - \theta) (s_t - \mu) dt + \frac{\sigma^2}{2}\Delta V(s_t) dt\right)\nonumber\\
+	(1 - \gamma) V(s_t) &= \BE\left[\tilde{r}(s_t)\right] + \frac{\partial V}{\partial s}(s_t) (\omega J - \theta) (s_t - \mu) + \frac{\sigma^2}{2}\Delta V(s_t).
+\end{align}
+This partial derivative equation is not directly solvable, but can be efficiently approximated
+using finite element methods.
+
+Important remarks:
+\begin{itemize}
+	\item Sampling i.i.d. from the trajectory seems to be extremely
+		important. Estimating gradients using consecutive samples fails
+		miserably (with NaNs early on, in setups where i.i.d. samples
+		lead to convergence).
+	\item Same goes for reward rescaling for high $\gamma$'s, an incorrect scaling
+		will lead to quick divergence. TODO: test if computing mean on the fly
+		suffices. TODO: Test if normalizing outputs suffices.
+\end{itemize}
+% }}}

### 2018-09-28 exp/data.py

@@ -27,7 +27,7 @@ def generate_trajectories(
     for t in range(T):
         noises[:, t] = noise.x
         noise.step()
-    return noises
+    return noises, noise
 
 if __name__ == '__main__':
     import matplotlib.pyplot as plt

### 2018-09-28 exp/femvfunction.py

@@ -1,57 +1,72 @@
+# pylint: disable=too-many-locals, too-many-arguments
 """ Attempt at FEMing bellman. """
+from typing import Tuple
 import numpy as np
 from skfem import MeshTri, InteriorBasis
 from skfem import ElementTriP1
 from skfem import bilinear_form, linear_form
 from skfem import asm, solve, condense
 
-m = MeshTri()
-scale = 10
-m.scale(scale)
-m.translate([-scale / 2, -scale / 2])
-m.refine(6)
+def solve_exact(sigma: float, theta: float, omega: float,
+                beta: float, gamma: float, thresh: float,
+                state_lim: Tuple[float, float]):
+    """ Solve Bellman equation directly using FEM. """
+    m = MeshTri()
+    scale = state_lim[1] - state_lim[0]
+    m.scale(scale)
+    m.translate([state_lim[0], state_lim[0]])
+    m.refine(6)
 
-e = ElementTriP1()
-basis = InteriorBasis(m, e)
+    e = ElementTriP1()
+    basis = InteriorBasis(m, e)
+    mu = 0
 
-sigma = .3
-theta = .01
-mu = 0
-omega = .1
-beta = 30
-gamma = .01
-reward_state_threshold = 2
-J = np.array(
-    [[0, 1],
-     [-1, 0]])
+    J = np.array(
+        [[0, -1],
+         [1, 0]])
 
-@bilinear_form
-def bellman_form(u, du, v, dv, w):
-    """ Laplace operator. """
-    laplace = sigma ** 2 / 2 * sum(du * dv)
-    shape = w.x.shape
-    state_trans = theta * (w.x - mu) + omega * (J @ w.x.reshape(2, -1)).reshape(*shape)
-    next_state_term = - sum(du * state_trans) * v
-    state_term = - (1 - gamma) * u * v
-    return -laplace + next_state_term + state_term
+    transition_matrix = omega * J - theta * np.eye(2)
+
+    @bilinear_form
+    def bellman_form(u, du, v, dv, w):
+        """ Laplace operator. """
+        laplace = -sigma ** 2 / 2 * sum(du * dv)
+        shape = w.x.shape
+        state_trans = (transition_matrix @ (w.x.reshape(2, -1) - mu)).reshape(shape)
+        next_state_term = sum(du * state_trans) * v
+        state_term = (1 - gamma) * u * v
[…]

### 2018-09-28 exp/model.py

@@ -8,10 +8,10 @@ class Model(nn.Module):
         nb_layers = 2
         self.core = nn.Sequential(*([
             nn.Linear(nb_input, nb_hidden),
-            nn.ReLU()
+            nn.ELU()
         ] + [
             nn.Linear(nb_hidden, nb_hidden),
-            nn.ReLU()
+            nn.ELU()
         ] * nb_layers + [
             nn.Linear(nb_hidden, 1)
         ]))

### 2018-09-28 exp/noise.py

@@ -39,28 +39,28 @@ class OrnsteinUlhenbeck(object): # pylint: disable=too-few-public-methods
     def stationary(self, samples: int) -> np.ndarray:
         """ Samples from the stationary distribution. """
         omega_coef = 0 if self.omega is None else self.omega
-        return (self.sigma / (2 * np.sqrt(self.theta ** 2 + omega_coef ** 2)) *
+        return (self.sigma / np.sqrt(2 * np.sqrt(self.theta ** 2 + omega_coef ** 2)) *
                 np.random.randn(samples, self.x.shape[-1]))
 
 
 if __name__ == "__main__":
     import matplotlib.pyplot as plt
+    np.random.seed(0)
 
-    plt.ion()
-    T = 100000
+    T = 10000
     dt = .001
     mu = np.array([0., 0.])
     sigma = .3
-    omega = .3
+    omega = .5
     x_0 = np.array([0., 0.])
 
     fig = plt.figure()
     ax = fig.add_subplot(111)
-    ax.set_xlim([-10, 10])
-    ax.set_ylim([-10, 10])
-    pline, = ax.plot([], [], alpha=.3)
+    ax.set_xlim([-2, 1])
+    ax.set_ylim([-3, 2])
+    pline, = ax.plot([], [], alpha=.7)
 
-    noise = OrnsteinUlhenbeck(dt, mu, sigma, x_0, omega, theta=1e-3)
+    noise = OrnsteinUlhenbeck(dt, mu, sigma, x_0, omega, theta=.03)
     line: Tuple[List[float], List[float]] = ([], [])
     for t in range(T):
         line[0].append(noise.x[0])
@@ -68,5 +68,4 @@ if __name__ == "__main__":
         noise.step()
         if t % int(1 / dt) == 0:
             pline.set_data(*line)
-            plt.show()
-            plt.pause(.01)
+    plt.savefig('../imgs/trajectory.pdf')

### 2018-09-28 exp/vfunction.py

@@ -6,7 +6,9 @@ import numpy as np
 import torch
 import torch.nn.functional as f
 from data import generate_trajectories, OrnsteinUlhenbeckParameters
+from noise import OrnsteinUlhenbeck
 from model import Model
+from femvfunction import solve_exact
 import matplotlib.pyplot as plt
 
 def to_tensor(x: np.ndarray, device) -> torch.Tensor:
@@ -15,6 +17,7 @@ def to_tensor(x: np.ndarray, device) -> torch.Tensor:
 
 def train(model: Model,
           optimizer: torch.optim.Optimizer,
+          scheduler: torch.optim.lr_scheduler.LambdaLR,
           data: np.ndarray,
           nb_steps: int,
           begin_index: int,
@@ -22,55 +25,86 @@ def train(model: Model,
           beta: float,
           dt: float,
           thresh: float,
+          ou_process: OrnsteinUlhenbeck,
           device) -> List[float]:
     """ Train. """
     losses = []
     for i in range(nb_steps):
-        np_next_state = data[:, begin_index + i + 1]
+        # index = begin_index + i
+        index = np.random.randint(0, data.shape[1] - 1)
+        np_next_state = data[:, index + 1]
         reward = to_tensor(reward_function(np_next_state, beta, dt, thresh), device)
-        state = to_tensor(data[:, begin_index + i], device)
+        state = to_tensor(data[:, index], device)
         next_state = to_tensor(np_next_state, device)
-        next_next_state = to_tensor(data[:, begin_index + i + 2], device)
+        # you need the copy, otherwise you modify data
+        ou_process.reset(data[:, index].copy())
+        fake_next_state = to_tensor(ou_process.step(), device)
 
         V = model(state).squeeze()
         V_next = model(next_state).squeeze()
-        V_next_next = model(next_next_state).squeeze()
-        expected_V = reward + true_gamma * V_next
-        hessian_loss = f.mse_loss((V_next_next - V_next).detach(), V_next - V) / (dt ** 4)
-        loss = f.mse_loss(V, expected_V.detach()) / (dt ** 2) # + 0 * hessian_loss
+        V_fake_next = model(fake_next_state).squeeze()
+        if True:
+            V_mixed_next = V_next.detach()
+        else:
[…]

### 2018-09-28 exp/vfunction.py
Commit

@@ -6,7 +6,6 @@ import numpy as np
 import torch
 import torch.nn.functional as f
 from data import generate_trajectories, OrnsteinUlhenbeckParameters
-from noise import OrnsteinUlhenbeck
 from model import Model
 from femvfunction import solve_exact
 import matplotlib.pyplot as plt
@@ -25,30 +24,23 @@ def train(model: Model,
           beta: float,
           dt: float,
           thresh: float,
-          ou_process: OrnsteinUlhenbeck,
           device) -> List[float]:
     """ Train. """
     losses = []
     for i in range(nb_steps):
-        # index = begin_index + i
-        index = np.random.randint(0, data.shape[1] - 1)
+        index = begin_index + i
+        # index = np.random.randint(0, data.shape[1] - 1)
         np_next_state = data[:, index + 1]
         reward = to_tensor(reward_function(np_next_state, beta, dt, thresh), device)
         state = to_tensor(data[:, index], device)
         next_state = to_tensor(np_next_state, device)
         # you need the copy, otherwise you modify data
-        ou_process.reset(data[:, index].copy())
-        fake_next_state = to_tensor(ou_process.step(), device)
 
         V = model(state).squeeze()
         V_next = model(next_state).squeeze()
-        V_fake_next = model(fake_next_state).squeeze()
-        if True:
-            V_mixed_next = V_next.detach()
-        else:
-            V_mixed_next = V_fake_next - V_fake_next.detach() + V_next.detach()
-        
-        loss = f.mse_loss(V - true_gamma * V_mixed_next, reward) / dt # + 0 * hessian_loss
+        V_mixed_next = V_next.detach()
+
+        loss = f.mse_loss(V - true_gamma * V_mixed_next, reward) # + 0 * hessian_loss
 
         optimizer.zero_grad()
         loss.backward()
@@ -120,7 +112,7 @@ def run(batch_size: int,
     device = torch.device('cuda' if torch.cuda.is_available() else 'cpu')
     dt = ou_params.dt
 
-    data, ou_process = generate_trajectories(
+    data, _ = generate_trajectories(
         batch_size=batch_size, T=T,
         w=w, ornstein_ulhenbeck_params=OrnsteinUlhenbeckParameters(dt, sigma, theta))
     print(f"Maximum and minimum values achieved in data: {data.max()}, {data.min()}")
[…]

### 2018-10-18 pendulum_exp/evaluation.py
Normalization change

@@ -5,7 +5,7 @@ from abstract import Policy, Env
 from policy import AdvantagePolicy
 from envs.vecenv import SubprocVecEnv
 from envs.pusher import PusherEnv
-from gym.envs.classic_control import PendulumEnv
+from envs.utils import WrapPendulum
 
 def specific_evaluation(
         epoch: int,
@@ -26,7 +26,8 @@ def specific_evaluation(
         plt.clf()
         plt.plot(state_space, vs)
         plt.pause(.1)
-    elif isinstance(env.envs[0], PendulumEnv):
+    elif isinstance(env.envs[0], WrapPendulum):
+        plt.figure()
         nb_pixels = 50
         theta_space = np.linspace(-np.pi, np.pi, nb_pixels)
         dtheta_space = np.linspace(-10, 10, nb_pixels)

### 2018-10-18 pendulum_exp/main.py

@@ -88,7 +88,7 @@ def main(
     obs = env.reset()
 
     def lr_decay(t):
-        return 1 / np.power(1 + dt * t, 1/2)
+        return 1 / np.power(1 + dt * t, 1/4)
 
     def noise_decay(_):
         return 1

### 2018-10-18 paper/main.tex

\documentclass[11pt]{article}

\usepackage{amsmath,amssymb}

\newcommand{\deltat}{{\delta t}}
\newcommand{\gauss}{\mathcal{N}}
\newcommand{\E}{\mathbb{E}}

\title{Scaling continuous reinforcement learning up}
\begin{document}
\maketitle
\begin{abstract}
	Most reinforcement learning algorithms are not resilient to a change of
	time discretization. Even worse, their time continuous limit is often
	ill defined.  While continuous reinforcement methods have been
	introduced in the past, they have been dismissed when scaling up from
	reinforcement learning to deep reinforcement learning. We introduce the
	first deep reinforcement learning that scales when faced with
	diminishing discretization steps.
\end{abstract}
\paragraph{Problem at hand}
Assume a stochastic continuous environment of the form
\begin{align}
	s_{t+\deltat} &= s_t + F(s_t, a_t) \deltat + \gauss(0, C(s_t, a_t)) \sqrt{\deltat}\\
	r(s_t, a_t) &= \tilde{r}(s_t, a_t) \deltat
\end{align}
with $F$ a deterministic drift, and $C$ a noise covariance matrix
associated to each $(s, a)$ pair. Our aim is to find, a policy $\pi$, that is a
(potentially stochastic) mapping from states to actions that maximizes the
average return, i.e. the expected average of rewards
\begin{align}
	R^\pi &= \lim\limits_{T\to+\infty}
	\frac{1}{T}
	\E_{s_0, a_0, \ldots, s_T, a_T\sim\pi}\left[
		\sum\limits_{t = 0}^T \tilde{r}(s_t, a_t)
	\right]
\end{align}
or the discounted sum of future rewards
\begin{align}
	R^\pi_\gamma &= \E_{s_0, a_0, \ldots, s_T, a_T \sim \pi}
	\left[
		\sum\limits_{t=0}^{+\infty}
		(\gamma^\deltat)^t \tilde{r}(s_t, a_t) 
	\right].
\end{align}
Further assume that for any policy, the environment admits an invariant
measure, that is that there exists a distribution on states $\rho^\pi$ such that for
any regular function $F$ and any infinite sequence $s_0, \ldots$ sampled using policy
$\pi$,
\begin{equation}
	\lim\limits_{T\to+\infty}
	\frac{1}{T}
	\sum\limits_{t=0}^T F(s_t) =
	\E_{s\sim \rho^\pi}\left[
		F(s)
	\right].
\end{equation}
\paragraph{Advantage updating}
\end{document}

### 2018-10-19 paper/main.tex
Commit

@@
 (potentially stochastic) mapping from states to actions that maximizes the
-average return, i.e. the expected average of rewards
+average return, i.e.\ the expected average of rewards
 \begin{align}
@@
 \end{align}
-or the discounted sum of future rewards
+or the discounted sum of future rewards for a physical discount factor of $\gamma$\footnote{
+	Having a physical discount factor of $\gamma$ induces a practical discretized discount
+	factors of $\gamma^\deltat$. Indeed, \eqref{eq:discounted} is the discretization of
+	\begin{equation}
+		R^\pi_\gamma = \int_{0}^{+\infty} \gamma^{t} r(t) dt.
+		\label{eq:discounted}
+	\end{equation}
+}
 \begin{align}
@@
 \end{align}
-Further assume that for any policy, the environment admits an invariant
+We have the following desiderata for our algorithms:
+\begin{itemize}
+	\item To be resilient to a change in discretization. Precisely, this
+		means that increasing the sampling rate should not decrease the
+		performances of the algorithm. Notably, this means that if we consider
+		two discretization steps $\deltat$ and $\deltat'$, and two associated
+		runs of the learning algorithm, the first run should learn in $\frac{1}{\deltat}$
+		samples approximately the same thing as the second in $\frac{1}{\deltat'}$ samples.
+	\item To be able to handle $\gamma$ coeficients arbitrarily close to $1$.
+		Such a condition turns out to be necessary if we require to
+		be able to handle constant physical discount factors.
+	\item To be offline, i.e.\ to learn a policy that is different from
+		the policy used for sampling trajectories. This is useful in
+		two ways. First it decorrelates the problem of learning a good
+		policy from the problem of exploring the world. Second it
+		allows for learning from a fixed dataset of trajectories, or more
+		precisely, in the fully observable case, from a dataset of
+		states, actions, rewards, next states quadruplets.
+\end{itemize}
+
+\paragraph{Assumptions and notations}
+We assume that for any policy, the environment admits an invariant
 measure, that is that there exists a distribution on states $\rho^\pi$ such that for
@@
 \end{equation}
-\paragraph{Advantage updating}
+\paragraph{Centralized advantage updating}
+The optimal centralized advantage function and the optimal centralized value
+function verify the following Bellman equation
+\begin{align}
+	A^*(s, a) &= \max\limits_{a} A^*(s, a) + \frac{
+		(\tilde{r}(s, a) - \E^*\left[
+			\tilde{r}
+		\right]
+		)\deltat + \gamma \E_s'\left[
+			V^*(s')
+		\right] - V^*(s)
+		}{
+		\deltat
+	}\\
+	\max\limits_{a} A^*(s, a) &= 0
+\end{align}
+
+\paragraph{Approximate advantage updating and learning rates scalings}
 \end{document}

### 2018-10-19 paper/main.tex
Paper

@@
 	\E_{s_0, a_0, \ldots, s_T, a_T\sim\pi}\left[
-		\sum\limits_{t = 0}^T \tilde{r}(s_t, a_t)
+		\sum\limits_{t = 0}^T \tilde{r}(s_t, a_t)\deltat
 	\right]
@@
 		\sum\limits_{t=0}^{+\infty}
-		(\gamma^\deltat)^t \tilde{r}(s_t, a_t) 
+		(\gamma^\deltat)^t \tilde{r}(s_t, a_t)\deltat 
 	\right].
@@
 \paragraph{Assumptions and notations}
+Value, value-state and advantage functions as well as there
+optimal counterparts are defined as usual, e.g.
+\begin{align}
+	V^\pi(s) &= \E_{a_0, s_1, \ldots \sim \pi}
+	\left[
+		\sum\limits_{t=0}^{+\infty}
+		(\gamma^\deltat)^t \tilde{r}(s_t, a_t) \deltat
+		\mid s=s_0
+	\right]\\
+	Q^\pi(s, a) &= \E_{a_1, s_1\ldots \sim \pi}
+	\left[
+		\sum\limits_{t=0}^{+\infty}
+		(\gamma^\deltat)^t \tilde{r}(s_t, a_t) \deltat
+		\mid s_0=s, a_0=a
+	\right]
+\end{align}
+and verify Bellman equations, e.g.
+\begin{align}
+	V^\pi(s) = \E_{a\sim\pi(\cdot\mid s)}
+	\left[
+		\tilde{r}(s, a) \deltat +
+		\gamma^\deltat \E_{s'\mid a} V(s')
+	\right]
+\end{align}
 We assume that for any policy, the environment admits an invariant
@@
 function verify the following Bellman equation
-\begin{align}
-	A^*(s, a) &= \max\limits_{a} A^*(s, a) + \frac{
+\begin{gather}
+	A^*(s, a) = \max\limits_{a'} A^*(s, a') + 
 		(\tilde{r}(s, a) - \E^*\left[
@@
 		\right]
-		)\deltat + \gamma \E_s'\left[
+		) + 
+		\frac{\gamma \E_{s'\mid a}\left[
 			V^*(s')
@@
 	}\\
-	\max\limits_{a} A^*(s, a) &= 0
-\end{align}
+	\max\limits_{a} A^*(s, a) = 0
+\end{gather}
 

### 2018-10-19 paper/main.tex
Commit

@@
 \usepackage{amsmath,amssymb}
+\usepackage{xcolor}
 
@@
 \newcommand{\E}{\mathbb{E}}
+\newcommand{\TODO}[1]{\textcolor{red}{#1}}
+\newcommand{\bigO}[1]{O(#1)}
 
@@
 		\mid s_0=s, a_0=a
-	\right]
+	\right]\\
+	A^\pi(s,a ) &= Q^\pi(s, a) - V^\pi(s)
 \end{align}
@@
 		\gamma^\deltat \E_{s'\mid a} V(s')
-	\right]
+	\right].
 \end{align}
@@
 \end{equation}
+
+\paragraph{The Q-function action independant when $\deltat \to 0$}
+Q-learning, which uses the optimal Bellman equation on the state value function
+to approximate the optimal Q-function is a natural candidate for off-policy
+continuous time learning. However, as stated in \TODO{refs}, the state value function
+becomes independant of its action parameter as the discretization step decreases towards
+$0$. Indeed, approximating to first order in $\deltat$ in the Bellman equation defining
+$Q^\pi$, 
+\begin{align}
+	Q^\pi(s, a) &= \tilde{r}(s, a)\deltat + \gamma^\deltat \E_{s_{t+\deltat} \mid a, s} 
+	V^\pi(s_{t+\deltat})\nonumber\\
+	&= \tilde{r}(s, a)\deltat + \gamma^\deltat V^\pi(s) + \gamma^\deltat \partial_s V(s) F(s, a) \deltat
+	+ \frac{\gamma^\deltat}{2} \text{tr}\left(C(s, a)\partial^2_{s} V(s)\right) \deltat\nonumber\\
+	&\to\limits_{\deltat \to 0} V^\pi(s).
+\end{align}
+
+Consequently Q-learning is extremely sensitive both to noise and to approximation errors.
+To extract the optimal action from an estimate of the optimal Q-function, you need an estimate
+whose approximation error is of order $\bigO{\deltat}$.
+Typically, when learning the state-value function only from monte-carlo trajectory samples, 
+an optimistic number of timesteps required to learn an estimate with precision $\bigO{\deltat}$ is
+$\bigO{\frac{1}{\deltat^2}}$.\footnote{
+	Monte-carlo estimates of the Q function have typical variance of order
+	$\frac{1}{\deltat\sqrt{2(1 - \gamma)}}$. To obtain an estimate with precision $1$ of the mean
+	of a quantity with variance $\sigma^2$ by averaging independant samples, you
+	require at least $\sigma^2$ samples.}
+This notably means that when $\deltat \to 0$, you would require an infinite number of samples
+per infinitesimal timestep to actually learn correct actions.
+
+\paragraph{The average reward dominates when $\gamma$ gets close to $1$}
+\TODO{Make this formal and clear.}\\
+When the time horizon is unbounded, $V^\pi(s)$ can be rewritten as
+\begin{align}
+	V^\pi(s) = \frac{\E_{s\sim\rho^\pi}\tilde{r}}{1 - \gamma} +
+	\sigma^\pi(s) + \varepsilon^\pi(s, \gamma)
+\end{align}
+with $\lim\limits_{\gamma\to 1} = 0$. This means that when $\gamma$ gets close to $1$,
+if $\rho^\pi(s)$ is non-zero, the 
 \paragraph{Centralized advantage updating}

### 2018-10-20 paper/main.tex
Changes

@@
 \newcommand{\bigO}[1]{O(#1)}
+\newcommand{\Tdo}{TD($0$)}
 
@@
 	\right].
+	\label{eq:ergodic}
 \end{equation}
@@
 whose approximation error is of order $\bigO{\deltat}$.
-Typically, when learning the state-value function only from monte-carlo trajectory samples, 
+Typically, when learning the state-action value function only from monte-carlo trajectory samples, 
 an optimistic number of timesteps required to learn an estimate with precision $\bigO{\deltat}$ is
-$\bigO{\frac{1}{\deltat^2}}$.\footnote{
+$\bigO{\frac{1}{\deltat^3}}$.\footnote{
 	Monte-carlo estimates of the Q function have typical variance of order
@@
 This notably means that when $\deltat \to 0$, you would require an infinite number of samples
-per infinitesimal timestep to actually learn correct actions.
+per infinitesimal timestep to actually learn correct actions. A more formal statement in a simple case
+is given in appendix.
 
@@
 \TODO{Make this formal and clear.}\\
-When the time horizon is unbounded, $V^\pi(s)$ can be rewritten as
+Under the ergodic condition eq.~\eqref{eq:ergodic}, $V^\pi(s)$ can be rewritten as
 \begin{align}
@@
 	\sigma^\pi(s) + \varepsilon^\pi(s, \gamma)
+	\label{eq:dominate}
 \end{align}
-with $\lim\limits_{\gamma\to 1} = 0$. This means that when $\gamma$ gets close to $1$,
-if $\rho^\pi(s)$ is non-zero, the 
+with $\lim\limits_{\gamma\to 1}\varepsilon^\pi(s, \gamma) = 0$. This means that
+when $\gamma$ gets close to $1$, if the average reward under the stationary
+policy is non zero, the value function, as well as the state-action value
+function, are dominated by a constant term, independant of both the state and
+the action. When $\gamma$ is exactly equal to $1$, the constant term becomes
+infinite, making all standard algorithms that direcetly learn the value
+function or the state-action value function impractical.
+
+Substracting arbitrary constants to all rewards does not change the relative
+performances of policies, and consequently does not change the optimal
+policy.\footnote{
+	This is only true for non episodic MDPs. For episodic MDPs, substracting
+	a constant factor to all rewards only leaves the relative performance of
+	policies unchanged if the agent if properly penalized (or rewarded) when
+	reaching a terminal state. More details are given in
+	\TODO{Probably in the appendix}.
+} If one is able to estimate the average reward of the current policy, substracting
+the average reward to all rewards negates the first term in eq.~\eqref{eq:dominate}.
+The average reward of the current policy is rarely readily available. When learning
+on-policy, the average reward can be estimated using running averages
+over trajectories\footnote{One must be aware that you cannot use TD(0) on the 
+	same trajectory as the one on which you are learning the average reward.
+	\TODO{Explain in more details: this creates a dependency of the average
+	reward on the current state, which biased learning.}
+}. When learning off policy, the average reward cannot be directly learnt from reward
+samples along the trajectories.
+
 \paragraph{Centralized advantage updating}
+The algorithm introduced hereby attends to both problematics in the following ways.
+It uses the advantage updating formulation introduced in~\TODO{ref} to learn
+in an offline fashion the advantage function. Contrary to the state-action value
+function, the difference of advantage between two actions does not vanish as the
+discretization step decreases. Combined with correctly scaled learning rates, this
+algorithm proves to be both efficient for high discretization steps, and resilient
+to wide changes in the discretization step, even for complex function approximators.
 The optimal centralized advantage function and the optimal centralized value
-function verify the following Bellman equation
+Second, by introducing a proper baseline parameter and adding a normalization term
+to the loss, the algorithm provide resilience to gamma values close to $1$, and
+equals to $1$ when eq.~\eqref{eq:ergodic} is verified.
+\subparagraph{Advantage updating}
+Advantage updating, as introduced in~\TODO{ref} relies on the fact that the advantage
+function and value function are the only functions that verify
 \begin{gather}
 	A^*(s, a) = \max\limits_{a'} A^*(s, a') + 
-		(\tilde{r}(s, a) - \E^*\left[
-			\tilde{r}
-		\right]
-		) + 
-		\frac{\gamma \E_{s'\mid a}\left[
+		\tilde{r}(s, a)
+		 + 
+		\frac{\gamma^\deltat \E_{s'\mid a}\left[
 			V^*(s')
@@
 	}\\
-	\max\limits_{a} A^*(s, a) = 0
+	\max\limits_{a} A^*(s, a) = 0.
 \end{gather}
+These equations can be used to learn approximations of both $A^*$ and $V^*$,
+$A_\theta$ and $V_\phi$. To that end, define the following residuals
+\begin{gather}
+	\delta^A_{\theta, \phi}(s, a) = A_\theta(s, a) -
+	\max\limits_{a'} A_\theta(s, a') + 
+	\tilde{r}(s, a)
+	+ 
+	\frac{\gamma^\deltat \E_{s'\mid a}\left[
+			V_\phi(s')
+		\right] - V_\phi(s)
+		}{
+		\deltat
+	}\\
+	\delta^\text{norm}_\theta(s, a) = \max\limits_{a} A_\theta(s, a).
+\end{gather}
+The mean squared residuals
+\begin{align}
+	\E_{s \sim \rho^\pi, a \sim \pi}\left[\delta^A_{\theta, \phi}(s, a)^2\right]\\
+	\E_{s \sim \rho^\pi, a \sim \pi}\left[\delta^\text{norm}_{\theta}(s, a)^2\right]
+\end{align}
+admit $V^*$ and $A^*$ as sole global minimizers.
 
-\paragraph{Approximate advantage updating and learning rates scalings}
+Estimates of the gradients of the first residual with respect to $\phi$ computed only
+from trajectories are not readily availables. Indeed, this would require computing two
+next states from each state, and this in turn requires a model of the world. Instead,
+the approximate \Tdo trick can be used, yielding gradient estimates of the form
+\begin{equation}
+	g^\phi_{\delta^A_{\theta, \phi}} =
+	\frac{1}{\deltat}
+	\partial_\phi V_\phi(s) \left(A_\theta(s, a) -
+		\max\limits_{a'} A_\theta(s, a') + 
+		\tilde{r}(s, a)
+		+ 
+		\frac{\gamma^\deltat 
+			V_\phi(s')
+			- V_\phi(s)
+			}{
+			\deltat
+	}\right).
+\end{equation}
+
+\subparagraph{Baseline centralization}
+As previously mentionned, using a non centralized reward may hurt learning, as the
+value function is bound to learn a very large constant which is not useful in the
+action selection process.
+
+To centralize rewards, we introduce a constant baseline $b$ the learning process.
+Consider the following set of equations on $A^*$, $V^*$ and $b$
+\begin{gather}
+	A^*(s, a) = \max\limits_{a'} A^*(s, a') + 
+		(\tilde{r}(s, a) - b)
+		 + 
+		\frac{\gamma^\deltat \E_{s'\mid a}\left[
+			V^*(s')
+		\right] - V^*(s)
+		}{
+		\deltat
+	}\\
+	\max\limits_{a} A^*(s, a) = 0\\
+	\E_{s\sim \rho^\pi}\left[V^*(s)\right] = 0
+	.
+\end{gather}
+It is unclear at first wether this system of equations admit a solution.
+Consider the two first equations. As we only substract a constant $b$ to all
+rewards, the only $V^*$ that verifies those equations takes the form
+\begin{equation}
+	V^*(s) = \frac{\E_{s\sim\rho^*}\tilde{r} - b}{1 - \gamma} +
+	\sigma^*(s) + \varepsilon^*(s, \gamma).
+\end{equation}
+Averaging under $\rho^\pi$, the following is obtained
+\begin{equation}
+	b = \E_{s\sim\rho^*}\tilde{r} +
+	(1 - \gamma) \E_{s\sim\rho^\pi}\left[\sigma^*(s) +
+		\varepsilon^*(s, \gamma) +
+		V^*(s)
+	\right].
+\end{equation}
+Consequently, for the third equation to hold, $b$ takes the value
+\begin{equation}
+	b = \E_{s\sim\rho^*}\tilde{r} +
+	(1 - \gamma) \E_{s\sim\rho^\pi}\left[\sigma^*(s) +
+		\varepsilon^*(s, \gamma)
+	\right]
+\end{equation}
+and notably, as $\gamma$ gets closer to $1$, $b$ gets closer to the
+average reward under the optimal policy.
+
+\TODO{More on that later}
+To approximately verify the three equations, the corresponding residuals are
+minimized over $b$, $\phi$ and $\theta$. Caution should be taken to minimize
+the third residual, as the gradient requires sampling of two different states.
+
+\paragraph{Learning rates scalings}
+\TODO{More on that later. In noisy setups we may need $\frac{1}{\deltat^\frac{3}{2}}$ for the last term of the loss.}
+Actual loss we minimize on both $\theta$ and $\phi$
+\begin{equation}
+	\E_{s \sim \rho^\pi, a \sim \pi}\left[\delta^A_{\theta, \phi}(s, a)^2\right] +
+	\E_{s \sim \rho^\pi, a \sim \pi}\left[\delta^\text{norm}_{\theta}(s, a)^2\right] +
+	+ \frac{1}{\deltat}\E_{s \sim \rho^\pi}\left[\delta^\text{mean}_{\theta}(s)^2\right]
+\end{equation}
+Learning rate on $\phi$ and $b$ are of order $\deltat$, learning rate on $\theta$ is of order $\deltat^2$. This is to have gradients moving of approximately $\sqrt{dt}
+
 \end{document}

### 2018-10-20 pendulum_exp/evaluation.py
Changing exact scalings

@@ -27,7 +27,6 @@ def specific_evaluation(
         plt.plot(state_space, vs)
         plt.pause(.1)
     elif isinstance(env.envs[0], WrapPendulum):
-        plt.figure()
         nb_pixels = 50
         theta_space = np.linspace(-np.pi, np.pi, nb_pixels)
         dtheta_space = np.linspace(-10, 10, nb_pixels)

### 2018-10-20 pendulum_exp/main.py

@@ -88,7 +88,7 @@ def main(
     obs = env.reset()
 
     def lr_decay(t):
-        return 1 / np.power(1 + dt * t, 1/4)
+        return 1 / np.power(1 + dt * t, 1/2)
 
     def noise_decay(_):
         return 1

### 2018-10-20 pendulum_exp/policy.py

@@ -107,7 +107,7 @@ class AdvantagePolicy(Policy):
 
             self._optimizers[0].zero_grad()
             self._optimizers[1].zero_grad()
-            (mean_loss / self._dt + loss).backward()
+            (mean_loss / np.power(self._dt, 3/2) + loss).backward()
             self._optimizers[0].step()
             self._schedulers[0].step()
             self._optimizers[1].step()

### 2018-10-20 paper/main.tex
Pap

@@
 \paragraph{Learning rates scalings}
-\TODO{More on that later. In noisy setups we may need $\frac{1}{\deltat^\frac{3}{2}}$ for the last term of the loss.}
+\TODO{More on that later. I am not sure if we should have a $\frac{1}{\deltat}$
+	or a $\frac{1}{\deltat^\frac{3}{2}}$ in front of the mean normalization
+term. My bet is $\frac{1}{\deltat^\frac{3}{2}}$}.
 Actual loss we minimize on both $\theta$ and $\phi$
@@
 	\E_{s \sim \rho^\pi, a \sim \pi}\left[\delta^\text{norm}_{\theta}(s, a)^2\right] +
-	+ \frac{1}{\deltat}\E_{s \sim \rho^\pi}\left[\delta^\text{mean}_{\theta}(s)^2\right]
+	+ \frac{1}{\deltat^\frac{3}{2}}\E_{s \sim \rho^\pi}\left[\delta^\text{mean}_{\theta}(s)^2\right].
 \end{equation}
-Learning rate on $\phi$ and $b$ are of order $\deltat$, learning rate on $\theta$ is of order $\deltat^2$. This is to have gradients moving of approximately $\sqrt{dt}
+Learning rates on $\phi$ and $b$ are of order $\deltat$, learning rate on $\theta$ is of order $\deltat^2$. This is to have parameters
+moving by a distance of order $\sqrt{\deltat}$ per iteration.
 

### 2018-10-20 pendulum_exp/policy.py
Reverting problem

@@ -107,7 +107,7 @@ class AdvantagePolicy(Policy):
 
             self._optimizers[0].zero_grad()
             self._optimizers[1].zero_grad()
-            (mean_loss / np.power(self._dt, 3/2) + loss).backward()
+            (mean_loss / self._dt + loss).backward()
             self._optimizers[0].step()
             self._schedulers[0].step()
             self._optimizers[1].step()

### 2018-10-22 paper/main.tex
Pap
Commit
continuous expressions, redefinition of advantage function

@@
 
-\usepackage{amsmath,amssymb}
+\usepackage[utf8]{inputenc}
+\usepackage[english]{babel}
+
 \usepackage{xcolor}
+\usepackage{amsfonts}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{amsthm}
 
@@
 with $F$ a deterministic drift, and $C$ a noise covariance matrix
-associated to each $(s, a)$ pair. Our aim is to find, a policy $\pi$, that is a
+associated to each $(s, a)$ pair. For a discretization $\deltat$, we use the notation $t = k \delta t$. An action is performed every $\deltat$ time step.
+Our aim is to find, a policy $\pi$, that is a
 (potentially stochastic) mapping from states to actions that maximizes the
@@
 \begin{align}
-	R^\pi &= \lim\limits_{T\to+\infty}
-	\frac{1}{T}
-	\E_{s_0, a_0, \ldots, s_T, a_T\sim\pi}\left[
-		\sum\limits_{t = 0}^T \tilde{r}(s_t, a_t)\deltat
+	R^\pi &= \lim\limits_{K\to+\infty}
+	\E_{s_0, a_0, \ldots, s_{\lfloor T/\deltat\rfloor\deltat}, a_{\lfloor T/\deltat\rfloor\deltat\sim\pi}}\left[
+		\frac{1}{K} \sum\limits_{k = 0}^K \tilde{r}(s_{k\deltat}, a_{k\deltat})
+                \right] \\
+              &= \lim\limits_{T\to+\infty}
+	\E_{s_0, a_0, \ldots, s_{\lfloor T/\deltat\rfloor\deltat}, a_{\lfloor T/\deltat\rfloor\deltat\sim\pi}}\left[
+		\frac{1}{T}\sum\limits_{k = 0}^{\lfloor T/\deltat\rfloor} \tilde{r}(s_{k\deltat}, a_{k\deltat})\deltat
+                \right]\\
+              &= \lim\limits_{T\to+\infty}	
+	\E_{s_0, a_0, \ldots, s_{\lfloor T/\deltat\rfloor\deltat}, a_{\lfloor T/\deltat\rfloor\deltat\sim\pi}}\left[
+		\frac{1}{T}\int\limits_{t = 0}^{T} \tilde{r}(s_{\lfloor t/\deltat\rfloor\deltat}, a_{\lfloor t/\deltat\rfloor\deltat})dt
 	\right]
@@
 \begin{align}
-	R^\pi_\gamma &= \E_{s_0, a_0, \ldots, s_T, a_T \sim \pi}
+	R^\pi_\gamma &= \E_{s_0, a_0, \ldots, s_K, aK \sim \pi}
 	\left[
-		\sum\limits_{t=0}^{+\infty}
-		(\gamma^\deltat)^t \tilde{r}(s_t, a_t)\deltat 
+		\sum\limits_{k=0}^{+\infty}
+		(\gamma^\deltat)^k \tilde{r}(s_k, a_k)\deltat 
+                       \right].\\
+                     &= \E_{s_0, a_0, \ldots, s_K, aK \sim \pi}
+	\left[
+		\int\limits_{t=0}^{+\infty}
+		\gamma^{\lfloor t/\deltat\rfloor \deltat} \tilde{r}(s_{\lfloor t/\deltat\rfloor\deltat}, a_{\lfloor t/\deltat \rfloor\deltat})dt 
 	\right].
@@
 	\left[
-		\sum\limits_{t=0}^{+\infty}
-		(\gamma^\deltat)^t \tilde{r}(s_t, a_t) \deltat
-		\mid s=s_0
+		\sum\limits_{k=0}^{+\infty}
+		(\gamma^{\deltat})^k \tilde{r}(s_{k\deltat}, a_{k\deltat}) \deltat
+		\mid s_0 = s 
                    \right]\\
-	Q^\pi(s, a) &= \E_{a_1, s_1\ldots \sim \pi}
+                 &= \E_{a_0, s_1, \ldots \sim \pi}
 	\left[
-		\sum\limits_{t=0}^{+\infty}
-		(\gamma^\deltat)^t \tilde{r}(s_t, a_t) \deltat
-		\mid s_0=s, a_0=a
+		\int\limits_{t=0}^{+\infty}
+		\gamma^{\lfloor t / \deltat \rfloor \deltat} \tilde{r}(s_{\lfloor t / \deltat \rfloor \deltat}, a_{\lfloor t / \deltat \rfloor \deltat}) dt
+		\mid s_0 = s 
+	\right]\\
+	Q^\pi(s, a) &= \E_{a_0, s_1, \ldots \sim \pi}
+	\left[
+		\sum\limits_{k=0}^{+\infty}
+		(\gamma^{\deltat})^k \tilde{r}(s_{k\deltat}, a_{k\deltat}) \deltat
+		\mid s_0 = s, a_0 = a
+                   \right]\\
+                 &= \E_{a_0, s_1, \ldots \sim \pi}
+	\left[
+		\int\limits_{t=0}^{+\infty}
+		\gamma^{\lfloor t / \deltat \rfloor \deltat} \tilde{r}(s_{\lfloor t / \deltat \rfloor \deltat}, a_{\lfloor t / \deltat \rfloor \deltat}) dt
+		\mid s_0 = s , a_0 = a
 	\right]\\
@@
 	+ \frac{\gamma^\deltat}{2} \text{tr}\left(C(s, a)\partial^2_{s} V(s)\right) \deltat\nonumber\\
-	&\to\limits_{\deltat \to 0} V^\pi(s).
+	&= V^\pi(s) + O_{\deltat \to 0}(\deltat).
 \end{align}
@@
 	Monte-carlo estimates of the Q function have typical variance of order
-	$\frac{1}{\deltat\sqrt{2(1 - \gamma)}}$. To obtain an estimate with precision $1$ of the mean
+	$\frac{1}{\deltat\sqrt{2(1 - \gamma)}}$\TODO{Pourquoi?}. To obtain an estimate with precision $1$ of the mean
 	of a quantity with variance $\sigma^2$ by averaging independant samples, you
@@
 	A^*(s, a) = \max\limits_{a'} A^*(s, a') + 
+		\tilde{r}(s, a)\deltat
+		 + 
+		\gamma^\deltat \E_{s'\mid a}\left[
+			V^*(s')
+		\right] - V^*(s) \\
+	\max\limits_{a} A^*(s, a) = 0.
+      \end{gather}
+Since we know that \TODO{REF EQ} the advantage is of order $O(\deltat)$, we define $\tilde{A} = A/\deltat$, and we have the following
+      \begin{gather}
+	\tilde{A}^*(s, a) = \max\limits_{a'} \tilde{A}^*(s, a') + 
 		\tilde{r}(s, a)
@@
 	}\\
-	\max\limits_{a} A^*(s, a) = 0.
+	\max\limits_{a} \tilde{A}^*(s, a) = 0.
 \end{gather}
-These equations can be used to learn approximations of both $A^*$ and $V^*$,
-$A_\theta$ and $V_\phi$. To that end, define the following residuals
+These equations can be used to learn approximations of both $\tilde{A}^*$ and $V^*$,
+$\tilde{A}_\theta$ and $V_\phi$. To that end, define the following residuals
 \begin{gather}
-	\delta^A_{\theta, \phi}(s, a) = A_\theta(s, a) -
-	\max\limits_{a'} A_\theta(s, a') + 
+	\delta^A_{\theta, \phi}(s, a) = \tilde{A}_\theta(s, a) -
+	\max\limits_{a'} \tilde{A}_\theta(s, a') + 
 	\tilde{r}(s, a)
@@
 	}\\
-	\delta^\text{norm}_\theta(s, a) = \max\limits_{a} A_\theta(s, a).
+	\delta^\text{norm}_\theta(s, a) = \max\limits_{a} \tilde{A}_\theta(s, a).
 \end{gather}
@@
 	\frac{1}{\deltat}
-	\partial_\phi V_\phi(s) \left(A_\theta(s, a) -
-		\max\limits_{a'} A_\theta(s, a') + 
+	\partial_\phi V_\phi(s) \left(\tilde{A}_\theta(s, a) -
+		\max\limits_{a'} \tilde{A}_\theta(s, a') + 
 		\tilde{r}(s, a)
@@
 \begin{gather}
-	A^*(s, a) = \max\limits_{a'} A^*(s, a') + 
+	\tilde{A}^*(s, a) = \max\limits_{a'} \tilde{A}^*(s, a') + 
 		(\tilde{r}(s, a) - b)
@@
 	}\\
-	\max\limits_{a} A^*(s, a) = 0\\
+	\max\limits_{a} \tilde{A}^*(s, a) = 0\\
 	\E_{s\sim \rho^\pi}\left[V^*(s)\right] = 0
@@
 	or a $\frac{1}{\deltat^\frac{3}{2}}$ in front of the mean normalization
-term. My bet is $\frac{1}{\deltat^\frac{3}{2}}$}.
+term. My bet is $\frac{1}{\deltat}$}.
 Actual loss we minimize on both $\theta$ and $\phi$
@@
 	\E_{s \sim \rho^\pi, a \sim \pi}\left[\delta^\text{norm}_{\theta}(s, a)^2\right] +
-	+ \frac{1}{\deltat^\frac{3}{2}}\E_{s \sim \rho^\pi}\left[\delta^\text{mean}_{\theta}(s)^2\right].
+	+ \frac{1}{\deltat}\E_{s \sim \rho^\pi}\left[\delta^\text{mean}_{\theta}(s)^2\right].
 \end{equation}
-Learning rates on $\phi$ and $b$ are of order $\deltat$, learning rate on $\theta$ is of order $\deltat^2$. This is to have parameters
-moving by a distance of order $\sqrt{\deltat}$ per iteration.
+Learning rates on $\phi$ and $b$ are of order $\deltat$, learning rate on
+$\theta$ is of order $\deltat^2$. This is to have parameters moving by a
+distance of order $\deltat$, with at most a $\sqrt{\deltat}$ centered noise per
+iteration.
 

### 2018-10-28 paper/main.tex

@@
 \end{abstract}
+\TODO{
+	\begin{itemize}
+		\item Write paragraph on advantage updating and notably why the equation on
+			$max_a A^*(s, a)$ is only a numerical trick not necessary to properly learn
+			$A^*$.
+		\item $Q$ est difficile à estimer à $\deltat$. $A$ est difficile à estimer en utilisant $Q$ et $V$. Mieux de le faire directement.
+		\item Pour que l'apprentissage en $(V(s') - V(s)) / \deltat$, pas hurler les gens, il faut bien expliquer que c'est d'ordre $\frac{1}{\sqrt{\deltat}}$, qui est "annulable" en $\frac{1}{\deltat}$ samples.
+		\item \'Ecrire algo vite.
+		\item Test algo avec 19 remplacé par sampling de 10 actions et prise de max.
+	\end{itemize}
+}
 \paragraph{Problem at hand}

### 2018-10-28 pendulum_exp/models.py
Learn policy faster than model

@@ -1,7 +1,6 @@
 """Define pytorch models."""
 import torch
 import torch.nn as nn
-import torch.nn.functional as f
 from abstract import ParametricFunction, Tensorable
 from convert import check_tensor
 
@@ -23,7 +22,7 @@ class MLP(nn.Module, ParametricFunction):
 class ContinuousPolicyMLP(MLP, ParametricFunction):
     """MLP with a Tanh on top..."""
     def forward(self, *inputs: Tensorable):
-        return f.tanh(super().forward(*inputs))
+        return torch.tanh(super().forward(*inputs))
 
 class ContinuousAdvantageMLP(MLP, ParametricFunction):
     """MLP with 2 inputs, 1 output."""

### 2018-10-28 pendulum_exp/policies/continuous.py

@@ -32,7 +32,7 @@ class AdvantagePolicy(Policy):
         self._optimizers = (
             torch.optim.SGD(chain(self._adv_function.parameters(), [self._baseline]), lr=lr * dt),
             torch.optim.SGD(self._val_function.parameters(), lr=lr * dt ** 2),
-            torch.optim.SGD(self._policy_function.parameters(), lr=lr * dt))
+            torch.optim.SGD(self._policy_function.parameters(), lr=2 * lr * dt))
         self._schedulers = (
             torch.optim.lr_scheduler.LambdaLR(self._optimizers[0], lr_decay),
             torch.optim.lr_scheduler.LambdaLR(self._optimizers[1], lr_decay),

### 2018-11-08 pendulum_exp/monitoring.py
solve bug log continuous, monitoring

@@ -36,10 +36,9 @@ def summary(logs):
     metrics = [("Avg_adv_loss", min), ("Return", max)]
     for met, f in metrics:
         metval = logs[met]
-        curr_ts, curr_val = max(metval, key=lambda x: x[0])
-        best_ts, best_val = f(metval, key=lambda x: x[1])
-        print("{}\tBest value: {:.2e} at ts {}\t Current value: {:.2e} at ts {}\t".format(\
-            met, best_val, curr_val, best_ts, curr_ts))
+        curr_ts, curr_val = max(metval.items(), key=lambda x: x[0])
+        best_ts, best_val = f(metval.items(), key=lambda x: x[1])
+        print(f"{met}\tBest value: {best_val:.2e} at ts {best_ts}\t Current value: {curr_val:.2e} at ts {curr_ts}\t")
 
 
 if __name__ == '__main__':

### 2018-11-08 pendulum_exp/mylog.py

@@ -1,14 +1,17 @@
 """Log facilities"""
 from os import makedirs
-from os.path import join, exists
+from os.path import join, exists, isfile
 from typing import Dict
 import pickle
 import numpy as np
+
+from logging import info
 
 class Logger:
     CURRENT = None # current logger
 
     def __init__(self):
+        assert Logger.CURRENT is None
         self._logs: Dict[str, Dict[int, float]] = dict()
         self._buffering = 10
         self._count = 0
@@ -32,6 +35,15 @@ class Logger:
 
     def set_dir(self, logdir):
         self._dir = logdir
+        if isfile(join(self._dir, 'logs.pkl')):
+            self.load()
+        info("logfile: {}".format(join(self._dir, 'logs.pkl')))
+
+    def load(self):
+        assert self._file is not None
+        with open(self.filename, 'rb') as f:
+            self._logs = pickle.load(f)
+            
 
     def dump(self):
         assert self._dir is not None
@@ -49,8 +61,6 @@ def log_video(tag: str, timestamp: int, frames):
     assert Logger.CURRENT is not None
     Logger.CURRENT.log_video(tag, timestamp, frames)
 
-def _configure():
+
+if Logger.CURRENT is None:
     Logger.CURRENT = Logger()
-
-
-_configure() # configure logger on import

### 2018-11-08 paper/main.tex

@@
 \newcommand{\Tdo}{TD($0$)}
+\newcommand{\tdres}{\delta_\text{TD}}
 
@@
 
+\paragraph{Resulting update equations}
+The resulting update equations on $\theta$, $\phi$ and $b$, for a transition
+quadruplet $(s, a, s', \tilde{r}(s, a))$ and a state $s''$ independant from the
+quadruplet are
+\begin{align}
+	\tdres &= A_\theta(s, a) - 
+	\max\limits_{a'} A_\theta(s, a')-
+	\tilde{r}(s, a) + b - \frac{\gamma^\deltat V_\phi(s') - V_\phi(s)}{\deltat}\\
+	\delta \theta &= \deltat \left(\alpha_1 \tdres \partial_\theta \left(A_\theta(s, a) - 
+		\max\limits_{a'} A_\theta(s, a')\right)
+	+ \alpha_2 \max\limits_{a'} A_\theta(s, a')\partial_\theta \max\limits_{a'} A_\theta(s, a')\right)\\
+	\delta \phi &= \deltat \left( \alpha_3 \tdres \partial_\phi V_\phi(s) + \alpha_4 V_\phi(s'')\partial_\phi V_\phi(s)\right)\\
+	\delta b &= \deltat \alpha_5 \tdres\\
+	\theta &\leftarrow \theta - \delta \theta\\
+	\phi &\leftarrow \phi - \delta \phi\\
+	b &\leftarrow b - \delta b\\
+\end{align}
+
+\paragraph{Alternate update rule for $b$}
+Averaging advantage updating bellman equation over the stationary distribution of the current policy $\pi$, $b$ can be shown to verify
+\begin{equation}
+	b = \E_{(s, a) \sim \rho^\pi \times \pi}\left[\tilde{r}(s, a) - A^*(s, a)\right]
+\end{equation}
+yielding a different update rule for $b$
+\begin{equation}
+	b \leftarrow (1 - \alpha_5 \deltat) b + \alpha_5 \deltat \left(\tilde{r}(s, a) - A^*(s, a)\right).
+\end{equation}
+
 \end{document}

### 2018-11-08 pendulum_exp/monitoring.py
add print for name plot file

@@ -28,6 +28,7 @@ def plots(logs, namefile):
         ax.set_xlabel("timestamps")
 
     plt.tight_layout()
+    print(f'Plot: {namefile}')
     plt.savefig(namefile, format="eps")
 
 

### 2018-11-08 pendulum_exp/evaluation.py
Seemingly working

@@ -2,8 +2,7 @@
 import numpy as np
 import matplotlib.pyplot as plt
 from abstract import Policy, Env
-from policies import AdvantagePolicy, ContinuousAdvantagePolicy
-from policies.wrappers import StateNormalization
+from policies import ContinuousAdvantagePolicy
 from envs.vecenv import SubprocVecEnv
 from envs.pusher import AbstractPusher, ContinuousPusherEnv
 from gym.envs.classic_control import PendulumEnv
@@ -14,8 +13,6 @@ def specific_evaluation(
         dt: float,
         env: Env,
         policy: Policy):
-    assert isinstance(policy, (AdvantagePolicy, ContinuousAdvantagePolicy, StateNormalization)), f"Incorrect policy type: {type(policy)}, "\
-        "AdvantagePolicy expected."
     assert isinstance(env, SubprocVecEnv), f"Incorrect environment type: {type(env)}, SubprocVecEnv expected." # type: ignore
 
     if isinstance(env.envs[0].unwrapped, AbstractPusher): # type: ignore
@@ -45,9 +42,10 @@ def specific_evaluation(
         dtheta_space = np.linspace(-10, 10, nb_pixels)
         theta, dtheta = np.meshgrid(theta_space, dtheta_space)
         state_space = np.stack([np.cos(theta), np.sin(theta), dtheta], axis=-1)
+        target_shape = state_space.shape[:2]
 
-        vs = policy.value(state_space).squeeze()
-        actions = policy.step(state_space).squeeze()
+        vs = policy.value(state_space.reshape(-1, 3)).reshape(target_shape).squeeze()
+        actions = policy.step(state_space.reshape(-1, 3)).reshape(target_shape).squeeze()
         plt.clf()
         plt.subplot(121)
         plt.imshow(actions, origin='lower')
@@ -55,5 +53,3 @@ def specific_evaluation(
         plt.imshow(vs, origin='lower')
         plt.colorbar()
         plt.pause(.1)
-        if epoch % log == log - 1:
-            plt.savefig(f'logs/results_{dt}_{epoch}.pdf')

### 2018-11-08 pendulum_exp/main.py

@@ -136,15 +136,15 @@ if __name__ == '__main__':
     parser.add_argument('--nb_epochs', type=int, default=50000)
     parser.add_argument('--nb_steps', type=int, default=100)
     parser.add_argument('--sigma_eval', type=float, default=0)
-    parser.add_argument('--sigma', type=float, default=.3)
-    parser.add_argument('--theta', type=float, default=1)
+    parser.add_argument('--sigma', type=float, default=1.5)
+    parser.add_argument('--theta', type=float, default=7.5)
     parser.add_argument('--nb_train_env', type=int, default=32)
     parser.add_argument('--nb_eval_env', type=int, default=16)
     parser.add_argument('--memory_size', type=int, default=10000)
     parser.add_argument('--learn_per_step', type=int, default=50)
     parser.add_argument('--cyclic_exploration', action='store_true')
     parser.add_argument('--normalize_state', action='store_true')
-    parser.add_argument('--lr', type=float, default=.003)
+    parser.add_argument('--lr', type=float, default=.03)
     parser.add_argument('--time_limit', type=float, default=None)
     parser.add_argument('--redirect_stdout', action='store_true')
     parser.add_argument('--policy_type', type=str, default='default')

### 2018-11-08 pendulum_exp/noise.py

@@ -67,7 +67,7 @@ class ParameterNoise(Noise):
             for name, p in function.named_parameters():
                 if 'ln' not in name:
                     p.copy_(p.data + self._p_noise[name])
-            perturbed_output = function(inputs)
+            perturbed_output = function(*inputs)
             for name, p in function.named_parameters():
                 if 'ln' not in name:
                     p.copy_(p.data - self._p_noise[name])

### 2018-11-08 pendulum_exp/policies/continuous.py

@@ -42,12 +42,17 @@ class SampledAdvantagePolicy(SharedAdvantagePolicy):
     def act(self, obs: Arrayable):
         with torch.no_grad():
             obs = check_array(obs)
+            batch_size = obs.shape[0]
+            o_dim = obs.ndim
+
             proposed_actions = np.random.uniform(
-                -1, 1, [self._nb_samples, obs.shape[0]] + self._action_shape)
+                -1, 1, [self._nb_samples, batch_size, *self._action_shape])
             advantages = self._adv_noise.perturb_output(
-                obs[np.newaxis, ...], proposed_actions, function=proposed_actions)
+                np.tile(obs, [self._nb_samples] + o_dim * [1]),
+                proposed_actions, function=self._adv_function).squeeze()
             self._adv_noise.step()
-            action = proposed_actions[np.argmax(advantages, axis=0)]
+            action_idx = np.argmax(advantages, axis=0)
+            action = proposed_actions[action_idx, np.arange(obs.shape[0])]
         return action
 
     def learn(self):
@@ -59,10 +64,10 @@ class SampledAdvantagePolicy(SharedAdvantagePolicy):
                     adv_a = self._adv_function(obs, action).squeeze()
 
                     proposed_actions = np.random.uniform(
-                        -1, 1, [self._nb_samples, obs.shape[0]] + self._action_shape)
+                        -1, 1, [self._nb_samples, obs.shape[0], *self._action_shape])
                     max_adv = torch.max(
                         self._adv_function(
-                            obs[np.newaxis, ...],
+                            np.tile(obs, [self._nb_samples] + obs.ndim * [1]),
                             proposed_actions).squeeze(), dim=0)[0]
 
                     if self._gamma == 1:
@@ -106,13 +111,11 @@ class SampledAdvantagePolicy(SharedAdvantagePolicy):
         self._train = True
         self._val_function.train()
         self._adv_function.train()
-        self._policy_function.train()
 
     def eval(self):
         self._train = False
         self._val_function.eval()
         self._adv_function.eval()
-        self._policy_function.eval()
 
     def value(self, obs: Arrayable):
         return th_to_arr(self._val_function(obs))

### 2018-11-09 paper/advup_uniquesol.tex

\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}

\usepackage{xcolor}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}

\newcommand{\deltat}{{\delta t}}
\newcommand{\gauss}{\mathcal{N}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\TODO}[1]{\textcolor{red}{#1}}
\newcommand{\bigO}[1]{O(#1)}
\newcommand{\Tdo}{TD($0$)}
\newcommand{\tdres}{\delta_\text{TD}}

\newcommand{\deq}{\mathrel{\mathop{:}}=}
\newcommand{\eqd}{=\mathrel{\mathop{:}}}

\title{Baseline centering problem}
\begin{document}
\maketitle

Let $\pi$ be any policy (or set of observed transitions) and let
$\rho^\pi$ be the associated distribution on states.

Let us start again from  the equations on $A^*$, $V^*$ and $b$:
\begin{gather}
	A(s, a) = \max\limits_{a'} A(s, a') + 
		(r(s, a) - b)
		 + 
		\gamma \E_{s'\mid a}\left[
			V(s')
		\right] - V(s)
	\\
	\max\limits_{a} A(s, a) = 0\\
	\E_{s\sim \rho^\pi}\left[V(s)\right] = 0
	.
\end{gather}

Let us prove existence and uniqueness of the solution over $(V,b)$ for a
given function $A$. Moreover, let us prove that when $\gamma\to 1$, $b$
tends to the average reward under the policy derived from $A$.

\paragraph{Existence of a solution.}

Let $\pi_A$ be the (or any in case of ties) policy described by the advantage function $A$,
namely, that returns an action $a^\ast$ that maximizes $A(s,a)$.

Applying the equation to $a^\ast$ we find
\begin{equation}
\label{eq:optbellman}
0=r(s,a^\ast)-b+\gamma \E_{s'|a^\ast} V(s')-V(s)
\end{equation}

This is the Bellman equation for the policy $\pi_A$ with rewards $r-b$.
Shifting rewards by $b$ shifts value functions by $b/(1-\gamma)$. So
under ergodicity, we know that the unique solution is
\begin{equation}
V(s)=V^{\pi_A}(s)-\frac{b}{1-\gamma}
\end{equation}
and averaging under the current empirical policy $\pi$ and using
$\E_{\rho^\pi} V=0$ we find
\begin{equation}
b=(1-\gamma)\E_{s\sim \rho^\pi} V^{\pi_A}(s)
\end{equation}

Conversely, the function $V(s)\deq V^{\pi_A}(s)-\frac{b}{1-\gamma}$ with
$b\deq (1-\gamma)\E_{s\sim \rho^\pi} V^{\pi_A}(s)$ satisfies 
both $\E_{\rho^\pi} V=0$ (by construction) and the Bellman equation for
every pair $(s,a^\ast)$, that is, the Bellman equation for $\pi_A$.

TODO Satisfying Bellman for every pair $(s,a)$ instead of just $(s,a^\ast)$
tells that $A$ is indeed the advantage function of $\pi_A$.

Thus, interestingly, the system provides a well-defined solution for $V$
and $b$ even if $A$ is not perfect (namely, the value function associated
with $\pi_A$). This provides some numerical protection.

\paragraph{The value of $b$ when $\gamma\to 1$.}
We know that when $\gamma\to 1$, the value function for $\pi_A$ takes the form
\begin{equation}
\label{eq:reladv}
	V^{\pi_A}(s) = \frac{\E_{(s,a^\ast)\sim\rho^{\pi_A}}r(s,a^\ast) }{1 - \gamma} +
	U(s) + O(1-\gamma)
\end{equation}
where $U(s)$ is the relative advantage function of the policy $\pi^A$.
The function $U$
is bounded when $\gamma\to 1$.

Since $b=(1-\gamma)\E_{\rho^\pi} V^{\pi_A}$ we just find
\begin{equation}
b=\E_{(s,a^\ast)\sim\rho^{\pi_A}}r(s,a^\ast)+O(1-\gamma)
\end{equation}
namely, $b$ is the average reward under policy $\pi^A$.

\end{document}

### 2018-11-09 paper/advup_uniquesol.tex

@@
 		 + 
-		\gamma \E_{s'\mid a}\left[
+		\gamma \E_{s'\mid s,a}\left[
 			V(s')
@@
 \label{eq:optbellman}
-0=r(s,a^\ast)-b+\gamma \E_{s'|a^\ast} V(s')-V(s)
+0=r(s,a^\ast)-b+\gamma \E_{s'|s,a^\ast} V(s')-V(s)
 \end{equation}
@@
 \end{equation}
-and averaging under the current empirical policy $\pi$ and using
-$\E_{\rho^\pi} V=0$ we find
+and averaging under the current empirical policy $\pi$, the only $b$ that
+satisfies
+$\E_{\rho^\pi} V=0$ is
 \begin{equation}
@@
 
-Conversely, the function $V(s)\deq V^{\pi_A}(s)-\frac{b}{1-\gamma}$ with
-$b\deq (1-\gamma)\E_{s\sim \rho^\pi} V^{\pi_A}(s)$ satisfies 
-both $\E_{\rho^\pi} V=0$ (by construction) and the Bellman equation for
+Conversely, defining 
+\begin{equation}
+V(s)\deq V^{\pi_A}(s)-\frac{b}{1-\gamma},\qquad
+b\deq (1-\gamma)\E_{s\sim \rho^\pi} V^{\pi_A}(s)
+\end{equation}
+these functions satisfy
+both the equation $\E_{\rho^\pi} V=0$ (by construction) and the Bellman equation for
 every pair $(s,a^\ast)$, that is, the Bellman equation for $\pi_A$.
-
-TODO Satisfying Bellman for every pair $(s,a)$ instead of just $(s,a^\ast)$
-tells that $A$ is indeed the advantage function of $\pi_A$.
 
@@
 
+[TODO: this assumes we only apply Bellman to $(s,a^\ast)$ not every
+$(s,a)$. SHOULD WE RESTRICT UPDATES OF $V$ and $b$ TO SUCH PAIRS? Using
+Bellman for $a\neq a^\ast$ might just pollute $V$ and $b$. Eg, if using
+eps-greedy, don't update $V$ when exploring, only update $A$?]
+
+TODO Satisfying Bellman for every pair $(s,a)$ instead of just $(s,a^\ast)$
+tells that $A$ is indeed the advantage function of $\pi_A$.
+
+TODO this still works if we have any heuristic procedure that returns a
+would-be argmax of $A$ instead of the true argmax!
+
+TODO Should we multiply the learning rate of $b$ by $1-\gamma$ if we want
+a well-defined algorithm when $\gamma\to 1$? I don't think it's needed: in
+Bellman, thanks to $\gamma V(s')-V(s)$, the homogeneity of $V$ is
+$(1-\gamma)$. So even if $V$ is initialized to a large value such as
+$V^{\pi_A}$ it should be fine.
+
 \paragraph{The value of $b$ when $\gamma\to 1$.}
@@
 \label{eq:reladv}
-	V^{\pi_A}(s) = \frac{\E_{(s,a^\ast)\sim\rho^{\pi_A}}r(s,a^\ast) }{1 - \gamma} +
+	V^{\pi_A}(s) = \frac{\E_{(s,a^\ast)\sim\rho^{\pi_A}}\,r(s,a^\ast) }{1 - \gamma} +
 	U(s) + O(1-\gamma)
@@
 \begin{equation}
-b=\E_{(s,a^\ast)\sim\rho^{\pi_A}}r(s,a^\ast)+O(1-\gamma)
+b=\E_{(s,a^\ast)\sim\rho^{\pi_A}}\,r(s,a^\ast)+O(1-\gamma)
 \end{equation}

### 2018-11-09 pendulum_exp/policies/continuous.py
Refactor done, needs testing

@@ -1,10 +1,10 @@
 """Define continuous policy."""
 from itertools import chain
 import torch
+from torch import Tensor
 import numpy as np
 from abstract import ParametricFunction, Arrayable, Noise, StateDict
-from convert import arr_to_th, th_to_arr, check_array
-from stats import penalize_mean
+from convert import th_to_arr, check_array
 from config import SampledAdvantagePolicyConfig, AdvantagePolicyConfig
 from policies.shared import SharedAdvantagePolicy
 from logging import info
@@ -18,23 +18,24 @@ class SampledAdvantagePolicy(SharedAdvantagePolicy):
                  policy_config: SampledAdvantagePolicyConfig,
                  action_shape,
                  device) -> None:
-        super().__init__(policy_config)
+        super().__init__(policy_config, val_function, device)
 
         self._adv_function = adv_function.to(device)
-        self._val_function = val_function.to(device)
-        self._baseline = torch.nn.Parameter(torch.Tensor([0.]).to(device))
         self._adv_noise = adv_noise
         self._nb_samples = policy_config.nb_samples
         self._action_shape = action_shape
 
         # optimization/storing
-        self._device = device
         self._optimizers = (
-            torch.optim.SGD(chain(self._adv_function.parameters(), [self._baseline]), lr=policy_config.lr * self._dt),
-            torch.optim.SGD(self._val_function.parameters(), lr=policy_config.lr * self._dt ** 2))
+            torch.optim.SGD(chain(self._adv_function.parameters(),
+                                  [self._baseline]), lr=policy_config.lr * self._dt),
+            torch.optim.SGD(self._val_function.parameters(),
+                            lr=policy_config.lr * self._dt ** 2))
         self._schedulers = (
-            torch.optim.lr_scheduler.LambdaLR(self._optimizers[0], policy_config.lr_decay),
-            torch.optim.lr_scheduler.LambdaLR(self._optimizers[1], policy_config.lr_decay))
+            torch.optim.lr_scheduler.LambdaLR(self._optimizers[0],
+                                              policy_config.lr_decay),
+            torch.optim.lr_scheduler.LambdaLR(self._optimizers[1],
+                                              policy_config.lr_decay))
 
         # logging
         self._cum_loss = 0
@@ -55,57 +56,39 @@ class SampledAdvantagePolicy(SharedAdvantagePolicy):
             action = proposed_actions[action_idx, np.arange(obs.shape[0])]
         return action
 
[…]

### 2018-11-09 pendulum_exp/policies/discrete.py

@@ -1,9 +1,9 @@
 """Define policies."""
 from itertools import chain
 import torch
+from torch import Tensor
 from abstract import ParametricFunction, Arrayable, Noise, StateDict
 from convert import arr_to_th, th_to_arr
-from stats import penalize_mean
 from config import PolicyConfig
 from policies.shared import SharedAdvantagePolicy
 from mylog import log
@@ -16,20 +16,21 @@ class AdvantagePolicy(SharedAdvantagePolicy):
                  adv_noise: Noise,
                  policy_config: PolicyConfig,
                  device) -> None:
-        super().__init__(policy_config)
+        super().__init__(policy_config, val_function, device)
         self._adv_function = adv_function.to(device)
-        self._val_function = val_function.to(device)
-        self._baseline = torch.nn.Parameter(torch.Tensor([0.]).to(device))
         self._adv_noise = adv_noise
 
         # optimization/storing
-        self._device = device
         self._optimizers = (
-            torch.optim.SGD(chain(self._adv_function.parameters(), [self._baseline]), lr=policy_config.lr * policy_config.dt),
-            torch.optim.SGD(self._val_function.parameters(), lr=policy_config.lr * policy_config.dt ** 2))
+            torch.optim.SGD(chain(self._adv_function.parameters(), [self._baseline]),
+                            lr=policy_config.lr * policy_config.dt),
+            torch.optim.SGD(self._val_function.parameters(),
+                            lr=policy_config.lr * policy_config.dt ** 2))
         self._schedulers = (
-            torch.optim.lr_scheduler.LambdaLR(self._optimizers[0], policy_config.lr_decay),
-            torch.optim.lr_scheduler.LambdaLR(self._optimizers[1], policy_config.lr_decay))
+            torch.optim.lr_scheduler.LambdaLR(self._optimizers[0],
+                                              policy_config.lr_decay),
+            torch.optim.lr_scheduler.LambdaLR(self._optimizers[1],
+                                              policy_config.lr_decay))
         # logging
         self._cum_loss = 0
         self._log = 100
@@ -41,54 +42,34 @@ class AdvantagePolicy(SharedAdvantagePolicy):
             self._adv_noise.step()
             return pre_action.argmax(axis=-1)
 
-    def learn(self):
-        if self._count % self._steps_btw_train == self._steps_btw_train - 1:
-            try:
-                for _ in range(self._learn_per_step):
-                    obs, action, next_obs, reward, done = self._sampler.sample()
-                    # for now, discrete actions
[…]

### 2018-11-09 pendulum_exp/policies/shared.py

@@ -1,11 +1,16 @@
 """ Define shared elements between continuous and discrete. """
 import numpy as np
-from abstract import Policy, Arrayable
+from abstract import Policy, Arrayable, ParametricFunction
 from config import PolicyConfig
 from memory import MemorySampler
+import torch
+from torch import Tensor
+from convert import arr_to_th, check_array
+from stats import penalize_mean
 
 class SharedAdvantagePolicy(Policy):
-    def __init__(self, policy_config: PolicyConfig) -> None:
+    def __init__(self, policy_config: PolicyConfig,
+                 val_function: ParametricFunction, device) -> None:
         self._train = True
         self.reset()
 
@@ -20,6 +25,11 @@ class SharedAdvantagePolicy(Policy):
             batch_size=policy_config.batch_size)
         self._count = 0
         self._learn_count = 0
+        self._device = device
+        self._val_function = val_function
+
+        # baseline (used in all advantage policies)
+        self._baseline = torch.nn.Parameter(torch.Tensor([0.]).to(device))
 
     def reset(self):
         # internals
@@ -54,3 +64,54 @@ class SharedAdvantagePolicy(Policy):
             self._sampler.push(
                 self._obs, self._action, self._next_obs, self._reward, self._done)
             self.learn()
+
+    def learn(self):
+        if self._count % self._steps_btw_train == self._steps_btw_train - 1:
+            try:
+                for _ in range(self._learn_per_step):
+                    obs, action, next_obs, reward, done = self._sampler.sample()
+                    done = arr_to_th(check_array(done).astype('float'), self._device)
+                    reward = arr_to_th(reward, self._device)
+
+                    v = self._val_function(obs).squeeze()
+                    next_v = self.compute_next_value(next_obs, done)
+                    adv, max_adv = self.compute_advantages(
+                        obs, action)
+
+                    expected_v = (reward - self._baseline) * self._dt + \
+                        self._gamma ** self._dt * next_v
+                    dv = (expected_v - v) / self._dt
+                    bell_residual = dv - adv + max_adv
+
+                    adv_update_loss = (bell_residual ** 2).mean()
+                    adv_norm_loss = (max_adv ** 2).mean()
+                    mean_loss = self._alpha * penalize_mean(v)
+                    bell_loss = adv_update_loss + adv_norm_loss
+
+                    self.optimize_value(bell_loss, mean_loss / self.dt)
+                    self.optimize_policy(max_adv)
+
[…]

### 2018-11-09 pendulum_exp/mylog.py
Refactor seemingly working

@@ -40,10 +40,9 @@ class Logger:
         info("logfile: {}".format(join(self._dir, 'logs.pkl')))
 
     def load(self):
-        assert self._file is not None
-        with open(self.filename, 'rb') as f:
+        assert self._dir is not None
+        with open(join(self._dir, 'logs.pkl'), 'rb') as f:
             self._logs = pickle.load(f)
-
 
     def dump(self):
         assert self._dir is not None

### 2018-11-09 pendulum_exp/policies/continuous.py

@@ -60,7 +60,7 @@ class SampledAdvantagePolicy(SharedAdvantagePolicy):
         obs = check_array(obs)
         action = check_array(action)
         proposed_actions = np.random.uniform(
-            -1, 1, [self._nb_samples, obs.shape[0], self._action_shape])
+            -1, 1, [self._nb_samples, obs.shape[0], *self._action_shape])
         adv = self._adv_function(obs, action).squeeze()
         max_adv = torch.max(
             self._adv_function(

### 2018-11-09 pendulum_exp/policies/shared.py

@@ -88,7 +88,7 @@ class SharedAdvantagePolicy(Policy):
                     mean_loss = self._alpha * penalize_mean(v)
                     bell_loss = adv_update_loss + adv_norm_loss
 
-                    self.optimize_value(bell_loss, mean_loss / self.dt)
+                    self.optimize_value(bell_loss, mean_loss / self._dt)
                     self.optimize_policy(max_adv)
 
                 self.log()

### 2018-11-09 paper/advup_uniquesol.tex

@@
 
-Let us start again from  the equations on $A^*$, $V^*$ and $b$:
+Let us start again from  the equations on $A$, $V$ and $b$:
 \begin{gather}
@@
 Shifting rewards by $b$ shifts value functions by $b/(1-\gamma)$. So
-under ergodicity, we know that the unique solution is
+the unique solution is
 \begin{equation}
@@
 [TODO: this assumes we only apply Bellman to $(s,a^\ast)$ not every
-$(s,a)$. SHOULD WE RESTRICT UPDATES OF $V$ and $b$ TO SUCH PAIRS? Using
+$(s,a)$! SHOULD WE RESTRICT UPDATES OF $V$ and $b$ TO SUCH PAIRS? Using
 Bellman for $a\neq a^\ast$ might just pollute $V$ and $b$. Eg, if using
@@
 
+\paragraph{Convergence of $V$ given $b$.} For robustness analysis, we now
+derive what happens to one variable when the other is wrong. Let us first
+study convergence of $V$ given $b$. There is a subtlety with the
+respective strength of the Bellman objective and the $\E V=0$ objective.
+This is because TD is not a gradient and behaves badly when added to the
+gradient of another loss.
+
+The loss corresponding to $\E_{\rho^\pi}V=0$ is $\frac12
+(\E_{\rho^\pi}V)^2$. We can estimate its gradient as $(\E_{\rho^\pi}\partial_\theta
+V)(\E_{\rho^\pi}V)$ where the two expectations require two independent
+samples from the current policy, eg, sampling twice from the training
+dataset.
+
+Assume that we learn $V$ via TD, to which we add $\alpha$ times the
+centering objective on $V$. Adding the TD step and the centering gradient
+step, the average step is
+\begin{equation}
+\E_{(s,a)\sim\rho^\pi}\partial_\theta
+V(s)(A(s,a^\ast)-A(s,a)+r(s,a)-b+\gamma \E_{s'|s,a}V(s')-V(s) - \alpha
+\E_{\rho^\pi}V)
+\end{equation}
+which corresponds to using TD to solve the modified Bellman equation
+with rewards $A(s,a^\ast)-A(s,a)+r(s,a)-b-\alpha \E_{\rho^\pi}V$. The
+solution is... TODO
+
+Let us consider the case when we just train $V$ on pairs $(s,a^\ast)$.
+Then the above uses TD to solve the modified Bellman equation with
+rewards $r(s,a)-b-\alpha \E_{\rho^\pi}V$. Thus we are trying to solve
+$r(s,a)-b-\alpha \E_{\rho^\pi}V+\gamma PV-V=0$, and the only fixed point is
+\begin{equation}
+V(s)=V^{\pi_A}(s)-\frac{b+\E_{\rho^\pi} V^{\pi_A}}{1+\alpha-\gamma}
+\end{equation}
+which looks different from the solution
+\begin{equation}
+V(s)=V^{\pi_A}(s)-\frac{b}{1-\gamma}
+\end{equation}
+considered above; however, these two may still coincide for the value of $b$ at
+equilibrium.
+Meanwhile, $b$ is trained to minimize $\E_{s\sim \rho^\pi}
+(r(s,a^\ast)-b+\gamma \E_{s'|s,a^\ast}V(s')-V(s))^2$ whose unique
+solution given $V$ is
+\begin{equation}
+b=\E_{s\sim \rho^\pi}(r(s,a^\ast)+\gamma \E_{s'|s,a^\ast}V(s')-V(s))
+\end{equation}
+and substituting the fixed point
+$V(s)=V^{\pi_A}(s)-\frac{b+\E V^{\pi_A}}{1-\gamma+\alpha}$ we find
+\begin{equation}
+b=-(\gamma-1) \frac{b+\E V^{\pi_A}}{1-\gamma+\alpha}
+\end{equation}
+which yields
+\begin{equation}
+b=\frac{1-\gamma}{\alpha} \E V^{\pi_A}
+\end{equation}
+contrary to $b=(1-\gamma)\E V^{\pi_A}$ we had before. Let us check that
+the constraint $\E V=0$ is satisfied:
+\begin{equation}
+\E V= \E V^{\pi_A} -\frac{b+\E_{\rho^\pi} V^{\pi_A}}{1+\alpha-\gamma}
+\end{equation}
+
 \end{document}

### 2018-11-09 paper/advup_uniquesol.tex

@@
 
+\paragraph{Only $\alpha=1$, i.e. penalty $\frac12
+(\E_{\rho^\pi}V)^2$, works for balancing the constraint on $\E V$
+and TD on $V$.}
+
 Let us consider the case when we just train $V$ on pairs $(s,a^\ast)$.
-Then the above uses TD to solve the modified Bellman equation with
+Then the combined TD+centering step on $V$, with centering loss
+$\frac{\alpha}{2} (\E_{\rho^\pi}V)^2$, is
+\begin{equation}
+\E_{s\sim\rho^\pi}\partial_\theta
+V(s)(r(s,a^\ast)-b+\gamma \E_{s'|s,a^\ast}V(s')-V(s) - \alpha
+\E_{\rho^\pi}V)
+\end{equation}
+
+This is equivalent to using TD
+to solve the modified Bellman equation with
 rewards $r(s,a)-b-\alpha \E_{\rho^\pi}V$. Thus we are trying to solve
@@
 \end{equation}
-considered above; however, these two may still coincide for the value of $b$ at
-equilibrium.
+considered above.
 Meanwhile, $b$ is trained to minimize $\E_{s\sim \rho^\pi}
@@
 \end{equation}
-contrary to $b=(1-\gamma)\E V^{\pi_A}$ we had before. Let us check that
-the constraint $\E V=0$ is satisfied:
+contrary to $b=(1-\gamma)\E V^{\pi_A}$ we had before. 
+This translates back to
 \begin{equation}
-\E V= \E V^{\pi_A} -\frac{b+\E_{\rho^\pi} V^{\pi_A}}{1+\alpha-\gamma}
+V(s)=V^{\pi_A}(s)-\frac{b+\E_{\rho^\pi} V^{\pi_A}}{1+\alpha-\gamma}
+=V^{\pi_A}(s)-\frac1\alpha \E_{\rho^\pi} V^{\pi_A}
 \end{equation}
+with this value of $b$.
+
+The constraint $\E V=0$ is satisfied if and only if $\alpha=1$.
+
+In conclusion, when $\alpha=1$, everything coincides with the desired
+solution; however for $\alpha\neq 1$ this is not the case.
 

### 2018-11-09 pendulum_exp/stats.py
Rescaling gradient penalty

@@ -42,4 +42,4 @@ def penalize_mean(arr: torch.Tensor):
     length = min(torch.sum(mask), torch.sum(1 - mask)).item()
     if length == 0:
         return torch.Tensor([0]).to(arr.device)
-    return (arr[mask][:length] * arr[1 - mask][:length]).mean()
+    return (arr[mask][:length] * arr[1 - mask][:length]).mean() / 2

### 2018-11-09 pendulum_exp/policies/shared.py
Small fix for gamma=1 (.byte() before .all())

@@ -111,7 +111,7 @@ class SharedAdvantagePolicy(Policy):
         done = arr_to_th(check_array(done).astype('float'), self._device)
         next_v = self._val_function(next_obs).squeeze().detach()
         if self._gamma == 1:
-            assert (1 - done).all(), "Gamma set to 1. with a potentially"\
+            assert (1 - done).byte().all(), "Gamma set to 1. with a potentially"\
                 "episodic problem..."
             return next_v
         return (1 - done) * next_v - done * self._baseline / (1 - self._gamma)

### 2018-11-10 paper/advup_uniquesol.tex

@@
 \begin{equation}
-V(s)\deq V^{\pi_A}(s)-\frac{b}{1-\gamma},\qquad
+V(s)\deq V^{\pi_A}(s)-\E_{s\sim \rho^\pi} V^{\pi_A}(s),\qquad
 b\deq (1-\gamma)\E_{s\sim \rho^\pi} V^{\pi_A}(s)
@@
 rewards $r(s,a)-b-\alpha \E_{\rho^\pi}V$. Thus we are trying to solve
-$r(s,a)-b-\alpha \E_{\rho^\pi}V+\gamma PV-V=0$, and the only fixed point is
+$r(s,a)-b-\alpha \E_{\rho^\pi}V+\gamma PV-V=0$, and the only fixed point
+of this gradient step is
 \begin{equation}
@@
 \end{equation}
-which looks different from the solution
-\begin{equation}
-V(s)=V^{\pi_A}(s)-\frac{b}{1-\gamma}
-\end{equation}
-considered above.
+% which looks different from the solution
+% \begin{equation}
+% V(s)=V^{\pi_A}(s)-\frac{b}{1-\gamma}
+% \end{equation}
+% considered above.
 Meanwhile, $b$ is trained to minimize $\E_{s\sim \rho^\pi}
 (r(s,a^\ast)-b+\gamma \E_{s'|s,a^\ast}V(s')-V(s))^2$ whose unique
-solution given $V$ is
+solution given $V$ is TODO WRONG I THINK WE HAVE A DOUBLE SAMPLING
+PROBLEM FOR $b$!!!
 \begin{equation}
@@
 
+\paragraph{A variant to make the learning of $A$ and $b$ insensitive to
+translations of $V$, and the dynamics independent from
+$\alpha_3/\alpha_5$.} Consider the modified off-policy Bellman equation
+for all states $s$,
+\begin{equation}
+0=\E \left(
+r(s,a)-b+\gamma V(s')-V(s)+(1-\gamma)V(\tilde s)
+\right)
+\end{equation}
+where $a$ is from some policy $\pi$, $s'$ is the next state given $s$ and
+$a$, and $\tilde s$ is a random state taken from any empirical
+distribution $\rho$. Minimize this by TD with $s$ taken from the same
+empirical distribution $\rho$, adding a gradient term to minimize
+$\frac{\alpha}{2} (\E_\rho V)^2$ as before.
+
+The TD step is
+\begin{equation}
+\E_s \,\partial_\theta V(s)\left(
+r(s,a)-b+\gamma V(s')-V(s)+(1-\gamma)V(\tilde s)-\alpha V(s)
+\right)
+\end{equation}
+and its fixed point solves $r(s,a)-b+\gamma V(s')-V(s)+(1-\gamma)V(\tilde
+s)-\alpha V(s)$ for all $s$. The unique solution for $V$ given $b$ is
+\begin{equation}
+V(s)=V^\pi(s)-\E_\rho V^\pi-\frac{b-(1-\gamma)\E_\rho V^\pi}{\alpha}
+\end{equation}
+
+Meanwhile, $b$ is trained to minimize $\E_\rho (\E)$
+
 \end{document}

### 2018-11-10 paper/advup_uniquesol.tex

@@
 % considered above.
-Meanwhile, $b$ is trained to minimize $\E_{s\sim \rho^\pi}
-(r(s,a^\ast)-b+\gamma \E_{s'|s,a^\ast}V(s')-V(s))^2$ whose unique
-solution given $V$ is TODO WRONG I THINK WE HAVE A DOUBLE SAMPLING
-PROBLEM FOR $b$!!!
+Meanwhile, $b$ is trained to minimize $\E_{s\sim
+\rho^\pi}\E_{s'|s,a^\ast}
+(r(s,a^\ast)-b+\gamma V(s')-V(s))^2$ whose unique
+solution given $V$ is 
 \begin{equation}
@@
 
-Meanwhile, $b$ is trained to minimize $\E_\rho (\E)$
+Meanwhile, $b$ is trained to minimize $\E_s\E_{s'|s,a}\E_{\tilde s} (r-b+\gamma
+V(s')-V(s)+(1-\gamma)V(\tilde s))^2$ whose unique solution is
+\begin{equation}
+b=\E_s\E_{s'|s,a}\E_{\tilde s}(r+\gamma
+V(s')-V(s)+(1-\gamma)V(\tilde s))
+\end{equation}
+Substituting the equilibrium value for $V$, $V=V^\pi+c$ yields
+\begin{equation}
+b=(1-\gamma)\E_{\tilde s} V^\pi(\tilde s)=(1-\gamma)\E_\rho V^\pi
+\end{equation}
+whatever $c$ is. Thus, for this optimal value of $b$, we have
+\begin{equation}
+V(s)=V^\pi(s)-\E_\rho V^\pi
+\end{equation}
 

### 2018-11-10 paper/advup_uniquesol.tex

@@
 $a$, and $\tilde s$ is a random state taken from any empirical
-distribution $\rho$. Minimize this by TD with $s$ taken from the same
+distribution $\rho$.
+
+With this expression, the Bellman gap is insensitive to constants on $V$.
+
+This expression does not change the optima of the original
+problem, since we have just added $(1-\gamma)\E_\rho V$ and the solution
+of the original problem makes this $0$ by definition.
+
+Let us minimize this by TD with $s$ taken from the same
 empirical distribution $\rho$, adding a gradient term to minimize
@@
 Meanwhile, $b$ is trained to minimize $\E_s\E_{s'|s,a}\E_{\tilde s} (r-b+\gamma
-V(s')-V(s)+(1-\gamma)V(\tilde s))^2$ whose unique solution is
+V(s')-V(s)+(1-\gamma)V(\tilde s))^2$. Given $V$, the unique solution is
 \begin{equation}
@@
 \end{equation}
-Substituting the equilibrium value for $V$, $V=V^\pi+c$ yields
+Substituting the equilibrium value for $V$, $V=V^\pi+\mathrm{cst}$ yields
 \begin{equation}
@@
 \end{equation}
-whatever $c$ is. Thus, for this optimal value of $b$, we have
+whatever $\mathrm{cst}$ is. Thus, for this optimal value of $b$, we have
 \begin{equation}

### 2018-11-10 paper/advup_uniquesol.tex

@@
 
+\paragraph{Sampling $\tilde s$ from a single trajectory.} Given a
+trajectory $(s_t)$ from any policy, we can keep online a set of unbiased
+samples of the trajectory, as follows. Just maintain a set of $k$ past
+states with the following replacement rule: at time $t$, with probability
+$k/t$, replace one of the previous states with the current state $s_t$.
+
 \end{document}

### 2018-11-10 paper/advup_uniquesol.tex

@@
 \E_s \,\partial_\theta V(s)\left(
-r(s,a)-b+\gamma V(s')-V(s)+(1-\gamma)V(\tilde s)-\alpha V(s)
+r(s,a)-b+\gamma V(s')-V(s)+(1-\gamma)V(\tilde s)-\alpha V(\tilde s)
 \right)
@@
 and its fixed point solves $r(s,a)-b+\gamma V(s')-V(s)+(1-\gamma)V(\tilde
-s)-\alpha V(s)$ for all $s$. The unique solution for $V$ given $b$ is
+s)-\alpha V(\tilde s)$ for all $s$. The unique solution for $V$ given $b$ is
 \begin{equation}
@@
 
+This means storing past states and recomputing the value function
+at every time, not just storing past value functions.
+
 \end{document}

### 2018-11-10 paper/advup_uniquesol.tex

@@
 
+Interestingly, the distributions for $s$ and $\tilde s$ do not even need
+to be the same. Namely, if $s\sim \rho$ and $\tilde s\sim \tilde \rho$,
+then $\partial_\theta V(s) V(\tilde s)$ is not a gradient of either
+$(\E_\rho V)^2$ or $(\E_{\tilde \rho}V)^2$. But we can still add $V(\tilde
+s)$ to the TD step at $s$ and look for a fixed point. The fixed point of
+the above is when, for any $s$, 
+\begin{equation}
+\E_{a,s',\tilde s|s} \left(r(s,a)-b+\gamma
+V(s')-V(s)+(1-\gamma-\alpha)V(\tilde s)\right)=0
+\end{equation}
+and this does not involve the distribution of $s$, so the unique solution is similar to the above, with expectations under
+$\tilde \rho$:
+\begin{equation}
+V(s)=V^\pi(s)-\E_{\tilde \rho} V^\pi-\frac{b-(1-\gamma)\E_{\tilde
+\rho}V^\pi}{\alpha}
+\end{equation}
+(And if $V_\theta$ is overparameterized, if we converge we converge to
+this.) Likewise, the solution for $b$ is as above, and if
+$V=V^\pi+\mathrm{cst}$ then 
+\begin{equation}
+b=(1-\gamma)\E_{\tilde s} V^\pi(\tilde s)=(1-\gamma)\E_{\tilde \rho} V^\pi
+\end{equation}
+
+We may even take a single value $s_0$ for $\tilde s$ as a reference point
+($\tilde \rho$ a Dirac mass at $s_0$): $V$ would converge to
+$V^\pi-V^\pi(s_0)$, and $b$ to $(1-\gamma)V^\pi(s_0)$. This may be the
+simplest practical solution. (If changing $s_0$ from time to time, $V$
+would have to be shifted by the change of $V^\pi(s_0)$.)
+
 \paragraph{Sampling $\tilde s$ from a single trajectory.} Given a

### 2018-11-10 paper/advup_uniquesol.tex

--- 2018-11-10 paper/advup_uniquesol.tex
@@
 where $a$ is from some policy $\pi$, $s'$ is the next state given $s$ and
-$a$, and $\tilde s$ is a random state taken from any empirical
-distribution $\rho$.
+$a$, and $\tilde s$ is a random state, independent from $s$, taken from any empirical
+distribution $\tilde \rho$.
 
@@
 
-Let us minimize this by TD with $s$ taken from the same
-empirical distribution $\rho$, adding a gradient term to minimize
-$\frac{\alpha}{2} (\E_\rho V)^2$ as before.
+Let us minimize this by TD with $s$ taken from some other empirical
+distribution $\rho$, and add a TD-style gradient term to minimize
+$\frac{\alpha}{2} (\E_{\tilde \rho} V)^2$. Interestingly we do not need
+to have $\rho=\tilde \rho$: in that case, we are not taking a true
+gradient of $\frac{\alpha}{2} (\E_{\tilde \rho} V)^2$ but a TD-style
+gradient, but it still has the right fixed point, as we show now.
 
@@
 \end{equation}
-and its fixed point solves $r(s,a)-b+\gamma V(s')-V(s)+(1-\gamma)V(\tilde
-s)-\alpha V(\tilde s)$ for all $s$. The unique solution for $V$ given $b$ is
+and its fixed point solves $0=\E[r(s,a)-b+\gamma V(s')-V(s)+(1-\gamma)V(\tilde
+s)-\alpha V(\tilde s)]$ for all $s$.\footnote{Here I assume the family
+$V_\theta$ is
+overparameterized: in that case, if the average TD step is $0$ then the
+right-hand side vanishes for every $s$. If not, the difference between $V(s)$ and the solution
+presented is orthogonal to the span of $\partial_\theta V$ wrt some weird
+metric.} The unique solution for $V$ given $b$ is
 \begin{equation}
-V(s)=V^\pi(s)-\E_\rho V^\pi-\frac{b-(1-\gamma)\E_\rho V^\pi}{\alpha}
+V(s)=V^\pi(s)-\E_{\tilde \rho} V^\pi-\frac{b-(1-\gamma)\E_{\tilde \rho} V^\pi}{\alpha}
 \end{equation}
@@
 \begin{equation}
-b=(1-\gamma)\E_{\tilde s} V^\pi(\tilde s)=(1-\gamma)\E_\rho V^\pi
+b=(1-\gamma)\E_{\tilde s} V^\pi(\tilde s)=(1-\gamma)\E_{\tilde \rho} V^\pi
 \end{equation}
@@
 \begin{equation}
-V(s)=V^\pi(s)-\E_\rho V^\pi
+V(s)=V^\pi(s)-\E_{\tilde \rho} V^\pi
 \end{equation}
+
+Notably the distributions of $s$ and $\tilde s$ do not have to match. We may even take a single value $s_0$ for $\tilde s$ as a reference point
+($\tilde \rho$ a Dirac mass at $s_0$): $V$ would converge to
+$V^\pi-V^\pi(s_0)$, and $b$ to $(1-\gamma)V^\pi(s_0)$. This may be the
+simplest practical solution. (If changing $s_0$ from time to time, $V$
+would have to be shifted by the change in $V^\pi(s_0)$.)
 

### 2018-11-10 pendulum_exp/policies/shared.py
Normpp implemented

@@ -70,22 +70,24 @@ class SharedAdvantagePolicy(Policy):
             try:
                 for _ in range(self._learn_per_step):
                     obs, action, next_obs, reward, done = self._sampler.sample()
+                    indep_obs, _, _, _, _ = self._sampler.sample()
                     done = arr_to_th(check_array(done).astype('float'), self._device)
                     reward = arr_to_th(reward, self._device)
 
                     v = self._val_function(obs).squeeze()
                     next_v = self.compute_next_value(next_obs, done)
+                    indep_v = self._val_function(indep_obs).squeeze().detach()
                     adv, max_adv = self.compute_advantages(
                         obs, action)
 
                     expected_v = (reward - self._baseline) * self._dt + \
                         self._gamma ** self._dt * next_v
                     dv = (expected_v - v) / self._dt
-                    bell_residual = dv - adv + max_adv
+                    bell_residual = dv - adv + max_adv + (1 - self._gamma) * indep_v.mean()
 
                     adv_update_loss = (bell_residual ** 2).mean()
                     adv_norm_loss = (max_adv ** 2).mean()
-                    mean_loss = self._alpha * penalize_mean(v)
+                    mean_loss = self._alpha * penalize_mean(v, indep_v)
                     bell_loss = adv_update_loss + adv_norm_loss
 
                     self.optimize_value(bell_loss, mean_loss / self._dt)

### 2018-11-10 pendulum_exp/stats.py

@@ -2,6 +2,7 @@
 from typing import Optional, Callable
 import numpy as np
 import torch
+from torch import Tensor
 
 class FloatingAvg:
     """ Computes floating average of a value. """
@@ -36,10 +37,13 @@ def epsilon_noise(arr: np.ndarray, epsilon: float) -> np.ndarray:
     mask = (np.random.uniform(0, 1, (arr.shape[0],)) < epsilon)
     return arr * (1 - mask) + (4 * np.random.randint(0, 2, (arr.shape[0],)) - 2) * mask
 
-def penalize_mean(arr: torch.Tensor):
+def penalize_mean_depr(arr: torch.Tensor):
     """ Introduce a mean penalization. """
     mask = (torch.zeros_like(arr).uniform_() >= .5)
     length = min(torch.sum(mask), torch.sum(1 - mask)).item()
     if length == 0:
         return torch.Tensor([0]).to(arr.device)
     return (arr[mask][:length] * arr[1 - mask][:length]).mean() / 2
+
+def penalize_mean(arr: torch.Tensor, indep_detach_arr: Tensor) -> Tensor:
+    return (arr * indep_detach_arr).mean()

### 2018-11-10 paper/advup_uniquesol.tex

@@
 \newcommand{\eqd}{=\mathrel{\mathop{:}}}
+\newcommand{\eps}{\varepsilon}
 
@@
 
+This also allows for computing $V^\pi$ while playing a regularized policy
+such as "$\pi$ with proba $1-\eps$ and exploration with proba $\eps$", by
+just training $V$ on transitions $(s,a,s')$ where $\pi$ was selected. The
+quantity $\eps$ does not have to be small and can be eg .9, as long as $1-\eps$ is not
+ridiculous.
+
 \paragraph{Sampling $\tilde s$ from a single trajectory.} Given a

### 2018-11-10 paper/advup_uniquesol.tex

@@
 
+To train $b$ it is important to either use a very small learning rate or
+experience replay from past transitions. Otherwise, a large learning rate
+for $b$ just results in $b$ being equal to the average of recent the
+Bellman gaps, thus erasing the training signal for $V$. One way to select
+transitions to train $b$ is as follows.
+
 \paragraph{Sampling $\tilde s$ from a single trajectory.} Given a

### 2018-11-10 paper/advup_uniquesol.tex

@@
 
-To train $b$ it is important to either use a very small learning rate or
-experience replay from past transitions. Otherwise, a large learning rate
+To train $b$ it is important to either use a very small learning rate, or
+experience replay from past transitions, or have several independent
+trajectories. Otherwise, a large learning rate
 for $b$ just results in $b$ being equal to the average of recent the
 Bellman gaps, thus erasing the training signal for $V$. One way to select
-transitions to train $b$ is as follows.
+transitions to train $b$ is as follows. ($b$ does not have to follow the
+same distribution as $s$, but it has to use the same policy... this is
+delicate of policy changes.)
 

### 2018-11-10 paper/advup_uniquesol.tex

--- 2018-11-10 paper/advup_uniquesol.tex
@@
 
+To train $b$ it is important to either use a very small learning rate, or
+experience replay from past transitions, or have several independent
+trajectories. Otherwise, a large learning rate
+for $b$ just results in $b$ being equal to the average of the recent
+Bellman gaps, thus erasing the training signal for $V$. One way to select
+transitions to train $b$ is as follows. ($b$ does not have to follow the
+same distribution for $s$, but it has to use the same policy for $a$
+given $s$... so reusing
+past transitions is
+delicate if policy changes.)
+
 \paragraph{Sampling $\tilde s$ from a single trajectory.} Given a

### 2018-11-10 paper/advup_uniquesol.tex

@@
 
+\paragraph{Removing $b$?} Let us just fix $b=0$ in the above. The unique
+solution for $V$ (with $\gamma<1$) is
+\begin{equation}
+V(s)=V^\pi(s)+\left(\frac{1-\gamma}{\alpha}-1\right)\E_{\tilde \rho} V^\pi
+\end{equation}
+and with $\alpha=1$ this is simply
+\begin{equation}
+V(s)=V^\pi(s)-\gamma \E_{\tilde \rho} V^\pi
+\end{equation}
+
+Let us check the behavior of this when $\gamma\to 1$. We know that
+\begin{equation}
+V^\pi(s)=\frac{\E_{\rho^\pi}r}{1-\gamma}+U^\pi(s)+O(1-\gamma)
+\end{equation}
+with $\rho^\pi$ the invariant distribution from $\pi$, and $U$ the
+relative value function of $\pi$. Thus, the fixed point for $V$ is
+\begin{equation}
+V(s)=U^\pi(s)-\gamma \E_{\tilde \rho}U^\pi +O(1-\gamma)
+\end{equation}
+which converges when $\gamma\to 1$???
+
 \end{document}

### 2018-11-10 paper/advup_uniquesol.tex

@@
 
+The TD step in that case is simply
+\begin{equation}
+\E_s \partial_\theta V(s) \left(
+r(s,a)+\gamma(V(s')-V(\tilde s))-V(s)\right)
+\end{equation}
+ie we just bring $V(s)$ closer to $V(s')-V(\tilde s)$. This is the
+simplest trick ever to remove constants...
+
+Would have to check what happens for training $A$, I think we have to
+take $\gamma(V(s')-V(\tilde s))-[V(s)-V(\tilde s)]$ if we want to remove
+constants on $A$.
+
 \end{document}

### 2018-11-15 pendulum_exp/benchmarkspolicies/dqn.py
dqn, and details

@@ -1,0 +1,105 @@
+from itertools import chain
+import torch
+from abstract import ParametricFunction, Arrayable, Noise, StateDict, Policy
+
+from convert import arr_to_th, th_to_arr
+from stats import penalize_mean
+from config import PolicyConfig
+from policies.shared import SharedAdvantagePolicy
+from mylog import log
+from logging import info
+
+
+class DiscreteDQNPolicy(Policy):
+    def __init__(self,
+                 qnet_function: ParametricFunction,
+                 target_function: ParametricFunction,
+                 qnet_noise: Noise,
+                 policy_config: PolicyConfig,
+                 device) -> None:
+        super().__init__(policy_config)
+        self._qnet_function = qnet_function.to(device)
+        self._target_function = target_function.to(device)
+        self._qnet_noise = qnet_noise
+
+        # optimization/storing
+        self._device = device
+        self._optimizer = torch.optim.SGD(self._qnet_function.parameters(), lr=policy_config.lr * policy_config.dt)
+        self._scheduler = torch.optim.lr_scheduler.LambdaLR(self._optimizer, policy_config.lr_decay)
+        # logging
+        self._cum_loss = 0
+        self._log = 100
+
+    def act(self, obs: Arrayable):
+        with torch.no_grad():
+            pre_action = self._adv_noise.perturb_output(
+                obs, function=self._qnet_function)
+            self._adv_noise.step()
+            return pre_action.argmax(axis=-1)
+
+    def learn(self):
+        if self._count % self._steps_btw_train == self._steps_btw_train - 1:
+            try:
+                for _ in range(self._learn_per_step):
+                    obs, action, next_obs, reward, done = self._sampler.sample()
+                    # for now, discrete actions
+                    indices = arr_to_th(action, self._device).long()
+
+                    q = self._qnet_function(obs).gather(1, indices.view(-1, 1)).squeeze()
+                    target = torch.max(self._target_function(next_obs), dim=1)[0].squeeze()
+
+                    exp_q = arr_to_th(reward, self._device) + self._gamma ** self._dt 
+                    if self._gamma == 1.:
+                        assert (1 - done).all(), "Gamma set to 1. with a potentially episodic problem..."
+                        exp_q += target.detach()
+                    else:
+                        done = arr_to_th(done.astype('float'), self._device)
+                        exp_q += self._gamma ** self._dt * target#.detach()
+
+                    loss = ((q - exp_q) ** 2).mean()
[…]

### 2018-11-15 pendulum_exp/envs/utils.py

@@ -198,6 +198,12 @@ def make_env(env_config: EnvConfig): # noqa: C901
         dt = env.dt
         info('The timestep dt is {}'.format(dt))
         # env.model.opt.timestep = dt / env.frame_skip
+    if env_id == 'ant':
+        # from gym.envs.mujoco import HalfCheetahEnv
+        env = gym.make('Ant-v2').unwrapped
+        dt = env.dt
+        info('The timestep dt is {}'.format(dt))
+        # env.model.opt.timestep = dt / env.frame_skip
     if env_id == 'pusher':
         env = DiscretePusherEnv()
         env.dt = dt

### 2018-11-15 pendulum_exp/main.py

@@ -41,7 +41,7 @@ def evaluate(
         time_limit: Optional[float]=None):
     """ Evaluate. """
     log_gap = int(.1 / dt)
-    video_log = 10
+    video_log = 1
     policy.eval()
 
     R = None

### 2018-11-15 pendulum_exp/monitoring.py

@@ -26,6 +26,9 @@ def plots(logs, namefile):
 
         ax.plot(x, y)
         ax.set_xlabel("timestamps")
+        ax.set_xlim(0., max(x))
+        l = len(x)
+        ax.set_ylim(min(y[l//10:]), max(y[l//10:]))
 
     plt.tight_layout()
     print(f'Plot: {namefile}')

### 2018-11-15 pendulum_exp/mylog.py

@@ -40,8 +40,8 @@ class Logger:
         info("logfile: {}".format(join(self._dir, 'logs.pkl')))
 
     def load(self):
-        assert self._file is not None
-        with open(self.filename, 'rb') as f:
+        assert self._dir is not None and isfile(join(self._dir, 'logs.pkl'))
+        with open(join(self._dir, 'logs.pkl'), 'rb') as f:
             self._logs = pickle.load(f)
 
 

### 2018-11-15 paper/advup_uniquesol.tex

@@
 \begin{equation}
-V(s)=U^\pi(s)-\gamma \E_{\tilde \rho}U^\pi +O(1-\gamma)
+V(s)=U^\pi(s)+\frac{1-\gamma-\alpha}{\alpha} \,\E_{\tilde \rho}U^\pi
++\frac1{\alpha} \E_{\rho^\pi}r+O(1-\gamma)
 \end{equation}
-which converges when $\gamma\to 1$???
+which converges to $V(s)=U^\pi(s)-\E_{\tilde \rho}U^\pi+\frac1{\alpha}
+\E_{\rho^\pi}r$ when $\gamma\to 1$.
 
-The TD step in that case is simply
+The TD step in that case, with $\alpha=1$, is simply
 \begin{equation}
@@
 \end{equation}
-ie we just bring $V(s)$ closer to $V(s')-V(\tilde s)$. This is the
-simplest trick ever to remove constants...
+ie we just bring $V(s)$ closer to $r+V(s')-V(\tilde s)$. This is the
+simplest trick ever to remove constants... Let's call this \emph{relative
+TD}. (TODO check if name is preoccupied.)
+
+(In particular, with $\gamma=1$ one
+has $\E_{\tilde \rho}V=\frac{1}{\alpha}\E_{\rho^\pi}r$. For instance, if $\tilde \rho$ is
+just a single reference
+state $\tilde s$ and $\alpha=1$ then 
+$V(\tilde s)=\E_{\rho^\pi}r$.)
 

### 2018-11-23 paper/advup_uniquesol.tex
Adding relative V terminal value

@@
 V(s)=V^\pi(s)+\left(\frac{1-\gamma}{\alpha}-1\right)\E_{\tilde \rho} V^\pi
+\label{eq:relative-td}
 \end{equation}
@@
 
+\paragraph{Relative $V$ terminal value in semi-episodic environments} When
+faced with a semi-episodic environment, i.e. an environment where terminal
+states exist, but some policies never reach them, the value of the relative
+$V$ function corresponding to a policy $\pi$ that do not reach terminal states,
+on a terminal state $t$ is given by
+\begin{equation}
+	V(t) = - \frac{\gamma}{1 - \gamma} \E_{\tilde{\rho}} V.
+\end{equation}
+The result is obtained from \eqref{eq:relative-td}, since $\E_{\tilde{\rho}} V = (1 - \gamma)
+\E_{\tilde{\rho}} V^\pi$, and $V^\pi(t) = 0$.
+
 \end{document}

### 2018-11-24 paper/tangotd.tex

\documentclass[11pt,a4paper]{article}
\input{header.en.tex}

\begin{document}

We consider the Bellman equation for a policy $\pi$ in a finite MDP with
expected reward $R$ and transition probability matrix $P$ ($P$ includes $\pi$ and
the environment response). The Bellman equation is
\begin{equation}
(\Id-P)V=R
\end{equation}

We assume we can sample transitions $(s,a,r,s')$ according to some
distribution $\rho$ on $s$. In the on-policy case, $\rho$ will be the
stationary distribution. Typically $a$ has to be taken from $\pi$ (or a
distribution close to $\pi$ after resampling/by using only those $a$
sampled from $\pi$; this covers $\eps$-greedy) so that we do get
transitions from
$P$. (Alternatively one could work with pairs $(s,a)$ TODO.) Thus, we
have access to unbiased samples of $\rho P$ and $\rho R$:
\begin{equation}
\rho P=\E[\1_{s}\transp{\1_{s'}}\,],\qquad \rho R=\E[r(s,a,s')\1_{s}]
\end{equation}

If we parameterize $V$ by some parameter $\theta$, we can apply
parametric TD. The expected TD step is
\begin{equation}
\theta\gets \theta -\eta \,\transp{\Phi}\rho (V-PV-R)
\end{equation}
where 
\begin{equation}
\Phi\deq \frac{\partial V}{\partial \theta}
\end{equation}
is the Jacobian matrix of $V$ with respect to its parameter. The linear
case $V=\Phi \theta$ corresponds to constant $\Phi$.

A fixed point of TD solves
\begin{equation}
\transp{\Phi}\rho(\Id-P)V=\transp{\Phi}\rho R
\end{equation}
and if the system is overparameterized ($\Phi$ invertible) and
well-sampled ($\rho>0$) this implies $(\Id-P)V=R$.

A local minimizer of the $L^2(\rho)$ error between $V$ and
$(\Id-P)^{-1}R$ solves
\begin{equation}
\transp{\Phi} \rho V=\transp{\Phi}\rho (\Id-P)^{-1} R
\end{equation}
which is not quite the same as the above. A local minimizer of the
$L^2(\rho)$ error between $(\Id-P)V$ and $R$ is
\begin{equation}
\transp{\Phi}\transp{(\Id-P)}\rho(\Id-P)V=\transp{\Phi}\transp{(\Id-P)}\rho R
\end{equation}
and in the linear case $V=\Phi\theta$, this is the standard solution to
the linear regression problem $(\Id-P)\Phi\theta\approx R$ in
$L^2(\rho)$.

The Dirichlet norm associated with $\rho$ is $\E_{ss'}
(f(s)-f(s'))^2$. Its matrix is $\rho+\rho'-\rho P-\transp{(\rho P)}$ where
$\rho'$ is the distribution of $s'$.
A local minimizer of the Dirichlet error between $V$ and
$(\Id-P)^{-1}R$ solves
\begin{equation}
\transp{\Phi} (\rho+\rho'-\rho P-\transp{(\rho P)})(V-(\Id-P)^{-1}R)=0
\end{equation}
If $\rho$ is the stationary distribution $\mu$ and $P$ is reversible, one
has $\transp{(\mu P)}=\mu P$ so $\rho+\rho'-\rho P-\transp{(\rho
P)}=2\mu(\Id-P)$ and this is equivalent to
\begin{equation}
\transp{\Phi}\mu ((\Id-P)V-R)=0
\end{equation}
namely, the same as a fixed point of TD.

\end{document}

### 2018-11-24 paper/tangotd.tex

@@
 and if the system is overparameterized ($\Phi$ invertible) and
-well-sampled ($\rho>0$) this implies $(\Id-P)V=R$.
+well-sampled ($\rho>0$) this implies $(\Id-P)V=R$. All terms in this
+equation are \emph{samplable}: we can sample from $\rho R$ and from
+$\rho(\Id-P)$.
 
@@
 
+\bigskip
+
+TD is the only one of those fixed point equations that is
+\emph{samplable} given that we can sample $\rho$, $\rho P$ and $\rho R$.
+\footnote{Actually the minimizer of the $L^2(\rho^2)$ error between $R$
+and $(\Id-P)V$ is also samplable:
+\begin{equation}
+\transp{\Phi}\transp{(\Id-P)}\rho^2(\Id-P)V=\transp{\Phi}\transp{(\Id-P)}\rho^2
+R
+\end{equation}
+however, the $\rho^2$ is a bit weird...
+}
+Thus, from now on we are going to focus on solving
+\begin{equation}
+\transp{\Phi}\rho (\Id-P)V=\transp{\Phi}\rho R
+\end{equation}
+using various SGD-like solvers.
+
+We now switch to the linear case and will de-linearize later. This
+becomes
+\begin{equation}
+\transp{\Phi}\rho (\Id-P)\Phi \theta=\transp{\Phi}\rho R
+\end{equation}
+to be solved in $\theta$.
+
+\paragraph{Solving $A\theta=b$ where $A$ and $b$ are samplable.} An
+obvious way to solve $A\theta=b$ via a stochastic gradient method is
+\begin{equation}
+\theta\gets \theta - \eta (A\theta-b)
+\end{equation}
+using samples for $A$ and $b$.
+This is TD. If $A$ is symmetric positive definite, this is also the
+gradient descent of the loss function
+$\transp{(A\theta-b)}A^{-1}(A\theta-b)$. However, we want to apply this
+to matrices $A$ that are not necessarily of this type.
+
+This works only if $A$ is \emph{stable}.
+
+\begin{defi}
+A matrix $A$ is \emph{stable} if one of the following equivalent
+conditions are satisfied:
+\begin{itemize}
+\item All the eigenvalues of $A$ have positive real part.
+\item The matrix $(\Id-\eps A)$ has spectral radius $<1$ for small enough
+$\eps$.
+\item The differential equation $\theta'=-A\theta$ converges to $0$ for
+any initial value.
+\item There exists a symmetric, positive definite matrix $L$ (Lyapunov
+function) such that $\transp{\theta}L\theta$ is decreasing along the
+solutions of the differential equation $\theta'=-A\theta$.
+\item There exists a symmetric, positive definite matrix $L$ such that
+\begin{equation}
+\transp{\theta} L A \theta>0
+\end{equation}
+for all $\theta\neq 0$. (This is the same $L$ as in the previous
+condition.)
+\end{itemize}
+\end{defi}
+
+In particular, a symmetric, definite positive matrix is stable. Since the
+solution of $\theta_t'=-A\theta_t$ is $\theta_t=e^{-tA}\theta_0$, a
+Lyapunov function that works is $\transp{\theta}L\theta=\int_{t \geq 0}
+\norm{\theta_t}^2$, namely
+$L=\int_{t\geq 0} \transp{(e^{-t A})}
+e^{-tA}$.
+
 \end{document}

### 2018-11-24 paper/tangotd.tex

@@
 is the Jacobian matrix of $V$ with respect to its parameter. The linear
-case $V=\Phi \theta$ corresponds to constant $\Phi$.
+case $V=\Phi \theta$ corresponds to constant $\Phi$. The tabular case is
+$\Phi=\Id$.
 
@@
 
-In particular, a symmetric, definite positive matrix is stable. Since the
+In particular, a symmetric, definite positive matrix is stable (last
+criterion with $L=\Id$).
+Stability is invariant by matrix similarity $A\gets B^{-1}A B$, since
+this preserves eigenvalues.
+
+Since the
 solution of $\theta_t'=-A\theta_t$ is $\theta_t=e^{-tA}\theta_0$, a
@@
 
+\paragraph{Making the TD iteration stable.} In linear TD the matrix $A$ is
+\begin{equation}
+A=\transp{\Phi}\rho (\Id-P)\Phi
+\end{equation}
+and we know it is stable in several cases: (TODO here I just remove
+constants)
+\begin{enumerate}
+\item On-policy: $\rho=\mu$, the invariant distribution. Then we know that $\rho(\Id-P)$ is positive
+(Dirichlet form). This implies that $\transp{\Phi}\rho (\Id-P)\Phi$ is
+positive too. So we recover the convergence result on-policy for linear
+TD.
+\item In the tabular case: $\Phi=\Id$. Indeed we have $A=\rho (\Id-P)$,
+which is not positive in general, but since $\mu (\Id-P)$ is, we can take
+$L=\mu/\rho$ so that $LA$ is positive.
+\end{enumerate}
+
+One can also \emph{precondition} the gradient descent. This can make it
+stable and/or make convergence faster. The step becomes
+\begin{equation}
+\theta\gets \theta-\eta B(A\theta-b)
+\end{equation}
+for some matrix $B$.
+
+Of course the values of $B$ that are most interesting are obtained as the
+inverse of something; for instance $B=A^{-1}$ solves the problem
+instantly... Two-timescale methods or TANGO-like algorithms can apply an
+inverse of something to $A\theta-b$, via an auxiliary gradient descent.
+TANGO works for stable matrices, if they are samplable \footnote{In
+TANGO, the consecutive steps absolutely have to be \emph{independent}. So even in
+the on-policy case, we will need a replay buffer from $\mu$ for the auxiliary
+gradient descent.}
+
+A few choices are:
+\begin{itemize}
+\item $B=A^{-1}$: obviously stable, jumps to the solution in one step if
+$\eta=1$. Requires computing the inverse of $A$, namely, solving the
+problem.
+\end{itemize}
+
 \end{document}

### 2018-11-24 pendulum_exp/monitoring.py
Monitoring floating average

@@ -6,7 +6,6 @@ from subprocess import call
 import matplotlib.pyplot as plt
 
 
-
 def plots(logs, namefile):
     fig, axes = plt.subplots(len(logs), figsize=(10., 15.))
 
@@ -19,7 +18,7 @@ def plots(logs, namefile):
         ax.plot(x, y)
         ax.set_xlabel("timestamps")
         ax.set_xlim(0., max(x))
-        l = len(x)
+        l = len(x) # flake8: noqa
         ax.set_ylim(min(y[l//10:]), max(y[l//10:]))
 
     plt.tight_layout()
@@ -29,7 +28,7 @@ def plots(logs, namefile):
 
 def summary(logs):
     print('Summary')
-    metrics = [("Avg_adv_loss", min), ("Return", max)]
+    metrics = [("Avg_adv_loss", min), ("Float_adv_loss", min), ("Return", max)]
     for met, f in metrics:
         metval = logs[met]
         curr_ts, curr_val = max(metval.items(), key=lambda x: x[0])

### 2018-11-24 pendulum_exp/policies/continuous.py

@@ -1,4 +1,5 @@
 """Define continuous policy."""
+from typing import Optional
 import torch
 from torch import Tensor
 import numpy as np
@@ -46,6 +47,7 @@ class SampledAdvantagePolicy(SharedAdvantagePolicy):
 
         # logging
         self._cum_loss = 0
+        self._float_loss: Optional[float] = None
 
     def act(self, obs: Arrayable):
         with torch.no_grad():
@@ -84,6 +86,11 @@ class SampledAdvantagePolicy(SharedAdvantagePolicy):
 
         # logging
         self._cum_loss += losses[0].item()
+        if self._float_loss is None:
+            self._float_loss = losses[0].item()
+        else:
+            self._float_loss = self._float_loss * self._gamma ** self._dt +\
+                (1 - self._gamma ** self._dt) * losses[0].item()
         self._learn_count += 1
 
     def optimize_policy(self, max_adv: Tensor):
@@ -92,8 +99,11 @@ class SampledAdvantagePolicy(SharedAdvantagePolicy):
     def log(self):
         log("Avg_adv_loss", self._cum_loss / self._learn_count,
             self._learn_count)
+        log("Float_adv_loss", self._float_loss, self._learn_count)
         info(f"At iteration {self._learn_count}, "
-             f"avg_loss: {self._cum_loss/self._learn_count}")
+             f'Float_adv_loss: {self._float_loss}, '
+             f'Avg_adv_loss: {self._cum_loss/self._learn_count}, '
+             f'Avg_policy_loss: {self._cum_policy_loss / self._learn_count}')
 
     def train(self):
         self._train = True
@@ -176,6 +186,7 @@ class AdvantagePolicy(SharedAdvantagePolicy):
 
         # logging
         self._cum_loss = 0
+        self._float_loss = None
         self._cum_policy_loss = 0
 
     def act(self, obs: Arrayable):
@@ -199,6 +210,11 @@ class AdvantagePolicy(SharedAdvantagePolicy):
 
         # logging
         self._cum_loss += losses[0].item()
+        if self._float_loss is None:
+            self._float_loss = losses[0].item()
+        else:
+            self._float_loss = self._float_loss * self._gamma ** self._dt +\
+                (1 - self._gamma ** self._dt) * losses[0].item()
         self._learn_count += 1
 
     def optimize_policy(self, max_adv: Tensor):
@@ -212,8 +228,10 @@ class AdvantagePolicy(SharedAdvantagePolicy):
 
[…]

### 2018-11-24 pendulum_exp/policies/discrete.py

@@ -1,4 +1,5 @@
 """Define policies."""
+from typing import Optional
 import torch
 from torch import Tensor
 from abstract import ParametricFunction, Arrayable, Noise, StateDict
@@ -42,6 +43,7 @@ class AdvantagePolicy(SharedAdvantagePolicy):
 
         # logging
         self._cum_loss = 0
+        self._float_loss: Optional[float] = None
         self._log = 100
 
     def act(self, obs: Arrayable):
@@ -68,6 +70,11 @@ class AdvantagePolicy(SharedAdvantagePolicy):
 
         # logging
         self._cum_loss += losses[0].item()
+        if self._float_loss is None:
+            self._float_loss = losses[0].item()
+        else:
+            self._float_loss = self._float_loss * self._gamma ** self._dt +\
+                (1 - self._gamma ** self._dt) * losses[0].item()
         self._learn_count += 1
 
     def optimize_policy(self, max_adv: Tensor):
@@ -79,7 +86,9 @@ class AdvantagePolicy(SharedAdvantagePolicy):
 
     def log(self):
         log("Avg_adv_loss", self._cum_loss / self._learn_count, self._learn_count)
+        log("Float_adv_loss", self._float_loss, self._learn_count)
         info(f"At iteration {self._learn_count}, "
+             f'Float_adv_loss: {self._float_loss}, '
              f"avg_loss: {self._cum_loss/self._learn_count}")
 
     def train(self):

### 2018-11-24 paper/tangotd.tex

@@
 $\eta=1$. Requires computing the inverse of $A$, namely, solving the
-problem.
+problem. But if $A$ is stable and samplable, this can be done via TANGO.
+
+\item $B=(\transp{\Phi}\rho\Phi)^{-1}$. Then we have to prove that
+$(\transp{\Phi}\Phi)^{-1}\transp{\Phi}\rho(\Id-P)\Phi$ is stable. Taking
+$L=\transp{\Phi}(\mu/\rho)\Phi$, and computing
+\begin{align}
+\transp{\theta}LBA\theta&=
+\transp{\theta}L(\transp{\Phi}\rho\Phi)^{-1}\transp{\Phi}\rho(\Id-P)\Phi\theta
+\end{align}
+and note that $(\transp{\Phi}\rho\Phi)^{-1}\transp{\Phi}\rho$ is the
+solution of linear regression with design matrix $\Phi$ in $L^2(\rho)$.
+This implies that
+\begin{equation}
+\Phi(\transp{\Phi}\rho\Phi)^{-1}\transp{\Phi}\rho=\Pi_\Phi
+\end{equation}
+where $\Pi$ is the orthogonal projector on the image of $\Phi$ in
+$L^2(\rho)$. Now taking
+\begin{equation}
+L=\transp{\Phi}(\mu/\rho)\Phi
+\end{equation}
+we have
+\begin{equation}
+\transp{\theta}LBA\theta=\transp{\theta} \transp{\Phi}(\mu/\rho)\Pi_\Phi
+(\Id-P)\Phi\theta
+\end{equation}
 \end{itemize}

### 2018-11-24 paper/tangotd.tex

--- 2018-11-24 paper/tangotd.tex
@@
 \begin{equation}
-A=\transp{\Phi}\rho (\Id-P)\Phi
+A\deq \transp{\Phi}\rho (\Id-P)\Phi
 \end{equation}
-and we know it is stable in several cases: (TODO here I just remove
-constants)
+and we want to solve
+\begin{equation}
+A\theta=b\,\qquad b\deq \transp{\Phi}\rho R
+\end{equation}
+If $A$ is stable, then iterating $\theta\gets \theta -\eta(A\theta-b)$ will
+converge.
+
+We know $A$ is stable in several cases: (TODO here I just ignore constant
+functions)
 \begin{enumerate}
@@
 \begin{itemize}
-\item $B=A^{-1}$: obviously stable, jumps to the solution in one step if
+\item GTD is a gradient descent of
+$\transp{(A\theta-b)}(A\theta-b)$ (Sutton 2009).
+This corresponds to $B=\transp{A}$. Then
+$BA=\transp{A}A$ is
+obviously stable. This helps with $A$ being non-symmetric.
+
+\item $B=A^{-1}$: obviously stable ($BA=\Id$), jumps to the solution in one step if
 $\eta=1$. Requires computing the inverse of $A$, namely, solving the
-problem.
+problem by first solving the problem. But if $A$ is stable and samplable,
+this can be done via TANGO. More on that later.
+
+\item $B=(\transp{\Phi}\rho\Phi)^{-1}$. This amounts to orthonormalizing
+the basis before applying TD. In general, this is not stable off-policy: take the
+standard two-point counterexample with a single $\Phi=\transp{(1,2)}$,
+orthonormalizing $\Phi$ is not going to help as $\transp{\Phi}\rho\Phi$
+is just a number.
+
+Still, this is stable
+on-policy or assuming overparameterization. (Indeed in the latter case we
+are back to the tabular case.) TODO
+% Then we have to prove th
+% $(\transp{\Phi}\Phi)^{-1}\transp{\Phi}\rho(\Id-P)\Phi$ is stable. Taking
+% $L=\transp{\Phi}(\mu/\rho)\Phi$, and computing
+% \begin{align}
+% \transp{\theta}LBA\theta&=
+% \transp{\theta}L(\transp{\Phi}\rho\Phi)^{-1}\transp{\Phi}\rho(\Id-P)\Phi\theta
+% \end{align}
+% and note that $(\transp{\Phi}\rho\Phi)^{-1}\transp{\Phi}\rho$ is the
+% solution of linear regression with design matrix $\Phi$ in $L^2(\rho)$.
+% This implies that
+% \begin{equation}
+% \Phi(\transp{\Phi}\rho\Phi)^{-1}\transp{\Phi}\rho=\Pi_\Phi
+% \end{equation}
+% where $\Pi$ is the orthogonal projector on the image of $\Phi$ in
+% $L^2(\rho)$. Now taking
+% \begin{equation}
+% L=\transp{\Phi}(\mu/\rho)\Phi
+% \end{equation}
+% we have
+% \begin{equation}
+% \transp{\theta}LBA\theta=\transp{\theta} \transp{\Phi}(\mu/\rho)\Pi_\Phi
+% (\Id-P)\Phi\theta
+% \end{equation}
+% TODO: on-policy or overparameterized
+
+\item The choice from Sutton 2009 "TDC" corresponds to a gradient descent
+of the projected norm of the Bellman gap on the subspace generated by
+$\Phi$, projected in $L^2(\rho)$. This is GTD with orthonormalization of
+$\Phi$, namely,
+\begin{equation}
+B=\transp{A} (\transp{\Phi}\rho\Phi)^{-1}=\transp{(\transp{\Phi}\rho(\Id-P)\Phi)}(\transp{\Phi}\rho\Phi)^{-1}
+\end{equation}
+from which it is clear that $BA$ is symmetric and stable. This is obtained as a
+gradient descent of the projected norm of the Bellman gap on the image of
+$\Phi$, projected in $L^2(\rho)$.
 \end{itemize}

### 2018-11-24 paper/tangotd.tex

@@
 \item In the tabular case: $\Phi=\Id$. Indeed we have $A=\rho (\Id-P)$,
-which is not positive in general, but since $\mu (\Id-P)$ is, we can take
+which is not positive in general. But $\mu (\Id-P)$ is positive, so we can take
 $L=\mu/\rho$ so that $LA$ is positive.
@@
 \end{equation}
-for some matrix $B$.
+for some matrix $B$. Some choices of $B$ correspond to natural choices of
+loss functions on $\theta$, as we will see.
 
@@
 of the projected norm of the Bellman gap on the subspace generated by
-$\Phi$, projected in $L^2(\rho)$. This is GTD with orthonormalization of
+$\Phi$, projected in $L^2(\rho)$. This is GTD with $L^2(\rho)$-orthonormalization of
 $\Phi$, namely,
@@
 $\Phi$, projected in $L^2(\rho)$.
+
+\item The choice $B=A^{-1}$ can also be written as
+$B=\transp{A}(A\transp{A})^{-1}$. The advantage is that we now have to
+invert a stable matrix, so, better behaved with TANGO.
 \end{itemize}

### 2018-11-24 paper/tangotd.tex

@@
 solution of $\theta_t'=-A\theta_t$ is $\theta_t=e^{-tA}\theta_0$, a
-Lyapunov function that works is $\transp{\theta}L\theta=\int_{t \geq 0}
-\norm{\theta_t}^2$, namely
+Lyapunov function that works is $\transp{\theta_0}L\theta_0\deq \int_{t \geq 0}
+\norm{\theta_t}^2$. Indeed this is decreasing, because
+$\transp{\theta_t}L\theta_t$ is just the same integral starting at $t$
+instead of $0$. Explicitly this is
 $L=\int_{t\geq 0} \transp{(e^{-t A})}
@@
 problem by first solving the problem. But if $A$ is stable and samplable,
-this can be done via TANGO. More on that later.
+this can be done via TANGO. More on that later. This should speed up
+things in situations where $A$ is stable in the first place (eg.,
+on-policy).
 
@@
 $B=\transp{A}(A\transp{A})^{-1}$. The advantage is that we now have to
-invert a stable matrix, so, better behaved with TANGO.
+invert a stable matrix, so, better behaved with TANGO. However, variance
+will be terrible: this requires four independent samples from $A$.
 \end{itemize}

### 2018-11-25 paper/tangotd.tex

@@
 \item On-policy: $\rho=\mu$, the invariant distribution. Then we know that $\rho(\Id-P)$ is positive
-(Dirichlet form). This implies that $\transp{\Phi}\rho (\Id-P)\Phi$ is
+(Dirichlet form). This implies that $A=\transp{\Phi}\rho (\Id-P)\Phi$ is
 positive too. So we recover the convergence result on-policy for linear
@@
 Still, this is stable
-on-policy or assuming overparameterization. (Indeed in the latter case we
-are back to the tabular case.) TODO
+on-policy (take $L=B^{-1}$, given that $A$ is stable on-policy) or assuming overparameterization. (Indeed in the latter case we
+are back to the tabular case.)
 % Then we have to prove th
@@
 will be terrible: this requires four independent samples from $A$.
+
+\item A gradient descent of
+$\transp{(A\theta-b)}(A+\transp{A})^{-1}(A\theta-b)$, corresponding to
+$B=\transp{A}(A+\transp{A})^{-1}$. This is probably the closest to
+unmodified TD in behavior: for instance when $A$ is symmetric ($\rho=\mu$
+and $P$ reversible) this coincides with TD.
+
 \end{itemize}

### 2018-11-25 paper/tangotd.tex

@@
 
+Only $B=A^{-1}$ and $B=\transp{A}(A\transp{A})^{-1}$ are truly
+second-order (namely, solve the problem in one step in the non-stochastic
+case). The other algorithms involving an inverse basically have the cost
+of second-order algorithms without the full benefits. They are
+second-order in feature-space (they orthonormalize $\Phi$) but not in
+Markov process space (they don't backpropagate rewards).
+
 \end{document}

### 2018-11-25 paper/tangotd.tex

@@
 \end{equation}
+and thus also to samples of $\rho(\Id-P)$. Thus, rewriting the equation
+as
+\begin{equation}
+\rho(\Id-P)V=\rho R
+\end{equation}
+(equivalent if $\rho$ is positive), both sides are samplable, and we can try to solve the above by
+stochastic gradient-like methods.
 
-If we parameterize $V$ by some parameter $\theta$, we can apply
+For instance, ff we parameterize $V$ by some parameter $\theta$, we can apply
 parametric TD. The expected TD step is
@@
 the linear regression problem $(\Id-P)\Phi\theta\approx R$ in
-$L^2(\rho)$.
+$L^2(\rho)$. However, the matrix on the left is not samplable.
 
@@
 second-order (namely, solve the problem in one step in the non-stochastic
-case). The other algorithms involving an inverse basically have the cost
-of second-order algorithms without the full benefits. They are
-second-order in feature-space (they orthonormalize $\Phi$) but not in
-Markov process space (they don't backpropagate rewards).
+case).
+
+The other algorithms involving an inverse basically have the cost of
+second-order algorithms without the full benefits.  They are second-order
+in feature-space (they orthonormalize $\Phi$) but not in Markov process
+space (they don't backpropagate rewards).
+
+TODO sampling: if $A=\E \tilde A$, can do $\transp{(\E \tilde A)}(\E
+\tilde A)$ or
+$\E(\transp{\tilde A}\tilde A)$, which is not equal to $\transp{A}A$ but
+is still definite positive (hence stable). This would avoid some
+quadruple-sampling problems with $B=\transp{A}(A\transp{A})^{-1}$. But
+breaks second-order correctness?
+
+TODO: define second-order with respect to $\Phi$ (second-order
+provided $\transp{\Delta}\Delta=\Id$, eg in a Fourier frequency-adjusted
+basis), with respect to the Markov chain (second-order provided
+$\transp{\Phi}\Phi=\Id$).
 

### 2018-11-25 paper/tangotd.tex

@@
 unmodified TD in behavior: for instance when $A$ is symmetric ($\rho=\mu$
-and $P$ reversible) this coincides with TD.
+and $P$ reversible) this coincides with TD. $A+\transp{A}$ corresponds to
+the Dirichlet form for $\rho$, namely, $\E_{\rho} (f(s)-f(s'))^2$ TODO
+CHECK.
 
@@
 
+One-and-a-half-timescale needs more independence than two-timescale. I
+\emph{think} that two-timescale can deal with taking the same sample for
+$A$ and $\transp{A}$ in $\transp{A}(...)^{-1}A$, and taking correlated
+consecutive samples (eg, online) for the matrix to be inverted. I think
+TANGO would need uncorrelated samples for the matrix to be inverted (eg,
+a replay buffer) and two samples for $A$ and $\transp{A}$, but not sure
+about the latter.
+
 \end{document}

### 2018-11-25 paper/tangotd.tex

@@
 \end{equation}
-(equivalent if $\rho$ is positive), both sides are samplable, and we can try to solve the above by
+(equivalent if $\rho$ is positive), both sides are \emph{samplable}, and we can try to solve the above by
 stochastic gradient-like methods.
 
-For instance, ff we parameterize $V$ by some parameter $\theta$, we can apply
+For instance, if we parameterize $V$ by some parameter $\theta$, we can apply
 parametric TD. The expected TD step is
@@
 \end{equation}
-however, the $\rho^2$ is a bit weird...
+however, the $\rho^2$ is a bit weird... but some of the algos we consider
+below actually amount to this.
 }

### 2018-11-25 paper/tangotd.tex

@@
 $BA=\transp{A}A$ is
-obviously stable. This helps with $A$ being non-symmetric.
+obviously stable. This helps with $A$ being non-symmetric. TODO: check if
+GTD uses the same sample for $\tilde{A}$ and $A$. If so, it will be
+faster but converge to a different stable point (still OK if
+overparameterized).
 
@@
 gradient descent of the projected norm of the Bellman gap on the image of
-$\Phi$, projected in $L^2(\rho)$.
+$\Phi$, projected in $L^2(\rho)$. TODO: TDC and GTD2 differ by using that
+$A$ has $(\Id-P)$ on  the right, and expanding the $\Id$ term and the $P$
+term separately, which results in simplifications and make TDC much
+closer to TD and less reliant on the inversion.
 
@@
 
+TODO: I'm more interested in an algo that's second-order wrt the Markov
+process (I don't necessarily want to orthonormalize features).
+
 \end{document}

### 2018-11-25 paper/tangotd.tex

@@
 obviously stable. This helps with $A$ being non-symmetric. TODO: check if
-GTD uses the same sample for $\tilde{A}$ and $A$. If so, it will be
-faster but converge to a different stable point (still OK if
-overparameterized).
+GTD uses the same sample for $\tilde{A}$ and $A$. If so, it will
+converge to a different stable point. TODO: Can be applied TANGO-style
+(by formally inverting $\Id$ in TANGO), advantage=fewer variance due to
+two-sampling. (TANGO uses two-sampling but on a sum of past terms, so
+that a past term gets TANGOified whenever we happen to reach a state that
+activates the same features in the future, which is probably fine). Can
+also be TANGOified with $\gamma\to 0$ (two-timescale) which now requires
+only single sampling (no replay buffers).
 

### 2018-11-25 paper/tangotd.tex

@@
 (by formally inverting $\Id$ in TANGO), advantage=fewer variance due to
-two-sampling. (TANGO uses two-sampling but on a sum of past terms, so
+two-sampling (Is this equivalent to eligibility traces without the
+variance??). (TANGO uses two-sampling but on a sum of past terms, so
 that a past term gets TANGOified whenever we happen to reach a state that
@@
 
+TODO: in the original TANGO paper: prove that the main learning rate
+doesn't have to be too small even for non-quadratic problems, namely,
+even if the Hessian changes, we're still applying a positive operator to
+earlier gradients so we still have a valid gradient direction.
+
 \end{document}

### 2018-11-25 paper/tangotd.tex

@@
 If $\rho$ is the stationary distribution $\mu$ and $P$ is reversible, one
-has $\transp{(\mu P)}=\mu P$ so $\rho+\rho'-\rho P-\transp{(\rho
+has $\transp{(\mu P)}=\mu P$ and $\rho'=\mu$ so $\rho+\rho'-\rho P-\transp{(\rho
 P)}=2\mu(\Id-P)$ and this is equivalent to
@@
 \end{equation}
-however, the $\rho^2$ is a bit weird... but some of the algos we consider
-below actually amount to this.
+however, the $\rho^2$ is a bit weird... moreover having
+$\transp{\Delta}\Delta$ means we are just facing the double sampling
+problem.
 }
@@
 
-One-and-a-half-timescale needs more independence than two-timescale. I
+Almost-one-timescale (such as TANGO) needs more independence than two-timescale. I
 \emph{think} that two-timescale can deal with taking the same sample for

### 2018-11-26 paper/tangotd.tex

@@
 
+Instead of applying the Dirichlet norm to $(V-(\Id-P)^{-1}R)$, one can
+apply the inverse Dirichlet norm to $((\Id-P)V-R)$. When When $P$ is
+reversible and $\rho=\mu$ this
+is the same. One tries to minimize
+$\transp{((\Id-P)V-R)}\rho\mathcal{D}^{-1}\rho((\Id-P)V-R)$ where
+$\mathcal{D}\deq \rho+\rho'-\rho P-\transp{(\rho P)}$ as above, and use
+TANGO to invert $\mathcal{D}$. However this suffers from a double
+sampling problem.
+
 \bigskip
@@
 $\transp{\Delta}\Delta$ means we are just facing the double sampling
-problem.
+problem. Even the right-hand-side involves double sampling.
+
+The minimizer of a weird norm $L^2(\rho \Phi\transp{\Phi}\rho)$ is better
+samplable, because now it's enough to sample states \emph{with the same
+features}, not just the exact same state. Use unclear. Should use a hash
+of states instead.
 }

### 2018-11-26 paper/tangotd.tex

@@
 
+\paragraph{Minimizers of errors.}
 A local minimizer of the $L^2(\rho)$ error between $V$ and
@@
 $L^2(\rho)$. However, the matrix on the left is not samplable.
+
+The minimizer of the $L^2(\rho^2)$ error between $R$
+and $(\Id-P)V$ is samplable:
+\begin{equation}
+\transp{\Phi}\transp{(\Id-P)}\rho^2(\Id-P)V=\transp{\Phi}\transp{(\Id-P)}\rho^2
+R
+\end{equation}
+however, the $\rho^2$ is a bit weird... moreover having
+$\transp{\Delta}\Delta$ means we are just facing the double sampling
+problem (we can sample $\transp{(\rho\Delta)}\rho\Delta$, but in practice the sample will always be $0$
+unless we are lucky and select two transitions $s\to s'$ that have the
+same $s$). Even the right-hand-side involves double sampling.
+
+The minimizer of a weird norm $L^2(\rho \Phi\transp{\Phi}\rho)$ is better
+samplable, because now it's enough to sample states \emph{with the same
+features}, not just the exact same state. This is basically what happens
+in the linear case. The features should not be orthogonal. Should use a
+hash of states instead?
 
@@
 \emph{samplable} given that we can sample $\rho$, $\rho P$ and $\rho R$.
-\footnote{Actually the minimizer of the $L^2(\rho^2)$ error between $R$
-and $(\Id-P)V$ is also samplable:
-\begin{equation}
-\transp{\Phi}\transp{(\Id-P)}\rho^2(\Id-P)V=\transp{\Phi}\transp{(\Id-P)}\rho^2
-R
-\end{equation}
-however, the $\rho^2$ is a bit weird... moreover having
-$\transp{\Delta}\Delta$ means we are just facing the double sampling
-problem. Even the right-hand-side involves double sampling.
-
-The minimizer of a weird norm $L^2(\rho \Phi\transp{\Phi}\rho)$ is better
-samplable, because now it's enough to sample states \emph{with the same
-features}, not just the exact same state. Use unclear. Should use a hash
-of states instead.
-}
 Thus, from now on we are going to focus on solving

### 2018-11-26 paper/tangotd.tex

@@
 \input{header.en.tex}
+
+\newcommand{\Dir}{\mathcal{D}}
 
@@
 in the linear case. The features should not be orthogonal. Should use a
-hash of states instead?
+hash of states instead? Or apply ReLU to orthogonal features.
 
@@
 Instead of applying the Dirichlet norm to $(V-(\Id-P)^{-1}R)$, one can
-apply the inverse Dirichlet norm to $((\Id-P)V-R)$. When When $P$ is
+apply the inverse Dirichlet norm to $((\Id-P)V-R)$. When $P$ is
 reversible and $\rho=\mu$ this
 is the same. One tries to minimize
-$\transp{((\Id-P)V-R)}\rho\mathcal{D}^{-1}\rho((\Id-P)V-R)$ where
-$\mathcal{D}\deq \rho+\rho'-\rho P-\transp{(\rho P)}$ as above, and use
-TANGO to invert $\mathcal{D}$. However this suffers from a double
-sampling problem.
+$\transp{((\Id-P)V-R)}\rho\Dir^{-1}\rho((\Id-P)V-R)$ where
+$\Dir\deq \rho+\rho'-\rho P-\transp{(\rho P)}$ as above, and use
+TANGO to invert $\Dir$. However this suffers again from a double
+sampling problem, in several places. Still, this can be rewritten as a
+correction to TD where only the correction suffers from double
+sampling/large variance. Namely, the gradient of the above with respect
+to $\theta$ is $\transp{\Phi}\transp{(\rho\Delta)} \Dir^{-1}\rho(\Delta V-R)$. Since
+$\rho\Delta$ is half of $\Dir$, this gradient is 
+\begin{gather}
+\transp{\Phi}\transp{(\rho\Delta)} \Dir^{-1}\rho(\Delta V-R)=
+\\
+\frac12\transp{\Phi}(\transp{(\rho \Delta)}+\rho'-\rho P)\Dir^{-1}\rho(\Delta
+V-R)+\frac12 \transp{\Phi}(\transp{(\rho \Delta)}+\rho P-\rho')\Dir^{-1}\rho(\Delta
+V-R)
+\\=\frac12\transp{\Phi}\rho(\Delta
+V-R)+\frac12 \transp{\Phi}(\transp{(\rho \Delta)}+\rho P-\rho')\Dir^{-1}\rho(\Delta
+V-R)
+\end{gather}
+the first part of which is TD. Thus one can write a correction to TD such
+that only the correction suffers from double-sampling. The correction vanishes for reversible
+chains. The double sampling can be simulated with high variance in finite
+state spaces (just take samples from $P$ and $\transp{P}$ until they
+match) but does not make sense in continuous state spaces. Use a model of
+transitions only for the correction? TODO: use the same trick on the
+right, double sampling becomes only on a term that is quadratic in the
+non-reversibility.
 

### 2018-11-26 pendulum_exp/mylog.py
Tensorboard logging

@@ -1,4 +1,5 @@
 """Log facilities"""
+from abc import ABC, abstractmethod
 from os import makedirs
 from os import remove
 from shutil import rmtree
@@ -9,17 +10,27 @@ import numpy as np
 
 from logging import info
 
-class Logger:
-    CURRENT = None # current logger
+class KVTWritter(ABC):
+    @abstractmethod
+    def writekvts(self, key: str, value: float, timestamp: int):
+        pass
 
+    @abstractmethod
+    def set_dir(self, logdir: str, reload: bool=False):
+        pass
+
+    @abstractmethod
+    def load(self):
+        pass
+
+class PickleKVTWriter(KVTWritter):
     def __init__(self):
-        assert Logger.CURRENT is None
         self._logs: Dict[str, Dict[int, float]] = dict()
         self._buffering = 500
         self._count = 0
         self._dir = None
 
-    def log(self, key: str, value: float, timestamp: int):
+    def writekvts(self, key: str, value: float, timestamp: int):
         if key not in self._logs:
             self._logs[key] = {timestamp: value}
         else:
@@ -28,12 +39,6 @@ class Logger:
         if self._count == self._buffering:
             self._count = 0
             self.dump()
-
-    def log_video(self, tag: str, timestamp: int, frames):
-        video_dir = join(self._dir, "videos")
-        if not exists(video_dir):
-            makedirs(video_dir)
-        np.savez(join(video_dir, f"{tag}_{timestamp}.npz"), frames)
 
     def set_dir(self, logdir: str, reload: bool=True):
         self._dir = logdir
@@ -55,6 +60,50 @@ class Logger:
     def dump(self):
         assert self._dir is not None
         pickle.dump(self._logs, open(join(self._dir, 'logs.pkl'), 'wb'))
+
+class TensorboardKVTWriter(KVTWritter):
+    def __init__(self):
+        self._writer = None
+        self._dir = None
+
+    def set_dir(self, logdir: str, reload: bool=True):
+        self._dir = join(logdir, 'train')
+        try:
+            makedirs(self._dir)
+        except OSError:
+            if not reload:
+                rmtree(self._dir, ignore_errors=True)
+                makedirs(self._dir)
+        self.load()
+
+    def load(self):
+        from tensorboardX import SummaryWriter
+        self.writer = SummaryWriter(self._dir)
+
+    def writekvts(self, key: str, value: float, timestamp: int):
[…]

### 2018-11-26 pendulum_exp/policies/continuous.py

@@ -1,5 +1,4 @@
 """Define continuous policy."""
-from typing import Optional
 import torch
 from torch import Tensor
 import numpy as np
@@ -47,7 +46,6 @@ class SampledAdvantagePolicy(SharedAdvantagePolicy):
 
         # logging
         self._cum_loss = 0
-        self._float_loss: Optional[float] = None
 
     def act(self, obs: Arrayable):
         with torch.no_grad():
@@ -86,24 +84,20 @@ class SampledAdvantagePolicy(SharedAdvantagePolicy):
 
         # logging
         self._cum_loss += losses[0].item()
-        if self._float_loss is None:
-            self._float_loss = losses[0].item()
-        else:
-            self._float_loss = self._float_loss * self._gamma ** self._dt +\
-                (1 - self._gamma ** self._dt) * losses[0].item()
+        self._log_step += 1
         self._learn_count += 1
 
     def optimize_policy(self, max_adv: Tensor):
         pass
 
+    def reset_log(self):
+        self._cum_loss = 0
+        self._log_step = 0
+
     def log(self):
-        log("Avg_adv_loss", self._cum_loss / self._learn_count,
-            self._learn_count)
-        log("Float_adv_loss", self._float_loss, self._learn_count)
+        log("adv_loss", self._cum_loss / self._log_step, self._learn_count)
         info(f"At iteration {self._learn_count}, "
-             f'Float_adv_loss: {self._float_loss}, '
-             f'Avg_adv_loss: {self._cum_loss/self._learn_count}, '
-             f'Avg_policy_loss: {self._cum_policy_loss / self._learn_count}')
+             f'adv_loss: {self._cum_loss/self._log_step}')
 
     def train(self):
         self._train = True
@@ -189,7 +183,6 @@ class AdvantagePolicy(SharedAdvantagePolicy):
 
         # logging
         self._cum_loss = 0
-        self._float_loss = None
         self._cum_policy_loss = 0
 
     def act(self, obs: Arrayable):
@@ -213,11 +206,7 @@ class AdvantagePolicy(SharedAdvantagePolicy):
 
         # logging
         self._cum_loss += losses[0].item()
-        if self._float_loss is None:
-            self._float_loss = losses[0].item()
-        else:
-            self._float_loss = self._float_loss * self._gamma ** self._dt +\
-                (1 - self._gamma ** self._dt) * losses[0].item()
+        self._log_step += 1
         self._learn_count += 1
 
[…]

### 2018-11-26 pendulum_exp/policies/discrete.py

@@ -1,5 +1,4 @@
 """Define policies."""
-from typing import Optional
 import torch
 from torch import Tensor
 from abstract import ParametricFunction, Arrayable, Noise, StateDict
@@ -43,8 +42,7 @@ class AdvantagePolicy(SharedAdvantagePolicy):
 
         # logging
         self._cum_loss = 0
-        self._float_loss: Optional[float] = None
-        self._log = 100
+        self._log_step = 0
 
     def act(self, obs: Arrayable):
         with torch.no_grad():
@@ -60,6 +58,10 @@ class AdvantagePolicy(SharedAdvantagePolicy):
         adv = adv_a.gather(1, indices.view(-1, 1)).squeeze()
         return adv, max_adv
 
+    def reset_log(self):
+        self._cum_loss = 0
+        self._log_step = 0
+
     def optimize_value(self, *losses: Tensor):
         assert len(losses) == 1
         self._optimizers[0].zero_grad()
@@ -70,11 +72,7 @@ class AdvantagePolicy(SharedAdvantagePolicy):
 
         # logging
         self._cum_loss += losses[0].item()
-        if self._float_loss is None:
-            self._float_loss = losses[0].item()
-        else:
-            self._float_loss = self._float_loss * self._gamma ** self._dt +\
-                (1 - self._gamma ** self._dt) * losses[0].item()
+        self._log_step += 1
         self._learn_count += 1
 
     def optimize_policy(self, max_adv: Tensor):
@@ -85,11 +83,9 @@ class AdvantagePolicy(SharedAdvantagePolicy):
         self._schedulers[1].step(eval_return)
 
     def log(self):
-        log("Avg_adv_loss", self._cum_loss / self._learn_count, self._learn_count)
-        log("Float_adv_loss", self._float_loss, self._learn_count)
+        log("adv_loss", self._cum_loss / self._log_step, self._learn_count)
         info(f"At iteration {self._learn_count}, "
-             f'Float_adv_loss: {self._float_loss}, '
-             f"avg_loss: {self._cum_loss/self._learn_count}")
+             f"adv_loss: {self._cum_loss/self._log_step}")
 
     def train(self):
         self._train = True

### 2018-11-26 pendulum_exp/policies/shared.py

@@ -28,6 +28,8 @@ class SharedAdvantagePolicy(Policy):
         self._schedule_params = dict(
             mode='max', factor=.5, patience=25)
 
+        self._log_step = 0
+
     def reset(self):
         # internals
         self._obs = np.array([])
@@ -75,6 +77,7 @@ class SharedAdvantagePolicy(Policy):
 
         if self._count % self._steps_btw_train == self._steps_btw_train - 1:
             try:
+                self.reset_log()
                 for _ in range(self._learn_per_step):
                     obs, action, next_obs, reward, done, weights, time_limit = \
                         self._sampler.sample()

### 2018-11-26 pendulum_exp/policies/benchmarkspolicies/dqn.py
dqn is working

@@ -1,9 +1,10 @@
 from itertools import chain
 import torch
+import numpy as np
 from abstract import ParametricFunction, Arrayable, Noise, StateDict, Policy
 
 import copy
-from convert import arr_to_th, th_to_arr
+from convert import arr_to_th, check_array, th_to_arr
 # from stats import penalize_mean
 from config import DQNConfig
 # from policies.shared import SharedAdvantagePolicy
@@ -32,6 +33,8 @@ class DQNPolicy(Policy):
         self._steps_btw_train = policy_config.steps_btw_train
         self._steps_btw_catchup = policy_config.steps_btw_catchup
         self._sampler = setup_memory(policy_config)
+        self._count = 0
+        self._learn_count = 0
         # optimization/storing
         self._device = device
         self._optimizer = setup_optimizer(
@@ -51,27 +54,60 @@ class DQNPolicy(Policy):
         self._log = 100
 
 
+    def reset(self):
+        # internals
+        self._obs = np.array([])
+        self._action = np.array([])
+        self._reward = np.array([])
+        self._next_obs = np.array([])
+        self._done = np.array([])
+
+    def step(self, obs: Arrayable):
+        if self._train:
+            self._obs = obs
+
+        action = self.act(obs)
+        if self._train:
+            self._action = action
+
+        return action
+
+    def observe(self,
+                next_obs: Arrayable,
+                reward: Arrayable,
+                done: Arrayable):
+        if self._train:
+            self._count += 1
+            self._next_obs = next_obs
+            self._reward = reward
+            self._done = done
+            self._sampler.push(
+                self._obs, self._action, self._next_obs, self._reward, self._done)
+            self.learn()
+
+
     def act(self, obs: Arrayable):
         with torch.no_grad():
-            pre_action = self._adv_noise.perturb_output(
+            pre_action = self._qnet_noise.perturb_output(
                 obs, function=self._qnet_function)
-            self._adv_noise.step()
+            self._qnet_noise.step()
             return pre_action.argmax(axis=-1)
 
     def learn(self):
         if self._count % self._steps_btw_train == self._steps_btw_train - 1:
             # TODO: enlever try except
-            print("Your are not dreaming, you are in DQN.")
             try:
                 for _ in range(self._learn_per_step):
-                    obs, action, next_obs, reward, done = self._sampler.sample()
+                    obs, action, next_obs, reward, done, weights = self._sampler.sample()
[…]

### 2018-11-26 paper/tangotd.tex

@@
 (f(s)-f(s'))^2$. Its matrix is $\rho+\rho'-\rho P-\transp{(\rho P)}$ where
-$\rho'$ is the distribution of $s'$.
+$\rho'$ is the distribution of $s'$ (and is samplable).
 A local minimizer of the Dirichlet error between $V$ and
@@
 transitions only for the correction? TODO: use the same trick on the
-right, double sampling becomes only on a term that is quadratic in the
-non-reversibility.
+right? No, because the $R$ term doesn't go away and I'd rather have a
+correction that vanishes when $\Delta V-R$ is small.
 

### 2018-11-28 paper/tangotd.tex

@@
 the linear regression problem $(\Id-P)\Phi\theta\approx R$ in
-$L^2(\rho)$. However, the matrix on the left is not samplable.
+$L^2(\rho)$. However, the matrix on the left is not samplable. We need
+double sampling for that.
+
+Combined with all this, is the problem that a Bellman solution on the
+support of $\rho$ may be very, very far from the global Bellman solution,
+even on the support of $\rho$ (take random walk on a circle, with $\rho$
+being everything but one point: then any linear function solves Bellman
+on the support of $\rho$).
 

### 2018-11-29 paper/tangotd.tex

@@
 
-This works only if $A$ is \emph{stable}.
+Under this update for $\theta$, the error $\theta-A^{-1}b$ is updated as
+\begin{equation}
+\theta-A^{-1}b \gets (\Id-\eta A)(\theta-A^{-1}b)
+\end{equation}
+
+This converges if only if $A$ is \emph{stable}.
 
@@
 $\eps$.
+\item For $\eps$ small enough, iterating $\theta\gets (\Id-\eps A)\theta$
+converges to $0$.
 \item The differential equation $\theta'=-A\theta$ converges to $0$ for
@@
 (Dirichlet form). This implies that $A=\transp{\Phi}\rho (\Id-P)\Phi$ is
-positive too. So we recover the convergence result on-policy for linear
-TD.
+positive too. So we can take $L=\Id$ and we recover convergence of linear
+TD on-policy.
 \item In the tabular case: $\Phi=\Id$. Indeed we have $A=\rho (\Id-P)$,
 which is not positive in general. But $\mu (\Id-P)$ is positive, so we can take
-$L=\mu/\rho$ so that $LA$ is positive.
+$L=\mu/\rho$ so that $LA$ is positive. Thus, tabular TD works
+off-policy.
 \end{enumerate}
@@
 
+\paragraph{Nonlinear case.} In the nonlinear case, we still set $\Phi\deq
+\partial V/\partial \theta$ and TD is
+\begin{equation}
+\theta \gets \theta-\eta \transp{\Phi}\rho (V-PV-R)
+\end{equation}
+and the resulting change of $V$ is
+\begin{equation}
+V\gets V-\eta \Phi\transp{\Phi}\rho (V-PV-R)+O(\eta^2)
+\end{equation}
+
+Preconditioning TD with a matrix $B$ amounts to using $B\transp{\Phi}$ instead of
+$\transp{\Phi}$ in the update of $\theta$, and $\Phi B \transp{\Phi}$ instead of
+$\Phi\transp{\Phi}$ in the update of $V$.
+
+Let us consider a loss function $\transp{(V-PV-R)}L(V-PV-R)$ on the
+Bellman gap, with $L$ a symmetric definite positive matrix. For small
+$\eta$, the derivative of this loss function wrt $\theta$ is
+\begin{equation}
+\transp{(V-PV-R)}L(\Phi-P\Phi)
+\end{equation}
+
+The loss function $L$ may not depend on $\theta$, otherwise we are
+decreasing a variable Lyapunov function which does not prove convergence.
+
+If the algorithm is some form of preconditioned TD, namely
+$\delta\theta=-\eta B\transp{\Phi}\rho (V-PV-R)$, the loss function will
+decrease if and only if
+\begin{equation}
+\transp{(V-PV-R)}L(\Phi-P\Phi)B\transp{\Phi}\rho (V-PV-R)
+\end{equation}
+is positive, namely, if $L\Delta \Phi B \transp{\Phi}\rho $ is a positive
+operator.
+
+In the reversible case, $\mu\Delta$ is symmetric. We can use the
+Dirichlet form which amounts to $L\deq\mu(\mu\Delta)^{-1}\mu$; then
+$L\Delta=\mu$ and we have to prove that $\mu\Phi B\transp{\Phi}\rho$ is
+positive.  On-policy, $\rho=\mu$ and this is obviously true, for any
+positive $B$. Thus, in the reversible case, nonlinear TD converges
+on-policy, for any $B$.
+
+In the non-reversible case, let $\mu\Delta=S+A$ with $S$ symmetric and
+$A$ antisymmetric. Let $L\deq \mu S^{-1} \mu$. Then we have to prove that
+$\mu S^{-1} \mu \Delta \Phi B\transp{\Phi}\rho$ is positive. For any
+function $f$,
+\begin{align}
+\transp{f} \mu S^{-1} \mu \Delta \Phi B\transp{\Phi}\rho f
+&=\transp{f} \mu S^{-1} (S+A) \Phi B\transp{\Phi}\rho f
+\\&=\transp{f} \mu \Phi B\transp{\Phi}\rho f + \transp{f} \mu S^{-1}
+A \Phi B\transp{\Phi}\rho f
+\end{align}
+
+Let us consider the on-policy case. Then the first term is $\transp{f}
+\mu \Phi B\transp{\Phi}\rho f$ and is always positive. For the second
+term, since $A$ is antisymmetric, we have $\transp{x}Ax=0$ for all $x$.
+Thus, we can add any multiple of $\transp{x}Ax=0$; let us take
+$x=S^{-1}\mu f$:
+\begin{equation}
+\transp{f} \mu S^{-1}
+A \Phi B\transp{\Phi}\mu f=
+\transp{f} \mu S^{-1}A (S^{-1}-\Phi B\transp{\Phi})\mu f
+\end{equation}
+
+In particular, if $S^{-1}=\Phi B\transp{\Phi}$ then this vanishes.
+The latter can only happen in the overparameterized case (indeed, the
+right-hand-side lives in the image of $\Phi$). In that case $\Phi$ is
+invertible and $B=(\transp{\Phi}S\Phi)^{-1}$ solves the problem; a
+two-timescale algorithm is required to compute $B$.
+
+Thus we have a two-timescale algorithm for non-linear TD in the on-policy
+overparameterized case.
+
 \end{document}

### 2018-11-29 paper/tangotd.tex

@@
 
-In the reversible case, $\mu\Delta$ is symmetric. We can use the
+In the reversible case, $\mu\Delta$ is symmetric.\footnote{$\rho \Delta$
+can be symmetric if and only if $\rho=\mu$. Indeed, if $\rho \Delta$ is
+symmetric then $\rho=\rho'$ and $\rho$ is the stationary distribution.} We can use the
 Dirichlet form which amounts to $L\deq\mu(\mu\Delta)^{-1}\mu$; then
@@
 
-In the non-reversible case, let $\mu\Delta=S+A$ with $S$ symmetric and
-$A$ antisymmetric. Let $L\deq \mu S^{-1} \mu$. Then we have to prove that
-$\mu S^{-1} \mu \Delta \Phi B\transp{\Phi}\rho$ is positive. For any
+In the non-reversible case, let $\rho\Delta=S+A$ with $S$ symmetric and
+$A$ antisymmetric. Let $L\deq \rho S^{-1} \rho$. Then we have to prove that
+$\rho S^{-1} \rho \Delta \Phi B\transp{\Phi}\rho$ is positive. For any
 function $f$,
 \begin{align}
-\transp{f} \mu S^{-1} \mu \Delta \Phi B\transp{\Phi}\rho f
-&=\transp{f} \mu S^{-1} (S+A) \Phi B\transp{\Phi}\rho f
-\\&=\transp{f} \mu \Phi B\transp{\Phi}\rho f + \transp{f} \mu S^{-1}
+\transp{f} \rho S^{-1} \rho \Delta \Phi B\transp{\Phi}\rho f
+&=\transp{f} \rho S^{-1} (S+A) \Phi B\transp{\Phi}\rho f
+\\&=\transp{f} \rho \Phi B\transp{\Phi}\rho f + \transp{f} \rho S^{-1}
 A \Phi B\transp{\Phi}\rho f
@@
 
-Let us consider the on-policy case. Then the first term is $\transp{f}
-\mu \Phi B\transp{\Phi}\rho f$ and is always positive. For the second
+The first term $\transp{f}
+\rho \Phi B\transp{\Phi}\rho f$ is always positive. For the second
 term, since $A$ is antisymmetric, we have $\transp{x}Ax=0$ for all $x$.
 Thus, we can add any multiple of $\transp{x}Ax=0$; let us take
-$x=S^{-1}\mu f$:
+$x=S^{-1}\rho f$:
 \begin{equation}
-\transp{f} \mu S^{-1}
-A \Phi B\transp{\Phi}\mu f=
-\transp{f} \mu S^{-1}A (S^{-1}-\Phi B\transp{\Phi})\mu f
+\transp{f} \rho S^{-1}
+A \Phi B\transp{\Phi}\rho f=
+\transp{f} \rho S^{-1}A (S^{-1}-\Phi B\transp{\Phi})\rho f
 \end{equation}
@@
 
-Thus we have a two-timescale algorithm for non-linear TD in the on-policy
+Thus we have a two-timescale algorithm for non-linear TD in the off-policy
 overparameterized case.
 
+(Still works for non-overparameterized if $A$ does not couple the image
+of $\Phi$ and its complement...)
+
 \end{document}

### 2018-11-29 paper/tangotd.tex

@@
 positive $B$. Thus, in the reversible case, nonlinear TD converges
-on-policy, for any $B$.
+on-policy, for any $B$. It does so by minimizing the Dirichlet norm.
 
@@
 A \Phi B\transp{\Phi}\rho f=
-\transp{f} \rho S^{-1}A (S^{-1}-\Phi B\transp{\Phi})\rho f
+\transp{f} \rho S^{-1}A (\Phi B\transp{\Phi}-S^{-1})\rho f
 \end{equation}
@@
 Thus we have a two-timescale algorithm for non-linear TD in the off-policy
-overparameterized case.
+overparameterized case; it minimizes the Dirichlet norm. TODO WRONG: $S$
+is not necessarily positive!!!!! is positive if $\rho=\mu$.
+
+TODO: Why do I end up using $\rho\Delta+\transp{(\rho \Delta)}$ while the
+Dirichlet form is $\rho \Delta + \rho'-\transp{(\rho P)}$ which is
+different, I think...
 
@@
 
+Nonlinear GTD2 etc use a function $L$ that depends on $\theta$ (via
+$\Phi$); this adds Hessian terms to the gradient of $L$, and it's not
+obvious how to interpret the local minima of $L$.
+
 \end{document}

### 2018-11-29 paper/learn_discontinuities.tex
Learning discontinuities for A

\documentclass[11pt]{article}

\usepackage{amsmath,amssymb}

\newcommand{\E}{\mathbb{E}}
\newcommand{\bern}[1]{\mathcal{B}\left(#1\right)}
\newcommand{\gauss}[1]{\mathcal{N}\left(#1\right)}
\newcommand{\argmax}{\text{amax}}
\newcommand{\bigO}{O}

\begin{document}
Consider the advantage updating set of equations in continuous time
\begin{align}
	A^*(s, a) &= \max_a A^*(s,a) + \frac{r_t dt + \gamma^{dt} \E_{s_{t+dt}} V^*(s_{t+dt}) - V^*(s_t)}{dt}\\
	\max_a A^*(s, a) &= 0.
	\label{eq:adv_up}
\end{align}
On ergodic reversible environments, those equations are mostly well defined in
term of orders of magnitude, as $Q^*(s,a) - V^*(s) = \bigO(dt)$ for all $s, a$.
However, when facing highly non reversible environments, e.g. episodic
environments, action state pairs where the advantage between two different actions
is not of order $dt$ exist, i.e., there can exist $s_0, a_0$ such that
$Q^*(s_0, a_0) - V^*(s_0) = \bigO(dt)$. In such cases, $A^*(s_0, a_0)$ is infinite,
and it is doubtful that an algorithm learning parametric approximations of
$V$ and $A$ based on~\eqref{eq:adv_up} will properly converge.

To palliate this behaviour, when learning parametric approximations of $V^*$ and $A^*$,
$V_\theta$, $A_\phi$, one can parameterize $A_\phi$ as a mixture of gaussians
\begin{align}
	A_\phi(s, a) = \bern{\mu_\phi(s, a)dt} \frac{\gauss{\tilde{A}_\phi(s, a), 1}}{dt} + \bern{1 - \mu_\phi(s, a)dt} \gauss{\bar{A}_\phi(s, a), 1}
\end{align}
where $\bern{p}$ denotes independant bernoulli random variables of parameter $p$, and
$\gauss{\mu, \sigma}$ independant gaussian random variables of mean $\mu$, variance
$\sigma$.

Suppose the value of $V^*(s)$ and $V^*(s')$ given on a dataset of transitions.
We want to compute the likelihood of our model of $A$.
The likelihood of the model of $A$ on a single datapoint is given by the log likelihood
of the following mixture density
\begin{align}
	&\bern{\mu(s, a)\mu(s, \argmax_{a'} \tilde{A}(s, a')dt^2} \frac{\gauss{\tilde{A}(s, a) - \max_a \tilde{A}(s, a), 2}}{dt} + \nonumber\\
	&\bern{(1 - \mu(s, a)dt)\mu(s, \argmax_{a'} \tilde{A}(s, a')dt} \frac{\gauss{\bar{A}(s, a)dt - \max_a \tilde{A}(s, a), 1 + dt^2}}{dt} + \nonumber\\
	&\bern{(1 - \mu(s, a)dt)(1 - \mu(s, \argmax_{a'} \bar{A}(s, a')dt)} \gauss{\bar{A}(s, a) - \max_a \bar{A}(s, a), 2} + \nonumber\\
	&\bern{\mu(s, a)dt(1 - \mu(s, \argmax_{a'} \bar{A}(s, a')dt)} \frac{\gauss{\tilde{A}(s, a) - \max_a \tilde{A}(s, a), 1 + dt^2}}{dt}
\end{align}
from which a gradient based learning procedure can be derived for $A$.
\end{document}

### 2018-11-29 paper/tangotd.tex

@@
 overparameterized case; it minimizes the Dirichlet norm. TODO WRONG: $S$
-is not necessarily positive!!!!! is positive if $\rho=\mu$.
+IS NOT NECESSARILY POSITIVE!!!!! $S$ is positive if $\rho=\mu$, so,
+correct on-policy.
+
+(Still works for non-overparameterized if $A$ does not couple the image
+of $\Phi$ and its complement...)
 
@@
 Dirichlet form is $\rho \Delta + \rho'-\transp{(\rho P)}$ which is
-different, I think...
-
-(Still works for non-overparameterized if $A$ does not couple the image
-of $\Phi$ and its complement...)
+different, I think...  Probably use the right combination of $\rho$ and
+$\rho'$ and the Dirichlet form rather than $\rho S^{-1}\rho$.
 
@@
 
+\paragraph{Why $S$ is not positive.} The Dirichlet form associated
+with the set of transitions $s\to s'$ is
+\begin{align}
+\Dir&=\rho+\rho'-\rho P-\transp{(\rho P)}
+\\&=\E[\1_{ss}+\1_{s's'}-\1_{ss'}-\1_{s's}]
+\end{align}
+where the expectation is over transitions in the dataset. The Dirichlet
+form satisfies
+\begin{equation}
+\qquad \transp{f}\Dir f=\E_{ss'}(f(s)-f(s'))^2
+\end{equation}
+which is obviously positive. Meanwhile,
+\begin{align}
+\rho\Delta&=\rho-\rho P=\E[\1_{ss}-\1_{ss'}]
+\\&=\frac12 \Dir+\frac12 \E[\1_{ss}-\1_{s's'}]+\frac12
+\E[\1_{s's}-\1_{ss'}]
+\end{align}
+and the first two terms are symmetric while the last term is
+antisymmetric, so
+\begin{equation}
+S=\frac12 \Dir+\frac12 \E[\1_{ss}-\1_{s's'}],\qquad A=\frac12
+\E[\1_{s's}-\1_{ss'}]
+\end{equation}
+and
+\begin{equation}
+\transp{f}Sf=\E f(s)^2-\E f(s)f(s')
+\end{equation}
+which in general is not positive (take $\rho$ concentrated on a single $s$, and a large value
+for $f(s')$).
+
 \end{document}

### 2018-11-29 paper/tangotd.tex

@@
 
+On-policy, a choice would seem to be $L=\mu\Phi B \transp{\Phi}\mu$ which
+would result in $(\mu\Phi B \transp{\Phi})(\mu \Delta) \mu\Phi B
+\transp{\Phi}\mu$ which is positive; however, this depends on $\theta$
+since $\Phi$ does. So this is not a suitable Lyapunov function.
+
 In the reversible case, $\mu\Delta$ is symmetric.\footnote{$\rho \Delta$
@@
 In the non-reversible case, let $\rho\Delta=S+A$ with $S$ symmetric and
-$A$ antisymmetric. Let $L\deq \rho S^{-1} \rho$. Then we have to prove that
+$A$ antisymmetric. Assume $S$ is positive. Let $L\deq \rho S^{-1} \rho$. Then we have to prove that
 $\rho S^{-1} \rho \Delta \Phi B\transp{\Phi}\rho$ is positive. For any
@@
 
-Thus we have a two-timescale algorithm for non-linear TD in the off-policy
-overparameterized case; it minimizes the Dirichlet norm. TODO WRONG: $S$
-IS NOT NECESSARILY POSITIVE!!!!! $S$ is positive if $\rho=\mu$, so,
-correct on-policy.
+Thus, assuming $S$ is positive, we have a two-timescale algorithm for non-linear TD in the off-policy
+overparameterized case; it minimizes the Dirichlet norm.
+However, in general $S$ is not positive. $S$ is positive if $\rho=\mu$, so,
+we have a two-timescale algorithm for non-linear on-policy
+overparameterized TD.
 

### 2018-11-29 pendulum_exp/config.py
Adding mixture policies. Still need to be entirely debugged.

@@ -44,6 +44,7 @@ class ApproximateAdvantagePolicyConfig(NamedTuple):
     weight_decay: float
     policy_lr: float
     optimizer: str
+    mixture: bool
 
 class SampledAdvantagePolicyConfig(NamedTuple):
     batch_size: int
@@ -81,7 +82,7 @@ class EnvConfig(NamedTuple):
 
 
 NoiseConfig = Union[ParameterNoiseConfig, ActionNoiseConfig]
-PolicyConfig = Union[SampledAdvantagePolicyConfig, AdvantagePolicyConfig, ApproximateAdvantagePolicyConfig]
+PolicyConfig = Union[SampledAdvantagePolicyConfig, AdvantagePolicyConfig, ApproximateAdvantagePolicyConfig, DQNConfig]
 
 def read_config(
         args,
@@ -107,13 +108,14 @@ def read_config(
     )
     if args.policy_lr is not None:
         policy_config_dict['policy_lr'] = args.policy_lr
+        policy_config_dict['mixture'] = args.algo == 'mdrau'
         policy_config: PolicyConfig = ApproximateAdvantagePolicyConfig(
             **policy_config_dict)
     elif args.nb_policy_samples is not None:
         policy_config_dict['nb_samples'] = args.nb_policy_samples
         policy_config = SampledAdvantagePolicyConfig(**policy_config_dict)
     else:
-        if args.algo == 'drau':
+        if 'drau' in args.algo:
             policy_config = AdvantagePolicyConfig(**policy_config_dict)
         elif args.algo == 'qlearn':
             policy_config = DQNConfig(

### 2018-11-29 pendulum_exp/envs/pusher.py

@@ -41,7 +41,7 @@ class AbstractPusher(gym.Env, Env):
 
         action = self.action(action)
         self._x -= action * self.dt
-        return self._x, np.exp(- self._x[0] ** 2), False, None
+        return self._x, np.exp(- self._x[0] ** 2), False, {}
 
     def reset(self):
         # without this, all thread have the same seed

### 2018-11-29 pendulum_exp/models.py

@@ -40,8 +40,8 @@ class ContinuousPolicyMLP(MLP, ParametricFunction):
 class ContinuousAdvantageMLP(MLP, ParametricFunction):
     """MLP with 2 inputs, 1 output."""
     def __init__(self, nb_state_feats: int, nb_actions: int,
-                 nb_layers: int, hidden_size: int) -> None:
-        super().__init__(nb_state_feats + nb_actions, 1,
+                 nb_layers: int, hidden_size: int, nb_outputs: int) -> None:
+        super().__init__(nb_state_feats + nb_actions, nb_outputs,
                          nb_layers, hidden_size)
         self._nb_state_feats = nb_state_feats
         self._nb_actions = nb_actions

### 2018-11-29 pendulum_exp/nn.py

@@ -1,0 +1,76 @@
+"""Some nn utilities."""
+from typing import Tuple
+import torch
+from torch import Tensor
+import torch.nn.functional as f
+from torch.distributions import Normal
+from abstract import ParametricFunction, Shape
+
+def gmm_loss(batch: Tensor,
+             mus: Tensor,
+             sigmas: Tensor,
+             logpi: Tensor,
+             reduce: bool=True) -> Tensor: # pylint: disable=too-many-arguments
+    """ Computes the gmm loss.
+    Compute minus the log probability of batch under the GMM model described
+    by mus, sigmas, pi. Precisely, with bs1, bs2, ... the sizes of the batch
+    dimensions (several batch dimension are useful when you have both a batch
+    axis and a time step axis), gs the number of mixtures and fs the number of
+    features.
+    :args batch: (bs1, bs2, *, fs) torch tensor
+    :args mus: (bs1, bs2, *, gs, fs) torch tensor
+    :args sigmas: (bs1, bs2, *, gs, fs) torch tensor
+    :args logpi: (bs1, bs2, *, gs) torch tensor
+    :args reduce: if not reduce, the mean in the following formula is ommited
+    :returns:
+    loss(batch) = - mean_{i1=0..bs1, i2=0..bs2, ...} log(
+        sum_{k=1..gs} pi[i1, i2, ..., k] * N(
+            batch[i1, i2, ..., :] | mus[i1, i2, ..., k, :], sigmas[i1, i2, ..., k, :]))
+    NOTE: The loss is not reduced along the feature dimension (i.e. it should scale ~linearily
+    with fs).
+    """
+    batch = batch.unsqueeze(-2)
+    normal_dist = Normal(mus, sigmas)
+    g_log_probs = normal_dist.log_prob(batch)
+    g_log_probs = logpi + torch.sum(g_log_probs, dim=-1)
+    max_log_probs = torch.max(g_log_probs, dim=-1, keepdim=True)[0]
+    g_log_probs = g_log_probs - max_log_probs
+
+    g_probs = torch.exp(g_log_probs)
+    probs = torch.sum(g_probs, dim=-1)
+
+    log_prob = max_log_probs.squeeze() + torch.log(probs)
+    if reduce:
+        return - torch.mean(log_prob)
+    return - log_prob
+
+class Scale(torch.nn.Module, ParametricFunction):
+    def __init__(self, core: ParametricFunction, scale: Tuple[float, ...]):
+        super().__init__()
+        self._core = core
[…]

### 2018-11-29 pendulum_exp/parse.py

@@ -4,7 +4,7 @@ import argload
 
 def setup_args():
     parser = argparse.ArgumentParser()
-    parser.add_argument('--algo', type=str, default='drau', choices=['drau', 'qlearn'])
+    parser.add_argument('--algo', type=str, default='drau', choices=['mdrau', 'drau', 'qlearn'])
     parser.add_argument('--dt', type=float, default=.05)
     parser.add_argument('--steps_btw_train', type=int, default=3)
     parser.add_argument('--steps_btw_catchup', type=int, default=10)
@@ -37,7 +37,7 @@ def setup_args():
                         default='sgd')
     parser.add_argument('--noreload', action='store_true')
     parser = argload.ArgumentLoader(parser, to_reload=[
-        'algo', 'dt', 'steps_btw_train', 'steps_btw_catchup', 'env_id', 'noise_type', 
+        'algo', 'dt', 'steps_btw_train', 'steps_btw_catchup', 'env_id', 'noise_type',
         'batch_size', 'hidden_size', 'nb_layers', 'gamma', 'nb_epochs', 'nb_steps',
         'sigma_eval', 'sigma', 'theta', 'nb_train_env', 'nb_eval_env', 'memory_size',
         'learn_per_step', 'cyclic_expliration', 'normalize_state', 'lr', 'time_limit',

### 2018-11-29 pendulum_exp/policies/__init__.py

@@ -4,6 +4,7 @@ from gym import Space
 from gym.spaces import Discrete, Box
 from abstract import Policy
 from policies.discrete import AdvantagePolicy
+from policies.mixture import ContinuousAdvantageMixturePolicy
 from policies.continuous import AdvantagePolicy as ContinuousAdvantagePolicy
 from policies.continuous import SampledAdvantagePolicy
 from policies.benchmarkspolicies.dqn import DQNPolicy
@@ -27,6 +28,11 @@ def setup_policy(observation_space: Space,
     # assume observation space has shape (F,) for now
     nb_state_feats = observation_space.shape[0]
 
+    if hasattr(policy_config, 'mixture') and policy_config.mixture:
+        nb_outputs = 4
+    else:
+        nb_outputs = 1
+
     # normalization
     if normalize_state:
         def maker(maker_cls, **kwargs):
@@ -36,12 +42,11 @@ def setup_policy(observation_space: Space,
         def maker(maker_cls, **kwargs):
             return maker_cls(**kwargs)
 
-
-    if isinstance(policy_config, (AdvantagePolicyConfig, SampledAdvantagePolicyConfig, 
+    if isinstance(policy_config, (AdvantagePolicyConfig, SampledAdvantagePolicyConfig,
                   ApproximateAdvantagePolicyConfig)):
         net_dict = dict(nb_layers=nb_layers, hidden_size=hidden_size)
 
-        val_function = maker(MLP, nb_inputs=nb_state_feats, nb_outputs=1, **net_dict).to(device)
+        val_function = maker(MLP, nb_inputs=nb_state_feats, nb_outputs=nb_outputs, **net_dict).to(device)
 
         policy_dict = dict(val_function=val_function, policy_config=policy_config, device=device)
 
@@ -62,7 +67,7 @@ def setup_policy(observation_space: Space,
         elif isinstance(policy_config, (ApproximateAdvantagePolicyConfig, SampledAdvantagePolicyConfig)):
             assert isinstance(action_space, Box)
             nb_actions = action_space.shape[-1]
-            adv_function = maker(ContinuousAdvantageMLP,
+            adv_function = maker(ContinuousAdvantageMLP, nb_outputs=nb_outputs,
                                  nb_state_feats=nb_state_feats, nb_actions=nb_actions,
                                  **net_dict).to(device)
             policy_dict["adv_function"] = adv_function
@@ -78,8 +83,12 @@ def setup_policy(observation_space: Space,
[…]

### 2018-11-29 pendulum_exp/policies/benchmarkspolicies/dqn.py

@@ -54,7 +54,6 @@ class DQNPolicy(Policy):
         self._cum_loss = 0
         self._log = 100
 
-
     def reset(self):
         # internals
         self._obs = np.array([])
@@ -63,7 +62,6 @@ class DQNPolicy(Policy):
         self._next_obs = np.array([])
         self._done = np.array([])
         self._time_limit = np.array([])
-
 
     def step(self, obs: Arrayable):
         if self._train:
@@ -79,7 +77,7 @@ class DQNPolicy(Policy):
                 next_obs: Arrayable,
                 reward: Arrayable,
                 done: Arrayable,
-                time_limit:Optional[Arrayable]=None):
+                time_limit: Optional[Arrayable]=None):
         if self._train:
             self._count += 1
             self._next_obs = next_obs
@@ -90,7 +88,6 @@ class DQNPolicy(Policy):
             self._sampler.push(
                 self._obs, self._action, self._next_obs, self._reward, self._done, self._time_limit)
             self.learn()
-
 
     def act(self, obs: Arrayable):
         with torch.no_grad():
@@ -113,13 +110,13 @@ class DQNPolicy(Policy):
                     q = self._qnet_function(obs).gather(1, indices.view(-1, 1)).squeeze()
                     target = torch.max(self._target_function(next_obs), dim=1)[0].squeeze()
 
-                    exp_q = reward + self._gamma ** self._dt 
+                    exp_q = reward + self._gamma ** self._dt
                     if self._gamma == 1.:
                         assert (1 - done).all(), "Gamma set to 1. with a potentially episodic problem..."
                         exp_q += target.detach()
                     else:
                         done = arr_to_th(done.astype('float'), self._device)
-                        exp_q += self._gamma ** self._dt * target#.detach()
+                        exp_q += self._gamma ** self._dt * target # .detach()
 
                     loss = (((q - exp_q) ** 2) * weights).mean()
                     self._optimizer.zero_grad()
@@ -136,9 +133,8 @@ class DQNPolicy(Policy):
                 # If not enough data in the buffer, do nothing
                 pass
 
-        if self._count % self._steps_btw_catchup == self._steps_btw_catchup -1:
+        if self._count % self._steps_btw_catchup == self._steps_btw_catchup - 1:
             self._target_function.load_state_dict(self._qnet_function.state_dict())
-
 
     def train(self):
         self._train = True
@@ -159,7 +155,7 @@ class DQNPolicy(Policy):
[…]

### 2018-11-29 pendulum_exp/policies/mixture.py

@@ -1,0 +1,270 @@
+import numpy as np
+import torch
+from torch import Tensor
+import torch.nn.functional as f
+from abstract import ParametricFunction, Arrayable, Noise, StateDict
+from policies.shared import SharedAdvantagePolicy
+from config import ApproximateAdvantagePolicyConfig
+from convert import arr_to_th, check_array, th_to_arr
+from optimizer import setup_optimizer
+from nn import gmm_loss
+from mylog import log
+from logging import info
+
+class ContinuousAdvantageMixturePolicy(SharedAdvantagePolicy):
+    def __init__(self, policy_config: ApproximateAdvantagePolicyConfig,
+                 val_function: ParametricFunction, adv_function: ParametricFunction,
+                 policy_function: ParametricFunction, policy_noise: Noise,
+                 device) -> None:
+        super().__init__(policy_config, val_function, device)
+        self._adv_function = adv_function
+        self._policy_function = policy_function.to(device)
+        self._policy_noise = policy_noise
+
+        self._optimizers = (
+            setup_optimizer(self._adv_function.parameters(),
+                            opt_name=policy_config.optimizer,
+                            lr=policy_config.lr, dt=self._dt,
+                            inverse_gradient_magnitude=1,
+                            weight_decay=policy_config.weight_decay),
+            setup_optimizer(self._val_function.parameters(),
+                            opt_name=policy_config.optimizer,
+                            lr=policy_config.lr, dt=self._dt,
+                            inverse_gradient_magnitude=1,
+                            weight_decay=policy_config.weight_decay),
+            setup_optimizer(self._policy_function.parameters(),
+                            opt_name=policy_config.optimizer,
+                            lr=policy_config.policy_lr, dt=self._dt,
+                            inverse_gradient_magnitude=1,
+                            weight_decay=policy_config.weight_decay))
+
+        self._schedulers = (
+            torch.optim.lr_scheduler.ReduceLROnPlateau(
+                self._optimizers[0], **self._schedule_params),
+            torch.optim.lr_scheduler.ReduceLROnPlateau(
+                self._optimizers[1], **self._schedule_params),
+            torch.optim.lr_scheduler.ReduceLROnPlateau(
[…]

### 2018-11-30 paper/baselines.tex

\documentclass[11pt,a4paper]{article}
\input{header.en.tex}

\begin{document}
Let us gather a few baselines for continuous-time RL, starting with the
most basic. The baselines below have no problem with the continuous-time
limit (provided $\gamma<1$).

All to be tested with the same exploration strategies etc. (I want to
test how much the problem lies with solving Bellman.)

\begin{enumerate}

\item Reinforce (online version). High variance. All generalization comes
from the policy. Therefore the policy had better have good features. Thus
the performance of Reinforce may improve significantly by providing
enhanced features to the policy, eg features coming from an auto-encoder
or world model. (I think this applies to all algorithms, actually, and
could be quite important.)

\item TD(1) then policy gradient or a variant of policy gradient.

\item TD(1) trained on a model of the reward, then some form of policy
gradient.

\item 4?
\end{enumerate}

Baselines that should not (in principle) pass to the limit:
\begin{itemize}
\item All fashionable deep RL algorithms, DQN, TRPO, PPO,...
\item DPG when applicable
\end{itemize}

\end{document}

### 2018-11-30 pendulum_exp/envs/hill.py
Correct hill env

@@ -8,30 +8,58 @@ import numpy as np
 from abstract import Env
 
 class HillEnv(gym.Env, Env):
-    def __init__(self):
-        self.observation_space = Box(low=-1, high=1, shape=(1,), dtype=np.float32)
-        self.action_space = Discrete(2)
+    def __init__(self, discrete: bool=True) -> None:
+        if discrete:
+            self._action_space = Discrete(2)
+        else:
+            self._action_space = Box(low=-1, high=1, shape=(1,), dtype=np.float32)
+        self._observation_space = Box(low=-1, high=1, shape=(1,), dtype=np.float32)
         self._x = None
         self.dt = .1
         self.seed()
 
-    def step(self, action):
-        assert self.action_space.contains(action)
-        done = False
-        if self._x[0] < 0:
-            done = True
-        elif self._x[0] < 10 * self.dt:
-            self._x = self._x - self.dt
-        else:
-            self._x += 2 * (action - 1) * self.dt
+    @property
+    def action_space(self):
+        return self._action_space
 
-        reward = 1
+    @property
+    def observation_space(self):
+        return self._observation_space
+
+    def action(self, action):
+        if isinstance(self._action_space, Box):
+            return action
+        else:
+            return 2 * action - 1
+
+    def step(self, action):
+        done = False
+
+        # perform action
+        if -1 < self._x[0] < -.25:
+            # if -1 < x < -.25, action has no impact
+            # we go left
+            self._x[0] = self._x[0] - self.dt
+        elif self._x[0] >= 1.:
+            self._x[0] = 1.
+        else:
+            self._x += self.action(action) * self.dt
+
+        # compute reward: if x < 0, r = 1, 0 otherwise
+        if self._x[0] < 0:
+            reward = 1
+        else:
+            reward = 0
+
+        # compute termination: if x <= -1, done
+        if self._x[0] <= -1:
+            done = True
 
         return self._x, reward, done, {}
 
     def reset(self):
         np.random.seed()
-        self._x = np.clip(np.random.normal(0, 1, (1,)), self.dt, 1)
+        self._x = np.clip(np.random.normal(0, 1, (1,)), 0, 1)
         return self._x
 
     def close(self):

### 2018-11-30 paper/tangotd.tex

@@
 
-Combined with all this, is the problem that a Bellman solution on the
-support of $\rho$ may be very, very far from the global Bellman solution,
-even on the support of $\rho$ (take random walk on a circle, with $\rho$
-being everything but one point: then any linear function solves Bellman
-on the support of $\rho$).
-
 The minimizer of the $L^2(\rho^2)$ error between $R$
@@
 \end{equation}
-however, the $\rho^2$ is a bit weird... moreover having
-$\transp{\Delta}\Delta$ means we are just facing the double sampling
+however, having
+$\transp{\Delta}\rho^2\Delta$ means we are just facing the double sampling
 problem (we can sample $\transp{(\rho\Delta)}\rho\Delta$, but in practice the sample will always be $0$
@@
 
-The minimizer of a weird norm $L^2(\rho \Phi\transp{\Phi}\rho)$ is better
-samplable, because now it's enough to sample states \emph{with the same
+If one has access to some features $\Psi$,
+the minimizer of a weird norm $L^2(\rho \Psi\transp{\Psi}\rho)$ is better
+samplable, because now in double sampling it's enough to sample states \emph{which share
+some
 features}, not just the exact same state. This is basically what happens
-in the linear case. The features should not be orthogonal. Should use a
+in the linear case, with $\Psi=\Phi$. In the nonlinear case taking
+$\Psi=\Phi$ is not acceptable because $\Phi$ depends on $\theta$, so the
+objective is not a fixed norm to be minimized. The features $\Psi$ should not be orthogonal. Should use a
 hash of states instead? Or apply ReLU to orthogonal features.
@@
 correction that vanishes when $\Delta V-R$ is small.
+
+[Combined with all these problems, is the problem that a Bellman solution on the
+support of $\rho$ may be very, very far from the global Bellman solution,
+even on the support of $\rho$ (take random walk on a circle, with $\rho$
+being everything but one point: then any linear function solves Bellman
+on the support of $\rho$).]
 

### 2018-12-04 paper/tangotd.tex

@@
 
-\paragraph{Minimizers of errors.}
-A local minimizer of the $L^2(\rho)$ error between $V$ and
-$(\Id-P)^{-1}R$ solves
-\begin{equation}
-\transp{\Phi} \rho V=\transp{\Phi}\rho (\Id-P)^{-1} R
-\end{equation}
-which is not quite the same as the above. A local minimizer of the
-$L^2(\rho)$ error between $(\Id-P)V$ and $R$ is
-\begin{equation}
-\transp{\Phi}\transp{(\Id-P)}\rho(\Id-P)V=\transp{\Phi}\transp{(\Id-P)}\rho R
-\end{equation}
-and in the linear case $V=\Phi\theta$, this is the standard solution to
-the linear regression problem $(\Id-P)\Phi\theta\approx R$ in
-$L^2(\rho)$. However, the matrix on the left is not samplable. We need
-double sampling for that.
-
-The minimizer of the $L^2(\rho^2)$ error between $R$
-and $(\Id-P)V$ is samplable:
-\begin{equation}
-\transp{\Phi}\transp{(\Id-P)}\rho^2(\Id-P)V=\transp{\Phi}\transp{(\Id-P)}\rho^2
-R
-\end{equation}
-however, having
-$\transp{\Delta}\rho^2\Delta$ means we are just facing the double sampling
-problem (we can sample $\transp{(\rho\Delta)}\rho\Delta$, but in practice the sample will always be $0$
-unless we are lucky and select two transitions $s\to s'$ that have the
-same $s$). Even the right-hand-side involves double sampling.
-
-If one has access to some features $\Psi$,
-the minimizer of a weird norm $L^2(\rho \Psi\transp{\Psi}\rho)$ is better
-samplable, because now in double sampling it's enough to sample states \emph{which share
-some
-features}, not just the exact same state. This is basically what happens
-in the linear case, with $\Psi=\Phi$. In the nonlinear case taking
-$\Psi=\Phi$ is not acceptable because $\Phi$ depends on $\theta$, so the
-objective is not a fixed norm to be minimized. The features $\Psi$ should not be orthogonal. Should use a
-hash of states instead? Or apply ReLU to orthogonal features.
-
-The Dirichlet norm associated with $\rho$ is $\E_{ss'}
-(f(s)-f(s'))^2$. Its matrix is $\rho+\rho'-\rho P-\transp{(\rho P)}$ where
-$\rho'$ is the distribution of $s'$ (and is samplable).
-A local minimizer of the Dirichlet error between $V$ and
-$(\Id-P)^{-1}R$ solves
-\begin{equation}
-\transp{\Phi} (\rho+\rho'-\rho P-\transp{(\rho P)})(V-(\Id-P)^{-1}R)=0
-\end{equation}
-If $\rho$ is the stationary distribution $\mu$ and $P$ is reversible, one
-has $\transp{(\mu P)}=\mu P$ and $\rho'=\mu$ so $\rho+\rho'-\rho P-\transp{(\rho
-P)}=2\mu(\Id-P)$ and this is equivalent to
-\begin{equation}
-\transp{\Phi}\mu ((\Id-P)V-R)=0
-\end{equation}
-namely, the same as a fixed point of TD.
-
-Instead of applying the Dirichlet norm to $(V-(\Id-P)^{-1}R)$, one can
-apply the inverse Dirichlet norm to $((\Id-P)V-R)$. When $P$ is
-reversible and $\rho=\mu$ this
-is the same. One tries to minimize
-$\transp{((\Id-P)V-R)}\rho\Dir^{-1}\rho((\Id-P)V-R)$ where
-$\Dir\deq \rho+\rho'-\rho P-\transp{(\rho P)}$ as above, and use
-TANGO to invert $\Dir$. However this suffers again from a double
-sampling problem, in several places. Still, this can be rewritten as a
-correction to TD where only the correction suffers from double
-sampling/large variance. Namely, the gradient of the above with respect
-to $\theta$ is $\transp{\Phi}\transp{(\rho\Delta)} \Dir^{-1}\rho(\Delta V-R)$. Since
-$\rho\Delta$ is half of $\Dir$, this gradient is 
-\begin{gather}
-\transp{\Phi}\transp{(\rho\Delta)} \Dir^{-1}\rho(\Delta V-R)=
-\\
-\frac12\transp{\Phi}(\transp{(\rho \Delta)}+\rho'-\rho P)\Dir^{-1}\rho(\Delta
-V-R)+\frac12 \transp{\Phi}(\transp{(\rho \Delta)}+\rho P-\rho')\Dir^{-1}\rho(\Delta
-V-R)
-\\=\frac12\transp{\Phi}\rho(\Delta
-V-R)+\frac12 \transp{\Phi}(\transp{(\rho \Delta)}+\rho P-\rho')\Dir^{-1}\rho(\Delta
-V-R)
-\end{gather}
-the first part of which is TD. Thus one can write a correction to TD such
-that only the correction suffers from double-sampling. The correction vanishes for reversible
-chains. The double sampling can be simulated with high variance in finite
-state spaces (just take samples from $P$ and $\transp{P}$ until they
-match) but does not make sense in continuous state spaces. Use a model of
-transitions only for the correction? TODO: use the same trick on the
-right? No, because the $R$ term doesn't go away and I'd rather have a
-correction that vanishes when $\Delta V-R$ is small.
-
-[Combined with all these problems, is the problem that a Bellman solution on the
-support of $\rho$ may be very, very far from the global Bellman solution,
-even on the support of $\rho$ (take random walk on a circle, with $\rho$
-being everything but one point: then any linear function solves Bellman
-on the support of $\rho$).]
-
-\bigskip
-
-TD is the only one of those fixed point equations that is
-\emph{samplable} given that we can sample $\rho$, $\rho P$ and $\rho R$.
 Thus, from now on we are going to focus on solving
@@
 \end{equation}
-using samples for $A$ and $b$.
-This is TD. If $A$ is symmetric positive definite, this is also the
+using samples for $A$ and $b$: the only fixed point solves $A\theta=b$.
+TD does exactly this, 
+with $A=\transp{\Phi}\rho(\Id-P)\Phi$ and $b=\transp{\Phi}\rho R$.
+
+If $A$ is symmetric positive definite, this is also the
 gradient descent of the loss function
@@
 \item All the eigenvalues of $A$ have positive real part.
-\item The matrix $(\Id-\eps A)$ has spectral radius $<1$ for small enough
-$\eps$.
-\item For $\eps$ small enough, iterating $\theta\gets (\Id-\eps A)\theta$
+\item The matrix $(\Id-\eta A)$ has spectral radius $<1$ for small enough
+$\eta$.
+\item For $\eta$ small enough, iterating $\theta\gets (\Id-\eta A)\theta$
 converges to $0$.
@@
 
+\paragraph{Minimizers of errors.} Rather than the fixed point of TD, one
+can look for minimizers of the error in various norms.
+
+A local minimizer of the $L^2(\rho)$ error between $V$ and
+$(\Id-P)^{-1}R$ solves
+\begin{equation}
+\transp{\Phi} \rho V=\transp{\Phi}\rho (\Id-P)^{-1} R
+\end{equation}
+which is not quite the same as the fixed point of TD. We have no access
+to the right-hand-side, indeed, computing $(\Id-P)^{-1} R$ is exactly the
+original problem.
+
+A local minimizer of the
+$L^2(\rho)$ error between $(\Id-P)V$ and $R$ is
+\begin{equation}
+\transp{\Phi}\transp{(\Id-P)}\rho(\Id-P)V=\transp{\Phi}\transp{(\Id-P)}\rho R
+\end{equation}
+and in the linear case $V=\Phi\theta$, this is the standard solution to
+the linear regression problem $(\Id-P)\Phi\theta\approx R$ in
+$L^2(\rho)$. However, the matrix on the left is not samplable: sampling
+$\transp{(\Id-P)}\rho(\Id-P)$ is exactly equivalent to double sampling
+from $s\sim \rho$.
+Indeed, $\transp{P}\rho P$ is equal to $\E[\1_{s' s''}]$ where $s'$ and
+$s''$ are two independent states from a transition $s\to s'$ with $s\sim
+\rho$.
+
+The minimizer of the $L^2(\rho^2)$ error between $R$
+and $(\Id-P)V$ is samplable:
+\begin{equation}
+\transp{\Phi}\transp{(\Id-P)}\rho^2(\Id-P)V=\transp{\Phi}\transp{(\Id-P)}\rho^2
+R
+\end{equation}
+Indeed, denoting $\Delta\deq \Id-P$, we can sample $\rho \Delta$ and
+therefore we can take two samples from $\rho \Delta$ which provides a
+sample from $\transp{(\rho\Delta)}\rho\Delta$. However, the double sampling
+problem resurfaces: Most of the time, taking the product
+$\transp{(\rho\Delta)}\rho\Delta$ from two samples of $\rho
+\Delta$ is $0$ unless we are lucky enough to sample two transitions $s\to
+s'$ with the same $s$.
+Even the right-hand-side $\rho^2 R$ involves double sampling.
+
+If one has access to some features $\Psi$, this can be mitigated:
+the minimizer of the norm $L^2(\rho \Psi\transp{\Psi}\rho)$ is better
+samplable, because now in double sampling it's enough to sample states \emph{which share
+some
+features}, not just the exact same state. This is basically what happens
+in the linear case, with $\Psi=\Phi$. In the nonlinear case taking
+$\Psi=\Phi$ is not acceptable because $\Phi$ depends on $\theta$, so the
+objective is not a fixed norm to be minimized. The features $\Psi$ should not be orthogonal. Should use a
+hash of states instead? Or apply ReLU to orthogonal features.
+
+The Dirichlet norm associated with $\rho$ is $\E_{ss'}
+(f(s)-f(s'))^2$. Its matrix is $\rho+\rho'-\rho P-\transp{(\rho P)}$ where
+$\rho'$ is the distribution of $s'$ (and is samplable).
+A local minimizer of the Dirichlet error between $V$ and
+$(\Id-P)^{-1}R$ solves
+\begin{equation}
+\transp{\Phi} (\rho+\rho'-\rho P-\transp{(\rho P)})(V-(\Id-P)^{-1}R)=0
+\end{equation}
+If $\rho$ is the stationary distribution $\mu$ and $P$ is reversible, one
+has $\transp{(\mu P)}=\mu P$ and $\rho'=\mu$ so $\rho+\rho'-\rho P-\transp{(\rho
+P)}=2\mu(\Id-P)$ and this is equivalent to
+\begin{equation}
+\transp{\Phi}\mu ((\Id-P)V-R)=0
+\end{equation}
+namely, the same as a fixed point of TD.
+
+Instead of applying the Dirichlet norm to $(V-(\Id-P)^{-1}R)$, one can
+apply the inverse Dirichlet norm to $((\Id-P)V-R)$. When $P$ is
+reversible and $\rho=\mu$ this
+is the same. One tries to minimize
+$\transp{((\Id-P)V-R)}\rho\Dir^{-1}\rho((\Id-P)V-R)$ where
+$\Dir\deq \rho+\rho'-\rho P-\transp{(\rho P)}$ as above, and use
+TANGO to invert $\Dir$. However this suffers again from a double
+sampling problem, in several places. Still, this can be rewritten as a
+correction to TD where only the correction suffers from double
+sampling/large variance. Namely, the gradient of the above with respect
+to $\theta$ is $\transp{\Phi}\transp{(\rho\Delta)} \Dir^{-1}\rho(\Delta V-R)$. Since
+$\rho\Delta$ is half of $\Dir$, this gradient is 
+\begin{gather}
+\transp{\Phi}\transp{(\rho\Delta)} \Dir^{-1}\rho(\Delta V-R)=
+\\
+\frac12\transp{\Phi}(\transp{(\rho \Delta)}+\rho'-\rho P)\Dir^{-1}\rho(\Delta
+V-R)+\frac12 \transp{\Phi}(\transp{(\rho \Delta)}+\rho P-\rho')\Dir^{-1}\rho(\Delta
+V-R)
+\\=\frac12\transp{\Phi}\rho(\Delta
+V-R)+\frac12 \transp{\Phi}(\transp{(\rho \Delta)}+\rho P-\rho')\Dir^{-1}\rho(\Delta
+V-R)
+\end{gather}
+the first part of which is TD. Thus one can write a correction to TD such
+that only the correction suffers from double-sampling. The correction vanishes for reversible
+chains. The double sampling can be simulated with high variance in finite
+state spaces (just take samples from $P$ and $\transp{P}$ until they
+match) but does not make sense in continuous state spaces. Use a model of
+transitions only for the correction? TODO: use the same trick on the
+right? No, because the $R$ term doesn't go away and I'd rather have a
+correction that vanishes when $\Delta V-R$ is small.
+
+[Combined with all these problems, is the problem that a Bellman solution on the
+support of $\rho$ may be very, very far from the global Bellman solution,
+even on the support of $\rho$ (take random walk on a circle, with $\rho$
+being everything but one point: then any linear function solves Bellman
+on the support of $\rho$).]
+
 \end{document}

### 2018-12-04 paper/tangotd.tex

@@
 
-Still, this is stable
-on-policy (take $L=B^{-1}$, given that $A$ is stable on-policy) or assuming overparameterization. (Indeed in the latter case we
-are back to the tabular case.)
+Still, this is stable in two cases:
+on-policy (take $L=B^{-1}$, given that $A$ is stable on-policy), or
+assuming overparameterization ($\Phi$ invertible: then we can take
+$L=\transp{\Phi}\mu \Phi$ so that $LBA=\transp{\Phi}\mu (\Id-P)\Phi$
+which is positive because $\mu(\Id-P)$ is positive; essentially, if
+$\Phi$ is invertible we are back to the tabular case.)
 % Then we have to prove th

### 2018-12-04 pendulum_exp/policies/mixture.py
Slightly different version of mixture

@@ -67,9 +67,15 @@ class ContinuousAdvantageMixturePolicy(SharedAdvantagePolicy):
             adv - max_adv,
             adv - torch.stack([max_adv[..., 1], max_adv[..., 0]], dim=-1)],
             dim=-1) # (b, 4)
+
+        # hacky hack to multiply by 1. / dt the contribution of points with high loss
+        # to the change of parameters in A
+        max_adv[..., 0] = (1 - 1 / self._dt) * max_adv[..., 0].detach() + max_adv[..., 0] / self._dt
+        adv[..., 0] = (1 - 1 / self._dt) * adv[..., 0].detach() + adv[..., 0] / self._dt
+
         sigmas = arr_to_th([[
-            np.sqrt(2 * self._dt), np.sqrt(2) * self._dt,
-            np.sqrt(self._dt + self._dt ** 2), np.sqrt(self._dt + self._dt ** 2)]], self._device) # (1, 4)
+            np.sqrt(2), np.sqrt(2) * self._dt,
+            np.sqrt(1 + self._dt ** 2), np.sqrt(1 + self._dt ** 2)]], self._device) # (1, 4)
         logpis = torch.cat([
             logpi + max_logpi,
             logpi + torch.stack([max_logpi[..., 1], max_logpi[..., 0]], dim=-1)],

### 2018-12-04 pendulum_exp/eval.py
Allowing nolog for monitoring

@@ -55,7 +55,10 @@ if __name__ == '__main__':
 
     policy_config, _, eval_noise_config, env_config = read_config(args)
 
-    logto(args.logdir)
+    try:
+        logto(args.logdir)
+    except Exception:
+        pass # Errors can occur with pickle. Still try to output video even if pickle is corrupted
     main(
         logdir=args.logdir, hidden_size=args.hidden_size, nb_layers=args.nb_layers,
         nb_eval_env=args.nb_eval_env, policy_config=policy_config,

### 2018-12-04 pendulum_exp/policies/mixture.py
Fixing inplace op

@@ -70,8 +70,12 @@ class ContinuousAdvantageMixturePolicy(SharedAdvantagePolicy):
 
         # hacky hack to multiply by 1. / dt the contribution of points with high loss
         # to the change of parameters in A
-        max_adv[..., 0] = (1 - 1 / self._dt) * max_adv[..., 0].detach() + max_adv[..., 0] / self._dt
-        adv[..., 0] = (1 - 1 / self._dt) * adv[..., 0].detach() + adv[..., 0] / self._dt
+        max_adv = torch.stack(
+            [(1 - 1 / self._dt) * max_adv[..., 0].detach() + max_adv[..., 0] / self._dt,
+             max_adv[..., 1]], dim=-1)
+        adv = torch.stack(
+            [(1 - 1 / self._dt) * adv[..., 0].detach() + adv[..., 0] / self._dt,
+             adv[..., 1]], dim=-1)
 
         sigmas = arr_to_th([[
             np.sqrt(2), np.sqrt(2) * self._dt,

### 2018-12-05 pendulum_exp/models.py
Only normalize states

@@ -98,12 +98,14 @@ class NormalizedMLP(nn.Module, ParametricFunction):
     def __init__(self, model: ParametricFunction) -> None:
         super().__init__()
         self._model = model
-        self._bns = nn.ModuleList([CustomBN(*feat) for feat in self._model.input_shape()])
+        # only normalize first input (is this what we want to do in the long go?)
+        self._bn = CustomBN(self._model.input_shape()[0][0])
 
     def forward(self, *inputs: Tensorable):
         device = next(self.parameters())
         tens_inputs = [check_tensor(inp, device) for inp in inputs]
-        return self._model(*[bn(inp) for (bn, inp) in zip(self._bns, tens_inputs)])
+        tens_inputs = [self._bn(tens_inputs[0])] + tens_inputs[1:]
+        return self._model(*tens_inputs)
 
     def input_shape(self) -> Shape:
         return self._model.input_shape()

### 2018-12-05 paper/icml_drau/icml_drau.tex

%%%%%%%% ICML 2018 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{amsmath, amssymb}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables
\usepackage{xcolor}

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2018} with \usepackage[nohyperref]{icml2018} above.
\usepackage{hyperref}

\input{macros.tex}

% Use the following line for the initial blind version submitted for review:
\usepackage{icml2018}

% If accepted, instead use the following line for the camera-ready submission:
%\usepackage[accepted]{icml2018}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Deep Relativistic Advantage Updating}

\begin{document}

\twocolumn[
\icmltitle{There are no small Advantages:
	Reinforcement learning in near continuous time
	with Deep RelAtive Mixture of Advantages
}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2018
% package.

% List of affiliations: The first argument should be a (short)
% identifier you will use later to specify author affiliations
% Academic affiliations should list Department, University, City, Region, Country
% Industry affiliations should list Company, City, Region, Country

% You can specify symbols, otherwise they are numbered in order.
% Ideally, you should not use this facility. Affiliations will be numbered
% in order of appearance and this is the preferred way.
\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
%\icmlauthor{Aeiau Zzzz}{equal,to}
%\icmlauthor{Bauiu C.~Yyyy}{equal,to,goo}
%\icmlauthor{Cieua Vvvvv}{goo}
%\icmlauthor{Iaesut Saoeu}{ed}
%\icmlauthor{Fiuea Rrrr}{to}
%\icmlauthor{Tateu H.~Yasehe}{ed,to,goo}
%\icmlauthor{Aaoeu Iasoh}{goo}
%\icmlauthor{Buiui Eueu}{ed}
%\icmlauthor{Aeuia Zzzz}{ed}
%\icmlauthor{Bieea C.~Yyyy}{to,goo}
%\icmlauthor{Teoau Xxxx}{ed}
%\icmlauthor{Eee Pppp}{ed}
\end{icmlauthorlist}

%\icmlaffiliation{to}{Department of Computation, University of Torontoland, Torontoland, Canada}
%\icmlaffiliation{goo}{Googol ShallowMind, New London, Michigan, USA}
%\icmlaffiliation{ed}{School of Computation, University of Edenborrow, Edenborrow, United Kingdom}
%
%\icmlcorrespondingauthor{Cieua Vvvvv}{c.vvvvv@googol.com}
%\icmlcorrespondingauthor{Eee Pppp}{ep@eden.co.uk}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Reinforcement learning}

\vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[

% This command actually creates the footnote in the first column
% listing the affiliations and the copyright notice.
% The command takes one argument, which is text to display at the start of the footnote.
% The \icmlEqualContribution command is standard text for equal contribution.
% Remove it (just {}) if you do not need this facility.

%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.

\begin{abstract}

\end{abstract}

\section{Introduction}
\label{sec:intro}

\section{Related work}
\label{sec:related}

\section{Continuous reinforcement learning and scaling problems}
\label{sec:continous}
In the following section, a restrictive framework for near continuous
time reinforcement learning is introduced, and the main challenges of this
setup are explained. \RKC{This is not what we want to write: we want to explain
why current reinforcement learning practices are ill suited for this framework}

\subsection{Framework}
Consider a fully observable \emph{Markov Decision Process} defined by the following
transitions and rewards
\begin{align}
	s_{t + \deltat} &= s_t + F(s_t, a_t) \deltat + \Sigma(s_t, a_t) \deltabt\\
	r_{t} &= \reward(s_t, a_t) \deltat + \Sigma_r(s_t, a_t) \deltabt^r
\end{align}
where the $s$'s are $n$-dimensional vectors, $F(s_t, a_t)$ is a $n$-dimensional
drift coeficient, $\Sigma(s_t, a_t)$ is a $n \times n$ diffusion matrix,
$\deltat$ is a small but finite discretization step, and $\deltabt$ is a sample
from $\gauss(0, \deltat I)$.

This framework is quite general, and most discretized deterministic continuous
control environments fall in this framework, with $\Sigma = 0$ and $\Sigma_r =
0$, as well as many stochastic control environments.

A deterministic policy $\pi$ is defined as a deterministic mapping from state
space to action space\footnote{We restrict ourselves to deterministic policies.
	In continuous time, defining stochastic policies with well defined
	state-value function is non trivial. In the discrete case, the
distribution of actions of a policy given a state only depends on the previous
state, and not on previous actions. In continuous time, such policies are akin
to white noise, and do not define stochastic processes.}.

Under a deterministic policy $\pi$, one can define the \emph{state value function} $V^\pi(s)$
and \emph{state action value function} $Q^\pi(s, a)$ as
\begin{align}

\end{align}

\section{Multi-chain continuous environments and biscale advantage}
\label{sec:multi-chain}

\section{Deep RelAtivistic Mixture of Advantage learning}
\label{sec:algo}

\section{Experiments}
\label{sec:exp}

\section{Conclusion}
\label{sec:concl}

\bibliography{icml_drau.bib}
\bibliographystyle{icml2018}
\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
% by Iain Murray in 2018. It was modified from a version from Dan Roy in
% 2017, which was based on a version from Lise Getoor and Tobias
% Scheffer, which was slightly modified from the 2010 version by
% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
% slightly modified from Prasad Tadepalli's 2007 version which is a
% lightly changed version of the previous year's version by Andrew
% Moore, which was in turn edited from those of Kristian Kersting and
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2018-12-06 paper/tangotd.tex

@@
 \begin{document}
+
+\section{Recovering known convergence results for TD, and a few ideas}
 
@@
 obviously stable. This helps with $A$ being non-symmetric. TODO: check if
-GTD uses the same sample for $\tilde{A}$ and $A$. If so, it will
+GTD uses the same sample for $\transp{A}$ and $A$. If so, it will
 converge to a different stable point. TODO: Can be applied TANGO-style
@@
 
-[Combined with all these problems, is the problem that a Bellman solution on the
+\section{A function that almost solves the Bellman equation can be very
+far from the true Bellman function}
+
+Combined with all these problems, is the problem that a Bellman solution on the
 support of $\rho$ may be very, very far from the global Bellman solution,
@@
 being everything but one point: then any linear function solves Bellman
-on the support of $\rho$).]
+on the support of $\rho$).
+
+TODO seems to be much worse off-policy.
+
+\section{Storing a summary of possible transitions?}
+
+The derivative of the average reward with respect to the policy
+parameters $\psi$ is equal to
+\begin{align}
+\partial_\psi \left(\E_\mu r\right)&=
+\E_{s,a,s'} \partial_\psi \log \pi(a|s)\left(
+r+V(s')
+\right)
+\\&=\E_{s,a}\partial_\psi \log \pi(a|s)\E_{s'|s,a}\sum_{k\geq 0}P^k R(s')
+\\&=\E_{s,a,s'}\partial_\psi \log \pi(a|s)(\Delta^{-1}R)(s')
+\end{align}
+by expanding $V(s')$ as a sum of expected future rewards (and attributing
+the reward of a transition $s\to s'$ to $s'$, which makes sense if $s'$
+is a sufficient statistic).
+(Here I'm neglecting constant functions again.)
+
+Thus if we denote
+\begin{equation}
+g(s')\deq \E_{s,a,s'}\partial_\psi \log \pi(a|s)\1_{s'}
+\end{equation}
+we have
+\begin{equation}
+\partial_\psi \left(\E_\mu r\right)=\transp{g}\Delta^{-1}R
+\end{equation}
+Let us define the row vector
+\begin{equation}
+G\deq \transp{g}\Delta^{-1},\qquad \partial_\psi \left(\E_\mu
+r\right)=GR
+\end{equation}
+then one has
+\begin{equation}
+G(s)=\partial_\psi \mu(s)
+\end{equation}
+namely, $G$ encodes the derivative of the invariant distribution with
+respect to the policy parameters. Thus, if we find a state $s$ with high
+reward $R(s)$, $G$ tells us how we can visit $s$ more often. This
+explains that the derivative of the average reward is $GR$.
+
+Eligibility traces are a random, unbiased estimator of $G$: if $s_t$ is a
+trajectory from the current policy then
+\begin{equation}
+\E(\sum_{k\geq 1}\partial_\psi \pi(a_{t-k}|s_{t-k})\1_{s_t})=G
+\end{equation}
+Moreover $G$ satisfies a backward Bellman equation on-policy,
+\begin{equation}
+G=\E_{s,a,s'} \partial_\psi \pi(a|s)\1_{s'}+GP
+\end{equation}
+thanks to $\Delta^{-1}=\sum P^k=\Id+\Delta^{-1}P$. Eligibility traces are
+updated this way at each step, via a sampling of $P$.
+
+$G$ cannot be stored but one can store an approximate model. For instance
+one could approximate $G$ as a sum of parameters times functions:
+\begin{equation}
+G\approx \sum_i \psi_i \phi_{\theta_i}(s)
+\end{equation}
+where each $\psi_i$ is a policy parameter, and each $\phi_{\theta_i}$ is
+a function of the state parameterized by some model. Then the reverse
+Bellman equation above allows us to learn $\psi_i$ and $\theta_i$ via 
+a TD-like algorithm.
+
+TODO: not the right representation, plus works only on-policy. Maybe learn a function $a(s,s')$ that
+returns the first action to be taken if we want to reach $s'$ from $s$?
 

### 2018-12-06 paper/icml_drau/icml_drau.tex
Biblio for paper

@@
 %\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
-\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
+%\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
 
@@
 \label{sec:related}
+\cite{adv_upd}
+\cite{adv_learn}
+\cite{dueling_nets}
 
@@
 \label{sec:continous}
-In the following section, a restrictive framework for near continuous
+In the following section, a restricted framework for near continuous
 time reinforcement learning is introduced, and the main challenges of this
@@
 
-Under a deterministic policy $\pi$, one can define the \emph{state value function} $V^\pi(s)$
-and \emph{state action value function} $Q^\pi(s, a)$ as
+Under a deterministic policy $\pi$, one can define the $\gamma$ discounted
+\emph{state value function} $V^\pi(s)$ and \emph{state action value function}
+$Q^\pi(s, a)$\footnote{
+	The values of $V^\pi$ and $Q^\pi$ are derived as discretization of e.g.
+	$V^\pi_\text{cont}(s) = \E\left[\int\limits_0^\infty \gamma^t \reward(s_t, \pi(s_t)) dt \mid s_0 = s\right]$.
+}as
 \begin{align}
+	V^\pi(s) &= \E\left[
+		\sum\limits_{k=0}^\infty \gamma^{k\deltat}
+		r_{k \deltat} \mid s_0 = s
+	\right]\\
+	Q^\pi(s, a) &= \E\left[
+		\sum\limits_{k=0}^\infty \gamma^{k\deltat}
+		r_{k \deltat} \mid s_0 = s, a_0 = a
+	\right].
+\end{align}
+Both $V^\pi$ and $Q^\pi$ verify Bellman equations, e.g.
+\begin{align}
+	V^\pi(s) = \reward(s, \pi(s)) \deltat + \gamma^{\deltat} \E_{s_{t+\deltat}\mid s_t=s} V^\pi(s_{t+\deltat}).
+\end{align}
 
-\end{align}
+\subsection{Scaling problems}
 

### 2018-12-06 paper/tangotd.tex

@@
 \item GTD is a gradient descent of
-$\transp{(A\theta-b)}(A\theta-b)$ (Sutton 2009).
+$\transp{(A\theta-b)}(A\theta-b)$ (Sutton 2009). This helps with $A$ being non-symmetric.
 This corresponds to $B=\transp{A}$. Then
 $BA=\transp{A}A$ is
-obviously stable. This helps with $A$ being non-symmetric. TODO: check if
+obviously stable, hence convergence.
+
+TODO: check if
 GTD uses the same sample for $\transp{A}$ and $A$. If so, it will
@@
 
-Thus if we denote
+Thus if we define the vector
 \begin{equation}
-g(s')\deq \E_{s,a,s'}\partial_\psi \log \pi(a|s)\1_{s'}
+g\deq \E_{s,a,s'}\partial_\psi \log \pi(a|s)\1_{s'}
 \end{equation}
@@
 
+[$g$ sums to $0$ on-policy, so we don't have to worry about removing
+constants and can directly take $\gamma=1$: $\transp{g}\Delta^{-1}$ is
+well-defined.]
+
 Eligibility traces are a random, unbiased estimator of $G$: if $s_t$ is a
@@
 \begin{equation}
-\E(\sum_{k\geq 1}\partial_\psi \pi(a_{t-k}|s_{t-k})\1_{s_t})=G
+\E\sum_{k\geq 1}\partial_\psi \log \pi(a_{t-k}|s_{t-k})\1_{s_t}=G
 \end{equation}
@@
 \begin{equation}
-G=\E_{s,a,s'} \partial_\psi \pi(a|s)\1_{s'}+GP
+G=\E_{s,a,s'} \partial_\psi \log \pi(a|s)\1_{s'}+GP
 \end{equation}
 thanks to $\Delta^{-1}=\sum P^k=\Id+\Delta^{-1}P$. Eligibility traces are
+precisely
 updated this way at each step, via a sampling of $P$.

### 2018-12-10 paper/tangotd.tex

@@
 
+\paragraph{Learning navigation.} Define a function $V(s,s')$ that
+represents the time spent in $s'$ if starting at $s$ and acting optimally
+towards $s'$. It is defined by the Bellman equation
+\begin{equation}
+V(s,s')=\1_{s=s'}+\gamma \max_a \E_{s_1|s,a} V(s_1,s')
+\end{equation}
+and also satisfies
+\begin{equation}
+V(s,s')=\1_{s=s'}+\gamma \sum_{s_{t-1}} V(s,s_{t-1})\max_a
+\E_{s_t|s_{t-1},a}\1_{s_t=s'}
+\end{equation}
+although it's not clear to me how to exploit the last expression. But I
+think we have to work from both ends if we want to represent
+compositionality of actions... (With the first equation, different values
+of $s'$ are just completely separate Bellman problems; the generalization
+abilities of the function approximator for $V$ may help, but only so
+much.)
+
+Then we may learn a goal-oriented policy $\pi_\theta(a|s,s')$ that returns an
+action $a$ at state $s$ which produces a move towards $s'$. Learning
+$\pi_\theta$ may be done by observing transitions and maximizing
+$V(s,s')$.
+
+It's not clear either how to exploit $V(s,s')$ to maximize a reward. If
+we were considering a fixed policy we would have
+$V(s)=\sum_{s'}V(s,s')R(s')$ (namely $V(s,s')$ would be the value
+function of the reward function $\1_{s'}$). But here $V(s,s')$ uses a
+strategy that depends on the goal $s'$.
+
 \end{document}

### 2018-12-10 paper/tangotd.tex

@@
 
-\paragraph{Learning navigation.} Define a function $V(s,s')$ that
+\paragraph{Learning to navigate.} Define a function $V(s,s')$ that
 represents the time spent in $s'$ if starting at $s$ and acting optimally
@@
 
+Another possibility is to learn a goal-oriented policy towards desired
+\emph{features} of the final state $s'$. Given a basis $(\Psi_i)$ of
+features, and a set of coefficient $\alpha_i$, one might want
+$\pi_\theta(a|s,(\alpha_i))$ to produce a policy that maximizes the
+expected reward $\sum \alpha_i \Psi_i$ (thus, simultaneously learning to
+solve a number of RL problems). Then we can learn an approximate
+decomposition of the actual reward over the $\Psi_i$, and we know what to
+do. (This assumes the reward is well approximated by the $\Psi_i$; or
+that we learn the complement.) There are many ways to build interesting
+$\Psi_i$'s: hidden features of a world model/reward model, eigenfunctions
+of $P$, or just setting a basis of random time-dependent functions on a
+trajectory and smoothing them via a NN (``zeitgeist'').
+
 \end{document}

### 2018-12-10 paper/tangotd.tex

@@
 do. (This assumes the reward is well approximated by the $\Psi_i$; or
-that we learn the complement.) There are many ways to build interesting
+that we learn the complement.) Advantage = even if there is no reward
+signal for some time, we still learn to navigate the environment.
+
+There are many ways to build interesting
 $\Psi_i$'s: hidden features of a world model/reward model, eigenfunctions

### 2018-12-10 paper/tangotd.tex

@@
 expected reward $\sum \alpha_i \Psi_i$ (thus, simultaneously learning to
-solve a number of RL problems). Then we can learn an approximate
-decomposition of the actual reward over the $\Psi_i$, and we know what to
-do. (This assumes the reward is well approximated by the $\Psi_i$; or
-that we learn the complement.) Advantage = even if there is no reward
-signal for some time, we still learn to navigate the environment.
+solve a number of RL problems). 
+This can be done either via learning a value function $V(s,(\alpha_i))$
+then actor-critic, or via an eligibility-trace-like thing, but
+
+Then we can learn an approximate decomposition of the actual reward over
+the $\Psi_i$, and we know what to do. (This assumes the reward is well
+approximated by the $\Psi_i$; or that we learn the complement.) Advantage
+= even if there is no reward signal for some time, we still learn to
+navigate the environment.
 
@@
 
+(The relationship between $V(s,s')$ and $V(s,(\alpha_i))$ is: the latter
+is more general, by taking a full basis of Dirac features $\1_{s'}$. Then
+$V(s,(\alpha_i))$ does not just learn how to go to individual states, but
+how to maximize time spent in an arbitrary weighted sum of states. This
+is why we can later extend to an arbitrary reward function.)
+
 \end{document}

### 2018-12-10 paper/tangotd.tex

@@
 This can be done either via learning a value function $V(s,(\alpha_i))$
-then actor-critic, or via an eligibility-trace-like thing, but
+then actor-critic, or via an eligibility-trace-like thing (but then it
+has to be on-policy, for which policy??)
 

### 2018-12-10 paper/tangotd.tex

@@
 represents the time spent in $s'$ if starting at $s$ and acting optimally
-towards $s'$. It is defined by the Bellman equation
+towards $s'$ (just as if the reward was $\1_{s'}$). It is defined by an iterated expectimax to reach $s'$, and
+satisfies the Bellman equation
 \begin{equation}
@@
 \end{equation}
-and also satisfies
+and it also satisfies
 \begin{equation}
@@
 think we have to work from both ends if we want to represent
-compositionality of actions... (With the first equation, different values
+compositionality of actions. Indeed with the first equation, different values
 of $s'$ are just completely separate Bellman problems; the generalization
 abilities of the function approximator for $V$ may help, but only so
-much.)
+much. Each time we observe a transition $s\to s'$, we should be able to
+update the probabilities to reach $s'$ based on the probabilities to
+reach $s$.
 

### 2018-12-10 paper/tangotd.tex

@@
 
-Eligibility traces are a random, unbiased estimator of $G$: if $s_t$ is a
+Eligibility traces are a random, unbiased estimator of $G$: if
+$(s_t)_{t\in \Z}$ is a
 trajectory from the current policy then
@@
 Bellman equation above allows us to learn $\psi_i$ and $\theta_i$ via 
-a TD-like algorithm.
+a TD-like algorithm, \emph{on-policy}.
 
@@
 \end{equation}
-although it's not clear to me how to exploit the last expression. But I
+although it's not clear to me how to exploit the last expression. 
+\footnote{On-policy it works, I think, because on-policy we can sample
+reverse transitions. But I think in that case $V(s,s')$ has to be defined
+as a density wrt $\mu(s')$.\\Should also work if we maintain a low-rank
+estimate of $V$ as a sum of Diracs over a few $s'$.}
+But I
 think we have to work from both ends if we want to represent

### 2018-12-10 paper/tangotd.tex

@@
 \item On-policy: $\rho=\mu$, the invariant distribution. Then we know that $\rho(\Id-P)$ is positive
-(Dirichlet form). This implies that $A=\transp{\Phi}\rho (\Id-P)\Phi$ is
+(Dirichlet form: $\transp{f}\rho(\Id-P)f=\E(f(s_t)-f(s_{t+1}))^2$). This implies that $A=\transp{\Phi}\rho (\Id-P)\Phi$ is
 positive too. So we can take $L=\Id$ and we recover convergence of linear
@@
 inverse of something; for instance $B=A^{-1}$ solves the problem
-instantly... Two-timescale methods or TANGO-like algorithms can apply an
+instantly... Two-timescale methods or TANGO-like algorithms
+\footnote{TANGO looks much like a two-timescale algorithm, but is
+``almost one-timescale'' in that the faster timescale can be fixed and
+does not need to tend to $0$.} can apply an
 inverse of something to $A\theta-b$, via an auxiliary gradient descent.
@@
 Still, this is stable in two cases:
-on-policy (take $L=B^{-1}$, given that $A$ is stable on-policy), or
-assuming overparameterization ($\Phi$ invertible: then we can take
+\emph{on-policy} (take $L=B^{-1}$, given that $A$ is stable on-policy), or
+assuming \emph{overparameterization} ($\Phi$ invertible: then we can take
 $L=\transp{\Phi}\mu \Phi$ so that $LBA=\transp{\Phi}\mu (\Id-P)\Phi$
 which is positive because $\mu(\Id-P)$ is positive; essentially, if
-$\Phi$ is invertible we are back to the tabular case.)
+$\Phi$ is invertible we are back to the tabular case).
 % Then we have to prove th
@@
 
-\item The choice from Sutton 2009 "TDC" corresponds to a gradient descent
+\item The choice from Sutton 2009 "TDC" combines the previous two tricks
+and corresponds to a gradient descent
 of the projected norm of the Bellman gap on the subspace generated by
@@
 $B=\transp{A}(A\transp{A})^{-1}$. The advantage is that we now have to
-invert a stable matrix, so, better behaved with TANGO. However, variance
+invert $(A\transp{A})$ which is a stable matrix, so, better behaved with
+TANGO or two-timescale. However, variance
 will be terrible: this requires four independent samples from $A$.
@@
 can be symmetric if and only if $\rho=\mu$. Indeed, if $\rho \Delta$ is
-symmetric then $\rho=\rho'$ and $\rho$ is the stationary distribution.} We can use the
+symmetric then $\rho=\rho'$ so that $\rho$ is the stationary distribution.} We can use the
 Dirichlet form which amounts to $L\deq\mu(\mu\Delta)^{-1}\mu$; then
@@
 overparameterized case; it minimizes the Dirichlet norm.
-However, in general $S$ is not positive. $S$ is positive if $\rho=\mu$, so,
-we have a two-timescale algorithm for non-linear on-policy
-overparameterized TD.
+However, in general $S$ is not positive.
 
-(Still works for non-overparameterized if $A$ does not couple the image
+Still, $S$ is positive if $\rho=\mu$, so,
+we have a two-timescale algorithm for \emph{non-linear on-policy
+overparameterized TD}.
+
+(Still works for non-overparameterized if the antisymmetric part $A$ does not couple the image
 of $\Phi$ and its complement...)
@@
 
+See \emph{Universal value function approximators},
+Schaul--Horgan--Gregor--Silver ICML 2015
+
 \end{document}

### 2018-12-10 paper/tangotd.tex

@@
 e^{-tA}$.
+
+\begin{table}
+\begin{tabular}{cccc}
+Algorithm & Conditions & Preconditioner & Lyapunov
+\\
+\hline
+\emph{Linear $V$} \\
+TD	& On-policy & $B=\Id$ & $L=\Id$ \\
+TD & Tabular & $B=\Id$ & $L=\mu/\rho$ \\
+GTD & 2ts & $B=\transp{A}$ & $L=\Id$\\
+– & On-policy, 2ts & $B=(\transp{\Phi}\rho\Phi)^{-1}$ & $L=B^{-1}$ \\
+– & Overparameterized, 2ts & $B=(\transp{\Phi}\rho\Phi)^{-1}$ &
+$L=\transp{\Phi}\mu\Phi$ \\
+TDC, GTD2 & 2ts & $B=\transp{A}(\transp{\Phi}\rho\Phi)^{-1}$ &
+$L=\Id$ \\
+2nd-order & On-policy, 2ts & $B=A^{-1}$ & $L=\Id$\\
+2nd-order & 2ts & $B=\transp{A}(A\transp{A})^{-1}$ & $L=\Id$\\
+– & 2ts & $B=\transp{A}(A+\transp{A})^{-1}$ & $L=\Id$\\
+\hline
+\emph{Non-linear $V$} \\
+TD & On-policy, reversible & any $B\succ 0$ & $L=\mu(\mu(\Id-P))^{-1}\mu$
+\\
+– & On-policy, overparameterized, 2ts & $B=(\transp{\Phi}S\Phi)^{-1}$ &
+$L= \mu S^{-1} \mu$ \\
+Non-lin GTD2 & 2ts, different objective &
+$B=\transp{A}(\transp{\Phi}\rho\Phi)^{-1}$  & $L=\Id$
+\\\hline
+\end{tabular}
+\caption{Various cases of convergence of TD-like algorithms, the
+preconditioner $B$ for the TD step, and the matrix $L$ that proves
+convergence via the Lyapunov criterion. 2ts=two-timescale.}
+\end{table}
 
@@
 
+See \emph{Universal value function approximators},
+Schaul--Horgan--Gregor--Silver ICML 2015
+
 Then we can learn an approximate decomposition of the actual reward over
@@
 
-See \emph{Universal value function approximators},
-Schaul--Horgan--Gregor--Silver ICML 2015
-
 \end{document}

### 2018-12-10 paper/tangotd.tex

@@
 
-\begin{defi}
+\begin{defi}[ (Stable matrix)]
 A matrix $A$ is \emph{stable} if one of the following equivalent
@@
 \end{tabular}
+\label{tab:TDlist}
 \caption{Various cases of convergence of TD-like algorithms, the
@@
 If $A$ is stable, then iterating $\theta\gets \theta -\eta(A\theta-b)$ will
-converge.
+converge. (TODO: I will ignore the problem with constant functions for
+$\gamma=1$.)
 
-We know $A$ is stable in several cases: (TODO here I just ignore constant
-functions)
+We now review several cases in which $A$ is known to be stable, or cases
+in which a modified TD algorithm makes the iteration stable (by
+preconditioning the gradient descent by a matrix $B$). See also
+Table~\ref{tab:TDlist}.
+
 \begin{enumerate}

### 2018-12-10 paper/tangotd.tex

@@
 \end{equation}
+namely, if $v$ is the eligibility trace vector, then the expectation of
+$\transp{v}\1_{s_t}$ is $G$, at every time $t$.
+
 Moreover $G$ satisfies a backward Bellman equation on-policy,
@@
 $G$ cannot be stored but one can store an approximate model. For instance
-one could approximate $G$ as a sum of parameters times functions:
+one could approximate $G$ via a neural network with input $s$ and output
+in the policy parameter space. \footnote{The policy parameter space may
+be quite large. This means the neural network should have a small
+next-to-last layer. Indeed, with a next-to-last layer of size $r$,
+the weights of the last layer are represented by $r$ policy parameters.
+This corresponds to a rank-$r$ approximation of $G$,
+as a sum of parameters times functions:
 \begin{equation}
-G\approx \sum_i \psi_i \phi_{\theta_i}(s)
+G\approx \sum_i \psi_i \phi_i(s)
 \end{equation}
-where each $\psi_i$ is a policy parameter, and each $\phi_{\theta_i}$ is
-a function of the state parameterized by some model. Then the reverse
-Bellman equation above allows us to learn $\psi_i$ and $\theta_i$ via 
+where each $\psi_i$ is a policy parameter, and each $\phi_i$ is
+a function of the state, corresponding to a feature in the network.} Then the reverse
+Bellman equation above allows us to learn this model via
 a TD-like algorithm, \emph{on-policy}.

### 2018-12-11 paper/tangotd.tex

@@
 
-Combined with all these problems, is the problem that a Bellman solution on the
-support of $\rho$ may be very, very far from the global Bellman solution,
-even on the support of $\rho$ (take random walk on a circle, with $\rho$
-being everything but one point: then any linear function solves Bellman
-on the support of $\rho$).
+Combined with all these problems, is the problem that an approximate
+solution to the Bellman equation may be very, very far from the true
+Bellman solution.  This is an absolutely major problem off-policy.
 
-TODO seems to be much worse off-policy.
+In a finite MDP, the solution to the Bellman equation is unique. However,
+in infinite MDPs this is not the case at all. And even in finite MDPs,
+the non-uniqueness problem becomes an ill-conditioning problem: some
+functions are almost perfect solutions to the Bellman equation but can be
+arbitrarily far from the true value function.
+
+As an example, consider a one-dimensional problem with state $s\in \Z$.
+Assume the policy $\pi$ can only jump to a nearest neighbor or stay in
+place (with arbitrary probabilities). Then the Bellman equation on $V$
+is, at each state $s$, a linear equation involving $V(s)$, $V(s+1)$ and
+$V(s-1)$. It is easy to build many solutions by choosing arbitrary
+values for $V(0)$ and $V(1)$, and then just solving the Bellman equation
+in $V(s+1)$ for $s\geq 1$ and in $V(s-1)$ for $s\leq -1$.
+
+For instance, imagine that the policy is a nearest-neighbor random walk
+on $\Z$ (with probabilities $1/2$ to jump right or left), and that
+$\gamma=1$. Then any linear function $f$ satisfies $f=Pf$, therefore,
+linear functions lie in the kernel of the Bellman gap operator.
+Consequently, if $V$ solves the Bellman equation, then $V+f$ is another
+solution, for any linear function $f$.
+
+What happens with linear functions is that the system ``believes''
+there's a source of reward at infinity, and this belief cannot be
+falsified via the Bellman equation alone.
+
+(If the policy and environment are ergodic with invariant measure $\mu$,
+and if rewards lie in $L^2(\mu)$, then the true value function is the
+only solution of Bellman in $L^2(\mu)$, so penalizing the $L^2$-norm might
+help a bit. However, some simple examples below are immune to this solution.)
+
+Let us take another example. The state is $s\in \Z$. With nearest-neighbor transitions, the Bellman function of the optimal
+policy satisfies (with $\gamma=1$)
+\begin{equation}
+\label{eq:Vopt}
+V(s)=R(s)+\max_{s'\in \{s-1,s,s+1\}} V(s')
+\end{equation}
+
+For instance, if the reward is $0$ at $0$ and $-1$ everywhere else, the
+optimal policy moves towards $0$, and the value function is
+$V(s)=-\abs{s}$. However, it is easy to check that
+$V(s)=\abs{s}+\1_{s=0}$ also solves the Bellman equation. This solution
+believes in ``rewards at infinity''.
+
+This ``almost'' happens in a finite state space. For instance, take a
+state space $[-n;n]$ with reward $0$ at the boundaries $\pm n$, and
+reward $-1$
+everywhere else. The optimal policy goes to the nearest boundary point,
+and the value function is $V(s)=\abs{s}-n$, shaped like a ``V''.
+
+This solution is unique; however, there are near-solutions that violate
+the Bellman equation only at one or a few points. For instance, $-V$ is a
+solution everywhere except at $s=0$.
+
+In an off-policy situation where we sample transitions $(s,s')$ and apply
+TD at $(s,s')$, it would take a long time to realize that $-V$ is
+not the correct solution, and convergence to the true solution would be
+slow. This solution ``believes'' there is a good reward at $s=0$, and
+this belief is consistent as long as we do not actually reach $s=0$.
+
+This phenomenon is fully general and not specific to 1D problems; in any
+MDP one can build near-Bellman solutions
+that ``believe'' there is a reward at a state $s_0$ and solve Bellman
+exactly everywhere except at $s_0$. Off-policy these violations would not be
+detected. On-policy, the policy would move towards $s_0$ and detect the
+problem. I believe this is a serious limitation of off-policy approaches.
+
+Are these pseudo-solutions frequent, or does a good initialization move
+away from such solutions?
+The continuous-state limit is interesting for this.
+
+We can view $[-n;n]$ as a
+discretization of a problem with state $s\in [-1;1]$, where the state is
+discretized with step $\eps=1/n$. Then the continuous-state limit of the
+Bellman equation \eqref{eq:Vopt} for the optimal policy (with $\gamma=1$)
+is just
+\begin{equation}
+\abs{\partial V(s)/\partial s}=-R(s)
+\end{equation}
+(as one can see by expanding $V(s+\eps)$ and $V(s-\eps)$ in the equation
+$V(s)=R(s)\eps +\max \{V(s+\eps),V(s-\eps)\}$; for simplicity I have
+discarded the possibility to stay in place). [One gets a constraint that
+rewards are negative because in such a situation, the optimal policy
+moves to the state with highest reward, so with $\gamma=1$ the value
+function is well-defined only if one shifts the rewards by the maximum
+reward so that the maximum reward becomes $0$.]
+
+This equation is wildly underdetermined. For instance, if $V$ is a
+solution then $-V$ is also a solution...
 

### 2018-12-11 paper/tangotd.tex

@@
 This equation is wildly underdetermined. For instance, if $V$ is a
-solution then $-V$ is also a solution...
+solution then $-V$ is also a solution.
+
+More generally, in dim $>1$, an equation such as $\abs{\partial
+V(s)/\partial s}=-R(s)$ leaves total freedom as to the direction of the
+gradient of $V$. Thus, with a randomly initialized $V$, my guess is that
+early TD-like optimization would ``freeze'' upon random choices of local
+directions of the gradient of $V$, each corresponding to a local belief
+that ``the optimum lies in this direction''. Regions with different
+directions of $\partial
+V(s)/\partial s$ would meet on a $(D-1)$-dimensional set where Bellman
+would be violated, but Bellman would still be satisfied on a set of
+measure $1$.
+
+Thus, off-policy algorithms based on sampling random transitions $(s,s')$
+from a dataset would be slow to detect a problem, while on-policy (or
+partly on-policy) algorithms would ``follow the local belief'' until
+reaching a state where it becomes clear that this belief is unfounded, or
+until they reach infinity.
 

### 2018-12-11 paper/tangotd.tex

@@
 \documentclass[11pt,a4paper]{article}
+\usepackage[colorlinks=true,linkcolor=blue]{hyperref}
 \input{header.en.tex}
@@
 \section{A function that almost solves the Bellman equation can be very
-far from the true Bellman function}
+far from the true value function}
 
@@
 solution to the Bellman equation may be very, very far from the true
-Bellman solution.  This is an absolutely major problem off-policy.
+Bellman solution.  I believe this is an absolutely major problem off-policy.
 
@@
 values for $V(0)$ and $V(1)$, and then just solving the Bellman equation
-in $V(s+1)$ for $s\geq 1$ and in $V(s-1)$ for $s\leq -1$.
+in $V(s+1)$ for $s\geq 2$ and solving in $V(s-1)$ for $s\leq -1$.
 
 For instance, imagine that the policy is a nearest-neighbor random walk
-on $\Z$ (with probabilities $1/2$ to jump right or left), and that
+on $\Z$, with probabilities $1/2$ to jump right or left, and that
 $\gamma=1$. Then any linear function $f$ satisfies $f=Pf$, therefore,
@@
 Consequently, if $V$ solves the Bellman equation, then $V+f$ is another
-solution, for any linear function $f$.
+solution, for any linear function $f$. (This works in any dimension.)
 
-What happens with linear functions is that the system ``believes''
+What happens with linear functions is that the solution ``believes''
 there's a source of reward at infinity, and this belief cannot be
@@
 
-For instance, if the reward is $0$ at $0$ and $-1$ everywhere else, the
+For instance, if the reward is $0$ at $0$ and $-1$ everywhere else
+(a typical episodic problem of reaching a specific state as fast as
+possible), the
 optimal policy moves towards $0$, and the value function is
@@
 everywhere else. The optimal policy goes to the nearest boundary point,
-and the value function is $V(s)=\abs{s}-n$, shaped like a ``V''.
+and the value function is the distance to the boundary, $V(s)=\abs{s}-n$, shaped like a ``V''.
 
@@
 the Bellman equation only at one or a few points. For instance, $-V$ is a
-solution everywhere except at $s=0$.
+solution everywhere except at $s=0$. More generally one can start with a
+solution and ``flip'' it
+vertically in a segment; this creates a few points at which Bellman is
+violated, but Bellman is still satisfied at most points.
 
+Assume we start with the pseudo-solution $-V$ in this example.
 In an off-policy situation where we sample transitions $(s,s')$ and apply
@@
 that ``believe'' there is a reward at a state $s_0$ and solve Bellman
-exactly everywhere except at $s_0$. Off-policy these violations would not be
+exactly everywhere except at $s_0$. Off-policy, these violations would
+rarely be
 detected. On-policy, the policy would move towards $s_0$ and detect the
@@
 $V(s)=R(s)\eps +\max \{V(s+\eps),V(s-\eps)\}$; for simplicity I have
-discarded the possibility to stay in place). [One gets a constraint that
-rewards are negative because in such a situation, the optimal policy
+discarded the possibility to stay in place). \footnote{One gets a constraint that
+rewards are negative for technical reasons due to $\gamma=1$: in such a situation, the optimal policy
 moves to the state with highest reward, so with $\gamma=1$ the value
 function is well-defined only if one shifts the rewards by the maximum
-reward so that the maximum reward becomes $0$.]
+reward so that the reward of the terminal state becomes $0$. This is the
+typical setup for many episodic environments.}
 

### 2018-12-11 paper/tangotd.tex

@@
 gradient of $V$. Thus, with a randomly initialized $V$, my guess is that
-early TD-like optimization would ``freeze'' upon random choices of local
+TD-like optimization would quickly ``freeze'' upon locally random choices of 
 directions of the gradient of $V$, each corresponding to a local belief

### 2018-12-11 paper/tangotd.tex

@@
 \begin{document}
+
+This contains: Section 1: remarks on recovering the known convergence results
+for TD in a single framework, plus ideas on systematic variants. Section
+2: the problem of the existence of near-solutions to the Bellman equation
+that are very far from the true value function (I think this is quite
+serious off-policy). Section 3: Ideas on how, even if there's no reward
+signal, we should still learn how to navigate in the Markov chain and
+learn which actions/policy to use to go to which states (this extends
+eligibility traces in a systematic way).
 

### 2018-12-12 paper/tangotd.tex

--- 2018-12-11 paper/tangotd.tex
@@
 \begin{document}
+
+This contains: Section 1: remarks on recovering the known convergence results
+for TD in a single framework, plus ideas on systematic variants around
+two-timescale algorithms or improved ``almost-one-timescale'' algorithms.
+
+Section 2: the problem of the existence of near-solutions to the Bellman
+equation that are very far from the true value function (I think this is
+quite serious off-policy).
+
+Section 3: Ideas on how, even if there's no reward signal, we should
+still learn how to navigate in the Markov chain and learn which
+actions/policy to use to go to which states. This extends eligibility
+traces in a systematic way.
 
@@
 
-The derivative of the average reward with respect to the policy
+In this paragraph I'm trying to store a summary of ``which actions lead
+where'' in the long term. Indeed, the value function can be written as
+$\Delta^{-1}R$ where $R$ is the reward, and the derivative of the average
+reward wrt the policy parameters can be written, by the actor-critic
+formula, as $\transp{g}V=\transp{g}\Delta^{-1}R$ where $g$ contains the
+log-derivatives of the probabilities of actions leading to a given state
+(see below).
+The Bellman equation first computes $\Delta^{-1}R$. But we could also
+first compute $\transp{g}\Delta^{-1}$: this (large) object contains
+information about which policy parameters increase the stationary
+distribution of which states. This does not depend on a reward signal and
+can be learned in a non-supervised way by observing trajectories.
+Eligibility traces are an unbiased, rank-one approximation of this.
+
+Also, when observing a transition $s\to s'$, eligibility traces updates
+the info about how to reach $s'$, while TD updates the info about the
+reward when starting from $s$. I feel that somehow we have to use both:
+intuitively when solving a problem we work ``from both ends'', expanding
+both from the starting point and around the target.
+
+Precisely, the derivative of the average reward with respect to the policy
 parameters $\psi$ is equal to

### 2018-12-12 paper/tangotd.tex

@@
 
-Also, when observing a transition $s\to s'$, eligibility traces updates
-the info about how to reach $s'$, while TD updates the info about the
-reward when starting from $s$. I feel that somehow we have to use both:
+Also, when observing a transition $s\to s'$, eligibility traces update
+the info about how to reach $s'$, while TD(0) updates the info about the
+reward when starting from $s$. I feel that somehow we should use both:
 intuitively when solving a problem we work ``from both ends'', expanding

### 2018-12-12 paper/tangotd.tex

@@
 
+In the end we should perhaps just learn a representation of $\Delta^{-1}$
+in a hidden state that represents distributions on $s$ (trained to
+faithfully reproduce averages of functions). $\Delta^{-1}$ satisfies the
+two-sided equation $\Delta^{-1}=\Id+\Delta^{-1}P=\Id+P\Delta^{-1}$, so
+when observing a transition $s\to s'$ we both update the model of
+how to arrive at $s'$ and of what happens when starting at $s$.
+
 \paragraph{Learning to navigate.} Define a function $V(s,s')$ that

### 2018-12-12 paper/tangotd.tex

@@
 
+\paragraph{Double sampling problem: using a model for one of the
+samples.} Double sampling allows us to minimize the $L^2$ norm of the
+Bellman gap. This is only possible if we have model of the environment,
+that allows for sampling $s'$ given $s$ and $a$.
+
+However, using Bellman on a model is dangerous because we compute the
+value function of the model, and even if the model is accurate for one
+step $s\to s'$, it may quickly diverge when iterated. Therefore its value
+function may be quite different from the true value function. The model
+needs to be accurate at time horizon $1/(1-\gamma)$.
+
+But in double-sampling, one could also use one real transition and one
+simulated transition. We show here that this is equivalent, at first order,
+to learning the value function of an environment in which every step is
+either a real or simulated step with probability $1/2$. This should be a
+bit more stable, as the real transitions may keep the model from
+hallucinating completely irrealistic trajectories.
+
+The $L^2(\rho)$ norm of the Bellman gap is $\transp{(\Delta
+V-R)}\rho(\Delta V-R)$. Its gradient with respect to $V$ is
+$2\transp{\Phi}\transp{\Delta}\rho (\Delta V-r)$. The presence of
+$\transp{\Delta}\rho\Delta$ indicates double sampling; this operator is
+not samplable.
+
+Let us assume that we have a model $\tilde P$ of the operator $P$. Thus
+we can sample $\transp{\tilde \Delta}\rho \Delta$ (just sample one
+transition $s\to s'$ from $\rho P$, and a second transition $s\to \tilde
+s'$ from the same $s$). We can also sample from $\transp{\tilde
+\Delta}\rho R$. Moreover, \emph{if rewards during $s\to s'$ are
+independent from $s'$}, we can sample from $\transp{\Delta}\rho R$. From
+now on, I will assume that the reward collected during a transition $s\to
+s'$ depends only on $s$. \footnote{This means the state $s$ is a
+sufficient statistic. If the reward depends only on $s'$ instead, we can
+just
+shift the rewards by one step and pretend they are collected at the next
+step; therefore we are back to a reward that depends only on $s$. The
+only case not covered is if the reward really depends on the pair
+$(s,s')$; in that case, the state $s'$ is not quite a sufficient
+statistic of the process.
+
+In the latter case, one would also need a model $\tilde R$ of rewards,
+and then we could sample from $\transp{\tilde \Delta}\rho R$ and from
+$\transp{\Delta}\rho\tilde R$.}
+
+Let us consider the norm
+\begin{equation}
+\norm{\Delta V-R}^2_{P,\tilde P}\deq \transp{(\tfrac12 (\Delta+\tilde
+\Delta)
+V-R)}\rho(\tfrac12 (\Delta+\tilde \Delta)V-R)
+\end{equation}
+which is the $L^2(\rho)$-norm of the Bellman gap of an environment
+$\frac12 (P+\tilde P)$ in which half of the transitions are real or
+simulated.
+
+The gradient of this norm is
+\begin{equation}
+\tfrac12 \transp{\Phi}\transp{(\Delta+\tilde
+\Delta)
+}\rho((\Delta+\tilde \Delta)V-R)
+\end{equation}
+
+We can easily compute
+\begin{equation}
+\tfrac12 \transp{\Phi}\transp{\Delta}
+\rho(\tilde \Delta V-R)
++
+\tfrac12 \transp{\Phi}\transp{\tilde \Delta}
+\rho(\Delta V-R)
+\end{equation}
+
 \section{A function that almost solves the Bellman equation can be very

### 2018-12-12 paper/tangotd.tex

@@
 s'$ from the same $s$). We can also sample from $\transp{\tilde
-\Delta}\rho R$. Moreover, \emph{if rewards during $s\to s'$ are
-independent from $s'$}, we can sample from $\transp{\Delta}\rho R$. From
-now on, I will assume that the reward collected during a transition $s\to
-s'$ depends only on $s$. \footnote{This means the state $s$ is a
-sufficient statistic. If the reward depends only on $s'$ instead, we can
-just
-shift the rewards by one step and pretend they are collected at the next
-step; therefore we are back to a reward that depends only on $s$. The
-only case not covered is if the reward really depends on the pair
-$(s,s')$; in that case, the state $s'$ is not quite a sufficient
-statistic of the process.
+\Delta}\rho R$, by sampling a reward during $s\to s'$ from the true
+model, and a simulated second
+transition from $s$. Note that this does not use a model of rewards, only of
+transitions.
+\footnote{
+Moreover, \emph{if rewards during $s\to s'$ are
+independent from $s'$}, we can sample from $\transp{\Delta}\rho R$. 
+(This also covers the case where rewards depend only on $s'$, by shifting
+rewards by one step.) Not covered is the case where the rewards really
+depends on the pair
+$(s,s')$ (in that case, the state $s'$ is not quite a sufficient
+statistic of the process); then we could have a model $\tilde R$ of rewards and
+sample from $\transp{\Delta}\rho \tilde{R}$.
 
-In the latter case, one would also need a model $\tilde R$ of rewards,
-and then we could sample from $\transp{\tilde \Delta}\rho R$ and from
-$\transp{\Delta}\rho\tilde R$.}
+Anyway we will not assume any of this: just using
+$\transp{\tilde
+\Delta}\rho R$ serves our purpose.}
 

### 2018-12-12 paper/tangotd.tex

@@
 
-Let us consider the norm
+Therefore, by various combinations of actual and simulated transitions,
+we can easily compute samples of
 \begin{equation}
-\norm{\Delta V-R}^2_{P,\tilde P}\deq \transp{(\tfrac12 (\Delta+\tilde
-\Delta)
-V-R)}\rho(\tfrac12 (\Delta+\tilde \Delta)V-R)
+\tfrac12 \transp{\Phi}(\transp{\Delta}\rho \tilde \Delta+\transp{\tilde
+\Delta}\rho\Delta)V
+-
+\transp{\Phi}\transp{\tilde \Delta}
+\rho R
 \end{equation}
-which is the $L^2(\rho)$-norm of the Bellman gap of an environment
-$\frac12 (P+\tilde P)$ in which half of the transitions are real or
-simulated.
-
-The gradient of this norm is
-\begin{equation}
-\tfrac12 \transp{\Phi}\transp{(\Delta+\tilde
-\Delta)
-}\rho((\Delta+\tilde \Delta)V-R)
-\end{equation}
-
-We can easily compute
+(variant 1) and
 \begin{equation}
@@
 \end{equation}
+(variant 2, under the conditions of the footnote).
+Setting
+\begin{equation}
+C\deq \tfrac12 (\transp{\Delta}\rho \tilde \Delta+\transp{\tilde
+\Delta}\rho\Delta)
+\end{equation}
+variant 1 is the gradient of
+\begin{equation}
+\norm{V-C^{-1}\transp{\tilde\Delta}\rho R}^2_C
+\end{equation}
+while variant 2 is the gradient of
+\begin{equation}
+\norm{V-\frac12 C^{-1}(\transp{\tilde\Delta}+\transp{\Delta})\rho R}^2_C
+\end{equation}
+
+Therefore, if $C$ is definite positive, then variant 1 and variant 2 will converge to
+the best approximation (in the parametric family $V$) of
+\begin{equation}
+\hat V_1\deq C^{-1}\transp{\tilde\Delta}\rho R \qquad\text{and}\qquad \hat
+V_2\deq \frac12
+C^{-1}(\transp{\tilde\Delta}+\transp{\Delta})\rho R
+\end{equation}
+respectively.
+If $\tilde \Delta=\Delta$, then both $\hat V_1$ and $\hat V_2$
+are equal to the desired target
+$\Delta^{-1}R$.
+
+Let us study how these fixed points approximate the true target
+$\Delta^{-1}R$ when $\tilde \Delta \approx \Delta$. First,
+\begin{equation}
+C=\frac14 \transp{(\tilde \Delta+\Delta)}\rho(\tilde \Delta+\Delta)-\frac14
+\transp{(\tilde
+\Delta-\Delta)}\rho(\tilde
+\Delta-\Delta)
+\end{equation}
+Therefore, $C$ is definite positive provided
+\begin{equation}
+\transp{(\tilde
+\Delta-\Delta)}\rho(\tilde
+\Delta-\Delta)\prec \transp{(\tilde \Delta+\Delta)}\rho(\tilde \Delta+\Delta)
+\end{equation}
+which is the case when the model $\tilde \Delta$ is close enough
+to the true environment $\Delta$ (moreover the left-hand-side is at
+second order in the error). Therefore, convergence is guaranteed in a relatively large
+zone around $\Delta$. But the method will diverge if $\tilde \Delta$ is
+too far from $\Delta$—contrary to double sampling using the model twice,
+which always converges to the value function of the model.
+
+Next, since $C=\frac14 \transp{(\tilde \Delta+\Delta)}\rho(\tilde
+\Delta+\Delta)+O(\tilde\Delta-\Delta)^2$, the fixed point $\hat V_2$ for variant 2
+is
+\begin{equation}
+\hat V_2=\left(\frac{\tilde
+\Delta+\Delta}{2}\right)^{-1}R+O(\tilde\Delta-\Delta)^2
+\end{equation}
+namely, at first order, the fixed point of variant 2 is the value
+function of a Markov chain with transitions $\tfrac12(\tilde P+P)$,
+namely, each transition is real or simulated with probability $1/2$.
+Intuitively this should keep reasonable properties of the true value
+function: even if the model $\tilde P$ diverges when iterated, the Markov chain $\tfrac12(\tilde
+P+P)$ will have a dynamics closer to the true environment.
 

### 2018-12-12 paper/tangotd.tex

@@
 
-This contains: Section 1: remarks on recovering the known convergence results
+This contains: Section~\ref{sec:tdconv}: remarks on recovering the known convergence results
 for TD in a single framework, plus ideas on systematic variants around
@@
 
-Section 2: the problem of the existence of near-solutions to the Bellman
+Section~\ref{sec:norms}: algorithms for convergence of the Bellman error
+in various norms. In particular, solving the double sampling problem by
+using a model for one of the samples, and a real transition for the other
+sample, has nice mathematical properties.
+
+Section~\ref{sec:quasibellman}: the problem of the existence of near-solutions to the Bellman
 equation that are very far from the true value function (I think this is
@@
 
-Section 3: Ideas on how, even if there's no reward signal, we should
+Section~\ref{sec:green}: Ideas on how, even if there's no reward signal, we should
 still learn how to navigate in the Markov chain and learn which
@@
 \section{Recovering known convergence results for TD, and a few ideas}
+\label{sec:tdconv}
 
@@
 
-\paragraph{Minimizers of errors.} Rather than the fixed point of TD, one
+\section{Minimizing the error in some norm via TD}
+\label{sec:norms}
+
+Rather than the fixed point of TD, one
 can look for minimizers of the error in various norms.
@@
 
-\paragraph{Double sampling problem: using a model for one of the
+\paragraph{Double sampling: using a model for one of the
 samples.} Double sampling allows us to minimize the $L^2$ norm of the
@@
 far from the true value function}
+\label{sec:quasibellman}
 
@@
 \section{Storing a summary of possible transitions?}
+\label{sec:green}
 

### 2018-12-12 paper/tangotd.tex

@@
 
-This contains: Section~\ref{sec:tdconv}: remarks on recovering the known convergence results
+These notes contain: Section~\ref{sec:tdconv}: remarks on recovering the known convergence results
 for TD in a single framework, plus ideas on systematic variants around
@@
 
+The fixed point $\hat V_1$ for variant 1 has less clear an
+interpretation. Defining the matrix $\eps$ via $\Delta \eqd(\Id+\eps\rho)\tilde \Delta$ where
+$\eps$ is some hopefully small model error, we find
+\begin{align}
+\hat V_2&=C^{-1}\transp{(\rho\tilde \Delta)}R
+\\&=\left(
+\Delta+\frac{(\transp{(\rho\tilde \Delta)})^{-1}\transp{\Delta} \rho \tilde
+\Delta-\Delta}{2}
+\right)^{-1}R
+\\&=\left(
+\Delta+\frac12(\transp{\eps}-\eps)\rho\tilde \Delta
+\right)^{-1}R
+\end{align}
+In particular, if $\eps$ is symmetric the error vanishes, but if $\eps$
+is antisymmetric we end up computing $\tilde \Delta^{-1}R$, namely the
+Bellman function of the model. Usually $\eps$ would be somewhere in
+between; I do not know if it is possible to force the model to favor
+symmetric values of $\eps$.
+
+In view of this, variant 2 seems more promising, although it relies on
+rewards depending only on $s$ or only on $s'$, not on the pair $(s,s')$.
+
 \section{A function that almost solves the Bellman equation can be very

### 2018-12-12 paper/tangotd.tex

@@
 \end{equation}
-(variant 2, under the conditions of the footnote).
+(variant 2, under the conditions of the footnote). Variant 2 corresponds
+to fully symmetrizing the roles of the synthetic and real samples in
+two-sampling.
+
 Setting
@@
 The fixed point $\hat V_1$ for variant 1 has less clear an
-interpretation. Defining the matrix $\eps$ via $\Delta \eqd(\Id+\eps\rho)\tilde \Delta$ where
-$\eps$ is some hopefully small model error, we find
+interpretation. Define the matrix $\eps$ via $\Delta \eqd(\Id+\eps\rho)\tilde \Delta$ where
+$\eps$ is some hopefully small model error. Then
 \begin{align}

### 2018-12-12 paper/tangotd.tex

@@
 
-Therefore, if $C$ is definite positive, then variant 1 and variant 2 will converge to
+Therefore, if $C$ is positive definite, then variant 1 and variant 2 will converge to
 the best approximation (in the parametric family $V$) of
@@
 \end{equation}
-Therefore, $C$ is definite positive provided
+Therefore, $C$ is positive definite provided
 \begin{equation}

### 2018-12-12 paper/tangotd.tex

@@
 which is the case when the model $\tilde \Delta$ is close enough
-to the true environment $\Delta$ (moreover the left-hand-side is at
-second order in the error). Therefore, convergence is guaranteed in a relatively large
+to the true environment $\Delta$. (Moreover the left-hand-side is of
+second order in the error, while the right-hand-side is of order $0$,
+therefore this condition should be relatively easy to satisfy.) Therefore, convergence is guaranteed in a relatively large
 zone around $\Delta$. But the method will diverge if $\tilde \Delta$ is

### 2018-12-13 paper/tangotd.tex

@@
 a function of the state, corresponding to a feature in the network.} Then the reverse
-Bellman equation above allows us to learn this model via
-a TD-like algorithm, \emph{on-policy}.
+Bellman equation above should allows us to learn this model via
+a TD-like algorithm, \emph{on-policy}. TODO: HOW? not clear how to
+sample from $\phi P$ if $\phi$ is a function. Would need a model of
+reverse transitions?
 

### 2018-12-13 paper/tangotd.tex

@@
 sample from $\phi P$ if $\phi$ is a function. Would need a model of
-reverse transitions?
+reverse transitions? \footnote{I think one has to learn $G\mu^{-1}$ and
+the equation $G=\Id+GP$ rewrites as $\mu
+(G\mu^{-1})\mu=\mu+\mu(G\mu^{-1})\mu P$ where everything is samplable.
+However I fear an implicit double sampling in there; probably
+featurize/dualize
+by asking that this is satisfied for any functions on the left and right.
+}
 

### 2018-12-13 paper/tangotd.tex

@@
 \newcommand{\Dir}{\mathcal{D}}
+\newcommand{\green}{\mathcal{G}}
 
@@
 
+\paragraph{A learned representation for $\Delta^{-1}$.} $\Delta^{-1}$
+contains all the necessary to do policy gradient or actor-critic with
+arbitrary reward functions, as seen above. It is feasible to store a
+compact and useful representation of $\Delta^{-1}$? Ideally, such a
+representation should be learned from the \emph{two-sided} Bellman
+equations
+\begin{equation}
+\Delta^{-1}=\Id + P \Delta^{-1},\qquad \Delta^{-1}=\Id + \Delta^{-1}P
+\end{equation}
+which represent that, when observing a transition $s\to s'$, we update
+both the ways to arrive at $s'$ and the ways to start at $s$.
+
+``Universal value function approximators'' do a bit of that, but only
+with the left (ordinary) Bellman equation.
+
+We propose to learn two functions $\green(s)$ and $f(s)$ from states into
+a state representation $E$, such that for any test function $\psi\from
+S\to \R$, for any $s_1\in S$,
+\begin{equation}
+\E_{s_2\sim \rho}\,\green(s_1)\cdot
+f(s_2)\psi(s_2)\approx \transp{\1_{s_1}}\,\Delta^{-1} \psi
+\end{equation}
+
+If the state space is finite and if we use a one-hot representation, a
+solution is indeed $E=\R^{\#S}$ with
+\begin{equation}
+f(s_2)=\1_{s_2},\qquad
+(\green(s_1))_{s_2}=(\Delta^{-1})_{s_1s_2}/\rho(s_2)
+\end{equation}
+
+The reason for this form is the following: $E$ is supposed to represent
+distributions/functions over states. $f(s_2)$ is supposed to be
+mapped to $\1_{s_2}$. $\green(s_1)$ is supposed to be mapped to the
+density wrt $\rho$ of the fundamental matrix at $s_1$ (the occupation
+distribution from $s_1$: $\transp{\1_{s_1}}\sum_t P^t$, where as usual I
+omit constants\footnote{To deal with constants, we probably have to be
+able to predict \emph{differences} $\psi(s_2)-\psi(s_2')$.}). In
+particular, $\E_{s_2\sim \rho}f(s_2)\psi(s_2)$ is supposed to represent
+the vector with components $\rho \psi(s_2)$. Then the scalar product in
+$E$ represents integration/summation with respect to $\rho$.
+
 \paragraph{Learning to navigate.} Define a function $V(s,s')$ that

### 2018-12-13 paper/tangotd.tex

@@
 \begin{equation}
+\label{eq:greenbell}
 \Delta^{-1}=\Id + P \Delta^{-1},\qquad \Delta^{-1}=\Id + \Delta^{-1}P
@@
 \begin{equation}
+\label{eq:greenrep}
 \E_{s_2\sim \rho}\,\green(s_1)\cdot
@@
 $E$ represents integration/summation with respect to $\rho$.
+
+Of course we cannot directly learn $\green$ from \eqref{eq:greenrep},
+since $\Delta^{-1}$ in the right-hand-side is exactly what we want to
+learn. But we can use the Bellman equations \eqref{eq:greenbell}; after
+applying \eqref{eq:greenbell} to an arbitrary state $s_1$ and test
+function $\psi$, and
+substituting the model $\green$ for $\Delta^{-1}$ via
+$\eqref{eq:greenrep}$, the left Bellman equation becomes
+\begin{gather}
+\transp{\1_{s_1}}\Delta^{-1}\psi=\transp{\1_{s_1}}\psi+\transp{\1_{s_1}}P\Delta^{-1}\psi
+\qquad \Longrightarrow
+\\
+\E_{s_2\sim \rho}\,\green(s_1)\cdot
+f(s_2)\psi(s_2)
+\approx \psi(s_1)+ \E_{s'\sim P(s'|s_1)}\E_{s_2\sim
+\rho}\,\green(s')\cdot
+f(s_2)\psi(s_2)
+\end{gather}
+and the right Bellman equation becomes
+\begin{gather}
+\transp{\1_{s_1}}\Delta^{-1}\psi=\transp{\1_{s_1}}\psi+\transp{\1_{s_1}}\Delta^{-1}P\psi
+\qquad \Longrightarrow
+\\
+\E_{s_2\sim \rho}\,\green(s_1)\cdot
+f(s_2)\psi(s_2)
+\approx \psi(s_1)+ \E_{s_2\sim \rho}\E_{s'_2\sim P(s'_2|s_2)}
+\,\green(s_1)\cdot
+f(s_2)\psi(s'_2)
+\end{gather}
+(indeed, 
+$P\psi$ is
+represented by $\E_{s_2\sim \rho} f(s_2)(P\psi)(s_2)$, and
+$(P\psi)(s_2)=\E_{s'_2\sim P(s'_2|s_2)}\psi(s_2)$).
+
+\emph{Both} these equations are samplable. Therefore, one may
+parameterize $f$ and $\green$ by neural networks and train the
+parameters. This works off-policy: we have not used that $\rho$ is the
+invariant distribution, just some distribution on states we can sample
+from.
+
+The value function is obtained with $\psi=R$, namely,
+\begin{equation}
+V(s_1)=\transp{\1_{s_1}}\Delta^{-1}R\approx
+\E_{s_2\sim \rho}\,\green(s_1)\cdot
+f(s_2)R(s_2)
+\end{equation}
+Of course it is not convenient to have to sample from $s_2$. Instead one
+might want to just learn the representation $f(R)$, defined by abuse of
+notation as
+\begin{equation}
+f(R)\deq \E_{s_2\sim \rho} \,f(s_2)R(s_2)
+\end{equation}
+(which can be computed online)
+and then
+\begin{equation}
+V(s_1)\approx \green(s_1)\cdot f(R)
+\end{equation}
+
+More generally, we can learn the representation of any function $\psi$.
+This could be useful if we want to train a few policy parameters to lead in
+certain directions/lead to states with certain features.
 

### 2018-12-13 paper/tangotd.tex

@@
 
-``Universal value function approximators'' do a bit of that, but only
-with the left (ordinary) Bellman equation.
+``Universal value function approximators'' (Schaul--Horgan--Gregor--Silver ICML 2015) do a bit of that, but only
+with the left (ordinary) Bellman equation. It amounts to solving $\#S$
+independent Bellman equations, and relying on the generalization
+abilities of the network, but the right equation is not exploited at all.
 
@@
 
+Another equation that might be enforced is commutation between
+$P$ and $\Delta^{-1}$; it is expressed similarly to the above, but I do
+not know if there's a gain. It might transfer left to right Bellman
+equations.
+
+$\green$ is basically a model of the world (summed over time). The
+difference with ordinary model-based is: $\green$ is trained as a fixed
+point, considering all timescales at once. In ordinary model-based, the
+model $\tilde P$ obviously commutes with its iterates ($\tilde P^t\tilde
+P=\tilde P\tilde P^t$), but does not provide an easy and accurate
+representation of the occupation measure $\tilde \Delta^{-1}=\sum_t \tilde P^t$. So even in
+the model-based case, the above might be useful.
+
 \paragraph{Learning to navigate.} Define a function $V(s,s')$ that

### 2018-12-13 paper/tangotd.tex

@@
 ``Universal value function approximators'' (Schaul--Horgan--Gregor--Silver ICML 2015) do a bit of that, but only
-with the left (ordinary) Bellman equation. It amounts to solving $\#S$
-independent Bellman equations, and relying on the generalization
+with the left (ordinary) Bellman equation. It amounts to solving a number
+of
+ordinary Bellman equations with different reward functions, and relying on the generalization
 abilities of the network, but the right equation is not exploited at all.
@@
 \end{equation}
+but with infinite or continuous state spaces, we have to find a more
+compact representation.
 
@@
 omit constants\footnote{To deal with constants, we probably have to be
-able to predict \emph{differences} $\psi(s_2)-\psi(s_2')$.}). In
+able to predict \emph{differences} $\psi(s_2)-\psi(s_2')$, or just use
+relative value functions and remove $\psi(s_0)$ for some $s_0$.}). In
 particular, $\E_{s_2\sim \rho}f(s_2)\psi(s_2)$ is supposed to represent
@@
 \end{gather}
-and the right Bellman equation becomes
+because $\transp{\1_{s_1}}P=\E_{s'\sim P(s'|s_1)}\transp{\1_{s'}}\,$.
+Meanwhile,
+the right Bellman equation becomes
 \begin{gather}
@@
 invariant distribution, just some distribution on states we can sample
-from.
+from. Of course, we should apply this to relevant test functions $\psi$,
+such as the reward function or any useful (predictive) features, namely,
+any function we're interested in predicting the future values of. 
+If just using a
+random function $\psi$, there is very little signal to train
+from. \footnote{An extreme case is to take a purely random $\psi$ at each
+training step; then $\psi(s_1)$ and $\psi(s_2)$ are independent unless $s_1=s_2$,
+which is a very unlikely event. This amounts to training with white noise
+features. Formally it works thanks to the rare case $s_1=s_2$, but
+variance is just terrible.}
 

### 2018-12-13 paper/tangotd.tex

@@
 \end{equation}
-which represent that, when observing a transition $s\to s'$, we update
-both the ways to arrive at $s'$ and the ways to start at $s$.
+With these two equations, when observing a transition $s\to s'$, we update
+both the model of what happens when we start at $s$ (left equation,
+standard Bellman), and the model of how to arrive at $s'$ (right
+equation, rarely used as far as I know).
 

### 2018-12-13 paper/tangotd.tex

@@
 contains all the necessary to do policy gradient or actor-critic with
-arbitrary reward functions, as seen above. It is feasible to store a
+arbitrary reward functions, as seen above. Is it feasible to store a
 compact and useful representation of $\Delta^{-1}$? Ideally, such a
@@
 equation, rarely used as far as I know).
+
+The left Bellman equation on $\Delta^{-1}$ is equivalent to
+simultaneously solving the Bellman equation with all possible reward
+functions, generated by $\1_{s'}$ for all possible target states $s'$.
 
@@
 invariant distribution, just some distribution on states we can sample
-from. Of course, we should apply this to relevant test functions $\psi$,
+from.
+
+Of course, we should apply this to relevant test functions $\psi$,
 such as the reward function or any useful (predictive) features, namely,

### 2018-12-13 paper/tangotd.tex

@@
 \label{eq:greenrep}
-\E_{s_2\sim \rho}\,\green(s_1)\cdot
+\green(s_1)\cdot \E_{s_2\sim \rho}
 f(s_2)\psi(s_2)\approx \transp{\1_{s_1}}\,\Delta^{-1} \psi
@@
 
-If the state space is finite and if we use a one-hot representation, a
-solution is indeed $E=\R^{\#S}$ with
-\begin{equation}
-f(s_2)=\1_{s_2},\qquad
-(\green(s_1))_{s_2}=(\Delta^{-1})_{s_1s_2}/\rho(s_2)
-\end{equation}
-but with infinite or continuous state spaces, we have to find a more
+If the state space is finite, a 
+solution is indeed $E=\R^{\#S}$ with a one-hot representation for $f$,
+and the true matrix $\Delta^{-1}$ for $\green$, namely,
+\begin{equation}
+f(s_2)=\1_{s_2}/\rho(s_2),\qquad
+(\green(s_1))_{s_2}=(\Delta^{-1})_{s_1s_2}
+\end{equation}
+or equivalently, $f(s_2)=\1_{s_2}$ and
+$(\green(s_1))_{s_2}=(\Delta^{-1})_{s_1s_2}/\rho(s_2)$. \footnote{The solution to
+such representation problems is never unique because $E$ is an arbitrary
+space and one may always make a change of basis in $E$, e.g., dividing
+the $s_2$-coordinate by $\rho(s_2)$.}
+But with infinite or continuous state spaces, we have to find a more
 compact representation.
@@
 The reason for this form is the following: $E$ is supposed to represent
-distributions/functions over states. $f(s_2)$ is supposed to be
-mapped to $\1_{s_2}$. $\green(s_1)$ is supposed to be mapped to the
-density wrt $\rho$ of the fundamental matrix at $s_1$ (the occupation
-distribution from $s_1$: $\transp{\1_{s_1}}\sum_t P^t$, where as usual I
+distributions or functions over states. $f$ is supposed to map $s_2$ to
+something like
+$\1_{s_2}$ (up to a factor $1/\rho$ to compensate for sampling
+probabilities). 
+In
+particular, $\E_{s_2\sim \rho}f(s_2)\psi(s_2)$ is supposed to represent
+the vector with components $\psi(s_2)$, namely, the function $\psi$. $\green$ is supposed to map each
+$s_1$ to the
+the occupation
+distribution of the random walk starting from $s_1$, namely,
+$\transp{\1_{s_1}}\sum_t P^t$, which is $\transp{\1_{s_1}}\,\Delta^{-1}$
+(as usual I
 omit constants\footnote{To deal with constants, we probably have to be
 able to predict \emph{differences} $\psi(s_2)-\psi(s_2')$, or just use
-relative value functions and remove $\psi(s_0)$ for some $s_0$.}). In
-particular, $\E_{s_2\sim \rho}f(s_2)\psi(s_2)$ is supposed to represent
-the vector with components $\rho \psi(s_2)$. Then the scalar product in
-$E$ represents integration/summation with respect to $\rho$.
+relative value functions and remove $\psi(s_0)$ for some $s_0$.}). Then the scalar product in
+$E$ represents integration/summation over the state space.
 
@@
 \begin{gather}
-\transp{\1_{s_1}}\Delta^{-1}\psi=\transp{\1_{s_1}}\psi+\transp{\1_{s_1}}P\Delta^{-1}\psi
+\transp{\1_{s_1}}\,\Delta^{-1}\psi=\transp{\1_{s_1}}\,\psi+\transp{\1_{s_1}}\,P\Delta^{-1}\psi
 \qquad \Longrightarrow
 \\
-\E_{s_2\sim \rho}\,\green(s_1)\cdot
+\green(s_1)\cdot\E_{s_2\sim \rho}
 f(s_2)\psi(s_2)
-\approx \psi(s_1)+ \E_{s'\sim P(s'|s_1)}\E_{s_2\sim
-\rho}\,\green(s')\cdot
+\approx \psi(s_1)+ \E_{s'\sim P(s'|s_1)}\green(s')\cdot
+\E_{s_2\sim \rho}\,
 f(s_2)\psi(s_2)
 \end{gather}
-because $\transp{\1_{s_1}}P=\E_{s'\sim P(s'|s_1)}\transp{\1_{s'}}\,$.
+because $\transp{\1_{s_1}}\,P=\E_{s'\sim P(s'|s_1)}\transp{\1_{s'}}\,$.
 Meanwhile,
@@
 \begin{gather}
-\transp{\1_{s_1}}\Delta^{-1}\psi=\transp{\1_{s_1}}\psi+\transp{\1_{s_1}}\Delta^{-1}P\psi
+\transp{\1_{s_1}}\,\Delta^{-1}\psi=\transp{\1_{s_1}}\,\psi+\transp{\1_{s_1}}\,\Delta^{-1}P\psi
 \qquad \Longrightarrow
 \\
-\E_{s_2\sim \rho}\,\green(s_1)\cdot
+\green(s_1)\cdot\E_{s_2\sim \rho}
 f(s_2)\psi(s_2)
-\approx \psi(s_1)+ \E_{s_2\sim \rho}\E_{s'_2\sim P(s'_2|s_2)}
-\,\green(s_1)\cdot
+\approx \psi(s_1)+ \green(s_1)\cdot\E_{s_2\sim \rho}\E_{s'_2\sim P(s'_2|s_2)}
 f(s_2)\psi(s'_2)
@@
 represented by $\E_{s_2\sim \rho} f(s_2)(P\psi)(s_2)$, and
-$(P\psi)(s_2)=\E_{s'_2\sim P(s'_2|s_2)}\psi(s_2)$).
+$(P\psi)(s_2)=\E_{s'_2\sim P(s'_2|s_2)}\psi(s'_2)$).
 
@@
 
+Thus we may learn $\green$ via TD-like algorithms. We may even have
+double sampling that works, thanks to featurization (using test functions
+$\psi$), at least for the right equation. TODO should we featurize on the
+left as well?
+
 The value function is obtained with $\psi=R$, namely,
@@
 V(s_1)=\transp{\1_{s_1}}\Delta^{-1}R\approx
-\E_{s_2\sim \rho}\,\green(s_1)\cdot
+\green(s_1)\cdot\E_{s_2\sim \rho}
 f(s_2)R(s_2)

### 2018-12-13 paper/tangotd.tex

@@
 
-If the state space is finite, a 
+If the state space is finite, an exact
 solution is indeed $E=\R^{\#S}$ with a one-hot representation for $f$,
@@
 since $\Delta^{-1}$ in the right-hand-side is exactly what we want to
-learn. But we can use the Bellman equations \eqref{eq:greenbell}; after
+learn. But $\Delta^{-1}$ is the unique solution to the Bellman equations \eqref{eq:greenbell}; after
 applying \eqref{eq:greenbell} to an arbitrary state $s_1$ and test

### 2018-12-13 paper/tangotd.tex

@@
 \paragraph{A learned representation for $\Delta^{-1}$.} $\Delta^{-1}$
-contains all the necessary to do policy gradient or actor-critic with
+contains all the necessary information to do policy gradient or actor-critic with
 arbitrary reward functions, as seen above. Is it feasible to store a
@@
 $\psi$), at least for the right equation. TODO should we featurize on the
-left as well?
+left as well? Or train only with the right equation?
 

### 2018-12-13 paper/tangotd.tex

@@
 standard Bellman), and the model of how to arrive at $s'$ (right
-equation, rarely used as far as I know).
+equation, rarely used as far as I know; somehow eligibility traces
+implement some stochastic form of it).
 

### 2018-12-13 paper/tangotd.tex

--- 2018-12-13 paper/tangotd.tex
@@
 \newcommand{\Dir}{\mathcal{D}}
-\newcommand{\green}{\mathcal{G}}
 
@@
 featurize/dualize
-by asking that this is satisfied for any functions on the left and right.
+by asking that this is satisfied for any test functions on the left and right.
+
+Another way to exploit the equation on the right is to impose that $G$
+and $P$ commute (indeed $G=\sum P^t$). This commutation equation should
+transfer left-learning to right-learning. It can be written in samplable
+terms
+via $\mu(G\mu^{-1})(\mu P\psi)=(\mu P)(G \mu^{-1})(\mu
+\psi)$ for any feature $\psi$. This apparently avoids double sampling!
 }
@@
 how to arrive at $s'$ and of what happens when starting at $s$.
-
-\paragraph{A learned representation for $\Delta^{-1}$.} $\Delta^{-1}$
-contains all the necessary to do policy gradient or actor-critic with
-arbitrary reward functions, as seen above. Is it feasible to store a
-compact and useful representation of $\Delta^{-1}$? Ideally, such a
-representation should be learned from the \emph{two-sided} Bellman
-equations
-\begin{equation}
-\label{eq:greenbell}
-\Delta^{-1}=\Id + P \Delta^{-1},\qquad \Delta^{-1}=\Id + \Delta^{-1}P
-\end{equation}
-With these two equations, when observing a transition $s\to s'$, we update
-both the model of what happens when we start at $s$ (left equation,
-standard Bellman), and the model of how to arrive at $s'$ (right
-equation, rarely used as far as I know).
-
-The left Bellman equation on $\Delta^{-1}$ is equivalent to
-simultaneously solving the Bellman equation with all possible reward
-functions, generated by $\1_{s'}$ for all possible target states $s'$.
-
-``Universal value function approximators'' (Schaul--Horgan--Gregor--Silver ICML 2015) do a bit of that, but only
-with the left (ordinary) Bellman equation. It amounts to solving a number
-of
-ordinary Bellman equations with different reward functions, and relying on the generalization
-abilities of the network, but the right equation is not exploited at all.
-
-We propose to learn two functions $\green(s)$ and $f(s)$ from states into
-a state representation $E$, such that for any test function $\psi\from
-S\to \R$, for any $s_1\in S$,
-\begin{equation}
-\label{eq:greenrep}
-\green(s_1)\cdot \E_{s_2\sim \rho}
-f(s_2)\psi(s_2)\approx \transp{\1_{s_1}}\,\Delta^{-1} \psi
-\end{equation}
-
-If the state space is finite, an exact
-solution is indeed $E=\R^{\#S}$ with a one-hot representation for $f$,
-and the true matrix $\Delta^{-1}$ for $\green$, namely,
-\begin{equation}
-f(s_2)=\1_{s_2}/\rho(s_2),\qquad
-(\green(s_1))_{s_2}=(\Delta^{-1})_{s_1s_2}
-\end{equation}
-or equivalently, $f(s_2)=\1_{s_2}$ and
-$(\green(s_1))_{s_2}=(\Delta^{-1})_{s_1s_2}/\rho(s_2)$. \footnote{The solution to
-such representation problems is never unique because $E$ is an arbitrary
-space and one may always make a change of basis in $E$, e.g., dividing
-the $s_2$-coordinate by $\rho(s_2)$.}
-But with infinite or continuous state spaces, we have to find a more
-compact representation.
-
-The reason for this form is the following: $E$ is supposed to represent
-distributions or functions over states. $f$ is supposed to map $s_2$ to
-something like
-$\1_{s_2}$ (up to a factor $1/\rho$ to compensate for sampling
-probabilities). 
-In
-particular, $\E_{s_2\sim \rho}f(s_2)\psi(s_2)$ is supposed to represent
-the vector with components $\psi(s_2)$, namely, the function $\psi$. $\green$ is supposed to map each
-$s_1$ to the
-the occupation
-distribution of the random walk starting from $s_1$, namely,
-$\transp{\1_{s_1}}\sum_t P^t$, which is $\transp{\1_{s_1}}\,\Delta^{-1}$
-(as usual I
-omit constants\footnote{To deal with constants, we probably have to be
-able to predict \emph{differences} $\psi(s_2)-\psi(s_2')$, or just use
-relative value functions and remove $\psi(s_0)$ for some $s_0$.}). Then the scalar product in
-$E$ represents integration/summation over the state space.
-
-Of course we cannot directly learn $\green$ from \eqref{eq:greenrep},
-since $\Delta^{-1}$ in the right-hand-side is exactly what we want to
-learn. But $\Delta^{-1}$ is the unique solution to the Bellman equations \eqref{eq:greenbell}; after
-applying \eqref{eq:greenbell} to an arbitrary state $s_1$ and test
-function $\psi$, and
-substituting the model $\green$ for $\Delta^{-1}$ via
-$\eqref{eq:greenrep}$, the left Bellman equation becomes
-\begin{gather}
-\transp{\1_{s_1}}\,\Delta^{-1}\psi=\transp{\1_{s_1}}\,\psi+\transp{\1_{s_1}}\,P\Delta^{-1}\psi
-\qquad \Longrightarrow
-\\
-\green(s_1)\cdot\E_{s_2\sim \rho}
-f(s_2)\psi(s_2)
-\approx \psi(s_1)+ \E_{s'\sim P(s'|s_1)}\green(s')\cdot
-\E_{s_2\sim \rho}\,
-f(s_2)\psi(s_2)
-\end{gather}
-because $\transp{\1_{s_1}}\,P=\E_{s'\sim P(s'|s_1)}\transp{\1_{s'}}\,$.
-Meanwhile,
-the right Bellman equation becomes
-\begin{gather}
-\transp{\1_{s_1}}\,\Delta^{-1}\psi=\transp{\1_{s_1}}\,\psi+\transp{\1_{s_1}}\,\Delta^{-1}P\psi
-\qquad \Longrightarrow
-\\
-\green(s_1)\cdot\E_{s_2\sim \rho}
-f(s_2)\psi(s_2)
-\approx \psi(s_1)+ \green(s_1)\cdot\E_{s_2\sim \rho}\E_{s'_2\sim P(s'_2|s_2)}
-f(s_2)\psi(s'_2)
-\end{gather}
-(indeed, 
-$P\psi$ is
-represented by $\E_{s_2\sim \rho} f(s_2)(P\psi)(s_2)$, and
-$(P\psi)(s_2)=\E_{s'_2\sim P(s'_2|s_2)}\psi(s'_2)$).
-
-\emph{Both} these equations are samplable. Therefore, one may
-parameterize $f$ and $\green$ by neural networks and train the
-parameters. This works off-policy: we have not used that $\rho$ is the
-invariant distribution, just some distribution on states we can sample
-from.
-
-Of course, we should apply this to relevant test functions $\psi$,
-such as the reward function or any useful (predictive) features, namely,
-any function we're interested in predicting the future values of. 
-If just using a
-random function $\psi$, there is very little signal to train
-from. \footnote{An extreme case is to take a purely random $\psi$ at each
-training step; then $\psi(s_1)$ and $\psi(s_2)$ are independent unless $s_1=s_2$,
-which is a very unlikely event. This amounts to training with white noise
-features. Formally it works thanks to the rare case $s_1=s_2$, but
-variance is just terrible.}
-
-Thus we may learn $\green$ via TD-like algorithms. We may even have
-double sampling that works, thanks to featurization (using test functions
-$\psi$), at least for the right equation. TODO should we featurize on the
-left as well?
-
-The value function is obtained with $\psi=R$, namely,
-\begin{equation}
-V(s_1)=\transp{\1_{s_1}}\Delta^{-1}R\approx
-\green(s_1)\cdot\E_{s_2\sim \rho}
-f(s_2)R(s_2)
-\end{equation}
-Of course it is not convenient to have to sample from $s_2$. Instead one
-might want to just learn the representation $f(R)$, defined by abuse of
-notation as
-\begin{equation}
-f(R)\deq \E_{s_2\sim \rho} \,f(s_2)R(s_2)
-\end{equation}
-(which can be computed online)
-and then
-\begin{equation}
-V(s_1)\approx \green(s_1)\cdot f(R)
-\end{equation}
-
-More generally, we can learn the representation of any function $\psi$.
-This could be useful if we want to train a few policy parameters to lead in
-certain directions/lead to states with certain features.
-
-Another equation that might be enforced is commutation between
-$P$ and $\Delta^{-1}$; it is expressed similarly to the above, but I do
-not know if there's a gain. It might transfer left to right Bellman
-equations.
-
-$\green$ is basically a model of the world (summed over time). The
-difference with ordinary model-based is: $\green$ is trained as a fixed
-point, considering all timescales at once. In ordinary model-based, the
-model $\tilde P$ obviously commutes with its iterates ($\tilde P^t\tilde
-P=\tilde P\tilde P^t$), but does not provide an easy and accurate
-representation of the occupation measure $\tilde \Delta^{-1}=\sum_t \tilde P^t$. So even in
-the model-based case, the above might be useful.
 

### 2018-12-13 paper/tangotd.tex

--- 2018-12-13 paper/tangotd.tex
@@
 
+Nico suggests the references around successor representation, proto-value
+functions, byt Marlos Machado.
+
 \paragraph{Learning to navigate.} Define a function $V(s,s')$ that

### 2018-12-13 paper/tangotd.tex

@@
 
+More precisely, $\Delta^{-1}_{ss'}$ is the number of paths from $s$ to
+$s'$ weighted by their probabilities. When observing a transition $s_1\to
+s_2$, the left equation updates paths from $s_1$; the right equation
+updates paths to $s_2$. I believe both are important. The right equation
+does not exist when working with a single ``target'' reward function $R$.
+
 The left Bellman equation on $\Delta^{-1}$ is equivalent to
@@
 functions, generated by $\1_{s'}$ for all possible target states $s'$.
+Using only the left Bellman equation treats all targets $s'$ as
+completely independent problems for learning.
 
@@
 such representation problems is never unique because $E$ is an arbitrary
-space and one may always make a change of basis in $E$, e.g., dividing
-the $s_2$-coordinate by $\rho(s_2)$.}
+space and one may always make a change of basis in $E$. More precisely,
+if $B$ is any invertible linear map on $E$, then changing $f\gets Bf$,
+$\green\gets (\transp{B})^{-1}\green$ produces a new solution; indeed
+this preserves $\green\cdot f$. This may
+be useful to ``renormalize'' once in a while, eg setting $f$ to have
+unit covariance, which may help with numerical conditioning.} 
 But with infinite or continuous state spaces, we have to find a more

### 2018-12-13 paper/tangotd.tex

@@
 this preserves $\green\cdot f$. This may
-be useful to ``renormalize'' once in a while, eg setting $f$ to have
-unit covariance, which may help with numerical conditioning.} 
+be useful to ``renormalize'' once in a while to help with numerical
+conditioning: eg setting $f$ to have
+unit covariance, or $f$ and $\green$ to have the same covariance.
+
+Actually the most symmetric solution may be something like
+$f(s_2)=\Delta^{-1/2}\1_{s_2}/\rho(s_2)$ and
+$\green(s_1)=\transp{(\Delta^{-1/2})}\1_{s_1}$. This may be the fastest
+to reach, by splitting eigenvalues of $\Delta$ halfway between $f$ and
+$\green$.}
 But with infinite or continuous state spaces, we have to find a more
@@
 
+When sampling for the right Bellman equation, it is wasteful to sample a
+state $s_2$ and apply a test function $\psi$ if $\psi(s_2)$ is very
+small. Instead one may want to sample $s_2$ according to
+$\rho_\psi\deq \rho(s_2)\abs{\psi(s_2)}/\E_{\rho}\abs{\psi}$, in which case
+$\E_{s_2\sim \rho} f(s_2)\psi(s_2)$ becomes $\E_{s_2\sim \rho_\psi}
+f(s_2)(\E_{\rho}\abs{\psi})$. This may significantly reduce variance.
+
 The value function is obtained with $\psi=R$, namely,
@@
 difference with ordinary model-based is: $\green$ is trained as a fixed
-point, considering all timescales at once. In ordinary model-based, the
+point, considering all timescales at once. $\Delta^{-1}$ contains all
+paths of all lengths from $s$ to $s'$. In ordinary model-based, the
 model $\tilde P$ obviously commutes with its iterates ($\tilde P^t\tilde
@@
 the model-based case, the above might be useful.
+
+One may also compute something like $\Delta^{-1}_{s,a,s'}$ (paths that start
+at $s$ with action $a$ and arrive at $s'$). Unclear if more useful.
 

### 2018-12-13 paper/tangotd.tex

@@
 $\rho_\psi\deq \rho(s_2)\abs{\psi(s_2)}/\E_{\rho}\abs{\psi}$, in which case
-$\E_{s_2\sim \rho} f(s_2)\psi(s_2)$ becomes $\E_{s_2\sim \rho_\psi}
-f(s_2)(\E_{\rho}\abs{\psi})$. This may significantly reduce variance.
+$\E_{s_2\sim \rho} f(s_2)\psi(s_2)$ is equal to $\E_{s_2\sim \rho_\psi}
+f(s_2)(\E_{\rho}\abs{\psi})$. This will significantly reduce variance,
+especially for test functions $\psi$ that are concentrated aroudn a few
+points.
 

### 2018-12-14 pendulum_exp/envs/utils.py
adding dart half cheetah

@@ -201,6 +201,9 @@ def make_env(env_id: str, dt: float, time_limit: float): # noqa: C901
         dt = env.dt
         info('The timestep dt is {}'.format(dt))
         # env.model.opt.timestep = dt / env.frame_skip
+    elif env_id == 'dart_cheetah':
+        env = gym.make('DartHalfCheetah-v1').unwrapped
+        env.dart_world.dt = dt / env.frame_skip
     elif env_id == 'pusher':
         env = DiscretePusherEnv()
         env.dt = dt

### 2018-12-14 paper/tangotd.tex

--- 2018-12-13 paper/tangotd.tex
@@
 \newcommand{\green}{\mathcal{G}}
+\DeclareMathOperator{\sgn}{sgn}
 
@@
 $\rho_\psi\deq \rho(s_2)\abs{\psi(s_2)}/\E_{\rho}\abs{\psi}$, in which case
-$\E_{s_2\sim \rho} f(s_2)\psi(s_2)$ becomes $\E_{s_2\sim \rho_\psi}
-f(s_2)(\E_{\rho}\abs{\psi})$. This may significantly reduce variance.
+$\E_{s_2\sim \rho} f(s_2)\psi(s_2)$ is equal to $(\E_{\rho}\abs{\psi})\E_{s_2\sim \rho_\psi}
+f(s_2)\sgn(\psi(s_2))$. This will significantly reduce variance,
+especially for test functions $\psi$ that are concentrated around a few
+points.
 

### 2018-12-17 pendulum_exp/critics/__init__.py
Order advantage

@@ -1,5 +1,7 @@
 from critics.advantage import AdvantageCritic
 from critics.value import ValueCritic
 from critics.mixture_advantage import MixtureAdvantageCritic
+from critics.order_advantage import OrderAdvantageCritic
 
-__all__ = ["AdvantageCritic", "ValueCritic", "MixtureAdvantageCritic"]
+__all__ = ["AdvantageCritic", "ValueCritic", "MixtureAdvantageCritic",
+           "OrderAdvantageCritic"]

### 2018-12-17 pendulum_exp/critics/order_advantage.py

@@ -1,0 +1,128 @@
+from abstract import Critic, Arrayable, ParametricFunction, Tensorable
+import torch
+from torch import Tensor
+import numpy as np
+from convert import arr_to_th, check_array, check_tensor
+from optimizer import setup_optimizer
+from gym import Space
+from gym.spaces import Box, Discrete
+from models import MLP, ContinuousAdvantageMLP, NormalizedMLP
+from stateful import CompoundStateful
+
+class OrderAdvantageCritic(CompoundStateful, Critic):
+    def __init__(self,
+                 dt: float, gamma: float, lr: float, optimizer: str,
+                 val_function: ParametricFunction, adv_function: ParametricFunction) -> None:
+        CompoundStateful.__init__(self)
+        self._reference_obs: Tensor = None
+        self._val_function = val_function
+        self._adv_function = adv_function
+
+        self._adv_optimizer = \
+            setup_optimizer(self._adv_function.parameters(),
+                            opt_name=optimizer, lr=lr, dt=dt,
+                            inverse_gradient_magnitude=1,
+                            weight_decay=0)
+        self._val_optimizer = \
+            setup_optimizer(self._val_function.parameters(),
+                            opt_name=optimizer, lr=lr, dt=dt,
+                            inverse_gradient_magnitude=dt,
+                            weight_decay=0)
+
+        self._dt = dt
+        self._gamma = gamma
+
+        self._device = 'cpu'
+
+    def optimize(self, obs: Arrayable, action: Arrayable, max_action: Tensor,
+                 next_obs: Arrayable, max_next_action: Tensor, reward: Arrayable,
+                 done: Arrayable, time_limit: Arrayable, weights: Arrayable) -> Tensor:
+        if self._reference_obs is None:
+            self._reference_obs = arr_to_th(obs, self._device)
+
+        obs = check_array(obs)
+        next_obs = check_array(next_obs)
+        action = arr_to_th(action, self._device).type_as(max_action)
+        reward = arr_to_th(reward, self._device)
+        done = arr_to_th(check_array(done).astype('float'), self._device)
+
+        reference_v = self._val_function(self._reference_obs).squeeze()
+        mean_v = reference_v.mean()
+
+        v = self._val_function(obs).squeeze()
+        bs = obs.shape[0]
+        advs, sigmas = self.compute_advantages(
[…]

### 2018-12-17 paper/tangotd.tex

@@
 because $\transp{\1_{s_1}}\,P=\E_{s'\sim P(s'|s_1)}\transp{\1_{s'}}\,$.
+For a fixed $f$, 
+this is basically a multiple Bellman equation on $\green$ with multiple
+rewards $\psi$, and features $f\psi$.
+
 Meanwhile,
@@
 features. Formally it works thanks to the rare case $s_1=s_2$, but
-variance is just terrible.}
-
-Thus we may learn $\green$ via TD-like algorithms. We may even have
+variance is just terrible.} Even training only on $\psi=R$ should be
+interesting: the left Bellman equation just becomes a particular
+parameterization of the value function as $V(s_1)=\green(s_1)\cdot
+\E_{s_2} R(s_2)f(s_2)$, so if the left Bellman equation is used, this
+will approximate the true $V$.
+
+Thus we may learn $\green$ and $f$ via TD-like algorithms. (Does it make
+sense to use the left equation for $\green$ and the right equation for
+$f$?) We may even have
 double sampling that works, thanks to featurization (using test functions
@@
 \end{equation}
+and of course it might be helpful to learn the remainder $V-\green\cdot
+f(R)$ via ordinary TD, in case $R$ has no component along the features
+learned.
 

### 2018-12-17 pendulum_exp/nn.py
big fix

@@ -10,7 +10,7 @@ def gmm_loss(batch: Tensor,
              mus: Tensor,
              sigmas: Tensor,
              logpi: Tensor,
-             reduce: bool=True) -> Tensor: # pylint: disable=too-many-arguments
+             reduce: bool = True) -> Tensor: # pylint: disable=too-many-arguments
     """ Computes the gmm loss.
     Compute minus the log probability of batch under the GMM model described
     by mus, sigmas, pi. Precisely, with bs1, bs2, ... the sizes of the batch
@@ -78,7 +78,7 @@ class LogSoftmax(torch.nn.Module, ParametricFunction):
 def soft_update(net: ParametricFunction, target_net: ParametricFunction, tau: float):
     with torch.no_grad():
         for target_param, param in zip(target_net.parameters(), net.parameters()):
-            target_param.add_(tau, param - target_param)
+            target_param.add_(1 - tau, param - target_param)
 
 def hard_update(net: ParametricFunction, target_net: ParametricFunction, tau: float):
     with torch.no_grad():

### 2018-12-17 paper/tangotd.tex

@@
 represented by $\E_{s_2\sim \rho} f(s_2)(P\psi)(s_2)$, and
-$(P\psi)(s_2)=\E_{s'_2\sim P(s'_2|s_2)}\psi(s'_2)$).
+$(P\psi)(s_2)=\E_{s'_2\sim P(s'_2|s_2)}\psi(s'_2)$). If we view this as
+an equation on $f$ for fixed $\green$, this imposes a kind of spatial
+regularity of features $f$.
 
@@
 Nico suggests the references around successor representation, proto-value
-functions, byt Marlos Machado.
+functions, by Marlos Machado.
+
+\paragraph{Enumerative combinatorics, the middle Bellman equation, and
+the Newton method for $\Delta^{-1}$.} As mentioned above, $(\Id-\gamma
+P)^{-1}$ counts the number of paths from state $s_1$ to $s_2$, weighted
+by their probability and $\gamma^\mathrm{length}$. When observing a new
+transition $s\to s'$, the left Bellman equation on $P$ produces new paths
+from $s$, by adding the edge $s\to s'$ to all paths from $s'$. Similarly
+the right Bellman equation produces new paths to $s'$.
+
+The Newton method for matrix inversion is a classical tool to compute
+$\Delta^{-1}$. It consists in iterating
+\begin{equation}
+A\gets 2A-A\Delta A
+\end{equation}
+It converges extremely fast if initialized reasonably close to $A^{-1}$ (namely if the spectral
+radius of $\Id-AX$ is less than $1$), but it can diverge if badly
+initialized. A suitable initialization is $A=\eps\transp{\Delta}$ for
+small enough $\eps$. For stochastic estimates $\tilde \Delta$ of $\Delta$ (as we have
+here through transitions), the stochastic update version with learning
+rate $\eta$ is
+\begin{equation}
+A\gets A+\eta A-\eta A\tilde \Delta A
+\end{equation}
+
+With $\Delta=\Id-\gamma P$, 
+this method corresponds to a ``middle'' Bellman equation
+\begin{equation}
+\Delta^{-1}\gets \Delta^{-1}+\eta\Delta^{-1}-\eta \Delta^{-2}+\eta \gamma \Delta^{-1}P\Delta^{-1}
+\end{equation}
+Importantly, when a new transition $s\to s'$ is observed from $P$, this equation
+produces new paths by connecting all paths that end at $s$ with all paths
+starting at $s'$, thus doubling path length (thanks to the $\Delta^{-1}$
+on the right and left of $P$). This doubling corresponds to the
+quadratic convergence of the Newton method. (However, this method does not
+produce new starting points or endpoints.)
+
+Here we do not have access to $\Delta$, but have access to transitions
+$s\to s'$, which are samples of
+$\rho\Delta$, namely, $\rho\Delta=\E_{s,s'} 1_s\transp{(\1_s-\gamma
+\1_{s'})}$. This corresponds to a stochastic version of the Newton
+method, with learning rate $\eta\ll 1$,
+\begin{equation}
+A\gets A+\eta A-\eta A 1_s\transp{(\1_s-\gamma \1_{s'})} A
+\end{equation}
+Indeed, in expectation over $s\to s'$ this corresponds to $A\gets A+\eta
+A-\eta A \rho\Delta A$, which converges to $A=\Delta^{-1}\rho^{-1}$. This
+also corresponds to the rank-one update formyla for matrix inversion,
+$(D+\eta u\transp{v})^{-1}=D^{-1}-\eta
+D^{-1}u\transp{v}D^{-1}+O(\eta^2)$. Namely,
+if we update our estimate of $\rho\Delta$ by
+$\rho\Delta\gets (1-\eta)\rho\Delta+\eta 1_s\transp{(\1_s-\gamma
+\1_{s'})}$ when a transition $s\to s'$ is observed, then
+$(\rho\Delta)^{-1}$ gets updated as above.
+
+With $A=\green\cdot \E_{\rho} f$ as above, everything is fully samplable.
+It is not clear to me how to mix the left, right and middle Bellman
+equations, but the correspondence with path combinatorics and the Newton
+method suggests this
+might be an important point for the navigation problem.
 

### 2018-12-17 paper/tangotd.tex

@@
 $s\to s'$, which are samples of
-$\rho\Delta$, namely, $\rho\Delta=\E_{s,s'} 1_s\transp{(\1_s-\gamma
+$\Delta$, namely, $\Delta=\sum_s\E_{s'\sim P(s'|s)} \1_s\transp{(\1_s-\gamma
 \1_{s'})}$. This corresponds to a stochastic version of the Newton
@@
 \begin{equation}
-A\gets A+\eta A-\eta A 1_s\transp{(\1_s-\gamma \1_{s'})} A
+A\gets A+\eta A-\eta A \E_{ss'} \frac{1}{\rho(s)}\1_s\transp{(\1_s-\gamma \1_{s'})} A
 \end{equation}
 Indeed, in expectation over $s\to s'$ this corresponds to $A\gets A+\eta
-A-\eta A \rho\Delta A$, which converges to $A=\Delta^{-1}\rho^{-1}$. This
+A-\eta A \Delta A$, which converges to $A=\Delta^{-1}$. This
 also corresponds to the rank-one update formyla for matrix inversion,
@@
 D^{-1}u\transp{v}D^{-1}+O(\eta^2)$. Namely,
-if we update our estimate of $\rho\Delta$ by
-$\rho\Delta\gets (1-\eta)\rho\Delta+\eta 1_s\transp{(\1_s-\gamma
+if we update our estimate of $\Delta$ by
+$\Delta\gets (1-\eta)\Delta+\eta \frac{1}{\rho(s)} \1_s\transp{(\1_s-\gamma
 \1_{s'})}$ when a transition $s\to s'$ is observed, then
-$(\rho\Delta)^{-1}$ gets updated as above.
-
-With $A=\green\cdot \E_{\rho} f$ as above, everything is fully samplable.
+$\Delta^{-1}$ gets updated as above.
+
+With $A=\green\cdot \E_{\rho} f$ as above, everything is fully samplable
+(thanks to the $\E_\rho$ which cancels the $1/\rho$ factors), see below.
 It is not clear to me how to mix the left, right and middle Bellman
@@
 might be an important point for the navigation problem.
+
+Namely, set
+\begin{equation}
+A=\transp{\green}F, \qquad \green=\sum_s \green(s)\1_s,\qquad F=\E_{s\sim
+\rho}
+f(s)\1_s
+\end{equation}
+which is the same representation as above, namely $A=\Delta^{-1}$ is
+equivalent to $\transp{\1_s}\green(s)\cdot \E_{s'\sim
+\rho} \psi(s')f(s')=\transp{\1_s}\Delta^{-1}\psi$ for all $s$ and test
+functions $\psi$.
 

### 2018-12-17 paper/tangotd.tex

@@
 \begin{equation}
-A=\transp{\green}F, \qquad \green=\sum_s \green(s)\1_s,\qquad F=\E_{s\sim
+A=\transp{\green}F, \qquad \green=\sum_s \green(s)\transp{\1_s},\qquad F=\E_{s\sim
 \rho}
-f(s)\1_s
+f(s)\transp{\1_s}
 \end{equation}
 which is the same representation as above, namely $A=\Delta^{-1}$ is
-equivalent to $\transp{\1_s}\green(s)\cdot \E_{s'\sim
+equivalent to $\green(s)\cdot \E_{s'\sim
 \rho} \psi(s')f(s')=\transp{\1_s}\Delta^{-1}\psi$ for all $s$ and test
-functions $\psi$.
+functions $\psi$. Then the Newton equation for $A$ reads
+\begin{equation}
+\transp{\green}F\gets (1+\eta)\transp{\green}F-\eta \transp{\green}F \E_{ss'} \frac{1}{\rho(s)}\1_s\transp{(\1_s-\gamma
+\1_{s'})} \transp{\green}F
+\end{equation}
+and since $F\1_s=\rho(s)f(s)$ and
+$\transp{\1_s}\transp{\green}=\transp{\green(s)}$, this simplifies to
+\begin{equation}
+\transp{\green}F\gets (1+\eta)\transp{\green}F-\eta
+\transp{\green} \E_{ss'} f(s)\transp{(\green(s)-\gamma
+\green(s'))}F
+\end{equation}
+and we propose to update separately for $\green$ given $f$ and $f$ given
+$\green$: the advantage being that for fixed $f$, we can obtain the
+update above by updating $\green$ via
+\begin{equation}
+\transp{\green}\gets (1+\eta)\transp{\green}-\eta
+\transp{\green} \E_{ss'} f(s)\transp{(\green(s)-\gamma
+\green(s'))}
+\end{equation}
+and likewise for $f$ given $\green$,
+\begin{equation}
+F\gets (1+\eta)F-\eta
+\E_{ss'} f(s)\transp{(\green(s)-\gamma
+\green(s'))}F
+\end{equation}
+both of which can be implemented via gradient descent on the parameters
+of $\green$ and $f$. There is no need to sample on test functions.
 

### 2018-12-19 paper/tangotd.tex

@@
 \begin{equation}
-A\gets 2A-A\Delta A
-\end{equation}
-It converges extremely fast if initialized reasonably close to $A^{-1}$ (namely if the spectral
-radius of $\Id-AX$ is less than $1$), but it can diverge if badly
-initialized. A suitable initialization is $A=\eps\transp{\Delta}$ for
+Z\gets 2Z-Z\Delta Z
+\end{equation}
+It converges extremely fast if initialized reasonably close to $Z^{-1}$ (namely if the spectral
+radius of $\Id-ZX$ is less than $1$), but it can diverge if badly
+initialized. A suitable initialization is $Z=\eps\transp{\Delta}$ for
 small enough $\eps$. For stochastic estimates $\tilde \Delta$ of $\Delta$ (as we have
@@
 \begin{equation}
-A\gets A+\eta A-\eta A\tilde \Delta A
+Z\gets Z+\eta Z-\eta Z\tilde \Delta Z
 \end{equation}
@@
 \begin{equation}
-A\gets A+\eta A-\eta A \E_{ss'} \frac{1}{\rho(s)}\1_s\transp{(\1_s-\gamma \1_{s'})} A
-\end{equation}
-Indeed, in expectation over $s\to s'$ this corresponds to $A\gets A+\eta
-A-\eta A \Delta A$, which converges to $A=\Delta^{-1}$. This
+Z\gets Z+\eta Z-\eta Z \E_{ss'} \frac{1}{\rho(s)}\1_s\transp{(\1_s-\gamma \1_{s'})} Z
+\end{equation}
+Indeed, in expectation over $s\to s'$ this corresponds to $Z\gets Z+\eta
+Z-\eta Z \Delta Z$, which converges to $Z=\Delta^{-1}$. This
 also corresponds to the rank-one update formyla for matrix inversion,
@@
 
-With $A=\green\cdot \E_{\rho} f$ as above, everything is fully samplable
+With $Z=\green\cdot \E_{\rho} f$ as above, everything is fully samplable
 (thanks to the $\E_\rho$ which cancels the $1/\rho$ factors), see below.
@@
 \begin{equation}
-A=\transp{\green}F, \qquad \green=\sum_s \green(s)\transp{\1_s},\qquad F=\E_{s\sim
+Z=\transp{\green}F, \qquad \green=\sum_s \green(s)\transp{\1_s},\qquad F=\E_{s\sim
 \rho}
@@
 \end{equation}
-which is the same representation as above, namely $A=\Delta^{-1}$ is
+which is the same representation as above, namely $Z=\Delta^{-1}$ is
 equivalent to $\green(s)\cdot \E_{s'\sim
 \rho} \psi(s')f(s')=\transp{\1_s}\Delta^{-1}\psi$ for all $s$ and test
-functions $\psi$. Then the Newton equation for $A$ reads
+functions $\psi$. Then the Newton equation for $Z$ reads
 \begin{equation}
@@
 both of which can be implemented via gradient descent on the parameters
-of $\green$ and $f$. There is no need to sample on test functions.
+of $\green$ and $f$. (This double the learning rate $\eta$.)
+There is no need to sample on test functions.
+
+(Equivalent to solving for $ \Id=\E_{ss'} f(s)\transp{(\green(s)-\gamma
+\green(s'))}$, namely, $\Id=F\Delta\transp{\green}$ in the representation space $E$, but via a different gradient
+descent.)
+
+The Newton version should not be used alone. Indeed, with a matrix $Z$ of
+rank $r$, it just settles on some $r$-dimensional subspace and inverts 
+$\Delta^{-1}$ on that subspace. The subspace in question does not
+change at all along optimization. (This is because the equations on $F$
+and $\green$ leave $\Ker F$ and $\Img \transp{\green}$ invariant; more
+precisely, $\Delta Z=\Id$ on $\Img \transp{\green}$ and $\Img(\Delta
+Z-\Id)\subset\Ker F$)
+From a purely algebraic point of view,
+all these $r$-dimensional subspaces are equally important: there are as
+many quasi-inverses as such subspaces, and the subspace is fixed at the
+initialization of $F$ and $\green$. (For instance, if $r=1$ one can
+choose any vector for $\transp{\green}$, and choose for $F$ any linear
+form that evaluates to $1$ on $\Delta \transp{\green}$).
+
+On the other hand the left and right Bellman equations prioritize
+inverting $\Delta$ on test functions $\psi$. So this introduces
+(implicitly) a quantitative criterion that breaks symmetry between
+possible quasi-inverses.
+
+Intuitively, the middle Bellman equation does not create new starting
+points or endpoints on paths, only works in the middle. (This corresponds
+to the fixed image and kernel.) Meanwhile the left Bellman equation
+creates starting points and the right Bellman equation creates endpoints
+when observing a new transition $s\to s'$. This introduces spatial
+smoothness over the image and kernel.
+
+I still don't know how to rightly combine these three equations.
 

### 2018-12-19 paper/tangotd.tex

@@
 $E$ represents integration/summation over the state space.
+This representation keeps linearity of the map $R\mapsto V$, while
+allowing for a nonlinear $f$ and $\green$.
 
@@
 points or endpoints on paths, only works in the middle. (This corresponds
-to the fixed image and kernel.) Meanwhile the left Bellman equation
+to the fixed image and kernel.) [TODO: though if I start with $Z=\Id$ (a
+path from every state to itself) and apply Newton then I do create paths,
+so the meaning of this is unclear and initialization might be important.
+It might be important to initialize $F$ and $\green$ so that $Z\approx \Id$.] Meanwhile, the left Bellman equation
 creates starting points and the right Bellman equation creates endpoints
@@
 I still don't know how to rightly combine these three equations.
+
+\paragraph{Variant.}
+I think
+everything works equally well if we just learn a kernel $Z(s_1,s_2)$ such
+that
+\begin{equation}
+\transp{\1_{s_1}}\Delta^{-1}\psi\approx \E_{s_2\sim \rho} Z(s_1,s_2)
+\psi(s_2)
+\end{equation}
+and this is still linear in $\psi$. This represents the inverse Laplace
+operator by
+\begin{equation}
+Z=\sum_{s_1,s_2}\rho(s_2)  Z(s_1,s_2)\1_{s_1}\transp{\1_{s_2}}
+\end{equation}
+
+The left Bellman equation is
+\begin{gather}
+\transp{\1_{s_1}}\,\Delta^{-1}\psi=\transp{\1_{s_1}}\,\psi+\transp{\1_{s_1}}\,P\Delta^{-1}\psi
+\qquad \Longrightarrow
+\\
+\E_{s_2\sim \rho}
+\,Z(s_1,s_2)\psi(s_2)
+\approx \psi(s_1)+ \E_{s'\sim P(s'|s_1)}
+\E_{s_2\sim \rho}\,
+Z(s',s_2)\psi(s_2)
+\end{gather}
+for any $s_1$ and $\psi$.
+The right Bellman equation becomes
+\begin{gather}
+\transp{\1_{s_1}}\,\Delta^{-1}\psi=\transp{\1_{s_1}}\,\psi+\transp{\1_{s_1}}\,\Delta^{-1}P\psi
+\qquad \Longrightarrow
+\\
+\E_{s_2\sim \rho}
+Z(s_1,s_2)\psi(s_2)
+\approx \psi(s_1)+ \E_{s_2\sim \rho}\E_{s'_2\sim P(s'_2|s_2)}
+Z(s_1,s_2)\psi(s'_2)
+\end{gather}
+These equations impose a kind of smoothness on $Z$ with respect to the
+random walk $P$.
+
+The Newton update is
+\begin{equation}
+Z\gets (1+\eta)Z-\eta Z^2+\eta \gamma ZPZ
+\end{equation}
+which after applying to $\1_{s_1}$ on the left yields
+\begin{multline}
+Z(s_1,s_2)\gets (1+\eta)Z(s_1,s_2)-\eta \E_{s_3\sim \rho}
+Z(s_1,s_3)Z(s_3,s_2)\,+\\\eta \gamma \E_{s_3\sim\rho, s'_3\sim P(s'_3|s_3)}
+Z(s_1,s_3)Z(s'_3,s_2)
+\end{multline}
+and we should find a way to initialize $Z$ so that $Z\approx \Id$ at time
+$0$.
 

### 2018-12-19 paper/tangotd.tex

@@
 \end{equation}
+and in particular the value function is
+\begin{equation}
+V(s_1)=\E_{s_2\sim \rho} Z(s_1,s_2) R(s_2)
+\end{equation}
+This representation of the value function does not allow for first
+computing $\E_{s_2} R(s_2)f(s_2)$ online, then taking a dot product with
+$\green$. Here we would have to learn an auxiliary network to imitate
+$\E_{s_2\sim \rho} Z(s_1,s_2) R(s_2)$ (but this is a supervised learning
+problem). Alternatively, policy gradient can just directly be obtained
+via
+\begin{equation}
+\partial \pi(a|s)\E_{s'|s,a} \E_{s_2\sim \rho} Z(s',s_2)R(s_2)
+\end{equation}
+(this works even if $\rho$ is \emph{not} the invariant distribution).
 

### 2018-12-19 paper/tangotd.tex

@@
 
-I still don't know how to rightly combine these three equations.
+I still don't know how to rightly combine these three equations. I'm not
+even sure this kernel/image thing is actually an issue: if $f$ and
+$\green$ are represented as neural networks with some input features,
+the kernel of $f$ is actually the set of functions/distributions on
+states which have the same average features, and the image of
+$\transp{\green}$ is the set of functions that can be represented as sums
+$V(s)=\sum_i \alpha_i\green_i(s)$ over the components of $\green$: if $V$
+indeed belongs to this class then we will compute it correctly. (But this
+class does not change during training...)
 

### 2018-12-19 paper/tangotd.tex

@@
 \end{equation}
-and in particular the value function is
+In particular the value function is
 \begin{equation}
@@
 \end{equation}
-(this works even if $\rho$ is \emph{not} the invariant distribution).
+(this works even if $\rho$ is \emph{not} the invariant distribution). We
+can apply this each time a new reward is observed at some point $s_2$. We
+can even set another arbitrary distribution on $s_1$: this policy
+gradient will maximize
+the reward obtained from a starting point from that distribution.
 

### 2018-12-19 paper/tangotd.tex

@@
 and we should find a way to initialize $Z$ so that $Z\approx \Id$ at time
-$0$.
+$0$ ($Z(s,s)=1/\rho(s)$, although starting with $Z(s,s)=1$ should be
+enough). In the end this Newton equation alone might be sufficient to
+encompass the left and right Bellman. It does not rely on test functions,
+but the smoothing is implicit via the model used for $Z(s_1,s_2)$.
 

### 2018-12-19 paper/tangotd.tex

@@
 
-\paragraph{Variant.}
+\paragraph{Better variant?}
 I think
@@
 \end{equation}
-which after applying to $\1_{s_1}$ on the left yields
+which after applying to $\1_{s_1}$ on the left and $\1_{s_2}/\rho(s_2)$
+on the right yields
 \begin{multline}
@@
 
-\paragraph{Learning to navigate.} Define a function $V(s,s')$ that
+\paragraph{Early version of the above: Learning to navigate.} Define a function $V(s,s')$ that
 represents the time spent in $s'$ if starting at $s$ and acting optimally

### 2018-12-19 paper/tangotd.tex

@@
 \end{equation}
-This representation of the value function does not allow for first
+Downside: this representation of the value function does not allow for first
 computing $\E_{s_2} R(s_2)f(s_2)$ online, then taking a dot product with
@@
 
-The Newton update is
+The Newton update/middle Bellman equation is
 \begin{equation}

### 2018-12-19 paper/tangotd.tex

@@
 \begin{equation}
-\transp{\1_{s_1}}\Delta^{-1}\psi\approx \E_{s_2\sim \rho} Z(s_1,s_2)
+\transp{\1_{s_1}}\Delta^{-1}\psi\approx \E_{s_2\sim \rho} \,Z(s_1,s_2)
 \psi(s_2)
@@
 \begin{equation}
-V(s_1)=\E_{s_2\sim \rho} Z(s_1,s_2) R(s_2)
+V(s_1)=\E_{s_2\sim \rho} \, Z(s_1,s_2) R(s_2)
 \end{equation}
 Downside: this representation of the value function does not allow for first
-computing $\E_{s_2} R(s_2)f(s_2)$ online, then taking a dot product with
+computing a representation of $R$ as $\E_{s_2} R(s_2)f(s_2)$ online, then taking a dot product with
 $\green$. Here we would have to learn an auxiliary network to imitate
@@
 \begin{equation}
-\partial \pi(a|s)\E_{s'|s,a} \E_{s_2\sim \rho} Z(s',s_2)R(s_2)
+\partial \pi(a|s)\E_{s'|s,a} \E_{s_2\sim \rho} \, Z(s',s_2)R(s_2)
 \end{equation}
@@
 \\
-\E_{s_2\sim \rho}
+\E_{s_2\sim \rho}\,
 Z(s_1,s_2)\psi(s_2)
-\approx \psi(s_1)+ \E_{s_2\sim \rho}\E_{s'_2\sim P(s'_2|s_2)}
+\approx \psi(s_1)+ \E_{s_2\sim \rho}\E_{s'_2\sim P(s'_2|s_2)}\,
 Z(s_1,s_2)\psi(s'_2)
@@
 \begin{multline}
-Z(s_1,s_2)\gets (1+\eta)Z(s_1,s_2)-\eta \E_{s_3\sim \rho}
-Z(s_1,s_3)Z(s_3,s_2)\,+\\\eta \gamma \E_{s_3\sim\rho, s'_3\sim P(s'_3|s_3)}
+Z(s_1,s_2)\gets (1+\eta)Z(s_1,s_2)-\eta \E_{s_3\sim \rho}\,
+Z(s_1,s_3)Z(s_3,s_2)\,+\\\eta \gamma \E_{s_3\sim\rho,\, s'_3\sim
+P(s'_3|s_3)}\,
 Z(s_1,s_3)Z(s'_3,s_2)
 \end{multline}
-and we should find a way to initialize $Z$ so that $Z\approx \Id$ at time
-$0$ ($Z(s,s)=1/\rho(s)$, although starting with $Z(s,s)=1$ should be
-enough). In the end this Newton equation alone might be sufficient to
+which is perfectly samplable. We can simulate the assignment
+$Z(s_1,s_2)\gets...$ via gradient descent on the parameters of $Z$.
+
+We should find a way to initialize $Z$ so that $Z\approx \Id$ at time
+$0$, namely $Z(s,s)=1/\rho(s)$, although starting with $Z(s,s)=1$ should be
+quite fine. $Z>0$ (unless we work with $\gamma=1$ and relative stuff
+TODO), so the parameterization could reflect that.
+
+In the end this Newton equation alone might be sufficient to
 encompass the left and right Bellman. It does not rely on test functions,
 but the smoothing is implicit via the model used for $Z(s_1,s_2)$.
+
+Newton has a potential to be numerically less stable: first because of
+the $(1+\eta)$ factor which makes $Z$ quickly grow when it's small (until
+the $-Z^2$ kicks in), then because of the $-Z^2$ factors which can be
+problematic when $Z$ is large. Small enough learning rates should be
+stable.
 

### 2018-12-19 paper/tangotd.tex

@@
 $0$, namely $Z(s,s)=1/\rho(s)$, although starting with $Z(s,s)=1$ should be
-quite fine. $Z>0$ (unless we work with $\gamma=1$ and relative stuff
+quite fine. (Small initializations of $Z$ are safer.) $Z>0$ (unless we work with $\gamma=1$ and relative stuff
 TODO), so the parameterization could reflect that.
@@
 stable.
+
+Ideally, given the transition $s_3\to s'_3$, I'd like to sample $s_1$ and
+$s_2$ proportionally to $Z$ (would cancel the $Z$ factors, and avoid
+applying the equation to states that have nothing to do with $s_3$). But
+I don't know how to build an efficient sampler from $Z(s_3,\cdot)$.
+Something like Metropolis? but this function $Z$ is not constant...
 

### 2018-12-19 paper/tangotd.tex

@@
 $s_2$ proportionally to $Z$ (would cancel the $Z$ factors, and avoid
-applying the equation to states that have nothing to do with $s_3$). But
+applying the equation to states that have nothing to do with
+$s_3$—although the $Z$ factors are different for the three terms...). But
 I don't know how to build an efficient sampler from $Z(s_3,\cdot)$.

### 2018-12-19 paper/tangotd.tex

@@
 \end{equation}
-which after applying to $\1_{s_1}$ on the left and $\1_{s_2}/\rho(s_2)$
-on the right yields
+which yields (after applying to $\1_{s_1}$ on the left and $\1_{s_2}/\rho(s_2)$
+on the right)
 \begin{multline}
@@
 Something like Metropolis? but this function $Z$ is not constant...
+On-policy, sampling from $Z(\cdot,s_3)$ is easy: just sample one of the
+recent states with geometric rates $\gamma$. This should divide the
+equation by $Z(s_1,s_3)$ (up to some $\gamma$-dependent factor). I'm not
+sure what to do with the resulting $Z(s_1,s_2)$. On-policy, another
+option is to sample triplets $(s_1,s_3,s_2)$ that are some time apart on
+the trajectory, e.g., starting with $s_3$ uniform along the past
+trajectory, and moving back for $s_1$ and forward for $s_2$.
 

### 2018-12-19 paper/tangotd.tex

@@
 equation by $Z(s_1,s_3)$ (up to some $\gamma$-dependent factor). I'm not
-sure what to do with the resulting $Z(s_1,s_2)$. On-policy, another
+sure what to do with the resulting $Z(s_1,s_2)/Z(s_1,s_3)$. On-policy, another
 option is to sample triplets $(s_1,s_3,s_2)$ that are some time apart on
 the trajectory, e.g., starting with $s_3$ uniform along the past
-trajectory, and moving back for $s_1$ and forward for $s_2$.
+trajectory, and moving back for $s_1$ and forward for $s_2$ (with factors
+$\gamma$). Looks a lot like eligibility traces?
 

### 2018-12-19 paper/tangotd.tex

@@
 which is perfectly samplable. We can simulate the assignment
-$Z(s_1,s_2)\gets...$ via gradient descent on the parameters of $Z$.
+$Z(s_1,s_2)\gets...$ via gradient descent on the parameters of
+$Z(s_1,s_2)$. (Or via double sampling, which is not problematic here!?)
 
@@
 Something like Metropolis? but this function $Z$ is not constant...
+
 On-policy, sampling from $Z(\cdot,s_3)$ is easy: just sample one of the
 recent states with geometric rates $\gamma$. This should divide the
-equation by $Z(s_1,s_3)$ (up to some $\gamma$-dependent factor). I'm not
-sure what to do with the resulting $Z(s_1,s_2)/Z(s_1,s_3)$. On-policy, another
+right part of the equation by $Z(s_1,s_3)$ (up to some $\gamma$-dependent
+factor). Thus we would sample two states $s_1,s_2$ along the trajectory,
+sample a state $s_3$ in between on the trajectory from $s_1$, and (with
+the right $\gamma$ factors) we would get a term with $Z(s_1,s_2)$, a term
+with $Z(s_3,s_2)$ and a term with $Z(s'_3,s_2)$. This looks suspiciously
+like a large-scale Bellman from $s_1$... Variance could be higher since
+we don't use the model $Z$ and use sampling instead. To be tested.
+On-policy, another
 option is to sample triplets $(s_1,s_3,s_2)$ that are some time apart on

### 2018-12-19 paper/tangotd.tex

@@
 $\E_{s_2\sim \rho} Z(s_1,s_2) R(s_2)$ (but this is a supervised learning
-problem). Alternatively, policy gradient can just directly be obtained
+problem; still, it looks suspiciously like $TD(1)$ except maybe for the
+$O(1/(1-\gamma))$ variance). Alternatively, policy gradient can just directly be obtained
 via
@@
 like a large-scale Bellman from $s_1$... Variance could be higher since
-we don't use the model $Z$ and use sampling instead. To be tested.
+we don't use the model $Z$ and use sampling instead. To be tested. Write
+carefully because $s'_3$ should be uncorrelated from $s_2$!! So not
+possible with a single
+trajectory?
 On-policy, another

### 2018-12-19 paper/tangotd.tex

@@
 
+On-policy, working with $\gamma\to 1$ is quite easy: indeed $Z(s_1,s_2)$ is
+a density with respect to $\mu(s_2)$ so that the asymptotics when
+$\gamma\to 1$ is $Z(s_1,s_2)=\frac{1}{1-\gamma}+Z_0(s_1,s_2)$, and we can
+directly learn $Z_0$ by inserting $\frac{1}{1-\gamma}+Z_0$ in the middle
+Bellman equation. TODO check for simplifications.
+
 \paragraph{Early version of the above: Learning to navigate.} Define a function $V(s,s')$ that

### 2018-12-19 paper/tangotd.tex

@@
 $Z(s_1,s_2)\gets...$ via gradient descent on the parameters of
-$Z(s_1,s_2)$. (Or via double sampling, which is not problematic here!?)
+$Z(s_1,s_2)$. \footnote{Should we adjust the size of the gradient descent
+to get the right change of $Z(s_1,s_2)$? If we just use SGD, since
+gradients may change from point to point, the speed may be different at
+various states and we depart from Newton. This is clear when $Z$ is very
+small: Newton grows exponentially while a standard gradient descent would
+make small steps.} (Or via double sampling, which is not problematic here!?)
 
@@
 like a large-scale Bellman from $s_1$... Variance could be higher since
-we don't use the model $Z$ and use sampling instead. To be tested. Write
+we don't use the model $Z$ and use sampling instead; plus, we lose any
+generalization introduced by $Z$. To be tested. Write
 carefully because $s'_3$ should be uncorrelated from $s_2$!! So not
@@
 directly learn $Z_0$ by inserting $\frac{1}{1-\gamma}+Z_0$ in the middle
-Bellman equation. TODO check for simplifications.
+Bellman equation. This yields (omitting the expectations)
+\begin{multline}
+Z_0(s_1,s_2)\gets (1+\eta)Z_0(s_1,s_2)-\eta
+Z_0(s_1,s_3)Z_0(s_3,s_2)+\\\eta\gamma Z_0(s_1,s_3)Z_0(s'_3,s_2)-\eta
+Z_0(s_1,s_3)+\frac{\eta}{1-\gamma}(\gamma Z_0(s'_3,s_2)-Z_0(s_3,s_2))
+\end{multline}
+TODO CHECK which we cannot directly apply with $\gamma=1$...
 

### 2018-12-19 paper/tangotd.tex

@@
 \paragraph{Better variant?}
-I think
-everything works equally well if we just learn a kernel $Z(s_1,s_2)$ such
+Everything works equally well if we just learn a kernel $Z(s_1,s_2)$ such
 that
@@
 Z_0(s_1,s_2)\gets (1+\eta)Z_0(s_1,s_2)-\eta
-Z_0(s_1,s_3)Z_0(s_3,s_2)+\\\eta\gamma Z_0(s_1,s_3)Z_0(s'_3,s_2)-\eta
+Z_0(s_1,s_3)(Z_0(s_3,s_2)-\gamma Z_0(s'_3,s_2))\\-\eta
 Z_0(s_1,s_3)+\frac{\eta}{1-\gamma}(\gamma Z_0(s'_3,s_2)-Z_0(s_3,s_2))
@@
 TODO CHECK which we cannot directly apply with $\gamma=1$...
+However ON-POLICY $s_3$ and
+$s'_3$ have the same law, so in expectation the last term is $-\eta
+\E_{s_3} Z_0(s_3,s_2)$. This yields the more symmetric
+\begin{multline}
+Z_0(s_1,s_2)\gets (1+\eta)Z_0(s_1,s_2)-\eta
+Z_0(s_1,s_3)\left(Z_0(s_3,s_2)-\gamma Z_0(s'_3,s_2)\right)\\-\eta
+Z_0(s_1,s_3)-\eta Z_0(s_3,s_2)
+\end{multline}
+on-policy, for any $\gamma\leq 1$.
 

### 2018-12-19 paper/tangotd.tex

@@
 $0$, namely $Z(s,s)=1/\rho(s)$, although starting with $Z(s,s)=1$ should be
-quite fine. (Small initializations of $Z$ are safer.) $Z>0$ (unless we work with $\gamma=1$ and relative stuff
-TODO), so the parameterization could reflect that.
+quite fine. (Small initializations of $Z$ are safer.) $Z>0$ (unless we
+work with $\gamma=1$ and relative stuff), so the parameterization should
+reflect that. Indeed the Newton iteration is unstable with negative
+values.
 
@@
 problematic when $Z$ is large. Small enough learning rates should be
-stable.
+stable. (TODO NOT TRUE: even in 1d, Newton is unstable if initialized to
+a negative number, for any learning rate. Stable only in some
+neighborhood of the true solution, but then, the stochastic noise might
+produce divergence...)
 
@@
 
+[TODO: Since the standard iteration is unstable for $Z<0$, this means the
+iteration on $Z_0$ will be unstable for $Z_0<-\frac{1}{1-\gamma}$.
+Use that somehow: clip at that value?]
+
 \paragraph{Early version of the above: Learning to navigate.} Define a function $V(s,s')$ that

### 2018-12-20 paper/tangotd.tex

@@
 
-We should find a way to initialize $Z$ so that $Z\approx \Id$ at time
+Initialization: A first idea would be to start with paths from $s$ to
+$s$, namely, initialize
+$Z$ so that $Z\approx \Id$ at time
 $0$, namely $Z(s,s)=1/\rho(s)$, although starting with $Z(s,s)=1$ should be
@@
 values.
+
+Another initialization might be $Z(s_1,s_2)\approx \frac{1}{1-\gamma}$
+for every pair $s_1,s_2$. Indeed, on-policy, $Z$ is the density with
+respect to the invariant distribution $\mu$, and for $\gamma$ close to
+$1$, $Z$ will be approximately $\frac{1}{1-\gamma}$ plus a smaller function (see
+below). Note that $Z(s_1,s_2)=\frac{1}{1-\gamma}$ is an exact solution of
+the middle Bellman equation (as is $Z=0$), but it is a repulsive fixed
+point, I think, so initializing to $\frac{1}{1-\gamma}$ plus noise should
+converge to the true solution. Initializing there should have the
+advantage of starting close to reasonable values.
 
@@
 neighborhood of the true solution, but then, the stochastic noise might
-produce divergence...)
+produce divergence... this is potentially a serious issue)
 

### 2018-12-20 paper/order_magnitude.tex
Text on order of magnitude critic + corresponding updates in the code

\documentclass[11pt]{article}
\usepackage{amssymb, amsmath}

\newcommand{\deltat}{\delta \hspace{-.1em}t}
\newcommand{\gauss}{\mathcal{N}}
\begin{document}
One can parameterize the advantage function as
\begin{equation}
	A_\phi(s, a) = (\sigma_\phi(s, a) A^m_\phi(s, a) + (1 - \sigma_\phi(s, a)) A^M_\phi(s, a)) \deltat^{\sigma_\phi(s, a)}.
\end{equation}
This way, $A_\phi$ takes macroscopic values when $\sigma_\phi = 0$ and microscopic values when $\sigma_\phi = 1$. The $A^m$ part of the function is used to model \emph{microscopic} values, while $A^M$ is used to model
\emph{macroscopic} values.

The full probabilistic model used is
\begin{equation}
	p(A(s, a) \mid \phi) = \gauss\left(A(s, a)\mid A_\phi(s, a), \sqrt{\deltat^{\sigma_\phi(s, a)}}^2\right).
\end{equation}
Given a dataset of triplets $(s, a, \tilde{A}(s, a))$, the loss on a single datapoint is,
up to an additive constant independant of $\phi$,
\begin{equation}
	\mathcal{L}_\phi = \log{p(\tilde{A}(s, a) \mid \phi)} = -\frac{\left(A_\phi(s, a) - \tilde{A}(s, a)\right)^2}{2 \deltat^{\sigma_\phi(s, a)}} - \frac{\sigma_\phi}{2} \log{\deltat}.
\end{equation}
Splitting $\phi$ into $(\phi^m, \phi^M, \theta)$, with $A^m$ (resp. $A^M$, $\sigma$) depending only on
$\phi^m$ (resp. $\phi^M$, $\theta$), gradients of the loss w.r.t. these parameters are
\begin{align}
	\partial_{\phi^m} \mathcal{L} =& - \sigma_\phi(s, a) \deltat^{\sigma_\phi(s,a)}
	\frac{\delta A_\phi(s, a)}{\deltat^{\sigma_\phi(s, a)}}
	\partial_{\phi^m} A^m_{\phi^m}(s, a)\\
	\partial_{\phi^M} \mathcal{L} =& - (1 - \sigma_\phi(s, a)) \deltat^{\sigma_\phi(s,a)}
	\frac{\delta A_\phi(s, a)}{\deltat^{\sigma_\phi(s, a)}}
	\partial_{\phi^M} A^M_{\phi^M}(s, a)\\
	\partial_{\theta} \mathcal{L} =&  \frac{1}{2}\left(\left(\left(\frac{\delta A_\phi(s, a)\log{\deltat}}{\deltat^{\sigma_\phi(s,a)}} - \left(A^m_\phi(s, a) - A^M_\phi(s, a)\right)\right)\delta A_\phi(s, a) - \log{\deltat}\right)\partial_\theta \sigma_\theta(s, a)\\
\end{align}
with $\delta A_\phi(s, a) = A_\phi(s, a) - \tilde{A}(s, a)$.

Now assume that $\tilde{A}(s, a)$ is or order $\deltat$ with probability roughly $1 - \deltat$, and
of order $1$ with probability $\deltat$. Further assume that $\sigma$ is close to optimal, in the sense
that $\sigma$ is close to $1$ when $\tilde{A}$ is of order $\deltat$, and close to $0$ when $\tilde{A}$
is of order $1$. Under these assumptions, $\delta A_\phi$ is of order $\deltat^\sigma(s, a)$, and the
following orders of magnitudes for the gradients are obtained for $\phi^m$ and $\phi^M$
\begin{align}
	\delta_{\phi^m} \mathcal{L} \approx& \deltat\\
	\delta_{\phi^M} \mathcal{L} \approx& \deltat.
\end{align}
The gradient with respect to $\theta$ could be problematic. When $\deltat$ is small, the recall term
$\log{\deltat} \partial_\theta \sigma_\theta$ dominates, meaning that $\sigma wi

\end{document}

### 2018-12-20 pendulum_exp/critics/order_advantage.py

@@ -13,16 +13,24 @@ from nn import soft_update
 class OrderAdvantageCritic(CompoundStateful, Critic):
     def __init__(self,
                  dt: float, gamma: float, lr: float, tau: float, optimizer: str,
-                 val_function: ParametricFunction, adv_function: ParametricFunction) -> None:
+                 val_function: ParametricFunction, adv_function: ParametricFunction,
+                 sigma_function: ParametricFunction) -> None:
         CompoundStateful.__init__(self)
         self._reference_obs: Tensor = None
         self._val_function = val_function
         self._adv_function = adv_function
+        self._sigma_function = sigma_function
         self._target_val_function = copy.deepcopy(val_function)
         self._target_adv_function = copy.deepcopy(adv_function)
+        self._target_sigma_function = copy.deepcopy(sigma_function)
 
         self._adv_optimizer = \
             setup_optimizer(self._adv_function.parameters(),
+                            opt_name=optimizer, lr=lr, dt=dt,
+                            inverse_gradient_magnitude=1,
+                            weight_decay=0)
+        self._sigma_optimizer = \
+            setup_optimizer(self._sigma_function.parameters(),
                             opt_name=optimizer, lr=lr, dt=dt,
                             inverse_gradient_magnitude=1,
                             weight_decay=0)
@@ -37,6 +45,23 @@ class OrderAdvantageCritic(CompoundStateful, Critic):
         self._tau = tau
 
         self._device = 'cpu'
+
+    def compute_advantages(self, obs: Arrayable, action: Tensorable, target: bool = False):
+        func = self._adv_function if not target else self._target_adv_function
+        sigma_func = self._sigma_function if not target else self._target_sigma_function
+        if len(func.input_shape()) == 2:
+            advs = func(obs, action).squeeze()
+            sigma = sigma_func(obs, action).squeeze()
+        else:
+            advs = func(obs)
+            sigma = sigma_func(obs)
+            action = check_tensor(action, self._device).long()
+            advs = advs.gather(1, action.view(-1, 1)).squeeze()
+            sigma = sigma.gather(1, action.view(-1, 1)).squeeze()
+        sigma = torch.sigmoid(sigma).squeeze()
+        adv = (advs * torch.stack([sigma, 1 - sigma], dim=-1)).sum(dim=-1) * \
[…]

### 2018-12-21 paper/order_magnitude.tex
Very quick aside on rmsprop (probably needs a bit more caution)

@@
 The gradient with respect to $\theta$ could be problematic. When $\deltat$ is small, the recall term
-$\log{\deltat} \partial_\theta \sigma_\theta$ dominates, meaning that $\sigma wi
+$\log{\deltat} \partial_\theta \sigma_\theta$ dominates. This gradient is always of order $1$. A
+consequence is that we should avoid parameter sharing between $\theta$ and the other parameters, as
+the gradient on $\sigma$ would almost always dominate.
+
+\paragraph{An aside on the use of RMSProp}
+Empirically, RMSProp helps a lot in term of convergence speed. Special care should be put on the batch
+size when using RMSProp. Recall that RMSProp updates network parameters as
+\begin{align}
+	H &\leftarrow (1 - \alpha) H + \alpha \text{diag}(g \odot g)\\
+	\phi &\leftarrow \phi - \eta H^{-1/2} g
+\end{align}
+with $g = \partial_\phi \mathcal{L}$.
+
+With batch size $1$ or close to $1$, in
+the one dimensional case and in term of orders of magnitude, for both $\phi^M$
+and $\phi^m$, $g \approx B \deltat + (1 - B)$ with $B\sim \mathcal{B}(1 -
+\deltat)$ a bernoulli with $\mathbb{P}(B = 1) = 1 - \deltat$. If the
+expectation of $H$ reaches a stationnary regime, it is easily verifiable that
+$\mathbb{E} H \approx \deltat$, and consequently that learning rates are divided
+by $\sqrt{\deltat}$.
+
+With batch sizes of order $1 / \deltat$, as the variance of $g$ is no longer of order
+$\deltat$, but of order $\deltat^2$, learning rates are divided
+by $\deltat$, resulting in update steps that are effectively of order $\eta$.
 

### 2018-12-21 paper/tangotd.tex

@@
 
+The above can be turned to an infinite-rank approximation by writing
+\begin{equation}
+\transp{1_{s_1}}Z\psi=\E_{s_2\sim \rho}\, \E_{\omega}\,
+\green(s_1,\omega)f(s_2,\omega)\psi(s_2)
+\end{equation}
+where integration is over $\omega$ in an arbitrary space (eg $\omega\in
+[0;1]$ or $\omega\in \R^d$ with Gaussian distribution). This amounts to
+sending each state to a function over $\omega$, and using the dot product
+from $L^2(\omega)$. This should solve some kernel/image issues and
+increase representational power. The equation
+$\Id=F\Delta\transp{\green}$ becomes $\delta_{\omega\omega'}=\E_{s\sim
+\rho, \,s'\sim P(s'|s)}
+f(s,\omega)(\green(s,\omega')-\green(s',\omega'))$ TODO CHECK, which is not easy to
+solve because of the singular $\delta_{\omega\omega'}$, but should be better with things on the right and left.
+
 \paragraph{Better variant?}

### 2018-12-21 paper/tangotd.tex

@@
 both of which can be implemented via gradient descent on the parameters
-of $\green$ and $f$. (This double the learning rate $\eta$.)
+of $\green$ and $f$. (This doubles the learning rate $\eta$.)
 There is no need to sample on test functions.
@@
 \end{equation}
+[TODO CHANGE NOTATION because currently the entries of $Z$ are not
+$Z(s_1,s_2)$ but $Z_{s_1s_2}=\rho(s_2)Z(s_1,s_2)$ which is confusing...]
 In particular the value function is

### 2018-12-21 paper/tangotd.tex

@@
 small: Newton grows exponentially while a standard gradient descent would
-make small steps.} (Or via double sampling, which is not problematic here!?)
+make small steps.} (Or via double sampling, which is not problematic
+here!? but would converge to trivial solutions like $Z=0$ or
+$Z=1/(1-\gamma)$, which are repulsive with Newton.)
 

### 2018-12-21 paper/tangotd.tex

@@
 f(s,\omega)(\green(s,\omega')-\green(s',\omega'))$ TODO CHECK, which is not easy to
-solve because of the singular $\delta_{\omega\omega'}$, but should be better with things on the right and left.
+solve because of the singular $\delta_{\omega\omega'}$, but should be
+better with things on the right and left. For instance, the Newton equation on
+$\transp{\green}$ is
+\begin{equation}
+\green(s_1,\omega)\gets
+(1+\eta)\green(s_1,\omega)-\eta\E_{s\to s'}\E_{\omega'}
+\green(s_1,\omega')f(s,\omega')(\green(s,\omega)-\gamma
+\green(s',\omega))
+\end{equation}
 
@@
 
+The Newton update $Z\gets 2Z-Z\Delta Z$ is stable provided it is
+initialized to some $Z$ such that $\Id-\Delta Z$
+has operator norm $< 1$. With $\Delta=\Id-\gamma P$, the initial value
+$Z=\Id$ satisfies this provided $\gamma<1$ (and probably $\gamma=1$ for
+the relative case). Here we use a stochastic update, which should follow
+the deterministic dynamics if the learning rate is small enough. However,
+in general we will not apply the update $Z\gets (1+\eta)Z-\eta Z\Delta
+Z$, even in expectation, because we will also sample $s_1$ and $s_2$.
+Sampling $s_1$ according to $\rho$ amounts to multiplying the update by
+$\rho$. This leaves the fixed point unchanged, but may change the
+stability TODO CHECK. Using a gradient descent on parameters will further
+introduce a preconditioner on the update. TODO CHECK.
+
+For discrete time but continuous state, the true $Z=\sum \gamma^t P^t$ is
+equal to $\Id$ plus a continuous component; the $\Id$ is singular, so
+learning $Z(s_1,s_2)$ using eg a neural network will be difficult around
+the diagonal (and the initialization will be difficult). We may directly
+write the dynamics on $Z-\Id$. (TODO not clear what this means in
+continuous time: not clear if $\Delta^{-1}-\Id$ is actually small; still,
+starting at $\Id$ is stable for Newton whatever $P$ is, so it should be
+stable in the limit.)
+
+Setting $Z=\Id+\zeta$ provides the update
+\begin{equation}
+\zeta\gets (1-\eta)\zeta+\eta\gamma P+\eta\gamma P\zeta
++\eta\gamma \zeta P-\eta\zeta(\Id-\gamma P)\zeta
+\end{equation}
+which we cannot quite sample (because of the $P$ term). But it mixes the
+left and right Bellman equations nicely. A way to sample it is to sample
+the left state $s_1$ according to $\rho$; this amounts to multiplying the
+update by $\rho$ (same fixed point, but maybe different dynamics TODO),
+and then $\rho P$ is samplable.
+
+The convergence of the update above to its fixed point $\gamma
+P(\Id-\gamma P)^{-1}$ can be studied by studying the
+remainder $r\deq \zeta-\gamma
+P(\Id-\gamma P)^{-1}$, whose dynamics is
+\begin{equation}
+r\gets (1-\eta) r-\eta r(\Id-\gamma P)r
+\end{equation}
+In particular, at first order for small $r$ this is just $r\gets
+(1-\eta)r$. This is contracting \emph{independently of the properties of
+$P$}, contrary to the Bellman equation. The stochastic
+Newton update with learning rate $\eta$ does not have quadratic
+convergence like the standard Newton method, but it still keeps better
+local convergence properties than Bellman. Is it worth the cost?
+
 \paragraph{Early version of the above: Learning to navigate.} Define a function $V(s,s')$ that

### 2019-01-07 paper/icml_drau/icml_drau.tex
Paper in progress

--- 2018-12-05 paper/icml_drau/icml_drau.tex
@@
 \usepackage{amsmath, amssymb}
+\usepackage{mathtools}
 \usepackage{microtype}
@@
 \usepackage{xcolor}
+\usepackage{algorithmic}
 
@@
 %\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
-\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
+%\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
 
 \begin{abstract}
-
+	Despite remarkable successes, \emph{Deep Reinforcement learning} remains sensitive to
+	hyperparameterization, implementation details and small environment changes~(\citealt{drl_matter}, \citealt{drl_matter_bis}). 
+	Overcoming this sensitivity is key to making DRL applicable to real world problems.
+	In this paper, we study the sensitivity of Deep Reinforcement learning algorithms to
+	time discretization. More precisely, we show that approaches based on estimations of the
+	Q-function, e.g. \emph{Deep Q-learning}~\cite{dqn} and \emph{Deep deterministic policy gradient}~\cite{ddpg}
+	are sensitive to variations of time discretization. We further show that a simple reparameterization of the
+	Q-function as a sum of a state value term and a \emph{small} action dependent advantage term yields an algorithm
+	much more resilient to variations of time discretization.
 \end{abstract}
@@
 \label{sec:intro}
+In recent years, deep learning based approaches to reinforcement learning have
+thrived and provided impressive results in a variety of domains, achieving superhuman
+performances with no expert knowledge in perfect information zero-sum
+games~\cite{alphazero}, reaching level of top players in video
+games~(\citealt{openai_five}, \citealt{dqn}), or learning dexterous manipulation
+from scratch without demonstrations~\cite{hand_control}.
+
+In spite of those successes, \emph{Deep reinforcement learning} approaches are
+sensitive to a number of factors, including hyperparameterization,
+implementation details or small changes in the environment
+parameters~(\citealt{drl_matter}, \citealt{drl_matter_bis}). This sensitivity,
+along with sample inefficiency, prevents DRL from being applied in most real
+world settings. High sensitivity to environment parameters notably prevents
+transfering abilities from imperfect simulators to real world scenarios.
+
+In this paper we focus our attention on the sensitivity to time discretization
+of DRL approaches, i.e. what happens when an agent receives $50$ observations
+and is expected to take $50$ actions per second instead of $10$. One expects
+that decreasing time discretization, or equivalently shortening reaction time,
+should only improve agent performances. This is not what happens in practice
+for approaches based on estimation of state action value functions, e.g.
+\emph{Deep Q-learning} (DQN~\citep{dqn} and \emph{Deep deterministic policy
+gradient} (DDPG~\citep{ddpg}). This is shown experimentally in Sec.~\ref{sec:exp}.
+
+We relate this sensitivity to the fact that, as the discretization timestep
+decreases, the effect of individual actions on the total return decreases, and
+can vanish when the discretization timestep becomes infinitesimal. The natural
+framework to study this phenomenon is Continuous time reinforcement learning 
+(Sec.~\ref{sec:continous}).
+
+To cope for this effect, we introduce a
+parameterization of the Q-function as a sum of a state value term and a
+\emph{small} action dependent advantage term (Sec.~\ref{subsec:reparam}). This
+reparameterized Q approximation can be trained with deep variants of Q-learning
+(Sec.~\ref{subsec:algorithm}).  To properly scale w.r.t. time discretization,
+particular attention has to be taken regarding both the exploration method, and
+the scaling of learning rates (Sec.~\ref{subsec:lr}, Sec.~\ref{subsec:explo}).
+The resulting algorithm is shown to provide near perfect invariance to time
+discretization on simple environments, and much better invariance properties
+than vanilla DQN or DDPG on more complex environments (Sec.~\ref{sec:exp}).
 
@@
 \label{sec:related}
+Continuous and near continuous reinforcement learning, which is at the core of
+our approach has been thoroughly studied in the deterministic case
+in~\cite{adv_upd} or \cite{cont_rl}, but has not been applied in the context of
+deep reinforcement learning.
 
-\section{Continuous reinforcement learning and scaling problems}
+The approach presented in this paper is closely related to \cite{adv_upd} and
+\cite{cont_rl}. \cite{adv_upd} already separated the learning procedure of the
+value and advantage functions, and introduced proper scalings between the two
+components. We extend on \cite{adv_upd} in three directions. We provide a
+formal argument on why the advantage contribution of the Q-function is of order
+$\bigO(\deltat)$, irrelevant of the stochasticity of the environment, in a quite
+general framework. We show that, provided with well suited learning rates and
+exploration method, Advantage updating provide near invariance to change of time
+discretization. We show that Advantage updating is viable in the context of deep
+reinforcement learning, while not using the true gradient of the residual error.
+
+More recently, \cite{dueling_nets} also introduced a parameterization separating
+the value and the advantage components of the Q-function. We additionally
+introduce a natural scaling coeficient between the two, that strongly improves
+the time discretization invariance properties.
+
+\section{Continuous and near continuous time reinforcement learning}
 \label{sec:continous}
-In the following section, a restrictive framework for near continuous
-time reinforcement learning is introduced, and the main challenges of this
-setup are explained. \RKC{This is not what we want to write: we want to explain
-why current reinforcement learning practices are ill suited for this framework}
+In this section, a continuous reinforcement learning framework is introduced,
+and it is shown that under this framework, the state-action value function does
+not depend on the action. We further show that in near continous domains, when
+the discretization timestep $\deltat$ is small, the difference in state-action
+value between two different actions in a fixed state is of order $\bigO(\deltat)$.
 
 \subsection{Framework}
+In what follows, $\dt$ denotes a truly infinitesimal time interval while $\deltat$
+denotes a small, but non infinitesimal time interval. Similarily, $\dbt$ denotes
+an infinitesimal (multi-dimensional) brownian step, while $\deltabt$ denotes its
+discretized equivalent. Informally, $\dbt$ is to be thought of as $\gauss(0, \sqrt{\dt}^2)$,
+and $\deltabt$ as $\gauss(0, \sqrt{\deltat}^2)$, where $\gauss(\mu, \sigma)$ denotes a gaussian
+random variable of mean $\mu$ and standard deviation (or covariance matrix in the multi-dimensional case)
+$\sigma$.
+
 Consider a fully observable \emph{Markov Decision Process} defined by the following
@@
 \begin{align}
-	s_{t + \deltat} &= s_t + F(s_t, a_t) \deltat + \Sigma(s_t, a_t) \deltabt\\
-	r_{t} &= \reward(s_t, a_t) \deltat + \Sigma_r(s_t, a_t) \deltabt^r
+	ds(s, a) &= F(s, a) dt + \Sigma(s, a) \dbt\\
+	dr(s, a) &= \reward(s, a) dt + \Sigma_r(s, a) \dbt.
 \end{align}
-where the $s$'s are $n$-dimensional vectors, $F(s_t, a_t)$ is a $n$-dimensional
-drift coeficient, $\Sigma(s_t, a_t)$ is a $n \times n$ diffusion matrix,
-$\deltat$ is a small but finite discretization step, and $\deltabt$ is a sample
-from $\gauss(0, \deltat I)$.
+The equation on $s$ is that of a diffusion process, with drift coeficient $F$
+and diffusion matrix $\Sigma$.  This framework is only moderately restrictive.
+For example, deterministic continuous control environments fall in this
+framework, with $\Sigma = 0$ and $\Sigma_r = 0$. Limitations of the framework
+are discussed in Sec.~\ref{sec:limitations}.
+This framework is naturally discretized to arbitrary $\deltat$ by replacing all
+$d$'s with $\delta$'s.
 
-This framework is quite general, and most discretized deterministic continuous
-control environments fall in this framework, with $\Sigma = 0$ and $\Sigma_r =
-0$, as well as many stochastic control environments.
+A continuous policy $\pi$ is defined as a deterministic mapping from states to
+actions.  In continuous time, defining stochastic policies with well defined
+state-value functions requires incorporating information from previous action into
+the state to obtain a temporally coherent policy. To avoid such inconvenience, the
+framework is restricted to deterministic policies. This is a mild constraint, as this
+only enforces determinism of the exploitation policy and not of exploration policies.
+In what follows, $\tau\sim\pi$ denotes a trajectory of states and actions sampled
+according to policy $\pi$, with $s_0$ sampled according to an arbitrary initial distribution
+on states $\rho_0$, and $dr_{t} = dr(s_t, a_t)$.
 
-A deterministic policy $\pi$ is defined as a deterministic mapping from state
-space to action space\footnote{We restrict ourselves to deterministic policies.
-	In continuous time, defining stochastic policies with well defined
-	state-value function is non trivial. In the discrete case, the
-distribution of actions of a policy given a state only depends on the previous
-state, and not on previous actions. In continuous time, such policies are akin
-to white noise, and do not define stochastic processes.}.
-
-Under a deterministic policy $\pi$, one can define the \emph{state value function} $V^\pi(s)$
-and \emph{state action value function} $Q^\pi(s, a)$ as
+Under a deterministic policy $\pi$, one can define the $\gamma$ discounted
+\emph{state value function} $V^\pi(s)$ and \emph{state action value function}
+$Q^\pi(s, a)$ as
 \begin{align}
-
+	V^\pi(s) &= \E_{\tau\sim\pi}\left[
+		\int\limits_{t=0}^\infty \gamma^{t}
+		dr_{t} \mid s_0 = s
+	\right]\\
+	Q^\pi(s, a) &= \E_{\tau\sim\pi}\left[
+		\int\limits_{t=0}^\infty \gamma^{t}
+		dr_{t} \mid s_0 = s, a_0 = a
+	\right]
+\end{align}
+which are naturally discretized as
+\begin{align}
+	V^\pi(s) &= \E_{\tau\sim\pi}\left[
+		\sum\limits_{k=0}^\infty \gamma^{k\deltat}
+		\delta r_{k\deltat} \mid s_0 = s
+	\right]\\
+	Q^\pi(s, a) &= \E_{\tau\sim\pi}\left[
+		\sum\limits_{k=0}^\infty \gamma^{k\deltat}
+		\delta r_{k\deltat} \mid s_0 = s, a_0 = a
+	\right].
+\end{align}
+Both $V^\pi$ and $Q^\pi$ verify Bellman equations, informally
+\begin{align}
+	V^\pi(s) &= \reward(s, \pi(s))\dt + \gamma^{\dt} \E_{ds \sim ds(s, \pi(s))} V^\pi(s + ds)\nonumber\\
+	Q^\pi(s, a) &= \reward(s, a)\dt + \gamma^{\dt} \E_{ds(s, a)} V^\pi(s + ds(s, a))
+	\label{eq:cont_bell}
+\end{align}
+which in continuous time translates, using Ito's lemma, into a Hamilton Jacobi equation
+\begin{align}
+	\reward + \partial_s V^\pi \cdot f + \frac{1}{2} \text{tr}\left(\Sigma^T\partial^2_{s^2} V^\pi\Sigma\right) = (1 - \gamma) V^\pi(s).
 \end{align}
 
-\section{Multi-chain continuous environments and biscale advantage}
-\label{sec:multi-chain}
+\subsection{Independence of $Q^\pi$ on $a$ in continuous time}
+Consider a discretization of Eq.~\eqref{eq:cont_bell} with time interval
+$\deltat \ll~1$, Ito's lemma yield
+\begin{align}
+	Q^\pi(s, a) &= \reward(s, a) \deltat + \gamma^{\deltat} \E_{\delta s \sim \delta s(s, a)}V^\pi(s + \delta s)\nonumber\\
+		    &= \reward(s, a) \deltat + \gamma^{\deltat} V^\pi(s) +
+		     \gamma^{\deltat} \partial_s V^\pi(s)f(s, a) \deltat\nonumber\\
+		    &+ \frac{1}{2}\text{tr}\left(\Sigma^T\partial^2_{s^2}V^\pi(s)\Sigma\right)\deltat + \bigO(\deltat^2)\nonumber\\
+		    &\xrightarrow[\deltat \to 0]{} V^\pi(s).
+		    \label{eq:Q_cont}
+\end{align}
+A consequence of Eq.~\eqref{eq:Q_cont} is that in exact continuous time,
+$Q^\pi$ does not bear any information on the hierarchy of actions, and
+thus cannot be used to select actions that yields higher future returns.
 
-\section{Deep RelAtivistic Mixture of Advantage learning}
-\label{sec:algo}
+As the difference between between $Q^\pi(s, a)$ and $V^\pi(s)$ is of order
+$\bigO(\deltat)$, it is natural to define a rescaled version of the advantage
+function, namely
+\begin{align}
+	A^\pi(s, a) &= \frac{Q^\pi(s,a) - V^\pi(s)}{\deltat}\\
+		    &= \reward(s, a) +
+		    \frac{\gamma^{\deltat} \E_{\delta s \sim \delta s(s, a)} V^\pi(s + \delta s) - V^\pi(s)}{\deltat}.
+    \label{eq:adv}
+\end{align}
+Contrary to the action state value function, the rescaled advantage function converges
+to an action dependent quantity when $\deltat$ goes to $0$
+\begin{align}
+	A^\pi(s, a) &\xrightarrow[\deltat \to 0]{} \reward(s, a) +
+	(\gamma - 1) V^\pi(s) + \partial_s V^\pi(s) f(s, a)\nonumber\\
+		    &+
+	\frac{\text{tr}\left(\Sigma^T(s, a)\partial^2_{s^2}V^\pi(s) \Sigma(s,a)\right)}{2},
+\end{align}
+and $Q^\pi$ can then be rewritten as
+\begin{equation}
+	Q^\pi(s, a) = V^\pi(s) + \deltat A^\pi(s, a).
+	\label{eq:reparam_q_pi}
+\end{equation}
+
+\section{Deep Advantage Updating}
+In continuous time, $Q^\pi$ is independant of actions and can thus not be used
+to select promising actions.  In near continuous time, $Q^\pi$ still depends on
+actions, and can still be used to choose promising actions under the current
+policy. However, when approximating $Q^\pi$, if the approximation error is much
+bigger than $\bigO(\deltat)$, approximation errors dominates, and the hierarchy
+of actions given by the approximation is likely to be erroneous.
+
+When the approximate Q-function is initialized, if the effect of actions on the
+Q-function is order of magnitudes higher than what it should be, approximating
+$Q^\pi$ is likely to be difficult. Besides, the error is likely to be further
+propagateed when the equation used to update our approximate $Q$ function
+relies on bootstraping, as is the case for usual temporal difference derived methods.
+
+\subsection{Reparameterizing $Q_\theta$}
+\label{subsec:reparam}
+To prevent overestimating the dependency on actions of the approximation of
+$Q^\pi$, $Q_\theta$, in an environment with time discretization $\deltat$,
+Eq.~\eqref{eq:reparam_q_pi} naturally yields the following reparameterization
+\begin{equation}
+	Q_\theta(s, a) = V_{\theta_1}(s) + \deltat A_{\theta_2}(s, a).
+\end{equation}
+To avoid cumbersome notations, in what follows, parameter indices are dropped.
+The insight behind this parameterization is that when $Q$ correctly
+approximates $Q^\pi$, $V$ should approximate the value function, while $A$
+should approximate the advantage function, as defined in Eq.~\eqref{eq:adv}.
+
+The proposed reparameterization does not, on its own, guarantee that
+when $Q$ correctly approximates $Q^\pi$, $A$ approximates $A^\pi$.  Indeed, for
+any given pair $V$, $A$, the transformed pair
+\begin{align}
+	\tilde{V}(s) &= V(s) - f(s)\\
+	\tilde{A}(s, a) &= A(s, a) + \frac{f(s)}{\deltat}
+\end{align}
+yields the exact same $Q$ function. Consequently, $Q$ can correctly approximate
+$Q^\pi$ even if $A$ is arbitrarily far from $A^\pi$.
+To enforce, identifiability of $A$, one must enforce that 
+\begin{equation}
+	A(s, \pi(s)) = 0.
+\end{equation}
+With this additional constraint, if $Q^\pi = Q$,
+\begin{align}
+	A^\pi(s, a) &= \frac{Q^\pi(s,a) - V^\pi(s)}{\deltat}\\
+		    &= \frac{Q(s,a) - Q(s, \pi(s))}{\deltat}\\
+		    &= A(s, a).
+\end{align}
+In the spirit of~\cite{dueling_nets}, this condition is enforced by writing $A$ as
+\begin{equation}
+	A(s, a) = \bar{A}(s, a) - \bar{A}(s, \pi(s)).
+\end{equation}
+
+\subsection{Time invariant exploration}
+\label{subsec:explo}
+To obtain a time invariant reinforcement learning algorithm, a time invariant
+exploration scheme is required.
+
+The exploration scheme used~\cite{ddpg}, which consists in adding an
+\emph{Orhnstein-Ullenbeck}~\cite{orn-ull} process to actions, is naturally time
+invariant in the sense that it is first defined as a time continuous stochastic
+process, and that discretization of this process are used in practice. Indeed,
+a continuous \emph{Orhnstein-Ullenbeck} process is defined as
+\begin{equation}
+	dx_t = - (x_t - \mu) \theta dt + \sigma \dbt
+\end{equation}
+and is naturally discretized by turning $d$'s into $\delta$'s, and using
+the same $\deltat$ for both the learning algorithm and the exploration scheme.
+Consequently, this exploration scheme can readily be used in continuous action
+environments.
+
+On the other hand, the usual exploration scheme used in discrete actions
+Q-learning, $\varepsilon$-greedy exploration~\cite{sutton} is not time
+invariant, in the sense that resulting state trajectories do not converge to a
+well defined stochastic process as $\deltat \rightarrow 0$.
+
+\subsection{Learning rates scalings}
+\label{subsec:lr}
+
+\subsection{Algorithms}
+\label{subsec:algorithm}
+We introduce two versions of \emph{Deep Advantage Updating} to cover both the
+continuous and discrete action cases. Both relies on applying either DDPG or DQN
+to the reparameterized Q-function and enforcing the identifiability of $A^\pi$ by
+making sure that $\max_a A^\pi(s, a) = 0$, either approximately or exactly for all
+states.
+\begin{algorithm}[ht]
+	\caption{Deep Advantage Updating (Discrete actions)}
+	\input{dauc.tex}
+\end{algorithm}
 
@@
 
+\section{Limitations}
+\label{sec:limitations}
 \section{Conclusion}

### 2019-01-07 pendulum_exp/critics/advantage.py

@@ -41,32 +41,26 @@ class AdvantageCritic(CompoundStateful, Critic):
     def optimize(self, obs: Arrayable, action: Arrayable, max_action: Tensor,
                  next_obs: Arrayable, max_next_action: Tensor, reward: Arrayable,
                  done: Arrayable, time_limit: Arrayable, weights: Arrayable) -> Tensor:
-        if self._reference_obs is None:
-            self._reference_obs = arr_to_th(obs, self._device)
 
+        obs = check_array(obs)
+        batch_size = obs.shape[0]
         action = arr_to_th(action, self._device).type_as(max_action)
         reward = arr_to_th(reward, self._device)
         weights = arr_to_th(check_array(weights), self._device)
         done = arr_to_th(check_array(done).astype('float'), self._device)
 
         v = self._val_function(obs).squeeze()
-        reference_v = self._val_function(self._reference_obs).squeeze()
-        mean_v = reference_v.mean()
-        next_v = (1 - done) * (
-            self._target_val_function(next_obs).squeeze() - self._dt * self._gamma ** (1 - self._dt) * mean_v) - \
-            done * self._gamma * mean_v / max(1 - self._gamma, 1e-5)
-
-        obs = check_array(obs)
-        batch_size = obs.shape[0]
-        advs = self.critic(
+        next_v = (1 - done) * self._target_val_function(next_obs).squeeze()
+        pre_advs = self.critic(
             np.concatenate([obs, obs], axis=0),
             torch.cat([action, max_action], dim=0))
-        adv, max_adv = advs[:batch_size], advs[batch_size:]
+        pre_adv, pre_max_adv = pre_advs[:batch_size], pre_advs[batch_size:]
+        adv = pre_adv - pre_max_adv
+        q = v + adv
+        # next_adv = 0 by definition
+        expected_q = (reward * self._dt + self._gamma ** self._dt * next_v).detach()
 
-        expected_v = (reward * self._dt + self._gamma ** self._dt * next_v).detach()
-        bell_residual = (expected_v - v) / self._dt - adv + max_adv
-
-        critic_loss = (bell_residual ** 2) + (max_adv ** 2)
+        critic_loss = (q - expected_q) ** 2
 
         self._val_optimizer.zero_grad()
         self._adv_optimizer.zero_grad()

### 2019-01-07 pendulum_exp/evaluation.py

@@ -2,9 +2,9 @@
 import numpy as np
 import matplotlib.pyplot as plt
 from abstract import Policy, Env
-from envs.vecenv import SubprocVecEnv
 from envs.hill import HillEnv
 from envs.pusher import AbstractPusher, ContinuousPusherEnv
+from critics import MixtureAdvantageCritic
 from convert import th_to_arr
 from gym.envs.classic_control import PendulumEnv
 from gym.spaces import Box
@@ -64,9 +64,9 @@ def specific_evaluation(
         critics = policy._critic.critic(state_space, actions).squeeze()
 
         plt.clf()
-        plt.subplot(1, 3, 1)
+        plt.subplot(1, 4, 1)
         plt.plot(state_space, th_to_arr(critics))
-        plt.subplot(1, 3, 2)
+        plt.subplot(1, 4, 2)
         plt.plot(state_space, th_to_arr(actions))
         if isinstance(env.envs[0].unwrapped.action_space, Box): # type: ignore
             action_space = np.linspace(-1, 1, nb_pixels)[:, np.newaxis]
@@ -74,6 +74,10 @@ def specific_evaluation(
             states = states[..., np.newaxis]
             actions = actions[..., np.newaxis]
             critics = policy._critic.critic(states, actions).squeeze()
-            plt.subplot(1, 3, 3)
+            if isinstance(policy._critic, MixtureAdvantageCritic):
+                _, adv_logpi, _ = policy._critic.compute_mixture_advantage(states, actions)
+                plt.subplot(1, 4, 4)
+                plt.imshow(th_to_arr(adv_logpi[..., 0]))
+            plt.subplot(1, 4, 3)
             plt.imshow(th_to_arr(critics))
         plt.pause(.1)

### 2019-01-07 paper/icml_drau/dauc.tex
Adding missing file

%! TEX root = icml_drau.tex
\begin{algorithmic}
	\STATE \textbf{Inputs:}
	\STATE $\theta_1$ and $\theta_2$, parameters of
	$V_{\theta_1}$ and $\bar{A}_{\theta_2}$.
	\STATE $\theta_1'$ parameter of target $V_{\theta_1'}$.
	\STATE $\pi_{\text{exp}}$ exploration policy.
	\STATE \textbf{opt}$_1$, \textbf{opt}$_2$, $\eta_1$ and $\eta_2$ optimizers and learning rates.
	\STATE $\mathcal{D}$, buffer of transitions $(s, a, r, d, s')$.
	\STATE $\deltat$ and $\gamma$, time discretization and discount factor.
	\STATE $\tau$, target network update factor.
	\STATE \textbf{nb\_epochs} number of epochs.
	\STATE \textbf{nb\_steps}, number of steps per epoch.
	\STATE
	\STATE Observe initial state $s^0$
	\STATE $t \gets 0$
	\FOR {$e=0, \textbf{nb\_epochs}$}
	\FOR {$k=1, \textbf{nb\_steps}$}
	\STATE Sample $a_t$ according to $\pi_{\text{exp}}$.
	\STATE Perform $a_t$ and observe $(r_{t+1}, d_{t+1}, s_{t+1})$.
	\STATE Store $(s_t, a_t, r_{t+1}, d_{t+1}, s_{t+1})$ in $\mathcal{D}$.
	\STATE Update $\pi_{\text{exp}}$ with $(a_t, s_{t+1})$.
	\STATE $t \gets t + 1$
	\ENDFOR
	\FOR {$k=0, \text{nb\_learn}$}
	\STATE \text{Sample a batch of $N$ random transitions from $\mathcal{D}$}
	\STATE $Q^i \gets V_{\theta_1}(s^i) + \deltat\hspace{-.17em}\left(
	\bar{A}_{\theta_2}(s^i, a^i) - \max\limits_{a'}\bar{A}_{\theta_2}(s^i, a')\right)$
	\STATE $\tilde{Q^i} \gets r^i + (1 - d^i) \gamma^{\deltat} V_{\theta_1}(s'^i)$
	\STATE $\Delta \theta_1 \gets \sum\limits_{i=1}^N \partial_{\theta_1} V_{\theta_1}(s^i) (Q^i - \tilde{Q^i})$
	\STATE $\Delta \theta_2 \gets \sum\limits_{i=1}^N \partial_{\theta_2} A_{\theta_2}(s^i, a^i) (Q^i - \tilde{Q^i})$
	\STATE Update $\theta_1$ using \textbf{opt}$_1$ with $\Delta \theta_1$ and learning rate $\eta_1$.
	\STATE Update $\theta_2$ using \textbf{opt}$_2$ with $\Delta \theta_2$ and learning rate $\eta_2$.
	\STATE $\theta_1' \gets \tau \theta_1' + (1 - \tau) \theta_1$
	\ENDFOR
	\ENDFOR
\end{algorithmic}

### 2019-01-07 paper/icml_drau/icml_drau.tex
Some additions to the text

@@
 The exploration scheme used~\cite{ddpg}, which consists in adding an
-\emph{Orhnstein-Ullenbeck}~\cite{orn-ull} process to actions, is naturally time
+\emph{Ornstein-Uhlenbeck}~\cite{orn-uhl} process to actions, is naturally time
 invariant in the sense that it is first defined as a time continuous stochastic
 process, and that discretization of this process are used in practice. Indeed,
-a continuous \emph{Orhnstein-Ullenbeck} process is defined as
+a continuous Ornstein-Uhlenbeck process is defined as
 \begin{equation}
 	dx_t = - (x_t - \mu) \theta dt + \sigma \dbt
+	\label{eq:orn_uhl}
 \end{equation}
@@
 invariant, in the sense that resulting state trajectories do not converge to a
-well defined stochastic process as $\deltat \rightarrow 0$.
+well defined stochastic process as $\deltat \rightarrow 0$. As $\dt$ goes to $0$,
+for any $\varepsilon$ independant of $\dt$, the behavior of state trajectories get
+closer to that of a white noise.
 
-\subsection{Learning rates scalings}
-\label{subsec:lr}
+To obtain a temporally coherent exploration scheme in discrete action environment
+when using Q-learning we instead directly apply a discretization of a time continuous
+stochastic process to the output of the approximate Q-function, and select actions
+by maximizing the resulting perturbed Q-function. Typically, the approximate Q-function
+values are also perturbed using a discretized Ornstein-Uhlenbeck noise.
 
@@
 to the reparameterized Q-function and enforcing the identifiability of $A^\pi$ by
-making sure that $\max_a A^\pi(s, a) = 0$, either approximately or exactly for all
-states.
+making sure that $\max_a A^\pi(s, a) = 0$ for all states.
+
+Both of these algorithms are off policy, and work by trying to minimize the
+optimal bellman gap on the Q-function. Minimizing the optimal Bellman Gap in
+an off policy setup amounts to minimizing the quantity
 \begin{algorithm}[ht]
@@
 \end{algorithm}
+
+\subsection{Learning rates scalings}
+\label{subsec:lr}
+To provide invariance to change of time discretization, one also needs to make
+sure that learning is possible in finite macroscopic time \TODO{The difference between
+macroscopic and microscopic must be defined earlier}, and that learning
+will take approximately as much macroscopic time for a coarse discretization than
+for a fine grained discretization. These two requirements means that learning a
+meaningful approximation of the advantage function as defined in~\ref{eq:adv},
+i.e. an approximation that is correct at a precision independant of the time
+discretization, should be possible with a number of samples of order at most
+$\frac{1}{\deltat}$. If this requirement is not satisfied, as the
+discretization timestep decreases, the macroscopic time needed to acquire
+enough data samples to learn a meaningful advantage function increases, meaning
+that with a very fine grained discretization, it would take a nearly infinite
+time to learn something that is easily learnable in a coarse discretization.
+
+Let us provide an informal derivation, in an on policy policy evaluation setup,
+of the orders of magnitudes on the learning rates of both $A$ and $V$. An
+example problem showing that these orders of magnitudes are indeed best case
+scenarios if we want a general learning algorithm is given in the supplementary
+material.
+
+In an on policy policy evaluation setup, when $A$ and $V$ are learnt using
+approximate TD(0), the update rules on $\theta_1$ and $\theta_2$ for a given
+transition $(s, a, \delta r, s + \delta s)$ are
+\begin{align}
+	\theta_1 &\leftarrow \theta_1 + \eta (\delta r + \gamma^{\deltat} V(s + \delta s) - V(s)) \partial_{\theta_1} V(s)\\
+	\theta_2 &\leftarrow \theta_2 + \eta (\deltat A(s, a) - \left(\delta r + \gamma^{\deltat} V(s + \delta s) - V(s)\right)) \partial_{\theta_2} A(s)
+\end{align}
 

### 2019-01-10 paper/icml_drau/icml_drau.tex
Paper update

--- 2019-01-07 paper/icml_drau/icml_drau.tex
@@
 \label{subsec:algorithm}
-We introduce two versions of \emph{Deep Advantage Updating} to cover both the
-continuous and discrete action cases. Both relies on applying either DDPG or DQN
-to the reparameterized Q-function and enforcing the identifiability of $A^\pi$ by
-making sure that $\max_a A^\pi(s, a) = 0$, either approximately or exactly for all
-states.
+Two variants of \emph{Deep Advantage Updating} are derived to cover both the continuous
+and discrete action cases. They both rely on variants of Q-learning. More precisely,
+we learn $V$ and $A$ using approximate version of the following bellman equation
+\begin{align}
+	V^{\pi}(s) + A^{\pi}(s, a) &= \E_{\delta r \sim \delta r(s, a)}\delta r\nonumber\\ 
+				   &+ \gamma^{\deltat}  \E_{\delta s \sim \delta s(s, a)} V^{\pi}(s + \delta s)\label{eq:bellman_A}\\
+	A^{\pi}(s, \pi(s)) &= 0\label{eq:max_A}
+\end{align}
+for a greedy exploitation policy $\pi(s) = \text{argmax}_{a'}A^\pi(s, a')$.
+Eq.~\eqref{eq:max_A} is directly verified owing to the parameterization of $A$.
+To approximately verify~\eqref{eq:bellman_A}, the corresponding squared residual
+is minimized by an approximate gradient descent.
+The corresponding update equations when learning from incoming transitions of
+an exploratory trajectory are
+\begin{align}
+	\delta Q_t &= A(s_t, a_t) - \frac{\delta r_t + \gamma^{\deltat} V(s_{t+\deltat}) - V(s_t)}{\deltat}\\
+	\theta^1_{t+\deltat} &= \theta^1_t + \eta_1 \partial_{\theta^1} V(s_t) \delta Q_t\\
+	\theta^2_{t+\deltat} &= \theta^2_t + \eta_2 \partial_{\theta^2} A(s_t, a_t) \delta Q_t.
+	\label{eq:approx_bellman_A}
+\end{align}
+where $\eta$'s are learning rates. We are now left to derive appropriate learning rates
+to provide invariance to time discretization.
 \begin{algorithm}[ht]
@@
 
+\subsection{Learning rates scalings}
+\label{subsec:lr}
+A necessary condition to guarantee invariance to time discretization is that
+the trajectory of the parameters of our function approximators should remain
+well behaved when the discretization parameter $\deltat$ decreases. This notably
+means that optimization steps should not be too large, in which case parameter
+trajectories could diverge in a single gradient step, or two small, in which case
+the trajectories would converge to single points as $\deltat$ goes to $0$.
+This in turn imposes conditions on the scalings of the learning rates of invariant
+algorithms with respect to the discretization timestep.
+
+If we optimize $\theta_1$ and $\theta_2$ according to Eq.~\eqref{eq:approx_bellman_A},
+both $\theta_1$ and $\theta_2$ verify update equations of the form
+\begin{align}
+	\
+	\delta \theta^1_t = \eta_1 \partial_{\theta_1} V(s_t) (
+\end{align}
+To have parameter steps that do not grow larger or shrink to $0$ as $\deltat
+\leftarrow 0$, both $\delta \theta^1$ and $\delta \theta^2$ should verify
+stochastic differential equations, and notably have a deterministic component
+of order $\deltat$. If the deterministic component 
 \section{Experiments}

### 2019-01-10 pendulum_exp/dataanalysis/dataloader.py
Paper update

@@ -1,141 +1,131 @@
-import os 
+import os
 import pickle as pkl
 from datetime import datetime
 
 class ExperimentData:
-	def __init__(self, experimentrunlist):
-		# experimentrunlist = [ExperimentRun(args_file, logs_file) for args_file, logs_file in argslogs_filelist]
+    def __init__(self, experimentrunlist):
+        # experimentrunlist = [ExperimentRun(args_file, logs_file) for args_file, logs_file in argslogs_filelist]
 
-		setting_list = []
-		for run in experimentrunlist:
-			args = run.args
-			for setting in setting_list:
-				if setting.args == args:
-					setting.push(run)
-					break
-			setting_list.append(ExperimentSetting([run]))
+        setting_list = []
+        for run in experimentrunlist:
+            args = run.args
+            for setting in setting_list:
+                if setting.args == args:
+                    # setting.push(run) do nothing for now
+                    break
+            setting_list.append(ExperimentSetting([run]))
 
-		self._setting_list = setting_list
+        self._setting_list = setting_list
 
+    @property
+    def deltakeys(self):
+        if not hasattr(self, "_deltaargs"):
+            args_list = [setting.args for setting in self._setting_list]
 
-	@property
-	def deltakeys(self):
-		if not hasattr(self, "_deltaargs"):
-			args_list = [setting.args for setting in self._setting_list]
+            expargsdict = {}
+            for args in args_list:
+                for k, v in args.items():
+                    if k not in expargsdict:
+                        expargsdict[k] = set()
+                    expargsdict[k].add(v)
 
-			expargsdict = {}
-			for args in args_list:
-				for k, v in args.items():
-					if k not in expargsdict:
-						expargsdict[k] = set()
-					expargsdict[k].add(v)
+            self._deltaargs = [k for k, vset in expargsdict.items() if len(vset) > 1]
+            self._sharedargs = dict((k, vset.pop()) for k, vset in expargsdict.items() if len(vset) == 1)
 
-			self._deltaargs = [k for k, vset in expargsdict.items() if len(vset) > 1]
-			self._sharedargs = dict((k, vset.pop()) for k, vset in expargsdict.items() if len(vset) == 1)
+        return self._deltaargs
 
-		return self._deltaargs
+    @property
+    def sharedargs(self):
[…]

### 2019-01-10 pendulum_exp/dataanalysis/plots.py

@@ -1,19 +1,22 @@
+import argparse
 import matplotlib
 # matplotlib.use('Agg')
 import matplotlib.pyplot as plt
-import numpy as np 
+import numpy as np
 from scipy.ndimage.filters import gaussian_filter1d
+import getpass
 
 
 from dataloader import loader_leonard, ExperimentData
-from datetime import datetime
-
-algolabeldict = {
-    'discrete_value':"qlearning",
-    'discrete_advantage':"advup"
-}
 
 def plot_learning_curves(expdata, key_list, namefile):
+    algolabeldict = {
+        'discrete_value': "qlearning",
+        'discrete_advantage': "advup",
+        'approximate_value': "qlearning",
+        'approximate_advantage': "advup"
+    }
+
     mint, maxt = 0, 40
 
     nlines, ncol = len(key_list), 1
@@ -28,27 +31,27 @@ def plot_learning_curves(expdata, key_list, namefile):
     for key, ax in zip(key_list, axes.flat):
         ax.set_title(key)
 
-        for setting in sorted(expdata._setting_list, key=lambda s:s.args['dt']):
+        for setting in sorted(expdata._setting_list, key=lambda s: s.args['dt']):
             args = setting.args
             dt = args['dt']
             algo = args["algo"]
             label = f"{algolabeldict[algo]}; dt={dt:.1e}"
-            if algo == 'discrete_value':
+            if 'value' in algo:
                 linestyle = '--'
-                c=cm.to_rgba(np.log(dt))
-                linewidth=1.
-                alpha=None
-            elif algo == 'discrete_advantage':
+                c = cm.to_rgba(np.log(dt))
+                linewidth = 1.
+                alpha = None
+            elif 'advantage' in algo:
                 linestyle = '-'
-                c=cm.to_rgba(np.log(dt))
-                linewidth=1.
+                c = cm.to_rgba(np.log(dt))
+                linewidth = 1.
                 alpha = None
             timeseq = setting.timeseq(key)
             if timeseq is None:
                 continue
 
-            xdata = np.array([i for (i,v) in timeseq.items()])
-            ydata = np.array([v for (i,v) in timeseq.items()])
+            xdata = np.array([i for (i, v) in timeseq.items()])
+            ydata = np.array([v for (i, v) in timeseq.items()])
 
[…]

### 2019-01-10 pendulum_exp/envs/utils.py
add support for mujoco with various dt

@@ -1,6 +1,7 @@
 """ Environment utilities """
 from abc import ABC, abstractmethod
 import gym
+
 from envs.wrappers import TimeLimit
 import numpy as np
 from envs.pusher import DiscretePusherEnv, ContinuousPusherEnv
@@ -190,16 +191,18 @@ def make_env(env_id: str, dt: float, time_limit: float): # noqa: C901
     elif env_id == 'bipedal_hardcore':
         env = WalkerHardcore(dt)
     elif env_id == 'half_cheetah':
-        # from gym.envs.mujoco import HalfCheetahEnv
-        env = gym.make('HalfCheetah-v2').unwrapped
-        dt = env.dt
-        info('The timestep dt is {}'.format(dt))
+        from gym.envs.mujoco import HalfCheetahEnv
+        # env = gym.make('HalfCheetah-v2').unwrapped
+        env = HalfCheetahEnv(dt)
+        assert dt == env.dt
         # env.model.opt.timestep = dt / env.frame_skip
     elif env_id == 'ant':
         # from gym.envs.mujoco import HalfCheetahEnv
-        env = gym.make('Ant-v2').unwrapped
-        dt = env.dt
-        info('The timestep dt is {}'.format(dt))
+        from gym.envs.mujoco.ant import AntEnv
+
+        env = AntEnv(dt)\
+        # env = gym.make('Ant-v2').unwrapped
+        assert env.dt == dt
         # env.model.opt.timestep = dt / env.frame_skip
     elif env_id == 'dart_cheetah':
         env = gym.make('DartHalfCheetah-v1').unwrapped

### 2019-01-10 pendulum_exp/dataanalysis/dataloader.py
.

@@ -3,155 +3,155 @@ import pickle as pkl
 from datetime import datetime
 
 class ExperimentData:
-	def __init__(self, experimentrunlist):
-		# experimentrunlist = [ExperimentRun(args_file, logs_file) for args_file, logs_file in argslogs_filelist]
+    def __init__(self, experimentrunlist):
+        # experimentrunlist = [ExperimentRun(args_file, logs_file) for args_file, logs_file in argslogs_filelist]
 
-		setting_list = []
-		for run in experimentrunlist:
-			args = run.args
-			for setting in setting_list:
-				if setting.args == args:
-					setting.push(run)
-					break
-			setting_list.append(ExperimentSetting([run]))
+        setting_list = []
+        for run in experimentrunlist:
+            args = run.args
+            for setting in setting_list:
+                if setting.args == args:
+                    setting.push(run)
+                    break
+            setting_list.append(ExperimentSetting([run]))
 
-		self._setting_list = setting_list
-
-
-	@property
-	def deltakeys(self):
-		if not hasattr(self, "_deltaargs"):
-			args_list = [setting.args for setting in self._setting_list]
-
-			expargsdict = {}
-			for args in args_list:
-				for k, v in args.items():
-					if k not in expargsdict:
-						expargsdict[k] = set()
-					expargsdict[k].add(v)
-
-			self._deltaargs = [k for k, vset in expargsdict.items() if len(vset) > 1]
-			self._sharedargs = dict((k, vset.pop()) for k, vset in expargsdict.items() if len(vset) == 1)
-
-		return self._deltaargs
-
-	@property
-	def sharedargs(self):
-		if not hasattr(self, "_sharedargs"):
-			self.deltakeys
-		return self._sharedargs
-	
+        self._setting_list = setting_list
 
 
-	def deltaitems(self):
-		if not hasattr(self, "_deltaitems"):
-			self._deltaitems = [dict((k, setting.args[k]) for k in self.deltakeys) for setting in self._setting_list]
-		return self._deltaitems
-	
-	def filter_settings(self, bool_function):
-		run_list = []
-		for setting in self._setting_list:
-			if bool_function(setting.args):
-				run_list.extend(setting._experimentrunlist)
-		return ExperimentData(run_list)
+    @property
+    def deltakeys(self):
+        if not hasattr(self, "_deltaargs"):
[…]

### 2019-01-10 pendulum_exp/dataanalysis/plots.py

@@ -10,11 +10,13 @@ from datetime import datetime
 
 algolabeldict = {
     'discrete_value':"qlearning",
-    'discrete_advantage':"advup"
+    'discrete_advantage':"advup",
+    'approximate_advantage':'advup',
+    'approximate_value':'ddpg',
 }
 
 def plot_learning_curves(expdata, key_list, namefile):
-    mint, maxt = 0, 40
+    mint, maxt = 0, 100
 
     nlines, ncol = len(key_list), 1
     fig, axes = plt.subplots(nlines, ncol, figsize=(5.*ncol, 4.*nlines))
@@ -32,17 +34,24 @@ def plot_learning_curves(expdata, key_list, namefile):
             args = setting.args
             dt = args['dt']
             algo = args["algo"]
-            label = f"{algolabeldict[algo]}; dt={dt:.1e}"
-            if algo == 'discrete_value':
+            label = f"{algolabeldict[algo]}; dt={dt:.0e}"
+            if algo == 'discrete_value' or algo == 'approximate_value':
                 linestyle = '--'
                 c=cm.to_rgba(np.log(dt))
                 linewidth=1.
                 alpha=None
-            elif algo == 'discrete_advantage':
+                if 'tau' in expdata.deltakeys:
+                    tau = setting.args['tau']
+                    label += f';tau{tau}'
+                    if tau != 0:
+                        linestyle = ':'
+            elif algo == 'discrete_advantage' or algo == 'approximate_advantage':
                 linestyle = '-'
                 c=cm.to_rgba(np.log(dt))
                 linewidth=1.
                 alpha = None
+            else:
+                raise ValueError
             timeseq = setting.timeseq(key)
             if timeseq is None:
                 continue
@@ -50,11 +59,12 @@ def plot_learning_curves(expdata, key_list, namefile):
             xdata = np.array([i for (i,v) in timeseq.items()])
             ydata = np.array([v for (i,v) in timeseq.items()])
 
+            kernelsize = max(maxt/100, 0.3) / dt
             x = np.arange(max(xdata)+1)
             yinterp = np.interp(x, xdata, ydata)
[…]

### 2019-01-10 pendulum_exp/dataanalysis/plots.py
separate script from plots

@@ -1,24 +1,21 @@
-import argparse
 import matplotlib
 # matplotlib.use('Agg')
 import matplotlib.pyplot as plt
-import numpy as np
+import numpy as np 
 from scipy.ndimage.filters import gaussian_filter1d
-import getpass
 
 
 from dataloader import loader_leonard, ExperimentData
 from datetime import datetime
 
-
-def plot_learning_curves(expdata, key_list, namefile):
-    algolabeldict = {
+algolabeldict = {
     'discrete_value':"qlearning",
     'discrete_advantage':"advup",
     'approximate_advantage':'advup',
     'approximate_value':'ddpg',
+}
 
-
+def plot_learning_curves(expdata, key_list, namefile):
     mint, maxt = 0, 100
 
     nlines, ncol = len(key_list), 1
@@ -33,12 +30,12 @@ def plot_learning_curves(expdata, key_list, namefile):
     for key, ax in zip(key_list, axes.flat):
         ax.set_title(key)
 
-        for setting in sorted(expdata._setting_list, key=lambda s: s.args['dt']):
+        for setting in sorted(expdata._setting_list, key=lambda s:s.args['dt']):
             args = setting.args
             dt = args['dt']
             algo = args["algo"]
             label = f"{algolabeldict[algo]}; dt={dt:.0e}"
-            if 'value' in algo:
+            if algo == 'discrete_value' or algo == 'approximate_value':
                 linestyle = '--'
                 c=cm.to_rgba(np.log(dt))
                 linewidth=1.
@@ -48,10 +45,10 @@ def plot_learning_curves(expdata, key_list, namefile):
                     label += f';tau{tau}'
                     if tau != 0:
                         linestyle = ':'
-            elif 'advantage' in algo:
+            elif algo == 'discrete_advantage' or algo == 'approximate_advantage':
                 linestyle = '-'
-                c = cm.to_rgba(np.log(dt))
-                linewidth = 1.
+                c=cm.to_rgba(np.log(dt))
+                linewidth=1.
                 alpha = None
             else:
                 raise ValueError
@@ -59,8 +56,8 @@ def plot_learning_curves(expdata, key_list, namefile):
             if timeseq is None:
                 continue
 
[…]

### 2019-01-10 pendulum_exp/dataanalysis/script_leonard.py

@@ -1,0 +1,28 @@
+from datetime import datetime
+
+
+
+from dataloader import loader_leonard, ExperimentData
+
+from plots import plot_learning_curves
+
+start_date = datetime.strptime('2019_01_09_05_44_37', "%Y_%m_%d_%H_%M_%S")
+# stop_date = datetime.strptime('2019_01_08_03_13_35', "%Y_%m_%d_%H_%M_%S")
+# start_date = 'last'
+stop_date = None
+runlist = loader_leonard('/private/home/leonardb/workdir', 'mujoco_continuous', 
+    start_date=start_date, stop_date=stop_date)
+
+expdata = ExperimentData(runlist)
+
+print(expdata.deltakeys)
+
+expdata_ant = expdata.filter_settings(lambda s: s['env_id'] == 'ant')
+plot_learning_curves(expdata_ant, ['Return', 'Return'], 'ant')
+
+expdata_cartpole = expdata.filter_settings(lambda s: s['env_id'] == 'cartpole')
+plot_learning_curves(expdata_cartpole, ['Return', 'Return'], 'cartpole')
+
+
+# expdata_tau = expdata.filter_settings(lambda s: s['algo'] == 'discrete_value')
+# expdata_tau.repr_rawlogs("Return", 5)

### 2019-01-10 paper/icml_drau/icml_drau.tex
split the files

@@
 
-\section{Continuous and near continuous time reinforcement learning}
+\section{Near continuous time reinforcement learning}
 \label{sec:continous}
-In this section, a continuous reinforcement learning framework is introduced,
-and it is shown that under this framework, the state-action value function does
-not depend on the action. We further show that in near continous domains, when
-the discretization timestep $\deltat$ is small, the difference in state-action
-value between two different actions in a fixed state is of order $\bigO(\deltat)$.
+% In this section, a continuous reinforcement learning framework is introduced,
+% and it is shown that under this framework, the state-action value function does
+% not depend on the action. We further show that in near continous domains, when
+% the discretization timestep $\deltat$ is small, the difference in state-action
+% value between two different actions in a fixed state is of order $\bigO(\deltat)$.
+TODO. Main ideas: near continuous time environments (control, mujoco, video games, ...)
+There is a time discretization $\deltat$ given by the problem.
+The uderlying process does not depend of $\deltat$.
+Need to see what is happening when changing $\deltat$.
+The problem is invariant to $\deltat$.
+The quantities we are looking at should be invariant to.
+The learning procedure should be invariant.
+Looking at the limits when $\deltat \rightarrow 0$.
+In simulated environments, changing $\deltat$ only means to change the FPS.
+In real environments, for example with robots, this could mean changing the quality of the sensors and motors.
 
 \subsection{Framework}
+
+Let assume that we have a Markov Decision Process (TODO:Cite?) $\langle {\cal S}, {\cal A}, T_{\deltat}, r, \gamma\rangle$.
+${\cal S}$ is the state-space, which is supposed to be continuous (TODO:not the good formulation).
+The action space is TODO.
+
+We assume that the transition is continuous.
+This means that ...
+TODO.
+Discussion on the hypothesis, in which environments it is true.
+Is the condition $\forall \deltat, \forall s, \E[\|s_{k+1} - s_k \||s_k = s] \leq ?$.
+
 In what follows, $\dt$ denotes a truly infinitesimal time interval while $\deltat$

### 2019-01-10 paper/icml_drau/advup.tex
split texte between files

--- 2019-01-10 paper/icml_drau/icml_drau.tex
@@
-%%%%%%%% ICML 2018 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
-
-\documentclass{article}
-
-% Recommended, but optional, packages for figures and better typesetting:
-\usepackage{amsmath, amssymb}
-\usepackage{mathtools}
-\usepackage{microtype}
-\usepackage{graphicx}
-\usepackage{subfigure}
-\usepackage{booktabs} % for professional tables
-\usepackage{xcolor}
-\usepackage{algorithmic}
-
-% hyperref makes hyperlinks in the resulting PDF.
-% If your build breaks (sometimes temporarily if a hyperlink spans a page)
-% please comment out the following usepackage line and replace
-% \usepackage{icml2018} with \usepackage[nohyperref]{icml2018} above.
-\usepackage{hyperref}
-
-\input{macros.tex}
-
-% Use the following line for the initial blind version submitted for review:
-\usepackage{icml2018}
-
-% If accepted, instead use the following line for the camera-ready submission:
-%\usepackage[accepted]{icml2018}
-
-% The \icmltitle you define below is probably too long as a header.
-% Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Deep Relativistic Advantage Updating}
-
-\begin{document}
-
-\twocolumn[
-\icmltitle{There are no small Advantages:
-	Reinforcement learning in near continuous time
-	with Deep RelAtive Mixture of Advantages
-}
-
-% It is OKAY to include author information, even for blind
-% submissions: the style file will automatically remove it for you
-% unless you've provided the [accepted] option to the icml2018
-% package.
-
-% List of affiliations: The first argument should be a (short)
-% identifier you will use later to specify author affiliations
-% Academic affiliations should list Department, University, City, Region, Country
-% Industry affiliations should list Company, City, Region, Country
-
-% You can specify symbols, otherwise they are numbered in order.
-% Ideally, you should not use this facility. Affiliations will be numbered
-% in order of appearance and this is the preferred way.
-\icmlsetsymbol{equal}{*}
-
-\begin{icmlauthorlist}
-%\icmlauthor{Aeiau Zzzz}{equal,to}
-%\icmlauthor{Bauiu C.~Yyyy}{equal,to,goo}
-%\icmlauthor{Cieua Vvvvv}{goo}
-%\icmlauthor{Iaesut Saoeu}{ed}
-%\icmlauthor{Fiuea Rrrr}{to}
-%\icmlauthor{Tateu H.~Yasehe}{ed,to,goo}
-%\icmlauthor{Aaoeu Iasoh}{goo}
-%\icmlauthor{Buiui Eueu}{ed}
-%\icmlauthor{Aeuia Zzzz}{ed}
-%\icmlauthor{Bieea C.~Yyyy}{to,goo}
-%\icmlauthor{Teoau Xxxx}{ed}
-%\icmlauthor{Eee Pppp}{ed}
-\end{icmlauthorlist}
-
-%\icmlaffiliation{to}{Department of Computation, University of Torontoland, Torontoland, Canada}
-%\icmlaffiliation{goo}{Googol ShallowMind, New London, Michigan, USA}
-%\icmlaffiliation{ed}{School of Computation, University of Edenborrow, Edenborrow, United Kingdom}
-%
-%\icmlcorrespondingauthor{Cieua Vvvvv}{c.vvvvv@googol.com}
-%\icmlcorrespondingauthor{Eee Pppp}{ep@eden.co.uk}
-
-% You may provide any keywords that you
-% find helpful for describing your paper; these are used to populate
-% the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{Reinforcement learning}
-
-\vskip 0.3in
-]
-
-% this must go after the closing bracket ] following \twocolumn[
-
-% This command actually creates the footnote in the first column
-% listing the affiliations and the copyright notice.
-% The command takes one argument, which is text to display at the start of the footnote.
-% The \icmlEqualContribution command is standard text for equal contribution.
-% Remove it (just {}) if you do not need this facility.
-
-%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
-%\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
-
-\begin{abstract}
-	Despite remarkable successes, \emph{Deep Reinforcement learning} remains sensitive to
-	hyperparameterization, implementation details and small environment changes~(\citealt{drl_matter}, \citealt{drl_matter_bis}). 
-	Overcoming this sensitivity is key to making DRL applicable to real world problems.
-	In this paper, we study the sensitivity of Deep Reinforcement learning algorithms to
-	time discretization. More precisely, we show that approaches based on estimations of the
-	Q-function, e.g. \emph{Deep Q-learning}~\cite{dqn} and \emph{Deep deterministic policy gradient}~\cite{ddpg}
-	are sensitive to variations of time discretization. We further show that a simple reparameterization of the
-	Q-function as a sum of a state value term and a \emph{small} action dependent advantage term yields an algorithm
-	much more resilient to variations of time discretization.
-\end{abstract}
-
-\section{Introduction}
-\label{sec:intro}
-In recent years, deep learning based approaches to reinforcement learning have
-thrived and provided impressive results in a variety of domains, achieving superhuman
-performances with no expert knowledge in perfect information zero-sum
-games~\cite{alphazero}, reaching level of top players in video
-games~(\citealt{openai_five}, \citealt{dqn}), or learning dexterous manipulation
-from scratch without demonstrations~\cite{hand_control}.
-
-In spite of those successes, \emph{Deep reinforcement learning} approaches are
-sensitive to a number of factors, including hyperparameterization,
-implementation details or small changes in the environment
-parameters~(\citealt{drl_matter}, \citealt{drl_matter_bis}). This sensitivity,
-along with sample inefficiency, prevents DRL from being applied in most real
-world settings. High sensitivity to environment parameters notably prevents
-transfering abilities from imperfect simulators to real world scenarios.
-
-In this paper we focus our attention on the sensitivity to time discretization
-of DRL approaches, i.e. what happens when an agent receives $50$ observations
-and is expected to take $50$ actions per second instead of $10$. One expects
-that decreasing time discretization, or equivalently shortening reaction time,
-should only improve agent performances. This is not what happens in practice
-for approaches based on estimation of state action value functions, e.g.
-\emph{Deep Q-learning} (DQN~\citep{dqn} and \emph{Deep deterministic policy
-gradient} (DDPG~\citep{ddpg}). This is shown experimentally in Sec.~\ref{sec:exp}.
-
-We relate this sensitivity to the fact that, as the discretization timestep
-decreases, the effect of individual actions on the total return decreases, and
-can vanish when the discretization timestep becomes infinitesimal. The natural
-framework to study this phenomenon is Continuous time reinforcement learning 
-(Sec.~\ref{sec:continous}).
-
-To cope for this effect, we introduce a
-parameterization of the Q-function as a sum of a state value term and a
-\emph{small} action dependent advantage term (Sec.~\ref{subsec:reparam}). This
-reparameterized Q approximation can be trained with deep variants of Q-learning
-(Sec.~\ref{subsec:algorithm}).  To properly scale w.r.t. time discretization,
-particular attention has to be taken regarding both the exploration method, and
-the scaling of learning rates (Sec.~\ref{subsec:lr}, Sec.~\ref{subsec:explo}).
-The resulting algorithm is shown to provide near perfect invariance to time
-discretization on simple environments, and much better invariance properties
-than vanilla DQN or DDPG on more complex environments (Sec.~\ref{sec:exp}).
-
-\section{Related work}
-\label{sec:related}
-Continuous and near continuous reinforcement learning, which is at the core of
-our approach has been thoroughly studied in the deterministic case
-in~\cite{adv_upd} or \cite{cont_rl}, but has not been applied in the context of
-deep reinforcement learning.
-
-The approach presented in this paper is closely related to \cite{adv_upd} and
-\cite{cont_rl}. \cite{adv_upd} already separated the learning procedure of the
-value and advantage functions, and introduced proper scalings between the two
-components. We extend on \cite{adv_upd} in three directions. We provide a
-formal argument on why the advantage contribution of the Q-function is of order
-$\bigO(\deltat)$, irrelevant of the stochasticity of the environment, in a quite
-general framework. We show that, provided with well suited learning rates and
-exploration method, Advantage updating provide near invariance to change of time
-discretization. We show that Advantage updating is viable in the context of deep
-reinforcement learning, while not using the true gradient of the residual error.
-
-More recently, \cite{dueling_nets} also introduced a parameterization separating
-the value and the advantage components of the Q-function. We additionally
-introduce a natural scaling coeficient between the two, that strongly improves
-the time discretization invariance properties.
-
-\section{Continuous and near continuous time reinforcement learning}
-\label{sec:continous}
-In this section, a continuous reinforcement learning framework is introduced,
-and it is shown that under this framework, the state-action value function does
-not depend on the action. We further show that in near continous domains, when
-the discretization timestep $\deltat$ is small, the difference in state-action
-value between two different actions in a fixed state is of order $\bigO(\deltat)$.
-
-\subsection{Framework}
-In what follows, $\dt$ denotes a truly infinitesimal time interval while $\deltat$
-denotes a small, but non infinitesimal time interval. Similarily, $\dbt$ denotes
-an infinitesimal (multi-dimensional) brownian step, while $\deltabt$ denotes its
-discretized equivalent. Informally, $\dbt$ is to be thought of as $\gauss(0, \sqrt{\dt}^2)$,
-and $\deltabt$ as $\gauss(0, \sqrt{\deltat}^2)$, where $\gauss(\mu, \sigma)$ denotes a gaussian
-random variable of mean $\mu$ and standard deviation (or covariance matrix in the multi-dimensional case)
-$\sigma$.
-
-Consider a fully observable \emph{Markov Decision Process} defined by the following
-transitions and rewards
-\begin{align}
-	ds(s, a) &= F(s, a) dt + \Sigma(s, a) \dbt\\
-	dr(s, a) &= \reward(s, a) dt + \Sigma_r(s, a) \dbt.
-\end{align}
-The equation on $s$ is that of a diffusion process, with drift coeficient $F$
-and diffusion matrix $\Sigma$.  This framework is only moderately restrictive.
-For example, deterministic continuous control environments fall in this
-framework, with $\Sigma = 0$ and $\Sigma_r = 0$. Limitations of the framework
-are discussed in Sec.~\ref{sec:limitations}.
-This framework is naturally discretized to arbitrary $\deltat$ by replacing all
-$d$'s with $\delta$'s.
-
-A continuous policy $\pi$ is defined as a deterministic mapping from states to
-actions.  In continuous time, defining stochastic policies with well defined
-state-value functions requires incorporating information from previous action into
-the state to obtain a temporally coherent policy. To avoid such inconvenience, the
-framework is restricted to deterministic policies. This is a mild constraint, as this
-only enforces determinism of the exploitation policy and not of exploration policies.
-In what follows, $\tau\sim\pi$ denotes a trajectory of states and actions sampled
-according to policy $\pi$, with $s_0$ sampled according to an arbitrary initial distribution
-on states $\rho_0$, and $dr_{t} = dr(s_t, a_t)$.
-
-Under a deterministic policy $\pi$, one can define the $\gamma$ discounted
-\emph{state value function} $V^\pi(s)$ and \emph{state action value function}
-$Q^\pi(s, a)$ as
-\begin{align}
-	V^\pi(s) &= \E_{\tau\sim\pi}\left[
-		\int\limits_{t=0}^\infty \gamma^{t}
-		dr_{t} \mid s_0 = s
-	\right]\\
-	Q^\pi(s, a) &= \E_{\tau\sim\pi}\left[
-		\int\limits_{t=0}^\infty \gamma^{t}
-		dr_{t} \mid s_0 = s, a_0 = a
-	\right]
-\end{align}
-which are naturally discretized as
-\begin{align}
-	V^\pi(s) &= \E_{\tau\sim\pi}\left[
-		\sum\limits_{k=0}^\infty \gamma^{k\deltat}
-		\delta r_{k\deltat} \mid s_0 = s
-	\right]\\
-	Q^\pi(s, a) &= \E_{\tau\sim\pi}\left[
-		\sum\limits_{k=0}^\infty \gamma^{k\deltat}
-		\delta r_{k\deltat} \mid s_0 = s, a_0 = a
-	\right].
-\end{align}
-Both $V^\pi$ and $Q^\pi$ verify Bellman equations, informally
-\begin{align}
-	V^\pi(s) &= \reward(s, \pi(s))\dt + \gamma^{\dt} \E_{ds \sim ds(s, \pi(s))} V^\pi(s + ds)\nonumber\\
-	Q^\pi(s, a) &= \reward(s, a)\dt + \gamma^{\dt} \E_{ds(s, a)} V^\pi(s + ds(s, a))
-	\label{eq:cont_bell}
-\end{align}
-which in continuous time translates, using Ito's lemma, into a Hamilton Jacobi equation
-\begin{align}
-	\reward + \partial_s V^\pi \cdot f + \frac{1}{2} \text{tr}\left(\Sigma^T\partial^2_{s^2} V^\pi\Sigma\right) = (1 - \gamma) V^\pi(s).
-\end{align}
-
-\subsection{Independence of $Q^\pi$ on $a$ in continuous time}
-Consider a discretization of Eq.~\eqref{eq:cont_bell} with time interval
-$\deltat \ll~1$, Ito's lemma yield
-\begin{align}
-	Q^\pi(s, a) &= \reward(s, a) \deltat + \gamma^{\deltat} \E_{\delta s \sim \delta s(s, a)}V^\pi(s + \delta s)\nonumber\\
-		    &= \reward(s, a) \deltat + \gamma^{\deltat} V^\pi(s) +
-		     \gamma^{\deltat} \partial_s V^\pi(s)f(s, a) \deltat\nonumber\\
-		    &+ \frac{1}{2}\text{tr}\left(\Sigma^T\partial^2_{s^2}V^\pi(s)\Sigma\right)\deltat + \bigO(\deltat^2)\nonumber\\
-		    &\xrightarrow[\deltat \to 0]{} V^\pi(s).
-		    \label{eq:Q_cont}
-\end{align}
-A consequence of Eq.~\eqref{eq:Q_cont} is that in exact continuous time,
-$Q^\pi$ does not bear any information on the hierarchy of actions, and
-thus cannot be used to select actions that yields higher future returns.
-
-As the difference between between $Q^\pi(s, a)$ and $V^\pi(s)$ is of order
-$\bigO(\deltat)$, it is natural to define a rescaled version of the advantage
-function, namely
-\begin{align}
-	A^\pi(s, a) &= \frac{Q^\pi(s,a) - V^\pi(s)}{\deltat}\\
-		    &= \reward(s, a) +
-		    \frac{\gamma^{\deltat} \E_{\delta s \sim \delta s(s, a)} V^\pi(s + \delta s) - V^\pi(s)}{\deltat}.
-    \label{eq:adv}
-\end{align}
-Contrary to the action state value function, the rescaled advantage function converges
-to an action dependent quantity when $\deltat$ goes to $0$
-\begin{align}
-	A^\pi(s, a) &\xrightarrow[\deltat \to 0]{} \reward(s, a) +
-	(\gamma - 1) V^\pi(s) + \partial_s V^\pi(s) f(s, a)\nonumber\\
-		    &+
-	\frac{\text{tr}\left(\Sigma^T(s, a)\partial^2_{s^2}V^\pi(s) \Sigma(s,a)\right)}{2},
-\end{align}
-and $Q^\pi$ can then be rewritten as
-\begin{equation}
-	Q^\pi(s, a) = V^\pi(s) + \deltat A^\pi(s, a).
-	\label{eq:reparam_q_pi}
-\end{equation}
-
 \section{Deep Advantage Updating}
@@
 of order $\deltat$. If the deterministic component
-\section{Experiments}
-\label{sec:exp}
-
-\section{Limitations}
-\label{sec:limitations}
-\section{Conclusion}
-\label{sec:concl}
-
-\bibliography{icml_drau.bib}
-\bibliographystyle{icml2018}
-\end{document}
-
-% This document was modified from the file originally made available by
-% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
-% by Iain Murray in 2018. It was modified from a version from Dan Roy in
-% 2017, which was based on a version from Lise Getoor and Tobias
-% Scheffer, which was slightly modified from the 2010 version by
-% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
-% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
-% slightly modified from Prasad Tadepalli's 2007 version which is a
-% lightly changed version of the previous year's version by Andrew
-% Moore, which was in turn edited from those of Kristian Kersting and
-% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2019-01-10 paper/icml_drau/framework.tex
split texte between files

--- 2019-01-10 paper/icml_drau/icml_drau.tex
@@
-%%%%%%%% ICML 2018 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
-
-\documentclass{article}
-
-% Recommended, but optional, packages for figures and better typesetting:
-\usepackage{amsmath, amssymb}
-\usepackage{mathtools}
-\usepackage{microtype}
-\usepackage{graphicx}
-\usepackage{subfigure}
-\usepackage{booktabs} % for professional tables
-\usepackage{xcolor}
-\usepackage{algorithmic}
-
-% hyperref makes hyperlinks in the resulting PDF.
-% If your build breaks (sometimes temporarily if a hyperlink spans a page)
-% please comment out the following usepackage line and replace
-% \usepackage{icml2018} with \usepackage[nohyperref]{icml2018} above.
-\usepackage{hyperref}
-
-\input{macros.tex}
-
-% Use the following line for the initial blind version submitted for review:
-\usepackage{icml2018}
-
-% If accepted, instead use the following line for the camera-ready submission:
-%\usepackage[accepted]{icml2018}
-
-% The \icmltitle you define below is probably too long as a header.
-% Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Deep Relativistic Advantage Updating}
-
-\begin{document}
-
-\twocolumn[
-\icmltitle{There are no small Advantages:
-	Reinforcement learning in near continuous time
-	with Deep RelAtive Mixture of Advantages
-}
-
-% It is OKAY to include author information, even for blind
-% submissions: the style file will automatically remove it for you
-% unless you've provided the [accepted] option to the icml2018
-% package.
-
-% List of affiliations: The first argument should be a (short)
-% identifier you will use later to specify author affiliations
-% Academic affiliations should list Department, University, City, Region, Country
-% Industry affiliations should list Company, City, Region, Country
-
-% You can specify symbols, otherwise they are numbered in order.
-% Ideally, you should not use this facility. Affiliations will be numbered
-% in order of appearance and this is the preferred way.
-\icmlsetsymbol{equal}{*}
-
-\begin{icmlauthorlist}
-%\icmlauthor{Aeiau Zzzz}{equal,to}
-%\icmlauthor{Bauiu C.~Yyyy}{equal,to,goo}
-%\icmlauthor{Cieua Vvvvv}{goo}
-%\icmlauthor{Iaesut Saoeu}{ed}
-%\icmlauthor{Fiuea Rrrr}{to}
-%\icmlauthor{Tateu H.~Yasehe}{ed,to,goo}
-%\icmlauthor{Aaoeu Iasoh}{goo}
-%\icmlauthor{Buiui Eueu}{ed}
-%\icmlauthor{Aeuia Zzzz}{ed}
-%\icmlauthor{Bieea C.~Yyyy}{to,goo}
-%\icmlauthor{Teoau Xxxx}{ed}
-%\icmlauthor{Eee Pppp}{ed}
-\end{icmlauthorlist}
-
-%\icmlaffiliation{to}{Department of Computation, University of Torontoland, Torontoland, Canada}
-%\icmlaffiliation{goo}{Googol ShallowMind, New London, Michigan, USA}
-%\icmlaffiliation{ed}{School of Computation, University of Edenborrow, Edenborrow, United Kingdom}
-%
-%\icmlcorrespondingauthor{Cieua Vvvvv}{c.vvvvv@googol.com}
-%\icmlcorrespondingauthor{Eee Pppp}{ep@eden.co.uk}
-
-% You may provide any keywords that you
-% find helpful for describing your paper; these are used to populate
-% the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{Reinforcement learning}
-
-\vskip 0.3in
-]
-
-% this must go after the closing bracket ] following \twocolumn[
-
-% This command actually creates the footnote in the first column
-% listing the affiliations and the copyright notice.
-% The command takes one argument, which is text to display at the start of the footnote.
-% The \icmlEqualContribution command is standard text for equal contribution.
-% Remove it (just {}) if you do not need this facility.
-
-%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
-%\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
-
-\begin{abstract}
-	Despite remarkable successes, \emph{Deep Reinforcement learning} remains sensitive to
-	hyperparameterization, implementation details and small environment changes~(\citealt{drl_matter}, \citealt{drl_matter_bis}). 
-	Overcoming this sensitivity is key to making DRL applicable to real world problems.
-	In this paper, we study the sensitivity of Deep Reinforcement learning algorithms to
-	time discretization. More precisely, we show that approaches based on estimations of the
-	Q-function, e.g. \emph{Deep Q-learning}~\cite{dqn} and \emph{Deep deterministic policy gradient}~\cite{ddpg}
-	are sensitive to variations of time discretization. We further show that a simple reparameterization of the
-	Q-function as a sum of a state value term and a \emph{small} action dependent advantage term yields an algorithm
-	much more resilient to variations of time discretization.
-\end{abstract}
-
-\section{Introduction}
-\label{sec:intro}
-In recent years, deep learning based approaches to reinforcement learning have
-thrived and provided impressive results in a variety of domains, achieving superhuman
-performances with no expert knowledge in perfect information zero-sum
-games~\cite{alphazero}, reaching level of top players in video
-games~(\citealt{openai_five}, \citealt{dqn}), or learning dexterous manipulation
-from scratch without demonstrations~\cite{hand_control}.
-
-In spite of those successes, \emph{Deep reinforcement learning} approaches are
-sensitive to a number of factors, including hyperparameterization,
-implementation details or small changes in the environment
-parameters~(\citealt{drl_matter}, \citealt{drl_matter_bis}). This sensitivity,
-along with sample inefficiency, prevents DRL from being applied in most real
-world settings. High sensitivity to environment parameters notably prevents
-transfering abilities from imperfect simulators to real world scenarios.
-
-In this paper we focus our attention on the sensitivity to time discretization
-of DRL approaches, i.e. what happens when an agent receives $50$ observations
-and is expected to take $50$ actions per second instead of $10$. One expects
-that decreasing time discretization, or equivalently shortening reaction time,
-should only improve agent performances. This is not what happens in practice
-for approaches based on estimation of state action value functions, e.g.
-\emph{Deep Q-learning} (DQN~\citep{dqn} and \emph{Deep deterministic policy
-gradient} (DDPG~\citep{ddpg}). This is shown experimentally in Sec.~\ref{sec:exp}.
-
-We relate this sensitivity to the fact that, as the discretization timestep
-decreases, the effect of individual actions on the total return decreases, and
-can vanish when the discretization timestep becomes infinitesimal. The natural
-framework to study this phenomenon is Continuous time reinforcement learning 
-(Sec.~\ref{sec:continous}).
-
-To cope for this effect, we introduce a
-parameterization of the Q-function as a sum of a state value term and a
-\emph{small} action dependent advantage term (Sec.~\ref{subsec:reparam}). This
-reparameterized Q approximation can be trained with deep variants of Q-learning
-(Sec.~\ref{subsec:algorithm}).  To properly scale w.r.t. time discretization,
-particular attention has to be taken regarding both the exploration method, and
-the scaling of learning rates (Sec.~\ref{subsec:lr}, Sec.~\ref{subsec:explo}).
-The resulting algorithm is shown to provide near perfect invariance to time
-discretization on simple environments, and much better invariance properties
-than vanilla DQN or DDPG on more complex environments (Sec.~\ref{sec:exp}).
-
-\section{Related work}
-\label{sec:related}
-Continuous and near continuous reinforcement learning, which is at the core of
-our approach has been thoroughly studied in the deterministic case
-in~\cite{adv_upd} or \cite{cont_rl}, but has not been applied in the context of
-deep reinforcement learning.
-
-The approach presented in this paper is closely related to \cite{adv_upd} and
-\cite{cont_rl}. \cite{adv_upd} already separated the learning procedure of the
-value and advantage functions, and introduced proper scalings between the two
-components. We extend on \cite{adv_upd} in three directions. We provide a
-formal argument on why the advantage contribution of the Q-function is of order
-$\bigO(\deltat)$, irrelevant of the stochasticity of the environment, in a quite
-general framework. We show that, provided with well suited learning rates and
-exploration method, Advantage updating provide near invariance to change of time
-discretization. We show that Advantage updating is viable in the context of deep
-reinforcement learning, while not using the true gradient of the residual error.
-
-More recently, \cite{dueling_nets} also introduced a parameterization separating
-the value and the advantage components of the Q-function. We additionally
-introduce a natural scaling coeficient between the two, that strongly improves
-the time discretization invariance properties.
-
 \section{Near continuous time reinforcement learning}
@@
 \end{equation}
-
-\section{Deep Advantage Updating}
-In continuous time, $Q^\pi$ is independant of actions and can thus not be used
-to select promising actions.  In near continuous time, $Q^\pi$ still depends on
-actions, and can still be used to choose promising actions under the current
-policy. However, when approximating $Q^\pi$, if the approximation error is much
-bigger than $\bigO(\deltat)$, approximation errors dominates, and the hierarchy
-of actions given by the approximation is likely to be erroneous.
-
-When the approximate Q-function is initialized, if the effect of actions on the
-Q-function is order of magnitudes higher than what it should be, approximating
-$Q^\pi$ is likely to be difficult. Besides, the error is likely to be further
-propagateed when the equation used to update our approximate $Q$ function
-relies on bootstraping, as is the case for usual temporal difference derived methods.
-
-\subsection{Reparameterizing $Q_\theta$}
-\label{subsec:reparam}
-To prevent overestimating the dependency on actions of the approximation of
-$Q^\pi$, $Q_\theta$, in an environment with time discretization $\deltat$,
-Eq.~\eqref{eq:reparam_q_pi} naturally yields the following reparameterization
-\begin{equation}
-	Q_\theta(s, a) = V_{\theta_1}(s) + \deltat A_{\theta_2}(s, a).
-\end{equation}
-To avoid cumbersome notations, in what follows, parameter indices are dropped.
-The insight behind this parameterization is that when $Q$ correctly
-approximates $Q^\pi$, $V$ should approximate the value function, while $A$
-should approximate the advantage function, as defined in Eq.~\eqref{eq:adv}.
-
-The proposed reparameterization does not, on its own, guarantee that
-when $Q$ correctly approximates $Q^\pi$, $A$ approximates $A^\pi$.  Indeed, for
-any given pair $V$, $A$, the transformed pair
-\begin{align}
-	\tilde{V}(s) &= V(s) - f(s)\\
-	\tilde{A}(s, a) &= A(s, a) + \frac{f(s)}{\deltat}
-\end{align}
-yields the exact same $Q$ function. Consequently, $Q$ can correctly approximate
-$Q^\pi$ even if $A$ is arbitrarily far from $A^\pi$.
-To enforce, identifiability of $A$, one must enforce that 
-\begin{equation}
-	A(s, \pi(s)) = 0.
-\end{equation}
-With this additional constraint, if $Q^\pi = Q$,
-\begin{align}
-	A^\pi(s, a) &= \frac{Q^\pi(s,a) - V^\pi(s)}{\deltat}\\
-		    &= \frac{Q(s,a) - Q(s, \pi(s))}{\deltat}\\
-		    &= A(s, a).
-\end{align}
-In the spirit of~\cite{dueling_nets}, this condition is enforced by writing $A$ as
-\begin{equation}
-	A(s, a) = \bar{A}(s, a) - \bar{A}(s, \pi(s)).
-\end{equation}
-
-\subsection{Time invariant exploration}
-\label{subsec:explo}
-To obtain a time invariant reinforcement learning algorithm, a time invariant
-exploration scheme is required.
-
-The exploration scheme used~\cite{ddpg}, which consists in adding an
-\emph{Ornstein-Uhlenbeck}~\cite{orn-uhl} process to actions, is naturally time
-invariant in the sense that it is first defined as a time continuous stochastic
-process, and that discretization of this process are used in practice. Indeed,
-a continuous Ornstein-Uhlenbeck process is defined as
-\begin{equation}
-	dx_t = - (x_t - \mu) \theta dt + \sigma \dbt
-	\label{eq:orn_uhl}
-\end{equation}
-and is naturally discretized by turning $d$'s into $\delta$'s, and using
-the same $\deltat$ for both the learning algorithm and the exploration scheme.
-Consequently, this exploration scheme can readily be used in continuous action
-environments.
-
-On the other hand, the usual exploration scheme used in discrete actions
-Q-learning, $\varepsilon$-greedy exploration~\cite{sutton} is not time
-invariant, in the sense that resulting state trajectories do not converge to a
-well defined stochastic process as $\deltat \rightarrow 0$. As $\dt$ goes to $0$,
-for any $\varepsilon$ independant of $\dt$, the behavior of state trajectories get
-closer to that of a white noise.
-
-To obtain a temporally coherent exploration scheme in discrete action environment
-when using Q-learning we instead directly apply a discretization of a time continuous
-stochastic process to the output of the approximate Q-function, and select actions
-by maximizing the resulting perturbed Q-function. Typically, the approximate Q-function
-values are also perturbed using a discretized Ornstein-Uhlenbeck noise.
-
-\subsection{Algorithms}
-\label{subsec:algorithm}
-Two variants of \emph{Deep Advantage Updating} are derived to cover both the continuous
-and discrete action cases. They both rely on variants of Q-learning. More precisely,
-we learn $V$ and $A$ using approximate version of the following bellman equation
-\begin{align}
-	V^{\pi}(s) + A^{\pi}(s, a) &= \E_{\delta r \sim \delta r(s, a)}\delta r\nonumber\\ 
-				   &+ \gamma^{\deltat}  \E_{\delta s \sim \delta s(s, a)} V^{\pi}(s + \delta s)\label{eq:bellman_A}\\
-	A^{\pi}(s, \pi(s)) &= 0\label{eq:max_A}
-\end{align}
-for a greedy exploitation policy $\pi(s) = \text{argmax}_{a'}A^\pi(s, a')$.
-Eq.~\eqref{eq:max_A} is directly verified owing to the parameterization of $A$.
-To approximately verify~\eqref{eq:bellman_A}, the corresponding squared residual
-is minimized by an approximate gradient descent.
-The corresponding update equations when learning from incoming transitions of
-an exploratory trajectory are
-\begin{align}
-	\delta Q_t &= A(s_t, a_t) - \frac{\delta r_t + \gamma^{\deltat} V(s_{t+\deltat}) - V(s_t)}{\deltat}\\
-	\theta^1_{t+\deltat} &= \theta^1_t + \eta_1 \partial_{\theta^1} V(s_t) \delta Q_t\\
-	\theta^2_{t+\deltat} &= \theta^2_t + \eta_2 \partial_{\theta^2} A(s_t, a_t) \delta Q_t.
-	\label{eq:approx_bellman_A}
-\end{align}
-where $\eta$'s are learning rates. We are now left to derive appropriate learning rates
-to provide invariance to time discretization.
-\begin{algorithm}[ht]
-	\caption{Deep Advantage Updating (Discrete actions)}
-	\input{dauc.tex}
-\end{algorithm}
-
-\subsection{Learning rates scalings}
-\label{subsec:lr}
-A necessary condition to guarantee invariance to time discretization is that
-the trajectory of the parameters of our function approximators should remain
-well behaved when the discretization parameter $\deltat$ decreases. This notably
-means that optimization steps should not be too large, in which case parameter
-trajectories could diverge in a single gradient step, or two small, in which case
-the trajectories would converge to single points as $\deltat$ goes to $0$.
-This in turn imposes conditions on the scalings of the learning rates of invariant
-algorithms with respect to the discretization timestep.
-
-If we optimize $\theta_1$ and $\theta_2$ according to Eq.~\eqref{eq:approx_bellman_A},
-both $\theta_1$ and $\theta_2$ verify update equations of the form
-\begin{align}
-	\
-	\delta \theta^1_t = \eta_1 \partial_{\theta_1} V(s_t) (
-\end{align}
-To have parameter steps that do not grow larger or shrink to $0$ as $\deltat
-\leftarrow 0$, both $\delta \theta^1$ and $\delta \theta^2$ should verify
-stochastic differential equations, and notably have a deterministic component
-of order $\deltat$. If the deterministic component 
-\section{Experiments}
-\label{sec:exp}
-
-\section{Limitations}
-\label{sec:limitations}
-\section{Conclusion}
-\label{sec:concl}
-
-\bibliography{icml_drau.bib}
-\bibliographystyle{icml2018}
-\end{document}
-
-% This document was modified from the file originally made available by
-% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
-% by Iain Murray in 2018. It was modified from a version from Dan Roy in
-% 2017, which was based on a version from Lise Getoor and Tobias
-% Scheffer, which was slightly modified from the 2010 version by
-% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
-% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
-% slightly modified from Prasad Tadepalli's 2007 version which is a
-% lightly changed version of the previous year's version by Andrew
-% Moore, which was in turn edited from those of Kristian Kersting and
-% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2019-01-10 paper/icml_drau/icml_drau.tex
split texte between files

@@
 
-\section{Deep Advantage Updating}
-In continuous time, $Q^\pi$ is independant of actions and can thus not be used
-to select promising actions.  In near continuous time, $Q^\pi$ still depends on
-actions, and can still be used to choose promising actions under the current
-policy. However, when approximating $Q^\pi$, if the approximation error is much
-bigger than $\bigO(\deltat)$, approximation errors dominates, and the hierarchy
-of actions given by the approximation is likely to be erroneous.
+\input{advup.tex}
 
-When the approximate Q-function is initialized, if the effect of actions on the
-Q-function is order of magnitudes higher than what it should be, approximating
-$Q^\pi$ is likely to be difficult. Besides, the error is likely to be further
-propagateed when the equation used to update our approximate $Q$ function
-relies on bootstraping, as is the case for usual temporal difference derived methods.
+\input{experiments.tex}
 
-\subsection{Reparameterizing $Q_\theta$}
-\label{subsec:reparam}
-To prevent overestimating the dependency on actions of the approximation of
-$Q^\pi$, $Q_\theta$, in an environment with time discretization $\deltat$,
-Eq.~\eqref{eq:reparam_q_pi} naturally yields the following reparameterization
-\begin{equation}
-	Q_\theta(s, a) = V_{\theta_1}(s) + \deltat A_{\theta_2}(s, a).
-\end{equation}
-To avoid cumbersome notations, in what follows, parameter indices are dropped.
-The insight behind this parameterization is that when $Q$ correctly
-approximates $Q^\pi$, $V$ should approximate the value function, while $A$
-should approximate the advantage function, as defined in Eq.~\eqref{eq:adv}.
+\input{discussions.tex}
 
-The proposed reparameterization does not, on its own, guarantee that
-when $Q$ correctly approximates $Q^\pi$, $A$ approximates $A^\pi$.  Indeed, for
-any given pair $V$, $A$, the transformed pair
-\begin{align}
-	\tilde{V}(s) &= V(s) - f(s)\\
-	\tilde{A}(s, a) &= A(s, a) + \frac{f(s)}{\deltat}
-\end{align}
-yields the exact same $Q$ function. Consequently, $Q$ can correctly approximate
-$Q^\pi$ even if $A$ is arbitrarily far from $A^\pi$.
-To enforce, identifiability of $A$, one must enforce that 
-\begin{equation}
-	A(s, \pi(s)) = 0.
-\end{equation}
-With this additional constraint, if $Q^\pi = Q$,
-\begin{align}
-	A^\pi(s, a) &= \frac{Q^\pi(s,a) - V^\pi(s)}{\deltat}\\
-		    &= \frac{Q(s,a) - Q(s, \pi(s))}{\deltat}\\
-		    &= A(s, a).
-\end{align}
-In the spirit of~\cite{dueling_nets}, this condition is enforced by writing $A$ as
-\begin{equation}
-	A(s, a) = \bar{A}(s, a) - \bar{A}(s, \pi(s)).
-\end{equation}
-
-\subsection{Time invariant exploration}
-\label{subsec:explo}
-To obtain a time invariant reinforcement learning algorithm, a time invariant
-exploration scheme is required.
-
-The exploration scheme used~\cite{ddpg}, which consists in adding an
-\emph{Ornstein-Uhlenbeck}~\cite{orn-uhl} process to actions, is naturally time
-invariant in the sense that it is first defined as a time continuous stochastic
-process, and that discretization of this process are used in practice. Indeed,
-a continuous Ornstein-Uhlenbeck process is defined as
-\begin{equation}
-	dx_t = - (x_t - \mu) \theta dt + \sigma \dbt
-	\label{eq:orn_uhl}
-\end{equation}
-and is naturally discretized by turning $d$'s into $\delta$'s, and using
-the same $\deltat$ for both the learning algorithm and the exploration scheme.
-Consequently, this exploration scheme can readily be used in continuous action
-environments.
-
-On the other hand, the usual exploration scheme used in discrete actions
-Q-learning, $\varepsilon$-greedy exploration~\cite{sutton} is not time
-invariant, in the sense that resulting state trajectories do not converge to a
-well defined stochastic process as $\deltat \rightarrow 0$. As $\dt$ goes to $0$,
-for any $\varepsilon$ independant of $\dt$, the behavior of state trajectories get
-closer to that of a white noise.
-
-To obtain a temporally coherent exploration scheme in discrete action environment
-when using Q-learning we instead directly apply a discretization of a time continuous
-stochastic process to the output of the approximate Q-function, and select actions
-by maximizing the resulting perturbed Q-function. Typically, the approximate Q-function
-values are also perturbed using a discretized Ornstein-Uhlenbeck noise.
-
-\subsection{Algorithms}
-\label{subsec:algorithm}
-Two variants of \emph{Deep Advantage Updating} are derived to cover both the continuous
-and discrete action cases. They both rely on variants of Q-learning. More precisely,
-we learn $V$ and $A$ using approximate version of the following bellman equation
-\begin{align}
-	V^{\pi}(s) + A^{\pi}(s, a) &= \E_{\delta r \sim \delta r(s, a)}\delta r\nonumber\\ 
-				   &+ \gamma^{\deltat}  \E_{\delta s \sim \delta s(s, a)} V^{\pi}(s + \delta s)\label{eq:bellman_A}\\
-	A^{\pi}(s, \pi(s)) &= 0\label{eq:max_A}
-\end{align}
-for a greedy exploitation policy $\pi(s) = \text{argmax}_{a'}A^\pi(s, a')$.
-Eq.~\eqref{eq:max_A} is directly verified owing to the parameterization of $A$.
-To approximately verify~\eqref{eq:bellman_A}, the corresponding squared residual
-is minimized by an approximate gradient descent.
-The corresponding update equations when learning from incoming transitions of
-an exploratory trajectory are
-\begin{align}
-	\delta Q_t &= A(s_t, a_t) - \frac{\delta r_t + \gamma^{\deltat} V(s_{t+\deltat}) - V(s_t)}{\deltat}\\
-	\theta^1_{t+\deltat} &= \theta^1_t + \eta_1 \partial_{\theta^1} V(s_t) \delta Q_t\\
-	\theta^2_{t+\deltat} &= \theta^2_t + \eta_2 \partial_{\theta^2} A(s_t, a_t) \delta Q_t.
-	\label{eq:approx_bellman_A}
-\end{align}
-where $\eta$'s are learning rates. We are now left to derive appropriate learning rates
-to provide invariance to time discretization.
-\begin{algorithm}[ht]
-	\caption{Deep Advantage Updating (Discrete actions)}
-	\input{dauc.tex}
-\end{algorithm}
-
-\subsection{Learning rates scalings}
-\label{subsec:lr}
-A necessary condition to guarantee invariance to time discretization is that
-the trajectory of the parameters of our function approximators should remain
-well behaved when the discretization parameter $\deltat$ decreases. This notably
-means that optimization steps should not be too large, in which case parameter
-trajectories could diverge in a single gradient step, or two small, in which case
-the trajectories would converge to single points as $\deltat$ goes to $0$.
-This in turn imposes conditions on the scalings of the learning rates of invariant
-algorithms with respect to the discretization timestep.
-
-If we optimize $\theta_1$ and $\theta_2$ according to Eq.~\eqref{eq:approx_bellman_A},
-both $\theta_1$ and $\theta_2$ verify update equations of the form
-\begin{align}
-	\
-	\delta \theta^1_t = \eta_1 \partial_{\theta_1} V(s_t) (
-\end{align}
-To have parameter steps that do not grow larger or shrink to $0$ as $\deltat
-\leftarrow 0$, both $\delta \theta^1$ and $\delta \theta^2$ should verify
-stochastic differential equations, and notably have a deterministic component
-of order $\deltat$. If the deterministic component 
-\section{Experiments}
-\label{sec:exp}
-
-\section{Limitations}
-\label{sec:limitations}
-\section{Conclusion}
-\label{sec:concl}
+\input{conclusion.tex}
 

### 2019-01-10 paper/icml_drau/intro.tex
split texte between files

--- 2019-01-10 paper/icml_drau/icml_drau.tex
@@
-%%%%%%%% ICML 2018 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
-
-\documentclass{article}
-
-% Recommended, but optional, packages for figures and better typesetting:
-\usepackage{amsmath, amssymb}
-\usepackage{mathtools}
-\usepackage{microtype}
-\usepackage{graphicx}
-\usepackage{subfigure}
-\usepackage{booktabs} % for professional tables
-\usepackage{xcolor}
-\usepackage{algorithmic}
-
-% hyperref makes hyperlinks in the resulting PDF.
-% If your build breaks (sometimes temporarily if a hyperlink spans a page)
-% please comment out the following usepackage line and replace
-% \usepackage{icml2018} with \usepackage[nohyperref]{icml2018} above.
-\usepackage{hyperref}
-
-\input{macros.tex}
-
-% Use the following line for the initial blind version submitted for review:
-\usepackage{icml2018}
-
-% If accepted, instead use the following line for the camera-ready submission:
-%\usepackage[accepted]{icml2018}
-
-% The \icmltitle you define below is probably too long as a header.
-% Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Deep Relativistic Advantage Updating}
-
-\begin{document}
-
-\twocolumn[
-\icmltitle{There are no small Advantages:
-	Reinforcement learning in near continuous time
-	with Deep RelAtive Mixture of Advantages
-}
-
-% It is OKAY to include author information, even for blind
-% submissions: the style file will automatically remove it for you
-% unless you've provided the [accepted] option to the icml2018
-% package.
-
-% List of affiliations: The first argument should be a (short)
-% identifier you will use later to specify author affiliations
-% Academic affiliations should list Department, University, City, Region, Country
-% Industry affiliations should list Company, City, Region, Country
-
-% You can specify symbols, otherwise they are numbered in order.
-% Ideally, you should not use this facility. Affiliations will be numbered
-% in order of appearance and this is the preferred way.
-\icmlsetsymbol{equal}{*}
-
-\begin{icmlauthorlist}
-%\icmlauthor{Aeiau Zzzz}{equal,to}
-%\icmlauthor{Bauiu C.~Yyyy}{equal,to,goo}
-%\icmlauthor{Cieua Vvvvv}{goo}
-%\icmlauthor{Iaesut Saoeu}{ed}
-%\icmlauthor{Fiuea Rrrr}{to}
-%\icmlauthor{Tateu H.~Yasehe}{ed,to,goo}
-%\icmlauthor{Aaoeu Iasoh}{goo}
-%\icmlauthor{Buiui Eueu}{ed}
-%\icmlauthor{Aeuia Zzzz}{ed}
-%\icmlauthor{Bieea C.~Yyyy}{to,goo}
-%\icmlauthor{Teoau Xxxx}{ed}
-%\icmlauthor{Eee Pppp}{ed}
-\end{icmlauthorlist}
-
-%\icmlaffiliation{to}{Department of Computation, University of Torontoland, Torontoland, Canada}
-%\icmlaffiliation{goo}{Googol ShallowMind, New London, Michigan, USA}
-%\icmlaffiliation{ed}{School of Computation, University of Edenborrow, Edenborrow, United Kingdom}
-%
-%\icmlcorrespondingauthor{Cieua Vvvvv}{c.vvvvv@googol.com}
-%\icmlcorrespondingauthor{Eee Pppp}{ep@eden.co.uk}
-
-% You may provide any keywords that you
-% find helpful for describing your paper; these are used to populate
-% the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{Reinforcement learning}
-
-\vskip 0.3in
-]
-
-% this must go after the closing bracket ] following \twocolumn[
-
-% This command actually creates the footnote in the first column
-% listing the affiliations and the copyright notice.
-% The command takes one argument, which is text to display at the start of the footnote.
-% The \icmlEqualContribution command is standard text for equal contribution.
-% Remove it (just {}) if you do not need this facility.
-
-%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
-%\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
-
-\begin{abstract}
-	Despite remarkable successes, \emph{Deep Reinforcement learning} remains sensitive to
-	hyperparameterization, implementation details and small environment changes~(\citealt{drl_matter}, \citealt{drl_matter_bis}). 
-	Overcoming this sensitivity is key to making DRL applicable to real world problems.
-	In this paper, we study the sensitivity of Deep Reinforcement learning algorithms to
-	time discretization. More precisely, we show that approaches based on estimations of the
-	Q-function, e.g. \emph{Deep Q-learning}~\cite{dqn} and \emph{Deep deterministic policy gradient}~\cite{ddpg}
-	are sensitive to variations of time discretization. We further show that a simple reparameterization of the
-	Q-function as a sum of a state value term and a \emph{small} action dependent advantage term yields an algorithm
-	much more resilient to variations of time discretization.
-\end{abstract}
-
 \section{Introduction}
@@
 than vanilla DQN or DDPG on more complex environments (Sec.~\ref{sec:exp}).
-
-\section{Related work}
-\label{sec:related}
-Continuous and near continuous reinforcement learning, which is at the core of
-our approach has been thoroughly studied in the deterministic case
-in~\cite{adv_upd} or \cite{cont_rl}, but has not been applied in the context of
-deep reinforcement learning.
-
-The approach presented in this paper is closely related to \cite{adv_upd} and
-\cite{cont_rl}. \cite{adv_upd} already separated the learning procedure of the
-value and advantage functions, and introduced proper scalings between the two
-components. We extend on \cite{adv_upd} in three directions. We provide a
-formal argument on why the advantage contribution of the Q-function is of order
-$\bigO(\deltat)$, irrelevant of the stochasticity of the environment, in a quite
-general framework. We show that, provided with well suited learning rates and
-exploration method, Advantage updating provide near invariance to change of time
-discretization. We show that Advantage updating is viable in the context of deep
-reinforcement learning, while not using the true gradient of the residual error.
-
-More recently, \cite{dueling_nets} also introduced a parameterization separating
-the value and the advantage components of the Q-function. We additionally
-introduce a natural scaling coeficient between the two, that strongly improves
-the time discretization invariance properties.
-
-\section{Near continuous time reinforcement learning}
-\label{sec:continous}
-% In this section, a continuous reinforcement learning framework is introduced,
-% and it is shown that under this framework, the state-action value function does
-% not depend on the action. We further show that in near continous domains, when
-% the discretization timestep $\deltat$ is small, the difference in state-action
-% value between two different actions in a fixed state is of order $\bigO(\deltat)$.
-TODO. Main ideas: near continuous time environments (control, mujoco, video games, ...)
-There is a time discretization $\deltat$ given by the problem.
-The uderlying process does not depend of $\deltat$.
-Need to see what is happening when changing $\deltat$.
-The problem is invariant to $\deltat$.
-The quantities we are looking at should be invariant to.
-The learning procedure should be invariant.
-Looking at the limits when $\deltat \rightarrow 0$.
-In simulated environments, changing $\deltat$ only means to change the FPS.
-In real environments, for example with robots, this could mean changing the quality of the sensors and motors.
-
-\subsection{Framework}
-
-Let assume that we have a Markov Decision Process (TODO:Cite?) $\langle {\cal S}, {\cal A}, T_{\deltat}, r, \gamma\rangle$.
-${\cal S}$ is the state-space, which is supposed to be continuous (TODO:not the good formulation).
-The action space is TODO.
-
-We assume that the transition is continuous.
-This means that ...
-TODO.
-Discussion on the hypothesis, in which environments it is true.
-Is the condition $\forall \deltat, \forall s, \E[\|s_{k+1} - s_k \||s_k = s] \leq ?$.
-
-In what follows, $\dt$ denotes a truly infinitesimal time interval while $\deltat$
-denotes a small, but non infinitesimal time interval. Similarily, $\dbt$ denotes
-an infinitesimal (multi-dimensional) brownian step, while $\deltabt$ denotes its
-discretized equivalent. Informally, $\dbt$ is to be thought of as $\gauss(0, \sqrt{\dt}^2)$,
-and $\deltabt$ as $\gauss(0, \sqrt{\deltat}^2)$, where $\gauss(\mu, \sigma)$ denotes a gaussian
-random variable of mean $\mu$ and standard deviation (or covariance matrix in the multi-dimensional case)
-$\sigma$.
-
-Consider a fully observable \emph{Markov Decision Process} defined by the following
-transitions and rewards
-\begin{align}
-	ds(s, a) &= F(s, a) dt + \Sigma(s, a) \dbt\\
-	dr(s, a) &= \reward(s, a) dt + \Sigma_r(s, a) \dbt.
-\end{align}
-The equation on $s$ is that of a diffusion process, with drift coeficient $F$
-and diffusion matrix $\Sigma$.  This framework is only moderately restrictive.
-For example, deterministic continuous control environments fall in this
-framework, with $\Sigma = 0$ and $\Sigma_r = 0$. Limitations of the framework
-are discussed in Sec.~\ref{sec:limitations}.
-This framework is naturally discretized to arbitrary $\deltat$ by replacing all
-$d$'s with $\delta$'s.
-
-A continuous policy $\pi$ is defined as a deterministic mapping from states to
-actions.  In continuous time, defining stochastic policies with well defined
-state-value functions requires incorporating information from previous action into
-the state to obtain a temporally coherent policy. To avoid such inconvenience, the
-framework is restricted to deterministic policies. This is a mild constraint, as this
-only enforces determinism of the exploitation policy and not of exploration policies.
-In what follows, $\tau\sim\pi$ denotes a trajectory of states and actions sampled
-according to policy $\pi$, with $s_0$ sampled according to an arbitrary initial distribution
-on states $\rho_0$, and $dr_{t} = dr(s_t, a_t)$.
-
-Under a deterministic policy $\pi$, one can define the $\gamma$ discounted
-\emph{state value function} $V^\pi(s)$ and \emph{state action value function}
-$Q^\pi(s, a)$ as
-\begin{align}
-	V^\pi(s) &= \E_{\tau\sim\pi}\left[
-		\int\limits_{t=0}^\infty \gamma^{t}
-		dr_{t} \mid s_0 = s
-	\right]\\
-	Q^\pi(s, a) &= \E_{\tau\sim\pi}\left[
-		\int\limits_{t=0}^\infty \gamma^{t}
-		dr_{t} \mid s_0 = s, a_0 = a
-	\right]
-\end{align}
-which are naturally discretized as
-\begin{align}
-	V^\pi(s) &= \E_{\tau\sim\pi}\left[
-		\sum\limits_{k=0}^\infty \gamma^{k\deltat}
-		\delta r_{k\deltat} \mid s_0 = s
-	\right]\\
-	Q^\pi(s, a) &= \E_{\tau\sim\pi}\left[
-		\sum\limits_{k=0}^\infty \gamma^{k\deltat}
-		\delta r_{k\deltat} \mid s_0 = s, a_0 = a
-	\right].
-\end{align}
-Both $V^\pi$ and $Q^\pi$ verify Bellman equations, informally
-\begin{align}
-	V^\pi(s) &= \reward(s, \pi(s))\dt + \gamma^{\dt} \E_{ds \sim ds(s, \pi(s))} V^\pi(s + ds)\nonumber\\
-	Q^\pi(s, a) &= \reward(s, a)\dt + \gamma^{\dt} \E_{ds(s, a)} V^\pi(s + ds(s, a))
-	\label{eq:cont_bell}
-\end{align}
-which in continuous time translates, using Ito's lemma, into a Hamilton Jacobi equation
-\begin{align}
-	\reward + \partial_s V^\pi \cdot f + \frac{1}{2} \text{tr}\left(\Sigma^T\partial^2_{s^2} V^\pi\Sigma\right) = (1 - \gamma) V^\pi(s).
-\end{align}
-
-\subsection{Independence of $Q^\pi$ on $a$ in continuous time}
-Consider a discretization of Eq.~\eqref{eq:cont_bell} with time interval
-$\deltat \ll~1$, Ito's lemma yield
-\begin{align}
-	Q^\pi(s, a) &= \reward(s, a) \deltat + \gamma^{\deltat} \E_{\delta s \sim \delta s(s, a)}V^\pi(s + \delta s)\nonumber\\
-		    &= \reward(s, a) \deltat + \gamma^{\deltat} V^\pi(s) +
-		     \gamma^{\deltat} \partial_s V^\pi(s)f(s, a) \deltat\nonumber\\
-		    &+ \frac{1}{2}\text{tr}\left(\Sigma^T\partial^2_{s^2}V^\pi(s)\Sigma\right)\deltat + \bigO(\deltat^2)\nonumber\\
-		    &\xrightarrow[\deltat \to 0]{} V^\pi(s).
-		    \label{eq:Q_cont}
-\end{align}
-A consequence of Eq.~\eqref{eq:Q_cont} is that in exact continuous time,
-$Q^\pi$ does not bear any information on the hierarchy of actions, and
-thus cannot be used to select actions that yields higher future returns.
-
-As the difference between between $Q^\pi(s, a)$ and $V^\pi(s)$ is of order
-$\bigO(\deltat)$, it is natural to define a rescaled version of the advantage
-function, namely
-\begin{align}
-	A^\pi(s, a) &= \frac{Q^\pi(s,a) - V^\pi(s)}{\deltat}\\
-		    &= \reward(s, a) +
-		    \frac{\gamma^{\deltat} \E_{\delta s \sim \delta s(s, a)} V^\pi(s + \delta s) - V^\pi(s)}{\deltat}.
-    \label{eq:adv}
-\end{align}
-Contrary to the action state value function, the rescaled advantage function converges
-to an action dependent quantity when $\deltat$ goes to $0$
-\begin{align}
-	A^\pi(s, a) &\xrightarrow[\deltat \to 0]{} \reward(s, a) +
-	(\gamma - 1) V^\pi(s) + \partial_s V^\pi(s) f(s, a)\nonumber\\
-		    &+
-	\frac{\text{tr}\left(\Sigma^T(s, a)\partial^2_{s^2}V^\pi(s) \Sigma(s,a)\right)}{2},
-\end{align}
-and $Q^\pi$ can then be rewritten as
-\begin{equation}
-	Q^\pi(s, a) = V^\pi(s) + \deltat A^\pi(s, a).
-	\label{eq:reparam_q_pi}
-\end{equation}
-
-\input{advup.tex}
-
-\input{experiments.tex}
-
-\input{discussions.tex}
-
-\input{conclusion.tex}
-
-\bibliography{icml_drau.bib}
-\bibliographystyle{icml2018}
-\end{document}
-
-% This document was modified from the file originally made available by
-% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
-% by Iain Murray in 2018. It was modified from a version from Dan Roy in
-% 2017, which was based on a version from Lise Getoor and Tobias
-% Scheffer, which was slightly modified from the 2010 version by
-% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
-% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
-% slightly modified from Prasad Tadepalli's 2007 version which is a
-% lightly changed version of the previous year's version by Andrew
-% Moore, which was in turn edited from those of Kristian Kersting and
-% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2019-01-10 paper/icml_drau/relatedworks.tex
split texte between files

--- 2019-01-10 paper/icml_drau/icml_drau.tex
@@
-%%%%%%%% ICML 2018 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
-
-\documentclass{article}
-
-% Recommended, but optional, packages for figures and better typesetting:
-\usepackage{amsmath, amssymb}
-\usepackage{mathtools}
-\usepackage{microtype}
-\usepackage{graphicx}
-\usepackage{subfigure}
-\usepackage{booktabs} % for professional tables
-\usepackage{xcolor}
-\usepackage{algorithmic}
-
-% hyperref makes hyperlinks in the resulting PDF.
-% If your build breaks (sometimes temporarily if a hyperlink spans a page)
-% please comment out the following usepackage line and replace
-% \usepackage{icml2018} with \usepackage[nohyperref]{icml2018} above.
-\usepackage{hyperref}
-
-\input{macros.tex}
-
-% Use the following line for the initial blind version submitted for review:
-\usepackage{icml2018}
-
-% If accepted, instead use the following line for the camera-ready submission:
-%\usepackage[accepted]{icml2018}
-
-% The \icmltitle you define below is probably too long as a header.
-% Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Deep Relativistic Advantage Updating}
-
-\begin{document}
-
-\twocolumn[
-\icmltitle{There are no small Advantages:
-	Reinforcement learning in near continuous time
-	with Deep RelAtive Mixture of Advantages
-}
-
-% It is OKAY to include author information, even for blind
-% submissions: the style file will automatically remove it for you
-% unless you've provided the [accepted] option to the icml2018
-% package.
-
-% List of affiliations: The first argument should be a (short)
-% identifier you will use later to specify author affiliations
-% Academic affiliations should list Department, University, City, Region, Country
-% Industry affiliations should list Company, City, Region, Country
-
-% You can specify symbols, otherwise they are numbered in order.
-% Ideally, you should not use this facility. Affiliations will be numbered
-% in order of appearance and this is the preferred way.
-\icmlsetsymbol{equal}{*}
-
-\begin{icmlauthorlist}
-%\icmlauthor{Aeiau Zzzz}{equal,to}
-%\icmlauthor{Bauiu C.~Yyyy}{equal,to,goo}
-%\icmlauthor{Cieua Vvvvv}{goo}
-%\icmlauthor{Iaesut Saoeu}{ed}
-%\icmlauthor{Fiuea Rrrr}{to}
-%\icmlauthor{Tateu H.~Yasehe}{ed,to,goo}
-%\icmlauthor{Aaoeu Iasoh}{goo}
-%\icmlauthor{Buiui Eueu}{ed}
-%\icmlauthor{Aeuia Zzzz}{ed}
-%\icmlauthor{Bieea C.~Yyyy}{to,goo}
-%\icmlauthor{Teoau Xxxx}{ed}
-%\icmlauthor{Eee Pppp}{ed}
-\end{icmlauthorlist}
-
-%\icmlaffiliation{to}{Department of Computation, University of Torontoland, Torontoland, Canada}
-%\icmlaffiliation{goo}{Googol ShallowMind, New London, Michigan, USA}
-%\icmlaffiliation{ed}{School of Computation, University of Edenborrow, Edenborrow, United Kingdom}
-%
-%\icmlcorrespondingauthor{Cieua Vvvvv}{c.vvvvv@googol.com}
-%\icmlcorrespondingauthor{Eee Pppp}{ep@eden.co.uk}
-
-% You may provide any keywords that you
-% find helpful for describing your paper; these are used to populate
-% the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{Reinforcement learning}
-
-\vskip 0.3in
-]
-
-% this must go after the closing bracket ] following \twocolumn[
-
-% This command actually creates the footnote in the first column
-% listing the affiliations and the copyright notice.
-% The command takes one argument, which is text to display at the start of the footnote.
-% The \icmlEqualContribution command is standard text for equal contribution.
-% Remove it (just {}) if you do not need this facility.
-
-%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
-%\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
-
-\begin{abstract}
-	Despite remarkable successes, \emph{Deep Reinforcement learning} remains sensitive to
-	hyperparameterization, implementation details and small environment changes~(\citealt{drl_matter}, \citealt{drl_matter_bis}). 
-	Overcoming this sensitivity is key to making DRL applicable to real world problems.
-	In this paper, we study the sensitivity of Deep Reinforcement learning algorithms to
-	time discretization. More precisely, we show that approaches based on estimations of the
-	Q-function, e.g. \emph{Deep Q-learning}~\cite{dqn} and \emph{Deep deterministic policy gradient}~\cite{ddpg}
-	are sensitive to variations of time discretization. We further show that a simple reparameterization of the
-	Q-function as a sum of a state value term and a \emph{small} action dependent advantage term yields an algorithm
-	much more resilient to variations of time discretization.
-\end{abstract}
-
-\section{Introduction}
-\label{sec:intro}
-In recent years, deep learning based approaches to reinforcement learning have
-thrived and provided impressive results in a variety of domains, achieving superhuman
-performances with no expert knowledge in perfect information zero-sum
-games~\cite{alphazero}, reaching level of top players in video
-games~(\citealt{openai_five}, \citealt{dqn}), or learning dexterous manipulation
-from scratch without demonstrations~\cite{hand_control}.
-
-In spite of those successes, \emph{Deep reinforcement learning} approaches are
-sensitive to a number of factors, including hyperparameterization,
-implementation details or small changes in the environment
-parameters~(\citealt{drl_matter}, \citealt{drl_matter_bis}). This sensitivity,
-along with sample inefficiency, prevents DRL from being applied in most real
-world settings. High sensitivity to environment parameters notably prevents
-transfering abilities from imperfect simulators to real world scenarios.
-
-In this paper we focus our attention on the sensitivity to time discretization
-of DRL approaches, i.e. what happens when an agent receives $50$ observations
-and is expected to take $50$ actions per second instead of $10$. One expects
-that decreasing time discretization, or equivalently shortening reaction time,
-should only improve agent performances. This is not what happens in practice
-for approaches based on estimation of state action value functions, e.g.
-\emph{Deep Q-learning} (DQN~\citep{dqn} and \emph{Deep deterministic policy
-gradient} (DDPG~\citep{ddpg}). This is shown experimentally in Sec.~\ref{sec:exp}.
-
-We relate this sensitivity to the fact that, as the discretization timestep
-decreases, the effect of individual actions on the total return decreases, and
-can vanish when the discretization timestep becomes infinitesimal. The natural
-framework to study this phenomenon is Continuous time reinforcement learning 
-(Sec.~\ref{sec:continous}).
-
-To cope for this effect, we introduce a
-parameterization of the Q-function as a sum of a state value term and a
-\emph{small} action dependent advantage term (Sec.~\ref{subsec:reparam}). This
-reparameterized Q approximation can be trained with deep variants of Q-learning
-(Sec.~\ref{subsec:algorithm}).  To properly scale w.r.t. time discretization,
-particular attention has to be taken regarding both the exploration method, and
-the scaling of learning rates (Sec.~\ref{subsec:lr}, Sec.~\ref{subsec:explo}).
-The resulting algorithm is shown to provide near perfect invariance to time
-discretization on simple environments, and much better invariance properties
-than vanilla DQN or DDPG on more complex environments (Sec.~\ref{sec:exp}).
-
 \section{Related work}
@@
 the time discretization invariance properties.
-
-\section{Near continuous time reinforcement learning}
-\label{sec:continous}
-% In this section, a continuous reinforcement learning framework is introduced,
-% and it is shown that under this framework, the state-action value function does
-% not depend on the action. We further show that in near continous domains, when
-% the discretization timestep $\deltat$ is small, the difference in state-action
-% value between two different actions in a fixed state is of order $\bigO(\deltat)$.
-TODO. Main ideas: near continuous time environments (control, mujoco, video games, ...)
-There is a time discretization $\deltat$ given by the problem.
-The uderlying process does not depend of $\deltat$.
-Need to see what is happening when changing $\deltat$.
-The problem is invariant to $\deltat$.
-The quantities we are looking at should be invariant to.
-The learning procedure should be invariant.
-Looking at the limits when $\deltat \rightarrow 0$.
-In simulated environments, changing $\deltat$ only means to change the FPS.
-In real environments, for example with robots, this could mean changing the quality of the sensors and motors.
-
-\subsection{Framework}
-
-Let assume that we have a Markov Decision Process (TODO:Cite?) $\langle {\cal S}, {\cal A}, T_{\deltat}, r, \gamma\rangle$.
-${\cal S}$ is the state-space, which is supposed to be continuous (TODO:not the good formulation).
-The action space is TODO.
-
-We assume that the transition is continuous.
-This means that ...
-TODO.
-Discussion on the hypothesis, in which environments it is true.
-Is the condition $\forall \deltat, \forall s, \E[\|s_{k+1} - s_k \||s_k = s] \leq ?$.
-
-In what follows, $\dt$ denotes a truly infinitesimal time interval while $\deltat$
-denotes a small, but non infinitesimal time interval. Similarily, $\dbt$ denotes
-an infinitesimal (multi-dimensional) brownian step, while $\deltabt$ denotes its
-discretized equivalent. Informally, $\dbt$ is to be thought of as $\gauss(0, \sqrt{\dt}^2)$,
-and $\deltabt$ as $\gauss(0, \sqrt{\deltat}^2)$, where $\gauss(\mu, \sigma)$ denotes a gaussian
-random variable of mean $\mu$ and standard deviation (or covariance matrix in the multi-dimensional case)
-$\sigma$.
-
-Consider a fully observable \emph{Markov Decision Process} defined by the following
-transitions and rewards
-\begin{align}
-	ds(s, a) &= F(s, a) dt + \Sigma(s, a) \dbt\\
-	dr(s, a) &= \reward(s, a) dt + \Sigma_r(s, a) \dbt.
-\end{align}
-The equation on $s$ is that of a diffusion process, with drift coeficient $F$
-and diffusion matrix $\Sigma$.  This framework is only moderately restrictive.
-For example, deterministic continuous control environments fall in this
-framework, with $\Sigma = 0$ and $\Sigma_r = 0$. Limitations of the framework
-are discussed in Sec.~\ref{sec:limitations}.
-This framework is naturally discretized to arbitrary $\deltat$ by replacing all
-$d$'s with $\delta$'s.
-
-A continuous policy $\pi$ is defined as a deterministic mapping from states to
-actions.  In continuous time, defining stochastic policies with well defined
-state-value functions requires incorporating information from previous action into
-the state to obtain a temporally coherent policy. To avoid such inconvenience, the
-framework is restricted to deterministic policies. This is a mild constraint, as this
-only enforces determinism of the exploitation policy and not of exploration policies.
-In what follows, $\tau\sim\pi$ denotes a trajectory of states and actions sampled
-according to policy $\pi$, with $s_0$ sampled according to an arbitrary initial distribution
-on states $\rho_0$, and $dr_{t} = dr(s_t, a_t)$.
-
-Under a deterministic policy $\pi$, one can define the $\gamma$ discounted
-\emph{state value function} $V^\pi(s)$ and \emph{state action value function}
-$Q^\pi(s, a)$ as
-\begin{align}
-	V^\pi(s) &= \E_{\tau\sim\pi}\left[
-		\int\limits_{t=0}^\infty \gamma^{t}
-		dr_{t} \mid s_0 = s
-	\right]\\
-	Q^\pi(s, a) &= \E_{\tau\sim\pi}\left[
-		\int\limits_{t=0}^\infty \gamma^{t}
-		dr_{t} \mid s_0 = s, a_0 = a
-	\right]
-\end{align}
-which are naturally discretized as
-\begin{align}
-	V^\pi(s) &= \E_{\tau\sim\pi}\left[
-		\sum\limits_{k=0}^\infty \gamma^{k\deltat}
-		\delta r_{k\deltat} \mid s_0 = s
-	\right]\\
-	Q^\pi(s, a) &= \E_{\tau\sim\pi}\left[
-		\sum\limits_{k=0}^\infty \gamma^{k\deltat}
-		\delta r_{k\deltat} \mid s_0 = s, a_0 = a
-	\right].
-\end{align}
-Both $V^\pi$ and $Q^\pi$ verify Bellman equations, informally
-\begin{align}
-	V^\pi(s) &= \reward(s, \pi(s))\dt + \gamma^{\dt} \E_{ds \sim ds(s, \pi(s))} V^\pi(s + ds)\nonumber\\
-	Q^\pi(s, a) &= \reward(s, a)\dt + \gamma^{\dt} \E_{ds(s, a)} V^\pi(s + ds(s, a))
-	\label{eq:cont_bell}
-\end{align}
-which in continuous time translates, using Ito's lemma, into a Hamilton Jacobi equation
-\begin{align}
-	\reward + \partial_s V^\pi \cdot f + \frac{1}{2} \text{tr}\left(\Sigma^T\partial^2_{s^2} V^\pi\Sigma\right) = (1 - \gamma) V^\pi(s).
-\end{align}
-
-\subsection{Independence of $Q^\pi$ on $a$ in continuous time}
-Consider a discretization of Eq.~\eqref{eq:cont_bell} with time interval
-$\deltat \ll~1$, Ito's lemma yield
-\begin{align}
-	Q^\pi(s, a) &= \reward(s, a) \deltat + \gamma^{\deltat} \E_{\delta s \sim \delta s(s, a)}V^\pi(s + \delta s)\nonumber\\
-		    &= \reward(s, a) \deltat + \gamma^{\deltat} V^\pi(s) +
-		     \gamma^{\deltat} \partial_s V^\pi(s)f(s, a) \deltat\nonumber\\
-		    &+ \frac{1}{2}\text{tr}\left(\Sigma^T\partial^2_{s^2}V^\pi(s)\Sigma\right)\deltat + \bigO(\deltat^2)\nonumber\\
-		    &\xrightarrow[\deltat \to 0]{} V^\pi(s).
-		    \label{eq:Q_cont}
-\end{align}
-A consequence of Eq.~\eqref{eq:Q_cont} is that in exact continuous time,
-$Q^\pi$ does not bear any information on the hierarchy of actions, and
-thus cannot be used to select actions that yields higher future returns.
-
-As the difference between between $Q^\pi(s, a)$ and $V^\pi(s)$ is of order
-$\bigO(\deltat)$, it is natural to define a rescaled version of the advantage
-function, namely
-\begin{align}
-	A^\pi(s, a) &= \frac{Q^\pi(s,a) - V^\pi(s)}{\deltat}\\
-		    &= \reward(s, a) +
-		    \frac{\gamma^{\deltat} \E_{\delta s \sim \delta s(s, a)} V^\pi(s + \delta s) - V^\pi(s)}{\deltat}.
-    \label{eq:adv}
-\end{align}
-Contrary to the action state value function, the rescaled advantage function converges
-to an action dependent quantity when $\deltat$ goes to $0$
-\begin{align}
-	A^\pi(s, a) &\xrightarrow[\deltat \to 0]{} \reward(s, a) +
-	(\gamma - 1) V^\pi(s) + \partial_s V^\pi(s) f(s, a)\nonumber\\
-		    &+
-	\frac{\text{tr}\left(\Sigma^T(s, a)\partial^2_{s^2}V^\pi(s) \Sigma(s,a)\right)}{2},
-\end{align}
-and $Q^\pi$ can then be rewritten as
-\begin{equation}
-	Q^\pi(s, a) = V^\pi(s) + \deltat A^\pi(s, a).
-	\label{eq:reparam_q_pi}
-\end{equation}
-
-\input{advup.tex}
-
-\input{experiments.tex}
-
-\input{discussions.tex}
-
-\input{conclusion.tex}
-
-\bibliography{icml_drau.bib}
-\bibliographystyle{icml2018}
-\end{document}
-
-% This document was modified from the file originally made available by
-% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
-% by Iain Murray in 2018. It was modified from a version from Dan Roy in
-% 2017, which was based on a version from Lise Getoor and Tobias
-% Scheffer, which was slightly modified from the 2010 version by
-% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
-% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
-% slightly modified from Prasad Tadepalli's 2007 version which is a
-% lightly changed version of the previous year's version by Andrew
-% Moore, which was in turn edited from those of Kristian Kersting and
-% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2019-01-10 paper/icml_drau/discussions.tex
Headers.

%! TEX root = icml_drau.tex
\section{Discussions}
\label{sec:discussions}

TODO

### 2019-01-10 paper/icml_drau/advup.tex
sec:continuous => sec:framework

@@
-\section{Deep Advantage Updating}
-In continuous time, $Q^\pi$ is independant of actions and can thus not be used
-to select promising actions.  In near continuous time, $Q^\pi$ still depends on
-actions, and can still be used to choose promising actions under the current
-policy. However, when approximating $Q^\pi$, if the approximation error is much
-bigger than $\bigO(\deltat)$, approximation errors dominates, and the hierarchy
-of actions given by the approximation is likely to be erroneous.
-
+%! TEX root = icml_drau.tex
+\section{Reinforcement learning with a continuous time limit}
+As seen in Sec.~\ref{sec:framework}, there is no continuous time limit to
+Q-learning. In continuous time, $Q^\pi$ is independant of actions and can thus
+not be used to select promising actions.  In near continuous time, $Q^\pi$
+still depends on actions, and can still be used to choose promising actions
+under the current policy. However, when approximating $Q^\pi$, if the
+approximation error is much bigger than $\bigO(\deltat)$, approximation errors
+dominates, and the hierarchy of actions given by the approximation is likely to
+be erroneous.
 When the approximate Q-function is initialized, if the effect of actions on the
@@
 $Q^\pi$ is likely to be difficult. Besides, the error is likely to be further
-propagateed when the equation used to update our approximate $Q$ function
+propagated when the equation used to update our approximate $Q$ function
 relies on bootstraping, as is the case for usual temporal difference derived methods.

### 2019-01-10 paper/icml_drau/framework.tex
Headers.
sec:continuous => sec:framework
...

@@
 % value between two different actions in a fixed state is of order $\bigO(\deltat)$.
-TODO. Main ideas: near continuous time environments (control, mujoco, video games, ...)
+\TODO{. Main ideas: near continuous time environments (control, mujoco, video games, ...)
 There is a time discretization $\deltat$ given by the problem.
@@
 In simulated environments, changing $\deltat$ only means to change the FPS.
-In real environments, for example with robots, this could mean changing the quality of the sensors and motors.
+In real environments, for example with robots, this could mean changing the quality of the sensors and motors.}
 
@@
 
-Let assume that we have a Markov Decision Process (TODO:Cite?) $\langle {\cal S}, {\cal A}, T_{\deltat}, r, \gamma\rangle$.
-${\cal S}$ is the state-space, which is supposed to be continuous (TODO:not the good formulation).
-The action space is TODO.
+Let assume that we have a Markov Decision Process \TODO{Cite?} $\langle {\cal S}, {\cal A}, T_{\deltat}, r_\deltat, \gamma_\deltat\rangle$.
+${\cal S}$ is the state-space, which is supposed to be continuous \TODO{not the good formulation}.
+The action space is \TODO{todo}.
 
@@
 This means that ...
-TODO.
+\TODO{todo}.
 Discussion on the hypothesis, in which environments it is true.
 Is the condition $\forall \deltat, \forall s, \E[\|s_{k+1} - s_k \||s_k = s] \leq ?$.
+
+In many situations, we can assume that there is an uderlying process:
+\begin{equation}
+  \label{eq:diffusion}
+  ds = F(s, a) + \Sigma(s, a)\dbt
+\end{equation}
+$F$ is the physics, and $\dbt$ is the noise.
+
+There is convergence to the continuous process when $\deltat \rightarrow 0$.
+
+What we want is an algorithm invariant to $\deltat$, so we need to build a discrete algorithm such that when $\deltat \rightarrow 0$, the algorithm converge to a limit algorithm doing what we want.
+
+TODO: Explain $a_t = \pi(s_t)$ deterministic.
+
+\subsection{Return, $\gamma$ and Value function in near continuous time}
+\label{sec:ret-gamma-v}
+
+For a continuous MDP we define the return as:
+\begin{equation}
+  \label{eq:continuous-return}
+  R(\tau) = \int_{t=0}^\infty\gamma^tr(s_t, a_t)dt
+\end{equation}
+
+The natural discretization of this, if we want to define a discretize return $R_\delta$ for each $\deltat$ is:
+\begin{equation}
+  \label{eq:discretized-return}
+  R_\deltat(\tau) = \sum_{k=0}^\infty\gamma^{k\deltat}r(s^\deltat_k, a^\deltat_k)\deltat
+\end{equation}
 

### 2019-01-10 paper/icml_drau/icml_drau.tex
Headers.

@@
 
-\section{Introduction}
-\label{sec:intro}
-In recent years, deep learning based approaches to reinforcement learning have
-thrived and provided impressive results in a variety of domains, achieving superhuman
-performances with no expert knowledge in perfect information zero-sum
-games~\cite{alphazero}, reaching level of top players in video
-games~(\citealt{openai_five}, \citealt{dqn}), or learning dexterous manipulation
-from scratch without demonstrations~\cite{hand_control}.
+\input{intro.tex}
 
-In spite of those successes, \emph{Deep reinforcement learning} approaches are
-sensitive to a number of factors, including hyperparameterization,
-implementation details or small changes in the environment
-parameters~(\citealt{drl_matter}, \citealt{drl_matter_bis}). This sensitivity,
-along with sample inefficiency, prevents DRL from being applied in most real
-world settings. High sensitivity to environment parameters notably prevents
-transfering abilities from imperfect simulators to real world scenarios.
+\input{relatedworks.tex}
 
-In this paper we focus our attention on the sensitivity to time discretization
-of DRL approaches, i.e. what happens when an agent receives $50$ observations
-and is expected to take $50$ actions per second instead of $10$. One expects
-that decreasing time discretization, or equivalently shortening reaction time,
-should only improve agent performances. This is not what happens in practice
-for approaches based on estimation of state action value functions, e.g.
-\emph{Deep Q-learning} (DQN~\citep{dqn} and \emph{Deep deterministic policy
-gradient} (DDPG~\citep{ddpg}). This is shown experimentally in Sec.~\ref{sec:exp}.
-
-We relate this sensitivity to the fact that, as the discretization timestep
-decreases, the effect of individual actions on the total return decreases, and
-can vanish when the discretization timestep becomes infinitesimal. The natural
-framework to study this phenomenon is Continuous time reinforcement learning 
-(Sec.~\ref{sec:continous}).
-
-To cope for this effect, we introduce a
-parameterization of the Q-function as a sum of a state value term and a
-\emph{small} action dependent advantage term (Sec.~\ref{subsec:reparam}). This
-reparameterized Q approximation can be trained with deep variants of Q-learning
-(Sec.~\ref{subsec:algorithm}).  To properly scale w.r.t. time discretization,
-particular attention has to be taken regarding both the exploration method, and
-the scaling of learning rates (Sec.~\ref{subsec:lr}, Sec.~\ref{subsec:explo}).
-The resulting algorithm is shown to provide near perfect invariance to time
-discretization on simple environments, and much better invariance properties
-than vanilla DQN or DDPG on more complex environments (Sec.~\ref{sec:exp}).
-
-\section{Related work}
-\label{sec:related}
-Continuous and near continuous reinforcement learning, which is at the core of
-our approach has been thoroughly studied in the deterministic case
-in~\cite{adv_upd} or \cite{cont_rl}, but has not been applied in the context of
-deep reinforcement learning.
-
-The approach presented in this paper is closely related to \cite{adv_upd} and
-\cite{cont_rl}. \cite{adv_upd} already separated the learning procedure of the
-value and advantage functions, and introduced proper scalings between the two
-components. We extend on \cite{adv_upd} in three directions. We provide a
-formal argument on why the advantage contribution of the Q-function is of order
-$\bigO(\deltat)$, irrelevant of the stochasticity of the environment, in a quite
-general framework. We show that, provided with well suited learning rates and
-exploration method, Advantage updating provide near invariance to change of time
-discretization. We show that Advantage updating is viable in the context of deep
-reinforcement learning, while not using the true gradient of the residual error.
-
-More recently, \cite{dueling_nets} also introduced a parameterization separating
-the value and the advantage components of the Q-function. We additionally
-introduce a natural scaling coeficient between the two, that strongly improves
-the time discretization invariance properties.
-
-\section{Near continuous time reinforcement learning}
-\label{sec:continous}
-% In this section, a continuous reinforcement learning framework is introduced,
-% and it is shown that under this framework, the state-action value function does
-% not depend on the action. We further show that in near continous domains, when
-% the discretization timestep $\deltat$ is small, the difference in state-action
-% value between two different actions in a fixed state is of order $\bigO(\deltat)$.
-TODO. Main ideas: near continuous time environments (control, mujoco, video games, ...)
-There is a time discretization $\deltat$ given by the problem.
-The uderlying process does not depend of $\deltat$.
-Need to see what is happening when changing $\deltat$.
-The problem is invariant to $\deltat$.
-The quantities we are looking at should be invariant to.
-The learning procedure should be invariant.
-Looking at the limits when $\deltat \rightarrow 0$.
-In simulated environments, changing $\deltat$ only means to change the FPS.
-In real environments, for example with robots, this could mean changing the quality of the sensors and motors.
-
-\subsection{Framework}
-
-Let assume that we have a Markov Decision Process (TODO:Cite?) $\langle {\cal S}, {\cal A}, T_{\deltat}, r, \gamma\rangle$.
-${\cal S}$ is the state-space, which is supposed to be continuous (TODO:not the good formulation).
-The action space is TODO.
-
-We assume that the transition is continuous.
-This means that ...
-TODO.
-Discussion on the hypothesis, in which environments it is true.
-Is the condition $\forall \deltat, \forall s, \E[\|s_{k+1} - s_k \||s_k = s] \leq ?$.
-
-In what follows, $\dt$ denotes a truly infinitesimal time interval while $\deltat$
-denotes a small, but non infinitesimal time interval. Similarily, $\dbt$ denotes
-an infinitesimal (multi-dimensional) brownian step, while $\deltabt$ denotes its
-discretized equivalent. Informally, $\dbt$ is to be thought of as $\gauss(0, \sqrt{\dt}^2)$,
-and $\deltabt$ as $\gauss(0, \sqrt{\deltat}^2)$, where $\gauss(\mu, \sigma)$ denotes a gaussian
-random variable of mean $\mu$ and standard deviation (or covariance matrix in the multi-dimensional case)
-$\sigma$.
-
-Consider a fully observable \emph{Markov Decision Process} defined by the following
-transitions and rewards
-\begin{align}
-	ds(s, a) &= F(s, a) dt + \Sigma(s, a) \dbt\\
-	dr(s, a) &= \reward(s, a) dt + \Sigma_r(s, a) \dbt.
-\end{align}
-The equation on $s$ is that of a diffusion process, with drift coeficient $F$
-and diffusion matrix $\Sigma$.  This framework is only moderately restrictive.
-For example, deterministic continuous control environments fall in this
-framework, with $\Sigma = 0$ and $\Sigma_r = 0$. Limitations of the framework
-are discussed in Sec.~\ref{sec:limitations}.
-This framework is naturally discretized to arbitrary $\deltat$ by replacing all
-$d$'s with $\delta$'s.
-
-A continuous policy $\pi$ is defined as a deterministic mapping from states to
-actions.  In continuous time, defining stochastic policies with well defined
-state-value functions requires incorporating information from previous action into
-the state to obtain a temporally coherent policy. To avoid such inconvenience, the
-framework is restricted to deterministic policies. This is a mild constraint, as this
-only enforces determinism of the exploitation policy and not of exploration policies.
-In what follows, $\tau\sim\pi$ denotes a trajectory of states and actions sampled
-according to policy $\pi$, with $s_0$ sampled according to an arbitrary initial distribution
-on states $\rho_0$, and $dr_{t} = dr(s_t, a_t)$.
-
-Under a deterministic policy $\pi$, one can define the $\gamma$ discounted
-\emph{state value function} $V^\pi(s)$ and \emph{state action value function}
-$Q^\pi(s, a)$ as
-\begin{align}
-	V^\pi(s) &= \E_{\tau\sim\pi}\left[
-		\int\limits_{t=0}^\infty \gamma^{t}
-		dr_{t} \mid s_0 = s
-	\right]\\
-	Q^\pi(s, a) &= \E_{\tau\sim\pi}\left[
-		\int\limits_{t=0}^\infty \gamma^{t}
-		dr_{t} \mid s_0 = s, a_0 = a
-	\right]
-\end{align}
-which are naturally discretized as
-\begin{align}
-	V^\pi(s) &= \E_{\tau\sim\pi}\left[
-		\sum\limits_{k=0}^\infty \gamma^{k\deltat}
-		\delta r_{k\deltat} \mid s_0 = s
-	\right]\\
-	Q^\pi(s, a) &= \E_{\tau\sim\pi}\left[
-		\sum\limits_{k=0}^\infty \gamma^{k\deltat}
-		\delta r_{k\deltat} \mid s_0 = s, a_0 = a
-	\right].
-\end{align}
-Both $V^\pi$ and $Q^\pi$ verify Bellman equations, informally
-\begin{align}
-	V^\pi(s) &= \reward(s, \pi(s))\dt + \gamma^{\dt} \E_{ds \sim ds(s, \pi(s))} V^\pi(s + ds)\nonumber\\
-	Q^\pi(s, a) &= \reward(s, a)\dt + \gamma^{\dt} \E_{ds(s, a)} V^\pi(s + ds(s, a))
-	\label{eq:cont_bell}
-\end{align}
-which in continuous time translates, using Ito's lemma, into a Hamilton Jacobi equation
-\begin{align}
-	\reward + \partial_s V^\pi \cdot f + \frac{1}{2} \text{tr}\left(\Sigma^T\partial^2_{s^2} V^\pi\Sigma\right) = (1 - \gamma) V^\pi(s).
-\end{align}
-
-\subsection{Independence of $Q^\pi$ on $a$ in continuous time}
-Consider a discretization of Eq.~\eqref{eq:cont_bell} with time interval
-$\deltat \ll~1$, Ito's lemma yield
-\begin{align}
-	Q^\pi(s, a) &= \reward(s, a) \deltat + \gamma^{\deltat} \E_{\delta s \sim \delta s(s, a)}V^\pi(s + \delta s)\nonumber\\
-		    &= \reward(s, a) \deltat + \gamma^{\deltat} V^\pi(s) +
-		     \gamma^{\deltat} \partial_s V^\pi(s)f(s, a) \deltat\nonumber\\
-		    &+ \frac{1}{2}\text{tr}\left(\Sigma^T\partial^2_{s^2}V^\pi(s)\Sigma\right)\deltat + \bigO(\deltat^2)\nonumber\\
-		    &\xrightarrow[\deltat \to 0]{} V^\pi(s).
-		    \label{eq:Q_cont}
-\end{align}
-A consequence of Eq.~\eqref{eq:Q_cont} is that in exact continuous time,
-$Q^\pi$ does not bear any information on the hierarchy of actions, and
-thus cannot be used to select actions that yields higher future returns.
-
-As the difference between between $Q^\pi(s, a)$ and $V^\pi(s)$ is of order
-$\bigO(\deltat)$, it is natural to define a rescaled version of the advantage
-function, namely
-\begin{align}
-	A^\pi(s, a) &= \frac{Q^\pi(s,a) - V^\pi(s)}{\deltat}\\
-		    &= \reward(s, a) +
-		    \frac{\gamma^{\deltat} \E_{\delta s \sim \delta s(s, a)} V^\pi(s + \delta s) - V^\pi(s)}{\deltat}.
-    \label{eq:adv}
-\end{align}
-Contrary to the action state value function, the rescaled advantage function converges
-to an action dependent quantity when $\deltat$ goes to $0$
-\begin{align}
-	A^\pi(s, a) &\xrightarrow[\deltat \to 0]{} \reward(s, a) +
-	(\gamma - 1) V^\pi(s) + \partial_s V^\pi(s) f(s, a)\nonumber\\
-		    &+
-	\frac{\text{tr}\left(\Sigma^T(s, a)\partial^2_{s^2}V^\pi(s) \Sigma(s,a)\right)}{2},
-\end{align}
-and $Q^\pi$ can then be rewritten as
-\begin{equation}
-	Q^\pi(s, a) = V^\pi(s) + \deltat A^\pi(s, a).
-	\label{eq:reparam_q_pi}
-\end{equation}
+\input{framework.tex}
 

### 2019-01-10 paper/tangotd.tex

@@
 
+Normalization of $Z$. Since $Z=\sum \gamma^t P^t$, we have that
+$(1-\gamma)Z$ is a stochastic matrix (thanks Yann). This might help with
+normalization/detecting divergence. This translates as $\E_{s_2\sim \rho}
+Z(s_1,s_2)=\frac{1}{1-\gamma}$ which is testable/enforceable. However, I
+fear it might make the fixed point $Z(s_1,s_2)=\frac{1}{1-\gamma}$
+attractive instead of repulsive, and it might also perturb initialization
+(initializing $Z=\Id$ does not satisfy this normalization, so enforcing
+it too early is basically like initializing to
+$Z(s_1,s_2)=\frac{1}{1-\gamma}$ which empirically diverges).
+
 \paragraph{Early version of the above: Learning to navigate.} Define a function $V(s,s')$ that

### 2019-01-10 paper/icml_drau/advup.tex

@@
 $Q^\pi$ is likely to be difficult. Besides, the error is likely to be further
-propagated when the equation used to update our approximate $Q$ function
+propagateed when the equation used to update our approximate $Q$ function
 relies on bootstraping, as is the case for usual temporal difference derived methods.

### 2019-01-10 paper/icml_drau/advup.tex
Advances.

--- 2019-01-10 paper/icml_drau/advup.tex
@@
 Q-learning. In continuous time, $Q^\pi$ is independant of actions and can thus
-not be used to select promising actions.  In near continuous time, $Q^\pi$
+not be used to select promising actions.  In near continuous time, $Q^\pi_\deltat$
 still depends on actions, and can still be used to choose promising actions
-under the current policy. However, when approximating $Q^\pi$, if the
+under the current policy. However, when approximating $Q^\pi_\deltat$, if the
 approximation error is much bigger than $\bigO(\deltat)$, approximation errors
 dominates, and the hierarchy of actions given by the approximation is likely to
-be erroneous.
-When the approximate Q-function is initialized, if the effect of actions on the
-Q-function is order of magnitudes higher than what it should be, approximating
-$Q^\pi$ is likely to be difficult. Besides, the error is likely to be further
-propagated when the equation used to update our approximate $Q$ function
-relies on bootstraping, as is the case for usual temporal difference derived methods.
+be erroneous.  When the approximate Q-function is initialized, if the effect of
+actions on the Q-function is order of magnitudes higher than what it should be,
+approximating $Q^\pi_\deltat$ is likely to be difficult. Besides, the error is likely
+to be further propagated when the equation used to update our approximate $Q$
+function relies on bootstraping, as is the case for usual temporal difference
+derived methods.
 
-\subsection{Reparameterizing $Q_\theta$}
+In what follows, a discrete algorithm that has a well defined time continuous
+limit is defined. The algorithm relies on three elements, namely, defining and
+learning a quantity that contains some information on the hierarchy of actions
+while admitting a time continuous limit, defining exploration methods that
+admit a continuous time limit and defining learning rates that induce well
+behaved parameter trajectories when $\deltat$ goes to $0$.
+
+\subsection{Advantage Updating}
 \label{subsec:reparam}
-To prevent overestimating the dependency on actions of the approximation of
-$Q^\pi$, $Q_\theta$, in an environment with time discretization $\deltat$,
-Eq.~\eqref{eq:reparam_q_pi} naturally yields the following reparameterization
+To define an object which contains the same information on actions as
+$Q^\pi_\deltat$, but admits an action dependant continuous time limit, , it is
+natural to define
+\begin{align}
+	A^\pi_\deltat(s, a) &= \frac{Q^\pi_\deltat(s,a) - V^\pi_\deltat(s)}{\deltat},
+    \label{eq:adv}
+\end{align}
+a rescaled version of the advantage function, as the difference between between
+$Q^\pi_\deltat(s, a)$ and $V^\pi\deltat(s)$ is of order $\bigO(\deltat)$.
+Contrary to the action state value function, the rescaled advantage function converges
+to an action dependent quantity when $\deltat$ goes to $0$
+\begin{align}
+	A^\pi_\deltat(s, a)=&\reward(s, a)+\frac{\gamma^{\deltat}\E_{\delta s \sim \delta s(s, a)}V^\pi_\deltat(s + \delta s)-V^\pi\deltat(s)}{\deltat}\nonumber\\
+	=&\reward(s, a)+(\gamma - 1) V^\pi(s)+\partial_s V^\pi(s)f(s, a)\nonumber\\
+         &+\frac{\text{tr}\left(\Sigma^T(s, a)\partial^2_{s^2}V^\pi(s) \Sigma(s,a)\right)}{2} + O(\deltat)\nonumber.
+\end{align}
+The discretized state action value function can then be rewritten as
+\begin{equation}
+	Q^\pi_\deltat(s, a) = V^\pi_\deltat(s) + \deltat A^\pi_\deltat(s, a).
+	\label{eq:reparam_q_pi}
+\end{equation}
+A natural way of approximating $V^\pi_\delta$ and $A^\pi_\deltat$ is to apply
+\emph{Sarsa} or Q-learning to a reparameterized Q-function approximator
 \begin{equation}
@@
 To avoid cumbersome notations, in what follows, parameter indices are dropped.
-The insight behind this parameterization is that when $Q$ correctly
-approximates $Q^\pi$, $V$ should approximate the value function, while $A$
-should approximate the advantage function, as defined in Eq.~\eqref{eq:adv}.
+At initialization, if both $V_{\theta_1}$ and $A_{\theta_2}$ are initialized
+independantly of $\deltat$, this parameterization reasonably scales the
+contribution of actions relative to states in the value of $Q$.
 
-The proposed reparameterization does not, on its own, guarantee that
-when $Q$ correctly approximates $Q^\pi$, $A$ approximates $A^\pi$.  Indeed, for
-any given pair $V$, $A$, the transformed pair
+However, the proposed reparameterization does not, on its own, guarantee that
+when $Q$ correctly approximates $Q^\pi_\deltat$, $A$ approximates $A^\pi_\deltat$.
+Indeed, for any given pair $V$, $A$,
 \begin{align}
@@
 $Q^\pi$ even if $A$ is arbitrarily far from $A^\pi$.
-To enforce, identifiability of $A$, one must enforce that 
+To enforce, identifiability of $A$, one must enforce the consistency equation
+\begin{equation}
+	Q^\pi_\deltat(s, \pi(s)) - V^\pi_\deltat(s) = 0
+\end{equation}
+on the approximate $A$ and $V$. This translates to
 \begin{equation}
@@
 \end{equation}
-With this additional constraint, if $Q^\pi = Q$,
+With this additional constraint, if $Q^\pi_\deltat = Q$,
 \begin{align}
-	A^\pi(s, a) &= \frac{Q^\pi(s,a) - V^\pi(s)}{\deltat}\\
+	A^\pi_\deltat(s, a) &= \frac{Q^\pi_\deltat(s,a) - V^\pi_\deltat(s)}{\deltat}\\
 		    &= \frac{Q(s,a) - Q(s, \pi(s))}{\deltat}\\

### 2019-01-10 paper/icml_drau/framework.tex
Advances.

@@
+%! TEX root = icml_drau.tex
 \section{Near continuous time reinforcement learning}
-\label{sec:continous}
+\label{sec:framework}
 % In this section, a continuous reinforcement learning framework is introduced,
@@
 thus cannot be used to select actions that yields higher future returns.
-
-As the difference between between $Q^\pi(s, a)$ and $V^\pi(s)$ is of order
-$\bigO(\deltat)$, it is natural to define a rescaled version of the advantage
-function, namely
-\begin{align}
-	A^\pi(s, a) &= \frac{Q^\pi(s,a) - V^\pi(s)}{\deltat}\\
-		    &= \reward(s, a) +
-		    \frac{\gamma^{\deltat} \E_{\delta s \sim \delta s(s, a)} V^\pi(s + \delta s) - V^\pi(s)}{\deltat}.
-    \label{eq:adv}
-\end{align}
-Contrary to the action state value function, the rescaled advantage function converges
-to an action dependent quantity when $\deltat$ goes to $0$
-\begin{align}
-	A^\pi(s, a) &\xrightarrow[\deltat \to 0]{} \reward(s, a) +
-	(\gamma - 1) V^\pi(s) + \partial_s V^\pi(s) f(s, a)\nonumber\\
-		    &+
-	\frac{\text{tr}\left(\Sigma^T(s, a)\partial^2_{s^2}V^\pi(s) \Sigma(s,a)\right)}{2},
-\end{align}
-and $Q^\pi$ can then be rewritten as
-\begin{equation}
-	Q^\pi(s, a) = V^\pi(s) + \deltat A^\pi(s, a).
-	\label{eq:reparam_q_pi}
-\end{equation}

### 2019-01-10 paper/icml_drau/framework.tex
very first version of the framework part, lots of thing missing but the main structure is there

@@
 \section{Near continuous time reinforcement learning}
-\label{sec:framework}
+\label{sec:continous}
 % In this section, a continuous reinforcement learning framework is introduced,
@@
 
-In what follows, $\dt$ denotes a truly infinitesimal time interval while $\deltat$
-denotes a small, but non infinitesimal time interval. Similarily, $\dbt$ denotes
-an infinitesimal (multi-dimensional) brownian step, while $\deltabt$ denotes its
-discretized equivalent. Informally, $\dbt$ is to be thought of as $\gauss(0, \sqrt{\dt}^2)$,
-and $\deltabt$ as $\gauss(0, \sqrt{\deltat}^2)$, where $\gauss(\mu, \sigma)$ denotes a gaussian
-random variable of mean $\mu$ and standard deviation (or covariance matrix in the multi-dimensional case)
-$\sigma$.
+We see that the good \TODO{rewrite} definitions of $\gamma_\deltat$ and $r_\deltat$ are:
+\begin{align}
+  \gamma_\deltat =& \gamma^\deltat\\
+  r_\deltat =& \deltat . r
+\end{align}
+Thus, we see that the natural discretization of a continuous MDP ${\cal M} = \langle {\cal S}, {\cal A}, T, r, \gamma\rangle$ is ${\cal M}_\deltat = \langle {\cal S}, {\cal A}, T_\deltat, r\deltat, \gamma^\deltat\rangle$.
 
-Consider a fully observable \emph{Markov Decision Process} defined by the following
-transitions and rewards
-\begin{align}
-	ds(s, a) &= F(s, a) dt + \Sigma(s, a) \dbt\\
-	dr(s, a) &= \reward(s, a) dt + \Sigma_r(s, a) \dbt.
-\end{align}
-The equation on $s$ is that of a diffusion process, with drift coeficient $F$
-and diffusion matrix $\Sigma$.  This framework is only moderately restrictive.
-For example, deterministic continuous control environments fall in this
-framework, with $\Sigma = 0$ and $\Sigma_r = 0$. Limitations of the framework
-are discussed in Sec.~\ref{sec:limitations}.
-This framework is naturally discretized to arbitrary $\deltat$ by replacing all
-$d$'s with $\delta$'s.
+Thus, we should define $\gamma_\deltat = \gamma^\deltat$, so we have $\lim_{\deltat \rightarrow 0} R_\deltat(\tau)  = R(\tau)$. \TODO{Should we state that more formally?}
 
-A continuous policy $\pi$ is defined as a deterministic mapping from states to
-actions.  In continuous time, defining stochastic policies with well defined
-state-value functions requires incorporating information from previous action into
-the state to obtain a temporally coherent policy. To avoid such inconvenience, the
-framework is restricted to deterministic policies. This is a mild constraint, as this
-only enforces determinism of the exploitation policy and not of exploration policies.
-In what follows, $\tau\sim\pi$ denotes a trajectory of states and actions sampled
-according to policy $\pi$, with $s_0$ sampled according to an arbitrary initial distribution
-on states $\rho_0$, and $dr_{t} = dr(s_t, a_t)$.
+This is consistent with what we expect for the time horizon.
+We know \TODO{Cite?} that for a MDP with discount factor $\gamma$, the time horizon is $\frac{1}{1-\gamma}$ steps.
+But if the duration of a step is $\deltat$, this means that the time horizon in \emph{physical time} (independant of $\deltat$) is $\frac{\deltat}{1-\gamma}$, which goes to 0 if $\deltat \rightarrow 0$.
+This means that if $\gamma$ is not adapted to $\deltat$, the return $R_\deltat(\tau) = \sum_{k=0}^\infty\gamma^{k}r(s^\deltat_k, a^\deltat_k)\deltat \rightarrow_{\deltat\rightarrow 0} r(s_0, a_0)$.
+\TODO{A changer pour ne pas ecrire de chose fausse} But in practice, for example on a control problem, we do not care about the time horizon as a number of steps but as a \emph{physical quantity}, independant of $\deltat$.
+With $\gamma_\deltat = \gamma^\deltat$, the time horizon in \emph{physical time} is:
+\begin{equation}
+  \label{eq:time-horizon}
+  \frac{\deltat}{1-\gamma^\deltat} \rightarrow_{\deltat \rightarrow 0} - \frac{1}{\log \gamma} \approx \frac{1}{1-\gamma}
+\end{equation}
+\TODO{Explain again that we have a limit so it is better}
 
-Under a deterministic policy $\pi$, one can define the $\gamma$ discounted
-\emph{state value function} $V^\pi(s)$ and \emph{state action value function}
-$Q^\pi(s, a)$ as
-\begin{align}
-	V^\pi(s) &= \E_{\tau\sim\pi}\left[
-		\int\limits_{t=0}^\infty \gamma^{t}
-		dr_{t} \mid s_0 = s
-	\right]\\
-	Q^\pi(s, a) &= \E_{\tau\sim\pi}\left[
-		\int\limits_{t=0}^\infty \gamma^{t}
-		dr_{t} \mid s_0 = s, a_0 = a
-	\right]
-\end{align}
-which are naturally discretized as
-\begin{align}
-	V^\pi(s) &= \E_{\tau\sim\pi}\left[
-		\sum\limits_{k=0}^\infty \gamma^{k\deltat}
-		\delta r_{k\deltat} \mid s_0 = s
-	\right]\\
-	Q^\pi(s, a) &= \E_{\tau\sim\pi}\left[
-		\sum\limits_{k=0}^\infty \gamma^{k\deltat}
-		\delta r_{k\deltat} \mid s_0 = s, a_0 = a
-	\right].
-\end{align}
-Both $V^\pi$ and $Q^\pi$ verify Bellman equations, informally
-\begin{align}
-	V^\pi(s) &= \reward(s, \pi(s))\dt + \gamma^{\dt} \E_{ds \sim ds(s, \pi(s))} V^\pi(s + ds)\nonumber\\
-	Q^\pi(s, a) &= \reward(s, a)\dt + \gamma^{\dt} \E_{ds(s, a)} V^\pi(s + ds(s, a))
-	\label{eq:cont_bell}
-\end{align}
-which in continuous time translates, using Ito's lemma, into a Hamilton Jacobi equation
-\begin{align}
-	\reward + \partial_s V^\pi \cdot f + \frac{1}{2} \text{tr}\left(\Sigma^T\partial^2_{s^2} V^\pi\Sigma\right) = (1 - \gamma) V^\pi(s).
-\end{align}
+The Value Function for a policy $\pi$ with a discretization $\deltat$ is
+\begin{equation}
+  V^\pi_\deltat(s) = \E_{\tau\sim\pi}\left[\sum\limits_{k=0}^\infty \gamma^{k\deltat}\deltat r_{k\deltat} \mid s_0 = s\right]
+\end{equation}
+and we have, for all $\sigma \in {\cal S}$, $\lim_{\deltat \rightarrow 0} V^\pi_\deltat(s) = V^\pi(s)$. \TODO{Theorem?}
 
-\subsection{Independence of $Q^\pi$ on $a$ in continuous time}
-Consider a discretization of Eq.~\eqref{eq:cont_bell} with time interval
-$\deltat \ll~1$, Ito's lemma yield
-\begin{align}
-	Q^\pi(s, a) &= \reward(s, a) \deltat + \gamma^{\deltat} \E_{\delta s \sim \delta s(s, a)}V^\pi(s + \delta s)\nonumber\\
-		    &= \reward(s, a) \deltat + \gamma^{\deltat} V^\pi(s) +
-		     \gamma^{\deltat} \partial_s V^\pi(s)f(s, a) \deltat\nonumber\\
-		    &+ \frac{1}{2}\text{tr}\left(\Sigma^T\partial^2_{s^2}V^\pi(s)\Sigma\right)\deltat + \bigO(\deltat^2)\nonumber\\
-		    &\xrightarrow[\deltat \to 0]{} V^\pi(s).
-		    \label{eq:Q_cont}
-\end{align}
-A consequence of Eq.~\eqref{eq:Q_cont} is that in exact continuous time,
+It follows the Bellman equation:
+\begin{equation}
+  \label{eq:bellman}
+  V^\pi_\deltat(s) = \reward(s, \pi(s))\deltat + \gamma^{\dt} \E_{s_{t+\deltat} | s_t = s} V^\pi(s_{t+\deltat})
+\end{equation}
+
+If the MDP is following the equation (\ref{eq:diffusion}), rhe limit of the Bellman equation for $V^\pi_\deltat$ when $\deltat \rightarrow 0$ is the Hamilton-Jacobi-Bellman equation on $V^\pi$:
+\begin{equation}
+  \label{eq:hamilton-jacobi-bellman}
+  r + \partial_s V^\pi \cdot F + \frac{1}{2} \text{tr}\left(\Sigma^T\partial^2_{s^2} V^\pi\Sigma\right) = (1 - \gamma) V^\pi(s)
+\end{equation}
+
+% In what follows, $\dt$ denotes a truly infinitesimal time interval while $\deltat$
+% denotes a small, but non infinitesimal time interval. Similarily, $\dbt$ denotes
+% an infinitesimal (multi-dimensional) brownian step, while $\deltabt$ denotes its
+% discretized equivalent. Informally, $\dbt$ is to be thought of as $\gauss(0, \sqrt{\dt}^2)$,
+% and $\deltabt$ as $\gauss(0, \sqrt{\deltat}^2)$, where $\gauss(\mu, \sigma)$ denotes a gaussian
+% random variable of mean $\mu$ and standard deviation (or covariance matrix in the multi-dimensional case)
+% $\sigma$.
+
+% Consider a fully observable \emph{Markov Decision Process} defined by the following
+% transitions and rewards
+% \begin{align}
+% 	ds(s, a) &= F(s, a) dt + \Sigma(s, a) \dbt\\
+% 	dr(s, a) &= \reward(s, a) dt + \Sigma_r(s, a) \dbt.
+% \end{align}
+% The equation on $s$ is that of a diffusion process, with drift coeficient $F$
+% and diffusion matrix $\Sigma$.  This framework is only moderately restrictive.
+% For example, deterministic continuous control environments fall in this
+% framework, with $\Sigma = 0$ and $\Sigma_r = 0$. Limitations of the framework
+% are discussed in Sec.~\ref{sec:limitations}.
+% This framework is naturally discretized to arbitrary $\deltat$ by replacing all
+% $d$'s with $\delta$'s.
+
+% A continuous policy $\pi$ is defined as a deterministic mapping from states to
+% actions.  In continuous time, defining stochastic policies with well defined
+% state-value functions requires incorporating information from previous action into
+% the state to obtain a temporally coherent policy. To avoid such inconvenience, the
+% framework is restricted to deterministic policies. This is a mild constraint, as this
+% only enforces determinism of the exploitation policy and not of exploration policies.
+% In what follows, $\tau\sim\pi$ denotes a trajectory of states and actions sampled
+% according to policy $\pi$, with $s_0$ sampled according to an arbitrary initial distribution
+% on states $\rho_0$, and $dr_{t} = dr(s_t, a_t)$.
+
+% Under a deterministic policy $\pi$, one can define the $\gamma$ discounted
+% \emph{state value function} $V^\pi(s)$ and \emph{state action value function}
+% $Q^\pi(s, a)$ as
+% \begin{align}
+% 	V^\pi(s) &= \E_{\tau\sim\pi}\left[
+% 		\int\limits_{t=0}^\infty \gamma^{t}
+% 		dr_{t} \mid s_0 = s
+% 	\right]\\
+% 	Q^\pi(s, a) &= \E_{\tau\sim\pi}\left[
+% 		\int\limits_{t=0}^\infty \gamma^{t}
+% 		dr_{t} \mid s_0 = s, a_0 = a
+% 	\right]
+% \end{align}
+% which are naturally discretized as
+% \begin{align}
+% 	V^\pi(s) &= \E_{\tau\sim\pi}\left[
+% 		\sum\limits_{k=0}^\infty \gamma^{k\deltat}
+% 		\delta r_{k\deltat} \mid s_0 = s
+% 	\right]\\
+% 	Q^\pi(s, a) &= \E_{\tau\sim\pi}\left[
+% 		\sum\limits_{k=0}^\infty \gamma^{k\deltat}
+% 		\delta r_{k\deltat} \mid s_0 = s, a_0 = a
+% 	\right].
+% \end{align}
+% Both $V^\pi$ and $Q^\pi$ verify Bellman equations, informally
+% \begin{align}
+% 	V^\pi(s) &= \reward(s, \pi(s))\dt + \gamma^{\dt} \E_{ds \sim ds(s, \pi(s))} V^\pi(s + ds)\nonumber\\
+% 	Q^\pi(s, a) &= \reward(s, a)\dt + \gamma^{\dt} \E_{ds(s, a)} V^\pi(s + ds(s, a))
+% 	\label{eq:cont_bell}
+% \end{align}
+% which in continuous time translates, using Ito's lemma, into a Hamilton Jacobi equation
+% \begin{align}
+% 	\reward + \partial_s V^\pi \cdot f + \frac{1}{2} \text{tr}\left(\Sigma^T\partial^2_{s^2} V^\pi\Sigma\right) = (1 - \gamma) V^\pi(s).
+% \end{align}
+
+\subsection{There is no Q-function in continuous time}
+
+Contrary to the Value function, the Action-Value function $Q$ is ill-defined for a continuous MDP. Moreover, this means that the discretized $Q$ function \TODO{explain}. This can be stated formally.
+\begin{theorem}
+  For all $s, a$, we have
+  \begin{equation}
+    \label{eq:q-discr}
+    Q^\pi_\deltat(s, a) = V^\pi(s) + \bigO\left(\sqrt\deltat\right)
+  \end{equation}
+
+  If we consider the continuous action-value function $Q^\pi$, we have:
+  \begin{equation}
+    \label{eq:q-cont}
+    Q^\pi(s, a) = V^\pi
+  \end{equation}
+  \label{th:q-cont}
+\end{theorem}
+% Consider a discretization of Eq.~\eqref{eq:cont_bell} with time interval
+% $\deltat \ll~1$, Ito's lemma yield
+% \begin{align}
+% 	Q^\pi(s, a) &= \reward(s, a) \deltat + \gamma^{\deltat} \E_{\delta s \sim \delta s(s, a)}V^\pi(s + \delta s)\nonumber\\
+% 		    &= \reward(s, a) \deltat + \gamma^{\deltat} V^\pi(s) +
+% 		     \gamma^{\deltat} \partial_s V^\pi(s)f(s, a) \deltat\nonumber\\
+% 		    &+ \frac{1}{2}\text{tr}\left(\Sigma^T\partial^2_{s^2}V^\pi(s)\Sigma\right)\deltat + \bigO(\deltat^2)\nonumber\\
+% 		    &\xrightarrow[\deltat \to 0]{} V^\pi(s).
+% 		    \label{eq:Q_cont}
+                      %   \end{align}
+
+A consequence of Theorem \ref{th:q-cont} is that in exact continuous time,
 $Q^\pi$ does not bear any information on the hierarchy of actions, and
 thus cannot be used to select actions that yields higher future returns.
+
+\begin{proof}
+    \TODO{TODO}
+  \end{proof}

### 2019-01-10 paper/tangotd.tex

--- 2018-12-21 paper/tangotd.tex
@@
 point, I think, so initializing to $\frac{1}{1-\gamma}$ plus noise should
-converge to the true solution. Initializing there should have the
+converge to the true solution [empirically I find this diverges]. Initializing there should have the
 advantage of starting close to reasonable values.
@@
 
+Normalization of $Z$. Since $Z=\sum \gamma^t P^t$, we have that
+$(1-\gamma)Z$ is a stochastic matrix (thanks Yann). This might help with
+normalization/detecting divergence. This translates as $\E_{s_2\sim \rho}
+Z(s_1,s_2)=\frac{1}{1-\gamma}$ which is testable/enforceable. However, I
+fear it might make the fixed point $Z(s_1,s_2)=\frac{1}{1-\gamma}$
+attractive instead of repulsive, and it might also perturb initialization
+(initializing $Z=\Id$ does not satisfy this normalization, so enforcing
+it too early is basically like initializing to
+$Z=\frac{1}{1-\gamma}\Id$ which empirically diverges).
+
 \paragraph{Early version of the above: Learning to navigate.} Define a function $V(s,s')$ that

### 2019-01-11 paper/icml_drau/advup.tex
Algorithm description.

@@
 \end{equation}
+With this parameterization, one can hope to learn an approximation of $A^\pi_\deltat$,
+from which a proper hiearchy of actions can be derived for any $\deltat$.
 
@@
 To obtain a time invariant reinforcement learning algorithm, a time invariant
-exploration scheme is required.
+exploration scheme is required. All the exploration schemes hereby considered
+generate sequences of actions as
+\begin{equation}
+	a^k_\deltat = \pi^\text{explore}_\deltat(s^k_\deltat, \nu^k_\deltat)
+\end{equation}
+with $(\nu^k_\deltat)_k$ a sequence of random variables, independant from the $a$'s and $s$'s.
+For instance, $\varepsilon$-greedy exploration naturally falls in this framework, with each $\nu^k_\deltat$
+being a pair composed of a bernoulli random variables $b^k_\deltat$ with probability of being $1$ equal to $\varepsilon$,
+and of a uniform categorical random variable $\xi^k_\deltat$ on actions. Exploratory actions are then selected as
+\begin{equation}
+	a^k_\deltat = (1 - b^k_\deltat) \text{argmax}_{a'} Q^\pi_\deltat(s, a') + b^k_\deltat \xi^k_\deltat.
+\end{equation}
+
+A sufficient condition for the exploratory policy to admit a time continuous
+limit is for $\pi^\text{explore}_\deltat$ to be a measurable function, and for
+$\nu_\deltat$ to admit one, i.e. that the sequence $\nu_\deltat$ converges to a
+well defined continuous stochastic process $\nu$ as $\deltat$ goes to $0$.
+\TODO{I'm convinced that this is true, reviewers might not be.}
 
 The exploration scheme used~\cite{ddpg}, which consists in adding an
-\emph{Ornstein-Uhlenbeck}~\cite{orn-uhl} process to actions, is naturally time
-invariant in the sense that it is first defined as a time continuous stochastic
-process, and that discretization of this process are used in practice. Indeed,
-a continuous Ornstein-Uhlenbeck process is defined as
+\emph{Ornstein-Uhlenbeck}~\cite{orn-uhl} process (OU process) to actions, is naturally time.
+invariant. It can be rewritten as
 \begin{equation}
-	dx_t = - (x_t - \mu) \theta dt + \sigma \dbt
+	a^k_\deltat = \pi(s^k_\deltat) + \nu^k_\deltat
+\end{equation}
+where $\nu^k_\deltat$ is the discretization of a time continuous \emph{Ornstein-Uhlenbeck}
+process, defined as
+\begin{equation}
+	d\nu_t = - \nu_t \theta dt + \sigma \dbt.
 	\label{eq:orn_uhl}
 \end{equation}
-and is naturally discretized by turning $d$'s into $\delta$'s, and using
-the same $\deltat$ for both the learning algorithm and the exploration scheme.
-Consequently, this exploration scheme can readily be used in continuous action
-environments.
 
-On the other hand, the usual exploration scheme used in discrete actions
-Q-learning, $\varepsilon$-greedy exploration~\cite{sutton} is not time
-invariant, in the sense that resulting state trajectories do not converge to a
-well defined stochastic process as $\deltat \rightarrow 0$. As $\dt$ goes to $0$,
-for any $\varepsilon$ independant of $\dt$, the behavior of state trajectories get
-closer to that of a white noise.
+On the other hand, the same argument cannot be applied to $\varepsilon$-greedy
+exploration, as the $\nu_\deltat$'s used to define it do not converge to a 
+well defined stochastic process, but to a white noise.\footnote{\TODO{$\varepsilon$-greedy
+		does not, in general, converges to a well defined exploration scheme in continuous time.
+		For instance, if actions directly affects position, $\varepsilon$-greedy results in white-noise
+trajectories in state space. Show this in more detail in appendix ?}}
 
 To obtain a temporally coherent exploration scheme in discrete action environment
-when using Q-learning we instead directly apply a discretization of a time continuous
-stochastic process to the output of the approximate Q-function, and select actions
-by maximizing the resulting perturbed Q-function. Typically, the approximate Q-function
-values are also perturbed using a discretized Ornstein-Uhlenbeck noise.
+$\nu_\deltat$ is chosen to be a discretization of a $|\mathcal{A}|$ dimensional
+continuous OU process, and set
+\begin{equation}
+	a^k_\deltat = \text{argmax}_{a'} \left(Q^\pi_\deltat(s, a') + \nu_\deltat^k[a']\right)
+\end{equation}
+where $\nu_\deltat^k[a']$ denotes the $a'$-th component of the discretized OU process at time
+$k$. The resulting exploration scheme converges to a proper time continuous exploration scheme.
 
@@
 \label{subsec:algorithm}
-Two variants of \emph{Deep Advantage Updating} are derived to cover both the continuous
-and discrete action cases. They both rely on variants of Q-learning. More precisely,
-we learn $V$ and $A$ using approximate version of the following bellman equation
+To learn $V_{\theta_1}$ and $A_{\theta_2}$, variants of Q-learning for continuous~\cite{ddpg} and
+discrete~\cite{dqn} action spaces are used. In details, $V_{\theta_1}$ and $A_{\theta_2}$ are learnt
+using approximate version of the following bellman equation
 \begin{align}
@@
 \end{align}
-for a greedy exploitation policy $\pi(s) = \text{argmax}_{a'}A^\pi(s, a')$.
-Eq.~\eqref{eq:max_A} is directly verified owing to the parameterization of $A$.
-To approximately verify~\eqref{eq:bellman_A}, the corresponding squared residual
-is minimized by an approximate gradient descent.
+for a greedy exploitation policy $\pi(s) = \text{argmax}_{a'}A^\pi(s, a')$,
+exactly implemented in the case of discrete actions, and approximated by a
+neural network $\pi_\phi(s)$, trained to maximize (an approximation of)
+$A^\pi(s, \pi_\phi(s))$ in the continuous case.  Eq.~\eqref{eq:max_A} is
+directly verified by $A_{\theta_1}$, owing to the reparameterization described
+in~\ref{subsec:reparam}.  To approximately verify~\eqref{eq:bellman_A}, the
+corresponding squared residual is minimized by an approximate gradient descent.
 The corresponding update equations when learning from incoming transitions of
@@
 \begin{align}
-	\delta Q_t &= A(s_t, a_t) - \frac{\delta r_t + \gamma^{\deltat} V(s_{t+\deltat}) - V(s_t)}{\deltat}\\
+	\delta Q_t &= A(s_t, a_t) - \frac{\delta r_t + \gamma^{\deltat} V(s_{t}+\delta s_{t}) - V(s_t)}{\deltat}\\
 	\theta^1_{t+\deltat} &= \theta^1_t + \eta_1 \partial_{\theta^1} V(s_t) \delta Q_t\\
@@
 \end{align}
-where $\eta$'s are learning rates. We are now left to derive appropriate learning rates
-to provide invariance to time discretization.
+where $\eta$'s are learning rates. In practice, $A$ and $V$ are learned using
+transitions sampled from a replay buffer instead of trajectory samples, to
+prevent excessive correlation between training samples~\cite{dqn}. Pseudocode
+for the discrete actions algorithm is given in Alg.~\ref{alg:dau}. Pseudocode
+for the continuous actions algorithm is given in the supplementary material.
 \begin{algorithm}[ht]
@@
 	\input{dauc.tex}
+	\label{alg:dau}
 \end{algorithm}

### 2019-01-11 paper/icml_drau/advup.tex
Algorithm section nearly finished.

@@
 	A^\pi_\deltat(s, a)=&\reward(s, a)+\frac{\gamma^{\deltat}\E_{\delta s \sim \delta s(s, a)}V^\pi_\deltat(s + \delta s)-V^\pi\deltat(s)}{\deltat}\nonumber\\
-	=&\reward(s, a)+(\gamma - 1) V^\pi(s)+\partial_s V^\pi(s)f(s, a)\nonumber\\
+	=&\reward(s, a)+\log(\gamma) V^\pi(s)+\partial_s V^\pi(s)f(s, a)\nonumber\\
          &+\frac{\text{tr}\left(\Sigma^T(s, a)\partial^2_{s^2}V^\pi(s) \Sigma(s,a)\right)}{2} + O(\deltat)\nonumber.
@@
 \begin{equation}
-	Q_\theta(s, a) = V_{\theta_1}(s) + \deltat A_{\theta_2}(s, a).
+	Q_\Theta(s, a) = V_{\theta}(s) + \deltat A_{\psi}(s, a).
 \end{equation}
 To avoid cumbersome notations, in what follows, parameter indices are dropped.
-At initialization, if both $V_{\theta_1}$ and $A_{\theta_2}$ are initialized
+At initialization, if both $V_{\theta}$ and $A_{\psi}$ are initialized
 independantly of $\deltat$, this parameterization reasonably scales the
@@
 \label{subsec:algorithm}
-To learn $V_{\theta_1}$ and $A_{\theta_2}$, variants of Q-learning for continuous~\cite{ddpg} and
-discrete~\cite{dqn} action spaces are used. In details, $V_{\theta_1}$ and $A_{\theta_2}$ are learnt
+To learn $V_{\theta}$ and $A_{\psi}$, variants of Q-learning for continuous~\cite{ddpg} and
+discrete~\cite{dqn} action spaces are used. In details, $V_{\theta}$ and $A_{\psi}$ are learnt
 using approximate version of the following bellman equation
 \begin{align}
-	V^{\pi}(s) + A^{\pi}(s, a) &= \E_{\delta r \sim \delta r(s, a)}\delta r\nonumber\\ 
-				   &+ \gamma^{\deltat}  \E_{\delta s \sim \delta s(s, a)} V^{\pi}(s + \delta s)\label{eq:bellman_A}\\
-	A^{\pi}(s, \pi(s)) &= 0\label{eq:max_A}
+	V^{\pi}_\deltat(s) + A^{\pi}_\deltat(s, a) &= \E_{\delta r \sim \delta r(s, a)}\delta r\nonumber\\ 
+				   &+ \gamma^{\deltat}  \E_{\delta s \sim \delta s(s, a)} V^{\pi}_\deltat(s + \delta s)\label{eq:bellman_A}\\
+	A^{\pi}_\deltat(s, \pi(s)) &= 0\label{eq:max_A}
 \end{align}
-for a greedy exploitation policy $\pi(s) = \text{argmax}_{a'}A^\pi(s, a')$,
+for a greedy exploitation policy $\pi(s) = \text{argmax}_{a'}A^\pi_\deltat(s, a')$,
 exactly implemented in the case of discrete actions, and approximated by a
 neural network $\pi_\phi(s)$, trained to maximize (an approximation of)
-$A^\pi(s, \pi_\phi(s))$ in the continuous case.  Eq.~\eqref{eq:max_A} is
-directly verified by $A_{\theta_1}$, owing to the reparameterization described
+$A^\pi_\deltat(s, \pi_\phi(s))$ in the continuous case.  Eq.~\eqref{eq:max_A} is
+directly verified by $A_{\theta}$, owing to the reparameterization described
 in~\ref{subsec:reparam}.  To approximately verify~\eqref{eq:bellman_A}, the
@@
 \begin{align}
-	\delta Q_t &= A(s_t, a_t) - \frac{\delta r_t + \gamma^{\deltat} V(s_{t}+\delta s_{t}) - V(s_t)}{\deltat}\\
-	\theta^1_{t+\deltat} &= \theta^1_t + \eta_1 \partial_{\theta^1} V(s_t) \delta Q_t\\
-	\theta^2_{t+\deltat} &= \theta^2_t + \eta_2 \partial_{\theta^2} A(s_t, a_t) \delta Q_t.
+	\delta Q^k_\deltat &= A(s^k, a^k) - \frac{r^k \deltat + \gamma^{\deltat} V(s^{k}+\delta s^{k}) - V(s^k)}{\deltat}\\
+	\theta^{k+1}_\deltat &= \theta^k_\deltat + \eta^1_\deltat \partial_{\theta} V(s^k) \delta Q^k_\deltat\\
+	\psi^{k+1}_\deltat &= \psi^k_\deltat + \eta^2_\deltat \partial_{\psi} A(s^k, a^k) \delta Q^k_\deltat.
 	\label{eq:approx_bellman_A}
@@
 \label{subsec:lr}
-A necessary condition to guarantee invariance to time discretization is that
-the trajectory of the parameters of our function approximators should remain
-well behaved when the discretization parameter $\deltat$ decreases. This notably
-means that optimization steps should not be too large, in which case parameter
-trajectories could diverge in a single gradient step, or two small, in which case
-the trajectories would converge to single points as $\deltat$ goes to $0$.
-This in turn imposes conditions on the scalings of the learning rates of invariant
-algorithms with respect to the discretization timestep.
+For the discrete algorithm to admit a continuous time limit, the trajectories
+of parameters must converge to well defined trajectories as $\deltat$ goes to
+$0$.  Intuitively, this means that optimization steps should not be too large,
+in which case parameter trajectories could diverge in a single gradient step,
+or too small, in which case the trajectories would converge to single points as
+$\deltat$ goes to $0$.  This in turn imposes conditions on the scalings of the
+learning rates of invariant algorithms with respect to the discretization
+timestep.
 
-If we optimize $\theta_1$ and $\theta_2$ according to Eq.~\eqref{eq:approx_bellman_A},
-both $\theta_1$ and $\theta_2$ verify update equations of the form
+The updates equations on $\theta$ and $\psi$ are described in
+Eq.~\eqref{eq:approx_bellman_A}.  To have parameter steps that do not grow
+larger or shrink to $0$ as $\deltat \leftarrow 0$, both $\delta \theta^1$ and
+$\delta \theta^2$ should verify stochastic differential equations, and notably
+have a deterministic component of order $\deltat$. As $\deltat$ goes to $0$,
+updates are rephrased as
 \begin{align}
-	\
-	\delta \theta^1_t = \eta_1 \partial_{\theta_1} V(s_t) (
+	\delta Q^k_\deltat &= A(s^k_\deltat, a^k_\deltat) - r^k_\deltat - \log(\gamma) V(s^k_\deltat) - \partial_s V(s^k_\deltat) \frac{\delta s^k_\deltat}{\deltat}\nonumber \\&- \frac{(\delta s^k_\deltat)^T \partial_{s^2}^2 V(s^k_\deltat) \delta s^k_\deltat}{2\deltat} + \smallo\left(\frac{\eta^1_\deltat}{\sqrt{\deltat}}\right)\\
+	\delta \theta^k_\deltat &= \eta^1_\deltat \partial_{\theta} V(s^k_\deltat) \delta Q^k_\deltat + \smallo\left(\frac{\eta^1_\deltat}{\sqrt{\deltat}}\right)\\
+	\delta \psi^k_\deltat &= \eta^2_\deltat \partial_{\psi} A(s^k_\deltat) \delta Q^k_\deltat + \smallo\left(\frac{\eta^2_\deltat}{\sqrt{\deltat}}\right).
 \end{align}
-To have parameter steps that do not grow larger or shrink to $0$ as $\deltat
-\leftarrow 0$, both $\delta \theta^1$ and $\delta \theta^2$ should verify
-stochastic differential equations, and notably have a deterministic component
-of order $\deltat$. If the deterministic component
+For parameter updates to follow a stochastic differential equations, one must
+have $\eta^1_\deltat = \alpha^1 \deltat$ and $\eta^2_\deltat = \alpha^2
+\deltat$. With this choice of learning rates, the limit SDEs are
+\begin{align}
+	F &= A(s_t, a_t) - r_t - \log(\gamma) V(s_t) - \partial_s V(s_t) f(s_t, a_t)\nonumber\\
+	G &= - \partial_s V(s_t, a_t) \Sigma(s_t, a_t)\nonumber\\
+	d\theta_t &= (\alpha^1 F  + D \dbt)\partial_{\theta} V(s_t)\nonumber\\
+	d\psi_t &= (\alpha^2 F  + D \dbt)\partial_{\psi} A(s_t, a_t)\nonumber.
+\end{align}

### 2019-01-11 paper/icml_drau/advup.tex
Update.

@@
 discrete~\cite{dqn} action spaces are used. In details, $V_{\theta}$ and $A_{\psi}$ are learnt
-using approximate version of the following bellman equation
+using approximate version of the following bellman equation, which holds for the true $A^\pi_\deltat$
+and $V^\pi_\deltat$,
 \begin{align}
-	V^{\pi}_\deltat(s) + A^{\pi}_\deltat(s, a) &= \E_{\delta r \sim \delta r(s, a)}\delta r\nonumber\\ 
-				   &+ \gamma^{\deltat}  \E_{\delta s \sim \delta s(s, a)} V^{\pi}_\deltat(s + \delta s)\label{eq:bellman_A}\\
+	V^{\pi}_\deltat(s) + \deltat A^{\pi}_\deltat(s, a) &= r\deltat + \gamma^{\deltat}  \E_{s'} V^{\pi}_\deltat(s')\label{eq:bellman_A}\\
 	A^{\pi}_\deltat(s, \pi(s)) &= 0\label{eq:max_A}
 \end{align}
-for a greedy exploitation policy $\pi(s) = \text{argmax}_{a'}A^\pi_\deltat(s, a')$,
-exactly implemented in the case of discrete actions, and approximated by a
-neural network $\pi_\phi(s)$, trained to maximize (an approximation of)
-$A^\pi_\deltat(s, \pi_\phi(s))$ in the continuous case.  Eq.~\eqref{eq:max_A} is
-directly verified by $A_{\theta}$, owing to the reparameterization described
+for a greedy exploitation policy $\pi(s) = \text{argmax}_{a'}A^\pi_\deltat(s,
+a')$, exactly implemented in the case of discrete actions, and approximated by
+a neural network $\pi_\phi(s)$, trained to maximize (an approximation of)
+$A^\pi_\deltat(s, \pi_\phi(s))$ in the continuous case.  Eq.~\eqref{eq:max_A}
+is directly verified by $A_{\psi}$, owing to the reparameterization
+$A_\psi(s, a) = \bar{A}_\psi(s, a) - \bar{A}_\psi(s, \pi(s))$, described
 in~\ref{subsec:reparam}.  To approximately verify~\eqref{eq:bellman_A}, the
 corresponding squared residual is minimized by an approximate gradient descent.
-The corresponding update equations when learning from incoming transitions of
+The corresponding update equations when learning from a transition $(s, a, r, s')$ of
 an exploratory trajectory are
 \begin{align}
-	\delta Q^k_\deltat &= A(s^k, a^k) - \frac{r^k \deltat + \gamma^{\deltat} V(s^{k}+\delta s^{k}) - V(s^k)}{\deltat}\\
-	\theta^{k+1}_\deltat &= \theta^k_\deltat + \eta^1_\deltat \partial_{\theta} V(s^k) \delta Q^k_\deltat\\
-	\psi^{k+1}_\deltat &= \psi^k_\deltat + \eta^2_\deltat \partial_{\psi} A(s^k, a^k) \delta Q^k_\deltat.
+	\delta Q_\deltat &\leftarrow A(s, a) \deltat - \left(r \deltat + \gamma^{\deltat} V(s') - V(s)\right)\\
+	\theta_\deltat &\leftarrow \theta_\deltat + \eta^1_\deltat \partial_{\theta} V(s) \frac{\delta Q_\deltat}{\deltat}\\
+	\psi_\deltat &\leftarrow \psi_\deltat + \eta^2_\deltat \partial_{\psi} A(s, a) \frac{\delta Q_\deltat}{\deltat}.
 	\label{eq:approx_bellman_A}
 \end{align}
-where $\eta$'s are learning rates. In practice, $A$ and $V$ are learned using
+where $\eta$'s are learning rates. As we will see in the following section,
+to obtain a well defined continuous time limit, the learning rates can be rewritten as
+$\eta^i_\deltat = \alpha^i \deltat$, where $\alpha^1$ and $\alpha^2$ are independant
+of $\deltat$. In practice, $A$ and $V$ are learned using
 transitions sampled from a replay buffer instead of trajectory samples, to
-prevent excessive correlation between training samples~\cite{dqn}. Pseudocode
-for the discrete actions algorithm is given in Alg.~\ref{alg:dau}. Pseudocode
-for the continuous actions algorithm is given in the supplementary material.
+prevent excessive correlation between training samples~\cite{dqn}.
 \begin{algorithm}[ht]
@@
 \begin{align}
-	\delta Q^k_\deltat &= A(s^k_\deltat, a^k_\deltat) - r^k_\deltat - \log(\gamma) V(s^k_\deltat) - \partial_s V(s^k_\deltat) \frac{\delta s^k_\deltat}{\deltat}\nonumber \\&- \frac{(\delta s^k_\deltat)^T \partial_{s^2}^2 V(s^k_\deltat) \delta s^k_\deltat}{2\deltat} + \smallo\left(\frac{\eta^1_\deltat}{\sqrt{\deltat}}\right)\\
-	\delta \theta^k_\deltat &= \eta^1_\deltat \partial_{\theta} V(s^k_\deltat) \delta Q^k_\deltat + \smallo\left(\frac{\eta^1_\deltat}{\sqrt{\deltat}}\right)\\
-	\delta \psi^k_\deltat &= \eta^2_\deltat \partial_{\psi} A(s^k_\deltat) \delta Q^k_\deltat + \smallo\left(\frac{\eta^2_\deltat}{\sqrt{\deltat}}\right).
+	\delta Q_\deltat &= A(s, a)\deltat - r\deltat - \log(\gamma) V(s)\deltat - \partial_s V(s) \delta s\nonumber \\&- \frac{(\delta s)^T \partial_{s^2}^2 V(s) \delta s}{2} + \smallo\left(\sqrt{\deltat}\right)\\
+	\delta \theta_\deltat &= \eta^1_\deltat \partial_{\theta} V(s) \frac{\delta Q_\deltat}{\deltat} + \smallo\left(\frac{\eta^1_\deltat}{\sqrt{\deltat}}\right)\\
+	\delta \psi_\deltat &= \eta^2_\deltat \partial_{\psi} A(s) \frac{\delta Q_\deltat}{\deltat} + \smallo\left(\frac{\eta^2_\deltat}{\sqrt{\deltat}}\right).
 \end{align}
@@
 \end{align}
+The full algorithm can now be derived.
+Pseudocode for the discrete actions algorithm is given in Alg.~\ref{alg:dau}.
+Pseudocode for the continuous actions algorithm is given in the supplementary
+material.
+\TODO{Be careful when using RMSprop. When you are facing a stochastic environment
+	RMSprop evaluates the second momentum of gradients and rescale gradients by its
+	square root. This second momentum will
+	be of order $\frac{1}{\deltat}$, due to the stochasticity in the environment. Consequently
+	all gradient steps are multiplied by $\sqrt{\deltat}$, and learning rates need only be
+multiplied by $\sqrt{\deltat}$ instead of $\deltat$. Taking batches of size proportional to $\frac{1}{\deltat}$
+applies a similar fix.}

### 2019-01-11 paper/tangotd.tex

--- 2018-12-21 paper/tangotd.tex
@@
 \paragraph{Better variant?}
-Everything works equally well if we just learn a kernel $Z(s_1,s_2)$ such
+Everything works equally well if we just learn a kernel $z(s_1,s_2)$ such
 that
 \begin{equation}
-\transp{\1_{s_1}}\Delta^{-1}\psi\approx \E_{s_2\sim \rho} \,Z(s_1,s_2)
+\transp{\1_{s_1}}\Delta^{-1}\psi\approx \E_{s_2\sim \rho} \,z(s_1,s_2)
 \psi(s_2)
@@
 \begin{equation}
-Z=\sum_{s_1,s_2}\rho(s_2)  Z(s_1,s_2)\1_{s_1}\transp{\1_{s_2}}
-\end{equation}
-[TODO CHANGE NOTATION because currently the entries of $Z$ are not
-$Z(s_1,s_2)$ but $Z_{s_1s_2}=\rho(s_2)Z(s_1,s_2)$ which is confusing...]
+\Delta^{-1}\approx Z=\sum_{s_1,s_2}\rho(s_2)  z(s_1,s_2)\1_{s_1}\transp{\1_{s_2}}
+\end{equation}
 In particular the value function is
 \begin{equation}
-V(s_1)=\E_{s_2\sim \rho} \, Z(s_1,s_2) R(s_2)
+V(s_1)=\E_{s_2\sim \rho} \, z(s_1,s_2) R(s_2)
 \end{equation}
@@
 $\green$. Here we would have to learn an auxiliary network to imitate
-$\E_{s_2\sim \rho} Z(s_1,s_2) R(s_2)$ (but this is a supervised learning
+$\E_{s_2\sim \rho} z(s_1,s_2) R(s_2)$ (but this is a supervised learning
 problem; still, it looks suspiciously like $TD(1)$ except maybe for the
@@
 \begin{equation}
-\partial \pi(a|s)\E_{s'|s,a} \E_{s_2\sim \rho} \, Z(s',s_2)R(s_2)
-\end{equation}
-(this works even if $\rho$ is \emph{not} the invariant distribution). We
+\partial \pi(a|s)\E_{s'|s,a} \E_{s_2\sim \rho} \, z(s',s_2)R(s_2)
+\end{equation}
+(this works even if $\rho$ is \emph{not} the invariant distribution) or,
+with a baseline,
+\begin{equation}
+\partial \pi(a|s)\E_{s'|s,a} \E_{s_2\sim \rho} \, (z(s',s_2)-z(s,s_2))R(s_2)
+\end{equation}
+which should have much better variance. We
 can apply this each time a new reward is observed at some point $s_2$. We
@@
 \E_{s_2\sim \rho}
-\,Z(s_1,s_2)\psi(s_2)
+\,z(s_1,s_2)\psi(s_2)
 \approx \psi(s_1)+ \E_{s'\sim P(s'|s_1)}
 \E_{s_2\sim \rho}\,
-Z(s',s_2)\psi(s_2)
+z(s',s_2)\psi(s_2)
 \end{gather}
@@
 \E_{s_2\sim \rho}\,
-Z(s_1,s_2)\psi(s_2)
+z(s_1,s_2)\psi(s_2)
 \approx \psi(s_1)+ \E_{s_2\sim \rho}\E_{s'_2\sim P(s'_2|s_2)}\,
-Z(s_1,s_2)\psi(s'_2)
+z(s_1,s_2)\psi(s'_2)
 \end{gather}
@@
 \begin{multline}
-Z(s_1,s_2)\gets (1+\eta)Z(s_1,s_2)-\eta \E_{s_3\sim \rho}\,
-Z(s_1,s_3)Z(s_3,s_2)\,+\\\eta \gamma \E_{s_3\sim\rho,\, s'_3\sim
+z(s_1,s_2)\gets (1+\eta)z(s_1,s_2)-\eta \E_{s_3\sim \rho}\,
+z(s_1,s_3)z(s_3,s_2)\,+\\\eta \gamma \E_{s_3\sim\rho,\, s'_3\sim
 P(s'_3|s_3)}\,
-Z(s_1,s_3)Z(s'_3,s_2)
+z(s_1,s_3)z(s'_3,s_2)
 \end{multline}
 which is perfectly samplable. We can simulate the assignment
-$Z(s_1,s_2)\gets...$ via gradient descent on the parameters of
-$Z(s_1,s_2)$. \footnote{Should we adjust the size of the gradient descent
-to get the right change of $Z(s_1,s_2)$? If we just use SGD, since
+$z(s_1,s_2)\gets...$ via gradient descent on the parameters of
+$z(s_1,s_2)$. \footnote{Should we adjust the size of the gradient descent
+to get the right change of $z(s_1,s_2)$? If we just use SGD, since
 gradients may change from point to point, the speed may be different at
@@
 $Z$ so that $Z\approx \Id$ at time
-$0$, namely $Z(s,s)=1/\rho(s)$, although starting with $Z(s,s)=1$ should be
+$0$, namely $z(s,s)=1/\rho(s)$, although starting with $z(s,s)=1$ should be
 quite fine. (Small initializations of $Z$ are safer.) $Z>0$ (unless we
@@
 
-Another initialization might be $Z(s_1,s_2)\approx \frac{1}{1-\gamma}$
-for every pair $s_1,s_2$. Indeed, on-policy, $Z$ is the density with
+Another initialization might be $z(s_1,s_2)\approx \frac{1}{1-\gamma}$
+for every pair $s_1,s_2$. Indeed, on-policy, $z$ is the density with
 respect to the invariant distribution $\mu$, and for $\gamma$ close to
-$1$, $Z$ will be approximately $\frac{1}{1-\gamma}$ plus a smaller function (see
-below). Note that $Z(s_1,s_2)=\frac{1}{1-\gamma}$ is an exact solution of
-the middle Bellman equation (as is $Z=0$), but it is a repulsive fixed
+$1$, $z$ will be approximately $\frac{1}{1-\gamma}$ plus a smaller function (see
+below). Note that $z(s_1,s_2)=\frac{1}{1-\gamma}$ is an exact solution of
+the middle Bellman equation (as is $z=0$), but it is a repulsive fixed
 point, I think, so initializing to $\frac{1}{1-\gamma}$ plus noise should
-converge to the true solution. Initializing there should have the
+converge to the true solution [empirically I find this diverges]. Initializing there should have the
 advantage of starting close to reasonable values.
@@
 encompass the left and right Bellman. It does not rely on test functions,
-but the smoothing is implicit via the model used for $Z(s_1,s_2)$.
+but the smoothing is implicit via the model used for $z(s_1,s_2)$.
 
@@
 $s_3$—although the $Z$ factors are different for the three terms...). But
-I don't know how to build an efficient sampler from $Z(s_3,\cdot)$.
+I don't know how to build an efficient sampler from $z(s_3,\cdot)$.
 Something like Metropolis? but this function $Z$ is not constant...
 
-On-policy, sampling from $Z(\cdot,s_3)$ is easy: just sample one of the
+On-policy, sampling from $z(\cdot,s_3)$ is easy: just sample one of the
 recent states with geometric rates $\gamma$. This should divide the
-right part of the equation by $Z(s_1,s_3)$ (up to some $\gamma$-dependent
+right part of the equation by $z(s_1,s_3)$ (up to some $\gamma$-dependent
 factor). Thus we would sample two states $s_1,s_2$ along the trajectory,
 sample a state $s_3$ in between on the trajectory from $s_1$, and (with
-the right $\gamma$ factors) we would get a term with $Z(s_1,s_2)$, a term
-with $Z(s_3,s_2)$ and a term with $Z(s'_3,s_2)$. This looks suspiciously
+the right $\gamma$ factors) we would get a term with $z(s_1,s_2)$, a term
+with $z(s_3,s_2)$ and a term with $z(s'_3,s_2)$. This looks suspiciously
 like a large-scale Bellman from $s_1$... Variance could be higher since
@@
 
-On-policy, working with $\gamma\to 1$ is quite easy: indeed $Z(s_1,s_2)$ is
+On-policy, working with $\gamma\to 1$ is quite easy: indeed $z(s_1,s_2)$ is
 a density with respect to $\mu(s_2)$ so that the asymptotics when
-$\gamma\to 1$ is $Z(s_1,s_2)=\frac{1}{1-\gamma}+Z_0(s_1,s_2)$, and we can
-directly learn $Z_0$ by inserting $\frac{1}{1-\gamma}+Z_0$ in the middle
+$\gamma\to 1$ is $z(s_1,s_2)=\frac{1}{1-\gamma}+z_0(s_1,s_2)$, and we can
+directly learn $z_0$ by inserting $\frac{1}{1-\gamma}+z_0$ in the middle
 Bellman equation. This yields (omitting the expectations)
 \begin{multline}
-Z_0(s_1,s_2)\gets (1+\eta)Z_0(s_1,s_2)-\eta
-Z_0(s_1,s_3)(Z_0(s_3,s_2)-\gamma Z_0(s'_3,s_2))\\-\eta
-Z_0(s_1,s_3)+\frac{\eta}{1-\gamma}(\gamma Z_0(s'_3,s_2)-Z_0(s_3,s_2))
+z_0(s_1,s_2)\gets (1+\eta)z_0(s_1,s_2)-\eta
+z_0(s_1,s_3)(z_0(s_3,s_2)-\gamma z_0(s'_3,s_2))\\-\eta
+z_0(s_1,s_3)+\frac{\eta}{1-\gamma}(\gamma z_0(s'_3,s_2)-z_0(s_3,s_2))
 \end{multline}
@@
 $s'_3$ have the same law, so in expectation the last term is $-\eta
-\E_{s_3} Z_0(s_3,s_2)$. This yields the more symmetric
+\E_{s_3} z_0(s_3,s_2)$. This yields the more symmetric
 \begin{multline}
-Z_0(s_1,s_2)\gets (1+\eta)Z_0(s_1,s_2)-\eta
-Z_0(s_1,s_3)\left(Z_0(s_3,s_2)-\gamma Z_0(s'_3,s_2)\right)\\-\eta
-Z_0(s_1,s_3)-\eta Z_0(s_3,s_2)
+z_0(s_1,s_2)\gets (1+\eta)z_0(s_1,s_2)-\eta
+z_0(s_1,s_3)\left(z_0(s_3,s_2)-\gamma z_0(s'_3,s_2)\right)\\-\eta
+z_0(s_1,s_3)-\eta z_0(s_3,s_2)
 \end{multline}
@@
 equal to $\Id$ plus a continuous component; the $\Id$ is singular, so
-learning $Z(s_1,s_2)$ using eg a neural network will be difficult around
+learning $z(s_1,s_2)$ using eg a neural network will be difficult around
 the diagonal (and the initialization will be difficult). We may directly
@@
 local convergence properties than Bellman. Is it worth the cost?
+
+Normalization of $Z$. Since $Z=\sum \gamma^t P^t$, we have that
+$(1-\gamma)Z$ is a stochastic matrix (thanks Yann). This might help with
+normalization/detecting divergence. This translates as $\E_{s_2\sim \rho}
+z(s_1,s_2)=\frac{1}{1-\gamma}$ which is testable/enforceable. However, I
+fear it might make the fixed point $z(s_1,s_2)=\frac{1}{1-\gamma}$
+attractive instead of repulsive, and it might also perturb initialization
+(initializing $Z=\Id$ does not satisfy this normalization, so enforcing
+it too early is basically like initializing to
+$Z=\frac{1}{1-\gamma}\Id$ which empirically diverges).
 

### 2019-01-11 paper/tangotd.tex

@@
 
+Convergence? The true Newton iteration $Z\gets 2Z-Z\Delta Z$ initialized
+with $Z=\Id$ converges, and so does the continuous-time version
+$Z'=Z-Z\Delta Z$. However, due to sampling of $s_1\sim \rho$, we will (in
+the tabular case) have something like $Z'=\rho(Z-Z\Delta Z)$ or (in the
+parametric case) $\theta'=\E_\rho (\partial_\theta Z.(Z-Z\Delta Z))$ and
+it is not obvious that those converge. (All the problems with TD appear
+here as well.) Assuming overparametrization and convergence of supervised
+learning for a parametric model $Z_\theta$, one might proceed by
+``semi-steps'' as follows: Start with a $\theta_0$, and take the full
+Newton step $2Z_{\theta_0}-Z_{\theta_0}\Delta Z_{\theta_0}$ as a
+supervised training target for $Z_\theta$; then it does not really matter
+what the sampling distribution for this supervised training is, as long
+as that sampling distribution covers everywhere.
+
 \paragraph{Early version of the above: Learning to navigate.} Define a function $V(s,s')$ that

### 2019-01-12 paper/tangotd.tex

@@
 
+Building a universal policy. One may make the policy depend on a goal
+representation $\omega$ as an additional input. One may simultaneously
+compute $Z(s_1,s_2,\omega)$, namely the $Z$ function when following the
+policy with goal $\omega$, and train at the same time a map $s_2\mapsto \omega=\Omega(s_2)$ and
+policy parameters $\pi(a|s_1,\omega)$ such that following
+$\pi(|\Omega(s_2))$ indeed goes towards $s_2$, namely, the policy
+gradient for every $s_2$ is
+\begin{equation}
+\partial \pi(a|s,\Omega(s_2))\E_{s'|s,a} \E_{s_2\sim \rho} \,
+(z(s',s_2,\Omega(s_2))-z(s,s_2,\Omega(s_2)))
+\end{equation}
+[TODO check this is indeed the policy gradient for the time spent at
+$s_2$ if following policy $\pi(|\Omega(s_2))$]. Advantage: diversity of
+policies should appear by itself, no need to manually force options to be
+distinct. Depends on a chosen distribution of goals $s_2$. TODO check:
+this is with a reward a Dirac at $s_2$, check there's no technical
+problem wrt the reference distribution $\rho(s_2)$, as we use $z$ not $Z$
+in the policy gradient and $z$ is the density of $Z$ wrt $\rho$ but I
+think it's correct. Actually, for goal representations, a goal is not
+necessarily a state itself, should be a feature of states, or some
+representation of a reward function, so we should figure out the way to
+give a meaning to $\Omega(R)$ for an arbitrary reward $R$. The
+representation of $Z$ as $Z=\transp{\green}f$ does that automatically,
+thanks to linearity wrt $R$.
+
 \paragraph{Early version of the above: Learning to navigate.} Define a function $V(s,s')$ that

### 2019-01-12 paper/icml_drau/pendulum_fig.tex
First figure

%! TEX root = icml_drau.tex
\newcommand{\iw}{1cm}
\begin{figure*}[!ht]
	\begin{tabular}{c|c}
		DDPG & AU \\
		\hline
		\begin{tabular}{l|lllll}
			$\delta t = 0.01$ & \includegraphics[width=\iw]{figs_data/pendulum/0.01/ddpg_act_2.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.01/ddpg_act_4.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.01/ddpg_act_6.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.01/ddpg_act_8.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.01/ddpg_act_10.png} \\
			$\delta t = 0.005$ &\includegraphics[width=\iw]{figs_data/pendulum/0.005/ddpg_act_4.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.005/ddpg_act_8.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.005/ddpg_act_12.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.005/ddpg_act_16.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.005/ddpg_act_20.png} \\
			$\delta t = 0.002$ &\includegraphics[width=\iw]{figs_data/pendulum/0.002/ddpg_act_10.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.002/ddpg_act_20.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.002/ddpg_act_30.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.002/ddpg_act_40.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.002/ddpg_act_50.png} \\
			$\delta t = 0.001$ &\includegraphics[width=\iw]{figs_data/pendulum/0.001/ddpg_act_20.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.001/ddpg_act_40.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.001/ddpg_act_60.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.001/ddpg_act_80.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.001/ddpg_act_100.png} \\
			$\delta t = 0.0005$ &\includegraphics[width=\iw]{figs_data/pendulum/0.0005/ddpg_act_40.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.0005/ddpg_act_80.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.0005/ddpg_act_120.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.0005/ddpg_act_160.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.0005/ddpg_act_200.png}
		\end{tabular} &
		\begin{tabular}{lllll}
			 \includegraphics[width=\iw]{figs_data/pendulum/0.01/au_act_2.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.01/au_act_4.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.01/au_act_6.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.01/au_act_8.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.01/au_act_10.png} \\
			\includegraphics[width=\iw]{figs_data/pendulum/0.005/au_act_4.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.005/au_act_8.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.005/au_act_12.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.005/au_act_16.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.005/au_act_20.png} \\
			\includegraphics[width=\iw]{figs_data/pendulum/0.002/au_act_10.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.002/au_act_20.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.002/au_act_30.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.002/au_act_40.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.002/au_act_50.png} \\
			\includegraphics[width=\iw]{figs_data/pendulum/0.001/au_act_20.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.001/au_act_40.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.001/au_act_60.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.001/au_act_80.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.001/au_act_100.png} \\
			\includegraphics[width=\iw]{figs_data/pendulum/0.0005/au_act_40.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.0005/au_act_80.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.0005/au_act_120.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.0005/au_act_160.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.0005/au_act_200.png}
		\end{tabular}
	\end{tabular}
	\caption{Policies obtained by DDPG and AU at different macroscopic instants of training. Each image represents the policy learnt by the policy network, with $x$-axis representing angle, and $y$-axis angular velocity. The lighter the pixel, the closer to $1$ the action, the darker, the closer to $-1$.}
\end{figure*}

### 2019-01-12 paper/icml_drau/framework.tex
advancing...

@@
 \section{Near continuous time reinforcement learning}
-\label{sec:continous}
+\label{sec:framework}
 % In this section, a continuous reinforcement learning framework is introduced,
@@
 % value between two different actions in a fixed state is of order $\bigO(\deltat)$.
+
 \TODO{. Main ideas: near continuous time environments (control, mujoco, video games, ...)
@@
 
-Let assume that we have a Markov Decision Process \TODO{Cite?} $\langle {\cal S}, {\cal A}, T_{\deltat}, r_\deltat, \gamma_\deltat\rangle$.
-${\cal S}$ is the state-space, which is supposed to be continuous \TODO{not the good formulation}.
-The action space is \TODO{todo}.
+We say that ${\cal M}_\deltat = \langle {\cal S}, {\cal A}, T_{\deltat}, r_\deltat, \gamma_\deltat\rangle$ is a \emph{near-continuous} Markov Decision Process with trajectories $(s_\deltat^k)_{k\in\mathbb{N}}$ if there is a \emph{physical time} $\deltat$ between two observations, and if $(s_\deltat^{k+1})$ is close to $s_\deltat^k$.
+We assume that the state space ${\cal S}$ is continuous \TODO{True?
+Ok?}.
+The action space ${\cal A}$ can be either continuous or discrete \TODO{except in the theorems}.
 
-We assume that the transition is continuous.
-This means that ...
-\TODO{todo}.
-Discussion on the hypothesis, in which environments it is true.
-Is the condition $\forall \deltat, \forall s, \E[\|s_{k+1} - s_k \||s_k = s] \leq ?$.
+Many practical Reinforcement Learning problem can be considered as near-continuous: For example, in a simulated physical environment in which the $\deltat$ is fixed by the simulator, or in a real environments, if the agent is a robot.
 
-In many situations, we can assume that there is an uderlying process:
+% , the $\deltat$ is fixed by the sensitivity of the sensors and the reactivity of the motors. We assume that the state space ${\cal S}$ is continuous.
+
+In many situations, we can assume that the agent is actually is following a diffusion process:
 \begin{equation}
   \label{eq:diffusion}
-  ds = F(s, a) + \Sigma(s, a)\dbt
+  ds = F(s, a)dt + \Sigma(s, a)\dbt
 \end{equation}
-$F$ is the physics, and $\dbt$ is the noise.
+where $F: {\cal S}\times{\cal A}\rightarrow {\cal S}$ describes the physics of the environment, and $\dbt$ is the noise of the environment.
+If the environment is deterministic as in classical control theory, or in many usual simulated environments (Mujoco, Gym, ...), we have $\Sigma = 0$.
 
-There is convergence to the continuous process when $\deltat \rightarrow 0$.
+In this work we only consider deterministic policies $\pi: {\cal S} \rightarrow {\cal A}$. Therefore, the diffusion equation becomes:
+\begin{equation}
+  \label{eq:diffusion-policy}
+  ds = F(s, \pi(s))dt + \Sigma(s, \pi(s))\dbt
+\end{equation}
+
+\begin{theorem}
+  Assume that $\pi: {\cal S} \rightarrow {\cal A}$ is a continuous and deterministic policy.
+Let $(s_t)_{t\in\mathbb{R}_+}$ be the continuous process by following \eqref{eq:diffusion}.
+Let $(s^\deltat_k)$ be the discrete process obtained if the agent only get observations every $\deltat$ \TODO{explain better}.
+Then:
+  \begin{equation}
+    \label{eq:conv-traj}
+    (s_{\lfloor \frac{t}{\deltat}\rfloor}^\deltat)_{t\in\mathbb{R}_+} \rightarrow_{\deltat \rightarrow 0} (s_t)_{t\in\mathbb{R}_+}
+  \end{equation}
+  as probability distributions.
+\end{theorem}
 
 What we want is an algorithm invariant to $\deltat$, so we need to build a discrete algorithm such that when $\deltat \rightarrow 0$, the algorithm converge to a limit algorithm doing what we want.
-
-TODO: Explain $a_t = \pi(s_t)$ deterministic.
 

### 2019-01-12 paper/icml_drau/dauc.tex
Algorithm description.

@@
 	\STATE \textbf{Inputs:}
-	\STATE $\theta_1$ and $\theta_2$, parameters of
-	$V_{\theta_1}$ and $\bar{A}_{\theta_2}$.
-	\STATE $\theta_1'$ parameter of target $V_{\theta_1'}$.
-	\STATE $\pi_{\text{exp}}$ exploration policy.
-	\STATE \textbf{opt}$_1$, \textbf{opt}$_2$, $\eta_1$ and $\eta_2$ optimizers and learning rates.
-	\STATE $\mathcal{D}$, buffer of transitions $(s, a, r, d, s')$.
+	\STATE $\theta$ and $\psi$, parameters of
+	$V_{\theta}$ and $\bar{A}_{\psi}$.
+	\STATE $\pi^{\text{explore}}$ and $\nu_\deltat$ defining an exploration policy.
+	\STATE \textbf{opt}$_1$, \textbf{opt}$_2$, $\alpha_1 \deltat$ and $\alpha_2 \deltat$ optimizers and learning rates.
+	\STATE $\mathcal{D}$, buffer of transitions $(s, a, r, d, s')$, with $d$ the episode termination signal.
 	\STATE $\deltat$ and $\gamma$, time discretization and discount factor.
-	\STATE $\tau$, target network update factor.
 	\STATE \textbf{nb\_epochs} number of epochs.
@@
 	\FOR {$e=0, \textbf{nb\_epochs}$}
-	\FOR {$k=1, \textbf{nb\_steps}$}
-	\STATE Sample $a_t$ according to $\pi_{\text{exp}}$.
-	\STATE Perform $a_t$ and observe $(r_{t+1}, d_{t+1}, s_{t+1})$.
-	\STATE Store $(s_t, a_t, r_{t+1}, d_{t+1}, s_{t+1})$ in $\mathcal{D}$.
-	\STATE Update $\pi_{\text{exp}}$ with $(a_t, s_{t+1})$.
-	\STATE $t \gets t + 1$
+	\FOR {$j=1, \textbf{nb\_steps}$}
+	\STATE $a^k \leftarrow \pi^{\text{explore}}(s^k, \nu^k_\deltat)$.
+	\STATE Perform $a^k$ and observe $(r^{k+1}, d^{k+1}, s^{k+1})$.
+	\STATE Store $(s^k, a^k, r^{k+1}, d^{k+1}, s^{k+1})$ in $\mathcal{D}$.
+	\STATE $k \gets k + 1$
 	\ENDFOR
@@
 	\STATE \text{Sample a batch of $N$ random transitions from $\mathcal{D}$}
-	\STATE $Q^i \gets V_{\theta_1}(s^i) + \deltat\hspace{-.17em}\left(
-	\bar{A}_{\theta_2}(s^i, a^i) - \max\limits_{a'}\bar{A}_{\theta_2}(s^i, a')\right)$
-	\STATE $\tilde{Q^i} \gets r^i + (1 - d^i) \gamma^{\deltat} V_{\theta_1}(s'^i)$
-	\STATE $\Delta \theta_1 \gets \sum\limits_{i=1}^N \partial_{\theta_1} V_{\theta_1}(s^i) (Q^i - \tilde{Q^i})$
-	\STATE $\Delta \theta_2 \gets \sum\limits_{i=1}^N \partial_{\theta_2} A_{\theta_2}(s^i, a^i) (Q^i - \tilde{Q^i})$
-	\STATE Update $\theta_1$ using \textbf{opt}$_1$ with $\Delta \theta_1$ and learning rate $\eta_1$.
-	\STATE Update $\theta_2$ using \textbf{opt}$_2$ with $\Delta \theta_2$ and learning rate $\eta_2$.
-	\STATE $\theta_1' \gets \tau \theta_1' + (1 - \tau) \theta_1$
+	\STATE $Q^i \gets V_{\theta}(s^i) + \deltat\hspace{-.17em}\left(
+	\bar{A}_{\psi}(s^i, a^i) - \max\limits_{a'}\bar{A}_{\psi}(s^i, a')\right)$
+	\STATE $\tilde{Q^i} \gets r^i\deltat + (1 - d^i) \gamma^{\deltat} V_{\theta}(s'^i)$
+	\STATE $\Delta \theta \gets \frac{1}{N}\sum\limits_{i=1}^N  \frac{\left(Q^i - \tilde{Q^i}\right)\partial_{\theta} V_{\theta}(s^i)}{\deltat}$
+	\STATE $\Delta \psi \gets \frac{1}{N}\sum\limits_{i=1}^N \frac{\left(Q^i - \tilde{Q^i}\right)\partial_{\psi} \left(\bar{A}_{\psi}(s^i, a^i) - \max\limits_{a'}\bar{A}_{\psi}(s^i, a')\right) }{\deltat}$
+	\STATE Update $\theta$ with \textbf{opt}$_1$, $\Delta \theta$ and learning rate $\alpha_1 \deltat$.
+	\STATE Update $\psi$ with \textbf{opt}$_2$, $\Delta \psi$ and learning rate $\alpha_2 \deltat$.
 	\ENDFOR

### 2019-01-12 paper/tangotd.tex

@@
 policies should appear by itself, no need to manually force options to be
-distinct. Depends on a chosen distribution of goals $s_2$. TODO check:
+distinct. Depends on a chosen distribution of goals $s_2$; we could learn
+to go to uniformly distributed states, or create a goal at points where
+we observe a high reward value (so learning to go to all states with a
+reward). Just fixing $\Omega(s)\deq s$ once and for all might work, in
+principle, but may be ambitious. TODO check:
 this is with a reward a Dirac at $s_2$, check there's no technical

### 2019-01-12 paper/tangotd.tex

@@
 reward). Just fixing $\Omega(s)\deq s$ once and for all might work, in
-principle, but may be ambitious. TODO check:
+principle, but may be ambitious (or use a summary/auto-encoded value of
+$s$). TODO check:
 this is with a reward a Dirac at $s_2$, check there's no technical
@@
 representation of $Z$ as $Z=\transp{\green}f$ does that automatically,
-thanks to linearity wrt $R$.
+thanks to linearity wrt $R$. This could even help with exploration:
+setting temporary goals towards already-seen past states solves the "pure
+random walk" problem by introducing temporary momentum; one can even
+increase temperature a bit on $\omega$, by using slightly larger/more
+variant values of $\omega$ as the goal.
 

### 2019-01-12 paper/tangotd.tex

@@
 
+It's a bit strange to train the policy for goal $s_2$ via policy gradient,
+instead of just selecting the next state to maximize
+$z(s,s_2;s_2)$ where $z(s,s_2;s_3)$ is the $z$ function associated with
+the policy that has goal $s_3$. Formally, computing the latter is
+feasible: one has to follow an exploration policy that has a non-constant
+goal,so at each step we observe a transition of the policy with some goal
+$s_3$; and hope that Newton--Bellman for computing $z(s,s_2;s_3)$
+generalizes well with respect to $z_3$.
+
 \paragraph{Early version of the above: Learning to navigate.} Define a function $V(s,s')$ that

### 2019-01-12 paper/tangotd.tex

@@
 
+What I don't like here: This still does not capture the composition of
+policies/goals (like: to go to $s_2$, first go to $s_1$ via the
+$s_1$-policy, then apply the known $s_2$-policy from $s_1$). The
+Newton--Bellman equation captures, \emph{for a fixed policy}, the
+composition of paths of that policy. But the above still treats
+$z(s_1,s_2;\omega)$ as a separate problem for each $\omega$.
+
 \paragraph{Early version of the above: Learning to navigate.} Define a function $V(s,s')$ that

### 2019-01-13 pendulum_exp/dataanalysis/plots.py
update plots (kernel size)

@@ -82,6 +82,26 @@ def plot_learning_curves(expdata, key_list, namefile):
     # plt.savefig(namefile+'.eps', format="eps")
     plt.savefig(namefile+'.png', format='png', dpi=1000)
 
+start_date = datetime.strptime('2019_01_09_05_44_37', "%Y_%m_%d_%H_%M_%S")
+# stop_date = datetime.strptime('2019_01_08_03_13_35', "%Y_%m_%d_%H_%M_%S")
+# start_date = 'last'
+stop_date = None
+runlist = loader_leonard('/private/home/leonardb/workdir', 'mujoco_continuous', 
+    start_date=start_date, stop_date=stop_date)
+
+expdata = ExperimentData(runlist)
+
+print(expdata.deltakeys)
+
+expdata_ant = expdata.filter_settings(lambda s: s['env_id'] == 'ant')
+plot_learning_curves(expdata_ant, ['Return', 'Return'], 'ant')
+
+expdata_cartpole = expdata.filter_settings(lambda s: s['env_id'] == 'cartpole')
+plot_learning_curves(expdata_cartpole, ['Return', 'Return'], 'cartpole')
+
+
+# expdata_tau = expdata.filter_settings(lambda s: s['algo'] == 'discrete_value')
+# expdata_tau.repr_rawlogs("Return", 5)
 
 
 

### 2019-01-13 pendulum_exp/dataanalysis/script_leonard.py

@@ -23,6 +23,9 @@ plot_learning_curves(expdata_ant, ['Return', 'Return'], 'ant')
 expdata_cartpole = expdata.filter_settings(lambda s: s['env_id'] == 'cartpole')
 plot_learning_curves(expdata_cartpole, ['Return', 'Return'], 'cartpole')
 
+expdata_cheetah = expdata.filter_settings(lambda s: s['env_id'] == 'half_cheetah')
+plot_learning_curves(expdata_cheetah, ['Return', 'Return'], 'cheetah')
 
-# expdata_tau = expdata.filter_settings(lambda s: s['algo'] == 'discrete_value')
-# expdata_tau.repr_rawlogs("Return", 5)
+
+expdata_tau = expdata.filter_settings(lambda s: s['algo'] == 'discrete_value')
+expdata_tau.repr_rawlogs("Return", 5)

### 2019-01-13 paper/icml_drau/framework.tex

@@
 
-\TODO{. Main ideas: near continuous time environments (control, mujoco, video games, ...)
-There is a time discretization $\deltat$ given by the problem.
-The uderlying process does not depend of $\deltat$.
-Need to see what is happening when changing $\deltat$.
-The problem is invariant to $\deltat$.
-The quantities we are looking at should be invariant to.
-The learning procedure should be invariant.
-Looking at the limits when $\deltat \rightarrow 0$.
-In simulated environments, changing $\deltat$ only means to change the FPS.
-In real environments, for example with robots, this could mean changing the quality of the sensors and motors.}
+Many reinforcement learning environments are discretizations or approximations of continuous environments.
+We say that these environments are \emph{near continuous}.
+This includes usual control environment (Mujoco, Gym, ...) \TODO{Cite}.
+In practice, these environments are considered as discrete, and a time discretization $\deltat$ is fixed by the simulator.
+In real environments, like in robotics, the time discretization $\deltat$ is fixed by the sensors or motors.
+Still, in these environments, there is an uderlying continuous process which do not depend on $\deltat$.
+Therefore, our learning methods should be invariant to $\deltat$, as well as all the meaningfull quantities we are using.
+
+In this paper, we try to analyze \emph{near-continuous} environments when  modifying $\deltat$, especially when $\deltat \rightarrow 0$.
+In order to design algorithms more robust to the choice of $\deltat$, we want that when $\deltat \rightarrow 0$, the quantities we are using and our algorithm do converge to meaningful limit quantities and algorithms.
+
+\TODO{We should explain that, if not invariant, harder to find good hyperparameters}
+
+In this section, we first introduce our framework.
+Then we show how we can define the return, discount factor and Value function such that it has a limit when $\deltat \rightarrow 0$.
+Finally, we show that there is no Action-Value function in continuous MDPs.
 
@@
 
-We say that ${\cal M}_\deltat = \langle {\cal S}, {\cal A}, T_{\deltat}, r_\deltat, \gamma_\deltat\rangle$ is a \emph{near-continuous} Markov Decision Process with trajectories $(s_\deltat^k)_{k\in\mathbb{N}}$ if there is a \emph{physical time} $\deltat$ between two observations, and if $(s_\deltat^{k+1})$ is close to $s_\deltat^k$.
+We say that ${\cal M}_\deltat = \langle {\cal S}, {\cal A}, T_{\deltat}, r_\deltat, \gamma_\deltat\rangle$ is a \emph{near-continuous} Markov Decision Process with trajectories $(s_\deltat^k)_{k\in\mathbb{N}}$ if there is a \emph{physical time} $\deltat$ between two observations, and if $s_\deltat^{k+1}$ is close to $s_\deltat^k$.
 We assume that the state space ${\cal S}$ is continuous \TODO{True?
@@
 
-Many practical Reinforcement Learning problem can be considered as near-continuous: For example, in a simulated physical environment in which the $\deltat$ is fixed by the simulator, or in a real environments, if the agent is a robot.
+% Many practical Reinforcement Learning are near-continuous: For example, in a simulated physical environment in which the $\deltat$ is fixed by the simulator, or in a real environments, if the agent is a robot.
 
@@
 
-In many situations, we can assume that the agent is actually is following a diffusion process:
+If the \emph{near continuous} MDP is the discretization, or a good approximation, of continuous MDP, Iwe can assume that the agent is actually is following a diffusion process:
 \begin{equation}
@@
 where $F: {\cal S}\times{\cal A}\rightarrow {\cal S}$ describes the physics of the environment, and $\dbt$ is the noise of the environment.
+In that case, the transition function is: $T_\deltat(s, a, s')$ is a probability distribution over $s'$, which is the distribution of $s_\deltat$ if $(s_t)_{t\in\mathbb{R_+}}$ is following equation (\ref{eq:diffusion}), $s_0 = s$ and the action is $a$ during the interval $[0, \deltat]$.
 If the environment is deterministic as in classical control theory, or in many usual simulated environments (Mujoco, Gym, ...), we have $\Sigma = 0$.
@@
 
+\TODO{Les deux paragraphes suivants sont redondants avec l'intro de la section}
+If we assume that the \emph{near continuous} MDP we are observing is a discretization of a continuous MDP \TODO{for example which could follow a diffusion process as in Equation (\ref{eq:diffusion})}, then we know that the time discretization $\deltat$ is not intrisically related to the problem we are tackling.
+Therefore, our algorithm should be robust to different choices of $\deltat$.
+For example, in a simulated environment where the $\deltat$ only depends on the FPS, or in a real environment in which the $\deltat$ is defined by the reactivity of the sensors and motors, we would like our algorithms to be robust if the FPS in changed, or if we use more reactive sensors.
+
+In this paper, we try to design RL methods for \emph{near continuous} MDP robust to the choice of $\deltat$.
+Our method is to analyse the different quantities we are looking at when $\deltat$ is changing, and especially when $\deltat \rightarrow 0$, and make sure that the quantities we are interested in and our algorithms do converge to meaningful limit quantities and algorithms.
+\TODO{OK?} The first formal statement we can make on discretized MDP with a time discretization $\deltat$ is that if $\deltat \rightarrow 0$, the distribution of trajectories do converge to the continuous trajectories.
+
 \begin{theorem}
   Assume that $\pi: {\cal S} \rightarrow {\cal A}$ is a continuous and deterministic policy.
-Let $(s_t)_{t\in\mathbb{R}_+}$ be the continuous process by following \eqref{eq:diffusion}.
-Let $(s^\deltat_k)$ be the discrete process obtained if the agent only get observations every $\deltat$ \TODO{explain better}.
-Then:
+Let $(s_t)_{t\in\mathbb{R}_+}$ be the continuous process by following \eqref{eq:diffusion} and $(s_\deltat^k)$ be the discrete process obtained if the agent only get observations every $\deltat$ \TODO{explain better}.
+
+  We define the time process $\tilde s_\deltat^t = s^{\lfloor \frac{t}{\deltat}\rfloor}_\deltat$ for $t\in\mathbb{R_+}$.\TODO{Explain}. Then:
   \begin{equation}
     \label{eq:conv-traj}
-    (s_{\lfloor \frac{t}{\deltat}\rfloor}^\deltat)_{t\in\mathbb{R}_+} \rightarrow_{\deltat \rightarrow 0} (s_t)_{t\in\mathbb{R}_+}
+    (\tilde s_\deltat^t)_{t\in\mathbb{R}_+} \rightarrow_{\deltat \rightarrow 0} (s_t)_{t\in\mathbb{R}_+}
   \end{equation}
@@
 
-What we want is an algorithm invariant to $\deltat$, so we need to build a discrete algorithm such that when $\deltat \rightarrow 0$, the algorithm converge to a limit algorithm doing what we want.
-
-\subsection{Return, $\gamma$ and Value function in near continuous time}
+\subsection{Return, discount factor  and Value function in near continuous time}
 \label{sec:ret-gamma-v}
 
-For a continuous MDP we define the return as:
+In this section, we show how to define the discount factor $\gamma_\deltat$ as a function of $\deltat$, such that the return and Value function do converge to the desired limit when $\deltat \rightarrow 0$.
+
+\paragraph{Return and Discount Factor} For a continuous MDP the return is defined as \TODO{Cite}:
 \begin{equation}
@@
 
-The natural discretization of this, if we want to define a discretize return $R_\delta$ for each $\deltat$ is:
+The natural discretization of this equation, if we want to define a discretize return $R_\delta$ for each $\deltat$ is:
 \begin{equation}
@@
 
-We see that the good \TODO{rewrite} definitions of $\gamma_\deltat$ and $r_\deltat$ are:
+We see that the natural definitions of $\gamma_\deltat$ and $r_\deltat$ such that the return defined in equation (\ref{eq:discretized-return}) corresponds to the discrete return for the MDP ${\cal M}_\deltat$ are:
 \begin{align}
+  \label{eq:def-gamma}
   \gamma_\deltat =& \gamma^\deltat\\
+  \label{eq:def-reward}
   r_\deltat =& \deltat . r
 \end{align}
-Thus, we see that the natural discretization of a continuous MDP ${\cal M} = \langle {\cal S}, {\cal A}, T, r, \gamma\rangle$ is ${\cal M}_\deltat = \langle {\cal S}, {\cal A}, T_\deltat, r\deltat, \gamma^\deltat\rangle$.
+% Thus, we see that the natural discretization of a continuous MDP ${\cal M} = \langle {\cal S}, {\cal A}, T, r, \gamma\rangle$ is ${\cal M}_\deltat = \langle {\cal S}, {\cal A}, T_\deltat, r\deltat, \gamma^\deltat\rangle$.
 
-Thus, we should define $\gamma_\deltat = \gamma^\deltat$, so we have $\lim_{\deltat \rightarrow 0} R_\deltat(\tau)  = R(\tau)$. \TODO{Should we state that more formally?}
+% Thus, we should define $\gamma_\deltat = \gamma^\deltat$, so we have $\lim_{\deltat \rightarrow 0} R_\deltat(\tau)  = R(\tau)$.
 
-This is consistent with what we expect for the time horizon.
+\paragraph{Time horizon} The definition of $\gamma_\deltat$ in equation (\ref{eq:def-gamma}) is consistent with what we expect for the time horizon.
 We know \TODO{Cite?} that for a MDP with discount factor $\gamma$, the time horizon is $\frac{1}{1-\gamma}$ steps.
-But if the duration of a step is $\deltat$, this means that the time horizon in \emph{physical time} (independant of $\deltat$) is $\frac{\deltat}{1-\gamma}$, which goes to 0 if $\deltat \rightarrow 0$.
-This means that if $\gamma$ is not adapted to $\deltat$, the return $R_\deltat(\tau) = \sum_{k=0}^\infty\gamma^{k}r(s^\deltat_k, a^\deltat_k)\deltat \rightarrow_{\deltat\rightarrow 0} r(s_0, a_0)$.
-\TODO{A changer pour ne pas ecrire de chose fausse} But in practice, for example on a control problem, we do not care about the time horizon as a number of steps but as a \emph{physical quantity}, independant of $\deltat$.
+But if the duration of a step is $\deltat$, the time horizon in \emph{physical time} is $\frac{\deltat}{1-\gamma}$, which goes to 0 if $\deltat \rightarrow 0$.
+% This means that if $\gamma$ is not adapted to $\deltat$, the return $R_\deltat(\tau) = \sum_{k=0}^\infty\gamma^{k}r(s^\deltat_k, a^\deltat_k)\deltat \rightarrow_{\deltat\rightarrow 0} r(s_0, a_0)$.
+\TODO{A changer pour ne pas ecrire de chose fausse}
+But in practice, for example on a control problem, we do not care about the time horizon as a number of steps but as a \emph{physical} quantity independant of $\deltat$.
 With $\gamma_\deltat = \gamma^\deltat$, the time horizon in \emph{physical time} is:
@@
 \end{equation}
-\TODO{Explain again that we have a limit so it is better}
 
-The Value Function for a policy $\pi$ with a discretization $\deltat$ is
+\paragraph{Value function}
+
+With the definition of $\gamma_\deltat$ and $r_\deltat$ in equations \eqref{eq:def-gamma} and \eqref{eq:def-reward}, the Value Function for a policy $\pi$ with a discretization $\deltat$ becomes:
 \begin{equation}
@@
 \end{equation}
-and we have, for all $\sigma \in {\cal S}$, $\lim_{\deltat \rightarrow 0} V^\pi_\deltat(s) = V^\pi(s)$. \TODO{Theorem?}
+We have the following convergence theorem of the discretized value function to the continuous value function:
+  \begin{theorem}
+    We have, for all $s \in {\cal S}$:
+    \begin{equation}
+      \label{eq:conv-value}
+      \lim_{\deltat \rightarrow 0} V^\pi_\deltat(s) = V^\pi(s)
+    \end{equation}
+  \end{theorem}
 
-It follows the Bellman equation:
+The discretized value function follows the Bellman equation:
 \begin{equation}
@@
 
-If the MDP is following the equation (\ref{eq:diffusion}), rhe limit of the Bellman equation for $V^\pi_\deltat$ when $\deltat \rightarrow 0$ is the Hamilton-Jacobi-Bellman equation on $V^\pi$:
+If the MDP is following the equation (\ref{eq:diffusion}), the limit of the Bellman equation for $V^\pi_\deltat$ when $\deltat \rightarrow 0$ is the Hamilton-Jacobi-Bellman equation on $V^\pi$:
 \begin{equation}
@@
 
-% In what follows, $\dt$ denotes a truly infinitesimal time interval while $\deltat$
-% denotes a small, but non infinitesimal time interval. Similarily, $\dbt$ denotes
-% an infinitesimal (multi-dimensional) brownian step, while $\deltabt$ denotes its
-% discretized equivalent. Informally, $\dbt$ is to be thought of as $\gauss(0, \sqrt{\dt}^2)$,
-% and $\deltabt$ as $\gauss(0, \sqrt{\deltat}^2)$, where $\gauss(\mu, \sigma)$ denotes a gaussian
-% random variable of mean $\mu$ and standard deviation (or covariance matrix in the multi-dimensional case)
-% $\sigma$.
-
-% Consider a fully observable \emph{Markov Decision Process} defined by the following
-% transitions and rewards
-% \begin{align}
-% 	ds(s, a) &= F(s, a) dt + \Sigma(s, a) \dbt\\
-% 	dr(s, a) &= \reward(s, a) dt + \Sigma_r(s, a) \dbt.
-% \end{align}
-% The equation on $s$ is that of a diffusion process, with drift coeficient $F$
-% and diffusion matrix $\Sigma$.  This framework is only moderately restrictive.
-% For example, deterministic continuous control environments fall in this
-% framework, with $\Sigma = 0$ and $\Sigma_r = 0$. Limitations of the framework
-% are discussed in Sec.~\ref{sec:limitations}.
-% This framework is naturally discretized to arbitrary $\deltat$ by replacing all
-% $d$'s with $\delta$'s.
-
-% A continuous policy $\pi$ is defined as a deterministic mapping from states to
-% actions.  In continuous time, defining stochastic policies with well defined
-% state-value functions requires incorporating information from previous action into
-% the state to obtain a temporally coherent policy. To avoid such inconvenience, the
-% framework is restricted to deterministic policies. This is a mild constraint, as this
-% only enforces determinism of the exploitation policy and not of exploration policies.
-% In what follows, $\tau\sim\pi$ denotes a trajectory of states and actions sampled
-% according to policy $\pi$, with $s_0$ sampled according to an arbitrary initial distribution
-% on states $\rho_0$, and $dr_{t} = dr(s_t, a_t)$.
-
-% Under a deterministic policy $\pi$, one can define the $\gamma$ discounted
-% \emph{state value function} $V^\pi(s)$ and \emph{state action value function}
-% $Q^\pi(s, a)$ as
-% \begin{align}
-% 	V^\pi(s) &= \E_{\tau\sim\pi}\left[
-% 		\int\limits_{t=0}^\infty \gamma^{t}
-% 		dr_{t} \mid s_0 = s
-% 	\right]\\
-% 	Q^\pi(s, a) &= \E_{\tau\sim\pi}\left[
-% 		\int\limits_{t=0}^\infty \gamma^{t}
-% 		dr_{t} \mid s_0 = s, a_0 = a
-% 	\right]
-% \end{align}
-% which are naturally discretized as
-% \begin{align}
-% 	V^\pi(s) &= \E_{\tau\sim\pi}\left[
-% 		\sum\limits_{k=0}^\infty \gamma^{k\deltat}
-% 		\delta r_{k\deltat} \mid s_0 = s
-% 	\right]\\
-% 	Q^\pi(s, a) &= \E_{\tau\sim\pi}\left[
-% 		\sum\limits_{k=0}^\infty \gamma^{k\deltat}
-% 		\delta r_{k\deltat} \mid s_0 = s, a_0 = a
-% 	\right].
-% \end{align}
-% Both $V^\pi$ and $Q^\pi$ verify Bellman equations, informally
-% \begin{align}
-% 	V^\pi(s) &= \reward(s, \pi(s))\dt + \gamma^{\dt} \E_{ds \sim ds(s, \pi(s))} V^\pi(s + ds)\nonumber\\
-% 	Q^\pi(s, a) &= \reward(s, a)\dt + \gamma^{\dt} \E_{ds(s, a)} V^\pi(s + ds(s, a))
-% 	\label{eq:cont_bell}
-% \end{align}
-% which in continuous time translates, using Ito's lemma, into a Hamilton Jacobi equation
-% \begin{align}
-% 	\reward + \partial_s V^\pi \cdot f + \frac{1}{2} \text{tr}\left(\Sigma^T\partial^2_{s^2} V^\pi\Sigma\right) = (1 - \gamma) V^\pi(s).
-% \end{align}
-
 \subsection{There is no Q-function in continuous time}
 
-Contrary to the Value function, the Action-Value function $Q$ is ill-defined for a continuous MDP. Moreover, this means that the discretized $Q$ function \TODO{explain}. This can be stated formally.
+Contrary to the Value function, the Action-Value function $Q$ is ill-defined for a continuous MDP.
+More precisely, for a continuous MDP, $Q^\pi(s, a) = V^\pi(s)$, so the continuous $Q$-function is independant of the actions.
+For a \emph{near-continuous} MDP, this means that the $Q$ is very close  to the $V$-function.\TODO{Say more ?} Moreover, this means that the discretized $Q$ function \TODO{explain}.
+This can be stated formally.
 \begin{theorem}
@@
     \label{eq:q-discr}
-    Q^\pi_\deltat(s, a) = V^\pi(s) + \bigO\left(\sqrt\deltat\right)
+    Q^\pi_\deltat(s, a) = V^\pi(s) + \bigO\left(\deltat\right)
   \end{equation}
@@
 \begin{proof}
-    \TODO{TODO}
+  \begin{align}
+    Q^\pi(s, a) &= r(s, a)\deltat + \gamma^\deltat \E_{s'|s, a}\left[V(s')\right] \\
+  \end{align}
+  But we know that $s'$ is close to $s$. If we define $\delta s = s' - s$, then:
+  \begin{align}
+    \|\E_{s'|s, a}\left[\delta s\right]\| &= \bigO(\deltat) \\
+    \E_{s'|s, a}\left[\|\delta s\|^2\right] &= \bigO(\deltat)
+  \end{align}
+  \TODO{Explain.} Therefore:
+  \begin{align}
+    |\E_{s'|s, a}\left[V(s')\right] - V(s)| &= \E_{\delta s|s, a}\left[\partial_s V\cdot \delta s +\bigO(\|\delta s\|^2)\right]\\
+                                            &\leq \|\partial_s V\|\cdot \|\E_{\delta s|s, a}\left[\delta s\right]|\| +\bigO(\delta t)\\
+    &= \bigO(\deltat)
+  \end{align}
+  Finally:
+  \begin{align}
+    Q^\pi(s, a) &= r(s, a)\deltat + \gamma^\deltat \E_{s'|s, a}\left[V(s')\right] \\
+                &= r(s, a)\deltat + \left(1 + \bigO(\deltat)\right)(V(s) + \bigO(\deltat))\\
+    &= V(s) + \bigO(\deltat)
+  \end{align}
   \end{proof}

### 2019-01-14 paper/icml_drau/framework.tex

@@
 
-Many reinforcement learning environments are discretizations or approximations of continuous environments.
-We say that these environments are \emph{near continuous}.
-This includes usual control environment (Mujoco, Gym, ...) \TODO{Cite}.
-In practice, these environments are considered as discrete, and a time discretization $\deltat$ is fixed by the simulator.
-In real environments, like in robotics, the time discretization $\deltat$ is fixed by the sensors or motors.
-Still, in these environments, there is an uderlying continuous process which do not depend on $\deltat$.
-Therefore, our learning methods should be invariant to $\deltat$, as well as all the meaningfull quantities we are using.
+Many reinforcement learning environments are discretizations or approximations
+of continuous environments. We say that these environments are \emph{near
+continuous}. This includes usual control environment (Mujoco, Gym, ...)
+\TODO{Cite}. In practice, these environments are considered as discrete, and a
+time discretization $\deltat$ is fixed by the simulator. In real environments,
+like in robotics, the time discretization $\deltat$ is fixed by the sensors or
+motors. Still, in these environments, there is an uderlying continuous process
+which do not depend on $\deltat$. Therefore, our learning methods should be
+invariant to $\deltat$, as well as all the meaningfull quantities we are using.
 
@@
 
+\VC{Many reinforcement learning environments are not intrinsically time
+	discrete, but discretization of an underlying continuous time
+	environment. For instance, many control environments, such as the
+	Mujoco~\cite{ddpg} environments or OpenAI Gym classic~!CITE! control
+	environments are discretization of continuous time control problems.
+	For real world environments, the discretization timestep, $\deltat$ is
+	most often imposed by the quality of sensors and actuators. For visual
+	simulated environments, it often corresponds to the number of frames
+	that the simulator generates per second. As $\deltat$ is not an
+	intrinsic parameter of the environment, but a constraint imposed by
+	computational ressources, we would expect that decreasing $\deltat$
+	could  only improve the performance of reinforcement learning agents.
+	To achieve this goal, reinforcement learning algorithms should be
+	resilient to a change of $\deltat$, and should remain viable when
+	$\deltat \rightarrow 0$. We are thus interested in the behavior of
+	reinforcement learning algorithms in discretized environments, when the
+	discretization timestep is small. We will refer to such environments as
+	\emph{near continuous environments}.
+
+	In what follows, a formalized view of near continuous environments is given,
+	along with $\deltat$ dependant definitions of return, discount factor and
+	value functions. With these definitions, each of these quantities is shown
+	to converge to a well defined continous time limit. Finally, it is shown
+	that the state-action value function collapses to the value function as
+	$\deltat$ goes to $0$, and consequently that there is no Q-learning in
+	continuous time.
+}
 \subsection{Framework}
 
-We say that ${\cal M}_\deltat = \langle {\cal S}, {\cal A}, T_{\deltat}, r_\deltat, \gamma_\deltat\rangle$ is a \emph{near-continuous} Markov Decision Process with trajectories $(s_\deltat^k)_{k\in\mathbb{N}}$ if there is a \emph{physical time} $\deltat$ between two observations, and if $s_\deltat^{k+1}$ is close to $s_\deltat^k$.
-We assume that the state space ${\cal S}$ is continuous \TODO{True?
-Ok?}.
-The action space ${\cal A}$ can be either continuous or discrete \TODO{except in the theorems}.
+We say that ${\cal M}_\deltat = \langle {\cal S}, {\cal A}, T_{\deltat},
+r_\deltat, \gamma_\deltat\rangle$ is a \emph{near-continuous} Markov Decision
+Process with trajectories $(s_\deltat^k)_{k\in\mathbb{N}}$ if there is a
+\emph{physical time} $\deltat$ between two observations, and if
+$s_\deltat^{k+1}$ is close to $s_\deltat^k$. We assume that the state space
+${\cal S}$ is continuous \TODO{True? Ok?}. The action space ${\cal A}$ can be
+either continuous or discrete \TODO{except in the theorems}.
 
@@
 
+\VC{Consider ${\cal S} = \mathbb{R}^d$ a set of states, and ${\cal A}$ a set of
+	actions. A familly of \emph{Markov decision processes} ${\cal
+	M}_\deltat = \langle {\cal S}, {\cal A}, T_{\deltat}, r_\deltat,
+	\gamma_\deltat\rangle$ is \emph{near continous} if for any
+	deterministic policy $\pi:{\cal S} \rightarrow {\cal A}$, trajectories
+	of ${\cal M}_\deltat$ converge to stochastic processes that satisfy the
+	Stochastic differential equation
+	\begin{equation}
+		\label{eq:diffusion}
+		ds = F(s, a)dt + \Sigma(s, a)\dbt.
+	\end{equation}
+	Informally $F: {\cal S}\times{\cal A}\rightarrow {\cal S}$ describes
+	the deterministic dynamic of the environment, while $\Sigma(s, a)\dbt$
+	describes its stochasticity. Notably, $\Sigma = 0$ if the environment is
+	deterministic, as in classical control theory, or in many usual simulated
+	environments (Mujoco, Gym, ...).
+
+	In what follows, ${\cal M}_\deltat$ is taken to be a direct discretization of
+	Eq.~\ref{eq:diffusion} with discretization timestep $\deltat$. The
+	corresponding transition function $T_\deltat(., s, a)$ is the distribution of
+	$s'$ when starting from point $s$, and following Eq.~\ref{eq:diffusion} while
+	maintaining $a$ constant on an interval of time $\deltat$.
+
+	In this paper, we try to design RL methods for \emph{near continuous} MDP
+	robust to the choice of $\deltat$. Our method is to analyse the different
+	quantities we are looking at when $\deltat$ is changing, and especially when
+	$\deltat \rightarrow 0$, and make sure that the quantities we are interested in
+	and our algorithms do converge to meaningful limit quantities and algorithms.
+	!OK?! The first formal statement we can make on discretized MDP with a
+	time discretization $\deltat$ is that if $\deltat \rightarrow 0$, the
+	distribution of trajectories do converge to the continuous trajectories.
+
+	\begin{theorem}
+		Assume that $\pi: {\cal S} \rightarrow {\cal A}$ is a continuous and
+		deterministic policy. Let $(s_t)_{t\in\mathbb{R}_+}$ be the
+		continuous process by following \eqref{eq:diffusion} and
+		$(s_\deltat^k)$ be the discrete process obtained if the agent only
+		get observations every $\deltat$ !explain better!.
+
+		We define the time process $\tilde s_\deltat^t = s^{\lfloor
+		\frac{t}{\deltat}\rfloor}_\deltat$ for
+		$t\in\mathbb{R_+}$.!Explain!. Then:
+		\begin{equation}
+			\label{eq:conv-traj}
+			(\tilde s_\deltat^t)_{t\in\mathbb{R}_+} 
+			\xrightarrow[\deltat \rightarrow 0]{} (s_t)_{t\in\mathbb{R}_+}
+		\end{equation}
+		in distribution.
+	\end{theorem}
+}
+
 \subsection{Return, discount factor  and Value function in near continuous time}
@@
 
-\paragraph{Return and Discount Factor} For a continuous MDP the return is defined as \TODO{Cite}:
+\VC{
+	In what follows, $\deltat$-dependant definitions of the discount factor
+	and the reward are given. These definitions fit the discrete case
+	scenario when $\deltat=1$, and yields well defined, non trivial returns
+	and value functions when $\deltat$ goes to $0$.
+}
+
+\paragraph{Return and Discount Factor}
+For a continuous MDP and a continuous trajectory $\tau = (s_t, a_t)_t$,
+the return is defined as~\cite{cont_rl}
 \begin{equation}
@@
 
-The natural discretization of this equation, if we want to define a discretize return $R_\delta$ for each $\deltat$ is:
+The natural discretization of this equation, if we want to define a discretized
+return $R_\delta$ for each $\deltat$ is:
 \begin{equation}
 	\label{eq:discretized-return}
-  R_\deltat(\tau) = \sum_{k=0}^\infty\gamma^{k\deltat}r(s^\deltat_k, a^\deltat_k)\deltat
+	R_\deltat(\tau) = \sum_{k=0}^\infty\gamma^{k\deltat}
+	r(s^\deltat_k, a^\deltat_k)\deltat
 \end{equation}
 
-We see that the natural definitions of $\gamma_\deltat$ and $r_\deltat$ such that the return defined in equation (\ref{eq:discretized-return}) corresponds to the discrete return for the MDP ${\cal M}_\deltat$ are:
+Natural definitions of $\gamma_\deltat$ and $r_\deltat$ such
+that the return defined in equation (\ref{eq:discretized-return}) corresponds
+to the discrete return for the MDP ${\cal M}_\deltat$ follow:
 \begin{align}
@@
 	\label{eq:def-reward}
-  r_\deltat =& \deltat . r
+	r_\deltat =& \deltat . r.
 \end{align}
@@
 
-\paragraph{Time horizon} The definition of $\gamma_\deltat$ in equation (\ref{eq:def-gamma}) is consistent with what we expect for the time horizon.
+\paragraph{Microscopic/Macroscopic time, Time horizon}
+\VC{In near continuous environments, there are two notions of time,
+	the number of observation received, or the number of actions taken,
+	which we call the microscopic time, and the actual amount of time spent
+	in the underlying continuous time environment, which we call the macroscopic
+	time. The two are related through
+	\begin{equation}
+		\tau^\text{macro} = \tau^\text{micro} \deltat.
+	\end{equation}
+
+	The notion of time horizon, which is, informally, the amount of time on
+	which the agent optimizes its return, is only meaningful in macroscopic
+	time. Having a time horizon of $10$ steps means something very different
+	if $\deltat = 0.1$ and if $\deltat = 0.0001$. A rule of thumb
+	is that the time horizon of an agent with discount factor $\gamma$ is
+	of order $\frac{1}{1 - \gamma}$. If $\gamma_\deltat$ was left constant
+	as $\deltat$ goes to $0$, the corresponding macroscopic time horizon 
+	would write as $\frac{\deltat}{1 - \gamma}$ which goes to $0$ when
+	$\deltat$ goes to $0$.
+
+	On the contrary, the definition of $\gamma_\deltat$ in
+	Eq.~\eqref{eq:def-gamma} yields an approximately constant macroscopic
+	time horizon when $\deltat$ goes to $0$. Indeed,
+	\begin{equation}
+		\label{eq:time-horizon}
+		\frac{\deltat}{1-\gamma^\deltat}
+		= - \frac{1}{\log \gamma} + \bigO(\deltat)
+		\approx \frac{1}{1-\gamma}
+	\end{equation}
+}
+The definition of $\gamma_\deltat$ in equation (\ref{eq:def-gamma}) is consistent with what we expect for the time horizon.
 We know \TODO{Cite?} that for a MDP with discount factor $\gamma$, the time horizon is $\frac{1}{1-\gamma}$ steps.
@@
 
-With the definition of $\gamma_\deltat$ and $r_\deltat$ in equations \eqref{eq:def-gamma} and \eqref{eq:def-reward}, the Value Function for a policy $\pi$ with a discretization $\deltat$ becomes:
+The definition of the return Eq.~\eqref{eq:continuous-return}, yields the
+following definition of the continuous value function
+\begin{equation}
+  V^\pi(s) = \E_{\tau\sim\pi}\left[\int\limits_{0}^\infty \gamma^{t}\deltat r_{t} \mid s_0 = s\right].
+\end{equation}
+The corresponding discretization for time discretization $\deltat$ is
 \begin{equation}
@@
 \end{equation}
+which fits the usual definition of the discrete V-function for $\gamma_\deltat$ and
+$r_\deltat$ defined in~(\ref{eq:def-gamma}, \ref{eq:def-reward}).
 We have the following convergence theorem of the discretized value function to the continuous value function:
@@
   \label{eq:hamilton-jacobi-bellman}
-  r + \partial_s V^\pi \cdot F + \frac{1}{2} \text{tr}\left(\Sigma^T\partial^2_{s^2} V^\pi\Sigma\right) = (1 - \gamma) V^\pi(s)
+  r + \partial_s V^\pi \cdot F + \frac{1}{2} \text{tr}\left(\Sigma^T\partial^2_{s^2} V^\pi\Sigma\right) = - \log(\gamma) V^\pi(s)
 \end{equation}
@@
 
-Contrary to the Value function, the Action-Value function $Q$ is ill-defined for a continuous MDP.
-More precisely, for a continuous MDP, $Q^\pi(s, a) = V^\pi(s)$, so the continuous $Q$-function is independant of the actions.
-For a \emph{near-continuous} MDP, this means that the $Q$ is very close  to the $V$-function.\TODO{Say more ?} Moreover, this means that the discretized $Q$ function \TODO{explain}.
-This can be stated formally.
+Contrary to the Value function, the Action-Value function $Q$ is ill-defined
+for a continuous MDP. More precisely, for a continuous MDP, $Q^\pi(s, a) =
+V^\pi(s)$, so the continuous $Q$-function is independant of the actions. For a
+\emph{near-continuous} MDP, this means that the $Q$-function is very close  to the
+$V$-function.\TODO{Say more ?} Moreover, this means that the discretized $Q$
+function \TODO{explain}. This can be stated formally.
 \begin{theorem}

### 2019-01-14 paper/tangotd.tex

@@
 \end{equation}
+This equation is used to train the policy parameters; $\Omega(s_2)$ is
+trained so that $\omega=\Omega(s_2)$ indeed maximizes $z(s,s_2,\omega)$
+(because the best way to reach $s_2$ is to select the $\omega$ with
+maximal $z(s,s_2,\omega)$ at each step; TODO: from which $s$? TODO: here
+I'm confusing constant action plans and time-variable action plans).
 [TODO check this is indeed the policy gradient for the time spent at
-$s_2$ if following policy $\pi(|\Omega(s_2))$]. Advantage: diversity of
+$s_2$ if following policy $\pi(\Omega(s_2))$]. Advantage: diversity of
 policies should appear by itself, no need to manually force options to be

### 2019-01-14 pendulum_exp/dataanalysis/plots.py
Adding parameters to plots

@@ -15,8 +15,7 @@ algolabeldict = {
     'approximate_value':'ddpg',
 }
 
-def plot_learning_curves(expdata, key_list, namefile):
-    mint, maxt = 0, 100
+def plot_learning_curves(expdata, key_list, namefile, mint: float = 0, maxt: float = 100):
 
     nlines, ncol = len(key_list), 1
     fig, axes = plt.subplots(nlines, ncol, figsize=(5.*ncol, 4.*nlines))

### 2019-01-14 pendulum_exp/dataanalysis/script_corentin.py

@@ -5,6 +5,8 @@ from plots import plot_learning_curves
 parser = argparse.ArgumentParser()
 parser.add_argument('--logdir', type=str, required=True)
 parser.add_argument('--exp_name', type=str, required=True)
+parser.add_argument('--min_t', type=float, default=0.)
+parser.add_argument('--max_t', type=float, default=100.)
 args = parser.parse_args()
 
 start_date = 'last'
@@ -13,4 +15,5 @@ runlist = loader_leonard(args.logdir, args.exp_name, start_date=start_date, stop
 
 expdata = ExperimentData(runlist)
 
-plot_learning_curves(expdata, ['Return', 'Return'], 'pendulum')
+plot_learning_curves(expdata, ['Return', 'Return'], args.exp_name, mint=args.min_t, maxt=args.max_t)
+expdata.repr_rawlogs("Return", 5)

### 2019-01-14 pendulum_exp/main.py
Adding nb_true_epochs as a parameter to end training

@@ -63,7 +63,7 @@ def main(args):
     logdir = args.logdir
     noreload = args.noreload
     dt = args.dt
-    nb_epochs = args.nb_epochs
+    nb_true_epochs = args.nb_true_epochs
     nb_steps = args.nb_steps
     time_limit = args.time_limit
     eval_gap = args.eval_gap
@@ -88,7 +88,7 @@ def main(args):
         policy.load_state_dict(state_dict)
     log_gap = int(.1 / dt)
 
-    for e in range(cur_e, nb_epochs):
+    for e in range(cur_e, int(nb_true_epochs / dt)):
         info(f"Epoch {e}...")
         obs = train(nb_steps, env, policy, obs)
         new_R = evaluate(

### 2019-01-14 paper/tangotd.tex

@@
 representation $\omega$ as an additional input. One may simultaneously
-compute $Z(s_1,s_2,\omega)$, namely the $Z$ function when following the
+compute $Z(s_1,s_2;\omega)$, namely the $Z$ function when following the
 policy with goal $\omega$, and train at the same time a map $s_2\mapsto \omega=\Omega(s_2)$ and
@@
 \partial \pi(a|s,\Omega(s_2))\E_{s'|s,a} \E_{s_2\sim \rho} \,
-(z(s',s_2,\Omega(s_2))-z(s,s_2,\Omega(s_2)))
+(z(s',s_2;\Omega(s_2))-z(s,s_2;\Omega(s_2)))
 \end{equation}

### 2019-01-14 paper/tangotd.tex

@@
 This equation is used to train the policy parameters; $\Omega(s_2)$ is
-trained so that $\omega=\Omega(s_2)$ indeed maximizes $z(s,s_2,\omega)$
+trained so that $\omega=\Omega(s_2)$ indeed maximizes $z(s,s_2;\omega)$
 (because the best way to reach $s_2$ is to select the $\omega$ with
-maximal $z(s,s_2,\omega)$ at each step; TODO: from which $s$? TODO: here
+maximal $z(s,s_2;\omega)$ at each step; TODO: from which $s$? TODO: here
 I'm confusing constant action plans and time-variable action plans).
@@
 \begin{equation}
+\label{eq:optrightbellman}
 V(s,s')=\1_{s=s'}+\gamma \sum_{s_{t-1}} V(s,s_{t-1})\max_a
@@
 
+By multiplying \eqref{eq:optrightbellman} by $V$, one finds the
+Newton--Bellman equation for the optimal function $V$ seen as a matrix:
+\begin{equation}
+\label{eq:optrightbellman}
+V^2_{s_1s_2}=V_{s_1s_2}+\gamma \sum_{s_3} V_{s_1s_3}\max_a
+\E_{s_4|s_3,a} V_{s_4s_2}
+\end{equation}
+TODO DOUBLE-CHECK
+and representing $V$ by its density wrt $\rho$ namely
+$V_{s_1s_2}=z(s_1,s_2)\rho(s_2)$ this suggests the Newton method
+\begin{equation}
+z(s_1,s_2)\gets 2z(s_1,s_2)-\E_{s_3\sim \rho} z(s_1,s_3)z(s_3,s_2)+\gamma
+\E_{s_3\sim \rho} z(s_1,s_3)\max_a \E_{s_4\sim P(s_4|s_3,a)} z(s_4,s_2)
+\end{equation}
+satisfied by the optimal $z$ function.
+
 \end{document}

### 2019-01-14 paper/tangotd.tex

@@
 \end{equation}
-TODO DOUBLE-CHECK
+TODO CHECK, MAY BE WRONG DUE TO MAX OF WRONG STUFF!!
 and representing $V$ by its density wrt $\rho$ namely

### 2019-01-14 paper/tangotd.tex

--- 2019-01-14 paper/tangotd.tex
@@
 \begin{equation}
+\label{eq:optleftbellman}
 V(s,s')=\1_{s=s'}+\gamma \max_a \E_{s_1|s,a} V(s_1,s')
@@
 
-By multiplying \eqref{eq:optrightbellman} by $V$, one finds the
+By multiplying \eqref{eq:optleftbellman} by $V$ on the left, one finds the
 Newton--Bellman equation for the optimal function $V$ seen as a matrix:
 \begin{equation}
-\label{eq:optrightbellman}
 V^2_{s_1s_2}=V_{s_1s_2}+\gamma \sum_{s_3} V_{s_1s_3}\max_a
@@
 \end{equation}
-TODO DOUBLE-CHECK
 and representing $V$ by its density wrt $\rho$ namely
-$V_{s_1s_2}=z(s_1,s_2)\rho(s_2)$ this suggests the Newton method
-\begin{equation}
-z(s_1,s_2)\gets 2z(s_1,s_2)-\E_{s_3\sim \rho} z(s_1,s_3)z(s_3,s_2)+\gamma
-\E_{s_3\sim \rho} z(s_1,s_3)\max_a \E_{s_4\sim P(s_4|s_3,a)} z(s_4,s_2)
-\end{equation}
+$V_{s_1s_2}=z^\ast(s_1,s_2)\rho(s_2)$ this suggests the Newton method
+\begin{multline}
+z^\ast(s_1,s_2)\gets 2z^\ast(s_1,s_2)-\E_{s_3\sim \rho}\,
+z^\ast(s_1,s_3)z^\ast(s_3,s_2)\\
++\gamma
+\E_{s_3\sim \rho} \,z^\ast(s_1,s_3)\max_a \E_{s_4\sim P(s_4|s_3,a)} z^\ast(s_4,s_2)
+\end{multline}
 satisfied by the optimal $z$ function.

### 2019-01-14 pendulum_exp/dataanalysis/dataloader.py
Plot type selection enabled

@@ -1,27 +1,30 @@
-import os 
+import os
 import pickle as pkl
 from datetime import datetime
+from typing import List, Callable, Any
+import numpy as np
 
 class ExperimentData:
-    def __init__(self, experimentrunlist):
+    def __init__(self, runs: List['ExperimentRun']) -> None:
         # experimentrunlist = [ExperimentRun(args_file, logs_file) for args_file, logs_file in argslogs_filelist]
 
-        setting_list = []
-        for run in experimentrunlist:
+        settings: List[ExperimentSetting] = []
+        for run in runs:
             args = run.args
-            for setting in setting_list:
+            for setting in settings:
                 if setting.args == args:
-                    # setting.push(run)
+                    setting.append(run)
                     break
-            setting_list.append(ExperimentSetting([run]))
+            else:
+                settings.append(ExperimentSetting([run]))
 
-        self._setting_list = setting_list
+        self._settings = settings
 
 
     @property
     def deltakeys(self):
         if not hasattr(self, "_deltaargs"):
-            args_list = [setting.args for setting in self._setting_list]
+            args_list = [setting.args for setting in self._settings]
 
             expargsdict = {}
             for args in args_list:
@@ -40,27 +43,26 @@ class ExperimentData:
         if not hasattr(self, "_sharedargs"):
             self.deltakeys
         return self._sharedargs
-    
 
 
     def deltaitems(self):
         if not hasattr(self, "_deltaitems"):
-            self._deltaitems = [dict((k, setting.args[k]) for k in self.deltakeys) for setting in self._setting_list]
+            self._deltaitems = [dict((k, setting.args[k]) for k in self.deltakeys) for setting in self._settings]
         return self._deltaitems
-    
-    def filter_settings(self, bool_function):
-        run_list = []
-        for setting in self._setting_list:
-            if bool_function(setting.args):
-                run_list.extend(setting._experimentrunlist)
-        return ExperimentData(run_list)
 
-    def repr_rawlogs(self, key, nlastvalues):
+    def filter_settings(self, predicate: Callable[[Any], bool]):
+        runs: List[ExperimentRun] = []
+        for setting in self._settings:
+            if predicate(setting.args):
+                runs.extend(setting._runs)
[…]

### 2019-01-14 pendulum_exp/dataanalysis/plots.py

@@ -1,26 +1,24 @@
 import matplotlib
 # matplotlib.use('Agg')
 import matplotlib.pyplot as plt
-import numpy as np 
+import numpy as np
 from scipy.ndimage.filters import gaussian_filter1d
+from typing import List
 
 
-from dataloader import loader_leonard, ExperimentData
-from datetime import datetime
-
-algolabeldict = {
-    'discrete_value':"qlearning",
-    'discrete_advantage':"advup",
-    'approximate_advantage':'advup',
-    'approximate_value':'ddpg',
-}
-
-def plot_learning_curves(expdata, key_list, namefile, mint: float = 0, maxt: float = 100):
+def plot_learning_curves(expdata, key_list: List[str], namefile: str,
+                         mint: float = 0, maxt: float = 100, gtype: str = 'time_std'):
+    algolabeldict = {
+        'discrete_value': "qlearning",
+        'discrete_advantage': "advup",
+        'approximate_advantage': 'advup',
+        'approximate_value': 'ddpg',
+    }
 
     nlines, ncol = len(key_list), 1
     fig, axes = plt.subplots(nlines, ncol, figsize=(5.*ncol, 4.*nlines))
 
-    dtlist = [np.log(setting.args['dt']) for setting in expdata._setting_list]
+    dtlist = [np.log(setting.args['dt']) for setting in expdata._settings]
     if len(dtlist) > 0:
         cnorm = matplotlib.colors.Normalize(vmin=min(dtlist), vmax=max(dtlist))
         cm = matplotlib.cm.ScalarMappable(norm=cnorm, cmap='plasma')
@@ -29,16 +27,16 @@ def plot_learning_curves(expdata, key_list, namefile, mint: float = 0, maxt: flo
     for key, ax in zip(key_list, axes.flat):
         ax.set_title(key)
 
-        for setting in sorted(expdata._setting_list, key=lambda s:s.args['dt']):
+        for setting in sorted(expdata._settings, key=lambda s: s.args['dt']):
             args = setting.args
             dt = args['dt']
             algo = args["algo"]
             label = f"{algolabeldict[algo]}; dt={dt:.0e}"
             if algo == 'discrete_value' or algo == 'approximate_value':
                 linestyle = '--'
-                c=cm.to_rgba(np.log(dt))
-                linewidth=1.
-                alpha=None
+                c = cm.to_rgba(np.log(dt))
+                linewidth = 1.
+                alpha = None
[…]

### 2019-01-14 pendulum_exp/dataanalysis/dataloader.py
Producing curves with std across runs

@@ -62,14 +62,16 @@ class ExperimentData:
         print(f'{nlastvalues} last values of key {key} for all settings')
         for setting in self._settings:
             print(' ; '.join(f'{key}: {setting.args[key]}' for key in deltakeys))
-            timeseq, timeseq_std = setting.timeseq(key)
-            if timeseq is None:
+            stats = setting.timeseq(key)
+            timeseq = stats['mean']
+            std = stats['std']
+            if stats is None:
                 print('No logs')
                 print('----')
                 continue
             idxs = sorted(list(timeseq.keys()))[-nlastvalues:]
             for i in idxs:
-                print(f"{i}: {timeseq[i]}")
+                print(f"{i}: {timeseq[i]}+-{std[i]}")
             print('----')
 
 
@@ -81,7 +83,7 @@ class ExperimentSetting:
         self._runs = runs
 
     def append(self, run: 'ExperimentRun'):
-        assert run == self.args
+        assert run.args == self.args
         self._runs.append(run)
 
     def timeseq(self, key: str):
@@ -90,10 +92,19 @@ class ExperimentSetting:
             return None
         timeseqs = {k: [r.logs[key][k] for r in self._runs if k in r.logs[key]]
                     for k in self._runs[0].logs[key]}
-        mean_seq = {k: np.mean(t) for k, t in timeseqs.items()}
-        std_seq = {k: np.std(t) for k, t in timeseqs.items()}
-        return (mean_seq, std_seq)
-
+        stats_seq = {k: {
+            'mean': np.mean(t),
+            'std': np.std(t),
+            'median': np.median(t),
+            'max': np.max(t),
+            'min': np.min(t)} for k, t in timeseqs.items()}
+        timeseq = {}
+        for k1 in stats_seq:
+            for k2 in stats_seq[k1]:
+                if k2 not in timeseq:
+                    timeseq[k2] = {}
+                timeseq[k2][k1] = stats_seq[k1][k2]
+        return timeseq
 
 class ExperimentRun:
     def __init__(self, args_file: str, logs_file: str) -> None:

### 2019-01-14 pendulum_exp/dataanalysis/plots.py

@@ -49,7 +49,8 @@ def plot_learning_curves(expdata, key_list: List[str], namefile: str,
                 alpha = None
             else:
                 raise ValueError
-            timeseq, timeseq_std = setting.timeseq(key)
+            timeseq_stats = setting.timeseq(key)
+            timeseq = timeseq_stats['mean']
             if timeseq is None:
                 continue
 
@@ -75,9 +76,12 @@ def plot_learning_curves(expdata, key_list: List[str], namefile: str,
                 ax.fill_between(x, y-sigma, y+sigma, facecolor=c, alpha=0.2)
                 # ax.plot(dt*xdata, dt*ydata, c=c, alpha=0.2, linestyle=linestyle, linewidth=linewidth)
             elif gtype == 'run_std':
-                std_data = np.array([v for (i, v) in timeseq_std.items()])
-                ax.plot(xdata, ydata, label=label, c=c, alpha=alpha, linestyle=linestyle, linewidth=linewidth)
-                ax.fill_between(xdata, ydata-std_data / 2, y+std_data / 2, facecolor=c, alpha=0.2)
+                std_data = np.array([v for (i, v) in timeseq_stats['std'].items()])
+                x = xdata * dt
+                y = ydata * dt
+                std = std_data * dt
+                ax.plot(x, y, label=label, c=c, alpha=alpha, linestyle=linestyle, linewidth=linewidth)
+                ax.fill_between(x, y - std / 2, y + std / 2, facecolor=c, alpha=0.2)
         first = False
 
     plt.tight_layout()

### 2019-01-14 pendulum_exp/dataanalysis/script_corentin.py

@@ -5,6 +5,7 @@ from plots import plot_learning_curves
 parser = argparse.ArgumentParser()
 parser.add_argument('--logdir', type=str, required=True)
 parser.add_argument('--exp_name', type=str, required=True)
+parser.add_argument('--std_type', type=str, choices=['time', 'run'])
 parser.add_argument('--min_t', type=float, default=0.)
 parser.add_argument('--max_t', type=float, default=100.)
 args = parser.parse_args()
@@ -15,5 +16,5 @@ runlist = loader_leonard(args.logdir, args.exp_name, start_date=start_date, stop
 
 expdata = ExperimentData(runlist)
 
-plot_learning_curves(expdata, ['Return', 'Return'], args.exp_name, mint=args.min_t, maxt=args.max_t)
+plot_learning_curves(expdata, ['Return', 'Return'], args.exp_name, mint=args.min_t, maxt=args.max_t, gtype=args.std_type + "_std")
 expdata.repr_rawlogs("Return", 5)

### 2019-01-14 paper/tangotd.tex

@@
 policies should appear by itself, no need to manually force options to be
-distinct. Depends on a chosen distribution of goals $s_2$; we could learn
+distinct. This could even help with exploration:
+setting temporary goals towards already-seen past states solves the "pure
+random walk" problem by introducing temporary momentum; one can even
+increase temperature a bit on $\omega$, by using slightly larger/more
+variant values of $\omega$ as the goal. Depends on a chosen distribution of goals $s_2$; we could learn
 to go to uniformly distributed states, or create a goal at points where
@@
 representation of $Z$ as $Z=\transp{\green}f$ does that automatically,
-thanks to linearity wrt $R$. This could even help with exploration:
-setting temporary goals towards already-seen past states solves the "pure
-random walk" problem by introducing temporary momentum; one can even
-increase temperature a bit on $\omega$, by using slightly larger/more
-variant values of $\omega$ as the goal.
+thanks to linearity wrt $R$.
 
@@
 composition of paths of that policy. But the above still treats
-$z(s_1,s_2;\omega)$ as a separate problem for each $\omega$.
+$z(s_1,s_2;\omega)$ as a separate problem for each $\omega$. The optimal
+Newton--Bellman equation for $z^\ast$ (see below) does capture
+composition of optimal actions, though.
 
@@
 +\gamma
-\E_{s_3\sim \rho} \,z^\ast(s_1,s_3)\max_a \E_{s_4\sim P(s_4|s_3,a)} z^\ast(s_4,s_2)
+\E_{s_3\sim \rho} \,z^\ast(s_1,s_3)\max_a \E_{s_4|s_3,a} z^\ast(s_4,s_2)
 \end{multline}
-satisfied by the optimal $z$ function.
+satisfied by the optimal $z$ function. This is not quite an instance of
+the Newton method for matrix inversion anymore (because the operator
+$\Id-\gamma \max_a \E_{\cdot|a}$ is not linear), so convergence
+is unclear.
+
+Moreover, this matrix $V$ does not contain all the information about
+optimizing rewards: the optimal strategy to optimize a reward $R$ is not
+an average of the optimal strategies for going to $s_2$ weighted by
+$R(s_2)$...
 

### 2019-01-14 pendulum_exp/dataanalysis/dataloader.py
Slight modifications for more general stats

@@ -115,7 +115,7 @@ class ExperimentRun:
     @property
     def args(self):
         if not hasattr(self, '_args'):
-            assert os.path.isfile(self._args_file)
+            assert os.path.isfile(self._args_file), f"Non existing args file {self._args_file}"
 
             with open(os.path.join(self._args_file), 'rb') as f:
                 args = pkl.load(f)
@@ -128,7 +128,7 @@ class ExperimentRun:
     @property
     def logs(self):
         if not hasattr(self, '_logs'):
-            assert os.path.isfile(self._logs_file)
+            assert os.path.isfile(self._logs_file), f"Non existing log file {self._logs_file}"
 
             with open(self._logs_file, 'rb') as f:
                 self._logs = pkl.load(f)

### 2019-01-14 pendulum_exp/dataanalysis/plots.py

@@ -49,10 +49,11 @@ def plot_learning_curves(expdata, key_list: List[str], namefile: str,
                 alpha = None
             else:
                 raise ValueError
+
             timeseq_stats = setting.timeseq(key)
-            timeseq = timeseq_stats['mean']
-            if timeseq is None:
+            if timeseq_stats is None:
                 continue
+            timeseq = timeseq_stats['mean']
 
             xdata = np.array([i for (i, v) in timeseq.items()])
             ydata = np.array([v for (i, v) in timeseq.items()])

### 2019-01-14 paper/icml_drau/advup.tex
lots in framework, remarks in the other places

@@
 To define an object which contains the same information on actions as
-$Q^\pi_\deltat$, but admits an action dependant continuous time limit, , it is
+$Q^\pi_\deltat$, but admits an action dependant continuous time limit, it is
 natural to define
@@
 Contrary to the action state value function, the rescaled advantage function converges
-to an action dependent quantity when $\deltat$ goes to $0$
+to an action dependent quantity when $\deltat$ goes to $0$ \LB{In next equation, tiny detail but I would write $\frac{1}{\deltat}(r\deltat+...)$, easier to read.}
 \begin{align}
-	A^\pi_\deltat(s, a)=&\reward(s, a)+\frac{\gamma^{\deltat}\E_{\delta s \sim \delta s(s, a)}V^\pi_\deltat(s + \delta s)-V^\pi\deltat(s)}{\deltat}\nonumber\\
+	A^\pi_\deltat(s, a)=&\reward(s, a)+\frac{\gamma^{\deltat}\E_{\delta s \sim \delta s(s, a)}V^\pi_\deltat(s + \delta s)-V^\pi_\deltat(s)}{\deltat}\nonumber\\
 	=&\reward(s, a)+\log(\gamma) V^\pi(s)+\partial_s V^\pi(s)f(s, a)\nonumber\\
@@
 \end{align}
+
+\LB{It might be my math bias, but I would really prefer to read something like:
+  Contrary to the action state value function, the rescaled advantage function converges
+  to an action dependent quantity when $\deltat$ goes to $0$, which contains all the necessary information to improve the strategy. We have the following theorem:
+  \begin{theorem}
+    There is function $A^\pi(s, a)$ such that for all $a, s$, we have:
+    \begin{equation}
+      A^\pi_\deltat(s, a) \rightarrow_{\deltat \rightarrow 0} A^\pi(s, a)
+    \end{equation}
+    Moreover, $A^\pi$ contains all the necessary information about actions. We have $\pi$ is the optimal strategy if and only if $A^\pi(s, \pi(s)) = 0$ for all $s \in {\cal S}$.
+    [[I don't know if this is true, but this is the kind of property we want for A, and this is the idea of not only it converges, but it converges to a meaningful limit]]
+  \end{theorem}
+  And the equation which is there before would only be in the appendix, in the proof.
+}
 The discretized state action value function can then be rewritten as
@@
 
-A sufficient condition for the exploratory policy to admit a time continuous
+\LB{All this subsection is not extremely clear to me, especially the next paragraph.
+And it is a detail, but I'm not sure I like $\nu$ as a notation for a random variable.} A sufficient condition for the exploratory policy to admit a time continuous
 limit is for $\pi^\text{explore}_\deltat$ to be a measurable function, and for
@@
 The exploration scheme used~\cite{ddpg}, which consists in adding an
-\emph{Ornstein-Uhlenbeck}~\cite{orn-uhl} process (OU process) to actions, is naturally time.
+\emph{Ornstein-Uhlenbeck}~\cite{orn-uhl} process (OU process) to actions, is naturally time
 invariant. It can be rewritten as
@@
 a neural network $\pi_\phi(s)$, trained to maximize (an approximation of)
-$A^\pi_\deltat(s, \pi_\phi(s))$ in the continuous case.  Eq.~\eqref{eq:max_A}
-is directly verified by $A_{\psi}$, owing to the reparameterization
+$A^\pi_\deltat(s, \pi_\phi(s))$ in the continuous case.  \LB{Maybe say here that it's the same than DDPG?}Eq.~\eqref{eq:max_A}
+is directly verified by $A_{\psi}$, owing to the reparametrization
 $A_\psi(s, a) = \bar{A}_\psi(s, a) - \bar{A}_\psi(s, \pi(s))$, described
@@
 \end{align}
+\LB{I don't like the notations $\alpha^1$ and $\alpha^2$. What about $\alpha^A$ and $\alpha^V$? Easier to read} 
 where $\eta$'s are learning rates. As we will see in the following section,

### 2019-01-14 paper/icml_drau/framework.tex

@@
 \label{sec:framework}
-% In this section, a continuous reinforcement learning framework is introduced,
-% and it is shown that under this framework, the state-action value function does
-% not depend on the action. We further show that in near continous domains, when
-% the discretization timestep $\deltat$ is small, the difference in state-action
-% value between two different actions in a fixed state is of order $\bigO(\deltat)$.
 
-Many reinforcement learning environments are discretizations or approximations
-of continuous environments. We say that these environments are \emph{near
-continuous}. This includes usual control environment (Mujoco, Gym, ...)
-\TODO{Cite}. In practice, these environments are considered as discrete, and a
-time discretization $\deltat$ is fixed by the simulator. In real environments,
-like in robotics, the time discretization $\deltat$ is fixed by the sensors or
-motors. Still, in these environments, there is an uderlying continuous process
-which do not depend on $\deltat$. Therefore, our learning methods should be
-invariant to $\deltat$, as well as all the meaningfull quantities we are using.
-
-In this paper, we try to analyze \emph{near-continuous} environments when  modifying $\deltat$, especially when $\deltat \rightarrow 0$.
-In order to design algorithms more robust to the choice of $\deltat$, we want that when $\deltat \rightarrow 0$, the quantities we are using and our algorithm do converge to meaningful limit quantities and algorithms.
-
-\TODO{We should explain that, if not invariant, harder to find good hyperparameters}
-
-In this section, we first introduce our framework.
-Then we show how we can define the return, discount factor and Value function such that it has a limit when $\deltat \rightarrow 0$.
-Finally, we show that there is no Action-Value function in continuous MDPs.
-
-\VC{Many reinforcement learning environments are not intrinsically time
+  Many reinforcement learning environments are not intrinsically time
 	discrete, but discretization of an underlying continuous time
@@
 	Mujoco~\cite{ddpg} environments or OpenAI Gym classic~!CITE! control
-	environments are discretization of continuous time control problems.
+	environments are discretization of continuous time control problems,
+        \LB{with a discretization timestep $\deltat$ fixed by the simulator}.
 	For real world environments, the discretization timestep, $\deltat$ is
@@
 	continuous time.
-}
+      \TODO{We should explain somwhere here that, if not invariant, harder to find good hyperparameters}
 \subsection{Framework}
 
-We say that ${\cal M}_\deltat = \langle {\cal S}, {\cal A}, T_{\deltat},
-r_\deltat, \gamma_\deltat\rangle$ is a \emph{near-continuous} Markov Decision
-Process with trajectories $(s_\deltat^k)_{k\in\mathbb{N}}$ if there is a
-\emph{physical time} $\deltat$ between two observations, and if
-$s_\deltat^{k+1}$ is close to $s_\deltat^k$. We assume that the state space
-${\cal S}$ is continuous \TODO{True? Ok?}. The action space ${\cal A}$ can be
-either continuous or discrete \TODO{except in the theorems}.
-
-% Many practical Reinforcement Learning are near-continuous: For example, in a simulated physical environment in which the $\deltat$ is fixed by the simulator, or in a real environments, if the agent is a robot.
-
-% , the $\deltat$ is fixed by the sensitivity of the sensors and the reactivity of the motors. We assume that the state space ${\cal S}$ is continuous.
-
-If the \emph{near continuous} MDP is the discretization, or a good approximation, of continuous MDP, Iwe can assume that the agent is actually is following a diffusion process:
-\begin{equation}
-  \label{eq:diffusion}
-  ds = F(s, a)dt + \Sigma(s, a)\dbt
-\end{equation}
-where $F: {\cal S}\times{\cal A}\rightarrow {\cal S}$ describes the physics of the environment, and $\dbt$ is the noise of the environment.
-In that case, the transition function is: $T_\deltat(s, a, s')$ is a probability distribution over $s'$, which is the distribution of $s_\deltat$ if $(s_t)_{t\in\mathbb{R_+}}$ is following equation (\ref{eq:diffusion}), $s_0 = s$ and the action is $a$ during the interval $[0, \deltat]$.
-If the environment is deterministic as in classical control theory, or in many usual simulated environments (Mujoco, Gym, ...), we have $\Sigma = 0$.
-
-In this work we only consider deterministic policies $\pi: {\cal S} \rightarrow {\cal A}$. Therefore, the diffusion equation becomes:
-\begin{equation}
-  \label{eq:diffusion-policy}
-  ds = F(s, \pi(s))dt + \Sigma(s, \pi(s))\dbt
-\end{equation}
-
-\TODO{Les deux paragraphes suivants sont redondants avec l'intro de la section}
-If we assume that the \emph{near continuous} MDP we are observing is a discretization of a continuous MDP \TODO{for example which could follow a diffusion process as in Equation (\ref{eq:diffusion})}, then we know that the time discretization $\deltat$ is not intrisically related to the problem we are tackling.
-Therefore, our algorithm should be robust to different choices of $\deltat$.
-For example, in a simulated environment where the $\deltat$ only depends on the FPS, or in a real environment in which the $\deltat$ is defined by the reactivity of the sensors and motors, we would like our algorithms to be robust if the FPS in changed, or if we use more reactive sensors.
-
-In this paper, we try to design RL methods for \emph{near continuous} MDP robust to the choice of $\deltat$.
-Our method is to analyse the different quantities we are looking at when $\deltat$ is changing, and especially when $\deltat \rightarrow 0$, and make sure that the quantities we are interested in and our algorithms do converge to meaningful limit quantities and algorithms.
-\TODO{OK?} The first formal statement we can make on discretized MDP with a time discretization $\deltat$ is that if $\deltat \rightarrow 0$, the distribution of trajectories do converge to the continuous trajectories.
-
-\begin{theorem}
-  Assume that $\pi: {\cal S} \rightarrow {\cal A}$ is a continuous and deterministic policy.
-Let $(s_t)_{t\in\mathbb{R}_+}$ be the continuous process by following \eqref{eq:diffusion} and $(s_\deltat^k)$ be the discrete process obtained if the agent only get observations every $\deltat$ \TODO{explain better}.
-
-  We define the time process $\tilde s_\deltat^t = s^{\lfloor \frac{t}{\deltat}\rfloor}_\deltat$ for $t\in\mathbb{R_+}$.\TODO{Explain}. Then:
-  \begin{equation}
-    \label{eq:conv-traj}
-    (\tilde s_\deltat^t)_{t\in\mathbb{R}_+} \rightarrow_{\deltat \rightarrow 0} (s_t)_{t\in\mathbb{R}_+}
-  \end{equation}
-  as probability distributions.
-\end{theorem}
-
-\VC{Consider ${\cal S} = \mathbb{R}^d$ a set of states, and ${\cal A}$ a set of
+\VC{
+  Consider ${\cal S} = \mathbb{R}^d$ a set of states, and ${\cal A}$ a set of
 	actions. A familly of \emph{Markov decision processes} ${\cal
@@
 
+      }
+
+      \LB{
+        [[What I don't like in that version is that it is too formal ...]]
+        Consider ${\cal S} = \mathbb{R}^d$ a set of states, and ${\cal A}$ a set of
+	actions. Assume that an agent is following the continuous Markov Decision Process defined by the Stochastic Differential Equation:
+        \begin{equation}
+		\label{eq:diffusion}
+		ds = F(s, a)dt + \Sigma(s, a)\dbt.
+              \end{equation}
+        Then, for any $\deltat$, we say that the discrete \emph{Markov decision processes} ${\cal
+	M}_\deltat = \langle {\cal S}, {\cal A}, T_{\deltat}, r_\deltat,
+      \gamma_\deltat\rangle$ is a discretization of the continuous MDP with \emph{time discretization} $\deltat$ if the transition probability $T_\deltat(s, a, s')$ is the probability distribution of $s_\deltat$, with $s_0 = s$ and the action is $a_t=a$ during the interval $[0, \deltat]$.
+We say the the MDP ${\cal M}_\deltat$ is \emph{near-continuous}.
+[[Pb here: we require an uderlying continuous process: there is no intrinsic near continuous process]]
+
+      Informally $F: {\cal S}\times{\cal A}\rightarrow {\cal S}$ describes
+	the deterministic dynamic of the environment, while $\Sigma(s, a)\dbt$
+	describes its stochasticity. Notably, $\Sigma = 0$ if the environment is
+	deterministic, as in classical control theory, or in many usual simulated
+	environments (Mujoco, Gym, ...).
+      }
+
         In this paper, we try to design RL methods for \emph{near continuous} MDP
@@
 	$\deltat \rightarrow 0$, and make sure that the quantities we are interested in
-	and our algorithms do converge to meaningful limit quantities and algorithms.
-	!OK?! The first formal statement we can make on discretized MDP with a
-	time discretization $\deltat$ is that if $\deltat \rightarrow 0$, the
-	distribution of trajectories do converge to the continuous trajectories.
+	and our algorithms do converge to meaningful limit quantities and algorithms.\TODO{Better formulation maybe?}
+
+        \LB{In that direction, the first thing to do is to check that the discretized
+          trajectory do converge to the continuous ones. This is stated in the following theorem. [[Tres tres mal dit, mais c'est l'idee]]
+        }
+        % The first formal statement we can make on discretized MDP with a
+	% time discretization $\deltat$ is that if $\deltat \rightarrow 0$, the
+	% distribution of trajectories do converge to the continuous trajectories.
 
@@
 	\end{theorem}
-}
 
@@
       }
-The definition of $\gamma_\deltat$ in equation (\ref{eq:def-gamma}) is consistent with what we expect for the time horizon.
-We know \TODO{Cite?} that for a MDP with discount factor $\gamma$, the time horizon is $\frac{1}{1-\gamma}$ steps.
-But if the duration of a step is $\deltat$, the time horizon in \emph{physical time} is $\frac{\deltat}{1-\gamma}$, which goes to 0 if $\deltat \rightarrow 0$.
-% This means that if $\gamma$ is not adapted to $\deltat$, the return $R_\deltat(\tau) = \sum_{k=0}^\infty\gamma^{k}r(s^\deltat_k, a^\deltat_k)\deltat \rightarrow_{\deltat\rightarrow 0} r(s_0, a_0)$.
-\TODO{A changer pour ne pas ecrire de chose fausse}
-But in practice, for example on a control problem, we do not care about the time horizon as a number of steps but as a \emph{physical} quantity independant of $\deltat$.
-With $\gamma_\deltat = \gamma^\deltat$, the time horizon in \emph{physical time} is:
+
+      \LB{ \paragraph{Number of steps / Physical time: Time horizon}
+        % [[I prefer to talk about ``number of steps'' VS ``Physical time'' than microscopic/Macroscopic]]
+        In near continuous environments, there are two notions of time,
+	the number of steps, or the number of actions taken, and the actual amount of time spent
+	in the underlying continuous time environment, which we call the physical
+	time. If $t$ is the real time and $k$ the number of steps, the two are related through $t = k \deltat$.
+
+	The notion of time horizon, which is, informally, the amount of time on
+	which the agent optimizes its return, is only meaningful in physical
+	time.
+
+        %Having a time horizon of $10$ steps means something very different
+	%if $\deltat = 0.1$ and if $\deltat = 0.0001$.
+
+        A rule of thumb
+	is that the time horizon of an agent with discount factor $\gamma$ is
+	of order $\frac{1}{1 - \gamma}$. If $\gamma_\deltat$ was left constant
+	as $\deltat$ goes to $0$, the corresponding macroscopic time horizon 
+	would write as $\frac{\deltat}{1 - \gamma}$ which goes to $0$ when
+	$\deltat$ goes to $0$.
+
+	On the contrary, the definition of $\gamma_\deltat$ in
+	Eq.~\eqref{eq:def-gamma} yields an approximately constant macroscopic
+	time horizon when $\deltat$ goes to $0$. Indeed,
 	\begin{equation}
           \label{eq:time-horizon}
-  \frac{\deltat}{1-\gamma^\deltat} \rightarrow_{\deltat \rightarrow 0} - \frac{1}{\log \gamma} \approx \frac{1}{1-\gamma}
+          \frac{\deltat}{1-\gamma^\deltat}
+          = - \frac{1}{\log \gamma} + \bigO(\deltat)
+          \approx \frac{1}{1-\gamma}
         \end{equation}
+
+      }
 
@@
 
+\LB{I did not change this part}
 The definition of the return Eq.~\eqref{eq:continuous-return}, yields the

### 2019-01-14 paper/icml_drau/icml_drau.tex
Advances.
very first version of the framework part, lots of thing missing but the main structure is there
First figure

@@
 % Recommended, but optional, packages for figures and better typesetting:
-\usepackage{amsmath, amssymb}
+\usepackage{amsmath, amssymb, amsthm}
 \usepackage{mathtools}
@@
 \usepackage{booktabs} % for professional tables
+\usepackage{array}
 \usepackage{xcolor}
@@
 \icmltitlerunning{Deep Relativistic Advantage Updating}
+
+\theoremstyle{plain}
+\newtheorem{theorem}{Theorem}
+\newtheorem{proposition}{Proposition}
 
@@
 	In this paper, we study the sensitivity of Deep Reinforcement learning algorithms to
-	time discretization. More precisely, we show that approaches based on estimations of the
+	time discretization \LB{in near-continuous time environments}. More precisely, we show that approaches based on estimations of the
 	Q-function, e.g. \emph{Deep Q-learning}~\cite{dqn} and \emph{Deep deterministic policy gradient}~\cite{ddpg}
@@
 	Q-function as a sum of a state value term and a \emph{small} action dependent advantage term yields an algorithm
-	much more resilient to variations of time discretization.
+	much more resilient to variations of time discretization.\LB{Change last sentence to: We introduce a more resilient algorithm to variations of time discretization [We could say more, but still less that what is already there]}
 \end{abstract}

### 2019-01-14 paper/icml_drau/intro.tex

@@
+%! TEX root = icml_drau.tex
+
 \section{Introduction}
@@
 games~(\citealt{openai_five}, \citealt{dqn}), or learning dexterous manipulation
-from scratch without demonstrations~\cite{hand_control}.
+from scratch without demonstrations~\cite{hand_control}. \LB{I would merge these two paragraphs.}
 
@@
 In this paper we focus our attention on the sensitivity to time discretization
-of DRL approaches, i.e. what happens when an agent receives $50$ observations
+of DRL approaches \LB{in \emph{near-continuous} environments (control, ...},
+i.e. what happens when an agent receives $50$ observations
 and is expected to take $50$ actions per second instead of $10$. One expects
@@
 framework to study this phenomenon is Continuous time reinforcement learning 
-(Sec.~\ref{sec:continous}).
+(Sec.~\ref{sec:framework}).
 
@@
 than vanilla DQN or DDPG on more complex environments (Sec.~\ref{sec:exp}).
+\LB{I think I would give less details, especially at the beginning, but insist on the methodology
+  (all quantities and algorithms should have meaningful limits when $\deltat \rightarrow 0$)}
+
+\LB{We should say somewhere that more invariance should mean less time spent to find hyperparameters}
+
+\LB{Our main contributions are:
+  \begin{itemize}
+  \item We show that the Q-function is ill-defined in continuous environments, and thus that
+    RL methods based on the Q-function (Q-learning, DDPG) are not robust to variations of time discretization.
+  \item We introduce an algorithm (NAME?) resilient to time discretization \TODO{say more}
+  \end{itemize}
+}

### 2019-01-15 paper/icml_drau/icml_drau.tex
working with CT

--- 2019-01-10 paper/icml_drau/icml_drau.tex
@@
 \begin{abstract}
-	Despite remarkable successes, \emph{Deep Reinforcement learning} remains sensitive to
+	Despite remarkable successes, \emph{Deep Reinforcement learning} (DRL) remains sensitive to
 	hyperparameterization, implementation details and small environment changes~(\citealt{drl_matter}, \citealt{drl_matter_bis}). 
 	Overcoming this sensitivity is key to making DRL applicable to real world problems.
-	In this paper, we study the sensitivity of Deep Reinforcement learning algorithms to
-	time discretization. More precisely, we show that approaches based on estimations of the
+	In this paper, we study the sensitivity of DRL to
+	time discretization in \emph{near continuous time} environments. More precisely, we show that approaches based on estimations of the
 	Q-function, e.g. \emph{Deep Q-learning}~\cite{dqn} and \emph{Deep deterministic policy gradient}~\cite{ddpg}
-	are sensitive to variations of time discretization. We further show that a simple reparameterization of the
-	Q-function as a sum of a state value term and a \emph{small} action dependent advantage term yields an algorithm
-	much more resilient to variations of time discretization.
+	are sensitive to variations of time discretization.
+        We detail a principled method to obtain an off-policy RL algorithm that is more robust to changes in time discretization, and demontrate this robustness empirically. 
+        % We further show that a simple reparameterization of the
+	% Q-function as a sum of a state value term and a \emph{small} action dependent advantage term yields an algorithm
+	% much more resilient to variations of time discretization.\LB{Change last sentence to: We introduce a more resilient algorithm to variations of time discretization [We could say more, but still less that what is already there]}
 \end{abstract}

### 2019-01-15 paper/icml_drau/intro.tex
working with CT

--- 2019-01-10 paper/icml_drau/intro.tex
@@
+%! TEX root = icml_drau.tex
+
 \section{Introduction}
 \label{sec:intro}
-In recent years, deep learning based approaches to reinforcement learning have
+In recent years, \emph{Deep Reinforcement Learning} (DRL) approaches have
 thrived and provided impressive results in a variety of domains, achieving superhuman
@@
 from scratch without demonstrations~\cite{hand_control}. 
-
-In spite of those successes, \emph{Deep reinforcement learning} approaches are
+In spite of those successes, DRL approaches are
 sensitive to a number of factors, including hyperparameterization,
@@
 In this paper we focus our attention on the sensitivity to time discretization
-of DRL approaches, i.e. what happens when an agent receives $50$ observations
+of DRL approaches,
+i.e. what happens when an agent receives $50$ observations
 and is expected to take $50$ actions per second instead of $10$. One expects
 that decreasing time discretization, or equivalently shortening reaction time,
-should only improve agent performances. This is not what happens in practice
-for approaches based on estimation of state action value functions, e.g.
+should only improve agent performances. Robustness to time discretization is notably relevant in \emph{near-continuous}
+environments, which includes most continuous control environments, robotics and many video games.
+\TODO{Explain that it also includes $\gamma \rightarrow 1$.}
+
+In practice, standard approaches based on estimation of state action value functions, e.g.
 \emph{Deep Q-learning} (DQN~\citep{dqn} and \emph{Deep deterministic policy
-gradient} (DDPG~\citep{ddpg}). This is shown experimentally in Sec.~\ref{sec:exp}.
-
-We relate this sensitivity to the fact that, as the discretization timestep
-decreases, the effect of individual actions on the total return decreases, and
+gradient} (DDPG~\citep{ddpg}) are not robust to changes in time discretization. This is shown experimentally in Sec.~\ref{sec:exp}. 
+This sensitivity can be understood as follows: as the discretization timestep
+decreases, the effect of individual actions on the total return decreases too, and
 can vanish when the discretization timestep becomes infinitesimal. The natural
 framework to study this phenomenon is Continuous time reinforcement learning 
-(Sec.~\ref{sec:continous}).
+(Sec.~\ref{sec:framework}) \cite{cont_rl, adv_upd}.
 
-To cope for this effect, we introduce a
-parameterization of the Q-function as a sum of a state value term and a
-\emph{small} action dependent advantage term (Sec.~\ref{subsec:reparam}). This
-reparameterized Q approximation can be trained with deep variants of Q-learning
-(Sec.~\ref{subsec:algorithm}).  To properly scale w.r.t. time discretization,
-particular attention has to be taken regarding both the exploration method, and
-the scaling of learning rates (Sec.~\ref{subsec:lr}, Sec.~\ref{subsec:explo}).
+A time discretization invariant algorithm should
+in particular admit a time continuous limit i.\,e. when the discretization timestep goes to $0$.
+This leads to precise design choices in term of
+agent architecture, exploration policy and learning rates scalings.
 The resulting algorithm is shown to provide near perfect invariance to time
@@
 than vanilla DQN or DDPG on more complex environments (Sec.~\ref{sec:exp}).
+On a new environment, as soon as the order of magnitude of the time discretization is known, our analysis readily provides salient scalings for a number of hyperparameters.
+
+Our contribution is threefold:
+\begin{itemize} 
+\item We formally show that the Q-function collapses to the V-function in \emph{near-continuous} time, and thus that
+    standard Q-learning is ill-behaved in this setting.
+  \item Our analysis of limit properties in near continuous environments leads us to define a robust methods off-policy algorithm. In particular,
+    it provides insights on how to design architectures, how to explore, and how to scale learning rates.
+  \item We empirically show that standard Q-learning methods are not robust to changes in time discretization, and that our algorithm provides huge gain in that domain.
+\end{itemize}

### 2019-01-15 paper/tangotd.tex

@@
 it too early is basically like initializing to
-$Z=\frac{1}{1-\gamma}\Id$ which empirically diverges).
+$Z=\frac{1}{1-\gamma}\Id$ which empirically diverges). Still, one might
+use whether $\E_{s_2\sim \rho}z(s_1,s_2)$ is significantly above
+$\frac1{1-\gamma}$ as a divergence criterion (same with negative values).
 
@@
 
-It's a bit strange to train the policy for goal $s_2$ via policy gradient,
-instead of just selecting the next state to maximize
-$z(s,s_2;s_2)$ where $z(s,s_2;s_3)$ is the $z$ function associated with
-the policy that has goal $s_3$. Formally, computing the latter is
-feasible: one has to follow an exploration policy that has a non-constant
-goal,so at each step we observe a transition of the policy with some goal
-$s_3$; and hope that Newton--Bellman for computing $z(s,s_2;s_3)$
-generalizes well with respect to $z_3$.
+It's a bit strange to train the policy for goal $s_2$ via policy gradient
+on the policy with parameter $\omega(s_2)$, instead of just selecting the
+policy parameter $\omega$ to maximize $z(s,s_2;\omega)$ where
+$z(s,s_2;\omega)$ is the $z$ function associated with the policy that has
+goal $\omega$.  This is why we propose to train the function $s_2\mapsto
+\omega$ so that $\omega(s_2)$ is indeed the parameter that maximizes
+$z(s,s_2;\omega)$.
+To compute $z(s,s_2;\omega)$ one has to follow an
+exploration policy that has a non-constant goal, so at each step we
+observe a transition of the policy with some goal $\omega$; and hope that
+Newton--Bellman for computing $z(s,s_2;\omega)$ generalizes well with
+respect to $\omega$.
 
@@
 and representing $V$ by its density wrt $\rho$ namely
-$V_{s_1s_2}=z^\ast(s_1,s_2)\rho(s_2)$ this suggests the Newton method
+$V_{s_1s_2}=z^\ast(s_1,s_2)\rho(s_2)$ this suggests the Newton-like method
 \begin{multline}

### 2019-01-15 pendulum_exp/dataanalysis/plots.py
Biped, to be tested

@@ -3,6 +3,7 @@ import matplotlib
 import matplotlib.pyplot as plt
 import numpy as np
 from scipy.ndimage.filters import gaussian_filter1d
+from scipy.interpolate import interp1d
 from typing import List
 
 
@@ -78,9 +79,12 @@ def plot_learning_curves(expdata, key_list: List[str], namefile: str,
                 # ax.plot(dt*xdata, dt*ydata, c=c, alpha=0.2, linestyle=linestyle, linewidth=linewidth)
             elif gtype == 'run_std':
                 std_data = np.array([v for (i, v) in timeseq_stats['std'].items()])
-                x = xdata * dt
-                y = ydata * dt
-                std = std_data * dt
+                x = np.linspace(min(xdata), max(xdata), 400)
+                y = interp1d(xdata, ydata, kind='cubic')(x)
+                std = interp1d(xdata, std_data, kind='cubic')(x)
+                x = x * dt
+                y = y * dt
+                std = std * dt
                 ax.plot(x, y, label=label, c=c, alpha=alpha, linestyle=linestyle, linewidth=linewidth)
                 ax.fill_between(x, y - std / 2, y + std / 2, facecolor=c, alpha=0.2)
         first = False

### 2019-01-15 paper/icml_drau/framework.tex

@@
 	discrete, but discretization of an underlying continuous time
-	environment. For instance, many control environments, such as the
-	Mujoco~\cite{ddpg} environments or OpenAI Gym classic~!CITE! control
+	environment. For instance, many simulated control environments, such as the
+	Mujoco environments~\cite{ddpg} or OpenAI Gym classic\TODO{CITE!} control
 	environments are discretization of continuous time control problems,
-        \LB{with a discretization timestep $\deltat$ fixed by the simulator}.
+        with a discretization timestep $\deltat$ fixed by the simulator.
 	For real world environments, the discretization timestep, $\deltat$ is
-	most often imposed by the quality of sensors and actuators. For visual
-	simulated environments, it often corresponds to the number of frames
-	that the simulator generates per second. As $\deltat$ is not an
+	most often imposed by the quality of sensors and actuators. As $\deltat$ is not an
 	intrinsic parameter of the environment, but a constraint imposed by
@@
 	discretization timestep is small. We will refer to such environments as
-	\emph{near continuous environments}.
+	near continuous environments.
 
@@
 	along with $\deltat$ dependant definitions of return, discount factor and
-	value functions. With these definitions, each of these quantities is shown
-	to converge to a well defined continous time limit. Finally, it is shown
+	value functions that are shown to converge to well defined continous time limits. Finally, it is shown
 	that the state-action value function collapses to the value function as
-	$\deltat$ goes to $0$, and consequently that there is no Q-learning in
-	continuous time.
+	$\deltat$ goes to $0$. Consequently there is no Q-learning in
+	continuous time, foreshadowing problematic behaviors of near continuous time Q-learning.
       \TODO{We should explain somwhere here that, if not invariant, harder to find good hyperparameters}
@@
 
-\VC{
-  Consider ${\cal S} = \mathbb{R}^d$ a set of states, and ${\cal A}$ a set of
-	actions. A familly of \emph{Markov decision processes} ${\cal
-	M}_\deltat = \langle {\cal S}, {\cal A}, T_{\deltat}, r_\deltat,
-	\gamma_\deltat\rangle$ is \emph{near continous} if for any
-	deterministic policy $\pi:{\cal S} \rightarrow {\cal A}$, trajectories
-	of ${\cal M}_\deltat$ converge to stochastic processes that satisfy the
-	Stochastic differential equation
+% \VC{
+%   Consider ${\cal S} = \mathbb{R}^d$ a set of states, and ${\cal A}$ a set of
+% 	actions. A familly of \emph{Markov decision processes} ${\cal
+% 	M}_\deltat = \langle {\cal S}, {\cal A}, T_{\deltat}, r_\deltat,
+% 	\gamma_\deltat\rangle$ is \emph{near continous} if for any
+% 	deterministic policy $\pi:{\cal S} \rightarrow {\cal A}$, trajectories
+% 	of ${\cal M}_\deltat$ converge to stochastic processes that satisfy the
+% 	Stochastic differential equation
+% 	\begin{equation}
+% 		\label{eq:diffusion}
+% 		ds = F(s, a)dt + \Sigma(s, a)\dbt.
+% 	\end{equation}
+% 	Informally $F: {\cal S}\times{\cal A}\rightarrow {\cal S}$ describes
+% 	the deterministic dynamic of the environment, while $\Sigma(s, a)\dbt$
+% 	describes its stochasticity. Notably, $\Sigma = 0$ if the environment is
+% 	deterministic, as in classical control theory, or in many usual simulated
+% 	environments (Mujoco, Gym, ...).
+
+% 	In what follows, ${\cal M}_\deltat$ is taken to be a direct discretization of
+% 	Eq.~\ref{eq:diffusion} with discretization timestep $\deltat$. The
+% 	corresponding transition function $T_\deltat(., s, a)$ is the distribution of
+% 	$s'$ when starting from point $s$, and following Eq.~\ref{eq:diffusion} while
+% 	maintaining $a$ constant on an interval of time $\deltat$.
+
+%       }
+
+        Let ${\cal S} = \mathbb{R}^d$ be a set of states, and ${\cal A}$ be a set of
+	actions. Consider the continuous \emph{Markov Decision Process} (MDP) defined by the \emph{Stochastic Differential Equation} (SDE)\TODO{Explain the intuition of Brownian in SDE}:
         \begin{equation}
@@
               \end{equation}
-	Informally $F: {\cal S}\times{\cal A}\rightarrow {\cal S}$ describes
-	the deterministic dynamic of the environment, while $\Sigma(s, a)\dbt$
-	describes its stochasticity. Notably, $\Sigma = 0$ if the environment is
-	deterministic, as in classical control theory, or in many usual simulated
-	environments (Mujoco, Gym, ...).
-
-	In what follows, ${\cal M}_\deltat$ is taken to be a direct discretization of
-	Eq.~\ref{eq:diffusion} with discretization timestep $\deltat$. The
-	corresponding transition function $T_\deltat(., s, a)$ is the distribution of
-	$s'$ when starting from point $s$, and following Eq.~\ref{eq:diffusion} while
-	maintaining $a$ constant on an interval of time $\deltat$.
-
-      }
-
-      \LB{
-        [[What I don't like in that version is that it is too formal ...]]
-        Consider ${\cal S} = \mathbb{R}^d$ a set of states, and ${\cal A}$ a set of
-	actions. Assume that an agent is following the continuous Markov Decision Process defined by the Stochastic Differential Equation:
-        \begin{equation}
-		\label{eq:diffusion}
-		ds = F(s, a)dt + \Sigma(s, a)\dbt.
-              \end{equation}
-        Then, for any $\deltat$, we say that the discrete \emph{Markov decision processes} ${\cal
+        For any $\deltat$, the discrete MDP ${\cal
 	M}_\deltat = \langle {\cal S}, {\cal A}, T_{\deltat}, r_\deltat,
-      \gamma_\deltat\rangle$ is a discretization of the continuous MDP with \emph{time discretization} $\deltat$ if the transition probability $T_\deltat(s, a, s')$ is the probability distribution of $s_\deltat$, with $s_0 = s$ and the action is $a_t=a$ during the interval $[0, \deltat]$.
-We say the the MDP ${\cal M}_\deltat$ is \emph{near-continuous}.
-[[Pb here: we require an uderlying continuous process: there is no intrinsic near continuous process]]
+      \gamma_\deltat\rangle$ is a discretization of the continuous MDP with \emph{time discretization} $\deltat$.
+The transition probability $T_\deltat(s, a, s')$ is the probability distribution of $s_\deltat$, when starting at $s_0 = s$ and maintaining $a_t=a$ constant for a time $\deltat$.
+      \footnote{Informally, $T_\deltat(s, a, s')$ is approximately the probability distribution of $s + F(s, a)\deltat + \sqrt{\deltat}\Sigma(s, a){\cal N}(0, I)$, which is the discretization of Eq.
+\eqref{eq:diffusion} \TODO{CITE}}.
+This corresponds to an agent evolving in the continuous environment described by Eq.
+\eqref{eq:diffusion}, but who is only making observations and choosing actions every $\deltat$.
+We say the the MDP ${\cal M}_\deltat$ is near-continuous.
 
@@
 	deterministic, as in classical control theory, or in many usual simulated
-	environments (Mujoco, Gym, ...).
-      }
+	environments.
 
-        In this paper, we try to design RL methods for \emph{near continuous} MDP
-	robust to the choice of $\deltat$. Our method is to analyse the different
-	quantities we are looking at when $\deltat$ is changing, and especially when
-	$\deltat \rightarrow 0$, and make sure that the quantities we are interested in
-	and our algorithms do converge to meaningful limit quantities and algorithms.\TODO{Better formulation maybe?}
+        In this paper, we try to design RL methods for near continuous MDP
+	robust to the choice of $\deltat$. For robustness, a necessary condition is to remain viable when $\deltat \rightarrow 0$. Our method is thus to analyse the different
+	quantities involved when $\deltat \rightarrow 0$, and make sure that they converge to meaningful limits.
 
-        \LB{In that direction, the first thing to do is to check that the discretized
-          trajectory do converge to the continuous ones. This is stated in the following theorem. [[Tres tres mal dit, mais c'est l'idee]]
-        }
-        % The first formal statement we can make on discretized MDP with a
-	% time discretization $\deltat$ is that if $\deltat \rightarrow 0$, the
-	% distribution of trajectories do converge to the continuous trajectories.
+        The first step in that direction is to check that the discretized
+          trajectories converge to the continuous ones.
 
@@
 		$(s_\deltat^k)$ be the discrete process obtained if the agent only
-		get observations every $\deltat$ !explain better!.
+		get observations every $\deltat$ \TODO{explain better}.
 
@@
 		\frac{t}{\deltat}\rfloor}_\deltat$ for
-		$t\in\mathbb{R_+}$.!Explain!. Then:
+		$t\in\mathbb{R_+}$.\TODO{Explain}. Then:
 		\begin{equation}
@@
 
-In this section, we show how to define the discount factor $\gamma_\deltat$ as a function of $\deltat$, such that the return and Value function do converge to the desired limit when $\deltat \rightarrow 0$.
-
-\VC{
-	In what follows, $\deltat$-dependant definitions of the discount factor
-	and the reward are given. These definitions fit the discrete case
+	In what follows, $\deltat$-dependant definitions of the discount factor $\gamma_\deltat$
+	and the reward $r_\deltat$ are given. These definitions fit the discrete case
 	scenario when $\deltat=1$, and yields well defined, non trivial returns
 	and value functions when $\deltat$ goes to $0$.
-}
 
@@
 
-\paragraph{Microscopic/Macroscopic time, Time horizon}
-\VC{In near continuous environments, there are two notions of time,
-	the number of observation received, or the number of actions taken,
-	which we call the microscopic time, and the actual amount of time spent
-	in the underlying continuous time environment, which we call the macroscopic
-	time. The two are related through
-	\begin{equation}
-		\tau^\text{macro} = \tau^\text{micro} \deltat.
-	\end{equation}
+% \paragraph{Microscopic/Macroscopic time, Time horizon}
+% \VC{In near continuous environments, there are two notions of time,
+% 	the number of observation received, or the number of actions taken,
+% 	which we call the microscopic time, and the actual amount of time spent
+% 	in the underlying continuous time environment, which we call the macroscopic
+% 	time. The two are related through
+% 	\begin{equation}
+% 		\tau^\text{macro} = \tau^\text{micro} \deltat.
+% 	\end{equation}
 
-	The notion of time horizon, which is, informally, the amount of time on
-	which the agent optimizes its return, is only meaningful in macroscopic
-	time. Having a time horizon of $10$ steps means something very different
-	if $\deltat = 0.1$ and if $\deltat = 0.0001$. A rule of thumb
-	is that the time horizon of an agent with discount factor $\gamma$ is
-	of order $\frac{1}{1 - \gamma}$. If $\gamma_\deltat$ was left constant
-	as $\deltat$ goes to $0$, the corresponding macroscopic time horizon 
-	would write as $\frac{\deltat}{1 - \gamma}$ which goes to $0$ when
-	$\deltat$ goes to $0$.
+% 	The notion of time horizon, which is, informally, the amount of time on
+% 	which the agent optimizes its return, is only meaningful in macroscopic
+% 	time. Having a time horizon of $10$ steps means something very different
+% 	if $\deltat = 0.1$ and if $\deltat = 0.0001$. A rule of thumb
+% 	is that the time horizon of an agent with discount factor $\gamma$ is
+% 	of order $\frac{1}{1 - \gamma}$. If $\gamma_\deltat$ was left constant
+% 	as $\deltat$ goes to $0$, the corresponding macroscopic time horizon 
+% 	would write as $\frac{\deltat}{1 - \gamma}$ which goes to $0$ when
+% 	$\deltat$ goes to $0$.
 
-	On the contrary, the definition of $\gamma_\deltat$ in
-	Eq.~\eqref{eq:def-gamma} yields an approximately constant macroscopic
-	time horizon when $\deltat$ goes to $0$. Indeed,
-	\begin{equation}
-		\label{eq:time-horizon}
-		\frac{\deltat}{1-\gamma^\deltat}
-		= - \frac{1}{\log \gamma} + \bigO(\deltat)
-		\approx \frac{1}{1-\gamma}
-	\end{equation}
-      }
+% 	On the contrary, the definition of $\gamma_\deltat$ in
+% 	Eq.~\eqref{eq:def-gamma} yields an approximately constant macroscopic
+% 	time horizon when $\deltat$ goes to $0$. Indeed,
+% 	\begin{equation}
+% 		\label{eq:time-horizon}
+% 		\frac{\deltat}{1-\gamma^\deltat}
+% 		= - \frac{1}{\log \gamma} + \bigO(\deltat)
+% 		\approx \frac{1}{1-\gamma}
+% 	\end{equation}
+%       }
 
-      \LB{ \paragraph{Number of steps / Physical time: Time horizon}
+      \paragraph{Number of steps / Physical time, Time horizon}
         % [[I prefer to talk about ``number of steps'' VS ``Physical time'' than microscopic/Macroscopic]]
@@
 	is that the time horizon of an agent with discount factor $\gamma$ is
-	of order $\frac{1}{1 - \gamma}$. If $\gamma_\deltat$ was left constant
-	as $\deltat$ goes to $0$, the corresponding macroscopic time horizon 
+	of order $\frac{1}{1 - \gamma}$ steps. If $\gamma_\deltat$ was left constant
+	as $\deltat$ goes to $0$, the corresponding time horizon in physical time
 	would write as $\frac{\deltat}{1 - \gamma}$ which goes to $0$ when
 	$\deltat$ goes to $0$.
-
         On the contrary, the definition of $\gamma_\deltat$ in
-	Eq.~\eqref{eq:def-gamma} yields an approximately constant macroscopic
-	time horizon when $\deltat$ goes to $0$. Indeed,
+	Eq.~\eqref{eq:def-gamma} yields an approximately constant
+	time horizon in physical time when $\deltat$ goes to $0$. Indeed,
 	\begin{equation}
@@
           = - \frac{1}{\log \gamma} + \bigO(\deltat)
-          \approx \frac{1}{1-\gamma}
+          \approx \frac{1}{1-\gamma},
         \end{equation}
-
-      }
+        approximately invariant to $\deltat$ when $\deltat$ goes to 0.
 
@@
 
-\LB{I did not change this part}
 The definition of the return Eq.~\eqref{eq:continuous-return}, yields the
 following definition of the continuous value function
-\begin{equation}
-  V^\pi(s) = \E_{\tau\sim\pi}\left[\int\limits_{0}^\infty \gamma^{t}\deltat r_{t} \mid s_0 = s\right].
-\end{equation}
+\begin{align}
+  V^\pi(s) =& \E_{\tau\sim\pi}\left[R(\tau) \mid s_0 = s\right]\\
+  =& \E_{\tau\sim\pi}\left[\int\limits_{0}^\infty \gamma^{t}\deltat r_{t} \mid s_0 = s\right].
+\end{align}
 The corresponding discretization for time discretization $\deltat$ is
-\begin{equation}
-  V^\pi_\deltat(s) = \E_{\tau\sim\pi}\left[\sum\limits_{k=0}^\infty \gamma^{k\deltat}\deltat r_{k\deltat} \mid s_0 = s\right]
-\end{equation}
+\begin{align}
+  V^\pi_\deltat(s) =& \E_{\tau\sim\pi}\left[R_\deltat(\tau) \mid s_0 = s\right]\\
+  =& \E_{\tau\sim\pi}\left[\sum\limits_{k=0}^\infty \gamma^{k\deltat}\deltat r_{k\deltat} \mid s_0 = s\right]
+\end{align}
 which fits the usual definition of the discrete V-function for $\gamma_\deltat$ and
@@
   \begin{theorem}
-    We have, for all $s \in {\cal S}$:
+   For all $s \in {\cal S}$:
     \begin{equation}
       \label{eq:conv-value}
-      \lim_{\deltat \rightarrow 0} V^\pi_\deltat(s) = V^\pi(s)
+      V^\pi_\deltat(s) = V^\pi(s) + \bigO(\deltat)
     \end{equation}
@@
   \label{eq:bellman}
-  V^\pi_\deltat(s) = \reward(s, \pi(s))\deltat + \gamma^{\dt} \E_{s_{t+\deltat} | s_t = s} V^\pi(s_{t+\deltat})
+  V^\pi_\deltat(s) = \reward(s, \pi(s))\deltat + \gamma^{\deltat} \E_{s_{\deltat}^{k+1} | s_{\deltat}^{k} = s} V^\pi_\deltat(s_{\deltat}^{k+1})
 \end{equation}
 
-If the MDP is following the equation (\ref{eq:diffusion}), the limit of the Bellman equation for $V^\pi_\deltat$ when $\deltat \rightarrow 0$ is the Hamilton-Jacobi-Bellman equation on $V^\pi$:
+If the MDP is following the Eq.
+\eqref{eq:diffusion}, the limit of the Bellman equation for $V^\pi_\deltat$ when $\deltat \rightarrow 0$ is the Hamilton-Jacobi-Bellman equation on $V^\pi$:\TODO{Cite ?}
 \begin{equation}
@@
 Contrary to the Value function, the Action-Value function $Q$ is ill-defined
-for a continuous MDP. More precisely, for a continuous MDP, $Q^\pi(s, a) =
-V^\pi(s)$, so the continuous $Q$-function is independant of the actions. For a
-\emph{near-continuous} MDP, this means that the $Q$-function is very close  to the
-$V$-function.\TODO{Say more ?} Moreover, this means that the discretized $Q$
-function \TODO{explain}. This can be stated formally.
+for a continuous MDP. More precisely, the Q-function  collapses to the V-function when $\deltat \rightarrow 0$. In near continuous time,
+the effect of individual actions on the Q-function is of order $\bigO(\deltat)$. This makes deriving a hierarchy of action from an approximate Q-function difficult.
+Formally:
 \begin{theorem}
@@
 $Q^\pi$ does not bear any information on the hierarchy of actions, and
-thus cannot be used to select actions that yields higher future returns.
+thus cannot be used to select actions that yields higher future returns. There is no time continuous Q-learning.
 
@@
   \begin{align}
-    Q^\pi(s, a) &= r(s, a)\deltat + \gamma^\deltat \E_{s'|s, a}\left[V(s')\right] \\
+    Q^\pi_\deltat(s, a) &= r(s, a)\deltat + \gamma^\deltat \E_{s'|s, a}\left[V^\pi_\deltat(s')\right] 
   \end{align}
   But we know that $s'$ is close to $s$. If we define $\delta s = s' - s$, then:
+  \begin{equation}
+    \E_{s'|s, a}\left[\delta s\right] = \bigO(\deltat) \quad \text{Var}(\delta s) = \bigO(\deltat)
+  \end{equation}
+  \TODO{Explain. And this is not exactly the Variance.} Therefore:
   \begin{align}
-    \|\E_{s'|s, a}\left[\delta s\right]\| &= \bigO(\deltat) \\
-    \E_{s'|s, a}\left[\|\delta s\|^2\right] &= \bigO(\deltat)
-  \end{align}
-  \TODO{Explain.} Therefore:
-  \begin{align}
-    |\E_{s'|s, a}\left[V(s')\right] - V(s)| &= \E_{\delta s|s, a}\left[\partial_s V\cdot \delta s +\bigO(\|\delta s\|^2)\right]\\
-                                            &\leq \|\partial_s V\|\cdot \|\E_{\delta s|s, a}\left[\delta s\right]|\| +\bigO(\delta t)\\
-    &= \bigO(\deltat)
+    |\E_{s'|s, a}\left[V^\pi_\deltat(s')\right] - V^\pi_\deltat(s)| &= \E_{\delta s|s, a}\left[\partial_s V^\pi_\deltat\cdot \delta s +\bigO(\|\delta s\|^2)\right]\nonumber\\
+                                            &\leq \|\partial_s V_\deltat^\pi\|\cdot \|\E_{\delta s|s, a}\left[\delta s\right]\| +\bigO(\delta t)\nonumber\\
+    &= \bigO(\deltat)\nonumber
   \end{align}
@@
   \begin{align}
-    Q^\pi(s, a) &= r(s, a)\deltat + \gamma^\deltat \E_{s'|s, a}\left[V(s')\right] \\
-                &= r(s, a)\deltat + \left(1 + \bigO(\deltat)\right)(V(s) + \bigO(\deltat))\\
-    &= V(s) + \bigO(\deltat)
+    Q^\pi_\deltat(s, a) &= r(s, a)\deltat + \gamma^\deltat \E_{s'|s, a}\left[V_\deltat^\pi(s')\right] \nonumber\\
+                &= r(s, a)\deltat + \left(1 + \bigO(\deltat)\right)(V^\pi(s) + \bigO(\deltat))\nonumber\\
+    &= V^\pi(s) + \bigO(\deltat)\nonumber
   \end{align}

### 2019-01-15 paper/tangotd.tex

@@
 \begin{equation}
-\partial \pi(a|s,\Omega(s_2))\E_{s'|s,a} \E_{s_2\sim \rho} \,
+\partial \pi(a|s,\Omega(s_2))\E_{s'|s,a} 
 (z(s',s_2;\Omega(s_2))-z(s,s_2;\Omega(s_2)))
@@
 Newton--Bellman equation for $z^\ast$ (see below) does capture
-composition of optimal actions, though.
+composition of optimal actions, though. TODO but I think I can prove that
+if $\Omega(s_2)$ contains the list of all optimal action paths to reach
+$s_2$, and the policy given $\Omega(s_2)$ follows that path, then
+$z(s,s_2;\Omega(s_2))=z^\ast(s,s_2)$; therefore, there would be no loss
+from working with the $z(s,s_2;\omega)$ formalism if we effectively find
+a solution to all equations. Still treats every $\omega$ as a separate
+problem though.
 
@@
 \end{multline}
-satisfied by the optimal $z$ function. This is not quite an instance of
-the Newton method for matrix inversion anymore (because the operator
-$\Id-\gamma \max_a \E_{\cdot|a}$ is not linear), so convergence
-is unclear.
+satisfied by the optimal $z$ function. A problem with this is how to
+estimate the max of an expectation; we would have to work with
+$Q$-function analogues, i.e., estimating $z^\ast(s_3,a,s_2)$. The
+advantage of this over working with $z(s_3,s_2;\omega)$ is unclear (I
+prefer goals/action plans to individual actions).
+
+This is not quite an instance of the Newton method for matrix inversion
+anymore (because the operator $\Id-\gamma \max_a \E_{\cdot|a}$ is not
+linear), so convergence is unclear.
 

### 2019-01-15 pendulum_exp/dataanalysis/plots.py
plots details

@@ -71,8 +71,7 @@ def plot_learning_curves(expdata, key_list: List[str], namefile: str,
 
                 if first:
                     ax.legend()
-                ax.set_xlim(mint, maxt)
-                ax.set_xlabel("time (second)")
+                
                 ax.plot(x, y, label=label, c=c, alpha=alpha, linestyle=linestyle, linewidth=linewidth)
                 ax.fill_between(x, y-sigma, y+sigma, facecolor=c, alpha=0.2)
                 # ax.plot(dt*xdata, dt*ydata, c=c, alpha=0.2, linestyle=linestyle, linewidth=linewidth)
@@ -82,7 +81,12 @@ def plot_learning_curves(expdata, key_list: List[str], namefile: str,
                 y = ydata * dt
                 std = std_data * dt
                 ax.plot(x, y, label=label, c=c, alpha=alpha, linestyle=linestyle, linewidth=linewidth)
-                ax.fill_between(x, y - std / 2, y + std / 2, facecolor=c, alpha=0.2)
+                ax.fill_between(x, y - std, y + std, facecolor=c, alpha=0.2)
+
+
+                ax.set_xlim(mint, maxt)
+                ax.set_xlabel("Physical time")
+                # ax.fill_between(x, y - std / 2, y + std / 2, facecolor=c, alpha=0.2)
         first = False
 
     plt.tight_layout()

### 2019-01-15 pendulum_exp/dataanalysis/script_leonard.py

@@ -6,9 +6,9 @@ from dataloader import loader_leonard, ExperimentData
 
 from plots import plot_learning_curves
 
-start_date = datetime.strptime('2019_01_09_05_44_37', "%Y_%m_%d_%H_%M_%S")
+# start_date = datetime.strptime('2019_01_09_05_44_37', "%Y_%m_%d_%H_%M_%S")
 # stop_date = datetime.strptime('2019_01_08_03_13_35', "%Y_%m_%d_%H_%M_%S")
-# start_date = 'last'
+start_date = 'last'
 stop_date = None
 runlist = loader_leonard('/private/home/leonardb/workdir', 'mujoco_continuous', 
     start_date=start_date, stop_date=stop_date)
@@ -18,14 +18,14 @@ expdata = ExperimentData(runlist)
 print(expdata.deltakeys)
 
 expdata_ant = expdata.filter_settings(lambda s: s['env_id'] == 'ant')
-plot_learning_curves(expdata_ant, ['Return', 'Return'], 'ant')
+plot_learning_curves(expdata_ant, ['Return', 'Return'], 'ant', gtype='run_std', mint=0, maxt=20)
 
-expdata_cartpole = expdata.filter_settings(lambda s: s['env_id'] == 'cartpole')
-plot_learning_curves(expdata_cartpole, ['Return', 'Return'], 'cartpole')
+# expdata_cartpole = expdata.filter_settings(lambda s: s['env_id'] == 'cartpole')
+# plot_learning_curves(expdata_cartpole, ['Return', 'Return'], 'cartpole')
 
-expdata_cheetah = expdata.filter_settings(lambda s: s['env_id'] == 'half_cheetah')
-plot_learning_curves(expdata_cheetah, ['Return', 'Return'], 'cheetah')
+# expdata_cheetah = expdata.filter_settings(lambda s: s['env_id'] == 'half_cheetah')
+# plot_learning_curves(expdata_cheetah, ['Return', 'Return'], 'cheetah')
 
 
-expdata_tau = expdata.filter_settings(lambda s: s['algo'] == 'discrete_value')
-expdata_tau.repr_rawlogs("Return", 5)
+# expdata_tau = expdata.filter_settings(lambda s: s['algo'] == 'discrete_value')
+# expdata_tau.repr_rawlogs("Return", 5)

### 2019-01-15 paper/icml_drau/advup.tex

--- 2019-01-11 paper/icml_drau/advup.tex
@@
 $Q^\pi_\deltat(s, a)$ and $V^\pi\deltat(s)$ is of order $\bigO(\deltat)$.
-Contrary to the action state value function, the rescaled advantage function converges
-to an action dependent quantity when $\deltat$ goes to $0$
-\begin{align}
-	A^\pi_\deltat(s, a)=&\reward(s, a)+\frac{\gamma^{\deltat}\E_{\delta s \sim \delta s(s, a)}V^\pi_\deltat(s + \delta s)-V^\pi\deltat(s)}{\deltat}\nonumber\\
-	=&\reward(s, a)+\log(\gamma) V^\pi(s)+\partial_s V^\pi(s)f(s, a)\nonumber\\
-         &+\frac{\text{tr}\left(\Sigma^T(s, a)\partial^2_{s^2}V^\pi(s) \Sigma(s,a)\right)}{2} + O(\deltat)\nonumber.
-\end{align}
-The discretized state action value function can then be rewritten as
+Contrary to the action state value function, the rescaled advantage function converges when $\deltat$ goes to $0$
+to an action dependent quantity which contains the necessary information for policy improvement:
+
+% \LB{In next equation, tiny detail but I would write $\frac{1}{\deltat}(r\deltat+...)$, easier to read.}
+% \begin{align}
+% 	A^\pi_\deltat(s, a)=&\reward(s, a)+\frac{\gamma^{\deltat}\E_{\delta s \sim \delta s(s, a)}V^\pi_\deltat(s + \delta s)-V^\pi_\deltat(s)}{\deltat}\nonumber\\
+% 	=&\reward(s, a)+\log(\gamma) V^\pi(s)+\partial_s V^\pi(s)f(s, a)\nonumber\\
+%          &+\frac{\text{tr}\left(\Sigma^T(s, a)\partial^2_{s^2}V^\pi(s) \Sigma(s,a)\right)}{2} + O(\deltat)\nonumber.
+% \end{align}
+
+  \begin{theorem}
+    There exists $A^\pi(s, a)$ such that for all $a, s$:
+    \begin{equation}
+      A^\pi_\deltat(s, a) \rightarrow_{\deltat \rightarrow 0} A^\pi(s, a)
+    \end{equation}
+    Moreover, $A^\pi$ contains all the necessary information for policy improvement. The policy $\pi$ optimal if and only if $A^\pi(s, \pi(s)) = \max_{a'}A^\pi(s, a')$ for all $s \in {\cal S}$.
+    \TODO{Is this true? What is true? Is this real life? We should probably add a reasonable hypothesis, for example $\pi_\deltat^* \rightarrow \pi^*$}
+  \end{theorem}
+
+The discretized Q-function can be rewritten as
 \begin{equation}
@@
 \end{equation}
-A natural way of approximating $V^\pi_\delta$ and $A^\pi_\deltat$ is to apply
+A natural way of approximating $V^\pi_\deltat$ and $A^\pi_\deltat$ is to apply
 \emph{Sarsa} or Q-learning to a reparameterized Q-function approximator
@@
 
-However, the proposed reparameterization does not, on its own, guarantee that
+\TODO{CT and LB are fighting each other about the importance which should be given to the next paragraph.
+A impartial judge should give his opinion.
+(CT is wrong)} However, the proposed reparameterization does not, on its own, guarantee that
 when $Q$ correctly approximates $Q^\pi_\deltat$, $A$ approximates $A^\pi_\deltat$.
@@
 \begin{equation}
-	a^k_\deltat = \pi^\text{explore}_\deltat(s^k_\deltat, \nu^k_\deltat)
+  \label{eq:explore}
+	a^k_\deltat = \pi^\text{explore}_\deltat(s^k_\deltat, z^k_\deltat)
 \end{equation}
-with $(\nu^k_\deltat)_k$ a sequence of random variables, independant from the $a$'s and $s$'s.
-For instance, $\varepsilon$-greedy exploration naturally falls in this framework, with each $\nu^k_\deltat$
+with $(z^k_\deltat)_k$ a sequence of random variables, independant from the $a$'s and $s$'s.
+For instance, $\varepsilon$-greedy exploration naturally falls in this framework, \TODO{[[CT would like to give these details, LB thinks it is not necessary:]] with each $z^k_\deltat$
 being a pair composed of a bernoulli random variables $b^k_\deltat$ with probability of being $1$ equal to $\varepsilon$,
@@
 	a^k_\deltat = (1 - b^k_\deltat) \text{argmax}_{a'} Q^\pi_\deltat(s, a') + b^k_\deltat \xi^k_\deltat.
-\end{equation}
+\end{equation}}
 
 A sufficient condition for the exploratory policy to admit a time continuous
-limit is for $\pi^\text{explore}_\deltat$ to be a measurable function, and for
-$\nu_\deltat$ to admit one, i.e. that the sequence $\nu_\deltat$ converges to a
-well defined continuous stochastic process $\nu$ as $\deltat$ goes to $0$.
+limit is for
+$z_\deltat$ to admit a limit, i.e. that the sequence $z_\deltat$ converges to a
+well defined continuous stochastic process $z$ as $\deltat$ goes to $0$.
 \TODO{I'm convinced that this is true, reviewers might not be.}
 
-The exploration scheme used~\cite{ddpg}, which consists in adding an
-\emph{Ornstein-Uhlenbeck}~\cite{orn-uhl} process (OU process) to actions, is naturally time.
-invariant. It can be rewritten as
+The exploration scheme used in~\cite{ddpg} for continuous actions, which consists in adding an
+\emph{Ornstein-Uhlenbeck}~\cite{orn-uhl} process (OU process) to actions, is naturally time
+invariant. It follows Eq. \eqref{eq:explore} with:
 \begin{equation}
-	a^k_\deltat = \pi(s^k_\deltat) + \nu^k_\deltat
+	\pi^\text{explore}(s_t, z_t) = \pi(s_t) + z_t
 \end{equation}
-where $\nu^k_\deltat$ is the discretization of a time continuous \emph{Ornstein-Uhlenbeck}
-process, defined as
+and $z_t$ a time continuous OU process, defined as
 \begin{equation}
-	d\nu_t = - \nu_t \theta dt + \sigma \dbt.
+	dz_t = - z_t \theta dt + \sigma \dbt.
 	\label{eq:orn_uhl}
@@
 On the other hand, the same argument cannot be applied to $\varepsilon$-greedy
-exploration, as the $\nu_\deltat$'s used to define it do not converge to a 
+exploration, as the $z_\deltat$'s used to define it do not converge to a 
 well defined stochastic process, but to a white noise.\footnote{\TODO{$\varepsilon$-greedy
@@
 To obtain a temporally coherent exploration scheme in discrete action environment
-$\nu_\deltat$ is chosen to be a discretization of a $|\mathcal{A}|$ dimensional
+$z_\deltat$ is chosen to be a discretization of a $|\mathcal{A}|$ dimensional
 continuous OU process, and set
 \begin{equation}
-	a^k_\deltat = \text{argmax}_{a'} \left(Q^\pi_\deltat(s, a') + \nu_\deltat^k[a']\right)
+  \pi^\text{explore}(s_t, z_t)= \text{argmax}_{a'} \left(Q(s, a') + z_t[a']\right)
 \end{equation}
-where $\nu_\deltat^k[a']$ denotes the $a'$-th component of the discretized OU process at time
+where $z_t[a']$ denotes the $a'$-th component of the discretized OU process at time
 $k$. The resulting exploration scheme converges to a proper time continuous exploration scheme.
@@
 for a greedy exploitation policy $\pi(s) = \text{argmax}_{a'}A^\pi_\deltat(s,
-a')$, exactly implemented in the case of discrete actions, and approximated by
-a neural network $\pi_\phi(s)$, trained to maximize (an approximation of)
-$A^\pi_\deltat(s, \pi_\phi(s))$ in the continuous case.  Eq.~\eqref{eq:max_A}
-is directly verified by $A_{\psi}$, owing to the reparameterization
+a')$. The argmax is exactly implemented in the case of discrete actions, and approximated by
+a neural network $\pi_\phi(s)$, trained to maximize
+$A(s, \pi_\phi(s))$ in the continuous case.  This is similar to what is done in \cite{ddpg}.
+
+Eq.~\eqref{eq:max_A}
+is directly verified by $A_{\psi}$, owing to the reparametrization
 $A_\psi(s, a) = \bar{A}_\psi(s, a) - \bar{A}_\psi(s, \pi(s))$, described
@@
 corresponding squared residual is minimized by an approximate gradient descent.
-The corresponding update equations when learning from a transition $(s, a, r, s')$ of
+The update equations when learning from a transition $(s, a, r, s')$ of
 an exploratory trajectory are
@@
 	\delta Q_\deltat &\leftarrow A(s, a) \deltat - \left(r \deltat + \gamma^{\deltat} V(s') - V(s)\right)\\
-	\theta_\deltat &\leftarrow \theta_\deltat + \eta^1_\deltat \partial_{\theta} V(s) \frac{\delta Q_\deltat}{\deltat}\\
-	\psi_\deltat &\leftarrow \psi_\deltat + \eta^2_\deltat \partial_{\psi} A(s, a) \frac{\delta Q_\deltat}{\deltat}.
+	\theta_\deltat &\leftarrow \theta_\deltat + \eta^V_\deltat \partial_{\theta} V(s) \frac{\delta Q_\deltat}{\deltat}\\
+	\psi_\deltat &\leftarrow \psi_\deltat + \eta^A_\deltat \partial_{\psi} A(s, a) \frac{\delta Q_\deltat}{\deltat}.
 	\label{eq:approx_bellman_A}
 \end{align} 
-where $\eta$'s are learning rates. As we will see in the following section,
-to obtain a well defined continuous time limit, the learning rates can be rewritten as
-$\eta^i_\deltat = \alpha^i \deltat$, where $\alpha^1$ and $\alpha^2$ are independant
-of $\deltat$. In practice, $A$ and $V$ are learned using
+where $\eta$'s are learning rates.
+Appropriate scalings for the learning rates $\eta^V_\deltat$ and $\eta^A_\deltat$ in term of $\deltat$ to obtain a well defined continuous limit are derived in the following section.
+\TODO{Change notations for the learning rates in the algorithm}
+% As we will see in the following section,
+% to obtain a well defined continuous time limit, the learning rates can be rewritten as
+% $\eta^i_\deltat = \alpha^i \deltat$, where $\alpha^1$ and $\alpha^2$ are independant
+% of $\deltat$.
+In practice, $A$ and $V$ are learned using
 transitions sampled from a replay buffer instead of trajectory samples, to
@@
 $\deltat$ goes to $0$.  This in turn imposes conditions on the scalings of the
-learning rates of invariant algorithms with respect to the discretization
-timestep.
+learning rates of algorithms robust to changes of the tine discretization.
 
 The updates equations on $\theta$ and $\psi$ are described in
-Eq.~\eqref{eq:approx_bellman_A}.  To have parameter steps that do not grow
-larger or shrink to $0$ as $\deltat \leftarrow 0$, both $\delta \theta^1$ and
-$\delta \theta^2$ should verify stochastic differential equations, and notably
+Eq.~\eqref{eq:approx_bellman_A}.  To have parameter steps that do not explode or shrink to $0$ too
+fast as $\deltat \rightarrow 0$, both $\delta \theta$ and
+$\delta \psi$ should verify stochastic differential equations, and notably
 have a deterministic component of order $\deltat$. As $\deltat$ goes to $0$,
@@
 \begin{align}
-	\delta Q_\deltat &= A(s, a)\deltat - r\deltat - \log(\gamma) V(s)\deltat - \partial_s V(s) \delta s\nonumber \\&- \frac{(\delta s)^T \partial_{s^2}^2 V(s) \delta s}{2} + \smallo\left(\sqrt{\deltat}\right)\\
-	\delta \theta_\deltat &= \eta^1_\deltat \partial_{\theta} V(s) \frac{\delta Q_\deltat}{\deltat} + \smallo\left(\frac{\eta^1_\deltat}{\sqrt{\deltat}}\right)\\
-	\delta \psi_\deltat &= \eta^2_\deltat \partial_{\psi} A(s) \frac{\delta Q_\deltat}{\deltat} + \smallo\left(\frac{\eta^2_\deltat}{\sqrt{\deltat}}\right).
+	\delta Q_\deltat &= A(s, a)\deltat - r\deltat - \log(\gamma) V(s)\deltat - \partial_s V(s) \delta s\nonumber \\&- \frac{(\delta s)^T \partial_{s^2}^2 V(s) \delta s}{2} + \smallo\left(\deltat\right)\\
+	\delta \theta_\deltat &= \eta^V_\deltat \partial_{\theta} V(s) \frac{\delta Q_\deltat}{\deltat} + \smallo\left(\eta^V_\deltat\right)\\
+	\delta \psi_\deltat &= \eta^A_\deltat \partial_{\psi} A(s) \frac{\delta Q_\deltat}{\deltat} + \smallo\left(\eta^A_\deltat\right).
 \end{align}
-For parameter updates to follow a stochastic differential equations, one must
-have $\eta^1_\deltat = \alpha^1 \deltat$ and $\eta^2_\deltat = \alpha^2
-\deltat$. With this choice of learning rates, the limit SDEs are
-\begin{align}
-	F &= A(s_t, a_t) - r_t - \log(\gamma) V(s_t) - \partial_s V(s_t) f(s_t, a_t)\nonumber\\
-	G &= - \partial_s V(s_t, a_t) \Sigma(s_t, a_t)\nonumber\\
-	d\theta_t &= (\alpha^1 F  + D \dbt)\partial_{\theta} V(s_t)\nonumber\\
-	d\psi_t &= (\alpha^2 F  + D \dbt)\partial_{\psi} A(s_t, a_t)\nonumber.
-\end{align}
+
+For parameter updates to follow a SDE, one must
+have $\eta^V_\deltat = \alpha^V \deltat$ and $\eta^A_\deltat = \alpha^A
+\deltat$:
+\begin{theorem}
+  If $\eta^V_\deltat = \alpha^V \deltat$ and $\eta^A_\deltat = \alpha^A\deltat$, the continuous limit trajectory for parameters follows an SDE.
+\end{theorem}
+
+% With this choice of learning rates, the limit SDEs are
+% \begin{align}
+% 	F &= A(s_t, a_t) - r_t - \log(\gamma) V(s_t) - \partial_s V(s_t) f(s_t, a_t)\nonumber\\
+% 	G &= - \partial_s V(s_t, a_t) \Sigma(s_t, a_t)\nonumber\\
+% 	d\theta_t &= (\alpha^V Fdt  + D \dbt)\partial_{\theta} V(s_t)\nonumber\\
+% 	d\psi_t &= (\alpha^A Fdt  + D \dbt)\partial_{\psi} A(s_t, a_t)\nonumber.
+% \end{align}
 The full algorithm can now be derived.

### 2019-01-15 paper/icml_drau/intro.tex

@@
 \begin{itemize} 
-\item We formally show that the Q-function collapses to the V-function in \emph{near-continuous} time, and thus that
+\item We formally show that the Q-function collapses to the V-function in near-continuous time, and thus that
     standard Q-learning is ill-behaved in this setting.

### 2019-01-16 pendulum_exp/parse.py
add adam to optimizers

@@ -36,7 +36,7 @@ def setup_args():
     parser.add_argument('--beta', type=float, default=None)
     parser.add_argument('--tau', type=float, default=.99)
     parser.add_argument('--noreference', action='store_true')
-    parser.add_argument('--optimizer', type=str, choices=['sgd', 'rmsprop'],
+    parser.add_argument('--optimizer', type=str, choices=['sgd', 'rmsprop', 'adam'],
                         default='sgd')
     parser.add_argument('--noreload', action='store_true')
     parser = argload.ArgumentLoader(parser, to_reload=[

### 2019-01-16 paper/icml_drau/icml_drau.tex
add tmp title
Commiting

@@
 % Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Deep Relativistic Advantage Updating}
+\icmltitlerunning{Making Q-learning robust to time discretization}
 
@@
 \newtheorem{proposition}{Proposition}
+\newtheorem{result}{Result}
 
@@
 \twocolumn[
-\icmltitle{There are no small Advantages:
-	Reinforcement learning in near continuous time
-	with Deep RelAtive Mixture of Advantages
-}
+\icmltitle{Making Q-learning robust to time discretization}
 
@@
 \begin{abstract}
-	Despite remarkable successes, \emph{Deep Reinforcement learning} (DRL) remains sensitive to
+	Despite remarkable successes, \emph{Deep Reinforcement learning} (DRL) remains not robust to
 	hyperparameterization, implementation details and small environment changes~(\citealt{drl_matter}, \citealt{drl_matter_bis}). 

### 2019-01-16 paper/icml_drau/intro.tex
Commiting

@@
 In recent years, \emph{Deep Reinforcement Learning} (DRL) approaches have
-thrived and provided impressive results in a variety of domains, achieving superhuman
-performances with no expert knowledge in perfect information zero-sum
+provided impressive results in a variety of domains, achieving superhuman
+performance with no expert knowledge in perfect information zero-sum
 games~\cite{alphazero}, reaching level of top players in video
@@
 parameters~(\citealt{drl_matter}, \citealt{drl_matter_bis}). This sensitivity,
-along with sample inefficiency, prevents DRL from being applied in most real
+along with sample inefficiency, prevents DRL from being applied in real
 world settings. High sensitivity to environment parameters notably prevents
-transfering abilities from imperfect simulators to real world scenarios.
+transfer of abilities from imperfect simulators to real world scenarios.
 
@@
 that decreasing time discretization, or equivalently shortening reaction time,
-should only improve agent performances. Robustness to time discretization is notably relevant in \emph{near-continuous}
+should only improve agent performance. Robustness to time discretization is notably relevant in \emph{near-continuous}
 environments, which includes most continuous control environments, robotics and many video games.
-\TODO{Explain that it also includes $\gamma \rightarrow 1$.}
+More generally, robusteness to time discretization is important in any setting in which the advantage
+an individual action over another is small compared to the total return, and notably on long time horizon
+problems.
 
@@
 
-A time discretization invariant algorithm should
-in particular admit a time continuous limit i.\,e. when the discretization timestep goes to $0$.
-This leads to precise design choices in term of
-agent architecture, exploration policy and learning rates scalings.
-The resulting algorithm is shown to provide near perfect invariance to time
-discretization on simple environments, and much better invariance properties
-than vanilla DQN or DDPG on more complex environments (Sec.~\ref{sec:exp}).
-On a new environment, as soon as the order of magnitude of the time discretization is known, our analysis readily provides salient scalings for a number of hyperparameters.
+A time discretization invariant algorithm should in particular admit a time
+continuous limit i.\,e. when the discretization timestep goes to $0$.  This
+leads to precise design choices in term of agent architecture, exploration
+policy and learning rates scalings.  The resulting algorithm is shown to
+provide better invariance to time discretization on many environments than
+vanilla DQN or DDPG (Sec.~\ref{sec:exp}).  On a new environment, as soon as the
+order of magnitude of the time discretization is known, our analysis readily
+provides relevant scalings for a number of hyperparameters.
 
@@
 \begin{itemize} 
-\item We formally show that the Q-function collapses to the V-function in near-continuous time, and thus that
+\item We show that the Q-function collapses to the V-function in near-continuous time, and thus that
     standard Q-learning is ill-behaved in this setting.
-  \item Our analysis of limit properties in near continuous environments leads us to define a robust methods off-policy algorithm. In particular,
-    it provides insights on how to design architectures, how to explore, and how to scale learning rates.
+  \item Our analysis of limit properties in near continuous environments leads to a robust off-policy algorithm. In particular,
+    it provides insights on how to design architectures, constraints exploration, as well as learning rates scalings.
   \item We empirically show that standard Q-learning methods are not robust to changes in time discretization, and that our algorithm provides huge gain in that domain.

### 2019-01-16 pendulum_exp/dataanalysis/plot_legend.py

@@ -1,0 +1,37 @@
+import matplotlib
+import matplotlib.pyplot as plt
+import numpy as np
+
+plt.style.use('ggplot')
+ax = plt.axes()
+ax.grid(False)
+ax.set_xticks([])
+ax.set_xlabel("Time discretization (log scale)")
+ax.set_yticks([])
+ax.margins(9, 0)
+MIN_DT = 5e-4
+MAX_DT = 5e-2
+bot_margin = .2
+width = 5
+
+color_range = np.linspace(np.log(MIN_DT), np.log(MAX_DT), 50)
+cnorm = matplotlib.colors.Normalize(vmin=np.log(MIN_DT) - 3, vmax=np.log(MAX_DT) + .1)
+cm1 = matplotlib.cm.ScalarMappable(norm=cnorm, cmap='Blues')
+cm2 = matplotlib.cm.ScalarMappable(norm=cnorm, cmap='Reds')
+line_color_blue = np.array([cm1.to_rgba(c) for c in color_range])[np.newaxis, ...]
+line_color_blue = np.tile(line_color_blue, (width, 1, 1))
+line_color_white = np.array([[1., 1., 1., 1.] for c in color_range])[np.newaxis, ...]
+line_color_white = np.tile(line_color_white, (width, 1, 1))
+line_color_red = np.array([cm2.to_rgba(c) for c in color_range])[np.newaxis, ...]
+line_color_red = np.tile(line_color_red, (width, 1, 1))
+line_color = np.concatenate([line_color_blue, line_color_white, line_color_red], axis=0)
+
+ax.text(0, 3 * width + bot_margin, f"{MIN_DT:3.2e}", color="dimgrey", verticalalignment='top', horizontalalignment='center')
+ax.text(50, 3 * width + bot_margin, f"{MAX_DT:3.2e}", color="dimgrey", verticalalignment='top', horizontalalignment='center')
+ax.text(-1, width / 2, "DDPG", color="dimgrey", verticalalignment='center', horizontalalignment='right', size=14)
+ax.text(-1, 2 * width + width / 2, "Advantage updating", color="dimgrey", verticalalignment='center', horizontalalignment='right', size=13)
+print(np.exp(color_range))
+
+plt.imshow(line_color)
+plt.show()
+input()

### 2019-01-17 paper/icml_drau/framework.tex
Commit

@@
 	environment. For instance, many simulated control environments, such as the
-	Mujoco environments~\cite{ddpg} or OpenAI Gym classic\TODO{CITE!} control
-	environments are discretization of continuous time control problems,
-        with a discretization timestep $\deltat$ fixed by the simulator.
-	For real world environments, the discretization timestep, $\deltat$ is
-	most often imposed by the quality of sensors and actuators. As $\deltat$ is not an
-	intrinsic parameter of the environment, but a constraint imposed by
+	Mujoco environments~\cite{ddpg} or OpenAI Gym classic control
+	environments~\cite{gym} are discretization of continuous time control problems.
+	In simulated environments, the time discretization is fixed by the simulator,
+	and is often used to approximate an underlying time continuous differential equation.
+	It can, for example, correspond to the inverse of the number of frames generated by second.
+	For real world environments, sensors and actuators have a fixed time precision: cameras can
+	only capture a fixed amount of frames per second, and physical limitations prevent actuators
+	from responding instantaneously. The quality of these components thus impose
+	a lower bound on the discretization timestep. As the time discretization 
+	$\deltat$ is not an intrinsic parameter of the environment, but a constraint imposed by
 	computational ressources, we would expect that decreasing $\deltat$
 	could  only improve the performance of reinforcement learning agents.
-	To achieve this goal, reinforcement learning algorithms should be
+	This implies that reinforcement learning algorithms should, at least, be
 	resilient to a change of $\deltat$, and should remain viable when
-	$\deltat \rightarrow 0$. We are thus interested in the behavior of
+	$\deltat \rightarrow 0$. Besides, designing a time discretization invariant algorithm
+	should alleviate tedious hyperparameterization by providing better defaults
+	for time horizon related parameters.
+	We are thus interested in the behavior of
 	reinforcement learning algorithms in discretized environments, when the
@@
 	continuous time, foreshadowing problematic behaviors of near continuous time Q-learning.
-      \TODO{We should explain somwhere here that, if not invariant, harder to find good hyperparameters}
+%      \TODO{We should explain somwhere here that, if not invariant, harder to find good hyperparameters} DONE
 \subsection{Framework}
@@
 
+% TODO: For now, differential equation is the default, SDE is a footnote.
         Let ${\cal S} = \mathbb{R}^d$ be a set of states, and ${\cal A}$ be a set of
-	actions. Consider the continuous \emph{Markov Decision Process} (MDP) defined by the \emph{Stochastic Differential Equation} (SDE)\TODO{Explain the intuition of Brownian in SDE}:
+	actions. Consider the continuous \emph{Markov Decision Process} (MDP) defined by the \emph{Differential Equation}: %\TODO{Explain the intuition of Brownian in SDE}:
         \begin{equation}
 		\label{eq:diffusion}
-		ds = F(s, a)dt + \Sigma(s, a)\dbt.
+		ds = F(s, a)dt. % + \Sigma(s, a)\dbt.
               \end{equation}
@@
 	M}_\deltat = \langle {\cal S}, {\cal A}, T_{\deltat}, r_\deltat,
-      \gamma_\deltat\rangle$ is a discretization of the continuous MDP with \emph{time discretization} $\deltat$.
-The transition probability $T_\deltat(s, a, s')$ is the probability distribution of $s_\deltat$, when starting at $s_0 = s$ and maintaining $a_t=a$ constant for a time $\deltat$.
-      \footnote{Informally, $T_\deltat(s, a, s')$ is approximately the probability distribution of $s + F(s, a)\deltat + \sqrt{\deltat}\Sigma(s, a){\cal N}(0, I)$, which is the discretization of Eq.
-\eqref{eq:diffusion} \TODO{CITE}}.
-This corresponds to an agent evolving in the continuous environment described by Eq.
-\eqref{eq:diffusion}, but who is only making observations and choosing actions every $\deltat$.
-We say the the MDP ${\cal M}_\deltat$ is near-continuous.
+      \gamma_\deltat\rangle$ is a discretization of the continuous MDP with
+      \emph{time discretization} $\deltat$. The transition probability
+      $T_\deltat(s, a, s')$ is a dirac on $s_\deltat$, where $s_\deltat$ is the
+      state obtained
+      when starting at $s_0 = s$ and maintaining $a_t=a$ constant for a time
+      $\deltat$.
+      This corresponds to an agent evolving in the continuous
+      environment described by Eq. \eqref{eq:diffusion}, but who is
+      only making observations and choosing actions every $\deltat$. We
+      say the the MDP ${\cal M}_\deltat$ is near-continuous.\footnote{
+	      We believe the results presented here to be true in a more general framework, where states
+	      follow a \emph{Stochastic Differential Equation} $ds = F(s, a)dt.  + \Sigma(s, a)\dbt$, with
+	      $B_t$ a multidimensional brownian motion and $\Sigma$ its associated covariance matrix. However,
+      formal treatment of SDEs is generally tedious and out of the scope of this paper.}
+      Informally $F: {\cal S}\times{\cal A}\rightarrow {\cal S}$ describes
+	the dynamic of the environment. The agent interacts with the environment through a deterministic policy function
+	$\pi \colon {\cal S} \rightarrow {\cal A}$. Determinism of the policy is assumed to prevent problematic definitions
+	of the value function in continuous time. This is only a mild requirement, as we introduce an off-policy algorithm,
+	and only impose determinism of the exploitation policy.
+	% This framework covers classical control theory, as well as many usual simulated environments. TODO: not convinced that this is relevant here.
 
-      Informally $F: {\cal S}\times{\cal A}\rightarrow {\cal S}$ describes
-	the deterministic dynamic of the environment, while $\Sigma(s, a)\dbt$
-	describes its stochasticity. Notably, $\Sigma = 0$ if the environment is
-	deterministic, as in classical control theory, or in many usual simulated
-	environments.
-
+%      \footnote{Informally, $T_\deltat(s, a, s')$ is approximately
+%	      the probability distribution of $s + F(s, a)\deltat +
+%	      \sqrt{\deltat}\Sigma(s, a){\cal N}(0, I)$, which is the
+%	      discretization of Eq. \eqref{eq:diffusion} \TODO{CITE}}.
         In this paper, we try to design RL methods for near continuous MDP
@@
           trajectories converge to the continuous ones. 
-
 	\begin{theorem}
@@
 		deterministic policy. Let $(s_t)_{t\in\mathbb{R}_+}$ be the
-		continuous process by following \eqref{eq:diffusion} and
+		continuous process obtained by following \eqref{eq:diffusion} and
 		$(s_\deltat^k)$ be the discrete process obtained if the agent only
-		get observations every $\deltat$ \TODO{explain better}.
+		get observations and interacts with the environment every $\deltat$ \TODO{explain better}.
 
@@
 			(\tilde s_\deltat^t)_{t\in\mathbb{R}_+} 
-			\xrightarrow[\deltat \rightarrow 0]{} (s_t)_{t\in\mathbb{R}_+}
+			\xrightarrow[\deltat \rightarrow 0]{} (s_t)_{t\in\mathbb{R}_+}.
 		\end{equation}
-		in distribution.
 	\end{theorem}

### 2019-01-17 paper/icml_drau/intro.tex
Commit

@@
 More generally, robusteness to time discretization is important in any setting in which the advantage
-an individual action over another is small compared to the total return, and notably on long time horizon
+of an individual action over another is small compared to the total return, and notably on long time horizon
 problems.

### 2019-01-17 paper/icml_drau/relatedworks.tex

@@
+%! TEX root = icml_drau.tex
 \section{Related work}
 \label{sec:related}
+
+Our approach largely builds on \cite{adv_upd}, who identified the
+collapse of Q-learning for small time steps and suggested advantage
+updating with proper $\deltat$ scalings as a solution.
+
 Continuous and near continuous reinforcement learning, which is at the core of

### 2019-01-17 paper/icml_drau/relatedworks.tex

@@
 
-Our approach largely builds on \cite{adv_upd}, who identified the
-collapse of Q-learning for small time steps and suggested advantage
-updating with proper $\deltat$ scalings as a solution.
+Our approach builds on \cite{adv_upd}, who identified the
+collapse of Q-learning for small time steps and suggested the Advantage
+Updating algorithm with proper $\deltat$ scalings as a solution; testing
+was only done on a quadratic-linear problem. Our Deep Advantage Updating
+reuses this idea with a different normalization step for $A$ (which
+forgoes the need to learn the normalization itself, thanks to the
+parameterization \eqref{eq:Aparam}), and tests Advantage Updating for the
+first time on a variety on RL environments using deep networks.
+Additionally, we provide formal proofs for the collapse
+of Q-learning when $\deltat\to 0$, and the non-collapse of Advantage
+Updating.
+
+\TODO{Say that Actor-critic has no problem when $\deltat\to 0$?}
 

### 2019-01-17 paper/icml_drau/relatedworks.tex

@@
 was only done on a quadratic-linear problem. Our Deep Advantage Updating
-reuses this idea with a different normalization step for $A$ (which
+expands this idea. Using a different normalization step for $A$ (which
 forgoes the need to learn the normalization itself, thanks to the
-parameterization \eqref{eq:Aparam}), and tests Advantage Updating for the
-first time on a variety on RL environments using deep networks.
+parameterization \eqref{eq:Aparam}), we test Advantage Updating for the
+first time on a variety on RL environments using deep networks,
+establishing as a viable algorithm in this setting.
 Additionally, we provide formal proofs for the collapse
-of Q-learning when $\deltat\to 0$, and the non-collapse of Advantage
+of Q-learning when $\deltat\to 0$, and for the non-collapse of Advantage
 Updating.

### 2019-01-17 paper/icml_drau/relatedworks.tex

--- 2019-01-17 paper/icml_drau/relatedworks.tex
@@
 
-Our approach largely builds on \cite{adv_upd}, who identified the
-collapse of Q-learning for small time steps and suggested advantage
-updating with proper $\deltat$ scalings as a solution.
+Our approach expands on \cite{adv_upd}, who identified the collapse of
+Q-learning for small time steps and suggested as a solution the Advantage Updating
+algorithm with proper $\deltat$ scalings for the $V$ part and the
+advantage part; testing was only
+done on a quadratic-linear problem. Using a different normalization step
+for $A$ (which forgoes the need to learn the normalization itself, thanks
+to the parameterization \eqref{eq:Aparam}), we test Advantage Updating
+for the first time on a variety on RL environments using deep networks,
+establishing Deep Advantage Updating as a viable algorithm in this
+setting.  Additionally, we provide formal proofs for the collapse of
+Q-learning when $\deltat\to 0$, and for the non-collapse of Advantage
+Updating, in a general setting.
+
+\TODO{Say that Actor-critic has no problem when $\deltat\to 0$?}
 
@@
 More recently, \cite{dueling_nets} also introduced a parameterization separating
-the value and the advantage components of the Q-function. We additionally
-introduce a natural scaling coeficient between the two, that strongly improves
-the time discretization invariance properties.
+the value and the advantage components of the Q-function. Contrary to
+\cite{advup}'s Advantage Updating, learning is still done in a standard
+way on the $Q$-function obtained from adding these two components. Thus this
+approach reparameterizes $Q$ but does not introduce the proper scalings
+and does not result in an invariant algorithm for small $\deltat$.%We additionally
+%introduce a natural scaling coeficient between the two, that strongly improves
+%the time discretization invariance properties.

### 2019-01-17 paper/icml_drau/framework.tex
.

@@
 
-\subsection{Return, discount factor  and Value function in near continuous time}
+\subsection{Return, discount factor and value function in near continuous time}
 \label{sec:ret-gamma-v}
 
-	In what follows, $\deltat$-dependant definitions of the discount factor $\gamma_\deltat$
+	In what follows, $\deltat$-dependant scalings of the discount factor $\gamma_\deltat$
 	and the reward $r_\deltat$ are given. These definitions fit the discrete case
@@
 	\label{eq:continuous-return}
-	R(\tau) = \int_{t=0}^\infty\gamma^tr(s_t, a_t)dt
+	R(\tau) = \int_{t=0}^\infty\gamma^tr(s_t, a_t)dt.
 \end{equation}
@@
 	R_\deltat(\tau) = \sum_{k=0}^\infty\gamma^{k\deltat}
-	r(s^\deltat_k, a^\deltat_k)\deltat
+	r(s_\deltat^k, a_\deltat^k)\deltat
 \end{equation}
@@
 	time.
-
         %Having a time horizon of $10$ steps means something very different
 	%if $\deltat = 0.1$ and if $\deltat = 0.0001$.
-
         A rule of thumb
@@
   V^\pi(s) =& \E_{\tau\sim\pi}\left[R(\tau) \mid s_0 = s\right]\\
-  =& \E_{\tau\sim\pi}\left[\int\limits_{0}^\infty \gamma^{t}\deltat r_{t} \mid s_0 = s\right].
+  =& \E_{\tau\sim\pi}\left[\int\limits_{0}^\infty \gamma^{t} r(s_t, a_t) dt \mid s_0 = s\right].
 \end{align}
@@
   V^\pi_\deltat(s) =& \E_{\tau\sim\pi}\left[R_\deltat(\tau) \mid s_0 = s\right]\\
-  =& \E_{\tau\sim\pi}\left[\sum\limits_{k=0}^\infty \gamma^{k\deltat}\deltat r_{k\deltat} \mid s_0 = s\right]
+  =& \E_{\tau\sim\pi}\left[\sum\limits_{k=0}^\infty \gamma^{k\deltat}r(s^k_{\deltat}, a^k_{\deltat})\deltat \mid s_0 = s\right]
 \end{align}
@@
 $r_\deltat$ defined in~(\ref{eq:def-gamma}, \ref{eq:def-reward}).
+The discretized value function follows the Bellman equation\footnote{
+	If the continuous MDP dynamic follows~\eqref{eq:diffusion}, the limit of the Bellman equation for $V^\pi_\deltat$ when $\deltat \rightarrow 0$ is the Hamilton-Jacobi-Bellman equation on $V^\pi$ \cite{cont_rl}
+\begin{equation}
+  \label{eq:hamilton-jacobi-bellman}
+  r + \partial_s V^\pi \cdot F = - \log(\gamma) V^\pi
+\end{equation}}
+
+\begin{equation}
+  \label{eq:bellman}
+  V^\pi_\deltat(s) = \reward(s, \pi(s))\deltat + \gamma^{\deltat} \E_{s_{\deltat}^{k+1} | s_{\deltat}^{k} = s} V^\pi_\deltat(s_{\deltat}^{k+1})
+\end{equation}
+
 We have the following convergence theorem of the discretized value function to the continuous value function:
@@
     \end{equation}
+    \label{th:conv-value}
   \end{theorem}
-
-The discretized value function follows the Bellman equation:
-\begin{equation}
-  \label{eq:bellman}
-  V^\pi_\deltat(s) = \reward(s, \pi(s))\deltat + \gamma^{\deltat} \E_{s_{\deltat}^{k+1} | s_{\deltat}^{k} = s} V^\pi_\deltat(s_{\deltat}^{k+1})
-\end{equation}
-
-If the MDP is following the Eq.
-\eqref{eq:diffusion}, the limit of the Bellman equation for $V^\pi_\deltat$ when $\deltat \rightarrow 0$ is the Hamilton-Jacobi-Bellman equation on $V^\pi$:\TODO{Cite ?}
-\begin{equation}
-  \label{eq:hamilton-jacobi-bellman}
-  r + \partial_s V^\pi \cdot F + \frac{1}{2} \text{tr}\left(\Sigma^T\partial^2_{s^2} V^\pi\Sigma\right) = - \log(\gamma) V^\pi(s)
-\end{equation}
 
@@
 Contrary to the Value function, the Action-Value function $Q$ is ill-defined
-for a continuous MDP. More precisely, the Q-function  collapses to the V-function when $\deltat \rightarrow 0$. In near continuous time,
-the effect of individual actions on the Q-function is of order $\bigO(\deltat)$. This makes deriving a hierarchy of action from an approximate Q-function difficult.
+for a continuous MDP. More precisely, the Q-function  collapses to the
+V-function when $\deltat \rightarrow 0$. In near continuous time, the effect of
+individual actions on the Q-function is of order $\bigO(\deltat)$. This makes
+deriving a ranking of action from an approximate Q-function difficult.
 Formally:
 \begin{theorem}
-  For all $s, a$, we have
+	Let $\pi$ be a deterministic policy such that for all $\deltat$, $V^\pi_\deltat$ is continuously
+	differentiable, then, for all $s, a$,
 	\begin{equation}
@@
 
-  If we consider the continuous action-value function $Q^\pi$, we have:
-  \begin{equation}
-    \label{eq:q-cont}
-    Q^\pi(s, a) = V^\pi
-  \end{equation}
   \label{th:q-cont}
@@
 A consequence of Theorem \ref{th:q-cont} is that in exact continuous time,
-$Q^\pi$ does not bear any information on the hierarchy of actions, and
+$Q^\pi$ does not bear any information on the ranking of actions, and
 thus cannot be used to select actions that yields higher future returns. There is no time continuous Q-learning.
@@
 \begin{proof}
+	Bellman equation for the discretized Q-function yields
   \begin{align}
@@
   \end{align}
-  But we know that $s'$ is close to $s$. If we define $\delta s = s' - s$, then:
+  From the environment dynamic equation, 
   \begin{equation}
-    \E_{s'|s, a}\left[\delta s\right] = \bigO(\deltat) \quad \text{Var}(\delta s) = \bigO(\deltat)
-  \end{equation}
-  \TODO{Explain. And this is not exactly the Variance.} Therefore:
+	  s' = s + f(s, a) \deltat + \smallo(\deltat).
+  \end{equation}As $V^\pi$ is assumed to be continuously differentiable,
   \begin{align}
-    |\E_{s'|s, a}\left[V^\pi_\deltat(s')\right] - V^\pi_\deltat(s)| &= \E_{\delta s|s, a}\left[\partial_s V^\pi_\deltat\cdot \delta s +\bigO(\|\delta s\|^2)\right]\nonumber\\
-                                            &\leq \|\partial_s V_\deltat^\pi\|\cdot \|\E_{\delta s|s, a}\left[\delta s\right]\| +\bigO(\delta t)\nonumber\\
-    &= \bigO(\deltat)\nonumber
+	  V^\pi_\deltat(s') &= V^\pi_\deltat(s) + \partial_s V^\pi_\deltat(s) F(s, a)\deltat + \smallo(\deltat)\\
+			    &= V^\pi_\deltat(s) + \bigO(\deltat)
   \end{align}
-  Finally:
+  Rewriting $Q^\pi_\deltat$ yields
   \begin{align}
-    Q^\pi_\deltat(s, a) &= r(s, a)\deltat + \gamma^\deltat \E_{s'|s, a}\left[V_\deltat^\pi(s')\right] \nonumber\\
-                &= r(s, a)\deltat + \left(1 + \bigO(\deltat)\right)(V^\pi(s) + \bigO(\deltat))\nonumber\\
-    &= V^\pi(s) + \bigO(\deltat)\nonumber
+	  Q^\pi_\deltat(s, a) &= r(s, a)\deltat + \gamma^\deltat (V^\pi_\deltat(s) + \bigO(\deltat))\\
+			      &= (1 + \bigO(\deltat)) (V^\pi_\deltat(s) + \bigO(\deltat)) + \bigO(\deltat)\\
+			      &= V^\pi_\deltat(s) + \bigO(\deltat).
   \end{align}
+  Now, using Thm.~\ref{th:conv-value},
+  \begin{align}
+	  Q^\pi_\deltat(s, a) = V^\pi(s) + \bigO(\deltat)
+  \end{align}
+  which ends the proof.
+%  But we know that $s'$ is close to $s$. If we define $\delta s = s' - s$, then:
+%  \begin{equation}
+%    \E_{s'|s, a}\left[\delta s\right] = \bigO(\deltat) \quad \text{Var}(\delta s) = \bigO(\deltat)
+%  \end{equation}
+%  \TODO{Explain. And this is not exactly the Variance.} Therefore:
+%  \begin{align}
+%    |\E_{s'|s, a}\left[V^\pi_\deltat(s')\right] - V^\pi_\deltat(s)| &= \E_{\delta s|s, a}\left[\partial_s V^\pi_\deltat\cdot \delta s +\bigO(\|\delta s\|^2)\right]\nonumber\\
+%                                            &\leq \|\partial_s V_\deltat^\pi\|\cdot \|\E_{\delta s|s, a}\left[\delta s\right]\| +\bigO(\delta t)\nonumber\\
+%    &= \bigO(\deltat)\nonumber
+%  \end{align}
+%  Finally:
+%  \begin{align}
+%    Q^\pi_\deltat(s, a) &= r(s, a)\deltat + \gamma^\deltat \E_{s'|s, a}\left[V_\deltat^\pi(s')\right] \nonumber\\
+%                &= r(s, a)\deltat + \left(1 + \bigO(\deltat)\right)(V^\pi(s) + \bigO(\deltat))\nonumber\\
+%    &= V^\pi(s) + \bigO(\deltat)\nonumber
+%  \end{align}
   \end{proof}

### 2019-01-17 paper/icml_drau/relatedworks.tex

@@
 
-Our approach expands on \cite{adv_upd}, who identified the collapse of
-Q-learning for small time steps and suggested as a solution the Advantage Updating
-algorithm with proper $\deltat$ scalings for the $V$ part and the
-advantage part; testing was only
-done on a quadratic-linear problem. Using a different normalization step
-for $A$ (which forgoes the need to learn the normalization itself, thanks
-to the parameterization \eqref{eq:Aparam}), we test Advantage Updating
-for the first time on a variety on RL environments using deep networks,
-establishing Deep Advantage Updating as a viable algorithm in this
-setting.  Additionally, we provide formal proofs for the collapse of
-Q-learning when $\deltat\to 0$, and for the non-collapse of Advantage
-Updating, in a general setting.
+Our approach builds on \cite{adv_upd}, who identified the collapse of
+Q-learning for small time steps and suggested as a solution the Advantage
+Updating algorithm with proper $\deltat$ scalings for the $V$ part and
+the advantage part; testing was only done on a quadratic-linear problem.
+We expand on \cite{adv_upd} in several directions. First, we modify the
+algorithm by
+using a different normalization step for $A$, which forgoes the need to
+learn the normalization itself, thanks to the parameterization
+\eqref{eq:Aparam}. Second, we test Advantage Updating for the first time on a
+variety on RL environments using deep networks, establishing Deep
+Advantage Updating as a viable algorithm in this setting.  Third,
+we provide formal proofs in a general setting for the collapse of Q-learning when $\deltat\to
+0$, and for the non-collapse of Advantage Updating.
+Fourth, we also discuss how to obtain $\deltat$-invariant exploration. Fifth, we
+provide stringent experimental tests of the actual robustness to changing
+$\deltat$.
 

### 2019-01-17 paper/icml_drau/intro.tex

@@
   \item Our analysis of limit properties in near continuous environments leads to a robust off-policy algorithm. In particular,
-    it provides insights on how to design architectures, constraints exploration, as well as learning rates scalings.
-  \item We empirically show that standard Q-learning methods are not robust to changes in time discretization, and that our algorithm provides huge gain in that domain.
+    it provides insights on how to design architectures, constraint exploration, and scale learning rates.
+  \item We empirically show that standard Q-learning methods are not robust to changes in time discretization, and that our algorithm provides notable gains in that domain.
 \end{itemize}

### 2019-01-17 paper/icml_drau/relatedworks.tex

@@
 $\deltat$.
-
-\TODO{Say that Actor-critic has no problem when $\deltat\to 0$?}
+%  \cite{adv_upd} already separated the learning procedure of the
+% value and advantage functions, and introduced proper scalings between the two
+% components. We extend on \cite{adv_upd} in three directions. We provide a
+% formal argument on why the advantage contribution of the Q-function is of order
+% $\bigO(\deltat)$, irrelevant of the stochasticity of the environment, in a quite
+% general framework. We show that, provided with well suited learning rates and
+% exploration method, Advantage updating provide near invariance to change of time
+% discretization. We show that Advantage updating is viable in the context of deep
+% reinforcement learning, while not using the true gradient of the residual error.
+% \TODO{Say that Actor-critic has no problem when $\deltat\to 0$?}
 
@@
 The approach presented in this paper is closely related to \cite{adv_upd} and
-\cite{cont_rl}. \cite{adv_upd} already separated the learning procedure of the
-value and advantage functions, and introduced proper scalings between the two
-components. We extend on \cite{adv_upd} in three directions. We provide a
-formal argument on why the advantage contribution of the Q-function is of order
-$\bigO(\deltat)$, irrelevant of the stochasticity of the environment, in a quite
-general framework. We show that, provided with well suited learning rates and
-exploration method, Advantage updating provide near invariance to change of time
-discretization. We show that Advantage updating is viable in the context of deep
-reinforcement learning, while not using the true gradient of the residual error.
+\cite{cont_rl}.
 

### 2019-01-17 paper/icml_drau/relatedworks.tex

@@
 we provide formal proofs in a general setting for the collapse of Q-learning when $\deltat\to
-0$, and for the non-collapse of Advantage Updating.
+0$, and for the non-collapse of Advantage Updating with the proper scalings.
 Fourth, we also discuss how to obtain $\deltat$-invariant exploration. Fifth, we

### 2019-01-17 paper/icml_drau/relatedworks.tex

@@
 % reinforcement learning, while not using the true gradient of the residual error.
-% \TODO{Say that Actor-critic has no problem when $\deltat\to 0$?}
 
-Continuous and near continuous reinforcement learning, which is at the core of
-our approach has been thoroughly studied in the deterministic case
-in~\cite{adv_upd} or \cite{cont_rl}, but has not been applied in the context of
-deep reinforcement learning.
-
-The approach presented in this paper is closely related to \cite{adv_upd} and
-\cite{cont_rl}.
+\TODO{Say that Actor-critic has no problem when $\deltat\to 0$?}
 
@@
 %the time discretization invariance properties.
+
+Continuous and near continuous reinforcement learning, which is at the core of
+our approach has been thoroughly studied in the deterministic case
+in~\cite{adv_upd} or \cite{cont_rl}, but has not been applied in the context of
+deep reinforcement learning.
+
+Reinforcement learning has been studied from a mathematical perspective
+when time and space are both continuous, in connection with optimal
+control and the Hamilton--Jacobi--Bellman equation (a PDE which
+characterizes the value function for continuous space-time). We refer to
+\cite{cont_rl} and the references therein. In particular \TODO{cite
+Munos, convergence of sol to HJB.}  \cite{cont_rl} focusses on practical
+algorithms for the continuous space-time case, including advantage
+updating.
+
+Here on the other hand we focus on generic RL algorithms that can handle
+both discrete and continuous time and space, without collapsing in
+continuous time, thus being robust to changing the timestep.
+
+%The approach presented in this paper is closely related to \cite{adv_upd} and \cite{cont_rl}.

### 2019-01-17 paper/icml_drau/relatedworks.tex

@@
 \cite{cont_rl} and the references therein. In particular \TODO{cite
-Munos, convergence of sol to HJB.}  \cite{cont_rl} focusses on practical
-algorithms for the continuous space-time case, including advantage
-updating.
+Munos, convergence of sol to HJB.}  The corresponding algorithms are
+often impractical \cite{cont_rl} for larger-dimensional problems.
+\cite{cont_rl} focusses on practical algorithms for the continuous
+space-time case, including advantage updating.
 
@@
 both discrete and continuous time and space, without collapsing in
-continuous time, thus being robust to changing the timestep.
+continuous time, thus being robust to arbitrary timesteps.
 

### 2019-01-17 paper/icml_drau/relatedworks.tex

@@
 
-Here on the other hand we focus on generic RL algorithms that can handle
+Here on the other hand we focus on generic deep RL algorithms that can handle
 both discrete and continuous time and space, without collapsing in

### 2019-01-17 paper/icml_drau/advup.tex

@@
 $Q^\pi_\deltat$, but admits an action dependant continuous time limit, it is
-natural to define
+natural to define \cite{adv_upd}
 \begin{align}
@@
 \begin{equation}
+\label{eq:Aparam}
 	A(s, a) = \bar{A}(s, a) - \bar{A}(s, \pi(s)).

### 2019-01-17 paper/icml_drau/relatedworks.tex

@@
 Our approach builds on \cite{adv_upd}, who identified the collapse of
-Q-learning for small time steps and suggested as a solution the Advantage
-Updating algorithm with proper $\deltat$ scalings for the $V$ part and
+Q-learning for small time steps and, as a solution, suggested the Advantage
+Updating algorithm, with proper $\deltat$ scalings for the $V$ part and
 the advantage part; testing was only done on a quadratic-linear problem.
@@
 
-Continuous and near continuous reinforcement learning, which is at the core of
-our approach has been thoroughly studied in the deterministic case
-in~\cite{adv_upd} or \cite{cont_rl}, but has not been applied in the context of
-deep reinforcement learning.
+% Continuous and near continuous reinforcement learning, which is at the core of
+% our approach has been thoroughly studied in the deterministic case
+% in~\cite{adv_upd} or \cite{cont_rl}, but has not been applied in the context of
+% deep reinforcement learning.
 
@@
 often impractical \cite{cont_rl} for larger-dimensional problems.
-\cite{cont_rl} focusses on practical algorithms for the continuous
+\cite{cont_rl} focusses on practical algorithms specific to the continuous
 space-time case, including advantage updating.

### 2019-01-17 paper/icml_drau/relatedworks.tex

@@
 the advantage part; testing was only done on a quadratic-linear problem.
+
 We expand on \cite{adv_upd} in several directions. First, we modify the
@@
 
-\TODO{Say that Actor-critic has no problem when $\deltat\to 0$?}
+\TODO{Say that Actor-critic has no problem when $\deltat\to 0$? With SGD
+but not with RMSProp, and needs robust exploration}
 
@@
 the value and the advantage components of the Q-function. Contrary to
-\cite{advup}'s Advantage Updating, learning is still done in a standard
+\cite{adv_upd}'s Advantage Updating, learning is still done in a standard
 way on the $Q$-function obtained from adding these two components. Thus this

### 2019-01-17 paper/icml_drau/relatedworks.tex

@@
 way on the $Q$-function obtained from adding these two components. Thus this
-approach reparameterizes $Q$ but does not introduce the proper scalings
+approach reparameterizes $Q$ but does not change scalings
 and does not result in an invariant algorithm for small $\deltat$.%We additionally

### 2019-01-17 paper/icml_drau/relatedworks.tex

@@
 when time and space are both continuous, in connection with optimal
-control and the Hamilton--Jacobi--Bellman equation (a PDE which
-characterizes the value function for continuous space-time). We refer to
-\cite{cont_rl} and the references therein. In particular \TODO{cite
-Munos, convergence of sol to HJB.}  The corresponding algorithms are
+control and the Hamilton--Jacobi--Bellman (HJB) equation (a PDE which
+characterizes the value function for continuous space-time). See for
+instance
+\cite{cont_rl,MunosBourgines98} and the references therein.
+\cite{MunosBourgines98} use a grid algorithm to provably solve the HJB
+equation when discretization tends to $0$. However,
+the corresponding algorithms are
 often impractical \cite{cont_rl} for larger-dimensional problems.

### 2019-01-17 paper/icml_drau/advup.tex
.

--- 2019-01-15 paper/icml_drau/advup.tex
@@
 \section{Reinforcement learning with a continuous time limit}
+
+In what follows, a discrete algorithm with a well defined time continuous
+limit is defined. The algorithm relies on three elements, namely, defining and
+learning a quantity that contains some information on the ranking of actions
+while admitting a time continuous limit, defining exploration methods that
+admit a continuous time limit and defining learning rates that induce well
+behaved parameter trajectories when $\deltat$ goes to $0$.
+
 As seen in Sec.~\ref{sec:framework}, there is no continuous time limit to
-Q-learning. In continuous time, $Q^\pi$ is independant of actions and can thus
+Q-learning. In continuous time, $Q^\pi$ is independent of actions and can thus
 not be used to select promising actions.  In near continuous time, $Q^\pi_\deltat$
@@
 approximation error is much bigger than $\bigO(\deltat)$, approximation errors
-dominates, and the hierarchy of actions given by the approximation is likely to
-be erroneous.  When the approximate Q-function is initialized, if the effect of
-actions on the Q-function is order of magnitudes higher than what it should be,
-approximating $Q^\pi_\deltat$ is likely to be difficult. Besides, the error is likely
-to be further propagated when the equation used to update our approximate $Q$
-function relies on bootstraping, as is the case for usual temporal difference
-derived methods.
-
-In what follows, a discrete algorithm that has a well defined time continuous
-limit is defined. The algorithm relies on three elements, namely, defining and
-learning a quantity that contains some information on the hierarchy of actions
-while admitting a time continuous limit, defining exploration methods that
-admit a continuous time limit and defining learning rates that induce well
-behaved parameter trajectories when $\deltat$ goes to $0$.
+dominates, and the ranking of actions given by the approximation is likely to
+be erroneous.
+% When the approximate Q-function is initialized, if the effect of
+% actions on the Q-function is order of magnitudes higher than what it should be,
+% approximating $Q^\pi_\deltat$ is likely to be difficult.
+%Besides, the error is likely
+%to be further propagated when the equation used to update our approximate $Q$
+%function relies on bootstraping, as is the case for usual temporal difference
+%derived methods.
 
@@
 \begin{align}
-	A^\pi_\deltat(s, a) &= \frac{Q^\pi_\deltat(s,a) - V^\pi_\deltat(s)}{\deltat},
+	A^\pi_\deltat(s, a) &\mathdef \frac{Q^\pi_\deltat(s,a) - V^\pi_\deltat(s)}{\deltat},
     \label{eq:adv}
 \end{align}
-a rescaled version of the advantage function, as the difference between between
+a rescaled version of the advantage function \cite{adv_upd}, as the difference between between
 $Q^\pi_\deltat(s, a)$ and $V^\pi\deltat(s)$ is of order $\bigO(\deltat)$.
@@
 \end{equation}
-To avoid cumbersome notations, in what follows, parameter indices are dropped.
-At initialization, if both $V_{\theta}$ and $A_{\psi}$ are initialized
-independantly of $\deltat$, this parameterization reasonably scales the
-contribution of actions relative to states in the value of $Q$.
+with $\Theta = (\theta, \pi)$. To avoid cumbersome notations, in what follows,
+parameter indices are dropped.  At initialization, if both $V_{\theta}$ and
+$A_{\psi}$ are initialized independantly of $\deltat$, this parameterization
+reasonably scales the contribution of actions relative to states in the value
+of $Q$.  Our goal is for $V_\theta$ to approximate $V^\pi_\deltat$ and for
+$A_{\psi}$ to approximate $A^\pi_\deltat$.
 
-\TODO{CT and LB are fighting each other about the importance which should be given to the next paragraph.
-A impartial judge should give his opinion.
-(CT is wrong)} However, the proposed reparameterization does not, on its own, guarantee that
-when $Q$ correctly approximates $Q^\pi_\deltat$, $A$ approximates $A^\pi_\deltat$.
+%\TODO{CT and LB are fighting each other about the importance which should be given to the next paragraph. A impartial judge should give his opinion. (CT is wrong)}
+However, the proposed reparameterization does not, on its own, guarantee that
+when $Q$ correctly approximates $Q^\pi_\deltat$, $A$ approximates $A^\pi_\deltat$\footnote{
+	Ensuring that $A$ do not differ from $A^\pi_\deltat$ by a state dependent function only affects
+	$A$'s interpretability. Indeed, for any function $f$, the ranking of actions given by
+	$A^\pi_\deltat(s, \cdot)$ and $A^\pi_\deltat(s, \cdot) + f(s)$ is the same.
+}.
 Indeed, for any given pair $V$, $A$,
@@
 $Q^\pi$ even if $A$ is arbitrarily far from $A^\pi$.
-To enforce, identifiability of $A$, one must enforce the consistency equation
+To enforce identifiability of $A$, one must enforce the consistency equation
 \begin{equation}
@@
 \end{equation}
-With this additional constraint, if $Q^\pi_\deltat = Q$,
+With this additional constraint, if $Q = Q^\pi_\deltat$, then $A = A^\pi_\deltat$ and $V = V^\pi_\deltat$.
 \begin{align}
@@
 To obtain a time invariant reinforcement learning algorithm, a time invariant
-exploration scheme is required. All the exploration schemes hereby considered
-generate sequences of actions as
+exploration scheme is required. 
+\cite{ddpg} already introduces an exploration scheme for continuous actions, which consists in adding an
+\emph{Ornstein-Uhlenbeck}~\cite{orn-uhl} process (OU process) to actions, that is naturally time
+discretization invariant. Formally, it is defined as
+\begin{equation}
+	\pi^\text{explore}(s^k_\deltat, z^k_\deltat) = \pi(s^k_\deltat) + z^k_\deltat
+\end{equation}
+with $z^k_\deltat$ the discretization of a time continuous OU process, defined as
+\begin{equation}
+	dz_t = - z_t \theta dt + \sigma \dbt.
+	\label{eq:orn_uhl}
+\end{equation}
+where $B_t$ is a brownian motion. Resulting discretized trajectories converge
+to non trivial continuous time trajectories.
+
+This exploration can be extended to schemes of the form
 \begin{equation}
@@
 with $(z^k_\deltat)_k$ a sequence of random variables, independant from the $a$'s and $s$'s.
-For instance, $\varepsilon$-greedy exploration naturally falls in this framework, \TODO{[[CT would like to give these details, LB thinks it is not necessary:]] with each $z^k_\deltat$
-being a pair composed of a bernoulli random variables $b^k_\deltat$ with probability of being $1$ equal to $\varepsilon$,
-and of a uniform categorical random variable $\xi^k_\deltat$ on actions. Exploratory actions are then selected as
-\begin{equation}
-	a^k_\deltat = (1 - b^k_\deltat) \text{argmax}_{a'} Q^\pi_\deltat(s, a') + b^k_\deltat \xi^k_\deltat.
-\end{equation}}
-
 A sufficient condition for the exploratory policy to admit a time continuous
@@
 well defined continuous stochastic process $z$ as $\deltat$ goes to $0$.
-\TODO{I'm convinced that this is true, reviewers might not be.}
+% For instance, $\varepsilon$-greedy exploration naturally falls in this framework, \TODO{[[CT would like to give these details, LB thinks it is not necessary:]] with each $z^k_\deltat$
+% being a pair composed of a bernoulli random variables $b^k_\deltat$ with probability of being $1$ equal to $\varepsilon$,
+% and of a uniform categorical random variable $\xi^k_\deltat$ on actions. Exploratory actions are then selected as
+% \begin{equation}
+% 	a^k_\deltat = (1 - b^k_\deltat) \text{argmax}_{a'} Q^\pi_\deltat(s, a') + b^k_\deltat \xi^k_\deltat.
+% \end{equation}}
 
-The exploration scheme used in~\cite{ddpg} for continuous actions, which consists in adding an
-\emph{Ornstein-Uhlenbeck}~\cite{orn-uhl} process (OU process) to actions, is naturally time
-invariant. It follows Eq. \eqref{eq:explore} with:
-\begin{equation}
-	\pi^\text{explore}(s_t, z_t) = \pi(s_t) + z_t
-\end{equation}
-and $z_t$ a time continuous OU process, defined as
-\begin{equation}
-	dz_t = - z_t \theta dt + \sigma \dbt.
-	\label{eq:orn_uhl}
-\end{equation}
+% \TODO{I'm convinced that this is true, reviewers might not be.}
 

### 2019-01-17 paper/icml_drau/relatedworks.tex

@@
 control and the Hamilton--Jacobi--Bellman (HJB) equation (a PDE which
-characterizes the value function for continuous space-time). See for
-instance
-\cite{cont_rl,MunosBourgines98} and the references therein.
-\cite{MunosBourgines98} use a grid algorithm to provably solve the HJB
-equation when discretization tends to $0$. However,
+characterizes the value function for continuous space-time). Continuous
+space-time algorithms are studied in
+\cite{cont_rl,MunosBourgines98} (see also the references therein).
+\cite{MunosBourgines98} use a grid approach to provably solve the HJB
+equation when discretization tends to $0$ (assuming every state in the
+grid is visited a large number of times). However,
 the corresponding algorithms are

### 2019-01-17 paper/icml_drau/relatedworks.tex

@@
 
-\TODO{Say that Actor-critic has no problem when $\deltat\to 0$? With SGD
+\TODO{Somewhere else: Say that Actor-critic has no problem when $\deltat\to 0$? With SGD
 but not with RMSProp, and needs robust exploration}
@@
 the corresponding algorithms are
-often impractical \cite{cont_rl} for larger-dimensional problems.
+impractical \cite{cont_rl} for larger-dimensional problems.
 \cite{cont_rl} focusses on practical algorithms specific to the continuous

### 2019-01-17 paper/icml_drau/advup.tex

@@
 % \end{equation}}
-
-% \TODO{I'm convinced that this is true, reviewers might not be.}
-
-On the other hand, the same argument cannot be applied to $\varepsilon$-greedy
-exploration, as the $z_\deltat$'s used to define it do not converge to a 
-well defined stochastic process, but to a white noise.\footnote{\TODO{$\varepsilon$-greedy
-		does not, in general, converges to a well defined exploration scheme in continuous time.
-		For instance, if actions directly affects position, $\varepsilon$-greedy results in white-noise
-trajectories in state space. Show this in more detail in appendix ?}}
-
 To obtain a temporally coherent exploration scheme in discrete action environment
@@
 where $z_t[a']$ denotes the $a'$-th component of the discretized OU process at time
-$k$. The resulting exploration scheme converges to a proper time continuous exploration scheme.
+$k$. The resulting exploration scheme converges to a proper time continuous exploration scheme,
+and is non trivial for most $Q$'s.
+
+Contrary to this discrete action exploration scheme, $\varepsilon$-greedy
+exploration is likely to not explore, i.e. to collapse to a deterministic
+policy when $\deltat$ goes to $0$. Indeed, one can prove that in a near
+continuous deterministic environment, the exploration policy resulting from
+applying $\varepsilon$ greedy exploration to a constant exploitation policy
+converges to a determinist policy as $\deltat$ goes to $0$. More precisely, let
+$(a_\deltat)$ a sequence of i.i.d. actions of distribution $p_A$, (for instance
+the sequence of actions generated by $\varepsilon$-greedy exploration on a
+constant policy). Then, as $\deltat$ goes to $0$, trajectories generated by
+interacting with the environment using the $a$'s converge to solutions of
+\begin{equation}
+	ds_t = \E_{a\sim p_A}\left[F(s_t, a)\right]dt
+\end{equation}
+where the expectation cancels the exploration noise.
 

### 2019-01-17 paper/icml_drau/advup.tex
.

@@
 \end{equation}
-where the expectation cancels the exploration noise.
+where the expectation cancels the exploration noise. The proof is given in the
+supplementary material.
 
@@
 \label{subsec:algorithm}
-To learn $V_{\theta}$ and $A_{\psi}$, variants of Q-learning for continuous~\cite{ddpg} and
-discrete~\cite{dqn} action spaces are used. In details, $V_{\theta}$ and $A_{\psi}$ are learnt
+To learn $V_{\theta}$ and $A_{\psi}$, variants of Q-learning for continuous and
+discrete action spaces are used. In details, $V_{\theta}$ and $A_{\psi}$ are learnt
 using approximate version of the following bellman equation, which holds for the true $A^\pi_\deltat$
-and $V^\pi_\deltat$,
+and $V^\pi_\deltat$, with a greedy exploitation policy $\pi(s) = \text{argmax}_{a'}A^\pi_\deltat(s, a')$
 \begin{align}
 	V^{\pi}_\deltat(s) + \deltat A^{\pi}_\deltat(s, a) &= r\deltat + \gamma^{\deltat}  \E_{s'} V^{\pi}_\deltat(s')\label{eq:bellman_A}\\
-	A^{\pi}_\deltat(s, \pi(s)) &= 0\label{eq:max_A}
+	A^{\pi}_\deltat(s, \pi(s)) &= 0\label{eq:max_A}.
 \end{align}
-for a greedy exploitation policy $\pi(s) = \text{argmax}_{a'}A^\pi_\deltat(s,
-a')$. The argmax is exactly implemented in the case of discrete actions, and approximated by
-a neural network $\pi_\phi(s)$, trained to maximize
-$A(s, \pi_\phi(s))$ in the continuous case.  This is similar to what is done in \cite{ddpg}.
+Maximization over actions is implemented exactly in the discrete actions case,
+and approximated by a policy neural network $\pi_\phi(s)$, trained to maximize $A(s,
+\pi_\phi(s))$ in the continuous case.  This is similar to what is done in
+\cite{ddpg}.
 
@@
 corresponding squared residual is minimized by an approximate gradient descent.
-The update equations when learning from a transition $(s, a, r, s')$ of
-an exploratory trajectory are
+The update equations when learning from a transition $(s, a, r, s')$, either from
+an exploratory trajectory or from a replay buffer~\cite{dqn} are
 \begin{align}
@@
 % of $\deltat$.
-In practice, $A$ and $V$ are learned using
-transitions sampled from a replay buffer instead of trajectory samples, to
-prevent excessive correlation between training samples~\cite{dqn}.
 \begin{algorithm}[ht]
@@
 have $\eta^V_\deltat = \alpha^V \deltat$ and $\eta^A_\deltat = \alpha^A
-\deltat$:
+\deltat$. More precisely
 \begin{theorem}
-  If $\eta^V_\deltat = \alpha^V \deltat$ and $\eta^A_\deltat = \alpha^A\deltat$, the continuous limit trajectory for parameters follows an SDE.
+	\TODO{We need a bit more details: we probably want to assume that states and actions are
+	drawn from a single fixed trajectory, and that we are learning for a fixed, constant policy.}
+	Let $\eta^V_\deltat = \alpha^V \deltat^\beta$ (resp. $\eta^A \deltat^\beta$) be the learning rate
+	of $V$ (resp. $A$), when training on a single exploratory trajectory. Then
+	\begin{itemize}
+		\item If $\beta = 1$ the discrete parameter trajectories converge to continuous parameter
+			trajectories as $\deltat$ goes to $0$.
+		\item If $\beta < 1$, parameters grow arbitrarily large in a single timestep as $\deltat$
+			goes to $0$.
+		\item If $\beta > 1$, parameter trajectories shrink to a single point as 
+			$\deltat$ goes to $0$.
+	\end{itemize}
 \end{theorem}

### 2019-01-18 paper/icml_drau/advup.tex
Commiting figures.

@@
 \label{subsec:algorithm}
+\begin{algorithm}[ht]
+	\caption{Deep Advantage Updating (Discrete actions)}
+	\input{dauc.tex}
+	\label{alg:dau}
+\end{algorithm}
+
 To learn $V_{\theta}$ and $A_{\psi}$, variants of Q-learning for continuous and
@@
 % of $\deltat$.
-\begin{algorithm}[ht]
-	\caption{Deep Advantage Updating (Discrete actions)}
-	\input{dauc.tex}
-	\label{alg:dau}
-\end{algorithm}
-
 \subsection{Learning rates scalings}
 \label{subsec:lr}
+\input{pendulum_fig.tex}
 For the discrete algorithm to admit a continuous time limit, the trajectories

### 2019-01-18 paper/icml_drau/pendulum_fig.tex
Commiting figures.

@@
 \newcommand{\iw}{1cm}
-\begin{figure*}[!ht]
-	\begin{tabular}{c|c}
-		DDPG & AU \\
-		\hline
-		\begin{tabular}{l|lllll}
-			$\delta t = 0.01$ & \includegraphics[width=\iw]{figs_data/pendulum/0.01/ddpg_act_2.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.01/ddpg_act_4.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.01/ddpg_act_6.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.01/ddpg_act_8.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.01/ddpg_act_10.png} \\
-			$\delta t = 0.005$ &\includegraphics[width=\iw]{figs_data/pendulum/0.005/ddpg_act_4.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.005/ddpg_act_8.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.005/ddpg_act_12.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.005/ddpg_act_16.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.005/ddpg_act_20.png} \\
-			$\delta t = 0.002$ &\includegraphics[width=\iw]{figs_data/pendulum/0.002/ddpg_act_10.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.002/ddpg_act_20.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.002/ddpg_act_30.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.002/ddpg_act_40.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.002/ddpg_act_50.png} \\
-			$\delta t = 0.001$ &\includegraphics[width=\iw]{figs_data/pendulum/0.001/ddpg_act_20.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.001/ddpg_act_40.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.001/ddpg_act_60.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.001/ddpg_act_80.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.001/ddpg_act_100.png} \\
-			$\delta t = 0.0005$ &\includegraphics[width=\iw]{figs_data/pendulum/0.0005/ddpg_act_40.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.0005/ddpg_act_80.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.0005/ddpg_act_120.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.0005/ddpg_act_160.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.0005/ddpg_act_200.png}
-		\end{tabular} &
-		\begin{tabular}{lllll}
-			 \includegraphics[width=\iw]{figs_data/pendulum/0.01/au_act_2.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.01/au_act_4.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.01/au_act_6.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.01/au_act_8.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.01/au_act_10.png} \\
-			\includegraphics[width=\iw]{figs_data/pendulum/0.005/au_act_4.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.005/au_act_8.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.005/au_act_12.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.005/au_act_16.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.005/au_act_20.png} \\
-			\includegraphics[width=\iw]{figs_data/pendulum/0.002/au_act_10.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.002/au_act_20.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.002/au_act_30.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.002/au_act_40.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.002/au_act_50.png} \\
-			\includegraphics[width=\iw]{figs_data/pendulum/0.001/au_act_20.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.001/au_act_40.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.001/au_act_60.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.001/au_act_80.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.001/au_act_100.png} \\
-			\includegraphics[width=\iw]{figs_data/pendulum/0.0005/au_act_40.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.0005/au_act_80.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.0005/au_act_120.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.0005/au_act_160.png} &\includegraphics[width=\iw]{figs_data/pendulum/0.0005/au_act_200.png}
-		\end{tabular}
-	\end{tabular}
+\begin{figure*}[ht]
+	\includegraphics[width=\textwidth]{figs_data/pendulum_fig.pdf}
 	\caption{Policies obtained by DDPG and AU at different macroscopic instants of training.

### 2019-01-18 pendulum_exp/dataanalysis/plot_legend.py

@@ -2,17 +2,19 @@ import matplotlib
 import matplotlib.pyplot as plt
 import numpy as np
 
+plt.figure(figsize=(11, 6))
 plt.style.use('ggplot')
 ax = plt.axes()
 ax.grid(False)
 ax.set_xticks([])
-ax.set_xlabel("Time discretization (log scale)")
 ax.set_yticks([])
 ax.margins(9, 0)
 MIN_DT = 5e-4
 MAX_DT = 5e-2
-bot_margin = .2
+bot_margin = 1
+left_margin = 2
 width = 5
+font_size = 35
 
 color_range = np.linspace(np.log(MIN_DT), np.log(MAX_DT), 50)
 cnorm = matplotlib.colors.Normalize(vmin=np.log(MIN_DT) - 3, vmax=np.log(MAX_DT) + .1)
@@ -26,12 +28,14 @@ line_color_red = np.array([cm2.to_rgba(c) for c in color_range])[np.newaxis, ...
 line_color_red = np.tile(line_color_red, (width, 1, 1))
 line_color = np.concatenate([line_color_blue, line_color_white, line_color_red], axis=0)
 
-ax.text(0, 3 * width + bot_margin, f"{MIN_DT:3.2e}", color="dimgrey", verticalalignment='top', horizontalalignment='center')
-ax.text(50, 3 * width + bot_margin, f"{MAX_DT:3.2e}", color="dimgrey", verticalalignment='top', horizontalalignment='center')
-ax.text(-1, width / 2, "DDPG", color="dimgrey", verticalalignment='center', horizontalalignment='right', size=14)
-ax.text(-1, 2 * width + width / 2, "Advantage updating", color="dimgrey", verticalalignment='center', horizontalalignment='right', size=13)
-print(np.exp(color_range))
+ax.text(0, 3 * width + bot_margin, f"{MIN_DT:3.2e}", color="dimgrey", verticalalignment='top', horizontalalignment='center', size=int(font_size * 2 / 3))
+ax.text(50, 3 * width + bot_margin, f"{MAX_DT:3.2e}", color="dimgrey", verticalalignment='top', horizontalalignment='center', size=int(font_size * 2 / 3))
+ax.text(0, width + bot_margin, f"{MIN_DT:3.2e}", color="dimgrey", verticalalignment='top', horizontalalignment='center', size=int(font_size * 2 / 3))
[…]

### 2019-01-18 paper/icml_drau/discussions.tex
.

@@
 \label{sec:discussions}
+The method derived in this work seems yield improved robustness to time discretization on 
+various environments, e.g. pendulum swing up, or simple locomotion tasks.
+However on certain environments, there is still much room for improvment. In what follows,
+we identify some of the issues that could explain this theoretical/practical discrepancy.
 
-TODO
+\paragraph{Regularity of the value function}~In all of our proofs, $V^\pi$ is
+assumed to be continuously differentiable for all $\pi$.  This is a strong
+hypothesis that is not always verified in practice. Very simple environments
+break this hypothesis, such as for instance, the pendulum swing up environment,
+with pendulum height as reward and torque as action.  Indeed, assume that $\pi$
+is the optimal policy for the pendulum swing up environment, with state
+representation $(\theta, \dot{\theta})$. The state space for this environment
+can be partitioned into two subspaces $\mathcal{G}$ and $\mathcal{B}$, with
+$\mathcal{G}$ the subspace where when following the optimal policy, the
+pendulum can be swang up without crossing the $y$-axis, and $\mathcal{B}$ where
+it have to cross the $y$-axis. The difference of V-function between a point in
+$\mathcal{G}$ and a point in $\mathcal{B}$ is lower bounded by a finite
+quantity $M$. Besides, one can find two points $s_1 \in \mathcal{G}$ and $s_2 \in \mathcal{B}$
+such that $s_1$ and $s_2$ are arbitrarily close. Consequently, $V^\pi$ is
+discontinuous on the frontier, which results in non infinitesimal advantages
+for actions taken on the frontier. \TODO{Non infinitesimal advantages are found,
+	more generally, as soon as a given policy has different regimes, depending on
+	where it starts in state space. For such policies, at the frontier between two regimes,
+the value function can be discontinuous, resulting in non infinitesimal advantages.}
+
+\paragraph{Memory buffer size scalings}~Thm.~\ref{th:cont-params} is stated for
+transitions sampled sequentially from a fixed infinite size sequence. In practice,
+transitions are sampled from a memory buffer, to prevent excessive correlations.
+The memory buffer used is a fixed size circular buffer, and is
+filled with samples coming from a single infinite exploratory trajectory. In
+our experiments, the same buffer size is used for all time discretization. When
+$\deltat$ goes to $0$, this corresponds to learning nearly online on the
+infinite sequence, since the buffer will only contain examples that are very
+close temporally to the end of the sequence in physical time. While the resulting
+algorithm \TODO{kind of} converges to a continuous time limit, it is still not
+time discretization invariant, as the freshness of samples in the buffer will
+vary with the time discretization.
+A different choice would be to scale the size of the buffer by a factor of
+$\frac{1}{\deltat}$. This would amount to always learning on the same time physical interval of
+the trajectory. Our analysis, however, does not cover this case.
+
+\paragraph{Near continuous time reinforcement learning and RMSprop}~Be careful
+when using RMSprop. When you are facing a stochastic environment RMSprop
+evaluates the second momentum of gradients and rescale gradients by its square
+root. This second momentum will be of order δt 1 , due to the stochasticity in
+the environment.  Consequently all √ gradient steps are multiplied by δt, and
+learning rates need √ only be multiplied by δt instead of δt. Taking batches of
+size proportional to δt 1 applies a similar fix.

### 2019-01-18 paper/icml_drau/icml_drau.tex
.

@@
 % Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Making Q-learning robust to time discretization}
+\icmltitlerunning{Making Deep Q-learning methods robust to time discretization}
 
@@
 \twocolumn[
-\icmltitle{Making Q-learning robust to time discretization}
+\icmltitle{Making Deep Q-learning Methods Robust to Time Discretization}
 
@@
 \begin{abstract}
-	Despite remarkable successes, \emph{Deep Reinforcement learning} (DRL) remains not robust to
-	hyperparameterization, implementation details and small environment changes~(\citealt{drl_matter}, \citealt{drl_matter_bis}). 
-	Overcoming this sensitivity is key to making DRL applicable to real world problems.
-	In this paper, we study the sensitivity of DRL to
-	time discretization in \emph{near continuous time} environments. More precisely, we show that approaches based on estimations of the
-	Q-function, e.g. \emph{Deep Q-learning}~\cite{dqn} and \emph{Deep deterministic policy gradient}~\cite{ddpg}
-	are sensitive to variations of time discretization.
-        We detail a principled method to obtain an off-policy RL algorithm that is more robust to changes in time discretization, and demontrate this robustness empirically. 
+	Despite remarkable successes, \emph{Deep Reinforcement Learning}
+	(DRL) is not robust to hyperparameterization, implementation
+	details, or small environment changes (\citealt{drl_matter},
+	\citealt{drl_matter_bis}).  Overcoming such sensitivity is key to
+	making DRL applicable to real world problems.  In this paper,
+	we identify
+	sensitivity to \emph{time discretization} in near
+	continuous-time environments as a critical factor; this covers, e.g., changing the number of
+	frames per second, or the action frequency of the controller.
+	Empirically, we find that
+	Q-learning-based approaches such as
+	\emph{Deep Q-learning}~\cite{dqn} and \emph{Deep Deterministic
+	Policy Gradient}~\cite{ddpg} %are highly sensitive to variations of time discretization and
+	collapse with small time steps. Formally, we prove that
+	$Q$-learning does not exist in
+	continuous time.
+	We detail a principled way to build an
+	off-policy RL algorithm that yields similar performances over a wide range
+	of time
+	discretizations, and confirm this robustness empirically. 
 	% We further show that a simple reparameterization of the
-	% Q-function as a sum of a state value term and a \emph{small} action dependent advantage term yields an algorithm
-	% much more resilient to variations of time discretization.\LB{Change last sentence to: We introduce a more resilient algorithm to variations of time discretization [We could say more, but still less that what is already there]}
+	% Q-function as a sum of a state value term and a \emph{small}
+	% action dependent advantage term yields an algorithm much more
+	% resilient to variations of time discretization.\LB{Change last
+	% sentence to: We introduce a more resilient algorithm to
+	% variations of time discretization [We could say more, but still
+	% less that what is already there]}
 \end{abstract}

### 2019-01-18 paper/icml_drau/intro.tex

@@
 parameters~(\citealt{drl_matter}, \citealt{drl_matter_bis}). This sensitivity,
-along with sample inefficiency, prevents DRL from being applied in real
-world settings. High sensitivity to environment parameters notably prevents
-transfer of abilities from imperfect simulators to real world scenarios.
+along with sample inefficiency, largely prevents DRL from being applied in real
+world settings. Notably, high sensitivity to environment parameters prevents
+transfer from imperfect simulators to real world scenarios.
 
-In this paper we focus our attention on the sensitivity to time discretization
-of DRL approaches,
-i.e. what happens when an agent receives $50$ observations
-and is expected to take $50$ actions per second instead of $10$. One expects
-that decreasing time discretization, or equivalently shortening reaction time,
-should only improve agent performance. Robustness to time discretization is notably relevant in \emph{near-continuous}
-environments, which includes most continuous control environments, robotics and many video games.
-More generally, robusteness to time discretization is important in any setting in which the advantage
-of an individual action over another is small compared to the total return, and notably on long time horizon
-problems.
+In this paper we focus on the sensitivity to time discretization of DRL
+approaches, such as what happens when an agent receives $50$ observations
+and is expected to take $50$ actions per second instead of $10$. One
+would expect that decreasing time discretization, or equivalently
+shortening reaction time, should only improve agent performance.
+Robustness to time discretization is especially relevant in
+\emph{near-continuous} environments, which includes most continuous
+control environments, robotics, and many video games.  (We focus on
+near-continuous time because this leads to a clear framework, but the
+approach below would be relevant in any setting in which the advantage of
+any single individual action is small compared to the total return.)%, for instance, long time horizon problems.
 

### 2019-01-18 paper/icml_drau/intro.tex

@@
 gradient} (DDPG~\citep{ddpg}) are not robust to changes in time discretization. This is shown experimentally in Sec.~\ref{sec:exp}. 
-This sensitivity can be understood as follows: as the discretization timestep
-decreases, the effect of individual actions on the total return decreases too, and
-can vanish when the discretization timestep becomes infinitesimal. The natural
-framework to study this phenomenon is Continuous time reinforcement learning 
-(Sec.~\ref{sec:framework}) \cite{cont_rl, adv_upd}.
+Intuitively, as the discretization timestep
+decreases, the effect of individual actions on the total return decreases
+too: $Q(s,a)$ is the value of playing action $a$ then playing optimally,
+and if $a$ is only maintained for a very short time its advantage over
+other actions will be accordingly small. (This occurs even with a
+well-adjusted decay factor $\gamma$.)
 
-A time discretization invariant algorithm should in particular admit a time
-continuous limit i.\,e. when the discretization timestep goes to $0$.  This
+The natural framework to study this phenomenon is Continuous time
+reinforcement learning (Sec.~\ref{sec:framework}) \cite{cont_rl,
+adv_upd}. Formally, the effect of a single action vanishes when the
+discretization timestep becomes infinitesimal: there is no
+continuous-time $Q$-function.
+
+A time discretization invariant algorithm should in particular admit a
+continuous-time
+limit when the discretization timestep goes to $0$.  This
 leads to precise design choices in term of agent architecture, exploration
-policy and learning rates scalings.  The resulting algorithm is shown to
+policy, and learning rates scalings.  The resulting algorithm is shown to
 provide better invariance to time discretization on many environments than

### 2019-01-18 paper/icml_drau/intro.tex

@@
 policy, and learning rates scalings.  The resulting algorithm is shown to
-provide better invariance to time discretization on many environments than
-vanilla DQN or DDPG (Sec.~\ref{sec:exp}).  On a new environment, as soon as the
+provide better invariance to time discretization than
+vanilla DQN or DDPG, on many environments (Sec.~\ref{sec:exp}).  On a new environment, as soon as the
 order of magnitude of the time discretization is known, our analysis readily

### 2019-01-18 paper/icml_drau/intro.tex

@@
 and if $a$ is only maintained for a very short time its advantage over
-other actions will be accordingly small. (This occurs even with a
+other actions will be accordingly small. (This continuity effect occurs even with a
 well-adjusted decay factor $\gamma$.)
@@
 \begin{itemize} 
-\item We show that the Q-function collapses to the V-function in near-continuous time, and thus that
-    standard Q-learning is ill-behaved in this setting.
+\item Building on \cite{adv_upd}, we formally show that the $Q$-function collapses to the $V$-function in near-continuous time, and thus that
+    standard $Q$-learning is ill-behaved in this setting.
   \item Our analysis of limit properties in near continuous environments leads to a robust off-policy algorithm. In particular,
-    it provides insights on how to design architectures, constraint exploration, and scale learning rates.
-  \item We empirically show that standard Q-learning methods are not robust to changes in time discretization, and that our algorithm provides notable gains in that domain.
+    it provides insights on how to design architectures, constrain exploration, and scale learning rates.
+  \item We empirically show that standard Q-learning methods are not
+  robust to changes in time discretization, while our algorithm provides
+  substantial robustness in that domain.
 \end{itemize}

### 2019-01-18 paper/icml_drau/intro.tex

@@
     it provides insights on how to design architectures, constrain exploration, and scale learning rates.
-  \item We empirically show that standard Q-learning methods are not
-  robust to changes in time discretization, while our algorithm provides
+  \item We empirically show that standard $Q$-learning methods are not
+  robust to changes in time discretization, with degraded performance, while our algorithm provides
   substantial robustness in that domain.

### 2019-01-18 paper/icml_drau/intro.tex

--- 2019-01-18 paper/icml_drau/intro.tex
@@
     it provides insights on how to design architectures, constrain exploration, and scale learning rates.
-  \item We empirically show that standard Q-learning methods are not
-  robust to changes in time discretization, while our algorithm provides
+  \item We empirically show that standard $Q$-learning methods are not
+  robust to changes in time discretization, exhibiting degraded performance, while our algorithm provides
   substantial robustness in that domain.

### 2019-01-18 paper/icml_drau/intro.tex

@@
 
-In practice, standard approaches based on estimation of state action value functions, e.g.
-\emph{Deep Q-learning} (DQN~\citep{dqn} and \emph{Deep deterministic policy
+Standard approaches based on estimation of state action
+value functions, such as
+\emph{Deep $Q$-learning} (DQN~\citep{dqn} and \emph{Deep deterministic policy
 gradient} (DDPG~\citep{ddpg}) are not robust to changes in time discretization. This is shown experimentally in Sec.~\ref{sec:exp}. 

### 2019-01-18 paper/icml_drau/relatedworks.tex

@@
 %! TEX root = icml_drau.tex
-\section{Related work}
+\section{Related Work}
 \label{sec:related}
@@
 Our approach builds on \cite{adv_upd}, who identified the collapse of
-Q-learning for small time steps and, as a solution, suggested the Advantage
+$Q$-learning for small time steps and, as a solution, suggested the Advantage
 Updating algorithm, with proper $\deltat$ scalings for the $V$ part and
@@
 Advantage Updating as a viable algorithm in this setting.  Third,
-we provide formal proofs in a general setting for the collapse of Q-learning when $\deltat\to
+we provide formal proofs in a general setting for the collapse of $Q$-learning when $\deltat\to
 0$, and for the non-collapse of Advantage Updating with the proper scalings.
@@
 % components. We extend on \cite{adv_upd} in three directions. We provide a
-% formal argument on why the advantage contribution of the Q-function is of order
+% formal argument on why the advantage contribution of the $Q$-function is of order
 % $\bigO(\deltat)$, irrelevant of the stochasticity of the environment, in a quite
@@
 \TODO{Somewhere else: Say that Actor-critic has no problem when $\deltat\to 0$? With SGD
-but not with RMSProp, and needs robust exploration}
+but not with RMSProp, and needs robust exploration, and limited to
+on-policy}
 
 More recently, \cite{dueling_nets} also introduced a parameterization separating
-the value and the advantage components of the Q-function. Contrary to
+the value and the advantage components of the $Q$-function. Contrary to
 \cite{adv_upd}'s Advantage Updating, learning is still done in a standard
@@
 control and the Hamilton--Jacobi--Bellman (HJB) equation (a PDE which
-characterizes the value function for continuous space-time). Continuous
-space-time algorithms are studied in
+characterizes the value function for continuous space-time). Explicit
+algorithms for continuous
+space-time can be found in
 \cite{cont_rl,MunosBourgines98} (see also the references therein).

### 2019-01-18 paper/icml_drau/relatedworks.tex

@@
 
-More recently, \cite{dueling_nets} also introduced a parameterization separating
-the value and the advantage components of the $Q$-function. Contrary to
+\cite{dueling_nets} also use a parameterization separating
+the value and advantage components of the $Q$-function. But contrary to
 \cite{adv_upd}'s Advantage Updating, learning is still done in a standard
@@
 and does not result in an invariant algorithm for small $\deltat$.%We additionally
-%introduce a natural scaling coeficient between the two, that strongly improves
+%introduce a natural scaling coefficient between the two, that strongly improves
 %the time discretization invariance properties.
@@
 % deep reinforcement learning.
+
+The problem studied here is a continuity effect quite distinct from
+multiscale RL approaches: indeed the issue arises even if there
+is only one timescale in the environment. Arguably, a small
+$\deltat$ can be seen as a mismatch between the algorithm's
+timescale and the physical system's timescale, but the collapse of the
+$Q$ function to the $V$ function is an intrinsic mathematical phenomenon from time
+continuity, and cannot obviously be addressed via existing multiscale
+techniques. \TODO{cite something!}
 

### 2019-01-18 paper/icml_drau/relatedworks.tex

@@
 grid is visited a large number of times). However,
-the corresponding algorithms are
+the resulting algorithms are
 impractical \cite{cont_rl} for larger-dimensional problems.
 \cite{cont_rl} focusses on practical algorithms specific to the continuous
-space-time case, including advantage updating.
+space-time case, including advantage updating and specifically modelling
+the time derivative of the environment.
 

### 2019-01-18 paper/icml_drau/framework.tex

@@
 %! TEX root = icml_drau.tex
-\section{Near continuous time reinforcement learning}
+\section{Near continuous-time reinforcement learning}
 \label{sec:framework}
 
-  Many reinforcement learning environments are not intrinsically time
-	discrete, but discretization of an underlying continuous time
-	environment. For instance, many simulated control environments, such as the
-	Mujoco environments~\cite{ddpg} or OpenAI Gym classic control
-	environments~\cite{gym} are discretization of continuous time control problems.
-	In simulated environments, the time discretization is fixed by the simulator,
-	and is often used to approximate an underlying time continuous differential equation.
-	It can, for example, correspond to the inverse of the number of frames generated by second.
-	For real world environments, sensors and actuators have a fixed time precision: cameras can
-	only capture a fixed amount of frames per second, and physical limitations prevent actuators
-	from responding instantaneously. The quality of these components thus impose
-	a lower bound on the discretization timestep. As the time discretization 
-	$\deltat$ is not an intrinsic parameter of the environment, but a constraint imposed by
-	computational ressources, we would expect that decreasing $\deltat$
-	could  only improve the performance of reinforcement learning agents.
-	This implies that reinforcement learning algorithms should, at least, be
-	resilient to a change of $\deltat$, and should remain viable when
-	$\deltat \rightarrow 0$. Besides, designing a time discretization invariant algorithm
-	should alleviate tedious hyperparameterization by providing better defaults
-	for time horizon related parameters.
-	We are thus interested in the behavior of
-	reinforcement learning algorithms in discretized environments, when the
-	discretization timestep is small. We will refer to such environments as
-	near continuous environments.
+Many reinforcement learning environments are not intrinsically
+time-discrete, but discretizations of an underlying continuous-time
+environment. For instance, many simulated control environments, such as
+the Mujoco environments~\cite{ddpg} or OpenAI Gym classic control
+environments~\cite{gym}, are discretization of continuous-time control
+problems.  In simulated environments, the time discretization is fixed by
+the simulator, and is often used to approximate an underlying
+differential equation.  For example, the timestep may correspond to the
+number of frames generated by second.  In real world environments,
+sensors and actuators have a fixed time precision: cameras can only
+capture a fixed amount of frames per second, and physical limitations
+prevent actuators from responding instantaneously. The quality of these
+components thus imposes a lower bound on the discretization timestep. As
+the timestep $\deltat$ is not an intrinsic parameter of the
+environment, but a constraint imposed by computational ressources, we
+would expect that decreasing $\deltat$ could only improve the performance
+of reinforcement learning agents (though it might make optimization
+harder).  Reinforcement
+learning algorithms should, at least, be resilient to a change of
+$\deltat$, and should remain viable when $\deltat \rightarrow 0$.
+Besides, designing a time discretization invariant algorithm could
+alleviate tedious hyperparameterization by providing better defaults for
+time-horizon-related parameters.
 
-	In what follows, a formalized view of near continuous environments is given,
-	along with $\deltat$ dependant definitions of return, discount factor and
-	value functions that are shown to converge to well defined continous time limits. Finally, it is shown
-	that the state-action value function collapses to the value function as
-	$\deltat$ goes to $0$. Consequently there is no Q-learning in
-	continuous time, foreshadowing problematic behaviors of near continuous time Q-learning.
-%      \TODO{We should explain somwhere here that, if not invariant, harder to find good hyperparameters} DONE
+We are thus interested in the behavior of reinforcement learning
+algorithms in discretized environments, when the discretization timestep
+is small. We will refer to such environments as \emph{near-continuous
+environments}.
+
+In what follows, a formalized view of near continuous environments is
+given, along with $\deltat$ dependant definitions of return, discount
+factor and value functions that are shown to converge to well defined
+continous time limits. Finally, it is shown that the state-action value
+function collapses to the value function as $\deltat$ goes to $0$.
+Consequently there is no Q-learning in continuous time, foreshadowing
+problematic behaviors of near continuous time Q-learning.
+%      \TODO{We should explain somwhere here that, if not invariant,
+%      harder to find good hyperparameters} DONE
+
 \subsection{Framework}

### 2019-01-18 paper/icml_drau/advup.tex
.

@@
 Appropriate scalings for the learning rates $\eta^V_\deltat$ and $\eta^A_\deltat$ in term of $\deltat$ to obtain a well defined continuous limit are derived in the following section.
-\TODO{Change notations for the learning rates in the algorithm}
 % As we will see in the following section,
@@
 	\end{itemize}
+	\label{th:cont-params}
 \end{theorem}
@@
 material.
-\TODO{Be careful when using RMSprop. When you are facing a stochastic environment
-	RMSprop evaluates the second momentum of gradients and rescale gradients by its
-	square root. This second momentum will
-	be of order $\frac{1}{\deltat}$, due to the stochasticity in the environment. Consequently
-	all gradient steps are multiplied by $\sqrt{\deltat}$, and learning rates need only be
-multiplied by $\sqrt{\deltat}$ instead of $\deltat$. Taking batches of size proportional to $\frac{1}{\deltat}$
-applies a similar fix.}

### 2019-01-18 paper/icml_drau/experiments.tex
Commiting figures.
.

--- 2019-01-10 paper/icml_drau/discussions.tex
@@
 %! TEX root = icml_drau.tex
-\section{Discussions}
-\label{sec:discussions}
+\section{Experiments}
+\label{sec:exp}
+\input{lc_fig.tex}
+The experiment set that we provide here is specifically aimed at showing that
+the proposed method can operate efficiently on a wide range of time
+discretization, without specific tuning, while standard deep Q-learning
+approaches can't.  Our method, which is referred to as \emph{Deep Advantage
+Updating}, is compared to DDPG in continuous action settings and to DQN in
+discrete action settings.
 
-TODO
+In all experimental settings, we use the learning algorithm described in
+alg.~\cite{alg:dau} and \TODO{supplementary, algo 1}. The variant of DDPG and DQN
+used are described in the supplementary material, as well as all the hyperparameters
+used.
+
+Let us stress that quantities used are rescaled to make comparison possible. For example,
+return results are given in term of macroscopic reward\footnote{This mostly amounts to scaling rewards
+by a factor $\deltat$ when this scaling is not naturally done in the environment. Environment specific
+details are given in the supplementary material.}, and, most notably, time elapsed is always given in
+normalized epochs, i.e. the true number of epochs multiplied by the discretization timestep.
+
+To provide qualitative results, and check that robustness to time
+discretization is provided both in term of raw return results, but also in term
+of convergence of approximate value function and policies to the optimal
+value function and policy, we first provide results on the simple pendulum environment
+provided in the openai gym classic control suite. Qualitative results are presented in
+fig.~\ref{fig:full_results}.
+% \begin{figure}[h]
+%   \centering
+%   \includegraphics[width=\columnwidth]{figs_data/cartpole_lc.png}
+%   \caption{Cartpole}
+%   \label{fig:cartpole-lc}
+% \end{figure}
+
+% \begin{figure}[h]
+%   \centering
+%   \includegraphics[width=\columnwidth]{figs_data/ant_lc.png}
+%   \caption{Ant}
+%   \label{fig:ant-lc}
+% \end{figure}
+% 
+% \begin{figure}[h]
+%   \centering
+%   \includegraphics[width=\columnwidth]{figs_data/pendulum_lc.png}
+%   \caption{Pendulum}
+%   \label{fig:pendulum-lc}
+% \end{figure}

### 2019-01-18 paper/icml_drau/framework.tex
.

--- 2019-01-17 paper/icml_drau/framework.tex
@@
 	      We believe the results presented here to be true in a more general framework, where states
-	      follow a \emph{Stochastic Differential Equation} $ds = F(s, a)dt.  + \Sigma(s, a)\dbt$, with
+	      follow a \emph{Stochastic Differential Equation} \begin{equation}
+		      ds = F(s, a)dt.  + \Sigma(s, a)\dbt
+		      \label{eq:sde}
+      \end{equation}, with
 	      $B_t$ a multidimensional brownian motion and $\Sigma$ its associated covariance matrix. However,

### 2019-01-18 paper/icml_drau/intro.tex

@@
 approaches, such as what happens when an agent receives $50$ observations
-and is expected to take $50$ actions per second instead of $10$. One
-would expect that decreasing time discretization, or equivalently
+and is expected to take $50$ actions per second instead of $10$.
+In principle, decreasing time discretization, or equivalently
 shortening reaction time, should only improve agent performance.
@@
 \emph{near-continuous} environments, which includes most continuous
-control environments, robotics, and many video games.  (We focus on
+control environments, robotics, and many video games.  
+(We focus on
 near-continuous time because this leads to a clear framework, but the
@@
 
-Standard approaches based on estimation of state action
+Standard approaches based on estimation of state-action
 value functions, such as
 \emph{Deep $Q$-learning} (DQN~\citep{dqn} and \emph{Deep deterministic policy
-gradient} (DDPG~\citep{ddpg}) are not robust to changes in time discretization. This is shown experimentally in Sec.~\ref{sec:exp}. 
+gradient} (DDPG~\citep{ddpg}) are not at all robust to changes in time discretization. This is shown experimentally in Sec.~\ref{sec:exp}. 
 Intuitively, as the discretization timestep

### 2019-01-18 paper/icml_drau/framework.tex

--- 2019-01-18 paper/icml_drau/framework.tex
@@
 
-In what follows, a formalized view of near continuous environments is
-given, along with $\deltat$ dependant definitions of return, discount
-factor and value functions that are shown to converge to well defined
-continous time limits. Finally, it is shown that the state-action value
-function collapses to the value function as $\deltat$ goes to $0$.
+A formalized view of near continuous environments is
+given below, along with $\deltat$-dependent definitions of return, discount
+factor and value functions, that converge to well defined
+continous-time limits. The state-action value
+function is shown to collapse to the value function as $\deltat$ goes to $0$.
 Consequently there is no Q-learning in continuous time, foreshadowing

### 2019-01-18 paper/icml_drau/relatedworks.tex

@@
 impractical \cite{cont_rl} for larger-dimensional problems.
-\cite{cont_rl} focusses on practical algorithms specific to the continuous
-space-time case, including advantage updating and specifically modelling
+\cite{cont_rl} focusses on algorithms specific to the continuous
+space-time case, including advantage updating and modelling
 the time derivative of the environment.

### 2019-01-18 paper/icml_drau/framework.tex

--- 2019-01-18 paper/icml_drau/framework.tex
@@
 environments}.
-
-A formalized view of near continuous environments is
+A formalized view of near-continuous environments is
 given below, along with $\deltat$-dependent definitions of return, discount
@@
 function is shown to collapse to the value function as $\deltat$ goes to $0$.
-Consequently there is no Q-learning in continuous time, foreshadowing
-problematic behaviors of near continuous time Q-learning.
+Consequently there is no $Q$-learning in continuous time, foreshadowing
+problematic behaviors of near continuous time $Q$-learning.
 %      \TODO{We should explain somwhere here that, if not invariant,
@@
 
+g
 %       }
@@
 Let ${\cal S} = \mathbb{R}^d$ be a set of states, and ${\cal A}$ be a set of
-	actions. Consider the continuous \emph{Markov Decision Process} (MDP) defined by the \emph{Differential Equation}: %\TODO{Explain the intuition of Brownian in SDE}:
+actions. Consider the continuous \emph{Markov Decision Process} (MDP) defined by the differential equation %\TODO{Explain the intuition of Brownian in SDE}:
         \begin{equation}
-		\label{eq:diffusion}
-		ds = F(s, a)dt. % + \Sigma(s, a)\dbt.
+g	\label{eq:diffusion}
+g	ds = F(s, a)dt. % + \Sigma(s, a)\dbt.
               \end{equation}
         For any $\deltat$, the discrete MDP ${\cal
-	M}_\deltat = \langle {\cal S}, {\cal A}, T_{\deltat}, r_\deltat,
+gM}_\deltat = \langle {\cal S}, {\cal A}, T_{\deltat}, r_\deltat,
       \gamma_\deltat\rangle$ is a discretization of the continuous MDP with
-      \emph{time discretization} $\deltat$. The transition probability
-      $T_\deltat(s, a, s')$ is a dirac on $s_\deltat$, where $s_\deltat$ is the
+      \emph{time discretization} $\deltat$. Here the deterministic
+      transition function
+      $T_\deltat(s, a, s')$ is %a dirac on $s_\deltat$, where $s_\deltat$ is 
+      the
       state obtained
@@
       This corresponds to an agent evolving in the continuous
-      environment described by Eq. \eqref{eq:diffusion}, but who is
+      environment described by Eq. \eqref{eq:diffusion}, but 
       only making observations and choosing actions every $\deltat$. We
-      say the the MDP ${\cal M}_\deltat$ is near-continuous.\footnote{
-	      We believe the results presented here to be true in a more general framework, where states
-	      follow a \emph{Stochastic Differential Equation} $ds = F(s, a)dt.  + \Sigma(s, a)\dbt$, with
-	      $B_t$ a multidimensional brownian motion and $\Sigma$ its associated covariance matrix. However,
-      formal treatment of SDEs is generally tedious and out of the scope of this paper.}
+      call such an  MDP ${\cal M}_\deltat$
+      near-continuous.\footnote{
+g      We believe the results presented here hold more
+g      generallly, assuming states
+g      follow a \emph{Stochastic Differential Equation} \begin{equation}
+g	      ds = F(s, a)dt  + \Sigma(s, a)\dbt
+g	      \label{eq:sde}
+      \end{equation} with
+g      $B_t$ a multidimensional Brownian motion and $\Sigma$ a covariance matrix. A
+      formal treatment of SDEs is beyond the scope of this paper.}
       Informally $F: {\cal S}\times{\cal A}\rightarrow {\cal S}$ describes
-	the dynamic of the environment. The agent interacts with the environment through a deterministic policy function
-	$\pi \colon {\cal S} \rightarrow {\cal A}$. Determinism of the policy is assumed to prevent problematic definitions
-	of the value function in continuous time. This is only a mild requirement, as we introduce an off-policy algorithm,
-	and only impose determinism of the exploitation policy.
-	% This framework covers classical control theory, as well as many usual simulated environments. TODO: not convinced that this is relevant here.
+gthe dynamic of the environment. The agent interacts with the environment through a deterministic policy function
+g$\pi \colon {\cal S} \rightarrow {\cal A}$. Determinism of the policy is assumed to prevent problematic definitions
+gof the value function in continuous time. This is only a mild requirement, as we introduce an off-policy algorithm,
+gand only impose determinism of the exploitation policy.
+g% This framework covers classical control theory, as well as many usual simulated environments. TODO: not convinced that this is relevant here.
 
@@
 %	      discretization of Eq. \eqref{eq:diffusion} \TODO{CITE}}.
-        In this paper, we try to design RL methods for near continuous MDP
-	robust to the choice of $\deltat$. For robustness, a necessary condition is to remain viable when $\deltat \rightarrow 0$. Our method is thus to analyse the different
-	quantities involved when $\deltat \rightarrow 0$, and make sure that they converge to meaningful limits.
+        %In this paper, we try to design RL methods for near continuous MDP
+g%robust to the choice of $\deltat$. 
+g
+gA necessary condition for robustness of an algorithm for
+gnear-continuous time MDPs is 
+gto remain viable when $\deltat \rightarrow 0$. Thus we will try to
+gmake sure the various quantities involved converge to meaningful
+glimits when $\deltat\to 0$.
 
@@
           trajectories converge to the continuous ones. 
-	\begin{theorem}
-		Assume that $\pi: {\cal S} \rightarrow {\cal A}$ is a continuous and
-		deterministic policy. Let $(s_t)_{t\in\mathbb{R}_+}$ be the
-		continuous process obtained by following \eqref{eq:diffusion} and
-		$(s_\deltat^k)$ be the discrete process obtained if the agent only
-		get observations and interacts with the environment every $\deltat$ \TODO{explain better}.
+g\begin{theorem}
+g	Assume that $\pi: {\cal S} \rightarrow {\cal A}$ is a continuous and
+g	deterministic policy. Let $(s_t)_{t\in\mathbb{R}_+}$ be the
+g	continuous process obtained by following \eqref{eq:diffusion} and
+g	$(s_\deltat^k)$ be the discrete process obtained if the agent only
+g	get observations and interacts with the environment every $\deltat$ \TODO{explain better}.
 
-		We define the time process $\tilde s_\deltat^t = s^{\lfloor
-		\frac{t}{\deltat}\rfloor}_\deltat$ for
-		$t\in\mathbb{R_+}$.\TODO{Explain}. Then:
-		\begin{equation}
-			\label{eq:conv-traj}
-			(\tilde s_\deltat^t)_{t\in\mathbb{R}_+} 
-			\xrightarrow[\deltat \rightarrow 0]{} (s_t)_{t\in\mathbb{R}_+}.
-		\end{equation}
-	\end{theorem}
+g	We define the time process $\tilde s_\deltat^t = s^{\lfloor
+g	\frac{t}{\deltat}\rfloor}_\deltat$ for
+g	$t\in\mathbb{R_+}$.\TODO{Explain}. Then:
+g	\begin{equation}
+g		\label{eq:conv-traj}
+g		(\tilde s_\deltat^t)_{t\in\mathbb{R}_+} 
+g		\xrightarrow[\deltat \rightarrow 0]{} (s_t)_{t\in\mathbb{R}_+}.
+g	\end{equation}
+g\end{theorem}
 
@@
 
-	In what follows, $\deltat$-dependant scalings of the discount factor $\gamma_\deltat$
-	and the reward $r_\deltat$ are given. These definitions fit the discrete case
-	scenario when $\deltat=1$, and yields well defined, non trivial returns
-	and value functions when $\deltat$ goes to $0$.
+gIn what follows, $\deltat$-dependant scalings of the discount factor $\gamma_\deltat$
+gand the reward $r_\deltat$ are given. These definitions fit the discrete case
+gscenario when $\deltat=1$, and yields well defined, non trivial returns
+gand value functions when $\deltat$ goes to $0$.
 
@@
 \begin{equation}
-	\label{eq:continuous-return}
-	R(\tau) = \int_{t=0}^\infty\gamma^tr(s_t, a_t)dt.
+g\label{eq:continuous-return}
+gR(\tau) = \int_{t=0}^\infty\gamma^tr(s_t, a_t)dt.
 \end{equation}
@@
 \begin{equation}
-	\label{eq:discretized-return}
-	R_\deltat(\tau) = \sum_{k=0}^\infty\gamma^{k\deltat}
-	r(s_\deltat^k, a_\deltat^k)\deltat
+g\label{eq:discretized-return}
+gR_\deltat(\tau) = \sum_{k=0}^\infty\gamma^{k\deltat}
+gr(s_\deltat^k, a_\deltat^k)\deltat
 \end{equation}
@@
 \begin{align}
-	\label{eq:def-gamma}
-	\gamma_\deltat =& \gamma^\deltat\\
-	\label{eq:def-reward}
-	r_\deltat =& \deltat . r.
+g\label{eq:def-gamma}
+g\gamma_\deltat =& \gamma^\deltat\\
+g\label{eq:def-reward}
+gr_\deltat =& \deltat . r.
 \end{align}
@@
         In near continuous environments, there are two notions of time,
-	the number of steps, or the number of actions taken, and the actual amount of time spent
-	in the underlying continuous time environment, which we call the physical
-	time. If $t$ is the real time and $k$ the number of steps, the two are related through $t = k \deltat$.
+gthe number of steps, or the number of actions taken, and the actual amount of time spent
+gin the underlying continuous time environment, which we call the physical
+gtime. If $t$ is the real time and $k$ the number of steps, the two are related through $t = k \deltat$.
 
-	The notion of time horizon, which is, informally, the amount of time on
-	which the agent optimizes its return, is only meaningful in physical
-	time.
+gThe notion of time horizon, which is, informally, the amount of time on
+gwhich the agent optimizes its return, is only meaningful in physical
+gtime.
         %Having a time horizon of $10$ steps means something very different
-	%if $\deltat = 0.1$ and if $\deltat = 0.0001$.
+g%if $\deltat = 0.1$ and if $\deltat = 0.0001$.
         A rule of thumb
-	is that the time horizon of an agent with discount factor $\gamma$ is
-	of order $\frac{1}{1 - \gamma}$ steps. If $\gamma_\deltat$ was left constant
-	as $\deltat$ goes to $0$, the corresponding time horizon in physical time
-	would write as $\frac{\deltat}{1 - \gamma}$ which goes to $0$ when
-	$\deltat$ goes to $0$.
+gis that the time horizon of an agent with discount factor $\gamma$ is
+gof order $\frac{1}{1 - \gamma}$ steps. If $\gamma_\deltat$ was left constant
+gas $\deltat$ goes to $0$, the corresponding time horizon in physical time
+gwould write as $\frac{\deltat}{1 - \gamma}$ which goes to $0$ when
+g$\deltat$ goes to $0$.
         On the contrary, the definition of $\gamma_\deltat$ in
-	Eq.~\eqref{eq:def-gamma} yields an approximately constant
-	time horizon in physical time when $\deltat$ goes to $0$. Indeed,
-	\begin{equation}
+gEq.~\eqref{eq:def-gamma} yields an approximately constant
+gtime horizon in physical time when $\deltat$ goes to $0$. Indeed,
+g\begin{equation}
           \label{eq:time-horizon}
@@
 The discretized value function follows the Bellman equation\footnote{
-	If the continuous MDP dynamic follows~\eqref{eq:diffusion}, the limit of the Bellman equation for $V^\pi_\deltat$ when $\deltat \rightarrow 0$ is the Hamilton-Jacobi-Bellman equation on $V^\pi$ \cite{cont_rl}
+gIf the continuous MDP dynamic follows~\eqref{eq:diffusion}, the limit of the Bellman equation for $V^\pi_\deltat$ when $\deltat \rightarrow 0$ is the Hamilton-Jacobi-Bellman equation on $V^\pi$ \cite{cont_rl}
 \begin{equation}
@@
 
-\subsection{There is no Q-function in continuous time}
+\subsection{There is no $Q$-function in continuous time}
 
 Contrary to the Value function, the Action-Value function $Q$ is ill-defined
-for a continuous MDP. More precisely, the Q-function  collapses to the
+for a continuous MDP. More precisely, the $Q$-function  collapses to the
 V-function when $\deltat \rightarrow 0$. In near continuous time, the effect of
-individual actions on the Q-function is of order $\bigO(\deltat)$. This makes
-deriving a ranking of action from an approximate Q-function difficult.
+individual actions on the $Q$-function is of order $\bigO(\deltat)$. This makes
+deriving a ranking of action from an approximate $Q$-function difficult.
 Formally:
 \begin{theorem}
-	Let $\pi$ be a deterministic policy such that for all $\deltat$, $V^\pi_\deltat$ is continuously
-	differentiable, then, for all $s, a$,
-	\begin{equation}
-		\label{eq:q-discr}
-		Q^\pi_\deltat(s, a) = V^\pi(s) + \bigO\left(\deltat\right)
-	\end{equation}
+gLet $\pi$ be a deterministic policy such that for all $\deltat$, $V^\pi_\deltat$ is continuously
+gdifferentiable, then, for all $s, a$,
+g\begin{equation}
+g	\label{eq:q-discr}
+g	Q^\pi_\deltat(s, a) = V^\pi(s) + \bigO\left(\deltat\right)
+g\end{equation}
 
@@
 $Q^\pi$ does not bear any information on the ranking of actions, and
-thus cannot be used to select actions that yields higher future returns. There is no time continuous Q-learning.
+thus cannot be used to select actions that yields higher future returns. There is no time continuous $Q$-learning.
 
 \begin{proof}
-	Bellman equation for the discretized Q-function yields
+gBellman equation for the discretized $Q$-function yields
   \begin{align}
@@
   \begin{equation}
-	  s' = s + f(s, a) \deltat + \smallo(\deltat).
+g  s' = s + f(s, a) \deltat + \smallo(\deltat).
   \end{equation}As $V^\pi$ is assumed to be continuously differentiable,
   \begin{align}
-	  V^\pi_\deltat(s') &= V^\pi_\deltat(s) + \partial_s V^\pi_\deltat(s) F(s, a)\deltat + \smallo(\deltat)\\
-			    &= V^\pi_\deltat(s) + \bigO(\deltat)
+g  V^\pi_\deltat(s') &= V^\pi_\deltat(s) + \partial_s V^\pi_\deltat(s) F(s, a)\deltat + \smallo(\deltat)\\
+g		    &= V^\pi_\deltat(s) + \bigO(\deltat)
   \end{align}
@@
   \begin{align}
-	  Q^\pi_\deltat(s, a) &= r(s, a)\deltat + \gamma^\deltat (V^\pi_\deltat(s) + \bigO(\deltat))\\
-			      &= (1 + \bigO(\deltat)) (V^\pi_\deltat(s) + \bigO(\deltat)) + \bigO(\deltat)\\
-			      &= V^\pi_\deltat(s) + \bigO(\deltat).
+g  Q^\pi_\deltat(s, a) &= r(s, a)\deltat + \gamma^\deltat (V^\pi_\deltat(s) + \bigO(\deltat))\\
+g		      &= (1 + \bigO(\deltat)) (V^\pi_\deltat(s) + \bigO(\deltat)) + \bigO(\deltat)\\
+g		      &= V^\pi_\deltat(s) + \bigO(\deltat).
   \end{align}
@@
   \begin{align}
-	  Q^\pi_\deltat(s, a) = V^\pi(s) + \bigO(\deltat)
+g  Q^\pi_\deltat(s, a) = V^\pi(s) + \bigO(\deltat)
   \end{align}

### 2019-01-18 paper/icml_drau/framework.tex

--- 2019-01-18 paper/icml_drau/framework.tex
@@
 Let ${\cal S} = \mathbb{R}^d$ be a set of states, and ${\cal A}$ be a set of
-	actions. Consider the continuous \emph{Markov Decision Process} (MDP) defined by the \emph{Differential Equation}: %\TODO{Explain the intuition of Brownian in SDE}:
+actions. Consider the continuous \emph{Markov Decision Process} (MDP) defined by the differential equation %\TODO{Explain the intuition of Brownian in SDE}:
         \begin{equation}
@@
       \gamma_\deltat\rangle$ is a discretization of the continuous MDP with
-      \emph{time discretization} $\deltat$. The transition probability
-      $T_\deltat(s, a, s')$ is a dirac on $s_\deltat$, where $s_\deltat$ is the
+      \emph{time discretization} $\deltat$. Here the deterministic
+      transition function
+      $T_\deltat(s, a, s')$ is %a dirac on $s_\deltat$, where $s_\deltat$ is 
+      the
       state obtained
@@
       This corresponds to an agent evolving in the continuous
-      environment described by Eq. \eqref{eq:diffusion}, but who is
+      environment described by Eq. \eqref{eq:diffusion}, but 
       only making observations and choosing actions every $\deltat$. We
-      say the the MDP ${\cal M}_\deltat$ is near-continuous.\footnote{
-	      We believe the results presented here to be true in a more general framework, where states
-	      follow a \emph{Stochastic Differential Equation} $ds = F(s, a)dt.  + \Sigma(s, a)\dbt$, with
-	      $B_t$ a multidimensional brownian motion and $\Sigma$ its associated covariance matrix. However,
-      formal treatment of SDEs is generally tedious and out of the scope of this paper.}
+      call such an  MDP ${\cal M}_\deltat$
+      near-continuous.\footnote{
+      We believe the results presented here hold more
+      generallly, assuming states
+      follow a \emph{Stochastic Differential Equation} \begin{equation}
+	      ds = F(s, a)dt  + \Sigma(s, a)\dbt
+	      \label{eq:sde}
+      \end{equation} with
+      $B_t$ a multidimensional Brownian motion and $\Sigma$ a covariance matrix. A
+      formal treatment of SDEs is beyond the scope of this paper.}
       Informally $F: {\cal S}\times{\cal A}\rightarrow {\cal S}$ describes
@@
 %	      discretization of Eq. \eqref{eq:diffusion} \TODO{CITE}}.
-        In this paper, we try to design RL methods for near continuous MDP
-	robust to the choice of $\deltat$. For robustness, a necessary condition is to remain viable when $\deltat \rightarrow 0$. Our method is thus to analyse the different
-	quantities involved when $\deltat \rightarrow 0$, and make sure that they converge to meaningful limits.
+        %In this paper, we try to design RL methods for near continuous MDP
+%robust to the choice of $\deltat$.
 
-        The first step in that direction is to check that the discretized
-          trajectories converge to the continuous ones. 
-	\begin{theorem}
-		Assume that $\pi: {\cal S} \rightarrow {\cal A}$ is a continuous and
-		deterministic policy. Let $(s_t)_{t\in\mathbb{R}_+}$ be the
-		continuous process obtained by following \eqref{eq:diffusion} and
-		$(s_\deltat^k)$ be the discrete process obtained if the agent only
-		get observations and interacts with the environment every $\deltat$ \TODO{explain better}.
+A necessary condition for robustness of an algorithm for
+near-continuous time MDPs is 
+to remain viable when $\deltat \rightarrow 0$. Thus we will try to
+make sure the various quantities involved converge to meaningful
+limits when $\deltat\to 0$.
 
-		We define the time process $\tilde s_\deltat^t = s^{\lfloor
-		\frac{t}{\deltat}\rfloor}_\deltat$ for
-		$t\in\mathbb{R_+}$.\TODO{Explain}. Then:
-		\begin{equation}
-			\label{eq:conv-traj}
-			(\tilde s_\deltat^t)_{t\in\mathbb{R}_+} 
-			\xrightarrow[\deltat \rightarrow 0]{} (s_t)_{t\in\mathbb{R}_+}.
-		\end{equation}
-	\end{theorem}
+By standard arguments involving time discretization of differential
+equations, the near-continuous time states themselves converge to the
+continuous-time states when $\deltat\to 0$. \TODO{ref to appendix+formal
+statement in appendix} We now turn to the RL-related quantities.
+
+%         The first step in that direction is to check that the discretized
+%           trajectories converge to the continuous ones. 
+% \begin{theorem}
+% 	Assume that $\pi: {\cal S} \rightarrow {\cal A}$ is a continuous and
+% 	deterministic policy. Let $(s_t)_{t\in\mathbb{R}_+}$ be the
+% 	continuous process obtained by following \eqref{eq:diffusion} and
+% 	$(s_\deltat^k)$ be the discrete process obtained if the agent only
+% 	get observations and interacts with the environment every $\deltat$ \TODO{explain better}.
+% 
+% 	We define the time process $\tilde s_\deltat^t = s^{\lfloor
+% 	\frac{t}{\deltat}\rfloor}_\deltat$ for
+% 	$t\in\mathbb{R_+}$.\TODO{Explain}. Then:
+% 	\begin{equation}
+% 		\label{eq:conv-traj}
+% 		(\tilde s_\deltat^t)_{t\in\mathbb{R}_+} 
+% 		\xrightarrow[\deltat \rightarrow 0]{} (s_t)_{t\in\mathbb{R}_+}.
+% 	\end{equation}
+% \end{theorem}
 
@@
 
-\subsection{There is no Q-function in continuous time}
+\subsection{There is no $Q$-function in continuous time}
 
 Contrary to the Value function, the Action-Value function $Q$ is ill-defined
-for a continuous MDP. More precisely, the Q-function  collapses to the
+for a continuous MDP. More precisely, the $Q$-function  collapses to the
 V-function when $\deltat \rightarrow 0$. In near continuous time, the effect of
-individual actions on the Q-function is of order $\bigO(\deltat)$. This makes
-deriving a ranking of action from an approximate Q-function difficult.
+individual actions on the $Q$-function is of order $\bigO(\deltat)$. This makes
+deriving a ranking of action from an approximate $Q$-function difficult.
 Formally:
@@
 $Q^\pi$ does not bear any information on the ranking of actions, and
-thus cannot be used to select actions that yields higher future returns. There is no time continuous Q-learning.
+thus cannot be used to select actions that yields higher future returns. There is no time continuous $Q$-learning.
 
 \begin{proof}
-	Bellman equation for the discretized Q-function yields
+Bellman equation for the discretized $Q$-function yields
   \begin{align}

### 2019-01-18 paper/icml_drau/framework.tex

@@
       This corresponds to an agent evolving in the continuous
-      environment described by Eq. \eqref{eq:diffusion}, but 
+      environment \eqref{eq:diffusion}, but 
       only making observations and choosing actions every $\deltat$. We
@@
 
-By standard arguments involving time discretization of differential
-equations, the near-continuous time states themselves converge to the
+By standard arguments for time discretizations of differential
+equations, the near-continuous-time states themselves converge to the
 continuous-time states when $\deltat\to 0$. \TODO{ref to appendix+formal

### 2019-01-18 paper/icml_drau/framework.tex

@@
       call such an  MDP ${\cal M}_\deltat$
-      near-continuous.\footnote{
+      \emph{near-continuous}.\footnote{
       We believe the results presented here hold more
       generallly, assuming states
-      follow a \emph{Stochastic Differential Equation} \begin{equation}
+      follow a \emph{stochastic} differential equation \begin{equation}
 	      ds = F(s, a)dt  + \Sigma(s, a)\dbt
@@
 By standard arguments for time discretizations of differential
-equations, the near-continuous-time states themselves converge to the
+equations, for a given policy the near-continuous states converge to the
 continuous-time states when $\deltat\to 0$. \TODO{ref to appendix+formal

### 2019-01-18 paper/icml_drau/framework.tex

@@
       environment \eqref{eq:diffusion}, but 
-      only making observations and choosing actions every $\deltat$. We
+      only making observations and choosing actions every $\deltat$. The
+      rewards and decay factor are specified below. We
       call such an  MDP ${\cal M}_\deltat$
@@
       formal treatment of SDEs is beyond the scope of this paper.}
-      Informally $F: {\cal S}\times{\cal A}\rightarrow {\cal S}$ describes
-the dynamic of the environment. The agent interacts with the environment through a deterministic policy function
+      Informally, $F\colon {\cal S}\times{\cal A}\rightarrow {\cal S}$ describes
+the dynamics of the environment. The agent interacts with the environment through a deterministic policy function
 $\pi \colon {\cal S} \rightarrow {\cal A}$. Determinism of the policy is assumed to prevent problematic definitions
 of the value function in continuous time. This is only a mild requirement, as we introduce an off-policy algorithm,
-and only impose determinism of the exploitation policy.
+and only impose determinism of the exploitation policy. \TODO{justify
+better or just say: for simplicity we assume that both the dynamics and
+policy are deterministic, or say the optimal policy can always be assumed
+to be deterministic.}
 % This framework covers classical control theory, as well as many usual simulated environments. TODO: not convinced that this is relevant here.

### 2019-01-19 paper/icml_drau/framework.tex

--- 2019-01-18 paper/icml_drau/framework.tex
@@
               \end{equation}
-        For any $\deltat$, the discrete MDP ${\cal
+Informally, $F\colon {\cal S}\times{\cal A}\rightarrow {\cal S}$ describes
+the dynamics of the environment. The agent interacts with the environment through a deterministic policy function
+$\pi \colon {\cal S} \rightarrow {\cal A}$, so that $a=\pi(s)$. 
+For simplicity we assume here that both the dynamics and exploitation policy are
+deterministic; the exploration policy will be random, but care must be
+taken to properly define random policies with
+discrete actions in continuous time, see below.
+
+For any $\deltat>0$, we can define an MDP ${\cal
 M}_\deltat = \langle {\cal S}, {\cal A}, T_{\deltat}, r_\deltat,
-      \gamma_\deltat\rangle$ is a discretization of the continuous MDP with
+      \gamma_\deltat\rangle$ as a discretization of the continuous MDP with
       \emph{time discretization} $\deltat$. Here the deterministic
@@
       formal treatment of SDEs is beyond the scope of this paper.}
-      Informally $F: {\cal S}\times{\cal A}\rightarrow {\cal S}$ describes
-the dynamic of the environment. The agent interacts with the environment through a deterministic policy function
-$\pi \colon {\cal S} \rightarrow {\cal A}$. Determinism of the policy is assumed to prevent problematic definitions
-of the value function in continuous time. This is only a mild requirement, as we introduce an off-policy algorithm,
-and only impose determinism of the exploitation policy.
       % This framework covers classical control theory, as well as many usual simulated environments. TODO: not convinced that this is relevant here.

### 2019-01-19 paper/icml_drau/framework.tex

@@
 	\label{eq:diffusion}
-	ds = F(s, a)dt. % + \Sigma(s, a)\dbt.
+	ds_t/dt = F(s_t, a) % + \Sigma(s, a)\dbt.
               \end{equation}
-Informally, $F\colon {\cal S}\times{\cal A}\rightarrow {\cal S}$ describes
+where $F\colon {\cal S}\times{\cal A}\rightarrow {\cal S}$ describes
 the dynamics of the environment. The agent interacts with the environment through a deterministic policy function
-$\pi \colon {\cal S} \rightarrow {\cal A}$, so that $a=\pi(s)$. 
+$\pi \colon {\cal S} \rightarrow {\cal A}$, so that $a=\pi(s_t)$. 
 For simplicity we assume here that both the dynamics and exploitation policy are

### 2019-01-19 paper/icml_drau/framework.tex

@@
 For simplicity we assume here that both the dynamics and exploitation policy are
-deterministic; the exploration policy will be random, but care must be
+deterministic; \footnote{
+      We believe the results presented here hold more
+      generally, assuming states
+      follow a \emph{stochastic} differential equation \begin{equation}
+	      ds = F(s, a)dt  + \Sigma(s, a)\dbt
+	      \label{eq:sde}
+      \end{equation} with
+      $B_t$ a multidimensional Brownian motion and $\Sigma$ a covariance matrix. A
+      formal treatment of SDEs is beyond the scope of this paper.}
+ the exploration policy will be random, but care must be
 taken to properly define random policies with
@@
       \gamma_\deltat\rangle$ as a discretization of the continuous MDP with
-      \emph{time discretization} $\deltat$. Here the deterministic
+      \emph{time discretization} $\deltat$. The %deterministic
       transition function
-      $T_\deltat(s, a, s')$ is %a dirac on $s_\deltat$, where $s_\deltat$ is 
+      %$T_\deltat(s, a, s')$ 
+      is %a dirac on $s_\deltat$, where $s_\deltat$ is 
       the
@@
       call such an  MDP ${\cal M}_\deltat$
-      \emph{near-continuous}.\footnote{
-      We believe the results presented here hold more
-      generallly, assuming states
-      follow a \emph{stochastic} differential equation \begin{equation}
-	      ds = F(s, a)dt  + \Sigma(s, a)\dbt
-	      \label{eq:sde}
-      \end{equation} with
-      $B_t$ a multidimensional Brownian motion and $\Sigma$ a covariance matrix. A
-      formal treatment of SDEs is beyond the scope of this paper.}
-      % This framework covers classical control theory, as well as many usual simulated environments. TODO: not convinced that this is relevant here.
+      \emph{near-continuous}.      % This framework covers classical control theory, as well as many usual simulated environments. TODO: not convinced that this is relevant here.
 
@@
 
+By standard arguments for time discretizations of differential
+equations, for a given policy the near-continuous states converge to the
+continuous-time states when $\deltat\to 0$. \TODO{ref to appendix+formal
+statement in appendix}% We now turn to the RL-related quantities.
+
 A necessary condition for robustness of an algorithm for
@@
 limits when $\deltat\to 0$.
-
-By standard arguments for time discretizations of differential
-equations, for a given policy the near-continuous states converge to the
-continuous-time states when $\deltat\to 0$. \TODO{ref to appendix+formal
-statement in appendix} We now turn to the RL-related quantities.
 
@@
 
-In what follows, $\deltat$-dependant scalings of the discount factor $\gamma_\deltat$
-and the reward $r_\deltat$ are given. These definitions fit the discrete case
-scenario when $\deltat=1$, and yields well defined, non trivial returns
+Suitable 
+$\deltat$-dependent scalings of the discount factor $\gamma_\deltat$
+and the reward $r_\deltat$ are as follows. These definitions fit the discrete case
+when $\deltat=1$, and provide well defined, non-trivial returns
 and value functions when $\deltat$ goes to $0$.
@@
 \label{eq:continuous-return}
-R(\tau) = \int_{t=0}^\infty\gamma^tr(s_t, a_t)dt.
+R(\tau) \deq \int_{t=0}^\infty\gamma^tr(s_t, a_t)dt.
 \end{equation}
 
-The natural discretization of this equation, if we want to define a discretized
-return $R_\delta$ for each $\deltat$ is:
+A natural time discretization is obtained  by defining the discretized return
+return $R_\delta$ for each $\deltat$ as
 \begin{equation}
 \label{eq:discretized-return}
-R_\deltat(\tau) = \sum_{k=0}^\infty\gamma^{k\deltat}
+R_\deltat(\tau) \deq \sum_{k=0}^\infty\gamma^{k\deltat}
 r(s_\deltat^k, a_\deltat^k)\deltat
 \end{equation}
-
-Natural definitions of $\gamma_\deltat$ and $r_\deltat$ such
-that the return defined in equation (\ref{eq:discretized-return}) corresponds
-to the discrete return for the MDP ${\cal M}_\deltat$ follow:
+and the discretized return will correspond to the continuous-time return
+if we set the decay factor $\gamma_\deltat$ and rewards $r_\deltat$ of
+the discretized MDP ${\cal M}_\deltat$ to
+% Natural definitions of $\gamma_\deltat$ and $r_\deltat$ such
+% that the return defined in equation (\ref{eq:discretized-return}) corresponds
+% to the discrete return for the MDP ${\cal M}_\deltat$ follow:
 \begin{align}
 \label{eq:def-gamma}
-\gamma_\deltat =& \gamma^\deltat\\
-\label{eq:def-reward}
-r_\deltat =& \deltat . r.
+%\label{eq:def-reward}
+\gamma_\deltat \deq \gamma^\deltat,\qquad
+r_\deltat \deq \deltat . r.
 \end{align}
@@
 which fits the usual definition of the discrete V-function for $\gamma_\deltat$ and
-$r_\deltat$ defined in~(\ref{eq:def-gamma}, \ref{eq:def-reward}).
+$r_\deltat$ defined in \eqref{eq:def-gamma}.
 The discretized value function follows the Bellman equation\footnote{

### 2019-01-19 paper/icml_drau/framework.tex

@@
 	\label{eq:diffusion}
-	ds_t/dt = F(s_t, a) % + \Sigma(s, a)\dbt.
+	ds_t/dt = F(s_t, a_t) % + \Sigma(s, a)\dbt.
               \end{equation}
@@
 the dynamics of the environment. The agent interacts with the environment through a deterministic policy function
-$\pi \colon {\cal S} \rightarrow {\cal A}$, so that $a=\pi(s_t)$. 
+$\pi \colon {\cal S} \rightarrow {\cal A}$, so that $a_t=\pi(s_t)$. 
 For simplicity we assume here that both the dynamics and exploitation policy are
@@
 
-\subsection{Return, discount factor and value function in near continuous time}
-\label{sec:ret-gamma-v}
+% \subsection{Return, discount factor and value function in near continuous time}
+% \label{sec:ret-gamma-v}
 
+\paragraph{Return and Discount Factor.}
 Suitable 
@@
 
-\paragraph{Return and Discount Factor}
 For a continuous MDP and a continuous trajectory $\tau = (s_t, a_t)_t$,
@@
 
-      \paragraph{Number of steps / Physical time, Time horizon}
+      \paragraph{Physical time vs algorithmic time, time horizon.}
         % [[I prefer to talk about ``number of steps'' VS ``Physical time'' than microscopic/Macroscopic]]
-        In near continuous environments, there are two notions of time,
-the number of steps, or the number of actions taken, and the actual amount of time spent
-in the underlying continuous time environment, which we call the physical
-time. If $t$ is the real time and $k$ the number of steps, the two are related through $t = k \deltat$.
+        In near continuous environments, there are two notions of time:
+the algorithmic time $k$ (number of steps or actions taken), and the
+physical time $t$ (time spent
+in the underlying continuous time environment), related via
+$t=k.\deltat$.
 
-The notion of time horizon, which is, informally, the amount of time on
-which the agent optimizes its return, is only meaningful in physical
-time.
+The time horizon is, informally, the time range over
+which the agent optimizes its return.
         %Having a time horizon of $10$ steps means something very different
 %if $\deltat = 0.1$ and if $\deltat = 0.0001$.
-        A rule of thumb
-is that the time horizon of an agent with discount factor $\gamma$ is
-of order $\frac{1}{1 - \gamma}$ steps. If $\gamma_\deltat$ was left constant
-as $\deltat$ goes to $0$, the corresponding time horizon in physical time
-would write as $\frac{\deltat}{1 - \gamma}$ which goes to $0$ when
-$\deltat$ goes to $0$.
-        On the contrary, the definition of $\gamma_\deltat$ in
-Eq.~\eqref{eq:def-gamma} yields an approximately constant
-time horizon in physical time when $\deltat$ goes to $0$. Indeed,
+As a rule of thumb,
+the time horizon of an agent with discount factor $\gamma$ is
+of order $\frac{1}{1 - \gamma}$ steps; beyond that, the decay factor
+kicks in and the influence of further rewards becomes small.
+
+The definition~\eqref{eq:def-gamma} of the decay factor $\gamma_\deltat$ in
+near-continuous environments keeps the time horizon constant in
+\emph{physical} time, by making $\gamma_\deltat$ close to $1$ in
+algorithmic time. Indeed, physical time horizon is $\deltat$ times the
+algorithmic time horizon, namely
 \begin{equation}
@@
           \frac{\deltat}{1-\gamma^\deltat}
-          = - \frac{1}{\log \gamma} + \bigO(\deltat)
+          %= - \frac{1}{\log \gamma} + \bigO(\deltat)
           \approx \frac{1}{1-\gamma},
         \end{equation}
-        approximately invariant to $\deltat$ when $\deltat$ goes to 0.
+which is thus stable when $\deltat\to 0$.
+On the other hand,
+if $\gamma_\deltat$ was left constant
+as $\deltat$ goes to $0$, the corresponding time horizon in physical time
+would be $\approx \frac{\deltat}{1 - \gamma}$ which goes to $0$ when
+$\deltat$ goes to $0$: such an agent would be increasingly short-sighted
+as $\deltat\to 0$.
 
-\paragraph{Value function}
+In the following, we use the suitably-scaled decay factor
+\eqref{eq:def-gamma} both for Deep Advantage Updating and for the
+classical deep $Q$-learning baselines.
+%         On the contrary, the definition of $\gamma_\deltat$ in
+% Eq.~\eqref{eq:def-gamma} yields an approximately constant
+% time horizon in physical time when $\deltat$ goes to $0$. Indeed,
+%         approximately invariant to $\deltat$ when $\deltat$ goes to 0.
+%
+
+\paragraph{Value function.}
 

### 2019-01-19 paper/icml_drau/framework.tex

@@
 \paragraph{Value function.}
-
-The definition of the return Eq.~\eqref{eq:continuous-return}, yields the
-following definition of the continuous value function
+The return \eqref{eq:continuous-return} yields the
+following continuous-time value function
 \begin{align}
-  V^\pi(s) =& \E_{\tau\sim\pi}\left[R(\tau) \mid s_0 = s\right]\\
-  =& \E_{\tau\sim\pi}\left[\int\limits_{0}^\infty \gamma^{t} r(s_t, a_t) dt \mid s_0 = s\right].
+  V^\pi(s) &= \E_{\tau\sim\pi}\left[R(\tau) \mid s_0 = s\right]\\
+  &= \E_{\tau\sim\pi}\left[\int\limits_{0}^\infty \gamma^{t} r(s_t, a_t) dt \mid s_0 = s\right].
 \end{align}
-The corresponding discretization for time discretization $\deltat$ is
+Meanwhile, the value function (in the ordinary sense) of the discrete MDP ${\cal
+M}_\deltat$ is
 \begin{align}
-  V^\pi_\deltat(s) =& \E_{\tau\sim\pi}\left[R_\deltat(\tau) \mid s_0 = s\right]\\
-  =& \E_{\tau\sim\pi}\left[\sum\limits_{k=0}^\infty \gamma^{k\deltat}r(s^k_{\deltat}, a^k_{\deltat})\deltat \mid s_0 = s\right]
+  V^\pi_\deltat(s) &= \E_{\tau\sim\pi}\left[R_\deltat(\tau) \mid s_0 = s\right]\\
+  &= \E_{\tau\sim\pi}\left[\sum\limits_{k=0}^\infty \gamma^{k\deltat}r(s^k_{\deltat}, a^k_{\deltat})\deltat \mid s_0 = s\right]
 \end{align}
-which fits the usual definition of the discrete V-function for $\gamma_\deltat$ and
-$r_\deltat$ defined in \eqref{eq:def-gamma}.
-The discretized value function follows the Bellman equation\footnote{
+% which fits the usual definition of the discrete V-function for $\gamma_\deltat$ and
+% $r_\deltat$ defined in \eqref{eq:def-gamma}.
+which obeys the Bellman equation
+\footnote{
 If the continuous MDP dynamic follows~\eqref{eq:diffusion}, the limit of the Bellman equation for $V^\pi_\deltat$ when $\deltat \rightarrow 0$ is the Hamilton-Jacobi-Bellman equation on $V^\pi$ \cite{cont_rl}
@@
   r + \partial_s V^\pi \cdot F = - \log(\gamma) V^\pi
-\end{equation}}
-
+\end{equation}
+}
 \begin{equation}
@@
 
-We have the following convergence theorem of the discretized value function to the continuous value function:
+When the timestep tends to $0$, one converges to the other.
   \begin{theorem}
-   For all $s \in {\cal S}$:
-    \begin{equation}
-      \label{eq:conv-value}
-      V^\pi_\deltat(s) = V^\pi(s) + \bigO(\deltat)
-    \end{equation}
     \label{th:conv-value}
+   Under suitable assumptions \TODO{ref appendix}, for all $s \in {\cal
+   S}$, one has
+    %\begin{equation}
+      %\label{eq:conv-value}
+      $V^\pi_\deltat(s) = V^\pi(s) + \bigO(\deltat)$
+    %\end{equation}
+    when $\deltat\to 0$.
   \end{theorem}

### 2019-01-19 paper/icml_drau/framework.tex

@@
 \end{align}
-% which fits the usual definition of the discrete V-function for $\gamma_\deltat$ and
+% which fits the usual definition of the discrete $V$-function for $\gamma_\deltat$ and
 % $r_\deltat$ defined in \eqref{eq:def-gamma}.
@@
 
-Contrary to the Value function, the Action-Value function $Q$ is ill-defined
-for a continuous MDP. More precisely, the $Q$-function  collapses to the
-V-function when $\deltat \rightarrow 0$. In near continuous time, the effect of
+Contrary to the value function, the action-value function $Q$ is ill-defined
+for continuous-time MDPs. More precisely, the $Q$-function  collapses to the
+$V$-function when $\deltat \rightarrow 0$. In near continuous time, the effect of
 individual actions on the $Q$-function is of order $\bigO(\deltat)$. This makes

### 2019-01-19 paper/icml_drau/framework.tex

@@
 $V$-function when $\deltat \rightarrow 0$. In near continuous time, the effect of
-individual actions on the $Q$-function is of order $\bigO(\deltat)$. This makes
-deriving a ranking of action from an approximate $Q$-function difficult.
+individual actions on the $Q$-function is of order $\bigO(\deltat)$. This
+will make 
+ranking of actions difficult, especially with an approximate
+$Q$-function.
+This argument appears informally in \cite{adv_upd}.
 Formally:
 \begin{theorem}
-Let $\pi$ be a deterministic policy such that for all $\deltat$, $V^\pi_\deltat$ is continuously
-differentiable, then, for all $s, a$,
+  \label{th:q-cont}
+The action-value function of a near-continuous MDP is related to its
+value function via
+% Let $\pi$ be a deterministic policy such that for all $\deltat$, $V^\pi_\deltat$ is continuously
+% differentiable, then, for all $s, a$,
 \begin{equation}
@@
 \end{equation}
-
-  \label{th:q-cont}
+when $\deltat \to 0$ for every $(s,a)$, under suitable technical
+assumptions
+\TODO{ref appendix}.
 \end{theorem}
@@
 
-A consequence of Theorem \ref{th:q-cont} is that in exact continuous time,
-$Q^\pi$ does not bear any information on the ranking of actions, and
-thus cannot be used to select actions that yields higher future returns. There is no time continuous $Q$-learning.
+As a consequence, in exact continuous time,
+$Q^\pi$ is equal to $V^\pi$:  it does not bear \emph{any} information on the ranking of actions, and
+thus cannot be used to select actions that yields higher future returns.
+There is no time continuous $Q$-learning.
 

### 2019-01-19 paper/icml_drau/framework.tex

@@
 
-A natural time discretization is obtained  by defining the discretized return
-return $R_\delta$ for each $\deltat$ as
+A natural time discretization is obtained  by defining the discretized
+return $R_\deltat$ of the MDP $\mathcal{M}_{\deltat}$ as
 \begin{equation}
 \label{eq:discretized-return}
-R_\deltat(\tau) \deq \sum_{k=0}^\infty\gamma^{k\deltat}
-r(s_\deltat^k, a_\deltat^k)\deltat
+R_\deltat(\tau) \deq \sum_{k=0}^\infty\gamma^{k\deltat}\,
+r(s_\deltat^k, a_\deltat^k)\,\deltat
 \end{equation}
@@
 when $\deltat \to 0$ for every $(s,a)$, under suitable technical
-assumptions
-\TODO{ref appendix}.
+assumptions.
 \end{theorem}
@@
 \begin{proof}
-Bellman equation for the discretized $Q$-function yields
+The discrete-time $Q$-function satisfies the Bellman equation
   \begin{align}
-    Q^\pi_\deltat(s, a) &= r(s, a)\deltat + \gamma^\deltat \E_{s'|s, a}\left[V^\pi_\deltat(s')\right] 
+    Q^\pi_\deltat(s, a) &= r(s, a)\,\deltat + \gamma^\deltat \E_{s'|s,
+    a}\left[V^\pi_\deltat(s')\right] .
   \end{align}
-  From the environment dynamic equation, 
+The dynamics \eqref{eq:diffusion} of the environment yields
   \begin{equation}
-  s' = s + f(s, a) \deltat + \smallo(\deltat).
-  \end{equation}As $V^\pi$ is assumed to be continuously differentiable,
+  s' = s + F(s, a) \,\deltat + \smallo(\deltat).
+  \end{equation}
+  Assuming that $V^\pi_\deltat$ is continuously differentiable with
+  respect to the state, and that its derivatives are uniformly bounded, we
+  obtain,
   \begin{align}
-  V^\pi_\deltat(s') &= V^\pi_\deltat(s) + \partial_s V^\pi_\deltat(s) F(s, a)\deltat + \smallo(\deltat)\\
+  V^\pi_\deltat(s') &= V^\pi_\deltat(s) + \nabla_s V^\pi_\deltat(s)\cdot
+  F(s, a)\,\deltat + \smallo(\deltat)\\
 		    &= V^\pi_\deltat(s) + \bigO(\deltat)

### 2019-01-19 paper/icml_drau/framework.tex

@@
 
-\paragraph{Return and Discount Factor.}
+\paragraph{Return and discount factor.}
 Suitable 
@@
 \label{eq:continuous-return}
-R(\tau) \deq \int_{t=0}^\infty\gamma^tr(s_t, a_t)dt.
+R(\tau) \deq \int_{t=0}^\infty\gamma^t\,r(s_t, a_t)\,dt.
 \end{equation}
@@
   V^\pi(s) &= \E_{\tau\sim\pi}\left[R(\tau) \mid s_0 = s\right]\\
-  &= \E_{\tau\sim\pi}\left[\int\limits_{0}^\infty \gamma^{t} r(s_t, a_t) dt \mid s_0 = s\right].
+  &= \E_{\tau\sim\pi}\left[\int\limits_{0}^\infty \gamma^{t}\, r(s_t,
+  a_t) \,dt \mid s_0 = s\right].
 \end{align}
@@
   V^\pi_\deltat(s) &= \E_{\tau\sim\pi}\left[R_\deltat(\tau) \mid s_0 = s\right]\\
-  &= \E_{\tau\sim\pi}\left[\sum\limits_{k=0}^\infty \gamma^{k\deltat}r(s^k_{\deltat}, a^k_{\deltat})\deltat \mid s_0 = s\right]
+  &= \E_{\tau\sim\pi}\left[\sum\limits_{k=0}^\infty
+  \gamma^{k\deltat}\,r(s^k_{\deltat}, a^k_{\deltat})\,\deltat \mid s_0 = s\right]
 \end{align}
@@
   \label{eq:bellman}
-  V^\pi_\deltat(s) = \reward(s, \pi(s))\deltat + \gamma^{\deltat} \E_{s_{\deltat}^{k+1} | s_{\deltat}^{k} = s} V^\pi_\deltat(s_{\deltat}^{k+1})
+  V^\pi_\deltat(s) = \reward(s, \pi(s))\,\deltat + \gamma^{\deltat}\, \E_{s_{\deltat}^{k+1} | s_{\deltat}^{k} = s} V^\pi_\deltat(s_{\deltat}^{k+1})
 \end{equation}
@@
 	\label{eq:q-discr}
-	Q^\pi_\deltat(s, a) = V^\pi(s) + \bigO\left(\deltat\right)
+	Q^\pi_\deltat(s, a) = V^\pi_\deltat(s) + \bigO\left(\deltat\right)
 \end{equation}
@@
   \end{align}
-  Rewriting $Q^\pi_\deltat$ yields
+  Expanding $V^\pi_\deltat(s')$ into $Q^\pi_\deltat$ yields
   \begin{align}
@@
   \end{align}
-  Now, using Thm.~\ref{th:conv-value},
-  \begin{align}
-  Q^\pi_\deltat(s, a) = V^\pi(s) + \bigO(\deltat)
-  \end{align}
+%   Now, using Thm.~\ref{th:conv-value},
+%   \begin{align}
+%   Q^\pi_\deltat(s, a) = V^\pi(s) + \bigO(\deltat)
+%   \end{align}
 which ends the proof.
-%  But we know that $s'$ is close to $s$. If we define $\delta s = s' - s$, then:
+% %  But we know that $s'$ is close to $s$. If we define $\delta s = s' - s$, then:
 %  \begin{equation}

### 2019-01-19 paper/icml_drau/framework.tex

@@
 $Q^\pi$ is equal to $V^\pi$:  it does not bear \emph{any} information on the ranking of actions, and
-thus cannot be used to select actions that yields higher future returns.
+thus cannot be used to select actions that yields higher returns.
 There is no time continuous $Q$-learning.
@@
   \begin{align}
-  Q^\pi_\deltat(s, a) &= r(s, a)\deltat + \gamma^\deltat (V^\pi_\deltat(s) + \bigO(\deltat))\\
-		      &= (1 + \bigO(\deltat)) (V^\pi_\deltat(s) + \bigO(\deltat)) + \bigO(\deltat)\\
+  Q^\pi_\deltat(s, a) &= r(s, a)\,\deltat + \gamma^\deltat (V^\pi_\deltat(s) + \bigO(\deltat))\\
+		      &= \bigO(\deltat)+(1 + \bigO(\deltat)) (V^\pi_\deltat(s) + \bigO(\deltat)) \\
 		      &= V^\pi_\deltat(s) + \bigO(\deltat).

### 2019-01-19 paper/icml_drau/framework.tex

@@
 \footnote{
-If the continuous MDP dynamic follows~\eqref{eq:diffusion}, the limit of the Bellman equation for $V^\pi_\deltat$ when $\deltat \rightarrow 0$ is the Hamilton-Jacobi-Bellman equation on $V^\pi$ \cite{cont_rl}
-\begin{equation}
-  \label{eq:hamilton-jacobi-bellman}
-  r + \partial_s V^\pi \cdot F = - \log(\gamma) V^\pi
-\end{equation}
+If the continuous MDP follows the dynamics \eqref{eq:diffusion}, the limit of
+the Bellman equation \eqref{eq:bellman} for $V^\pi_\deltat$ when $\deltat \rightarrow 0$ is
+the \emph{Hamilton--Jacobi--Bellman} equation on $V^\pi$ \cite{cont_rl},
+namely,
+%\begin{equation}
+  %\label{eq:hamilton-jacobi-bellman}
+  $r + \nabla_s V^\pi \cdot F = - \log(\gamma) V^\pi$.
+%\end{equation}
 }

### 2019-01-19 paper/icml_drau/experiments.tex
experiments

--- 2019-01-10 paper/icml_drau/discussions.tex
@@
 %! TEX root = icml_drau.tex
-\section{Discussions}
-\label{sec:discussions}
+\section{Experiments}
+\label{sec:exp}
+\input{lc_fig.tex}
+The experiment set that we provide here is specifically aimed at showing that
+the proposed method can operate efficiently on a wide range of time
+discretization, without specific tuning, while standard deep Q-learning
+approaches can't.  Our method, which is referred to as \emph{Deep Advantage
+Updating}\TODO{Introduce the name of the algorith, earlier}, is compared to DDPG in continuous action settings and to DQN in
+discrete action settings.
 
-TODO
+In all experimental settings, we use the learning algorithm described in
+alg.~\ref{alg:dau} and \TODO{supplementary, algo 1}. The variant of DDPG and DQN
+used are described in the supplementary material, as well as all the hyperparameters
+used. To make comparison fair, we still use rescaled discount factor $\gamma_\deltat$ and reward $r_\deltat$  as defined in equations \TODO{ref}, as well as scaled learning rates for DDPG and DQN.
+
+Let us stress that quantities shown are rescaled to make comparison possible. For example,
+return results are given in term of discretized return $R_\deltat$ as defined in Eq. \eqref{eq:discretized-return},\footnote{This mostly amounts to scaling rewards
+by a factor $\deltat$ when this scaling is not naturally done in the environment. Environment specific
+details are given in the supplementary material.} and, most notably, time elapsed is always given in
+normalized epochs, i.e. the true number of epochs multiplied by the discretization timestep.
+
+\paragraph{Qualitative experiments: Visualizing policies and values}
+To provide qualitative results, and check that robustness to time
+discretization is provided both in term of raw return results, but also in term
+of convergence of approximate value function and policies, we first provide results on the simple pendulum environment
+provided in the openai gym classic control suite.
+In the simple pendulum environment, the state space is of dimension $2$.
+A visualization of both the learnt value and policy functions by plotting for each point of the phase diagram $(\theta, \dot{\theta})$ its value and policy.
+Qualitative results are presented in
+Fig.~\ref{fig:pend} for the policy and in Fig.\TODO{ref} in Supplementary \TODO{ref} for the value function.
+
+We observe the learnt policy at several instants in physical time (or normalized epochs) during training, for various time discretizations $\deltat$, for both DAU and DDPG.
+With DAU, the agent's policy quickly converges for every time discretization without specific hyperparameter tuning.
+On the contrary, with DDPG, the agent is only able to learn for the two highest time discretization, and learning gets harder as $\deltat$ decreases.
+
+\paragraph{Quantitative experiments}
+We benchmark DAU against DDPG on classic control benchmarks: Pendulum, Cartpole, BipedalWalker, Ant and Half-Cheetah environments from Open AI Gym.
+For Pendulum and Ant, DDPG is able to learn for the highest $\deltat$ but fails for the others, while DAU remains reasonably invariant to the time discretization.
+On Bipedal Walker, DDPG is never able to learn while DAU is solving the environment for every $\deltat$.
+On Cartpole and Half Cheetah, the results are less clear.
+On Cartpole, noise dominates, which makes interpretation unclear.
+On Half Cheetah, DAU is not clearly more invariant to time discretization than DDPG.
+This could be explained by the multiple suboptimal regimes that coexist in the half cheetah environment (walking on the head, walking on the back), and that DAU may fail to correctly weight the performance of each regime (as explained in the Section \ref{sec:discussions}).
+
+% \begin{figure}[h]
+%   \centering
+%   \includegraphics[width=\columnwidth]{figs_data/cartpole_lc.png}
+%   \caption{Cartpole}
+%   \label{fig:cartpole-lc}
+% \end{figure}
+
+% \begin{figure}[h]
+%   \centering
+%   \includegraphics[width=\columnwidth]{figs_data/ant_lc.png}
+%   \caption{Ant}
+%   \label{fig:ant-lc}
+% \end{figure}
+% 
+% \begin{figure}[h]
+%   \centering
+%   \includegraphics[width=\columnwidth]{figs_data/pendulum_lc.png}
+%   \caption{Pendulum}
+%   \label{fig:pendulum-lc}
+% \end{figure}

### 2019-01-19 paper/icml_drau/framework.tex

@@
 $Q^\pi$ is equal to $V^\pi$:  it does not bear \emph{any} information on the ranking of actions, and
-thus cannot be used to select actions that yields higher returns.
-There is no time continuous $Q$-learning.
+thus cannot be used to select actions with higher returns.
+There is no continuous-time $Q$-learning.
 

### 2019-01-19 paper/icml_drau/intro.tex
relecture intro

@@
 control environments, robotics, and many video games.  
-(We focus on
+\TODO{Remove parenthesis.} (We focus on
 near-continuous time because this leads to a clear framework, but the
@@
 value functions, such as
-\emph{Deep $Q$-learning} (DQN~\citep{dqn} and \emph{Deep deterministic policy
-gradient} (DDPG~\citep{ddpg}) are not at all robust to changes in time discretization. This is shown experimentally in Sec.~\ref{sec:exp}. 
+\emph{Deep $Q$-learning} (DQN,~\citealt{dqn}) and \emph{Deep deterministic policy
+gradient} (DDPG,~\citealt{ddpg}) are not at all \TODO{Remove familiar?}robust to changes in time discretization. This is shown experimentally in Sec.~\ref{sec:exp}. 
 Intuitively, as the discretization timestep
 decreases, the effect of individual actions on the total return decreases
-too: $Q(s,a)$ is the value of playing action $a$ then playing optimally,
+too: $Q^*(s,a)$ is the value of playing action $a$ then playing optimally,
 and if $a$ is only maintained for a very short time its advantage over
 other actions will be accordingly small. (This continuity effect occurs even with a
-well-adjusted decay factor $\gamma$.)
+well-adjusted decay factor $\gamma$.) \TODO{Really relevant in the introduction?}
 
@@
 
-A time discretization invariant algorithm should in particular admit a
+\TODO{Time discretization invariant algo is never  mentioned before.
+Maybe start by ``We seek a time discretization invariant algorithm.
+It should ...''} A time discretization invariant algorithm should \TODO{remain viable when the time discretization is small and} in particular admit a
 continuous-time
@@
   robust to changes in time discretization, exhibiting degraded performance, while our algorithm provides
-  substantial robustness in that domain.
+  substantial robustness in that domain. \TODO{Remove ``in that domain''?}
 \end{itemize}

### 2019-01-19 paper/icml_drau/framework.tex
relecture framework

@@
 the simulator, and is often used to approximate an underlying
-differential equation.  For example, the timestep may correspond to the
+differential equation.  \TODO{(For example) $->$ In this case}, the timestep may correspond to the
 number of frames generated by second.  In real world environments,
@@
 would expect that decreasing $\deltat$ could only improve the performance
-of reinforcement learning agents (though it might make optimization
-harder).  Reinforcement
-learning algorithms should, at least, be resilient to a change of
+of RL agents \TODO{We would remove: (though it might make optimization
+harder) It is not in the spririt of what is said here}.  RL algorithms should, at least, be resilient to a change of
 $\deltat$, and should remain viable when $\deltat \rightarrow 0$.
@@
 
-We are thus interested in the behavior of reinforcement learning
+We are thus interested in the behavior of RL
 algorithms in discretized environments, when the discretization timestep
@@
       \emph{time discretization} $\deltat$. The %deterministic
-      transition function
+      transition function of a state $s$
       %$T_\deltat(s, a, s')$ 
@@
         \end{equation}
-which is thus stable when $\deltat\to 0$.
+which is thus stable when $\deltat\to 0$. \TODO{We need to add that $\gamma$ is close to 1}
 On the other hand,
@@
 \begin{proof}
-The discrete-time $Q$-function satisfies the Bellman equation
+The discrete-time $Q$-function of the MDP $\mathcal{M}_\deltat$ satisfies the Bellman equation
   \begin{align}

### 2019-01-19 paper/icml_drau/advup.tex
experiments
relecture advup

@@
 \end{align}
-a rescaled version of the advantage function \cite{adv_upd}, as the difference between between
-$Q^\pi_\deltat(s, a)$ and $V^\pi\deltat(s)$ is of order $\bigO(\deltat)$.
+a rescaled version of the advantage function, as the difference between between
+$Q^\pi_\deltat(s, a)$ and $V^\pi_\deltat(s)$ is of order $\bigO(\deltat)$.
 Contrary to the action state value function, the rescaled advantage function converges when $\deltat$ goes to $0$
@@
 when $Q$ correctly approximates $Q^\pi_\deltat$, $A$ approximates $A^\pi_\deltat$\footnote{
-	Ensuring that $A$ do not differ from $A^\pi_\deltat$ by a state dependent function only affects
+	Ensuring that $A$ does not differ from $A^\pi_\deltat$ by a state dependent function only affects
 	$A$'s interpretability. Indeed, for any function $f$, the ranking of actions given by
@@
 With this parameterization, one can hope to learn an approximation of $A^\pi_\deltat$,
-from which a proper hiearchy of actions can be derived for any $\deltat$.
+from which a proper ranking of actions can be derived for any $\deltat$.
 
@@
 \label{subsec:explo}
-To obtain a time invariant reinforcement learning algorithm, a time invariant
+To obtain a time invariant RL algorithm, a time invariant
 exploration scheme is required. 
-\cite{ddpg} already introduces an exploration scheme for continuous actions, which consists in adding an
-\emph{Ornstein-Uhlenbeck}~\cite{orn-uhl} process (OU process) to actions, that is naturally time
-discretization invariant. Formally, it is defined as
+\cite{ddpg} already introduces a time discretization invariant exploration scheme for continuous actions, which consists in adding an
+\emph{Ornstein-Uhlenbeck}~\cite{orn-uhl} process (OU process) to actions. Formally, it is defined as
 \begin{equation}
@@
 \end{equation}
-with $z^k_\deltat$ the discretization of a time continuous OU process, defined as
+with $z^k_\deltat$ the discretization of a time continuous OU process, 
 \begin{equation}
-	dz_t = - z_t \theta dt + \sigma \dbt.
+	dz_t = - z_t \kappa dt + \sigma \dbt.
 	\label{eq:orn_uhl}
 \end{equation}
-where $B_t$ is a brownian motion. Resulting discretized trajectories converge
+where $B_t$ is a brownian motion, $\kappa$ a stiffness parameter and $\sigma$ a noise scaling parameter. Resulting discretized trajectories converge
 to non trivial continuous time trajectories.
@@
 continuous deterministic environment, the exploration policy resulting from
-applying $\varepsilon$ greedy exploration to a constant exploitation policy
+applying $\varepsilon$-greedy exploration to a constant exploitation policy
 converges to a determinist policy as $\deltat$ goes to $0$. More precisely, let
-$(a_\deltat)$ a sequence of i.i.d. actions of distribution $p_A$, (for instance
+$(a_\deltat)$ be a sequence of i.i.d. actions of distribution drawn from $p_A$, (for instance
 the sequence of actions generated by $\varepsilon$-greedy exploration on a
@@
 where the expectation cancels the exploration noise. The proof is given in the
-supplementary material.
+supplementary material. \TODO{Is there a stronger theorm? Should we talk more about that?}
 
@@
   \caption{Deep Advantage Updating (Discrete actions)}
+  \label{alg:dau}
 	\input{dauc.tex}
-	\label{alg:dau}
 \end{algorithm}
@@
 		\item If $\beta > 1$, parameter trajectories shrink to a single point as 
-			$\deltat$ goes to $0$.
+			$\deltat$ goes to $0$. \TODO{Is that clear}
 	\end{itemize}

### 2019-01-19 paper/icml_drau/advup.tex
relecture experiments

@@
 
-In what follows, a discrete algorithm with a well defined time continuous
-limit is defined. The algorithm relies on three elements, namely, defining and
-learning a quantity that contains some information on the ranking of actions
-while admitting a time continuous limit, defining exploration methods that
-admit a continuous time limit and defining learning rates that induce well
-behaved parameter trajectories when $\deltat$ goes to $0$.
+We now define a discrete algorithm with a well-defined continuous-time
+limit.  It relies on three elements: defining and learning a quantity
+that still contains information on action rankings in the limit, using
+exploration methods with a meaningful limit, and scaling learning rates
+to induce well-behaved parameter trajectories when $\deltat$ goes to $0$.
 
-As seen in Sec.~\ref{sec:framework}, there is no continuous time limit to
-Q-learning. In continuous time, $Q^\pi$ is independent of actions and can thus
-not be used to select promising actions.  In near continuous time, $Q^\pi_\deltat$
-still depends on actions, and can still be used to choose promising actions
-under the current policy. However, when approximating $Q^\pi_\deltat$, if the
-approximation error is much bigger than $\bigO(\deltat)$, approximation errors
-dominates, and the ranking of actions given by the approximation is likely to
-be erroneous.
+As seen above, there is no continuous time limit to Q-learning. In
+continuous time, $Q^\pi$ is independent of actions and can thus not be
+used to select promising actions.  In near-continuous time,
+$Q^\pi_\deltat$ still depends on actions, and could still be used to
+choose promising actions under the current policy. However, when
+approximating $Q^\pi_\deltat$, if the approximation error is much bigger
+than $\bigO(\deltat)$, approximation errors dominates, and the ranking of
+actions given by the approximation is likely to be erroneous.
+
 % When the approximate Q-function is initialized, if the effect of
@@
 % \end{align}
-The full algorithm can now be derived.
+The full algorithm, that is referred to as \emph{Deep Advantage Updating} (DAU), can now be derived.
 Pseudocode for the discrete actions algorithm is given in Alg.~\ref{alg:dau}.

### 2019-01-19 paper/icml_drau/advup.tex

@@
 
-As seen above, there is no continuous time limit to Q-learning. In
-continuous time, $Q^\pi$ is independent of actions and can thus not be
-used to select promising actions.  In near-continuous time,
-$Q^\pi_\deltat$ still depends on actions, and could still be used to
-choose promising actions under the current policy. However, when
-approximating $Q^\pi_\deltat$, if the approximation error is much bigger
-than $\bigO(\deltat)$, approximation errors dominates, and the ranking of
-actions given by the approximation is likely to be erroneous.
-
-% When the approximate Q-function is initialized, if the effect of
-% actions on the Q-function is order of magnitudes higher than what it should be,
+% When the approximate $Q$-function is initialized, if the effect of
+% actions on the $Q$-function is order of magnitudes higher than what it should be,
 % approximating $Q^\pi_\deltat$ is likely to be difficult.
@@
 \label{subsec:reparam}
+
+As seen above, there is no continuous time limit to $Q$-learning, because
+$Q^\pi$ becomes independent of actions and thus cannot be
+used to select actions.  In near-continuous time,
+$Q^\pi_\deltat$ still depends on actions, and could still be used to
+choose actions. However, when
+approximating $Q^\pi_\deltat$, if the approximation error is much larger
+than $\bigO(\deltat)$, this error dominates, the ranking of
+actions given by the approximated $Q^\pi_\deltat$ is likely to be erroneous.
+
 To define an object which contains the same information on actions as
-$Q^\pi_\deltat$, but admits an action dependant continuous time limit, it is
+$Q^\pi_\deltat$, but admits a learnable action-dependent limit, it is
 natural to define \cite{adv_upd}
@@
 a rescaled version of the advantage function, as the difference between between
-$Q^\pi_\deltat(s, a)$ and $V^\pi_\deltat(s)$ is of order $\bigO(\deltat)$.
-Contrary to the action state value function, the rescaled advantage function converges when $\deltat$ goes to $0$
-to an action dependent quantity which contains the necessary information for policy improvement:
+$Q^\pi_\deltat(s, a)$ and $V^\pi_\deltat(s)$ is of order
+$\bigO(\deltat)$. This amounts to splitting $Q$ into value and advantage,
+and observing that these scale very differently when $\deltat\to 0$.
+%Contrary to the action state value function,
+This
+rescaled advantage function converges when $\deltat\to 0$
+to an action-dependent quantity which, if it can be learned, contains the necessary information for policy improvement:
 
@@
 
-The discretized Q-function can be rewritten as
+The discretized $Q$-function rewrites as
 \begin{equation}
@@
 A natural way of approximating $V^\pi_\deltat$ and $A^\pi_\deltat$ is to apply
-\emph{Sarsa} or Q-learning to a reparameterized Q-function approximator
+\emph{Sarsa} or $Q$-learning to a reparameterized $Q$-function approximator
 \begin{equation}
@@
 
-To learn $V_{\theta}$ and $A_{\psi}$, variants of Q-learning for continuous and
+To learn $V_{\theta}$ and $A_{\psi}$, variants of $Q$-learning for continuous and
 discrete action spaces are used. In details, $V_{\theta}$ and $A_{\psi}$ are learnt

### 2019-01-19 paper/icml_drau/intro.tex

@@
 \emph{Deep $Q$-learning} (DQN,~\citealt{dqn}) and \emph{Deep deterministic policy
-gradient} (DDPG,~\citealt{ddpg}) are not at all \TODO{Remove familiar?}robust to changes in time discretization. This is shown experimentally in Sec.~\ref{sec:exp}. 
+gradient} (DDPG,~\citealt{ddpg}) are not at all \TODO{Remove familiar?
+NDY it's not particularly familiar}robust to changes in time discretization. This is shown experimentally in Sec.~\ref{sec:exp}. 
 Intuitively, as the discretization timestep

### 2019-01-19 paper/icml_drau/intro.tex

@@
 other actions will be accordingly small. (This continuity effect occurs even with a
-well-adjusted decay factor $\gamma$.) \TODO{Really relevant in the introduction?}
+well-adjusted decay factor $\gamma$.) \TODO{Really relevant in the
+introduction? NDY Yes, I predict the first thing some people would say is
+"you just put the wrong gamma, how stupid of you..."}
 

### 2019-01-19 paper/icml_drau/advup.tex

@@
 \end{equation}
-A natural way of approximating $V^\pi_\deltat$ and $A^\pi_\deltat$ is to apply
+A natural way to approximate $V^\pi_\deltat$ and $A^\pi_\deltat$ is to apply
 \emph{Sarsa} or $Q$-learning to a reparameterized $Q$-function approximator
@@
 \end{equation}
-with $\Theta = (\theta, \pi)$. To avoid cumbersome notations, in what follows,
+with $\Theta = (\theta, \pi)$. To avoid cumbersome notation, in what follows,
 parameter indices are dropped.  At initialization, if both $V_{\theta}$ and
 $A_{\psi}$ are initialized independantly of $\deltat$, this parameterization
-reasonably scales the contribution of actions relative to states in the value
-of $Q$.  Our goal is for $V_\theta$ to approximate $V^\pi_\deltat$ and for
+provides reasonable scaling of the contribution of actions versus states
+in $Q$.
+Our goal is for $V_\theta$ to approximate $V^\pi_\deltat$ and for
 $A_{\psi}$ to approximate $A^\pi_\deltat$.
@@
 %\TODO{CT and LB are fighting each other about the importance which should be given to the next paragraph. A impartial judge should give his opinion. (CT is wrong)}
-However, the proposed reparameterization does not, on its own, guarantee that
+Still, this reparameterization does not, on its own, guarantee that
 when $Q$ correctly approximates $Q^\pi_\deltat$, $A$ approximates $A^\pi_\deltat$\footnote{
@@
 	$A$'s interpretability. Indeed, for any function $f$, the ranking of actions given by
-	$A^\pi_\deltat(s, \cdot)$ and $A^\pi_\deltat(s, \cdot) + f(s)$ is the same.
+	$A^\pi_\deltat(s, \cdot)$ and $A^\pi_\deltat(s, \cdot) + f(s)$ is
+	the same. \TODO{still unclear why we'd care about identifiability}
 }.
-Indeed, for any given pair $V$, $A$,
-\begin{align}
-	\tilde{V}(s) &= V(s) - f(s)\\
-	\tilde{A}(s, a) &= A(s, a) + \frac{f(s)}{\deltat}
-\end{align}
+Indeed, for any given pair $(V,\,A)$, the pair $(V(s) -
+f(s),\,A(s,a)+f(s)/\deltat)$ (for an arbitrary $f$)
+% \begin{align}
+% 	\tilde{V}(s) &= V(s) - f(s),\qquad
+% 	\tilde{A}(s, a) = A(s, a) + \frac{f(s)}{\deltat}
+% \end{align}
 yields the exact same $Q$ function. Consequently, $Q$ can correctly approximate
@@
 \end{equation}
-With this additional constraint, if $Q = Q^\pi_\deltat$, then $A = A^\pi_\deltat$ and $V = V^\pi_\deltat$.
+With this additional constraint, if $Q = Q^\pi_\deltat$, then $A =
+A^\pi_\deltat$ and $V = V^\pi_\deltat$: indeed
 \begin{align}
@@
 \label{eq:Aparam}
-	A(s, a) = \bar{A}(s, a) - \bar{A}(s, \pi(s)).
+	A(s, a) \deq \bar{A}(s, a) - \bar{A}(s, \pi(s)).
 \end{equation}

### 2019-01-19 paper/icml_drau/advup.tex

@@
 \begin{equation}
-	Q_\Theta(s, a) = V_{\theta}(s) + \deltat A_{\psi}(s, a).
+	Q_\Theta(s, a) \deq V_{\theta}(s) + \deltat A_{\psi}(s, a).
 \end{equation}
-with $\Theta = (\theta, \pi)$. To avoid cumbersome notation, in what follows,
-parameter indices are dropped.  At initialization, if both $V_{\theta}$ and
+with $\Theta \deq (\theta, \psi)$. (To avoid cumbersome notation, 
+parameter indices are dropped in what follows.)  At initialization, if both $V_{\theta}$ and
 $A_{\psi}$ are initialized independantly of $\deltat$, this parameterization

### 2019-01-19 paper/icml_drau/advup.tex

@@
 and observing that these scale very differently when $\deltat\to 0$.
-%Contrary to the action state value function,
-This
+
+Contrary to the action state value function,
+this
 rescaled advantage function converges when $\deltat\to 0$
-to an action-dependent quantity which, if it can be learned, contains the necessary information for policy improvement:
+to an action-dependent quantity which, if it can be learned, contains the
+necessary information for policy improvement \TODO{See Thm... in the appendix; we
+should put the theorem in the appendix, because otherwise the theorem is
+just a paraphrase of the previous sentence}.
 
@@
 
-  \begin{theorem}
-    There exists $A^\pi(s, a)$ such that for all $a, s$:
-    \begin{equation}
-      A^\pi_\deltat(s, a) \rightarrow_{\deltat \rightarrow 0} A^\pi(s, a)
-    \end{equation}
-    Moreover, $A^\pi$ contains all the necessary information for policy improvement. The policy $\pi$ optimal if and only if $A^\pi(s, \pi(s)) = \max_{a'}A^\pi(s, a')$ for all $s \in {\cal S}$.
-    \TODO{Is this true? What is true? Is this real life? We should probably add a reasonable hypothesis, for example $\pi_\deltat^* \rightarrow \pi^*$}
-  \end{theorem}
+%   \begin{theorem}
+%     There exists $A^\pi(s, a)$ such that for all $a, s$:
+%     \begin{equation}
+%       A^\pi_\deltat(s, a) \rightarrow_{\deltat \rightarrow 0} A^\pi(s, a)
+%     \end{equation}
+%     Moreover, $A^\pi$ contains all the necessary information for policy improvement. The policy $\pi$ optimal if and only if $A^\pi(s, \pi(s)) = \max_{a'}A^\pi(s, a')$ for all $s \in {\cal S}$.
+%     \TODO{Is this true? What is true? Is this real life? We should probably add a reasonable hypothesis, for example $\pi_\deltat^* \rightarrow \pi^*$}
+%   \end{theorem}
 
@@
 parameter indices are dropped in what follows.)  At initialization, if both $V_{\theta}$ and
-$A_{\psi}$ are initialized independantly of $\deltat$, this parameterization
+$A_{\psi}$ are initialized independently of $\deltat$, this parameterization
 provides reasonable scaling of the contribution of actions versus states
@@
 %\TODO{CT and LB are fighting each other about the importance which should be given to the next paragraph. A impartial judge should give his opinion. (CT is wrong)}
-Still, this reparameterization does not, on its own, guarantee that
-when $Q$ correctly approximates $Q^\pi_\deltat$, $A$ approximates $A^\pi_\deltat$\footnote{
-	Ensuring that $A$ does not differ from $A^\pi_\deltat$ by a state dependent function only affects
-	$A$'s interpretability. Indeed, for any function $f$, the ranking of actions given by
-	$A^\pi_\deltat(s, \cdot)$ and $A^\pi_\deltat(s, \cdot) + f(s)$ is
-	the same. \TODO{still unclear why we'd care about identifiability}
-}.
+Still, this reparameterization does not, on its own, guarantee that $A$
+correctly approximates $A^\pi_\deltat$ if
+$Q$ approximates $Q^\pi_\deltat$% \footnote{
+% 	Ensuring that $A$ does not differ from $A^\pi_\deltat$ by a state dependent function only affects
+% 	$A$'s interpretability. Indeed, for any function $f$, the ranking of actions given by
+% 	$A^\pi_\deltat(s, \cdot)$ and $A^\pi_\deltat(s, \cdot) + f(s)$ is
+% 	the same. \TODO{still unclear why we'd care about identifiability}
+%}
+.
 Indeed, for any given pair $(V,\,A)$, the pair $(V(s) -
@@
 % \end{align}
-yields the exact same $Q$ function. Consequently, $Q$ can correctly approximate
-$Q^\pi$ even if $A$ is arbitrarily far from $A^\pi$.
+yields the exact same $Q$ function. This new $A$ still defines %the same $Q$ and
+the same ranking of actions, yet this phenomenon might cause
+numerical problems or instability of $A$ when $\deltat\to 0$, and prevents direct
+interpretation of the learned $A$.
+% Consequently, $Q$ can correctly approximate
+% $Q^\pi$ even if $A$ is arbitrarily far from $A^\pi$.
 To enforce identifiability of $A$, one must enforce the consistency equation
@@
 \end{equation}
-With this parameterization, one can hope to learn an approximation of $A^\pi_\deltat$,
+With this parameterization, one can hope to learn an approximation of $A^\pi$,
 from which a proper ranking of actions can be derived for any $\deltat$.

### 2019-01-19 pendulum_exp/dataanalysis/pend_plot.py
.

@@ -1,0 +1,83 @@
+"""Retrieves pendulum figure from the last version of pendulum_final exp."""
+import pickle
+from typing import List, Dict, Tuple
+import os
+from os.path import join, isdir
+from datetime import datetime
+import matplotlib.pyplot as plt
+import matplotlib.image as mpimg
+
+timesteps = [2, 4, 6, 8, 10]
+nb_ts = len(timesteps)
+exp_dir = 'local_logs/pendulum_final/'
+max_time = datetime.strptime("1999_01_01_01_01_01", "%Y_%m_%d_%H_%M_%S")
+max_dir = None
+
+for sub_dir in os.listdir(exp_dir):
+    time = datetime.strptime(sub_dir, "%Y_%m_%d_%H_%M_%S")
+    max_dir = sub_dir if time > max_time or not max_dir else max_dir
+
+assert max_dir
+exp_dir = join(exp_dir, max_dir)
+
+exp_dirs: List[str] = []
+for sub_dir in os.listdir(exp_dir):
+    full_dir = join(exp_dir, sub_dir)
+    if isdir(full_dir):
+        exp_dirs.append(full_dir)
+
+dt_dict: Dict[Tuple[str, float], str] = {}
+max_dt = 0.
+nb_dt = 0
+for sub_dir in exp_dirs:
+    args = pickle.load(open(join(sub_dir, 'args'), 'rb'))
+    key = (args.algo, args.dt)
+    max_dt = max(args.dt, max_dt)
+    if key not in dt_dict:
+        dt_dict[key] = sub_dir
+        nb_dt += 1
+    else:
+        dt_dict[key] = sub_dir
+assert nb_dt % 2 == 0
+nb_dt = nb_dt // 2
+
+val_img_dict: Dict[Tuple[str, float], List[str]] = {k: [] for k in dt_dict}
+act_img_dict: Dict[Tuple[str, float], List[str]] = {k: [] for k in dt_dict}
+for key, value in dt_dict.items():
+    for ts in timesteps:
+        val_img_dict[key].append(join(value, 'imgs', f'val_{int(ts * max_dt / key[1])}.png'))
+        act_img_dict[key].append(join(value, 'imgs', f'act_{int(ts * max_dt / key[1])}.png'))
+
+dts = sorted(list(set([k[1] for k in dt_dict])))
+print(f"List of dts: {', '.join([str(dt) for dt in dts])}")
+
[…]

### 2019-01-19 paper/icml_drau/dauc.tex

@@
 	\STATE $\pi^{\text{explore}}$ and $\nu_\deltat$ defining an exploration policy.
-	\STATE \textbf{opt}$_1$, \textbf{opt}$_2$, $\alpha_1 \deltat$ and $\alpha_2 \deltat$ optimizers and learning rates.
+	\STATE \textbf{opt}$_1$, \textbf{opt}$_2$, $\alpha^V \deltat$ and $\alpha^V \deltat$ optimizers and learning rates.
 	\STATE $\mathcal{D}$, buffer of transitions $(s, a, r, d, s')$, with $d$ the episode termination signal.
@@
 	\STATE $\Delta \psi \gets \frac{1}{N}\sum\limits_{i=1}^N \frac{\left(Q^i - \tilde{Q^i}\right)\partial_{\psi} \left(\bar{A}_{\psi}(s^i, a^i) - \max\limits_{a'}\bar{A}_{\psi}(s^i, a')\right) }{\deltat}$
-	\STATE Update $\theta$ with \textbf{opt}$_1$, $\Delta \theta$ and learning rate $\alpha_1 \deltat$.
-	\STATE Update $\psi$ with \textbf{opt}$_2$, $\Delta \psi$ and learning rate $\alpha_2 \deltat$.
+	\STATE Update $\theta$ with \textbf{opt}$_1$, $\Delta \theta$ and learning rate $\alpha^V \deltat$.
+	\STATE Update $\psi$ with \textbf{opt}$_2$, $\Delta \psi$ and learning rate $\alpha^A \deltat$.
 	\ENDFOR

### 2019-01-19 paper/icml_drau/discussions.tex

@@
 A different choice would be to scale the size of the buffer by a factor of
-$\frac{1}{\deltat}$. This would amount to always learning on the same time physical interval of
+$\frac{1}{\deltat}$. This would amount to always learning on the same physical time interval of
 the trajectory. Our analysis, however, does not cover this case.
 
-\paragraph{Near continuous time reinforcement learning and RMSprop}~Be careful
-when using RMSprop. When you are facing a stochastic environment RMSprop
-evaluates the second momentum of gradients and rescale gradients by its square
-root. This second momentum will be of order δt 1 , due to the stochasticity in
-the environment.  Consequently all √ gradient steps are multiplied by δt, and
-learning rates need √ only be multiplied by δt instead of δt. Taking batches of
-size proportional to δt 1 applies a similar fix.
+\paragraph{Near continuous time reinforcement learning and RMSprop}~RMSprop~\cite{rmsprop}
+estimates a batched version of the second momentum of gradients using moving
+averages, and divides gradient steps by the square root of this estimate. In
+deterministic environments, gradient noise typically scales as $\bigO(1)$ in
+term of $\deltat$, as seen in Eq.~\eqref{eq:approx_bellman_A}.  Consequently, RMSprop
+preconditioning have no effect on the order of magnitude of required learning
+rates. However, in near continuous stochastic environments, as in
+Eq.~\ref{eq:sde}, variance of the gradient noise typically scales as
+$\bigO\left(\frac{1}{\deltat}\right)$. If batch size does not depend on
+$\deltat$, RMSprop preconditioning thus multiplies gradients by a factor $\bigO(\sqrt{\deltat})$,
+and learning rates need only be scaled as $\alpha \sqrt{\deltat}$ instead of
+$\alpha \deltat$.

### 2019-01-19 paper/icml_drau/experiments.tex

@@
 The experiment set that we provide here is specifically aimed at showing that
-the proposed method can operate efficiently on a wide range of time
+the proposed method DAU can operate efficiently on a wide range of time
 discretization, without specific tuning, while standard deep Q-learning
-approaches can't.  Our method, which is referred to as \emph{Deep Advantage
-Updating}\TODO{Introduce the name of the algorith, earlier}, is compared to DDPG in continuous action settings and to DQN in
+approaches can't. DAU is compared to DDPG in continuous action settings and to DQN in
 discrete action settings.
@@
 of convergence of approximate value function and policies, we first provide results on the simple pendulum environment
-provided in the openai gym classic control suite.
-In the simple pendulum environment, the state space is of dimension $2$.
-A visualization of both the learnt value and policy functions by plotting for each point of the phase diagram $(\theta, \dot{\theta})$ its value and policy.
-Qualitative results are presented in
-Fig.~\ref{fig:pend} for the policy and in Fig.\TODO{ref} in Supplementary \TODO{ref} for the value function.
+from the OpenAI Gym classic control suite.
+This environment's state space is of dimension $2$.
+A visualization of both the learnt value and policy functions is obtained by plotting for each point of the phase diagram $(\theta, \dot{\theta})$ its value and policy.
+results are presented in
+Fig.~\ref{fig:pend} for the policy function and in Fig.\TODO{ref} in Supplementary \TODO{ref} for the value function.
 
@@
 \paragraph{Quantitative experiments}
-We benchmark DAU against DDPG on classic control benchmarks: Pendulum, Cartpole, BipedalWalker, Ant and Half-Cheetah environments from Open AI Gym.
+We benchmark DAU against DDPG on classic control benchmarks: Pendulum, Cartpole, BipedalWalker, Ant and Half-Cheetah environments from OpenAI Gym.
 For Pendulum and Ant, DDPG is able to learn for the highest $\deltat$ but fails for the others, while DAU remains reasonably invariant to the time discretization.
-On Bipedal Walker, DDPG is never able to learn while DAU is solving the environment for every $\deltat$.
-On Cartpole and Half Cheetah, the results are less clear.
-On Cartpole, noise dominates, which makes interpretation unclear.
+On Bipedal Walker, DDPG is never able to learn while DAU is achieving decent performance for all the $\deltat$'s.
+On Cartpole and Half Cheetah, results are less clear.
+On Cartpole, noise dominates, which makes interpretation difficult.
 On Half Cheetah, DAU is not clearly more invariant to time discretization than DDPG.
-This could be explained by the multiple suboptimal regimes that coexist in the half cheetah environment (walking on the head, walking on the back), and that DAU may fail to correctly weight the performance of each regime (as explained in the Section \ref{sec:discussions}).
+This could be explained by the multiple suboptimal regimes that coexist in the half cheetah environment (walking on the head, walking on the back), and that DAU may fail to correctly weight the performance of each regime (as explained in Section \ref{sec:discussions}).
 

### 2019-01-19 paper/icml_drau/framework.tex

@@
 of RL agents \TODO{We would remove: (though it might make optimization
-harder) It is not in the spririt of what is said here}.  RL algorithms should, at least, be resilient to a change of
+harder) It is not in the spririt of what is said here. NDY keep it.
+Anyone with a bit of experience in traditional optim would scream and say exactly
+that, if it was not included.}.  RL algorithms should, at least, be resilient to a change of
 $\deltat$, and should remain viable when $\deltat \rightarrow 0$.

### 2019-01-19 paper/icml_drau/relatedworks.tex

@@
 the advantage part; testing was only done on a quadratic-linear problem.
+\TODO{mention that Baird did not experimentally study robustness}
 
@@
 timescale and the physical system's timescale, but the collapse of the
-$Q$ function to the $V$ function is an intrinsic mathematical phenomenon from time
+$Q$ function to the $V$ function is an intrinsic mathematical phenomenon \TODO{arising} from time
 continuity, and cannot obviously be addressed via existing multiscale

### 2019-01-19 paper/icml_drau/advup.tex

@@
 With this additional constraint, if $Q = Q^\pi_\deltat$, then $A =
-A^\pi_\deltat$ and $V = V^\pi_\deltat$: indeed
+A^\pi_\deltat$ and $V = V^\pi_\deltat$: indeed \TODO{keep or  leave as
+an exercise?}
 \begin{align}
@@
 
-\subsection{Time invariant exploration}
+\subsection{Timestep-invariant exploration}
 \label{subsec:explo}
-To obtain a time invariant RL algorithm, a time invariant
+
+To obtain a timestep-invariant RL algorithm, a timestep-invariant
 exploration scheme is required. 
-\cite{ddpg} already introduces a time discretization invariant exploration scheme for continuous actions, which consists in adding an
-\emph{Ornstein-Uhlenbeck}~\cite{orn-uhl} process (OU process) to actions. Formally, it is defined as
+For continuous actions, \cite{ddpg} already introduced a time discretization invariant exploration scheme, which consists in adding an
+\emph{Ornstein--Uhlenbeck} \cite{orn-uhl} (OU) process to actions. Formally, it is defined as
 \begin{equation}
-	\pi^\text{explore}(s^k_\deltat, z^k_\deltat) = \pi(s^k_\deltat) + z^k_\deltat
+	\pi^\text{explore}(s^k_\deltat, z^k_\deltat) \deq \pi(s^k_\deltat) + z^k_\deltat
 \end{equation}
-with $z^k_\deltat$ the discretization of a time continuous OU process, 
+with $z^k_\deltat$ the discretization of a continuous-time OU process,
 \begin{equation}
-	dz_t = - z_t \kappa dt + \sigma \dbt.
+	dz_t = - z_t \,\kappa\, dt + \sigma \,\dbt.
 	\label{eq:orn_uhl}
 \end{equation}
-where $B_t$ is a brownian motion, $\kappa$ a stiffness parameter and $\sigma$ a noise scaling parameter. Resulting discretized trajectories converge
-to non trivial continuous time trajectories.
+where $B_t$ is a brownian motion, $\kappa$ a stiffness parameter and
+$\sigma$ a noise scaling parameter. The discretized trajectories converge
+to non trivial continuous time trajectories, exhibiting a Brownian
+behavior with a recall force towards $0$.
 
@@
 \end{equation}
-with $(z^k_\deltat)_k$ a sequence of random variables, independant from the $a$'s and $s$'s.
-A sufficient condition for the exploratory policy to admit a time continuous
-limit is for
-$z_\deltat$ to admit a limit, i.e. that the sequence $z_\deltat$ converges to a
-well defined continuous stochastic process $z$ as $\deltat$ goes to $0$.
+with $(z^k_\deltat)_{k\geq 0}$ a sequence of random variables independent from the $a$'s and $s$'s.
+A sufficient condition for this policy to admit a continuous-time
+limit is for the sequence
+$z_\deltat$ to converge in law to a
+well-defined continuous stochastic process $z$ as $\deltat$ goes to $0$.
 % For instance, $\varepsilon$-greedy exploration naturally falls in this framework, \TODO{[[CT would like to give these details, LB thinks it is not necessary:]] with each $z^k_\deltat$
@@
 % \end{equation}}
-To obtain a temporally coherent exploration scheme in discrete action environment
-$z_\deltat$ is chosen to be a discretization of a $|\mathcal{A}|$ dimensional
-continuous OU process, and set
+
+Thus, for discrete actions we can obtain a temporally consistent
+exploration scheme by taking
+$z_\deltat$ to be a discretization of an $|\mathcal{A}|$-dimensional
+continuous OU process, and setting
 \begin{equation}
-  \pi^\text{explore}(s_t, z_t)= \text{argmax}_{a'} \left(Q(s, a') + z_t[a']\right)
+  \pi^\text{explore}(s_t, z_t)\deq \text{argmax}_{a'} \left(Q(s, a') + z_t[a']\right)
 \end{equation}
-where $z_t[a']$ denotes the $a'$-th component of the discretized OU process at time
-$k$. The resulting exploration scheme converges to a proper time continuous exploration scheme,
+where $z_t[a']$ denotes the $a'$-th component of $z_t$. Namely, we
+perturb $Q$-values by a random process. The resulting exploration scheme
+converges in continuous time,
 and is non trivial for most $Q$'s.

### 2019-01-19 paper/icml_drau/advup.tex

@@
 \begin{equation}
-  \pi^\text{explore}(s_t, z_t)\deq \text{argmax}_{a'} \left(Q(s, a') + z_t[a']\right)
+  \pi^\text{explore}(s_t, z_t)\deq \text{argmax}_{a'} \left(A(s, a') + z_t[a']\right)
 \end{equation}
 where $z_t[a']$ denotes the $a'$-th component of $z_t$. Namely, we
-perturb $Q$-values by a random process. The resulting exploration scheme
-converges in continuous time,
-and is non trivial for most $Q$'s.
+perturb advantage values by a random process before selecting an action. The resulting scheme
+converges in continuous time to a nontrivial exploration scheme.
 

### 2019-01-19 paper/icml_drau/advup.tex

@@
 
-Contrary to this discrete action exploration scheme, $\varepsilon$-greedy
-exploration is likely to not explore, i.e. to collapse to a deterministic
-policy when $\deltat$ goes to $0$. Indeed, one can prove that in a near
-continuous deterministic environment, the exploration policy resulting from
-applying $\varepsilon$-greedy exploration to a constant exploitation policy
-converges to a determinist policy as $\deltat$ goes to $0$. More precisely, let
-$(a_\deltat)$ be a sequence of i.i.d. actions of distribution drawn from $p_A$, (for instance
+On the other hand,
+$\varepsilon$-greedy
+exploration is likely \emph{not} to explore, i.e. to collapse to a deterministic
+policy, when $\deltat$ goes to $0$.
+% Indeed, in a
+% near-continuous deterministic environment, %the exploration policy resulting from
+% applying $\varepsilon$-greedy exploration to a given exploitation policy
+% converges to a determinist policy as $\deltat$ goes to $0$.
+Intuitively,
+with very small $\deltat$, changing the action at random every $\deltat$ time
+step just averages out the randomness due to the law of large numbers.
+More precisely, let
+$(a_\deltat)$ be a sequence of i.i.d.\ actions drawn from a distribution $p_A$, (for instance
 the sequence of actions generated by $\varepsilon$-greedy exploration on a
-constant policy). Then, as $\deltat$ goes to $0$, trajectories generated by
-interacting with the environment using the $a$'s converge to solutions of
+fixed policy) in a near-continuous environment. Then, as $\deltat$ goes
+to $0$, the trajectories generated by
+this policy tend to solutions of the deterministic equation
 \begin{equation}
-	ds_t = \E_{a\sim p_A}\left[F(s_t, a)\right]dt
+	ds_t/dt  = \E_{a\sim p_A}\left[F(s_t, a)\right]
 \end{equation}

### 2019-01-19 paper/icml_drau/advup.tex

@@
 
-To learn $V_{\theta}$ and $A_{\psi}$, variants of $Q$-learning for continuous and
-discrete action spaces are used. In details, $V_{\theta}$ and $A_{\psi}$ are learnt
-using approximate version of the following bellman equation, which holds for the true $A^\pi_\deltat$
-and $V^\pi_\deltat$, with a greedy exploitation policy $\pi(s) = \text{argmax}_{a'}A^\pi_\deltat(s, a')$
+$V_{\theta}$ and $A_{\psi}$ are learned by a variants of $Q$-learning for continuous and
+discrete action spaces are used. Namely, 
+the true $A^\pi_\deltat$
+and $V^\pi_\deltat$ of a near-continuous MDP with greedy exploitation
+policy $\pi(s) \deq
+\text{argmax}_{a'}A^\pi_\deltat(s, a')$
+satisfy the Bellman equation
 \begin{align}
@@
 \end{align}
+\TODO{cite Baird or argue this is a simple rewriting?} and $V_{\theta}$
+and $A_{\psi}$ are trained to approximately solve this equation.
+
 Maximization over actions is implemented exactly in the discrete actions case,

### 2019-01-19 paper/icml_drau/advup.tex

@@
 
-Maximization over actions is implemented exactly in the discrete actions case,
+Maximization over actions in $\pi$ is implemented exactly for discrete actions,
 and approximated by a policy neural network $\pi_\phi(s)$, trained to maximize $A(s,
-\pi_\phi(s))$ in the continuous case.  This is similar to what is done in
+\pi_\phi(s))$ for continuous actions. This is similar to
 \cite{ddpg}.
@@
 The update equations when learning from a transition $(s, a, r, s')$, either from
-an exploratory trajectory or from a replay buffer~\cite{dqn} are
+an exploratory trajectory or from a replay buffer~\cite{dqn}, are
 \begin{align}
@@
 \end{align} 
-where $\eta$'s are learning rates.
-Appropriate scalings for the learning rates $\eta^V_\deltat$ and $\eta^A_\deltat$ in term of $\deltat$ to obtain a well defined continuous limit are derived in the following section.
+where the $\eta$'s are learning rates.
+Appropriate scalings for the learning rates $\eta^V_\deltat$ and
+$\eta^A_\deltat$ in term of $\deltat$ to obtain a well defined continuous
+limit are derived next.
 % As we will see in the following section,
@@
 % of $\deltat$.
-\subsection{Learning rates scalings}
+
+\subsection{Scaling the Learning Rates}
 \label{subsec:lr}

### 2019-01-19 paper/icml_drau/advup.tex

@@
 %! TEX root = icml_drau.tex
-\section{Reinforcement learning with a continuous time limit}
+\section{Reinforcement Learning with a Continuous-Time Limit}
 
@@
 
-\subsection{Timestep-invariant exploration}
+\subsection{Timestep-Invariant Exploration}
 \label{subsec:explo}
@@
 
-\subsection{Algorithms}
+\subsection{Algorithms for Deep Advantage Updating}
 \label{subsec:algorithm}
@@
 \text{argmax}_{a'}A^\pi_\deltat(s, a')$
-satisfy the Bellman equation
+are the unique solution to the Bellman and identifiability equations
 \begin{align}
-	V^{\pi}_\deltat(s) + \deltat A^{\pi}_\deltat(s, a) &= r\deltat + \gamma^{\deltat}  \E_{s'} V^{\pi}_\deltat(s')\label{eq:bellman_A}\\
+	V^{\pi}_\deltat(s) + \deltat\, A^{\pi}_\deltat(s, a) &=
+	r\,\deltat + \gamma^{\deltat}  \,\E_{s'} V^{\pi}_\deltat(s')\label{eq:bellman_A}\\
 	A^{\pi}_\deltat(s, \pi(s)) &= 0\label{eq:max_A}.
 \end{align}
-\TODO{cite Baird or argue this is a simple rewriting?} and $V_{\theta}$
+as seen in \ref{subsec:reparam}. Then $V_{\theta}$
 and $A_{\psi}$ are trained to approximately solve this equation.
@@
 and approximated by a policy neural network $\pi_\phi(s)$, trained to maximize $A(s,
-\pi_\phi(s))$ for continuous actions. This is similar to
+\pi_\phi(s))$ for continuous actions. This point is similar to
 \cite{ddpg}.
@@
 \begin{align}
-	\delta Q_\deltat &\leftarrow A(s, a) \deltat - \left(r \deltat + \gamma^{\deltat} V(s') - V(s)\right)\\
-	\theta_\deltat &\leftarrow \theta_\deltat + \eta^V_\deltat \partial_{\theta} V(s) \frac{\delta Q_\deltat}{\deltat}\\
-	\psi_\deltat &\leftarrow \psi_\deltat + \eta^A_\deltat \partial_{\psi} A(s, a) \frac{\delta Q_\deltat}{\deltat}.
+	\delta Q_\deltat &\leftarrow A(s, a)\, \deltat - \left(r
+	\,\deltat + \gamma^{\deltat}\, V(s') - V(s)\right)\\
+	\theta_\deltat &\leftarrow \theta_\deltat + \eta^V_\deltat
+	\,\partial_{\theta} V(s) \,\frac{\delta Q_\deltat}{\deltat}\\
+	\psi_\deltat &\leftarrow \psi_\deltat + \eta^A_\deltat
+	\,\partial_{\psi} A(s, a) \,\frac{\delta Q_\deltat}{\deltat}.
 	\label{eq:approx_bellman_A}
@@
 % of $\deltat$.
+
+\TODO{say somewhere around here that we assume $\deltat$ is known}
 

### 2019-01-19 paper/icml_drau/advup.tex

@@
 \text{argmax}_{a'}A^\pi_\deltat(s, a')$
-are the unique solution to the Bellman and identifiability equations
+are the unique solution to the Bellman and consistency equations
 \begin{align}
@@
 as seen in \ref{subsec:reparam}. Then $V_{\theta}$
-and $A_{\psi}$ are trained to approximately solve this equation.
+and $A_{\psi}$ are trained to approximately solve these equations.
 

### 2019-01-19 paper/icml_drau/advup.tex

@@
 % \begin{align}
-% 	A^\pi_\deltat(s, a)=&\reward(s, a)+\frac{\gamma^{\deltat}\E_{\delta s \sim \delta s(s, a)}V^\pi_\deltat(s + \delta s)-V^\pi_\deltat(s)}{\deltat}\nonumber\\
+% 	A^\pi_\deltat(s, a)=&\reward(s, a)+\frac{\gamma^{\deltat}\E_{\deltas \sim \deltas(s, a)}V^\pi_\deltat(s + \deltas)-V^\pi_\deltat(s)}{\deltat}\nonumber\\
 % 	=&\reward(s, a)+\log(\gamma) V^\pi(s)+\partial_s V^\pi(s)f(s, a)\nonumber\\
@@
 of parameters must converge to well defined trajectories as $\deltat$ goes to
-$0$.  Intuitively, this means that optimization steps should not be too large,
-in which case parameter trajectories could diverge in a single gradient step,
-or too small, in which case the trajectories would converge to single points as
+$0$.  The optimization steps should not be too large,
+in which case parameter trajectories could jump in a single gradient step,
+or too small, in which case the trajectories would become stationary as
 $\deltat$ goes to $0$.  This in turn imposes conditions on the scalings of the
@@
 
-The updates equations on $\theta$ and $\psi$ are described in
-Eq.~\eqref{eq:approx_bellman_A}.  To have parameter steps that do not explode or shrink to $0$ too
+The updates of $\theta$ and $\psi$ are described in
+\eqref{eq:approx_bellman_A}.  To have parameter updates that do not explode or shrink to $0$ too
 fast as $\deltat \rightarrow 0$, both $\delta \theta$ and
-$\delta \psi$ should verify stochastic differential equations, and notably
+$\delta \psi$ should verify stochastic differential equations in the
+limit, and notably,
 have a deterministic component of order $\deltat$. As $\deltat$ goes to $0$,
-updates are rephrased as
+a Taylor expansion yields (with $\deltas\deq s'-s$)
 \begin{align}
-	\delta Q_\deltat &= A(s, a)\deltat - r\deltat - \log(\gamma) V(s)\deltat - \partial_s V(s) \delta s\nonumber \\&- \frac{(\delta s)^T \partial_{s^2}^2 V(s) \delta s}{2} + \smallo\left(\deltat\right)\\
-	\delta \theta_\deltat &= \eta^V_\deltat \partial_{\theta} V(s) \frac{\delta Q_\deltat}{\deltat} + \smallo\left(\eta^V_\deltat\right)\\
-	\delta \psi_\deltat &= \eta^A_\deltat \partial_{\psi} A(s) \frac{\delta Q_\deltat}{\deltat} + \smallo\left(\eta^A_\deltat\right).
+	\delta Q_\deltat &= A(s, a)\,\deltat - r\,\deltat - \log(\gamma)
+	\,V(s)\,\deltat - \partial_s V(s) \,\deltas\nonumber \\&- \frac{(\deltas)^T \partial_{s}^2 V(s) \deltas}{2} + \smallo\left(\deltat\right)\\
+	\delta \theta_\deltat &= \eta^V_\deltat \partial_{\theta} V(s)
+	\frac{\delta Q_\deltat}{\deltat} \\%+ \smallo\left(\eta^V_\deltat\right)\\
+	\delta \psi_\deltat &= \eta^A_\deltat \partial_{\psi} A(s)
+	\frac{\delta Q_\deltat}{\deltat} %+ \smallo\left(\eta^A_\deltat\right).
 \end{align}
 
-For parameter updates to follow a SDE, one must
-have $\eta^V_\deltat = \alpha^V \deltat$ and $\eta^A_\deltat = \alpha^A
-\deltat$. More precisely
+If we set 
+$\eta^V_\deltat = \alpha^V \deltat$ and $\eta^A_\deltat = \alpha^A
+\deltat$, the parameter updates will follow a Euler discretization scheme
+for a continuous-time equation, thus presumably converging. This is
+actually the only possibility; more precisely,
+
 \begin{theorem}

### 2019-01-19 paper/icml_drau/experiments.tex

@@
 \input{lc_fig.tex}
-The experiment set that we provide here is specifically aimed at showing that
+The experiment set provided here is specifically aimed at showing that
 the proposed method DAU can operate efficiently on a wide range of time
-discretization, without specific tuning, while standard deep Q-learning
-approaches can't. DAU is compared to DDPG in continuous action settings and to DQN in
-discrete action settings.
+discretizations, without specific tuning, while standard deep Q-learning
+approaches cannot. DAU is compared to DDPG in continuous action settings and to DQN in
+discrete action settings. \TODO{(We only test off-policy
+algorithms.)}
 

### 2019-01-19 paper/icml_drau/experiments.tex

@@
 
-In all experimental settings, we use the learning algorithm described in
-alg.~\ref{alg:dau} and \TODO{supplementary, algo 1}. The variant of DDPG and DQN
+In all settings, we use the learning algorithm described in
+Alg.~\ref{alg:dau} and \TODO{supplementary, algo 1}. The variants of DDPG and DQN
 used are described in the supplementary material, as well as all the hyperparameters
-used. To make comparison fair, we still use rescaled discount factor $\gamma_\deltat$ and reward $r_\deltat$  as defined in equations \TODO{ref}, as well as scaled learning rates for DDPG and DQN.
+used. To make comparison fair, we use the properly rescaled discount
+factor $\gamma_\deltat$ and reward $r_\deltat$ from \eqref{eq:def-gamma}, as well as scaled learning rates, for DDPG and DQN.
 

### 2019-01-19 paper/icml_drau/experiments.tex

@@
 used. To make comparison fair, we use the properly rescaled discount
-factor $\gamma_\deltat$ and reward $r_\deltat$ from \eqref{eq:def-gamma}, as well as scaled learning rates, for DDPG and DQN.
+factor $\gamma_\deltat$ and reward $r_\deltat$ from \eqref{eq:def-gamma},
+as well as scaled learning rates, for DDPG and DQN. \TODO{NDY not
+convincing: precisely because DDPG and DQN are not invariant, they might
+benefit from a different scaling.}
 
-Let us stress that quantities shown are rescaled to make comparison possible. For example,
-return results are given in term of discretized return $R_\deltat$ as defined in Eq. \eqref{eq:discretized-return},\footnote{This mostly amounts to scaling rewards
-by a factor $\deltat$ when this scaling is not naturally done in the environment. Environment specific
+Let us stress that the quantities plotted are rescaled to make comparison
+possible across different timesteps. For example,
+return results are given in term of discretized return $R_\deltat$ as defined in \eqref{eq:discretized-return},\footnote{This mostly amounts to scaling rewards
+by a factor $\deltat$ when this scaling is not naturally done in the environment. Environment-specific
 details are given in the supplementary material.} and, most notably, time elapsed is always given in
-normalized epochs, i.e. the true number of epochs multiplied by the discretization timestep.
+normalized epochs (physical time), i.e. the true number of epochs multiplied by the
+discretization timestep.
 

### 2019-01-19 paper/icml_drau/experiments.tex

@@
 
-\paragraph{Qualitative experiments: Visualizing policies and values}
-To provide qualitative results, and check that robustness to time
-discretization is provided both in term of raw return results, but also in term
-of convergence of approximate value function and policies, we first provide results on the simple pendulum environment
-from the OpenAI Gym classic control suite.
-This environment's state space is of dimension $2$.
-A visualization of both the learnt value and policy functions is obtained by plotting for each point of the phase diagram $(\theta, \dot{\theta})$ its value and policy.
-results are presented in
+\paragraph{Qualitative experiments: Visualizing policies and values.}
+To provide qualitative results, and check robustness to time
+discretization both in terms of raw return results and in terms
+of convergence of the approximate value function and policies, we first provide results on the simple pendulum environment
+from the OpenAI Gym classic control suite.  Its state space is of
+dimension $2$. We visualize both the learnt value and policy functions by
+plotting, for each point of the phase diagram $(\theta, \dot{\theta})$,
+its value and policy. The results are presented in
 Fig.~\ref{fig:pend} for the policy function and in Fig.\TODO{ref} in Supplementary \TODO{ref} for the value function.

### 2019-01-19 paper/icml_drau/experiments.tex

@@
 its value and policy. The results are presented in
-Fig.~\ref{fig:pend} for the policy function and in Fig.\TODO{ref} in Supplementary \TODO{ref} for the value function.
+Fig.~\ref{fig:pend} (policy function) and in Fig.\TODO{ref} in
+Supplementary \TODO{ref} (value function).
 
-We observe the learnt policy at several instants in physical time (or normalized epochs) during training, for various time discretizations $\deltat$, for both DAU and DDPG.
-With DAU, the agent's policy quickly converges for every time discretization without specific hyperparameter tuning.
-On the contrary, with DDPG, the agent is only able to learn for the two highest time discretization, and learning gets harder as $\deltat$ decreases.
+We plot the learnt policy at several instants in physical time (or
+normalized epochs) during training, for various time discretizations
+$\deltat$, for both DAU and DDPG. With DAU, the agent's policy quickly
+converges for every time discretization without specific hyperparameter
+tuning. On the contrary, with DDPG, the agent is only able to learn for
+the two largest time discretizations; DDPG learning gets harder as $\deltat$ decreases.
 

### 2019-01-19 paper/icml_drau/experiments.tex

@@
 
-\paragraph{Quantitative experiments}
-We benchmark DAU against DDPG on classic control benchmarks: Pendulum, Cartpole, BipedalWalker, Ant and Half-Cheetah environments from OpenAI Gym.
-For Pendulum and Ant, DDPG is able to learn for the highest $\deltat$ but fails for the others, while DAU remains reasonably invariant to the time discretization.
-On Bipedal Walker, DDPG is never able to learn while DAU is achieving decent performance for all the $\deltat$'s.
-On Cartpole and Half Cheetah, results are less clear.
-On Cartpole, noise dominates, which makes interpretation difficult.
-On Half Cheetah, DAU is not clearly more invariant to time discretization than DDPG.
-This could be explained by the multiple suboptimal regimes that coexist in the half cheetah environment (walking on the head, walking on the back), and that DAU may fail to correctly weight the performance of each regime (as explained in Section \ref{sec:discussions}).
+\paragraph{Quantitative experiments.}
+We benchmark DAU against DDPG on classic control benchmarks: Pendulum,
+Cartpole, BipedalWalker, Ant, and Half-Cheetah environments from OpenAI
+Gym. For Pendulum and Ant, DDPG is able to learn for the highest
+$\deltat$ but fails for the others, while DAU remains reasonably
+invariant to the time discretization. On BipedalWalker, DDPG is never
+able to learn while DAU is achieving decent performance for all the
+$\deltat$'s. On Cartpole and Half-Cheetah, the results are less clear. On
+Cartpole, noise dominates, which makes interpretation difficult. On Half
+Cheetah, DAU is not clearly more invariant to time discretization than
+DDPG. This could be explained by the multiple suboptimal regimes that
+coexist in the half cheetah environment (walking on the head, walking on
+the back), and that DAU may fail to correctly weight the performance of
+each regime (as explained in Section \ref{sec:discussions}).
 

### 2019-01-19 paper/icml_drau/experiments.tex

@@
 $\deltat$'s. On Cartpole and Half-Cheetah, the results are less clear. On
-Cartpole, noise dominates, which makes interpretation difficult. On Half
-Cheetah, DAU is not clearly more invariant to time discretization than
+Cartpole, noise dominates, which makes interpretation difficult. On Half-Cheetah, DAU is not clearly more invariant to time discretization than
 DDPG. This could be explained by the multiple suboptimal regimes that
-coexist in the half cheetah environment (walking on the head, walking on
-the back), and that DAU may fail to correctly weight the performance of
-each regime (as explained in Section \ref{sec:discussions}).
+coexist in the Half-Cheetah environment (walking on the head, walking on
+the back). \TODO{Removed, was unclear: ``and that DAU may fail to correctly weight the performance of
+each regime (as explained in Section \ref{sec:discussions}).''}
 

### 2019-01-19 paper/icml_drau/experiments.tex

@@
 the back). \TODO{Removed, was unclear: ``and that DAU may fail to correctly weight the performance of
-each regime (as explained in Section \ref{sec:discussions}).''}
+each regime (as explained in Section \ref{sec:discussions}).'' Anyway
+this regimes exist for all algorithms...}
 

### 2019-01-19 paper/icml_drau/discussions.tex

@@
 %! TEX root = icml_drau.tex
-\section{Discussions}
+\section{Discussion}
 \label{sec:discussions}
-The method derived in this work seems yield improved robustness to time discretization on 
-various environments, e.g. pendulum swing up, or simple locomotion tasks.
-However on certain environments, there is still much room for improvment. In what follows,
-we identify some of the issues that could explain this theoretical/practical discrepancy.
+
+The method derived in this work seems to yield improved robustness to
+time discretization on various environments, e.g. pendulum swing up, or
+simple locomotion tasks.  However on certain environments, there is still
+room for improvment. In what follows, we identify some of the issues
+that could explain this theoretical/practical discrepancy.
 

### 2019-01-19 paper/icml_drau/discussions.tex

@@
 
-\paragraph{Regularity of the value function}~In all of our proofs, $V^\pi$ is
-assumed to be continuously differentiable for all $\pi$.  This is a strong
-hypothesis that is not always verified in practice. Very simple environments
-break this hypothesis, such as for instance, the pendulum swing up environment,
-with pendulum height as reward and torque as action.  Indeed, assume that $\pi$
-is the optimal policy for the pendulum swing up environment, with state
-representation $(\theta, \dot{\theta})$. The state space for this environment
+\paragraph{Smoothness of the value function.} In all of our proofs, $V^\pi$ is
+assumed to be continuously differentiable for all $\pi$.  This %is a strong
+hypothesis is not always satisfied in practice. Very simple environments
+break this hypothesis, such as the pendulum swing-up environment,
+with pendulum height as reward and torque as action.  Indeed, 
+% assume that $\pi$
+% is the optimal policy for the pendulum swing-up environment, with state
+% representation $(\theta, \dot{\theta})$. 
+the state space $(\theta, \dot{\theta})$ for this environment
 can be partitioned into two subspaces $\mathcal{G}$ and $\mathcal{B}$, with
-$\mathcal{G}$ the subspace where when following the optimal policy, the
-pendulum can be swang up without crossing the $y$-axis, and $\mathcal{B}$ where
-it have to cross the $y$-axis. The difference of V-function between a point in
+$\mathcal{G}$ the subspace where the optimal policy 
+can swing up the pendulum without crossing the $y$-axis, and $\mathcal{B}$ where
+it has to cross the $y$-axis. The difference of $V$-function between a point in
 $\mathcal{G}$ and a point in $\mathcal{B}$ is lower bounded by a finite
-quantity $M$. Besides, one can find two points $s_1 \in \mathcal{G}$ and $s_2 \in \mathcal{B}$
-such that $s_1$ and $s_2$ are arbitrarily close. Consequently, $V^\pi$ is
-discontinuous on the frontier, which results in non infinitesimal advantages
+quantity. 
+%Besides, one can find two points $s_1 \in \mathcal{G}$ and $s_2 \in \mathcal{B}$
+%such that $s_1$ and $s_2$ are arbitrarily close.
+Consequently, $V^\pi$ is
+discontinuous on the frontier between $\mathcal{G}$ and $\mathcal{B}$, which results in non infinitesimal advantages
 for actions taken on the frontier. \TODO{Non infinitesimal advantages are found,

### 2019-01-19 paper/icml_drau/discussions.tex

@@
 %such that $s_1$ and $s_2$ are arbitrarily close.
-Consequently, $V^\pi$ is
+Consequently, the value function is
 discontinuous on the frontier between $\mathcal{G}$ and $\mathcal{B}$, which results in non infinitesimal advantages
 for actions taken on the frontier. \TODO{Non infinitesimal advantages are found,
-	more generally, as soon as a given policy has different regimes, depending on
-	where it starts in state space. For such policies, at the frontier between two regimes,
-the value function can be discontinuous, resulting in non infinitesimal advantages.}
+more generally, as soon as a given policy has different regimes depending on
+where it starts in state space. For such policies the value
+function is typically discontinuous at the frontier between two
+regimes, resulting in non infinitesimal advantages. The
+continuous-time DAU properties would only hold almost-everywhere.}
 
@@
 transitions are sampled from a memory buffer, to prevent excessive correlations.
-The memory buffer used is a fixed size circular buffer, and is
-filled with samples coming from a single infinite exploratory trajectory. In
+The memory buffer used is a fixed size circular buffer, filled
+with samples from a single infinite exploratory trajectory. In
 our experiments, the same buffer size is used for all time discretization. When
@@
 infinite sequence, since the buffer will only contain examples that are very
-close temporally to the end of the sequence in physical time. While the resulting
-algorithm \TODO{kind of} converges to a continuous time limit, it is still not
+close temporally to the end of the sequence in physical time. \TODO{this
+contradicts the use of the buffer against correlations} While the resulting
+algorithm converges to a continuous time limit, in the strictest sense it is not
 time discretization invariant, as the freshness of samples in the buffer will
@@
 $\frac{1}{\deltat}$. This would amount to always learning on the same physical time interval of
-the trajectory. Our analysis, however, does not cover this case.
+the trajectory. Our theoretical analysis does not cover this case.
 

### 2019-01-19 paper/icml_drau/discussions.tex

@@
 the trajectory. Our theoretical analysis does not cover this case.
+\TODO{a bit too many details here?}
 
-\paragraph{Near continuous time reinforcement learning and RMSprop}~RMSprop~\cite{rmsprop}
-estimates a batched version of the second momentum of gradients using moving
-averages, and divides gradient steps by the square root of this estimate. In
+\paragraph{Near continuous time reinforcement learning and RMSprop.} RMSprop~\cite{rmsprop}
+estimates a batched version of the second moment of gradients via moving
+averages, and divides gradient steps by the square root of this estimate.
+This may interact with the learning rate scaling discussed above. In
 deterministic environments, gradient noise typically scales as $\bigO(1)$ in
-term of $\deltat$, as seen in Eq.~\eqref{eq:approx_bellman_A}.  Consequently, RMSprop
-preconditioning have no effect on the order of magnitude of required learning
-rates. However, in near continuous stochastic environments, as in
+term of $\deltat$, as seen in \eqref{eq:approx_bellman_A}.  In that case, RMSprop
+preconditioning has no effect on the suitable order of magnitude for learning
+rates. However, in near continuous \emph{stochastic} environments, as in
 Eq.~\ref{eq:sde}, variance of the gradient noise typically scales as

### 2019-01-20 pendulum_exp/critics/value.py
Adding noscale argument

@@ -13,21 +13,29 @@ from nn import soft_update
 class ValueCritic(CompoundStateful, Critic):
     def __init__(self,
                  dt: float, gamma: float, lr: float, optimizer: str,
-                 q_function: ParametricFunction, tau: float) -> None:
+                 q_function: ParametricFunction, tau: float, noscale: bool) -> None:
         CompoundStateful.__init__(self)
         self._reference_obs: Tensor = None
         self._q_function = q_function
         self._target_q_function = copy.deepcopy(self._q_function)
         self._tau = tau
 
+        ref_dt = 0.02
+        if noscale:
+            self._gamma = gamma ** (dt / ref_dt)
+        else:
+            self._gamma = gamma
+
+        if noscale:
+            self._dt = ref_dt
+        else:
+            self._dt = dt
+
         self._q_optimizer = \
             setup_optimizer(self._q_function.parameters(),
                             opt_name=optimizer, lr=lr, dt=dt,
                             inverse_gradient_magnitude=dt,
                             weight_decay=0)
-
-        self._dt = dt
-        self._gamma = gamma
 
         self._device = 'cpu'
 
@@ -89,7 +97,7 @@ class ValueCritic(CompoundStateful, Critic):
     def configure(dt: float, gamma: float, lr: float, optimizer: str,
                   action_space: Space, observation_space: Space,
                   nb_layers: int, hidden_size: int, normalize: bool,
-                  tau: float, **kwargs):
+                  tau: float, noscale: bool, **kwargs):
         assert isinstance(observation_space, Box)
         nb_state_feats = observation_space.shape[-1]
         net_dict = dict(nb_layers=nb_layers, hidden_size=hidden_size)
@@ -104,4 +112,4 @@ class ValueCritic(CompoundStateful, Critic):
                 **net_dict)
         if normalize:
             q_function = NormalizedMLP(q_function)
-        return ValueCritic(dt, gamma, lr, optimizer, q_function, tau)
+        return ValueCritic(dt, gamma, lr, optimizer, q_function, tau, noscale)

### 2019-01-20 pendulum_exp/noise.py

@@ -7,8 +7,9 @@ from convert import th_to_arr, check_tensor
 
 def setup_noise(
         noise_type: str, sigma: float, theta: float, dt: float,
-        sigma_decay: DecayFunction, **kwargs) -> Noise:
-    keywords_args = dict(sigma=sigma, theta=theta, dt=dt, sigma_decay=sigma_decay)
+        sigma_decay: DecayFunction, noscale: bool, **kwargs) -> Noise:
+    keywords_args = dict(sigma=sigma, theta=theta, dt=dt, noscale=noscale,
+                         sigma_decay=sigma_decay)
 
     if noise_type == 'parameter':
         return ParameterNoise(**keywords_args) # type: ignore
@@ -23,12 +24,17 @@ class ParameterNoise(Noise):
                  theta: float,
                  sigma: float,
                  dt: float,
+                 noscale: bool,
                  sigma_decay: Optional[Callable[[int], float]] = None) -> None:
         self._p_noise: Dict[str, Any] = {}
         self._theta = theta
         self._sigma = sigma
         self._sigma_decay = sigma_decay
-        self._dt = dt
+        ref_dt = .02
+        if noscale:
+            self._dt = ref_dt
+        else:
+            self._dt = dt
         self._count = 0
         self._device = torch.device('cpu')
 
@@ -78,12 +84,17 @@ class ActionNoise(Noise): # pylint: disable=too-few-public-methods
                  theta: float,
                  sigma: float,
                  dt: float,
+                 noscale: bool,
                  sigma_decay: Optional[Callable[[int], float]] = None) -> None:
         self.noise = torch.empty(())
         self._theta = theta
         self._sigma = sigma
         self._sigma_decay = sigma_decay
-        self._dt = dt
+        ref_dt = .02
+        if noscale:
+            self._dt = ref_dt
+        else:
+            self._dt = dt
         self._count = 0
         self._device = torch.device('cpu')
 

### 2019-01-20 paper/icml_drau/advup.tex

@@
 		\item If $\beta < 1$, parameters grow arbitrarily large in a single timestep as $\deltat$
-			goes to $0$.
-		\item If $\beta > 1$, parameter trajectories shrink to a single point as 
-			$\deltat$ goes to $0$. \TODO{Is that clear}
+			goes to $0$. \TODO{not quite true, in a single
+			step the update is bounded. But the parameters
+			become arbitrary large in an arbitrarily small
+			physical time.}
+		\item If $\beta > 1$, parameter trajectories become
+		stationary as
+			$\deltat$ goes to $0$.
 	\end{itemize}
@@
 % \end{align}
-The full algorithm, that is referred to as \emph{Deep Advantage Updating} (DAU), can now be derived.
+
+This results in a fullly specified algorithm,
+\emph{Deep Advantage Updating} (DAU).
 Pseudocode for the discrete actions algorithm is given in Alg.~\ref{alg:dau}.
-Pseudocode for the continuous actions algorithm is given in the supplementary
+Pseudocode for continuous actions is given in the supplementary
 material.

### 2019-01-20 paper/icml_drau/discussions.tex

@@
 This may interact with the learning rate scaling discussed above. In
-deterministic environments, gradient noise typically scales as $\bigO(1)$ in
+deterministic environments, gradients typically scale as $\bigO(1)$ in
 term of $\deltat$, as seen in \eqref{eq:approx_bellman_A}.  In that case, RMSprop
 preconditioning has no effect on the suitable order of magnitude for learning
-rates. However, in near continuous \emph{stochastic} environments, as in
-Eq.~\ref{eq:sde}, variance of the gradient noise typically scales as
-$\bigO\left(\frac{1}{\deltat}\right)$. If batch size does not depend on
-$\deltat$, RMSprop preconditioning thus multiplies gradients by a factor $\bigO(\sqrt{\deltat})$,
-and learning rates need only be scaled as $\alpha \sqrt{\deltat}$ instead of
-$\alpha \deltat$.
+rates. However, in near continuous \emph{stochastic} environments (Eq.~\ref{eq:sde}), variance of the gradients typically scales as
+$\bigO\left(\frac{1}{\deltat}\right)$. With a fixed batch size,
+RMSprop will multiply gradients by a factor $\bigO(\sqrt{\deltat})$. In
+that case,
+learning rates need only be scaled as $\alpha \,\sqrt{\deltat}$ instead of
+$\alpha \,\deltat$.

### 2019-01-20 paper/icml_drau/supplementary.tex
.

\documentclass[11pt]{article}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}

\newtheorem{theorem}{Theorem}
\input{macros.tex}

\begin{document}
\section{Proofs}
We hereby give proofs for all the results presented in the paper.
The first result presented is a proof of convergence for discretized
trajectories.
\begin{theorem}
	Let $F: \cal{S} \times \cal{A} \rightarrow \bb{R}^n$ and
	$\pi: \cal{S} \rightarrow \cal{A}$ be the dynamic and policy
	functions. Assume that $s \rightarrow F(s, \pi(s))$ is 
	$\cal{C}^1$ and $K$-lipschitz. By the 
	a

	\label{th:traj-conv}
\end{theorem}
\begin{proof}
	\input{proof_traj.tex}
\end{proof}
\begin{theorem}
	Under suitable assumptions \TODO{ref appendix}, for all $s \in {\cal
	S}$, one has
	%\begin{equation}
	%\label{eq:conv-value}
	$V^\pi_\deltat(s) = V^\pi(s) + \bigO(\deltat)$
	%\end{equation}
	when $\deltat\to 0$.
	\label{th:conv-value}
\end{theorem}
\begin{proof}
	\input{proof_conv_value.tex}
\end{proof}
\begin{theorem}
	$A^\pi_\deltat$ convergence theorem.
\end{theorem}
\begin{proof}
	\input{proof_conv_adv.tex}
\end{proof}
\begin{theorem}
	Discrete parameter trajectories converge to continuous trajectories.
\end{theorem}
\begin{proof}
	\input{proof_conv_params.tex}
\end{proof}
\section{Implementation details}
\end{document}

### 2019-01-21 pendulum_exp/dataanalysis/script_leonard.py
.

@@ -7,7 +7,7 @@ from dataloader import loader_leonard, ExperimentData
 from plots import plot_learning_curves
 
 # 
-start_date = datetime.strptime('2019_01_14_10_22_54', "%Y_%m_%d_%H_%M_%S")
+start_date = datetime.strptime('2019_01_20_06_09_29', "%Y_%m_%d_%H_%M_%S")
 # stop_date = datetime.strptime('2019_01_14_10_22_55', "%Y_%m_%d_%H_%M_%S")
 
 # start_date = 'last'
@@ -27,6 +27,10 @@ plot_learning_curves(expdata_ant, ['Return', 'Return'], 'ant', gtype='run_std', 
 
 expdata_cheetah = expdata.filter_settings(lambda s: s['env_id'] == 'half_cheetah')
 plot_learning_curves(expdata_cheetah, ['Return', 'Return'], 'cheetah', gtype='run_std', mint=0, maxt=20)
+
+expdata_bipedal = expdata.filter_settings(lambda s: s['env_id'] == 'bipedal_walker')
+plot_learning_curves(expdata_cheetah, ['Return'], 'bipedal_walker', gtype='run_std', mint=0, maxt=20)
+
 
 
 # expdata_tau = expdata.filter_settings(lambda s: s['algo'] == 'discrete_value')

### 2019-01-21 paper/icml_drau/supplementary.tex

@@
 \usepackage{amsthm}
+\usepackage{mathtools}
+\usepackage{xcolor}
 
@@
 \begin{theorem}
-	Let $F: \cal{S} \times \cal{A} \rightarrow \bb{R}^n$ and
-	$\pi: \cal{S} \rightarrow \cal{A}$ be the dynamic and policy
-	functions. Assume that $s \rightarrow F(s, \pi(s))$ is 
-	$\cal{C}^1$ and $K$-lipschitz. By the 
-	a
-
+	Let $F: {\cal S} \times {\cal A} \rightarrow \bb{R}^n$ and $\pi: {\cal S}
+	\rightarrow {\cal A}$ be the dynamic and policy functions. Assume that,
+	for any $a$, $s \rightarrow F(s, a)$ and $s \rightarrow F(s, \pi(s))$
+	are ${\cal C}^1$, bounded and $K$-lipschitz.  For
+	a given $s_0$, define the trajectory $(s_t)_{t\geq 0}$ as the unique
+	solution of the differential equation
+	\begin{equation}
+		\frac{ds_t}{dt} = F(s_t, \pi(s_t)).
+		\label{eq:diff}
+	\end{equation}
+	For any $\deltat > 0$, define the discretized trajectory
+	$(s_\deltat^k)_k$ recurrently as $s_\deltat^0 = s_0$,
+	$s_\deltat^{k + 1}$ is the value at time $\deltat$ of
+	the unique solution of
+	\begin{equation}
+		\frac{d\tilde{s}_t}{dt} = F(\tilde{s}_t, \pi(s_\deltat^k))
+	\end{equation}
+	with initial point $s_\deltat^k$.
+	Then, for every $t \geq 0$
+	\begin{equation}
+		\|s_t - s_\deltat^{\lfloor \frac{t}{\deltat} \rfloor}\|
+		\leq 
+	\end{equation}
 	\label{th:traj-conv}

### 2019-01-21 paper/icml_drau/proof_traj.tex

%! TEX root = icml_drau.tex
For any $k$, define
\begin{equation}
	e_\deltat^k = \|s_\deltat^k - s_{\deltat k}\|.
\end{equation}
Taylor integral formula gives
\begin{align}
	s_\deltat^{k + 1} &= s_\deltat^k + F(s_\deltat^k, \pi(s_\deltat^k)) \deltat +
	\frac{1}{2}\int\limits_{0}^\deltat (\deltat - t) \frac{d^2 \tilde{s}_t}{dt^2} dt\\
	s_{\deltat(k + 1)} &= s_{\deltat k} + F(s_{\deltat k}, \pi(s_{\deltat k})) \deltat +
	\frac{1}{2}\int\limits_0^\deltat (\deltat - t) \frac{d^2 s_{t + \deltat k}}{dt^2} dt.
\end{align}
Now, both $d^2 s_t/dt^2$ and $d^2 \tilde{s}_t/dt^2$ are uniformly bounded, by
boundedness and lipschitzianity of $s \rightarrow F(s, \pi(s))$ and $s
\rightarrow F(s, \pi(s_\deltat^k))$. Consequently, there exists $C$ such that
\begin{align}
	e_\deltat^{k + 1} &\leq \|s_\deltat^k - s_{\deltat k}\| + \|F(s_\deltat^k, \pi(s_\deltat^k)) - F(s_{\deltat k}, \pi(s_{\deltat k}))\| \deltat + C \deltat^2\\
			  &\leq (1 + K \deltat) e_\deltat^k + C \deltat^2.
\end{align}
Now, it is easy to prove by recurrence that
\begin{equation}
	e_\deltat^k \leq (1 + K \deltat)^k (e_\deltat^0 + \frac{C}{K} \deltat) - \frac{C}{K}\deltat.
\end{equation}
As $e_\deltat^0 = 0$, this translates to
\begin{align}
	e_\deltat^k &\leq ((1 + K \deltat)^k - 1) \deltat \frac{C}{K}\\
		    &\leq (e^{K \deltat k} - 1) \deltat \frac{C}{K}.
\end{align}
Consequently,
\begin{equation}
	e_\deltat^\lfloor t / \deltat \rfloor \leq (e^{K(t + \deltat)} - 1) \deltat \frac{C}{K}
\end{equation}

### 2019-01-21 paper/icml_drau/supplementary.tex
.

@@
 	with initial point $s_\deltat^k$.
-	Then, for every $t \geq 0$
+	Then, there exists $C > 0$ such that, for every $t \geq 0$
 	\begin{equation}
 		\|s_t - s_\deltat^{\lfloor \frac{t}{\deltat} \rfloor}\|
-		\leq 
+		\leq \deltat \frac{C}{K}e^{Kt}.
 	\end{equation}
+	Notably, discretized trajectories converge pointwise to continuous trajectories.
 	\label{th:traj-conv}
@@
 \end{proof}
+In what follows, we will assume that the reward function $r: {\cal S} \times {\cal A} \rightarrow \bb{R}$
+is bounded, to ensure existence of $V^\pi$ and $V^\pi_\deltat$ for all $\deltat$. We further assume that
+both $V^\pi$ and $V^\pi_\deltat$ are continuously differentiable, and that 
 \begin{theorem}

### 2019-01-21 paper/icml_drau/supplementary.tex
.

--- 2019-01-21 paper/icml_drau/supplementary.tex
@@
 \end{proof}
-\begin{theorem}
+In what follows, we will assume that the reward function $r: {\cal S} \times {\cal A} \rightarrow \bb{R}$
+is bounded, to ensure existence of $V^\pi$ and $V^\pi_\deltat$ for all $\deltat$.\begin{theorem}
 	Under suitable assumptions \TODO{ref appendix}, for all $s \in {\cal
@@
 \end{proof}
+
+For the following proof, we further assume that both $V^\pi$ and
+$V^\pi_\deltat$ are continuously differentiable, and that the gradient and
+hessian of $V^\pi_\deltat$ w.r.t. $s$ are uniformly bounded in both $s$ and $\deltat$.
+We also assume convergence of $\partial_s V^\pi_\deltat(s)$ to $\partial_s V^\pi(s)$ for
+all $s$.
 \begin{theorem}
-	$A^\pi_\deltat$ convergence theorem.
+	Under the above hypothesis, there exists $A^\pi: {\cal S} \rightarrow
+	\bb{R}$ such that $A^\pi_\deltat$ converges pointwise to $A^\pi$ as
+	$\deltat$ goes to $0$.
 \end{theorem}
@@
 	\input{proof_conv_adv.tex}
+\end{proof}
+
+We now show that policy improvement works with the continous time advantage function, i.e.\
+\begin{theorem}
+	Let $\pi$ and $\pi'$ be two policies such that both $s \rightarrow r(s, \pi(s))$ and
+	$s \rightarrow r(s, \pi'(s))$ are continuous.
+	Assume that both $V^\pi$ and $V^{\pi'}$ are continuously differentiable.
+	Define the advantage function for policies $\pi$ and $\pi'$ as in Eq.~\eqref{eq:adv_function}.
+
+	If for all $s$, $A^\pi(s, \pi'(s)) \geq 0$, then for all $s$, $V^\pi(s) \leq V^{\pi'}(s)$.
+\end{theorem}
+\begin{proof}
+	\input{policy_improvement.tex}
 \end{proof}

### 2019-01-21 paper/icml_drau/cdau.tex
.

--- 2019-01-19 paper/icml_drau/dauc.tex
@@
-%! TEX root = icml_drau.tex
+%! TEX root = supplementary.tex
+
 \begin{algorithmic}
 	\STATE \textbf{Inputs:}
-	\STATE $\theta$ and $\psi$, parameters of
-	$V_{\theta}$ and $\bar{A}_{\psi}$.
+	\STATE $\theta$, $\psi$ and $\phi$, parameters of
+	$V_{\theta}$, $\bar{A}_{\psi}$ and $\pi_\phi$.
 	\STATE $\pi^{\text{explore}}$ and $\nu_\deltat$ defining an exploration policy.
-	\STATE \textbf{opt}$_1$, \textbf{opt}$_2$, $\alpha^V \deltat$ and $\alpha^V \deltat$ optimizers and learning rates.
+	\STATE \textbf{opt}$_V$, \textbf{opt}$_A$, \textbf{opt}$_\pi$, $\alpha^V \deltat$, $\alpha^A \deltat$ and $\alpha^\pi\deltat$, optimizers and learning rates.
 	\STATE $\mathcal{D}$, buffer of transitions $(s, a, r, d, s')$, with $d$ the episode termination signal.
@@
 	\STATE $Q^i \gets V_{\theta}(s^i) + \deltat\hspace{-.17em}\left(
-	\bar{A}_{\psi}(s^i, a^i) - \max\limits_{a'}\bar{A}_{\psi}(s^i, a')\right)$
+	\bar{A}_{\psi}(s^i, a^i) - \bar{A}_{\psi}(s^i, \pi_\phi(s^i))\right)$
 	\STATE $\tilde{Q^i} \gets r^i\deltat + (1 - d^i) \gamma^{\deltat} V_{\theta}(s'^i)$
 	\STATE $\Delta \theta \gets \frac{1}{N}\sum\limits_{i=1}^N  \frac{\left(Q^i - \tilde{Q^i}\right)\partial_{\theta} V_{\theta}(s^i)}{\deltat}$
-	\STATE $\Delta \psi \gets \frac{1}{N}\sum\limits_{i=1}^N \frac{\left(Q^i - \tilde{Q^i}\right)\partial_{\psi} \left(\bar{A}_{\psi}(s^i, a^i) - \max\limits_{a'}\bar{A}_{\psi}(s^i, a')\right) }{\deltat}$
-	\STATE Update $\theta$ with \textbf{opt}$_1$, $\Delta \theta$ and learning rate $\alpha^V \deltat$.
-	\STATE Update $\psi$ with \textbf{opt}$_2$, $\Delta \psi$ and learning rate $\alpha^A \deltat$.
+	\STATE $\Delta \psi \gets \frac{1}{N}\sum\limits_{i=1}^N \frac{\left(Q^i - \tilde{Q^i}\right)\partial_{\psi} \left(\bar{A}_{\psi}(s^i, a^i) - \bar{A}_{\psi}(s^i, \pi_\phi(s^i))\right) }{\deltat}$
+	\STATE $\Delta \phi \gets \frac{1}{N} \sum\limits_{i=1}^N \partial_a \bar{A}_\psi(s^i, \pi_\phi(s^i)) \partial_\phi \pi_\phi(s^i)$
+	\STATE Update $\theta$ with \textbf{opt}$_V$, $\Delta \theta$ and learning rate $\alpha^V \deltat$.
+	\STATE Update $\psi$ with \textbf{opt}$_A$, $\Delta \psi$ and learning rate $\alpha^A \deltat$.
+	\STATE Update $\phi$ with \textbf{opt}$_\pi$, $\Delta \phi$ and learning rate $\alpha^\pi \deltat$.
 	\ENDFOR

### 2019-01-21 paper/icml_drau/impl_details.tex
.

%! TEX root = supplementary.tex
All the details specifying our implementation are hereby given. We first give precise pseudo code
descriptions for both \emph{Continuous Deep Advantage Updating} (Alg.~\ref{alg:cdau}), as well as the variants of DDPG (Alg.~\ref{alg:ddpg})
and DQN (Alg.~\ref{alg:dqn}) used.

\begin{algorithm}
	\input{cdau.tex}
	\caption{Continuous DAU}
	\label{alg:cdau}
\end{algorithm}
\begin{algorithm}
	\input{ddpg.tex}
	\caption{DDPG}
	\label{alg:ddpg}
\end{algorithm}
\begin{algorithm}
	\input{dqn.tex}
	\caption{DQN}
	\label{alg:dqn}
\end{algorithm}

For DDPG and DQN, two different settings were experimented with:
\begin{itemize}
	\item One with time discretization scalings, to keep the comparison
		fair. In this setting, the discount factor is still scaled as $\gamma^\deltat$,
		rewards are scaled as $r \deltat$, and learning rates are scaled to obtain parameter
		updates of order $\deltat$. As RMSprop is used for all experiments, this amounts
		to using a learning rate scaling as $\alpha^Q = \tilde{\alpha}^Q \deltat$,
		$\alpha^\pi = \tilde{\alpha}^\pi \deltat$.
	\item One without discretization scalings. In that case, only the discount factor is scaled
		as $\gamma^\deltat$, to prevent unfair shortsightedness. All other
		parameters are set with a reference $\deltat_0 = 1e-2$. For instance,
		for all $\deltat$'s, the reward perceived is $r * \deltat_0$, and
		similarily for learning rates, $\alpha^Q = \tilde{\alpha}^Q
		\deltat_0$, $\alpha^\pi = \tilde{\alpha}^Q \deltat_0$. These scalings
		don't depend on the discretization, but perform decently at least for
		the highest discretization.
\end{itemize}
\subsection{Global hyperparameters}
The following hyperparameters are maintained constant throughout all our experiments,
\begin{itemize}
	\item All networks used are of the form
		\begin{verbatim}
		Sequential(
		    Linear(nb_inputs, 256),
		    LayerNorm(256),
		    ReLU(),
		    Linear(256, 256),
		    LayerNorm(256),
		    ReLU(),
		    Linear(256, nb_outputs)
		).
		\end{verbatim}
	Policy networks have an additional $\tanh$ layer to constraint action range. On certain
	environments, network inputs are normalized by applying a mean-std normalization, with
	mean and standard deviations computed on each individual input features, on all previously
	encountered samples.
	\item $\mathcal{D}$ is a cyclic buffer of size $1000000$.
	\item $\textbf{nb\_steps}$ is set to $10$, and $256$ environments are run in parallel to
		accelerate the training procedure, totalling $2560$ environment interactions between
		learning steps.
	\item $\textbf{nb\_learn}$ is set to $50$.
	\item The physical $\gamma$ is set to $0.8$. It is always scaled as $\gamma^\deltat$ (even for
		unscaled DQN and DDPG).
	\item $N$, the batch size is set to $256$.
	\item RMSprop is used as an optimizer without momentum, and with
		$\alpha=1 - \deltat$ (or $1 - \deltat_0$ for unscaled DDPG and
		DQN).
	\item Exploration is always performed as described in the main text. The OU process used as
		parameters $\kappa = 7.5$, $\sigma = 1.5$.
	\item Unless otherwise stated, $\alpha_1 \deq \tilde{\alpha}^Q = \alpha^V = \alpha^A = 0.1$, $\alpha_2 \deq \tilde{\alpha}^\pi =
		\alpha^\pi = 0.03$.
	\item $\tau = 0.9$
\end{itemize}
\subsection{Environment dependent hyperparameters}
We hereby list the hyperparameters used for each environment. Continuous actions environments are marked with a
(C), discrete actions environments with a (D).
\begin{itemize}
	\item {\bf Ant (C)}: State normalization is used. Discretization range: $[0.05, 0.02, 0.01, 0.005, 0.002]$.
	\item {\bf Cheetah (C)}: State normalization is used. Discretization range: $[0.05, 0.02, 0.01, 0.005, 0.002]$
	\item {\bf Bipedal Walker (C)}: State normalization is used, $\alpha_2 = 0.02$. Discretization range: $[0.01, 0.005, 0.002, 0.001]$.
	\item {\bf Cartpole (D)}: $\alpha_2 = 0.02$, $\tau = 0$. Discretization range: $[0.01, 0.005, 0.002, 0.001, 0.0005]$.
	\item {\bf Pendulum (C)}: $\alpha_2 = 0.02$, $\tau = 0$. Discretization range: $[0.01, 0.005, 0.002, 0.001, 0.0005]$.

\end{itemize}

### 2019-01-21 paper/icml_drau/policy_improvement.tex
.

%! TEX root = supplementary.tex
Let $(s_t)_{t\geq 0}$ be a trajectory sampled from $\pi'$ i.e.\ solution of the equation
\begin{equation}
	ds_t / dt = F(s_t, \pi'(s_t))
\end{equation}
with initial condition $s_0 = s$.

Define
\begin{equation}
	B(T) = \int_{t=0}^T \gamma^t r(s_t, \pi'(s_t)) dt + \gamma^T V^\pi(s_T).
\end{equation}
This function if continuously differentiable, and its derivative is
\begin{align}
	\dot{B}(T) &= \gamma^T r(s_T, \pi'(s_T)) + \gamma^T \partial_s V^\pi(s) F(s, \pi'(s)) + \gamma^T \ln(\gamma) V^\pi(s_T)\\
		   &= \gamma^T A^\pi(s_T, \pi'(s))
		   &\geq 0.
\end{align}
Thus $B$ is increasing, and $B(0) = V^\pi(s)$, $\lim\limits_{T\rightarrow \infty} B(t) = V^{\pi'}(s)$.
Consequently, $V^\pi(s) \leq V^{\pi'}(s)$, which ends the proof.

### 2019-01-22 paper/icml_drau/proof_conv_adv.tex

%! TEX root = supplementary.tex
To make things more precise, we denote $\tilde{s}_\deltat^t(s_0)$ the
evaluation at instant $t$ of the solution of $d\tilde{s}_t/dt = F(\tilde{s}_t,
\pi(s_0))$ with starting point $s_0$.

Bellman equation on $Q^\pi_\deltat$ yields
\begin{align}
	Q^\pi_\deltat(s, a) &= r(s, a) \deltat + \gamma^\deltat V^\pi_\deltat(\tilde{s}_\deltat^\deltat(s)).
\end{align}

For all $s$, a first order taylor expansion yields
\begin{equation}
	\tilde{s}_\deltat^\deltat(s) = s + F(s, a)\deltat + \bigO(\deltat^2)
\end{equation}
thus, by uniform boundedness of the hessian of $V^\pi_\deltat$,\footnote{
	Without boundedness of the hessian, we cannot write the second order taylor expansion of
	$V^\pi_\deltat(\tilde{s}_\deltat^\deltat(s))$ in term of $\deltat$.
}
\begin{equation}
	Q^\pi_\deltat(s, a) = r(s, a) \deltat + (1 + \ln(\gamma) \deltat + \bigO(\deltat^2)) (V^\pi_\deltat(s) + \deltat \partial_s V^\pi_\deltat(s) F(s, a) + \bigO(\deltat^2).
\end{equation}
Now, this yields
\begin{equation}
	A^\pi_\deltat(s, a) = r(s, a) + \ln(\gamma) V^\pi_\deltat(s) + \partial_s V^\pi_\deltat(s) F(s, a) + \bigO(\deltat),
\end{equation}
and using the convergence of $V^\pi_\deltat(s)$ to $V^\pi(s)$ (Thm.~\ref{th:conv-value}) and $\partial_s V^\pi_\deltat(s)$ to $\partial_s V^\pi(s)$ (hypothesis) yields the result with 
\begin{equation}
	A^\pi(s, a) = r(s, a) + \ln(\gamma) V^\pi(s) + \partial_s V^\pi(s) F(s, a).
	\label{eq:adv_function}
\end{equation}

### 2019-01-22 paper/icml_drau/advup.tex
.

@@
 to an action-dependent quantity which, if it can be learned, contains the
-necessary information for policy improvement \TODO{See Thm... in the appendix; we
-should put the theorem in the appendix, because otherwise the theorem is
-just a paraphrase of the previous sentence}.
+necessary information for policy improvement. A formal statement is given in the supplementary material.
 
@@
 \begin{theorem}
-	\TODO{We need a bit more details: we probably want to assume that states and actions are
-	drawn from a single fixed trajectory, and that we are learning for a fixed, constant policy.}
-	Let $\eta^V_\deltat = \alpha^V \deltat^\beta$ (resp. $\eta^A \deltat^\beta$) be the learning rate
-	of $V$ (resp. $A$), when training on a single exploratory trajectory. Then
+	\TODO{Assume a continuous action space. Let $(s_t, a_t)_{t\geq 0}$ be a
+	time continuous exploratory trajectory. Let $\pi$ be a sufficiently
+regular policy, and $\bar{A}$ and $\bar{V}$ be regular function approximators.}
+	Let $\eta^V_\deltat = \alpha^V \deltat^\beta$ (resp. $\eta^A = \alpha^A \deltat^\beta$) be the learning rate
+	of $V$ (resp. $\bar{A}$). Then, when learning on $(s_t, a_t)_{t\geq 0}$, 
 	\begin{itemize}
@@
 			trajectories as $\deltat$ goes to $0$.
-		\item If $\beta < 1$, parameters grow arbitrarily large in a single timestep as $\deltat$
-			goes to $0$. \TODO{not quite true, in a single
+		\item If $\beta < 1$, parameters grow arbitrarily large in after an arbitrarily small physical time when $\deltat$ goes to $0$. \TODO{not quite true, in a single
 			step the update is bounded. But the parameters
@@
 material.
+
+%%% Local Variables:
+%%% TeX-master: "icml_drau"
+%%% End:

### 2019-01-22 paper/icml_drau/experiments.tex
.

@@
 In all settings, we use the learning algorithm described in
-Alg.~\ref{alg:dau} and \TODO{supplementary, algo 1}. The variants of DDPG and DQN
+Alg.~\ref{alg:dau} and supplementary, Algo 1. The variants of DDPG and DQN
 used are described in the supplementary material, as well as all the hyperparameters
-used. To make comparison fair, we use the properly rescaled discount
+used. To make comparison fair, we tested to different scalings for DDPG and DQN.
+One where we use the properly rescaled discount
 factor $\gamma_\deltat$ and reward $r_\deltat$ from \eqref{eq:def-gamma},
-as well as scaled learning rates, for DDPG and DQN. \TODO{NDY not
+as well as scaled learning rates. One where we only scaled the discount factor
+(to avoid shortsightedness), but left all other hyperaparameters constant
+across time discretizations. Experiments on the latter yielded better result
+and is presented hereby. Experiments on the first variant are presented in the
+supplementary material. \TODO{NDY not
 convincing: precisely because DDPG and DQN are not invariant, they might
@@
 % \end{figure}
+
+%%% Local Variables:
+%%% TeX-master: "icml_drau"
+%%% End:

### 2019-01-22 paper/icml_drau/framework.tex
.

@@
 make sure the various quantities involved converge to meaningful
-limits when $\deltat\to 0$.
+limits when $\deltat\to 0$. \TODO{Formal statements are made in the following section. Proofs and technical assumptions can be found in the supplementary material.}
 
@@
     \label{th:conv-value}
-   Under suitable assumptions \TODO{ref appendix}, for all $s \in {\cal
-   S}$, one has
+    Assume that $r: {\cal S} \times {\cal A} \rightarrow \bb R$ is bounded, and
+    that $s \rightarrow r(s, \pi(s))$ is lipschitz continuous, then
+    for all $s \in {\cal S}$, one has
     %\begin{equation}
     %\label{eq:conv-value}
-      $V^\pi_\deltat(s) = V^\pi(s) + \bigO(\deltat)$
+    $V^\pi_\deltat(s) = V^\pi(s) + \smallo(\deltat)$
     %\end{equation}
@@
 
-\subsection{There is no $Q$-function in continuous time}
+\subsection{There is No $Q$-Function in Continuous Time}
 
@@
   \end{proof}
+
+%%% Local Variables:
+%%% TeX-master: "icml_drau"
+%%% End:

### 2019-01-22 paper/icml_drau/proof_conv_params.tex
.

%! TEX root = supplementary.tex
Let $(s_t, a_t)_{t\geq 0}$ be the trajectory on which parameters are
learnt. To simplify notations, define
\begin{equation}
	A_\psi(s, a) = \bar{A}_\psi(s, a) - \bar{A}_\psi(s, \pi(s)).
\end{equation}
Define $F$ as
\begin{align}
	F^\theta(\theta, \psi, s, a) &= \alpha^V (r(s, a) + \ln(\gamma) V_\theta(s) + \partial_s V_\theta(s) F(s, a) - A_\psi(s, a))\partial_\theta V_\theta(s)\\
	F^\psi(\theta, \psi, s, a) &= \alpha^A (r(s, a) + \ln(\gamma) V_\theta(s) + \partial_s V_\theta(s) F(s, a) - A_\psi(s, a))\partial_\psi A_\psi(s, a).
\end{align}
Both $V$ and $A$ are uniformly lipschitz continuous in $\theta$ and $\psi$, thus
$F$ is lipschitz continuous.

The discrete equations for parameters updates with learning rates $\alpha^V \deltat^\beta$ and
$\alpha^A \deltat^\beta$ are
\begin{align}
	\delta Q &= r(s_{k \deltat}, a_{k \deltat}) \deltat + 
		\gamma^\deltat V_{\theta^{k}_\deltat}(s_{(k + 1)\deltat}) - 
		V_{\theta^k_\deltat}(s_{k\deltat}) - A_\psi(s_{k \deltat}, a_{k \deltat})\\
	\theta^{k+1}_\deltat &= \theta^k_\deltat + 
	\alpha^V \deltat^\beta \frac{\delta Q
	}{\deltat}\partial_\theta V_{\theta^k_\deltat}(s_{k\deltat})\\
	\psi^{k+1}_\deltat &= \psi^k_\deltat + 
	\alpha^A \deltat^\beta \frac{\delta Q
	}{\deltat}
	\partial_\psi A_{\theta^k_\deltat}(s_{k\deltat}, a_{k\deltat})
\end{align}
Under uniform boundedness of the hessian of $s \rightarrow V_\theta(s)$,
one can show
\begin{equation}
	\begin{pmatrix}
		\theta_\deltat^{k+1} \\
		\psi_\deltat^{k+1}
	\end{pmatrix} =
	\begin{pmatrix}
		\theta_\deltat^{k} \\
		\psi_\deltat^{k}
	\end{pmatrix} + \deltat^\beta F(\theta_\deltat^k, \psi_\deltat^k, s_{k \deltat}, a_{k\deltat}) + \bigO(\deltat^\beta \deltat),
\end{equation}
with a $\bigO$ independent of $k$. With the additional hypothesis that the gradient of
$(s, a) \rightarrow \bar{A}_\psi(s, a)$ is uniformly bounded, we have
\begin{itemize}
	\item For $\beta = 1$, a proof schema similar to that of Thm.~\ref{th:traj-conv} shows that
discrete trajectories converge pointwise to continuous trajectories defined by the differential equation
\begin{equation}
	\frac{d}{dt}\begin{pmatrix}
		\theta_t \\
		\psi_t
	\end{pmatrix} =
	F(\theta_t, \psi_t, s_t, a_t),
\end{equation}
which admits unique solutions for all initial parameters, since $F$ is uniformly lipschitz continuous.
\item Similarly, for $\beta > 1$, the proof schema of Thm.~\ref{th:traj-conv} shows that
discrete trajectories converge pointwise to continuous trajectories defined by the differential equation
\begin{equation}
	\frac{d}{dt}\begin{pmatrix}
		\theta_t \\
		\psi_t
	\end{pmatrix} = 0
\end{equation}
and thus that trajectories shrink to a single point as $\deltat$ goes to $0$.
\end{itemize}

### 2019-01-22 paper/icml_drau/supplementary.tex
.

@@
 \usepackage{xcolor}
+\usepackage{algorithm}
+\usepackage{algorithmic}
 
@@
 In what follows, we will assume that the reward function $r: {\cal S} \times {\cal A} \rightarrow \bb{R}$
-is bounded, to ensure existence of $V^\pi$ and $V^\pi_\deltat$ for all $\deltat$.\begin{theorem}
-	Under suitable assumptions \TODO{ref appendix}, for all $s \in {\cal
-	S}$, one has
+is bounded, to ensure existence of $V^\pi$ and $V^\pi_\deltat$ for all $\deltat$
+.\begin{theorem}
+	Assume that $r: {\cal S} \times {\cal A} \rightarrow \bb R$ is bounded, and
+	that $s \rightarrow r(s, \pi(s))$ is $L_r$-lipschitz continuous, then
+	for all $s \in {\cal S}$, one has
 	%\begin{equation}
 	%\label{eq:conv-value}
-	$V^\pi_\deltat(s) = V^\pi(s) + \bigO(\deltat)$
+	$V^\pi_\deltat(s) = V^\pi(s) + \smallo(\deltat)$
 	%\end{equation}
@@
 \section{Implementation details}
+\input{impl_details.tex}
 \end{document}

### 2019-01-22 paper/icml_drau/supplementary.tex

@@
 \section{Proofs}
-We hereby give proofs for all the results presented in the paper.
+We now give proofs for all the results presented in the paper.
 The first result presented is a proof of convergence for discretized
@@
 \begin{theorem}
-	Let $F: {\cal S} \times {\cal A} \rightarrow \bb{R}^n$ and $\pi: {\cal S}
+	Let $F\from {\cal S} \times {\cal A} \rightarrow \bb{R}^n$ and $\pi\from {\cal S}
 	\rightarrow {\cal A}$ be the dynamic and policy functions. Assume that,
@@
 	For any $\deltat > 0$, define the discretized trajectory
-	$(s_\deltat^k)_k$ recurrently as $s_\deltat^0 = s_0$,
+	$(s_\deltat^k)_k$ which amounts to maintaining each action for a
+	time interval $\deltat$; it is defined by induction as $s_\deltat^0 = s_0$,
 	$s_\deltat^{k + 1}$ is the value at time $\deltat$ of
@@
 	\end{equation}
-	Notably, discretized trajectories converge pointwise to continuous trajectories.
+	Therefore, discretized trajectories converge pointwise to continuous trajectories.
 	\label{th:traj-conv}
@@
 \end{proof}
-In what follows, we will assume that the reward function $r: {\cal S} \times {\cal A} \rightarrow \bb{R}$
+
+In what follows, we assume that the continuous-time reward function $r\from {\cal S} \times {\cal A} \rightarrow \bb{R}$
 is bounded, to ensure existence of $V^\pi$ and $V^\pi_\deltat$ for all $\deltat$
 .\begin{theorem}
-	Assume that $r: {\cal S} \times {\cal A} \rightarrow \bb R$ is bounded, and
-	that $s \rightarrow r(s, \pi(s))$ is $L_r$-lipschitz continuous, then
+	Assume that $r\from {\cal S} \times {\cal A} \rightarrow \bb R$ is bounded, and
+	that $s \rightarrow r(s, \pi(s))$ is $L_r$-Lipschitz continuous, then
 	for all $s \in {\cal S}$, one has
@@
 	%\end{equation}
-	when $\deltat\to 0$.
+	Lhen $\deltat\to 0$.
 	\label{th:conv-value}
@@
 \begin{theorem}
-	Under the above hypothesis, there exists $A^\pi: {\cal S} \rightarrow
+	Under the hypothesis above, there exists $A^\pi\from {\cal S} \rightarrow
 	\bb{R}$ such that $A^\pi_\deltat$ converges pointwise to $A^\pi$ as
@@
       \end{proof}
+
+      \begin{theorem}
+	Exploration strategy
+\end{theorem}
+\begin{proof}
+	\input{proof_conv_expl.tex}
+\end{proof}
 \section{Implementation details}

### 2019-01-22 pendulum_exp/dataanalysis/script_corentin.py
.

@@ -1,10 +1,11 @@
 import argparse
 from dataloader import loader_leonard, ExperimentData
 from plots import plot_learning_curves
+from typing import List
 
 parser = argparse.ArgumentParser()
 parser.add_argument('--logdir', type=str, required=True)
-parser.add_argument('--exp_name', type=str, required=True)
+parser.add_argument('--exp_names', nargs='+', type=str, required=True)
 parser.add_argument('--std_type', type=str, choices=['time', 'run'])
 parser.add_argument('--min_t', type=float, default=0.)
 parser.add_argument('--max_t', type=float, default=100.)
@@ -12,9 +13,18 @@ args = parser.parse_args()
 
 start_date = 'last'
 stop_date = None
-runlist = loader_leonard(args.logdir, args.exp_name, start_date=start_date, stop_date=stop_date)
+runlist: List[ExperimentData] = []
+for exp_name in args.exp_names:
+    runlist.append(loader_leonard(args.logdir, exp_name, start_date=start_date, stop_date=stop_date))
 
 expdata = ExperimentData(runlist)
+
+def filter(args) -> bool:
+    return (hasattr(args, 'noscale') and args.noscale and 'value' in args.algo) \
+        or 'advantage' in args.algo
+
+
+expdata = expdata.filter_settings(filter)
 
 plot_learning_curves(expdata, ['Return'], args.exp_name, mint=args.min_t, maxt=args.max_t, gtype=args.std_type + "_std")
 expdata.repr_rawlogs("Return", 5)

### 2019-01-22 paper/icml_drau/proof_traj.tex
.

@@
-%! TEX root = icml_drau.tex
+%! TEX root = supplementary.tex
 For any $k$, define
@@
 \end{equation}
-Taylor integral formula gives
+The Taylor integral formula gives \TODO{explain piecewise $C^2$}
 \begin{align}
 	s_\deltat^{k + 1} &= s_\deltat^k + F(s_\deltat^k, \pi(s_\deltat^k)) \deltat +
-	\frac{1}{2}\int\limits_{0}^\deltat (\deltat - t) \frac{d^2 \tilde{s}_t}{dt^2} dt\\
+	\int_{0}^\deltat (\deltat - t) \frac{d^2 \tilde{s}_t}{dt^2} dt\\
 	s_{\deltat(k + 1)} &= s_{\deltat k} + F(s_{\deltat k}, \pi(s_{\deltat k})) \deltat +
-	\frac{1}{2}\int\limits_0^\deltat (\deltat - t) \frac{d^2 s_{t + \deltat k}}{dt^2} dt.
+	\int_0^\deltat (\deltat - t) \frac{d^2 s_{t + \deltat k}}{dt^2} dt.
 \end{align}
@@
 \end{align}
-Now, it is easy to prove by recurrence that
+Now, it is easy to prove by induction that
 \begin{equation}
@@
 \begin{equation}
-	e_\deltat^\lfloor t / \deltat \rfloor \leq (e^{K(t + \deltat)} - 1) \deltat \frac{C}{K}
+	e_\deltat^{\lfloor t / \deltat \rfloor} \leq (e^{K(t + \deltat)} - 1) \deltat \frac{C}{K}.
+	\label{eq:bound_err}
 \end{equation}
+Finally, by boundedness, of $s \rightarrow F(s, \pi(s))$, there exists $C'$ such that
+\begin{equation}
+	\|s_{\deltat\lfloor t / \deltat \rfloor} - s_t \| \leq \deltat C'.
+	\label{eq:close_s}
+\end{equation}
+Combining Eq.~\eqref{eq:close_s} with Eq.~\eqref{eq:bound_err}, one can find $C''$ such that
+\begin{equation}
+	\|s_t - s_\deltat^{\lfloor t / \deltat \rfloor}\| \leq \deltat \frac{C''}{K} e^{Kt}.
+\end{equation}

### 2019-01-22 paper/icml_drau/supplementary.tex

@@
 $V^\pi_\deltat$ are continuously differentiable, and that the gradient and
-hessian of $V^\pi_\deltat$ w.r.t. $s$ are uniformly bounded in both $s$ and $\deltat$.
+Hessian of $V^\pi_\deltat$ w.r.t. $s$ are uniformly bounded in both $s$ and $\deltat$.
 We also assume convergence of $\partial_s V^\pi_\deltat(s)$ to $\partial_s V^\pi(s)$ for
 all $s$.
+
 \begin{theorem}
@@
 	\bb{R}$ such that $A^\pi_\deltat$ converges pointwise to $A^\pi$ as
-	$\deltat$ goes to $0$.
+	$\deltat$ goes to $0$. \TODO{give expression for $A^\pi$ (used
+	later + interesting per se)}
 \end{theorem}
+
 \begin{proof}
@@
 
-	If for all $s$, $A^\pi(s, \pi'(s)) \geq 0$, then for all $s$, $V^\pi(s) \leq V^{\pi'}(s)$.
+	If for all $s$, $A^\pi(s, \pi'(s)) \geq 0$, then for all $s$,
+	$V^\pi(s) \leq V^{\pi'}(s)$. \TODO{Moreover, if
+	$V^{\pi'}(s)>V^{\pi}(s)$, then $A^\pi(s,\pi'(s))>0$ for some state $s$.}
 \end{theorem}
+
 \begin{proof}

### 2019-01-22 paper/icml_drau/proof_traj.tex

@@
 %! TEX root = supplementary.tex
+The proof mostly follos the classical argument for convergence of the
+Euler scheme for differential equations.
 For any $k$, define
@@
 Now, both $d^2 s_t/dt^2$ and $d^2 \tilde{s}_t/dt^2$ are uniformly bounded, by
-boundedness and lipschitzianity of $s \rightarrow F(s, \pi(s))$ and $s
+boundedness and Lipschitzness of $s \rightarrow F(s, \pi(s))$ and $s
 \rightarrow F(s, \pi(s_\deltat^k))$. Consequently, there exists $C$ such that

### 2019-01-22 paper/icml_drau/supplementary.tex

@@
 \section{Proofs}
-We now give proofs for all the results presented in the paper.
+
+We now give proofs for all the results presented in the paper. Most
+proofs follow standard patterns from calculus and numerical schemes for
+differential equations, except for
+Theorem~\ref{thm:policyimprovement}, which uses a nice argument specific
+to reinforcement learning to prove that the continuous-time advantage
+function contains all the necessary information for policy values.
+
 The first result presented is a proof of convergence for discretized
@@
 \begin{theorem}
+\label{thm:policyimprovement}
 	Let $\pi$ and $\pi'$ be two policies such that both $s \rightarrow r(s, \pi(s))$ and

### 2019-01-22 paper/icml_drau/proof_conv_params.tex

@@
 \end{align}
-Under uniform boundedness of the hessian of $s \rightarrow V_\theta(s)$,
+Under uniform boundedness of the Hessian of $s \mapsto V_\theta(s)$,
 one can show
@@
 \begin{itemize}
-	\item For $\beta = 1$, a proof schema similar to that of Thm.~\ref{th:traj-conv} shows that
+	\item For $\beta = 1$, a proof scheme identical to that of Thm.~\ref{th:traj-conv} shows that
 discrete trajectories converge pointwise to continuous trajectories defined by the differential equation

### 2019-01-22 paper/icml_drau/proof_conv_params.tex

@@
 which admits unique solutions for all initial parameters, since $F$ is uniformly lipschitz continuous.
-\item Similarly, for $\beta > 1$, the proof schema of Thm.~\ref{th:traj-conv} shows that
+\item Similarly, for $\beta > 1$, the proof scheme of Thm.~\ref{th:traj-conv} shows that
 discrete trajectories converge pointwise to continuous trajectories defined by the differential equation

### 2019-01-22 paper/icml_drau/supplementary.tex

@@
 proofs follow standard patterns from calculus and numerical schemes for
-differential equations, except for
-Theorem~\ref{thm:policyimprovement}, which uses a nice argument specific
-to reinforcement learning to prove that the continuous-time advantage
-function contains all the necessary information for policy values.
+differential equations, except for Theorem~\ref{thm:policyimprovement},
+which uses an argument specific to reinforcement learning to prove that
+the continuous-time advantage function contains all the necessary
+information for policy values.
 

### 2019-01-22 paper/icml_drau/discussions.tex

@@
 
-\paragraph{Near continuous time reinforcement learning and RMSprop.} RMSprop~\cite{rmsprop}
+\paragraph{Near continuous time reinforcement learning and RMSProp.} RMSProp~\cite{rmsprop}
 estimates a batched version of the second moment of gradients via moving
@@
 deterministic environments, gradients typically scale as $\bigO(1)$ in
-term of $\deltat$, as seen in \eqref{eq:approx_bellman_A}.  In that case, RMSprop
+term of $\deltat$, as seen in \eqref{eq:approx_bellman_A}.  In that case, RMSProp
 preconditioning has no effect on the suitable order of magnitude for learning
@@
 $\bigO\left(\frac{1}{\deltat}\right)$. With a fixed batch size,
-RMSprop will multiply gradients by a factor $\bigO(\sqrt{\deltat})$. In
+RMSProp will multiply gradients by a factor $\bigO(\sqrt{\deltat})$. In
 that case,
@@
 $\alpha \,\deltat$.
+
+\TODO{+ discuss the use of LSTMs to provide memory/deal with
+non-observability; Such LSTMs should take timescales into account, eg
+chrono init?}

### 2019-01-22 paper/icml_drau/intro.tex

--- 2019-01-19 paper/icml_drau/intro.tex
@@
 control environments, robotics, and many video games.  
-\TODO{Remove parenthesis.} (We focus on
+We focus on
 near-continuous time because this leads to a clear framework, but the
 approach below would be relevant in any setting in which the advantage of
-any single individual action is small compared to the total return.)%, for instance, long time horizon problems.
+any single individual action is small compared to the total return.%, for instance, long time horizon problems.
 
@@
 \emph{Deep $Q$-learning} (DQN,~\citealt{dqn}) and \emph{Deep deterministic policy
-gradient} (DDPG,~\citealt{ddpg}) are not at all \TODO{Remove familiar?}robust to changes in time discretization. This is shown experimentally in Sec.~\ref{sec:exp}. 
+gradient} (DDPG,~\citealt{ddpg}) are not at all robust to changes in time discretization. This is shown experimentally in Sec.~\ref{sec:exp}. 
 Intuitively, as the discretization timestep
@@
 other actions will be accordingly small. (This continuity effect occurs even with a
-well-adjusted decay factor $\gamma$.) \TODO{Really relevant in the introduction?}
+suitably adjusted decay factor $\gamma$.)
 
@@
 
-\TODO{Time discretization invariant algo is never  mentioned before.
-Maybe start by ``We seek a time discretization invariant algorithm.
-It should ...''} A time discretization invariant algorithm should \TODO{remain viable when the time discretization is small and} in particular admit a
+We are looking for an algorithm that would be as invariant as possible to
+changing the discretization timestep. Such an 
+algorithm should remain viable when this timestep is small, and in particular admit a
 continuous-time
@@
   \item We empirically show that standard $Q$-learning methods are not
-  robust to changes in time discretization, exhibiting degraded performance, while our algorithm provides
-  substantial robustness in that domain. \TODO{Remove ``in that domain''?}
+  robust to changes in time discretization, exhibiting degraded
+  performance, while our algorithm demonstrates
+  substantial robustness.
 \end{itemize}
+
+%%% Local Variables:
+%%% TeX-master: "icml_drau"
+%%% End:

### 2019-01-22 paper/icml_drau/framework.tex

@@
 the simulator, and is often used to approximate an underlying
-differential equation.  \TODO{(For example) $->$ In this case}, the timestep may correspond to the
+differential equation.  In this case, the timestep may correspond to the
 number of frames generated by second.  In real world environments,
@@
 would expect that decreasing $\deltat$ could only improve the performance
-of RL agents \TODO{We would remove: (though it might make optimization
-harder) It is not in the spririt of what is said here. NDY keep it.
-Anyone with a bit of experience in traditional optim would scream and say exactly
-that, if it was not included.}.  RL algorithms should, at least, be resilient to a change of
+of RL agents (though it might make optimization
+harder).  RL algorithms should, at least, be resilient to a change of
 $\deltat$, and should remain viable when $\deltat \rightarrow 0$.

### 2019-01-22 paper/icml_drau/framework.tex

@@
 make sure the various quantities involved converge to meaningful
-limits when $\deltat\to 0$. \TODO{Formal statements are made in the following section. Proofs and technical assumptions can be found in the supplementary material.}
+limits when $\deltat\to 0$. The proofs and technical assumptions for the
+formal statements below can be found in the supplementary material.
 
@@
         \end{equation}
-which is thus stable when $\deltat\to 0$. \TODO{We need to add that $\gamma$ is close to 1}
+which is thus stable when $\deltat\to 0$. 
 On the other hand,
@@
     Assume that $r: {\cal S} \times {\cal A} \rightarrow \bb R$ is bounded, and
-    that $s \rightarrow r(s, \pi(s))$ is lipschitz continuous, then
+    that $s \mapsto r(s, \pi(s))$ is lipschitz continuous. Then
     for all $s \in {\cal S}$, one has

### 2019-01-22 paper/icml_drau/framework.tex

@@
 make sure the various quantities involved converge to meaningful
-limits when $\deltat\to 0$. The proofs and technical assumptions for the
-formal statements below can be found in the supplementary material.
+limits when $\deltat\to 0$.
+
+We give semi-formal statements below; the full statements, proofs, and
+technical assumptions (typically, differentiability assumptions)
+can be found in the supplementary material.
 
@@
     \label{th:conv-value}
-    Assume that $r: {\cal S} \times {\cal A} \rightarrow \bb R$ is bounded, and
-    that $s \mapsto r(s, \pi(s))$ is lipschitz continuous. Then
-    for all $s \in {\cal S}$, one has
-    %\begin{equation}
-    %\label{eq:conv-value}
-    $V^\pi_\deltat(s) = V^\pi(s) + \smallo(\deltat)$
-    %\end{equation}
-    when $\deltat\to 0$.
+%     Assume that $r: {\cal S} \times {\cal A} \rightarrow \bb R$ is bounded, and
+%     that $s \mapsto r(s, \pi(s))$ is lipschitz continuous. Then
+%     for all $s \in {\cal S}$, one has
+%     %\begin{equation}
+%     %\label{eq:conv-value}
+%     $V^\pi_\deltat(s) = V^\pi(s) + \smallo(\deltat)$
+%     %\end{equation}
+%     when $\deltat\to 0$.
+Under suitable smoothness assumptions, $V^\pi_\deltat(s)$ converges to
+$V^\pi(s)$ when $\deltat\to 0$.
    \end{theorem}
@@
   \label{th:q-cont}
+  Under suitable smoothness assumptions,
 The action-value function of a near-continuous MDP is related to its
@@
 \end{equation}
-when $\deltat \to 0$ for every $(s,a)$, under suitable technical
-assumptions.
+when $\deltat \to 0$ for every $(s,a)$.
 \end{theorem}

### 2019-01-22 paper/icml_drau/advup.tex

@@
 	A^\pi_\deltat(s, a) &= \frac{Q^\pi_\deltat(s,a) - V^\pi_\deltat(s)}{\deltat}\\
-		    &= \frac{Q(s,a) - Q(s, \pi(s))}{\deltat}\\
-		    &= A(s, a).
+		    &= \frac{Q(s,a) - Q(s, \pi(s))}{\deltat}
+		    %\\ &
+		    = A(s, a).
 \end{align}
-In the spirit of~\cite{dueling_nets}, this condition is enforced by writing $A$ as
+In the spirit of~\cite{dueling_nets}, this condition is enforced by
+parameterizing $A$ through another function $\bar{A}$ and setting
 \begin{equation}
@@
 \end{equation}
-With this parameterization, one can hope to learn an approximation of $A^\pi$,
-from which a proper ranking of actions can be derived for any $\deltat$.
+
+This approach will lead to $\deltat$-robust algorithms for
+approximating $A^\pi$, from which a ranking of actions can be
+derived.
 

### 2019-01-22 paper/icml_drau/advup.tex

@@
 This approach will lead to $\deltat$-robust algorithms for
-approximating $A^\pi$, from which a ranking of actions can be
+approximating $A^\pi_\deltat$, from which a ranking of actions can be
 derived.
@@
 step just averages out the randomness due to the law of large numbers.
-More precisely, let
-$(a_\deltat)$ be a sequence of i.i.d.\ actions drawn from a distribution $p_A$, (for instance
-the sequence of actions generated by $\varepsilon$-greedy exploration on a
-fixed policy) in a near-continuous environment. Then, as $\deltat$ goes
-to $0$, the trajectories generated by
-this policy tend to solutions of the deterministic equation
+More precisely:%, let
+% $(a_\deltat)$ be a sequence of i.i.d.\ actions drawn from a distribution $p_A$, (for instance
+% the sequence of actions generated by $\varepsilon$-greedy exploration on a
+% fixed policy) in a near-continuous environment. Then, as $\deltat$ goes
+% to $0$, the trajectories generated by
+% this policy tend to solutions of the deterministic equation
+% \begin{equation}
+% 	ds_t/dt  = \E_{a\sim p_A}\left[F(s_t, a)\right]
+% \end{equation}
+% where the expectation cancels the exploration noise. The proof is given in the
+% supplementary material. \TODO{Is there a stronger theorm? Should we talk more about that?}
+
+\begin{theorem}
+Consider a near-continuous MDP in which an agent selects an
+action according to an $\eps$-greedy policy that mixes a deterministic
+exploitation policy $\pi$ with an action taken from a noise policy
+$\pi^\text{noise}(a|s)$ with probability $\eps$ at each step. Then the
+agent's trajectories converge when $\deltat\to 0$ to the solutions of the
+\emph{deterministic} equation
 \begin{equation}
-	ds_t/dt  = \E_{a\sim p_A}\left[F(s_t, a)\right]
+d s_t/dt= (1-\eps) F(s_t,\pi(s_t))+\eps \E_{a\sim
+\pi^\text{noise}(a|s)}F(s_t,a)
 \end{equation}
-where the expectation cancels the exploration noise. The proof is given in the
-supplementary material. \TODO{Is there a stronger theorm? Should we talk more about that?}
+\end{theorem}
 

### 2019-01-22 paper/icml_drau/advup.tex

@@
 
+\TODO{
+\begin{theorem}
+Under suitable smoothness assumptions, $A^\pi_\deltat(s,a)$ has a limit
+$A^\pi(s,a)$ when $\deltat\to 0$. The limit is informative about
+actions, namely, if a policy $\pi'$ satisfies $V^{\pi'}>V^{\pi}$ for all states, then
+$A^\pi(s,\pi'(s))>0$ for some $s$.
+\end{theorem}
+}
+
 \subsection{Timestep-Invariant Exploration}
@@
 
+\TODO{
 \begin{theorem}
@@
 \end{theorem}
+}
 
@@
 
-$V_{\theta}$ and $A_{\psi}$ are learned by a variants of $Q$-learning for continuous and
-discrete action spaces are used. Namely, 
+We learn $V_{\theta}$ and $A_{\psi}$ via suitable variants of $Q$-learning for continuous and
+discrete action spaces. Namely, 
 the true $A^\pi_\deltat$

### 2019-01-22 paper/icml_drau/advup.tex

@@
 	\,\deltat + \gamma^{\deltat}\, V(s') - V(s)\right)\\
+	\label{eq:approx_deltaQ}
 	\theta_\deltat &\leftarrow \theta_\deltat + \eta^V_\deltat
@@
 \begin{theorem}
-	\TODO{Assume a continuous action space. Let $(s_t, a_t)_{t\geq 0}$ be a
-	time continuous exploratory trajectory. Let $\pi$ be a sufficiently
-regular policy, and $\bar{A}$ and $\bar{V}$ be regular function approximators.}
-	Let $\eta^V_\deltat = \alpha^V \deltat^\beta$ (resp. $\eta^A = \alpha^A \deltat^\beta$) be the learning rate
-	of $V$ (resp. $\bar{A}$). Then, when learning on $(s_t, a_t)_{t\geq 0}$, 
+Let $(s_t,a_t)$ be some exploration trajectory in a near-continuous MDP. Set the learning rates to $\eta^V_\deltat =
+\alpha^V \deltat^\beta$ and $\eta^A = \alpha^A \deltat^\beta$ for some
+$\beta\geq 0$, and learn the parameters $\theta$ and $\psi$ by iterating
+\eqref{eq:approx_deltaQ}--\eqref{eq:approx_bellman_A}. Then, when
+$\deltat\to 0$:
+% 	\TODO{Assume a continuous action space. Let $(s_t, a_t)_{t\geq 0}$ be a
+% 	time continuous exploratory trajectory. Let $\pi$ be a sufficiently
+% regular policy, and $\bar{A}$ and $\bar{V}$ be regular function approximators.}
+% 	Let $\eta^V_\deltat = \alpha^V \deltat^\beta$ (resp. $\eta^A = \alpha^A \deltat^\beta$) be the learning rate
+% 	of $V$ (resp. $\bar{A}$). Then, when learning on $(s_t, a_t)_{t\geq 0}$, 
 	\begin{itemize}
 		\item If $\beta = 1$ the discrete parameter trajectories converge to continuous parameter
-			trajectories as $\deltat$ goes to $0$.
-		\item If $\beta < 1$, parameters grow arbitrarily large in after an arbitrarily small physical time when $\deltat$ goes to $0$. \TODO{not quite true, in a single
-			step the update is bounded. But the parameters
-			become arbitrary large in an arbitrarily small
-			physical time.}
-		\item If $\beta > 1$, parameter trajectories become
-		stationary as
-			$\deltat$ goes to $0$.
+			trajectories.% as $\deltat$ goes to $0$.
+		\item If $\beta>1$ the parameter trajectories become
+		stationary.% as $\deltat$ goes to $0$.
+		\item If $\beta < 1$, the parameters can reach infinity
+		in arbitrarily small physical time.%	parameters grow arbitrarily large in after an arbitrarily small physical time when $\deltat$ goes to $0$.
 	\end{itemize}

### 2019-01-22 paper/icml_drau/advup.tex

@@
 \begin{theorem}
+	\label{th:cont-params}
 Let $(s_t,a_t)$ be some exploration trajectory in a near-continuous MDP. Set the learning rates to $\eta^V_\deltat =
@@
 $\beta\geq 0$, and learn the parameters $\theta$ and $\psi$ by iterating
-\eqref{eq:approx_deltaQ}--\eqref{eq:approx_bellman_A}. Then, when
+\eqref{eq:approx_deltaQ}--\eqref{eq:approx_bellman_A} along the
+trajectory $(s_t,a_t)$. Then, when
 $\deltat\to 0$:
@@
 	\end{itemize}
-	\label{th:cont-params}
 \end{theorem}
@@
 
-This results in a fullly specified algorithm,
+\TODO{Thus, the right learning rate scaling } results in a fullly specified algorithm,
 \emph{Deep Advantage Updating} (DAU).

### 2019-01-22 paper/icml_drau/discussions.tex

@@
 assumed to be continuously differentiable for all $\pi$.  This %is a strong
-hypothesis is not always satisfied in practice. Very simple environments
-break this hypothesis, such as the pendulum swing-up environment,
-with pendulum height as reward and torque as action.  Indeed, 
-% assume that $\pi$
-% is the optimal policy for the pendulum swing-up environment, with state
-% representation $(\theta, \dot{\theta})$. 
-the state space $(\theta, \dot{\theta})$ for this environment
-can be partitioned into two subspaces $\mathcal{G}$ and $\mathcal{B}$, with
-$\mathcal{G}$ the subspace where the optimal policy 
-can swing up the pendulum without crossing the $y$-axis, and $\mathcal{B}$ where
-it has to cross the $y$-axis. The difference of $V$-function between a point in
-$\mathcal{G}$ and a point in $\mathcal{B}$ is lower bounded by a finite
-quantity. 
-%Besides, one can find two points $s_1 \in \mathcal{G}$ and $s_2 \in \mathcal{B}$
-%such that $s_1$ and $s_2$ are arbitrarily close.
-Consequently, the value function is
-discontinuous on the frontier between $\mathcal{G}$ and $\mathcal{B}$, which results in non infinitesimal advantages
-for actions taken on the frontier. \TODO{Non infinitesimal advantages are found,
+hypothesis is not always satisfied in practice. 
+\TODO{For instance, in the pendulum swing-up environment, depending on initial
+position and momentum, the optimal policy may need to turn round a
+certain number of times before reaching the desired state. At the
+points of state space where the number of turns changes, the value is
+discontinuous.
+% Very simple environments
+% break this hypothesis, such as the pendulum swing-up environment,
+% with pendulum height as reward and torque as action.  Indeed, 
+% % assume that $\pi$
+% % is the optimal policy for the pendulum swing-up environment, with state
+% % representation $(\theta, \dot{\theta})$. 
+% the state space $(\theta, \dot{\theta})$ for this environment
+% can be partitioned into two subspaces $\mathcal{G}$ and $\mathcal{B}$, with
+% $\mathcal{G}$ the subspace where the optimal policy 
+% can swing up the pendulum without crossing the $y$-axis, and $\mathcal{B}$ where
+% it has to cross the $y$-axis. The difference of $V$-function between a point in
+% $\mathcal{G}$ and a point in $\mathcal{B}$ is lower bounded by a finite
+% quantity. 
+% %Besides, one can find two points $s_1 \in \mathcal{G}$ and $s_2 \in \mathcal{B}$
+% %such that $s_1$ and $s_2$ are arbitrarily close.
+% Consequently, the value function is
+%discontinuous on the frontier between $\mathcal{G}$ and $\mathcal{B}$, which 
+This results in non infinitesimal advantages
+for actions taken on the frontier. This occurs,
 more generally, as soon as a given policy has different regimes depending on
@@
 function is typically discontinuous at the frontier between two
-regimes, resulting in non infinitesimal advantages. The
+regimes, resulting in non-infinitesimal advantages. The
 continuous-time DAU properties would only hold almost-everywhere.}

### 2019-01-22 paper/icml_drau/advup.tex

@@
 \input{pendulum_fig.tex}
-For the discrete algorithm to admit a continuous time limit, the trajectories
+For the algorithm to admit a continuous-time limit, the discrete-time trajectories
 of parameters must converge to well defined trajectories as $\deltat$ goes to
-$0$.  The optimization steps should not be too large,
-in which case parameter trajectories could jump in a single gradient step,
-or too small, in which case the trajectories would become stationary as
-$\deltat$ goes to $0$.  This in turn imposes conditions on the scalings of the
-learning rates of algorithms robust to changes of the tine discretization.
+$0$.
+% The optimization steps should not be too large,
+% in which case parameter trajectories could jump in a single gradient step,
+% or too small, in which case the trajectories would become stationary as
+% $\deltat$ goes to $0$.
+This in turn imposes conditions on the scalings of the
+learning rates of algorithms robust to changes of the time discretization.
 
-The updates of $\theta$ and $\psi$ are described in
-\eqref{eq:approx_bellman_A}.  To have parameter updates that do not explode or shrink to $0$ too
-fast as $\deltat \rightarrow 0$, both $\delta \theta$ and
-$\delta \psi$ should verify stochastic differential equations in the
-limit, and notably,
-have a deterministic component of order $\deltat$. As $\deltat$ goes to $0$,
-a Taylor expansion yields (with $\deltas\deq s'-s$)
-\begin{align}
-	\delta Q_\deltat &= A(s, a)\,\deltat - r\,\deltat - \log(\gamma)
-	\,V(s)\,\deltat - \partial_s V(s) \,\deltas\nonumber \\&- \frac{(\deltas)^T \partial_{s}^2 V(s) \deltas}{2} + \smallo\left(\deltat\right)\\
-	\delta \theta_\deltat &= \eta^V_\deltat \partial_{\theta} V(s)
-	\frac{\delta Q_\deltat}{\deltat} \\%+ \smallo\left(\eta^V_\deltat\right)\\
-	\delta \psi_\deltat &= \eta^A_\deltat \partial_{\psi} A(s)
-	\frac{\delta Q_\deltat}{\deltat} %+ \smallo\left(\eta^A_\deltat\right).
-\end{align}
+Informally, in the updates of $\theta$ and $\psi$, the quantity $\delta
+Q_\deltat$ is of order $O(\deltat)$, because $s'=s+O(\deltat)$ in a
+near-continuous system. Therefore, $\delta
+Q_\deltat/\deltat$ is of order $O(1)$, so that the gradients used to
+update $\theta$ and $\psi$ are of order $O(1)$. Therefore, if the
+learning rates are of order $\deltat$, one would expect 
+the parameters $\theta$ and $\psi$ to move by $O(\deltat)$ in
+each time interval $\deltat$, thus hopefully converging to smooth
+continuous-time trajectories.
 
-If we set 
+% The updates of $\theta$ and $\psi$ are described in
+% \eqref{eq:approx_bellman_A}.  To have parameter updates that do not explode or shrink to $0$ too
+% fast as $\deltat \rightarrow 0$, both $\delta \theta$ and
+% $\delta \psi$ should verify stochastic differential equations in the
+% limit, and notably,
+% have a deterministic component of order $\deltat$. As $\deltat$ goes to $0$,
+% a Taylor expansion yields (with $\deltas\deq s'-s$)
+% \begin{align}
+% 	\delta Q_\deltat &= A(s, a)\,\deltat - r\,\deltat - \log(\gamma)
+% 	\,V(s)\,\deltat - \partial_s V(s) \,\deltas\nonumber \\&- \frac{(\deltas)^T \partial_{s}^2 V(s) \deltas}{2} + \smallo\left(\deltat\right)\\
+% 	\delta \theta_\deltat &= \eta^V_\deltat \partial_{\theta} V(s)
+% 	\frac{\delta Q_\deltat}{\deltat} \\%+ \smallo\left(\eta^V_\deltat\right)\\
+% 	\delta \psi_\deltat &= \eta^A_\deltat \partial_{\psi} A(s)
+% 	\frac{\delta Q_\deltat}{\deltat} %+ \smallo\left(\eta^A_\deltat\right).
+% \end{align}
+
+The next theorem confirms this analysis: if we set 
 $\eta^V_\deltat = \alpha^V \deltat$ and $\eta^A_\deltat = \alpha^A
-\deltat$, the parameter updates will follow a Euler discretization scheme
-for a continuous-time equation, thus presumably converging. This is
-actually the only possibility; more precisely,
+\deltat$, the parameter updates follow a Euler discretization scheme
+for a continuous-time equation and have a well-defined limit behavior. On
+the other hand, with smaller learning rates the parameters do not move,
+and can explode instantaneously with larger learning rates.
 
@@
 
-\TODO{Thus, the right learning rate scaling } results in a fullly specified algorithm,
+\TODO{Thus, the right learning rate scaling } results in a fully specified algorithm,
 \emph{Deep Advantage Updating} (DAU).

### 2019-01-22 paper/icml_drau/advup.tex

@@
 	\delta Q_\deltat &\leftarrow A(s, a)\, \deltat - \left(r
-	\,\deltat + \gamma^{\deltat}\, V(s') - V(s)\right)\\
-	\label{eq:approx_deltaQ}
+	\,\deltat + \gamma^{\deltat}\, V(s') - V(s)\right)
+	\label{eq:approx_deltaQ}\\
 	\theta_\deltat &\leftarrow \theta_\deltat + \eta^V_\deltat
-	\,\partial_{\theta} V(s) \,\frac{\delta Q_\deltat}{\deltat}\\
+	\,\partial_{\theta} V(s) \,\frac{\delta Q_\deltat}{\deltat}
+	\label{eq:approx_bellman_V}\\
 	\psi_\deltat &\leftarrow \psi_\deltat + \eta^A_\deltat
@@
 
-Informally, in the updates of $\theta$ and $\psi$, the quantity $\delta
+Informally, in the parameter updates
+\eqref{eq:approx_deltaQ}--\eqref{eq:approx_bellman_A}, the quantity $\delta
 Q_\deltat$ is of order $O(\deltat)$, because $s'=s+O(\deltat)$ in a
 near-continuous system. Therefore, $\delta
-Q_\deltat/\deltat$ is of order $O(1)$, so that the gradients used to
-update $\theta$ and $\psi$ are of order $O(1)$. Therefore, if the
+Q_\deltat/\deltat$ is $O(1)$, so that the gradients used to
+update $\theta$ and $\psi$ are $O(1)$. Therefore, if the
 learning rates are of order $\deltat$, one would expect 
@@
 each time interval $\deltat$, thus hopefully converging to smooth
-continuous-time trajectories.
+continuous-time trajectories. The next theorem formally confirms that
+learning rates of order $\deltat$ are the only possibility.
 
@@
 
-The next theorem confirms this analysis: if we set 
-$\eta^V_\deltat = \alpha^V \deltat$ and $\eta^A_\deltat = \alpha^A
-\deltat$, the parameter updates follow a Euler discretization scheme
-for a continuous-time equation and have a well-defined limit behavior. On
-the other hand, with smaller learning rates the parameters do not move,
-and can explode instantaneously with larger learning rates.
+% The next theorem confirms this analysis: with
+% $\eta^V_\deltat = \alpha^V \deltat$ and $\eta^A_\deltat = \alpha^A
+% \deltat$, the parameter updates follow a Euler discretization scheme
+% for a continuous-time equation and have a well-defined limit behavior. On
+% the other hand, with smaller learning rates the parameters do not move,
+% and can explode instantaneously with larger learning rates.\TODO{keep? or
+% just say "the next theorem confirms this"?}
 
@@
 			trajectories.% as $\deltat$ goes to $0$.
-		\item If $\beta>1$ the parameter trajectories become
-		stationary.% as $\deltat$ goes to $0$.
+		\item If $\beta>1$ the parameters stay at their initial
+		values.
+		%trajectories become
+		%stationary.% as $\deltat$ goes to $0$.
 		\item If $\beta < 1$, the parameters can reach infinity
@@
 
-\TODO{Thus, the right learning rate scaling } results in a fully specified algorithm,
-\emph{Deep Advantage Updating} (DAU).
-Pseudocode for the discrete actions algorithm is given in Alg.~\ref{alg:dau}.
-Pseudocode for continuous actions is given in the supplementary
-material.
+The resulting algorithm with suitable scalings,
+\emph{Deep Advantage Updating} (DAU), is fully specified in Alg.~\ref{alg:dau} for
+discrete actions, and in the supplementary material for continuous
+actions.
 

### 2019-01-22 paper/icml_drau/discussions.tex

@@
 \TODO{For instance, in the pendulum swing-up environment, depending on initial
-position and momentum, the optimal policy may need to turn round a
-certain number of times before reaching the desired state. At the
-points of state space where the number of turns changes, the value is
-discontinuous.
+position and momentum, the optimal policy needs to turn round a
+certain number of times before reaching the target state. The value
+function is discontinuous at the points of state space where the number
+of turns changes.
 % Very simple environments
@@
 This results in non infinitesimal advantages
-for actions taken on the frontier. This occurs,
-more generally, as soon as a given policy has different regimes depending on
-where it starts in state space. For such policies the value
-function is typically discontinuous at the frontier between two
-regimes, resulting in non-infinitesimal advantages. The
-continuous-time DAU properties would only hold almost-everywhere.}
+for actions on the boundary. In such environments where a given policy
+has different regimes depending on the initial state, the continuous-time
+analysis would only hold almost-everywhere.}
+% more generally, as soon as a given policy has different regimes depending on
+% where it starts in state space. For such policies the value
+% function is typically discontinuous at the frontier between two
+% regimes, resulting in non-infinitesimal advantages. The
+% continuous-time DAU properties would only hold almost-everywhere.
 
+\TODO{
 \paragraph{Memory buffer size scalings}~Thm.~\ref{th:cont-params} is stated for
-transitions sampled sequentially from a fixed infinite size sequence. In practice,
+transitions sampled sequentially from a fixed trajectory. In practice,
 transitions are sampled from a memory buffer, to prevent excessive correlations.
-The memory buffer used is a fixed size circular buffer, filled
-with samples from a single infinite exploratory trajectory. In
-our experiments, the same buffer size is used for all time discretization. When
-$\deltat$ goes to $0$, this corresponds to learning nearly online on the
-infinite sequence, since the buffer will only contain examples that are very
-close temporally to the end of the sequence in physical time. \TODO{this
-contradicts the use of the buffer against correlations} While the resulting
-algorithm converges to a continuous time limit, in the strictest sense it is not
-time discretization invariant, as the freshness of samples in the buffer will
-vary with the time discretization.
-A different choice would be to scale the size of the buffer by a factor of
-$\frac{1}{\deltat}$. This would amount to always learning on the same physical time interval of
-the trajectory. Our theoretical analysis does not cover this case.
-\TODO{a bit too many details here?}
+We used a fixed size circular buffer, filled
+with samples from a single growing exploratory trajectory. In
+our experiments, the same buffer size is used for all time discretizations. 
+Thus the physical-time freshness of samples in the buffer varies with the
+time discretization, and in the strictest sense using a fixed-size buffer
+breaks timestep invariance. A memory-intensive option would be to use a
+buffer of size $\frac{1}{\deltat}$ (fixed physical time); our theoretical
+analysis does not cover this case.
+% When
+% $\deltat$ goes to $0$, this corresponds to learning nearly online on the
+% infinite sequence, since the buffer will only contain examples that are very
+% close temporally to the end of the sequence in physical time. \TODO{this
+% contradicts the use of the buffer against correlations} While the resulting
+% algorithm converges to a continuous time limit, in the strictest sense it is not
+% time discretization invariant, as the freshness of samples in the buffer will
+% vary with the time discretization.
+% A different choice would be to scale the size of the buffer by a factor of
+% $\frac{1}{\deltat}$. This would amount to always learning on the same physical time interval of
+% the trajectory. Our theoretical analysis does not cover this case.
+% \TODO{a bit too many details here?}
+}
 
-\paragraph{Near continuous time reinforcement learning and RMSProp.} RMSProp~\cite{rmsprop}
+\paragraph{Near-continuous reinforcement learning and RMSProp.} RMSProp~\cite{rmsprop}
 estimates a batched version of the second moment of gradients via moving

### 2019-01-22 paper/icml_drau/advup.tex
.

--- 2019-01-22 paper/icml_drau/advup.tex
@@
 	A^\pi_\deltat(s, a) &= \frac{Q^\pi_\deltat(s,a) - V^\pi_\deltat(s)}{\deltat}\\
-		    &= \frac{Q(s,a) - Q(s, \pi(s))}{\deltat}
-		    %\\ &
-		    = A(s, a).
+		    &= \frac{Q(s,a) - Q(s, \pi(s))}{\deltat}\\
+		    &= A(s, a).
 \end{align}
-In the spirit of~\cite{dueling_nets}, this condition is enforced by
-parameterizing $A$ through another function $\bar{A}$ and setting
+In the spirit of~\cite{dueling_nets}, this condition is enforced by writing $A$ as
 \begin{equation}
@@
 \end{equation}
-
-This approach will lead to $\deltat$-robust algorithms for
-approximating $A^\pi_\deltat$, from which a ranking of actions can be
-derived.
-
-\TODO{
-\begin{theorem}
-Under suitable smoothness assumptions, $A^\pi_\deltat(s,a)$ has a limit
-$A^\pi(s,a)$ when $\deltat\to 0$. The limit is informative about
-actions, namely, if a policy $\pi'$ satisfies $V^{\pi'}>V^{\pi}$ for all states, then
-$A^\pi(s,\pi'(s))>0$ for some $s$.
-\end{theorem}
-}
+With this parameterization, one can hope to learn an approximation of $A^\pi$,
+from which a proper ranking of actions can be derived for any $\deltat$.
 
@@
 step just averages out the randomness due to the law of large numbers.
-More precisely:%, let
-% $(a_\deltat)$ be a sequence of i.i.d.\ actions drawn from a distribution $p_A$, (for instance
-% the sequence of actions generated by $\varepsilon$-greedy exploration on a
-% fixed policy) in a near-continuous environment. Then, as $\deltat$ goes
-% to $0$, the trajectories generated by
-% this policy tend to solutions of the deterministic equation
-% \begin{equation}
-% 	ds_t/dt  = \E_{a\sim p_A}\left[F(s_t, a)\right]
-% \end{equation}
-% where the expectation cancels the exploration noise. The proof is given in the
-% supplementary material. \TODO{Is there a stronger theorm? Should we talk more about that?}
-
-\TODO{
-\begin{theorem}
-Consider a near-continuous MDP in which an agent selects an
-action according to an $\eps$-greedy policy that mixes a deterministic
-exploitation policy $\pi$ with an action taken from a noise policy
-$\pi^\text{noise}(a|s)$ with probability $\eps$ at each step. Then the
-agent's trajectories converge when $\deltat\to 0$ to the solutions of the
-\emph{deterministic} equation
+More precisely, let
+$(a_\deltat)$ be a sequence of i.i.d.\ actions drawn from a distribution $p_A$, (for instance
+the sequence of actions generated by $\varepsilon$-greedy exploration on a
+fixed policy) in a near-continuous environment. Then, as $\deltat$ goes
+to $0$, the trajectories generated by
+this policy tend to solutions of the deterministic equation
 \begin{equation}
-d s_t/dt= (1-\eps) F(s_t,\pi(s_t))+\eps \E_{a\sim
-\pi^\text{noise}(a|s)}F(s_t,a)
+	ds_t/dt  = \E_{a\sim p_A}\left[F(s_t, a)\right]
 \end{equation}
-\end{theorem}
-}
+where the expectation cancels the exploration noise. The proof is given in the
+supplementary material. \TODO{Is there a stronger theorm? Should we talk more about that?}
 
@@
 
-We learn $V_{\theta}$ and $A_{\psi}$ via suitable variants of $Q$-learning for continuous and
-discrete action spaces. Namely, 
+$V_{\theta}$ and $A_{\psi}$ are learned by a variants of $Q$-learning for continuous and
+discrete action spaces are used. Namely, 
 the true $A^\pi_\deltat$
@@
 	\,\deltat + \gamma^{\deltat}\, V(s') - V(s)\right)\\
-	\label{eq:approx_deltaQ}
 	\theta_\deltat &\leftarrow \theta_\deltat + \eta^V_\deltat
@@
 \begin{theorem}
-	\label{th:cont-params}
-Let $(s_t,a_t)$ be some exploration trajectory in a near-continuous MDP. Set the learning rates to $\eta^V_\deltat =
-\alpha^V \deltat^\beta$ and $\eta^A = \alpha^A \deltat^\beta$ for some
-$\beta\geq 0$, and learn the parameters $\theta$ and $\psi$ by iterating
-\eqref{eq:approx_deltaQ}--\eqref{eq:approx_bellman_A} along the
-trajectory $(s_t,a_t)$. Then, when
-$\deltat\to 0$:
-% 	\TODO{Assume a continuous action space. Let $(s_t, a_t)_{t\geq 0}$ be a
-% 	time continuous exploratory trajectory. Let $\pi$ be a sufficiently
-% regular policy, and $\bar{A}$ and $\bar{V}$ be regular function approximators.}
-% 	Let $\eta^V_\deltat = \alpha^V \deltat^\beta$ (resp. $\eta^A = \alpha^A \deltat^\beta$) be the learning rate
-% 	of $V$ (resp. $\bar{A}$). Then, when learning on $(s_t, a_t)_{t\geq 0}$, 
+	\TODO{Assume a continuous action space. Let $(s_t, a_t)_{t\geq 0}$ be a
+	time continuous exploratory trajectory. Let $\pi$ be a sufficiently
+regular policy, and $\bar{A}$ and $\bar{V}$ be regular function approximators.}
+	Let $\eta^V_\deltat = \alpha^V \deltat^\beta$ (resp. $\eta^A = \alpha^A \deltat^\beta$) be the learning rate
+	of $V$ (resp. $\bar{A}$). Then, when learning on $(s_t, a_t)_{t\geq 0}$, 
 	\begin{itemize}
 		\item If $\beta = 1$ the discrete parameter trajectories converge to continuous parameter
-			trajectories.% as $\deltat$ goes to $0$.
-		\item If $\beta>1$ the parameter trajectories become
-		stationary.% as $\deltat$ goes to $0$.
-		\item If $\beta < 1$, the parameters can reach infinity
-		in arbitrarily small physical time.%	parameters grow arbitrarily large in after an arbitrarily small physical time when $\deltat$ goes to $0$.
+			trajectories as $\deltat$ goes to $0$.
+		\item If $\beta < 1$, parameters grow arbitrarily large after an arbitrarily small physical time when $\deltat$ goes to $0$.		\item If $\beta > 1$, parameter trajectories become
+		stationary as
+			$\deltat$ goes to $0$.
 	\end{itemize}
+	\label{th:cont-params}
 \end{theorem}
@@
 
-\TODO{Thus, the right learning rate scaling } results in a fullly specified algorithm,
+This results in a fullly specified algorithm,
 \emph{Deep Advantage Updating} (DAU).

### 2019-01-22 paper/icml_drau/supplementary.tex
.

@@
 		\frac{d\tilde{s}_t}{dt} = F(\tilde{s}_t, \pi(s_\deltat^k))
+		\label{eq:diff_2}
 	\end{equation}
@@
 	\bb{R}$ such that $A^\pi_\deltat$ converges pointwise to $A^\pi$ as
-	$\deltat$ goes to $0$. \TODO{give expression for $A^\pi$ (used
-	later + interesting per se)}
-\end{theorem}
-
+	$\deltat$ goes to $0$. Besides,
+	\begin{equation}
+		A^\pi(s, a) = r(s, a) + \partial_s V^\pi(s) F(s, a) + \log \gamma V^\pi(s).
+	\end{equation}\end{theorem}
 \begin{proof}
@@
 	If for all $s$, $A^\pi(s, \pi'(s)) \geq 0$, then for all $s$,
-	$V^\pi(s) \leq V^{\pi'}(s)$. \TODO{Moreover, if
-	$V^{\pi'}(s)>V^{\pi}(s)$, then $A^\pi(s,\pi'(s))>0$ for some state $s$.}
+	$V^\pi(s) \leq V^{\pi'}(s)$. Moreover, if for all $s$, 
+	$V^{\pi'}(s)>V^{\pi}(s)$, then there exists $s'$ such that $A^\pi(s',\pi'(s'))>0$.
 \end{theorem}
@@
 \begin{theorem}
-	Discrete parameter trajectories converge to continuous trajectories.
+	Let $\cal A = \bb R^A$ be the action space, and let ${\cal P}_1 = \bb
+	R^{p_1}$ and ${\cal P}_2 = \bb R^{p_2}$ be parameter spaces.  Let
+	$A\colon {\cal P}_1 \times {\cal S} \times {\cal A} \rightarrow \bb R$
+	and $V\colon {\cal P}_2 \times {\cal S} \rightarrow \bb R$ be ${\cal
+	C}^2$ function approximators with bounded gradients and Hessians. Let
+	$(a_t)_{t\geq 0}$ be a ${\cal C}^1$ exploratory actions trajectory and
+	$(s_t)_{t\geq 0}$ the resulting state trajectory, when starting from $s_0$ and
+	verifying Eq.~\eqref{eq:diff}.  Let $\theta^k_\deltat$ and
+	$\psi^k_\deltat$ be the discrete parameter trajectories resulting from
+	the gradient descent steps in the main text, with
+	learning rates $\eta^V = \alpha^V \deltat^\beta$ and $\eta^A = \alpha^A \deltat^\beta$. Then,
+	\begin{itemize}
+		\item If $\beta = 1$ the discrete parameter trajectories converge to continuous parameter
+			trajectories as $\deltat$ goes to $0$.
+		\item If $\beta > 1$, parameter trajectories become
+		stationary as
+			$\deltat$ goes to $0$.
+		\item If $\beta < 1$, parameters can grow arbitrarily large after an arbitrarily small physical time when $\deltat$ goes to $0$.
+	\end{itemize}
 \end{theorem}
@@
 	\input{proof_conv_params.tex}
+	\input{no_conv_beta.tex}
 \end{proof}

### 2019-01-22 paper/icml_drau/proof_conv_params.tex

@@
 \end{align}
-Both $V$ and $A$ are uniformly lipschitz continuous in $\theta$ and $\psi$, thus
-$F$ is lipschitz continuous.
+Both $V$ and $A$ are uniformly Lipschitz continuous in $\theta$ and $\psi$, thus
+$F$ is Lipschitz continuous. \TODO{need 1 more derivative assumption on
+$V, A$,
+because $F$ uses derivatives of $V, A$.}
 

### 2019-01-22 paper/icml_drau/relatedworks.tex
emacs config details

--- 2019-01-18 paper/icml_drau/relatedworks.tex
@@
 the advantage part; testing was only done on a quadratic-linear problem.
+%\TODO{mention that Baird did not experimentally study robustness. NDY:
+%fifth point below}
 
@@
 timescale and the physical system's timescale, but the collapse of the
-$Q$ function to the $V$ function is an intrinsic mathematical phenomenon from time
+$Q$ function to the $V$ function is an intrinsic mathematical phenomenon arising from time
 continuity, and cannot obviously be addressed via existing multiscale
@@
 %The approach presented in this paper is closely related to \cite{adv_upd} and \cite{cont_rl}.
+
+%%% Local Variables:
+%%% TeX-master: "icml_drau"
+%%% End:

### 2019-01-23 paper/icml_drau/advup.tex
.

--- 2019-01-22 paper/icml_drau/advup.tex
@@
 
-\TODO{
 \begin{theorem}
@@
 \end{theorem}
-}
 
@@
 
-\TODO{say somewhere around here that we assume $\deltat$ is known}
-
 \subsection{Scaling the Learning Rates}
@@
 \input{pendulum_fig.tex}
+\TODO{
 For the algorithm to admit a continuous-time limit, the discrete-time trajectories
@@
 learning rates of algorithms robust to changes of the time discretization.
+}
 

### 2019-01-23 paper/icml_drau/discussions.tex
.

@@
 
-\TODO{
 \paragraph{Memory buffer size scalings}~Thm.~\ref{th:cont-params} is stated for
@@
 % \TODO{a bit too many details here?}
-}
 
@@
 
-\TODO{+ discuss the use of LSTMs to provide memory/deal with
-non-observability; Such LSTMs should take timescales into account, eg
-chrono init?}
+More generally, the continuous-time analysis should in principle be repeated for every component of  
+a system.
+For instance, if a recurrent model is used to handle state memory or partial observability, care should be taken that the model is able to maintain moemry for a non-infinitesimal physical-time when $\deltat\to 0$ (see for instance \TODO{chronornn}).

### 2019-01-23 paper/icml_drau/experiments.tex
.

--- 2019-01-19 paper/icml_drau/experiments.tex
@@
 approaches cannot. DAU is compared to DDPG in continuous action settings and to DQN in
-discrete action settings. \TODO{(We only test off-policy
-algorithms.)}
+discrete action settings.
+We work off-policy; this excludes methods such as \TODO{(A3C, PPO, TRPO, ...)}.
 
 In all settings, we use the learning algorithm described in
-Alg.~\ref{alg:dau} and \TODO{supplementary, algo 1}. The variants of DDPG and DQN
+Alg.~\ref{alg:dau} and supplementary, Algo 1. The variants of DDPG and DQN
 used are described in the supplementary material, as well as all the hyperparameters
-used. To make comparison fair, we use the properly rescaled discount
-factor $\gamma_\deltat$ and reward $r_\deltat$ from \eqref{eq:def-gamma},
-as well as scaled learning rates, for DDPG and DQN. \TODO{NDY not
-convincing: precisely because DDPG and DQN are not invariant, they might
-benefit from a different scaling.}
+used. We tested two different setups for DDPG and DQN.
+In one, we scaled the discount factor (to avoid shortsightedness with small $\deltat$), but
+left all other hyperparameters constant across time discretizations.
+In the other, we used the properly rescaled discount
+factor and reward from Eq.~\eqref{eq:def-gamma},
+as well as $O(\deltat)$ learning rates for RMSProp.  The former yields slightly
+better results, and is presented in the main text. Experiments on the second
+variant are presented in the
+supplementary material.
 
@@
 coexist in the Half-Cheetah environment (walking on the head, walking on
-the back). \TODO{Removed, was unclear: ``and that DAU may fail to correctly weight the performance of
-each regime (as explained in Section \ref{sec:discussions}).''}
-
+the back).
 % \begin{figure}[h]

### 2019-01-23 paper/icml_drau/pendulum_fig.tex
.

@@
 \begin{figure*}[ht]
-	\includegraphics[width=\textwidth]{figs_data/pendulum_fig.pdf}
-	\caption{Policies obtained by DDPG and AU at different macroscopic instants of training.
+	\includegraphics[width=\textwidth]{figs_data/pendulum_unscaled/pendulum_fig_val_unscaled.pdf}
+	\caption{Value functions obtained by DDPG (unscaled version) and AU at different instants in physical time of training on the pendulum swing-up environment.
 Each image represents the policy learnt by the policy network, with $x$-axis representing angle, and $y$-axis angular velocity.
 The lighter the pixel, the closer to $1$ the action, the darker, the closer to $-1$.}
+        \label{fig:pend}
 \end{figure*}

### 2019-01-23 paper/icml_drau/pendulum_figs_suppl.tex
.

%! TEX root = supplementary.tex
\newcommand{\iw}{1cm}
\newcommand{\cw}{\textwidth}
\begin{figure*}[h]
	\includegraphics[width=\cw]{figs_data/pendulum_unscaled/pendulum_fig_act_unscaled.pdf}
	\caption{Policies obtained by DDPG (unscaled version) and AU at different instants in physical time of training on the pendulum swing-up environment. Each image represents the policy learnt by the policy network, with $x$-axis representing angle, and $y$-axis angular velocity. The lighter the pixel, the closer to $1$ the action, the darker, the closer to $-1$.}
        \label{fig:pend}
\end{figure*}
\begin{figure*}[h]
	\includegraphics[width=\cw]{figs_data/pendulum_scaled/pendulum_fig_act_scaled.pdf}
	\caption{Policies obtained by DDPG (scaled version) and AU at different instants in physical time of training on the pendulum swing-up environment. Each image represents the policy learnt by the policy network, with $x$-axis representing angle, and $y$-axis angular velocity. The lighter the pixel, the closer to $1$ the action, the darker, the closer to $-1$.}
        \label{fig:pend}
\end{figure*}
\begin{figure*}[h]
	\includegraphics[width=\cw]{figs_data/pendulum_scaled/pendulum_fig_val_scaled.pdf}
	\caption{Value functions obtained by DDPG (scaled version) and AU at different instants in physical time of training on the pendulum swing-up environment. Each image represents the policy learnt by the policy network, with $x$-axis representing angle, and $y$-axis angular velocity. The lighter the pixel, the closer to $1$ the action, the darker, the closer to $-1$.}
        \label{fig:pend}
\end{figure*}

### 2019-01-23 paper/icml_drau/relatedworks.tex
.

@@
 
-\TODO{Somewhere else: Say that Actor-critic has no problem when $\deltat\to 0$? With SGD
-but not with RMSProp, and needs robust exploration, and limited to
-on-policy}
+Our study focuses on off-policy algorithms. Some on-policy algorithms, such as
+A3C \TODO{cite}, or PPO \TODO{cite} may be time discretization invariant in
+specific setups. This would require a separate study.
 
@@
 $Q$ function to the $V$ function is an intrinsic mathematical phenomenon arising from time
-continuity, and cannot obviously be addressed via existing multiscale
-techniques. \TODO{cite something!}
+continuity.
 

### 2019-01-23 paper/icml_drau/pendulum_fig.tex
.

@@
 	\caption{Value functions obtained by DDPG (unscaled version) and AU at different instants in physical time of training on the pendulum swing-up environment.
-Each image represents the policy learnt by the policy network, with $x$-axis representing angle, and $y$-axis angular velocity.
-The lighter the pixel, the closer to $1$ the action, the darker, the closer to $-1$.}
+Each image represents the value function learnt, with $x$-axis representing angle, and $y$-axis angular velocity.
+The lighter the pixel, the higher the value.}
         \label{fig:pend}

### 2019-01-23 pendulum_exp/dataanalysis/plot_legend.py

@@ -34,7 +34,7 @@ ax.text(0, width + bot_margin, f"{MIN_DT:3.2e}", color="dimgrey", verticalalignm
 ax.text(50, width + bot_margin, f"{MAX_DT:3.2e}", color="dimgrey", verticalalignment='top', horizontalalignment='center', size=int(font_size * 2 / 3))
 ax.text(25, 3 * width + 3 * bot_margin, "Time discretization (log scale)", verticalalignment='top', horizontalalignment='center', size=int(font_size * 3 / 4), color="dimgrey")
 ax.text(-left_margin, width / 2, "DDPG", color="dimgrey", verticalalignment='center', horizontalalignment='right', size=font_size)
-ax.text(-left_margin, 2 * width + width / 2, "AU", color="dimgrey", verticalalignment='center', horizontalalignment='right', size=font_size)
+ax.text(-left_margin, 2 * width + width / 2, "DAU", color="dimgrey", verticalalignment='center', horizontalalignment='right', size=font_size)
 
 plt.imshow(line_color)
 plt.subplots_adjust(left=.2, right=.9)

### 2019-01-23 paper/icml_drau/discussions.tex
.

@@
 a system.
-For instance, if a recurrent model is used to handle state memory or partial observability, care should be taken that the model is able to maintain moemry for a non-infinitesimal physical-time when $\deltat\to 0$ (see for instance \TODO{chronornn}).
+For instance, if a recurrent model is used to handle state memory or partial observability, care should be taken that the model is able to maintain memory for a non-infinitesimal physical-time when $\deltat\to 0$ (see for instance~\citealt{chronornn}).

### 2019-01-23 paper/icml_drau/experiments.tex
.

@@
 discrete action settings.
-We work off-policy; this excludes methods such as \TODO{(A3C, PPO, TRPO, ...)}.
+Since our focus is the study of Q-learning based off-policy methods, we do not
+compare against on-policy methods such as A3C~\cite{a3c}, PPO~\cite{ppo} or
+TRPO~\cite{trpo}.
 
@@
 its value and policy. The results are presented in
-Fig.~\ref{fig:pend} (policy function) and in Fig.\TODO{ref} in
-Supplementary \TODO{ref} (value function).
+Fig.~\ref{fig:pend} (value function) and in Fig.~1, 2 and 3 in
+Supplementary.
 

### 2019-01-23 paper/icml_drau/relatedworks.tex
.

@@
 Our study focuses on off-policy algorithms. Some on-policy algorithms, such as
-A3C \TODO{cite}, or PPO \TODO{cite} may be time discretization invariant in
-specific setups. This would require a separate study.
+A3C~\cite{a3c}, or PPO~\cite{ppo} may be time discretization invariant in
+specific setups. This is out of the scope of our work and would require a
+separate study.
 

### 2019-01-23 paper/icml_drau/intro.tex

@@
 control environments, robotics, and many video games.
-We focus on
-near-continuous time because this leads to a clear framework, but the
-approach below would be relevant in any setting in which the advantage of
-any single individual action is small compared to the total return.%, for instance, long time horizon problems.
 
@@
 continuous-time $Q$-function.
+
+We focus on near-continuous time because this leads to a clear framework,
+but our observations make sense in any setting in which the value
+results from a taking a large number of small individual actions. Our
+results suggest naive $Q$-learning will fail in such settings without a
+delicate balance of hyperparameter scalings.
+%the
+%advantage of any single individual action is small compared to the total
+%return.%, for instance, long time horizon problems.
 

### 2019-01-23 paper/icml_drau/intro.tex

@@
 
+\todo{
 We focus on near-continuous time because this leads to a clear framework,
@@
 results suggest naive $Q$-learning will fail in such settings without a
-delicate balance of hyperparameter scalings.
+delicate balance of hyperparameter scalings.}
 %the

### 2019-01-23 paper/icml_drau/intro.tex

@@
 
-\todo{
-We focus on near-continuous time because this leads to a clear framework,
+\TODO{
+We focus on continuous time because this leads to a clear theoretical framework,
 but our observations make sense in any setting in which the value

### 2019-01-23 paper/icml_drau/intro.tex

--- 2019-01-23 paper/icml_drau/intro.tex
@@
 
-We focus on near-continuous time because this leads to a clear framework,
+\TODO{
+We focus on continuous time because this leads to a clear theoretical framework,
 but our observations make sense in any setting in which the value
-results from a taking a large number of small individual actions. Our
+comes from accumulating a large number of small individual actions. Our
 results suggest naive $Q$-learning will fail in such settings without a
-delicate balance of hyperparameter scalings.
+delicate balance of hyperparameter scalings and reparameterizations.}
 %the

### 2019-01-23 paper/icml_drau/intro.tex

--- 2019-01-23 paper/icml_drau/intro.tex
@@
 
-\todo{
-We focus on near-continuous time because this leads to a clear framework,
+\TODO{
+We focus on continuous time because this leads to a clear theoretical framework,
 but our observations make sense in any setting in which the value
-results from a taking a large number of small individual actions. Our
-results suggest naive $Q$-learning will fail in such settings without a
-delicate balance of hyperparameter scalings.}
+results from taking a large number of small individual actions. Our
+results suggest standard $Q$-learning will fail in such settings without a
+delicate balance of hyperparameter scalings and reparameterizations.}
 %the

### 2019-01-23 paper/icml_drau/advup.tex

@@
 $A^\pi(s,a)$ when $\deltat\to 0$. The limit is informative about
-actions, namely, if a policy $\pi'$ satisfies $V^{\pi'}>V^{\pi}$ for all states, then
-$A^\pi(s,\pi'(s))>0$ for some $s$.
+actions: namely, if a policy $\pi'$ strictly dominates $\pi$, 
+%satisfies $V^{\pi'}>V^{\pi}$ for all states,
+then
+$A^\pi(s,\pi'(s))>0$ for some state $s$.
 \end{theorem}
@@
 For the algorithm to admit a continuous-time limit, the discrete-time trajectories
-of parameters must converge to well defined trajectories as $\deltat$ goes to
+of parameters must converge to well-defined trajectories as $\deltat$ goes to
 $0$.
@@
 % $\deltat$ goes to $0$.
-This in turn imposes conditions on the scalings of the
-learning rates of algorithms robust to changes of the time discretization.
-}
+This in turn imposes precise conditions on the scalings of the
+learning rates.}% of algorithms robust to changes of the time discretization.  }
 

### 2019-01-23 pendulum_exp/dataanalysis/script_leonard.py
add test option in the main

@@ -22,7 +22,7 @@ def predicat_noscale(s):
 	return ('noscale' in s and s['noscale'])
 
 
-scaled = False
+scaled = True
 
 expdata = expdata.filter_settings(lambda s: (predicat_noscale(s) != scaled) or 'advantage' in s['algo'])
 

### 2019-01-23 pendulum_exp/main.py

@@ -27,7 +27,7 @@ def train(nb_steps: int, env: Env, policy: Policy, start_obs: Arrayable):
 
 def evaluate(dt: float, epoch: int, env: Env, policy: Policy, eval_gap: float,
              time_limit: Optional[float] = None, eval_return: bool = False,
-             progress_bar: bool = False, video: bool = False, no_log: bool = False):
+             progress_bar: bool = False, video: bool = False, no_log: bool = False, test=False):
     """ Evaluate. """
     log_gap = int(eval_gap / dt)
     policy.eval()
@@ -50,7 +50,10 @@ def evaluate(dt: float, epoch: int, env: Env, policy: Policy, eval_gap: float,
                            np.stack(dones, axis=0))
         info(f"At epoch {epoch}, return: {R}")
         if not no_log:
-            log("Return", R, epoch) # don't log when outputing video
+            if not test:
+                log("Return", R, epoch) # don't log when outputing video
+            else:
+                log("Return_test", R, epoch)
         if video:
             log_video("demo", epoch, np.stack(imgs, axis=0))
 
@@ -98,10 +101,13 @@ def main(args):
             eval_gap,
             time_limit,
             eval_return=e % log_gap == log_gap - 1,
+            test=False
         )
         if new_R is not None:
             # policy.observe_evaluation(new_R)
             if new_R > R:
+                evaluate(dt, e, eval_env, policy, eval_gap,
+                    time_limit, eval_return=True, test=True)
                 info(f"Saving new policy with return {new_R}")
                 state_dict = policy.state_dict()
                 state_dict["return"] = new_R

### 2019-01-23 paper/icml_drau/intro.tex

@@
 and if $a$ is only maintained for a very short time its advantage over
-other actions will be accordingly small. (This continuity effect occurs even with a
+other actions will be accordingly small. (This occurs even with a
 suitably adjusted decay factor $\gamma$.)
 
-The natural framework to study this phenomenon is Continuous time
-reinforcement learning (Sec.~\ref{sec:framework}) \cite{cont_rl,
-adv_upd}. Formally, the effect of a single action vanishes when the
-discretization timestep becomes infinitesimal: there is no
-continuous-time $Q$-function.
+\TODO{
+If the discretization timestep becomes infinitesimal, the effect of every
+individual action vanishes: there is no continuous-time $Q$-function
+(Thm.~\ref{th:q-cont}). These statements can be fully formalized in the framework
+of continuous-time reinforcement learning (Sec.~\ref{sec:framework})
+\cite{cont_rl, adv_upd}.
 
-\TODO{
 We focus on continuous time because this leads to a clear theoretical framework,

### 2019-01-23 paper/icml_drau/intro.tex

@@
 individual action vanishes: there is no continuous-time $Q$-function
-(Thm.~\ref{th:q-cont}). These statements can be fully formalized in the framework
+(Thm.~\ref{th:q-cont}), hence the poor performance of $Q$-learning with
+small time steps. These statements can be fully formalized in the framework
 of continuous-time reinforcement learning (Sec.~\ref{sec:framework})

### 2019-01-23 paper/icml_drau/advup.tex
.

@@
 \end{equation}
-with $\Theta \deq (\theta, \psi)$. (To avoid cumbersome notation, 
-parameter indices are dropped in what follows.)  At initialization, if both $V_{\theta}$ and
+with $\Theta \deq (\theta, \psi)$. At initialization, if both $V_{\theta}$ and
 $A_{\psi}$ are initialized independently of $\deltat$, this parameterization
@@
 correctly approximates $A^\pi_\deltat$ if
-$Q$ approximates $Q^\pi_\deltat$% \footnote{
+$Q_\Theta$ approximates $Q^\pi_\deltat$% \footnote{
 % 	Ensuring that $A$ does not differ from $A^\pi_\deltat$ by a state dependent function only affects
@@
 .
-Indeed, for any given pair $(V,\,A)$, the pair $(V(s) -
-f(s),\,A(s,a)+f(s)/\deltat)$ (for an arbitrary $f$)
+Indeed, for any given pair $(V_\theta,\,A_\psi)$, the pair $(V_\theta(s) -
+f(s),\,A_\psi(s,a)+f(s)/\deltat)$ (for an arbitrary $f$)
 % \begin{align}
@@
 % \end{align}
-yields the exact same $Q$ function. This new $A$ still defines %the same $Q$ and
+yields the exact same $Q_\Theta$ function. This new $A_\psi$ still defines %the same $Q$ and
 the same ranking of actions, yet this phenomenon might cause
-numerical problems or instability of $A$ when $\deltat\to 0$, and prevents direct
-interpretation of the learned $A$.
+numerical problems or instability of $A_\psi$ when $\deltat\to 0$, and prevents direct
+interpretation of the learned $A_\psi$.
 % Consequently, $Q$ can correctly approximate
 % $Q^\pi$ even if $A$ is arbitrarily far from $A^\pi$.
-To enforce identifiability of $A$, one must enforce the consistency equation
+To enforce identifiability of $A_\psi$, one must enforce the consistency equation
 \begin{equation}
@@
 \end{equation}
-on the approximate $A$ and $V$. This translates to
+on the approximate $A_\psi$ and $V_\theta$. This translates to
 \begin{equation}
-	A(s, \pi(s)) = 0.
+	A_\psi(s, \pi(s)) = 0.
 \end{equation}
-With this additional constraint, if $Q = Q^\pi_\deltat$, then $A =
-A^\pi_\deltat$ and $V = V^\pi_\deltat$: indeed \TODO{keep or  leave as
+With this additional constraint, if $Q_\Theta = Q^\pi_\deltat$, then $A_\psi =
+A^\pi_\deltat$ and $V_\theta = V^\pi_\deltat$: indeed \TODO{keep or  leave as
 an exercise?}
@@
 	A^\pi_\deltat(s, a) &= \frac{Q^\pi_\deltat(s,a) - V^\pi_\deltat(s)}{\deltat}\\
-		    &= \frac{Q(s,a) - Q(s, \pi(s))}{\deltat}
+		    &= \frac{Q_\Theta(s,a) - Q_\Theta(s, \pi(s))}{\deltat}
 		    %\\ &
-		    = A(s, a).
+		    = A_\psi(s, a).
 \end{align}
-In the spirit of~\cite{dueling_nets}, this condition is enforced by
-parameterizing $A$ through another function $\bar{A}$ and setting
+In the spirit of~\cite{dueling_nets}, instead of directly parameterizing $A_\psi$,
+we define a parametric function $\bar{A}_\psi$ (typically a neural network),
+and use $\bar{A}_\psi$ to define $A_\psi$ as
 \begin{equation}
 \label{eq:Aparam}
-	A(s, a) \deq \bar{A}(s, a) - \bar{A}(s, \pi(s)).
+	A_\psi(s, a) \deq \bar{A}_\psi(s, a) - \bar{A}_\psi(s, \pi(s)).
 \end{equation}
+% Rq Harsh
+With this definition, $A_\psi$ directly verifies the normalization condition.
 
@@
 Under suitable smoothness assumptions, $A^\pi_\deltat(s,a)$ has a limit
-$A^\pi(s,a)$ when $\deltat\to 0$. The limit is informative about
+$A^\pi(s,a)$ when $\deltat\to 0$. The limit keeps information about
 actions: namely, if a policy $\pi'$ strictly dominates $\pi$, 
@@
 \begin{equation}
-  \pi^\text{explore}(s_t, z_t)\deq \text{argmax}_{a'} \left(A(s, a') + z_t[a']\right)
+  \pi^\text{explore}(s^k_\deltat, z^k_\deltat)\deq \text{argmax}_{a'} \left(A_\psi(s^k_\deltat, a') + z^k_\deltat[a']\right)
 \end{equation}
-where $z_t[a']$ denotes the $a'$-th component of $z_t$. Namely, we
-perturb advantage values by a random process before selecting an action. The resulting scheme
-converges in continuous time to a nontrivial exploration scheme.
+where $z^k_\deltat[a']$ denotes the $a'$-th component of $z^k_\deltat$. Namely,
+we perturb advantage values by a random process before selecting an action. The
+resulting scheme converges in continuous time to a nontrivial exploration
+scheme.
 
@@
 Maximization over actions in $\pi$ is implemented exactly for discrete actions,
-and approximated by a policy neural network $\pi_\phi(s)$, trained to maximize $A(s,
+and approximated by a policy neural network $\pi_\phi(s)$, trained to maximize $A_\psi(s,
 \pi_\phi(s))$ for continuous actions. This point is similar to
@@
 \begin{align}
-	\delta Q_\deltat &\leftarrow A(s, a)\, \deltat - \left(r
-	\,\deltat + \gamma^{\deltat}\, V(s') - V(s)\right)
+	\delta Q_\deltat &\leftarrow A_\psi(s, a)\, \deltat - \left(r
+	\,\deltat + \gamma^{\deltat}\, V_\theta(s') - V_\theta(s)\right)
 	\label{eq:approx_deltaQ}\\
 	\theta_\deltat &\leftarrow \theta_\deltat + \eta^V_\deltat
-	\,\partial_{\theta} V(s) \,\frac{\delta Q_\deltat}{\deltat}
+	\,\partial_{\theta} V_\theta(s) \,\frac{\delta Q_\deltat}{\deltat}
 	\label{eq:approx_bellman_V}\\
 	\psi_\deltat &\leftarrow \psi_\deltat + \eta^A_\deltat
-	\,\partial_{\psi} A(s, a) \,\frac{\delta Q_\deltat}{\deltat}.
+	\,\partial_{\psi} A_\psi(s, a) \,\frac{\delta Q_\deltat}{\deltat}.
 	\label{eq:approx_bellman_A}

### 2019-01-23 paper/icml_drau/intro.tex
.

--- 2019-01-23 paper/icml_drau/intro.tex
@@
 suitably adjusted decay factor $\gamma$.)
-
-\TODO{
 If the discretization timestep becomes infinitesimal, the effect of every
 individual action vanishes: there is no continuous-time $Q$-function
-(Thm.~\ref{th:q-cont}). These statements can be fully formalized in the framework
+(Thm.~\ref{th:q-cont}), hence the poor performance of $Q$-learning with
+small time steps. These statements can be fully formalized in the framework
 of continuous-time reinforcement learning (Sec.~\ref{sec:framework})
@@
 results suggest standard $Q$-learning will fail in such settings without a
-delicate balance of hyperparameter scalings and reparameterizations.}
+delicate balance of hyperparameter scalings and reparameterizations.
 %the

### 2019-01-23 paper/icml_drau/supplementary.tex
.

@@
 	C}^2$ function approximators with bounded gradients and Hessians. Let
-	$(a_t)_{t\geq 0}$ be a ${\cal C}^1$ exploratory actions trajectory and
+	$(a_t)_{t\geq 0}$ be a ${\cal C}^1$ exploratory action trajectory and
 	$(s_t)_{t\geq 0}$ the resulting state trajectory, when starting from $s_0$ and
-	verifying Eq.~\eqref{eq:diff}.  Let $\theta^k_\deltat$ and
+	following $ds_t/dt=F(s_t,a_t)$.  Let $\theta^k_\deltat$ and
 	$\psi^k_\deltat$ be the discrete parameter trajectories resulting from
@@
 \input{impl_details.tex}
+\section{Additional results}
+Additional results mentionned in the text are presented in this section.
+\input{pendulum_figs_suppl.tex}
 \end{document}

### 2019-01-23 paper/icml_drau/supplementary.tex

@@
 	the gradient descent steps in the main text, with
-	learning rates $\eta^V = \alpha^V \deltat^\beta$ and $\eta^A = \alpha^A \deltat^\beta$. Then,
+	learning rates $\eta^V = \alpha^V \deltat^\beta$ and $\eta^A =
+	\alpha^A \deltat^\beta$ for some $\beta\geq 0$. Then,
 	\begin{itemize}

### 2019-01-23 paper/icml_drau/advup.tex
.

@@
 \input{pendulum_fig.tex}
-\TODO{
 For the algorithm to admit a continuous-time limit, the discrete-time trajectories
@@
 This in turn imposes precise conditions on the scalings of the
-learning rates.}% of algorithms robust to changes of the time discretization.  }
+learning rates.% of algorithms robust to changes of the time discretization.  }
 

### 2019-01-23 paper/icml_drau/experiments.tex
.

@@
 In all settings, we use the learning algorithm described in
-Alg.~\ref{alg:dau} and supplementary, Algo 1. The variants of DDPG and DQN
+Alg.~\ref{alg:dau} and supplementary, Alg.~1. The variants of DDPG and DQN
 used are described in the supplementary material, as well as all the hyperparameters
@@
 details are given in the supplementary material.} and, most notably, time elapsed is always given in
-normalized epochs (physical time), i.e. the true number of epochs multiplied by the
-discretization timestep.
+physical time, i.e. the amount of time that the agent spent interacting with the environment (this is not the computational time).
 
@@
 normalized epochs) during training, for various time discretizations
-$\deltat$, for both DAU and DDPG. With DAU, the agent's policy quickly
-converges for every time discretization without specific hyperparameter
-tuning. On the contrary, with DDPG, the agent is only able to learn for
-the two largest time discretizations; DDPG learning gets harder as $\deltat$ decreases.
+$\deltat$, for both DAU and DDPG. With DAU, the agent's policy and value function quickly
+converge for every time discretization without specific
+tuning. On the contrary, with DDPG, learning of both value function
+and policy greatly vary from one discretization to another.
 
@@
 We benchmark DAU against DDPG on classic control benchmarks: Pendulum,
-Cartpole, BipedalWalker, Ant, and Half-Cheetah environments from OpenAI
-Gym. For Pendulum and Ant, DDPG is able to learn for the highest
-$\deltat$ but fails for the others, while DAU remains reasonably
-invariant to the time discretization. On BipedalWalker, DDPG is never
-able to learn while DAU is achieving decent performance for all the
-$\deltat$'s. On Cartpole and Half-Cheetah, the results are less clear. On
-Cartpole, noise dominates, which makes interpretation difficult. On Half-Cheetah, DAU is not clearly more invariant to time discretization than
-DDPG. This could be explained by the multiple suboptimal regimes that
-coexist in the Half-Cheetah environment (walking on the head, walking on
-the back).
+Cartpole, BipedalWalker, Ant, and Half-Cheetah environments from OpenAI Gym. On
+Pendulum, Bipedal Walker and Ant, DAU is quite robust to variations of
+$\deltat$ and displays reasonable performances in all cases. On the other hand,
+DDPG's performance varies with $\deltat$; performance are either degraded as
+$\deltat$ decreases (Ant, Cheetah), or become more variable as learning
+progresses (Pendulum) with small $\deltat$'s. On Cartpole, noise dominates,
+which makes interpretation difficult. On Half-Cheetah, DAU is not clearly
+invariant to time discretization. This could be explained by the multiple
+suboptimal regimes that coexist in the Half-Cheetah environment (walking on the
+head, walking on the back).
 % \begin{figure}[h]

### 2019-01-23 paper/icml_drau/advup.tex

--- 2019-01-23 paper/icml_drau/advup.tex
@@
 to an action-dependent quantity which, if it can be learned, contains the
-necessary information for policy improvement. A formal statement is given in the supplementary material.
+necessary information for policy improvement (Thm.~\ref{thm:Alimit}).
 
@@
 \begin{theorem}
+\label{thm:Alimit}
 Under suitable smoothness assumptions, $A^\pi_\deltat(s,a)$ has a limit

### 2019-01-23 paper/icml_drau/discussions.tex

@@
 hypothesis is not always satisfied in practice. 
-\TODO{For instance, in the pendulum swing-up environment, depending on initial
-position and momentum, the optimal policy needs to turn round a
-certain number of times before reaching the target state. The value
-function is discontinuous at the points of state space where the number
-of turns changes.
+For instance, in the pendulum swing-up environment, depending on initial
+position and momentum, the optimal policy may need to oscillate before
+it is able to reach the target state. The optimal value
+function is discontinuous at the frontier between states where
+oscillations are needed and those where they are not.
 % Very simple environments
@@
 has different regimes depending on the initial state, the continuous-time
-analysis would only hold almost-everywhere.}
+analysis would only hold almost-everywhere.
 % more generally, as soon as a given policy has different regimes depending on

### 2019-01-23 paper/icml_drau/advup.tex

@@
 With this additional constraint, if $Q_\Theta = Q^\pi_\deltat$, then $A_\psi =
-A^\pi_\deltat$ and $V_\theta = V^\pi_\deltat$: indeed \TODO{keep or  leave as
-an exercise?}
+A^\pi_\deltat$ and $V_\theta = V^\pi_\deltat$: indeed
 \begin{align}
@@
 and approximated by a policy neural network $\pi_\phi(s)$, trained to maximize $A_\psi(s,
-\pi_\phi(s))$ for continuous actions. This point is similar to
+\pi_\phi(s))$ for continuous actions, similarly to
 \cite{ddpg}.
@@
 The resulting algorithm with suitable scalings,
-\emph{Deep Advantage Updating} (DAU), is fully specified in Alg.~\ref{alg:dau} for
-discrete actions, and in the supplementary material for continuous
+\emph{Deep Advantage Updating} (DAU), is specified in Alg.~\ref{alg:dau} for
+discrete actions, and in the Supplementary for continuous
 actions.

### 2019-01-23 paper/icml_drau/discussions.tex

@@
 
-\paragraph{Smoothness of the value function.} In all of our proofs, $V^\pi$ is
-assumed to be continuously differentiable for all $\pi$.  This %is a strong
+\paragraph{Smoothness of the value function.} In our proofs, $V^\pi$ is
+assumed to be continuously differentiable.  This %is a strong
 hypothesis is not always satisfied in practice. 
@@
 position and momentum, the optimal policy may need to oscillate before
-it is able to reach the target state. The optimal value
-function is discontinuous at the frontier between states where
+reaching the target state. The optimal value
+function is discontinuous at the boundary between states where
 oscillations are needed and those where they are not.
@@
 
-\paragraph{Near-continuous reinforcement learning and RMSProp.} RMSProp~\cite{rmsprop}
-estimates a batched version of the second moment of gradients via moving
-averages, and divides gradient steps by the square root of this estimate.
+\paragraph{Near-continuous reinforcement learning and RMSProp.}
+RMSProp~\cite{rmsprop} divides gradient steps by the square root of a
+moving average estimate of the second moment of gradients.
 This may interact with the learning rate scaling discussed above. In
@@
 More generally, the continuous-time analysis should in principle be repeated for every component of  
-a system.
-For instance, if a recurrent model is used to handle state memory or partial observability, care should be taken that the model is able to maintain memory for a non-infinitesimal physical-time when $\deltat\to 0$ (see for instance~\citealt{chronornn}).
+a system. For instance, if a recurrent model is used to handle state
+memory or partial observability, care should be taken that the model is
+able to maintain memory for a non-infinitesimal physical-time when
+$\deltat\to 0$ (see e.g.~\citealt{chronornn}).

### 2019-01-23 paper/icml_drau/supplementary.tex
proof exploration

--- 2019-01-22 paper/icml_drau/supplementary.tex
@@
 \newtheorem{theorem}{Theorem}
+\newtheorem{lemma}{Lemma}
 \input{macros.tex}
@@
 trajectories.
-\begin{theorem}
+\begin{lemma}
 	Let $F\from {\cal S} \times {\cal A} \rightarrow \bb{R}^n$ and $\pi\from {\cal S}
@@
 	\label{th:traj-conv}
-\end{theorem}
+\end{lemma}
 \begin{proof}
@@
 	%\end{equation}
-	Lhen $\deltat\to 0$.
+	when $\deltat\to 0$.
 	\label{th:conv-value}
@@
 \begin{theorem}
-	Exploration strategy
+  Let $F\from {\cal S} \times {\cal A} \rightarrow \bb{R}^n$ be the dynamic, and $\pi\from {\cal S} \times {\cal A} \rightarrow [0,1]$ be the policy, such that $\pi(s, \cdot)$ is a probability distribution over ${\cal A}$.
+Assume that
+  $F$ is $C^1$ with bounded derivatives, and that $\pi$ is $C^1$ and bounded.
+  For any $\deltat > 0$, define the discretized trajectory
+	$(s_\deltat^k)_k$ which amounts to sample an action from $\pi(s, \cdot)$ and maintaining each action for a
+	time interval $\deltat$; it is defined by induction as $s_\deltat^0 = s_0$,
+	$s_\deltat^{k + 1}$ is the value at time $\deltat$ of
+	the unique solution of
+	\begin{equation}
+		\frac{d\tilde{s}_t}{dt} = F(\tilde{s}_t, a_k)
+		\label{eq:diff_2}
+	\end{equation}
+	with $a_k \sim \pi(s_\deltat^k, \cdot)$ and initial point $s_\deltat^k$.
+
+        Then the
+        agent's trajectories converge when $\deltat\to 0$ to the solutions of the
+        \emph{deterministic} equation:
+        \begin{equation}
+		\frac{ds_t}{dt} = \E_{a\sim \pi(s_t, \cdot)}F(s_t, a).
+		\label{eq:diff}
+              \end{equation}
+              Notably, if $\pi$ is an epsilon greedy strategy that mixes a deterministic exploitation policy $\pi^{\text{deterministic}}$ with an action taken from a noise policy $\pi^{\text{noise}}$ with probability $\varepsilon$ at, the trajectory converge to the solutions of the equation:
+        \begin{equation}
+          d s_t/dt= (1-\eps) F(s_t,\pi^\text{deterministic}(s_t))+\eps \E_{a\sim
+            \pi^\text{noise}(a|s)}F(s_t,a)
+	\end{equation}
 \end{theorem}

### 2019-01-23 paper/icml_drau/experiments.tex

@@
 \input{lc_fig.tex}
-The experiment set provided here is specifically aimed at showing that
-the proposed method DAU can operate efficiently on a wide range of time
-discretizations, without specific tuning, while standard deep Q-learning
-approaches cannot. DAU is compared to DDPG in continuous action settings and to DQN in
-discrete action settings.
-Since our focus is the study of Q-learning based off-policy methods, we do not
-compare against on-policy methods such as A3C~\cite{a3c}, PPO~\cite{ppo} or
-TRPO~\cite{trpo}.
+The experiments provided here are specifically aimed at showing that
+the proposed method, DAU, works efficiently over a wide range of time
+discretizations, without specific tuning, while standard deep $Q$-learning
+approaches do not. DAU is compared to DDPG for continuous actions and to
+DQN for
+discrete actions.
+Since we focus on Q-learning-based off-policy methods, we do not
+compare against on-policy methods (A3C~\cite{a3c}, PPO~\cite{ppo},
+TRPO~\cite{trpo}...).
 
-In all settings, we use the learning algorithm described in
-Alg.~\ref{alg:dau} and supplementary, Alg.~1. The variants of DDPG and DQN
-used are described in the supplementary material, as well as all the hyperparameters
-used. We tested two different setups for DDPG and DQN.
+In all setups, we use the algorithms described in
+Alg.~\ref{alg:dau} and Supplementary Alg.~1. The variants of DDPG and DQN
+used are described in the Supplementary, as well as all hyperparameters. We tested two different setups for DDPG and DQN.
 In one, we scaled the discount factor (to avoid shortsightedness with small $\deltat$), but
@@
 possible across different timesteps. For example,
-return results are given in term of discretized return $R_\deltat$ as defined in \eqref{eq:discretized-return},\footnote{This mostly amounts to scaling rewards
+returns are given in term of discretized return $R_\deltat$ as defined in \eqref{eq:discretized-return},\footnote{This mostly amounts to scaling rewards
 by a factor $\deltat$ when this scaling is not naturally done in the environment. Environment-specific
@@
 To provide qualitative results, and check robustness to time
-discretization both in terms of raw return results and in terms
+discretization both in terms of returns and in terms
 of convergence of the approximate value function and policies, we first provide results on the simple pendulum environment
-from the OpenAI Gym classic control suite.  Its state space is of
+from the OpenAI Gym classic control suite.  The state space is of
 dimension $2$. We visualize both the learnt value and policy functions by
@@
 its value and policy. The results are presented in
-Fig.~\ref{fig:pend} (value function) and in Fig.~1, 2 and 3 in
+Fig.~\ref{fig:pend} (value function) and Figs.~1, 2,  3 in
 Supplementary.

### 2019-01-23 paper/icml_drau/pendulum_fig.tex

@@
 	\includegraphics[width=\textwidth]{figs_data/pendulum_unscaled/pendulum_fig_val_unscaled.pdf}
-	\caption{Value functions obtained by DDPG (unscaled version) and AU at different instants in physical time of training on the pendulum swing-up environment.
-Each image represents the value function learnt, with $x$-axis representing angle, and $y$-axis angular velocity.
-The lighter the pixel, the higher the value.}
+	\caption{Value functions obtained by DDPG (unscaled version) and
+	DAU at different instants in physical time of training on the
+	pendulum swing-up environment. Each image represents the learnt
+	value function (the $x$-axis is the angle, and the $y$-axis the angular velocity). The lighter the pixel, the higher the value.}
         \label{fig:pend}

### 2019-01-23 paper/icml_drau/supplementary.tex

@@
 	C}^2$ function approximators with bounded gradients and Hessians. Let
-	$(a_t)_{t\geq 0}$ be a ${\cal C}^1$ exploratory actions trajectory and
+	$(a_t)_{t\geq 0}$ be a ${\cal C}^1$ exploratory action trajectory and
 	$(s_t)_{t\geq 0}$ the resulting state trajectory, when starting from $s_0$ and
-	verifying Eq.~\eqref{eq:diff}.  Let $\theta^k_\deltat$ and
+	following $ds_t/dt=F(s_t,a_t)$.  Let $\theta^k_\deltat$ and
 	$\psi^k_\deltat$ be the discrete parameter trajectories resulting from
 	the gradient descent steps in the main text, with
-	learning rates $\eta^V = \alpha^V \deltat^\beta$ and $\eta^A = \alpha^A \deltat^\beta$. Then,
+	learning rates $\eta^V = \alpha^V \deltat^\beta$ and $\eta^A =
+	\alpha^A \deltat^\beta$ for some $\beta\geq 0$. Then,
 	\begin{itemize}

### 2019-01-23 paper/icml_drau/framework.tex
.

@@
 %! TEX root = icml_drau.tex
-\section{Near continuous-time reinforcement learning}
+\section{Near Continuous-Time Reinforcement Learning}
 \label{sec:framework}
@@
 the Mujoco environments~\cite{ddpg} or OpenAI Gym classic control
-environments~\cite{gym}, are discretization of continuous-time control
+environments~\cite{gym}, are discretizations of continuous-time control
 problems.  In simulated environments, the time discretization is fixed by
@@
 components thus imposes a lower bound on the discretization timestep. As
-the timestep $\deltat$ is not an intrinsic parameter of the
-environment, but a constraint imposed by computational ressources, we
-would expect that decreasing $\deltat$ could only improve the performance
+the timestep $\deltat$ is 
+%not an intrinsic parameter of the environment, but
+largely a constraint imposed by computational ressources, we
+would expect that decreasing $\deltat$ would only improve the performance
 of RL agents (though it might make optimization
@@
 Consequently there is no $Q$-learning in continuous time, foreshadowing
-problematic behaviors of near continuous time $Q$-learning.
+problematic behavior of $Q$-learning with small timesteps.
 %      \TODO{We should explain somwhere here that, if not invariant,
@@
 Let ${\cal S} = \mathbb{R}^d$ be a set of states, and ${\cal A}$ be a set of
-actions. Consider the continuous \emph{Markov Decision Process} (MDP) defined by the differential equation %\TODO{Explain the intuition of Brownian in SDE}:
+actions. Consider the continuous-time \emph{Markov Decision Process} (MDP) defined by the differential equation %\TODO{Explain the intuition of Brownian in SDE}:
         \begin{equation}
@@
 the dynamics of the environment. The agent interacts with the environment through a deterministic policy function
-$\pi \colon {\cal S} \rightarrow {\cal A}$, so that $a_t=\pi(s_t)$. 
+$\pi \colon {\cal S} \rightarrow {\cal A}$, so that $a_t=\pi(s_t)$;
+actions can be discrete or continuous. 
 For simplicity we assume here that both the dynamics and exploitation policy are
@@
  the exploration policy will be random, but care must be
-taken to properly define random policies with
-discrete actions in continuous time, see below.
+taken to define proper random policies 
+in continuous time, especially with discrete actions (Sec.~\ref{subsec:explo}).
 
-For any $\deltat>0$, we can define an MDP ${\cal
+For any timestep $\deltat>0$, we can define an MDP ${\cal
 M}_\deltat = \langle {\cal S}, {\cal A}, T_{\deltat}, r_\deltat,
-      \gamma_\deltat\rangle$ as a discretization of the continuous MDP with
+      \gamma_\deltat\rangle$ as a discretization of the continuous-time MDP with
       \emph{time discretization} $\deltat$. The %deterministic
@@
 
-By standard arguments for time discretizations of differential
-equations, for a given policy the near-continuous states converge to the
-continuous-time states when $\deltat\to 0$. \TODO{ref to appendix+formal
-statement in appendix}% We now turn to the RL-related quantities.
+% By standard arguments for time discretizations of differential
+% equations, for a given policy the near-continuous states converge to the
+% continuous-time states when $\deltat\to 0$. \TODO{ref to appendix+formal
+% statement in appendix}% We now turn to the RL-related quantities.
 
@@
           \frac{\deltat}{1-\gamma^\deltat}
-          %= - \frac{1}{\log \gamma} + \bigO(\deltat)
+          = - \frac{1}{\log \gamma} + \bigO(\deltat)
           \approx \frac{1}{1-\gamma},
@@
 \end{equation}
-when $\deltat \to 0$ for every $(s,a)$.
+when $\deltat \to 0$, for every $(s,a)$.
 \end{theorem}

### 2019-01-23 paper/icml_drau/intro.tex

@@
 performance with no expert knowledge in perfect information zero-sum
-games~\cite{alphazero}, reaching level of top players in video
+games~\cite{alphazero}, reaching top player level in video
 games~(\citealt{openai_five}, \citealt{dqn}), or learning dexterous manipulation
@@
     standard $Q$-learning is ill-behaved in this setting.
-  \item Our analysis of limit properties in near continuous environments leads to a robust off-policy algorithm. In particular,
-    it provides insights on how to design architectures, constrain exploration, and scale learning rates.
+  \item Our analysis of properties in the continuous-time limit leads to a robust off-policy algorithm. In particular,
+    it provides insights on architecture design, and constrains
+    exploration schemes and learning rates scalings.
   \item We empirically show that standard $Q$-learning methods are not

### 2019-01-23 paper/icml_drau/relatedworks.tex

@@
 $Q$-learning for small time steps and, as a solution, suggested the Advantage
-Updating algorithm, with proper $\deltat$ scalings for the $V$ part and
-the advantage part; testing was only done on a quadratic-linear problem.
+Updating algorithm, with proper scalings for the $V$ and
+advantage parts depending on timescale $\deltat$; testing was only done on a quadratic-linear problem.
 %\TODO{mention that Baird did not experimentally study robustness. NDY:
@@
 Advantage Updating as a viable algorithm in this setting.  Third,
-we provide formal proofs in a general setting for the collapse of $Q$-learning when $\deltat\to
-0$, and for the non-collapse of Advantage Updating with the proper scalings.
+we provide formal proofs in a general setting for the collapse of
+$Q$-learning when the timescale $\deltat$ tends to $0$,
+and for the non-collapse of Advantage Updating with the proper scalings.
 Fourth, we also discuss how to obtain $\deltat$-invariant exploration. Fifth, we
@@
 
-Our study focuses on off-policy algorithms. Some on-policy algorithms, such as
-A3C~\cite{a3c}, or PPO~\cite{ppo} may be time discretization invariant in
-specific setups. This is out of the scope of our work and would require a
-separate study.
+Our study focuses on off-policy algorithms. Some on-policy algorithms,
+such as A3C~\cite{a3c}, or PPO~\cite{ppo} may be time discretization
+invariant with specific setups. This is out of the scope of our work and
+would require a separate study.
 

### 2019-01-23 paper/icml_drau/framework.tex

@@
 $\deltat$-dependent scalings of the discount factor $\gamma_\deltat$
-and the reward $r_\deltat$ are as follows. These definitions fit the discrete case
-when $\deltat=1$, and provide well defined, non-trivial returns
+and reward $r_\deltat$ are as follows. These definitions fit the discrete case
+when $\deltat=1$, and provide well-defined, non-trivial returns
 and value functions when $\deltat$ goes to $0$.

### 2019-01-23 paper/icml_drau/framework.tex

@@
 R_\deltat(\tau) \deq \sum_{k=0}^\infty\gamma^{k\deltat}\,
-r(s_\deltat^k, a_\deltat^k)\,\deltat
+r(s_{k\deltat}, a_{k\deltat})\,\deltat
 \end{equation}
@@
         % [[I prefer to talk about ``number of steps'' VS ``Physical time'' than microscopic/Macroscopic]]
-        In near continuous environments, there are two notions of time:
+        In near-continuous environments, there are two notions of time:
 the algorithmic time $k$ (number of steps or actions taken), and the
@@
 \paragraph{Value function.}
-The return \eqref{eq:continuous-return} yields the
+The return \eqref{eq:continuous-return} leads to the
 following continuous-time value function
@@
   &= \E_{\tau\sim\pi}\left[\sum\limits_{k=0}^\infty
-  \gamma^{k\deltat}\,r(s^k_{\deltat}, a^k_{\deltat})\,\deltat \mid s_0 = s\right]
+  \gamma^{k\deltat}\,r(s_{k\deltat}, a_{k\deltat})\,\deltat \mid s_0 = s\right]
 \end{align}

### 2019-01-23 paper/icml_drau/advup.tex

@@
 $Q^\pi$ becomes independent of actions and thus cannot be
-used to select actions.  In near-continuous time,
+used to select actions.  With small but nonzero $\deltat$,
 $Q^\pi_\deltat$ still depends on actions, and could still be used to
@@
 
-Contrary to the action state value function,
+Contrary to the $Q$-function,
 this
 rescaled advantage function converges when $\deltat\to 0$
-to an action-dependent quantity which, if it can be learned, contains the
-necessary information for policy improvement (Thm.~\ref{thm:Alimit}).
+to a non-degenerate action-dependent quantity.
+
+\begin{theorem}
+\label{thm:Alimit}
+Under suitable smoothness assumptions, $A^\pi_\deltat(s,a)$ has a limit
+$A^\pi(s,a)$ when $\deltat\to 0$. The limit keeps information about
+actions: namely, if a policy $\pi'$ strictly dominates $\pi$, 
+%satisfies $V^{\pi'}>V^{\pi}$ for all states,
+then
+$A^\pi(s,\pi'(s))>0$ for some state $s$.
+\end{theorem}
 
@@
 
-The discretized $Q$-function rewrites as
+\paragraph{Learning $A^\pi$.} The discretized $Q$-function rewrites as
 \begin{equation}
@@
 % \end{align}
-yields the exact same $Q_\Theta$ function. This new $A_\psi$ still defines %the same $Q$ and
+yields the exact same function $Q_\Theta$. This new $A_\psi$ still defines %the same $Q$ and
 the same ranking of actions, yet this phenomenon might cause
@@
 derived.
-
-\begin{theorem}
-\label{thm:Alimit}
-Under suitable smoothness assumptions, $A^\pi_\deltat(s,a)$ has a limit
-$A^\pi(s,a)$ when $\deltat\to 0$. The limit keeps information about
-actions: namely, if a policy $\pi'$ strictly dominates $\pi$, 
-%satisfies $V^{\pi'}>V^{\pi}$ for all states,
-then
-$A^\pi(s,\pi'(s))>0$ for some state $s$.
-\end{theorem}
 

### 2019-01-23 paper/icml_drau/advup.tex

@@
 \begin{equation}
-	Q^\pi_\deltat(s, \pi(s)) - V^\pi_\deltat(s) = 0
+	V^\pi_\deltat(s) = Q^\pi_\deltat(s,
+	\pi(s))
 \end{equation}
@@
 % Rq Harsh
-With this definition, $A_\psi$ directly verifies the normalization condition.
+so that $A_\psi$ directly verifies the consistency condition.
 
@@
 exploration scheme is required. 
-For continuous actions, \cite{ddpg} already introduced a time discretization invariant exploration scheme, which consists in adding an
+For \emph{continuous} actions, \cite{ddpg} already introduced such a scheme, 
+by adding an
 \emph{Ornstein--Uhlenbeck} \cite{orn-uhl} (OU) process to actions. Formally, it is defined as
 \begin{equation}
-	\pi^\text{explore}(s^k_\deltat, z^k_\deltat) \deq \pi(s^k_\deltat) + z^k_\deltat
+	\pi^\text{explore}(s_{k\deltat}, z_{k\deltat}) \deq
+	\pi(s_{k\deltat}) + z_{k\deltat}
 \end{equation}
-with $z^k_\deltat$ the discretization of a continuous-time OU process,
+with $z_{k\deltat}$ the discretization of a continuous-time OU process,
 \begin{equation}

### 2019-01-23 paper/icml_drau/icml_drau.tex
emacs
.

@@
 % Use the following line for the initial blind version submitted for review:
-\usepackage{icml2018}
+\usepackage{icml2019}
 
@@
 \begin{icmlauthorlist}
+	\icmlauthor{Corentin Tallec}{tau}
+	\icmlauthor{L\'eonard Blier}{tau,fair}
+	\icmlauthor{Yann Ollivier}{fair}
 %\icmlauthor{Aeiau Zzzz}{equal,to}
@@
 
-%\icmlaffiliation{to}{Department of Computation, University of Torontoland, Torontoland, Canada}
-%\icmlaffiliation{goo}{Googol ShallowMind, New London, Michigan, USA}
+\icmlaffiliation{tau}{TAckling the Underspecified, Universit\'e Paris Sud}
+\icmlaffiliation{fair}{Facebook Artificial Intelligence Research}
 %\icmlaffiliation{ed}{School of Computation, University of Edenborrow, Edenborrow, United Kingdom}
@@
 Empirically, we find that
-	Q-learning-based approaches such as
-	\emph{Deep Q-learning}~\cite{dqn} and \emph{Deep Deterministic
+$Q$-learning-based approaches such as
+\emph{Deep $Q$-learning}~\cite{dqn} and \emph{Deep Deterministic
 Policy Gradient}~\cite{ddpg} %are highly sensitive to variations of time discretization and
@@
 \input{intro.tex}
-
 \input{relatedworks.tex}
-
 \input{framework.tex}
-
 \input{advup.tex}
-
 \input{experiments.tex}
-
 \input{discussions.tex}
-
 \input{conclusion.tex}
-
+\pagebreak
 \bibliography{icml_drau.bib}
-\bibliographystyle{icml2018}
+\bibliographystyle{icml2019}
 \end{document}
@@
 % Codrina Lauth. Alex Smola contributed to the algorithmic style files.
+
+%%% Local Variables:
+%%% TeX-master: "icml_drau"
+%%% End:

### 2019-01-23 paper/icml_drau/advup.tex

@@
 by adding an
-\emph{Ornstein--Uhlenbeck} \cite{orn-uhl} (OU) process to actions. Formally, it is defined as
+\emph{Ornstein--Uhlenbeck} \cite{orn-uhl} (OU) random process to the
+actions. Formally, this is defined as
 \begin{equation}
@@
 where $B_t$ is a brownian motion, $\kappa$ a stiffness parameter and
-$\sigma$ a noise scaling parameter. The discretized trajectories converge
-to non trivial continuous time trajectories, exhibiting a Brownian
+$\sigma$ a noise scaling parameter. The discretized trajectories of $z$ converge
+to nontrivial continuous-time trajectories, exhibiting Brownian
 behavior with a recall force towards $0$.
@@
   \label{eq:explore}
-	a^k_\deltat = \pi^\text{explore}_\deltat(s^k_\deltat, z^k_\deltat)
+	a_{k\deltat} = \pi^\text{explore}_\deltat(s_{k\deltat},
+	z_{k\deltat})
 \end{equation}
-with $(z^k_\deltat)_{k\geq 0}$ a sequence of random variables independent from the $a$'s and $s$'s.
+with $(z_{k\deltat})_{k\geq 0}$ a sequence of random variables independent from the $a$'s and $s$'s.
 A sufficient condition for this policy to admit a continuous-time
 limit is for the sequence
-$z_\deltat$ to converge in law to a
-well-defined continuous stochastic process $z$ as $\deltat$ goes to $0$.
+$z_{k\deltat}$ to converge in law to a
+well-defined continuous stochastic process $z_t$ as $\deltat$ goes to $0$.
 % For instance, $\varepsilon$-greedy exploration naturally falls in this framework, \TODO{[[CT would like to give these details, LB thinks it is not necessary:]] with each $z^k_\deltat$
@@
 
-Thus, for discrete actions we can obtain a temporally consistent
+Thus, for \emph{discrete} actions we can obtain a consistent
 exploration scheme by taking
-$z_\deltat$ to be a discretization of an $|\mathcal{A}|$-dimensional
+$z_\deltat$ to be a discretization of an $(\#\mathcal{A})$-dimensional
 continuous OU process, and setting
 \begin{equation}
-  \pi^\text{explore}(s^k_\deltat, z^k_\deltat)\deq \text{argmax}_{a'} \left(A_\psi(s^k_\deltat, a') + z^k_\deltat[a']\right)
+  \pi^\text{explore}(s_{k\deltat}, z_{k\deltat})\deq \text{argmax}_{a}
+  \left(A_\psi(s_{k\deltat}, a) + z_{k\deltat}[a]\right)
 \end{equation}
-where $z^k_\deltat[a']$ denotes the $a'$-th component of $z^k_\deltat$. Namely,
-we perturb advantage values by a random process before selecting an action. The
+where $z_{k\deltat}[a]$ denotes the $a$-th component of $z_{k\deltat}$. Namely,
+we perturb the \emph{advantage values} by a random process before selecting an action. The
 resulting scheme converges in continuous time to a nontrivial exploration

### 2019-01-23 paper/icml_drau/advup.tex

@@
 \begin{equation}
-  \pi^\text{explore}(s_{k\deltat}, z_{k\deltat})\deq \text{argmax}_{a}
+  \pi^\text{explore}(s_{k\deltat}, z_{k\deltat})\deq \argmax_{a}
   \left(A_\psi(s_{k\deltat}, a) + z_{k\deltat}[a]\right)
@@
 $\varepsilon$-greedy
-exploration is likely \emph{not} to explore, i.e. to collapse to a deterministic
+exploration is likely \emph{not} to explore, i.e., to collapse to a deterministic
 policy, when $\deltat$ goes to $0$.

### 2019-01-23 paper/icml_drau/framework.tex

@@
 the dynamics of the environment. The agent interacts with the environment through a deterministic policy function
-$\pi \colon {\cal S} \rightarrow {\cal A}$, so that $a_t=\pi(s_t)$;
-actions can be discrete or continuous. 
+$\pi \colon {\cal S} \rightarrow {\cal A}$, so that $a_t=\pi(s_t)$.
+Actions can be discrete or continuous. 
 For simplicity we assume here that both the dynamics and exploitation policy are
@@
   \label{eq:bellman}
-  V^\pi_\deltat(s) = \reward(s, \pi(s))\,\deltat + \gamma^{\deltat}\, \E_{s_{\deltat}^{k+1} | s_{\deltat}^{k} = s} V^\pi_\deltat(s_{\deltat}^{k+1})
+  V^\pi_\deltat(s) = \reward(s, \pi(s))\,\deltat + \gamma^{\deltat}\,
+  \E_{s_{(k+1)\deltat} | s_{k\deltat} = s} V^\pi_\deltat(s_{(k+1)\deltat})
 \end{equation}

### 2019-01-23 paper/icml_drau/experiments.tex

@@
 discrete actions.
-Since we focus on Q-learning-based off-policy methods, we do not
+Since we focus on $Q$-learning-like off-policy methods, we do not
 compare against on-policy methods (A3C~\cite{a3c}, PPO~\cite{ppo},
@@
 factor and reward from Eq.~\eqref{eq:def-gamma},
-as well as $O(\deltat)$ learning rates for RMSProp.  The former yields slightly
-better results, and is presented in the main text. Experiments on the second
-variant are presented in the
-supplementary material.
+as well as $O(\deltat)$ learning rates for RMSProp.  The first variant yields slightly
+better results, and is presented here, with the second variant in the
+Supplementary.
 
@@
 possible across different timesteps. For example,
-returns are given in term of discretized return $R_\deltat$ as defined in \eqref{eq:discretized-return},\footnote{This mostly amounts to scaling rewards
+returns are given in terms of the discretized return $R_\deltat$ as defined in \eqref{eq:discretized-return},\footnote{This mostly amounts to scaling rewards
 by a factor $\deltat$ when this scaling is not naturally done in the environment. Environment-specific
-details are given in the supplementary material.} and, most notably, time elapsed is always given in
-physical time, i.e. the amount of time that the agent spent interacting with the environment (this is not the computational time).
+details are given in the Supplementary.} and, most notably, time elapsed is always given in
+\emph{physical} time, i.e., the amount of time that the agent spent
+interacting with the environment (this is not the number of steps).
 
@@
 
-We plot the learnt policy at several instants in physical time (or
-normalized epochs) during training, for various time discretizations
-$\deltat$, for both DAU and DDPG. With DAU, the agent's policy and value function quickly
-converge for every time discretization without specific
-tuning. On the contrary, with DDPG, learning of both value function
-and policy greatly vary from one discretization to another.
+We plot the learnt policy at several instants in physical time
+(normalized epochs) during training, for various time discretizations
+$\deltat$, for both DAU and DDPG. With DAU, the agent's policy and value
+function quickly converge for every time discretization without specific
+tuning. On the contrary, with DDPG, learning of both value function and
+policy vary greatly from one discretization to another.
 

### 2019-01-23 paper/icml_drau/experiments.tex

@@
 Pendulum, Bipedal Walker and Ant, DAU is quite robust to variations of
-$\deltat$ and displays reasonable performances in all cases. On the other hand,
-DDPG's performance varies with $\deltat$; performance are either degraded as
-$\deltat$ decreases (Ant, Cheetah), or become more variable as learning
-progresses (Pendulum) with small $\deltat$'s. On Cartpole, noise dominates,
-which makes interpretation difficult. On Half-Cheetah, DAU is not clearly
+$\deltat$ and displays reasonable performance in all cases. On the other hand,
+DDPG's performance varies with $\deltat$; performance either degrades as
+$\deltat$ decreases (Ant, Cheetah), or becomes more variable as learning
+progresses (Pendulum) for small $\deltat$. On Cartpole, noise dominates,
+making interpretation difficult. On Half-Cheetah, DAU is not clearly
 invariant to time discretization. This could be explained by the multiple
 suboptimal regimes that coexist in the Half-Cheetah environment (walking on the
-head, walking on the back).
+head, walking on the back), which create discontinuities in the value
+function (see Discussion).
 % \begin{figure}[h]

### 2019-01-23 paper/icml_drau/discussions.tex

@@
 
-The method derived in this work seems to yield improved robustness to
-time discretization on various environments, e.g. pendulum swing up, or
-simple locomotion tasks.  However on certain environments, there is still
-room for improvment. In what follows, we identify some of the issues
+The method derived in this work is theoretically invariant to time
+discretization, and
+indeed seems to yield improved timestep robustness
+on various environments, e.g.,
+simple locomotion tasks.  However, on some environments there is still
+room for improvment. We discuss some of the issues
 that could explain this theoretical/practical discrepancy.
@@
 %discontinuous on the frontier between $\mathcal{G}$ and $\mathcal{B}$, which 
-This results in non infinitesimal advantages
+This results in non-infinitesimal advantages
 for actions on the boundary. In such environments where a given policy
 has different regimes depending on the initial state, the continuous-time
-analysis would only hold almost-everywhere.
+analysis only holds almost-everywhere.
 % more generally, as soon as a given policy has different regimes depending on
@@
 
-\paragraph{Memory buffer size scalings}~Thm.~\ref{th:cont-params} is stated for
+\paragraph{Memory buffer size.}~Thm.~\ref{th:cont-params} is stated for
 transitions sampled sequentially from a fixed trajectory. In practice,
-transitions are sampled from a memory buffer, to prevent excessive correlations.
-We used a fixed size circular buffer, filled
+transitions are sampled from a memory replay buffer, to prevent excessive correlations.
+We used a fixed-size circular buffer, filled
 with samples from a single growing exploratory trajectory. In
-our experiments, the same buffer size is used for all time discretizations. 
+our experiments, the same buffer size was used for all time discretizations. 
 Thus the physical-time freshness of samples in the buffer varies with the
@@
 breaks timestep invariance. A memory-intensive option would be to use a
-buffer of size $\frac{1}{\deltat}$ (fixed physical time); our theoretical
-analysis does not cover this case.
+buffer of size $\frac{1}{\deltat}$ (fixed memory in physical time).
+%; our theoretical analysis does not cover this case.
 % When
@@
 
+\paragraph{Knowing $\deltat$.} Applying Alg.~\ref{alg:dau} requires
+knowledge of the timestep $\deltat$. In most environments, this is
+readily available, or even directly set by the practitioner: depending on
+the environment it is given by the frame rate, the maximum frequency of
+actuators or observation acquisition, etc.
+
 \paragraph{Near-continuous reinforcement learning and RMSProp.}
@@
 deterministic environments, gradients typically scale as $\bigO(1)$ in
-term of $\deltat$, as seen in \eqref{eq:approx_bellman_A}.  In that case, RMSProp
+terms of $\deltat$, as seen in \eqref{eq:approx_bellman_A}.  In that case, RMSProp
 preconditioning has no effect on the suitable order of magnitude for learning
-rates. However, in near continuous \emph{stochastic} environments (Eq.~\ref{eq:sde}), variance of the gradients typically scales as
-$\bigO\left(\frac{1}{\deltat}\right)$. With a fixed batch size,
+rates. However, in near continuous \emph{stochastic} environments
+(Eq.~\ref{eq:sde}), variance of $\delta
+Q_{\deltat}/\deltat$ and of the gradients typically scales as
+$\bigO\left(1/\deltat\right)$. With a fixed batch size,
 RMSProp will multiply gradients by a factor $\bigO(\sqrt{\deltat})$. In
 that case,
-learning rates need only be scaled as $\alpha \,\sqrt{\deltat}$ instead of
-$\alpha \,\deltat$.
+learning rates need only be scaled as $\sqrt{\deltat}$ instead of
+$\deltat$.
 
@@
 memory or partial observability, care should be taken that the model is
-able to maintain memory for a non-infinitesimal physical-time when
+able to maintain memory for a non-infinitesimal physical time when
 $\deltat\to 0$ (see e.g.~\citealt{chronornn}).

### 2019-01-23 paper/icml_drau/experiments.tex
Removed Normalized epoch

@@
 discretizations, without specific tuning, while standard deep $Q$-learning
-approaches do not. DAU is compared to DDPG for continuous actions and to
-DQN for
-discrete actions.
-Since we focus on $Q$-learning-like off-policy methods, we do not
-compare against on-policy methods (A3C~\cite{a3c}, PPO~\cite{ppo},
-TRPO~\cite{trpo}...).
+approaches do not. DAU is compared to DDPG for continuous actions and to DQN
+for discrete actions.  As mentionned earlier, we do not study the time
+discretization invariance of on-policy methods (A3C, PPO, TRPO...).
 
@@
 We plot the learnt policy at several instants in physical time
-(normalized epochs) during training, for various time discretizations
+during training, for various time discretizations
 $\deltat$, for both DAU and DDPG. With DAU, the agent's policy and value

### 2019-01-28 paper/thirdbell.tex

\documentclass[11pt,a4paper]{article}
\input{header.en.tex}

\title{The Third Bellman Equation}
\author{}

\newcommand{\drho}{\!\setminus\hspace{-.75em}\rho}
\newcommand{\ztar}{z^\mathrm{tar}}
\newcommand{\Nnewt}{N_\mathrm{steps/Newt}}

\begin{document}
\maketitle

\paragraph{Approximating $\Delta^{-1}$ via the Bellman--Newton equation.}
We learn an approximation $Z$ of $\Delta^{-1}$ via the Newton iteration
(V.Pan and J.Reif, \emph{Fast and efficient parallel solution of dense
linear systems}, 1989)
\begin{equation}
Z\gets 2Z-Z\Delta Z=2Z-Z^2+\gamma ZPZ
\end{equation}
which, if properly initialized, converges to $\Delta^{-1}$. A criterion
for correct initialization is if $\norm{\Id-Z\Delta}<1$ where $\norm{}$
is the operator norm. With $\Delta=\Id-\gamma P$ this is satisfied, for
instance, with $Z=\Id$ (for $\gamma<1$).

A stochastic version can be
\begin{equation}
Z\gets (1+\eta) Z-Z\tilde \Delta Z
\end{equation}
with a stochastic estimate $\tilde \Delta$ of $\Delta$ (obtained from a
transition observed in the MDP). Usually in RL we can only sample
transitions $s\to s'$ from some initial state $s$ distributed wrt some
distribution $\rho$ (which we may not control). We then parameterize
\begin{equation}
Z_{s_1s_2}\eqd \,\E_{s_2\sim \rho} z(s_1,s_2)
\end{equation}
namely, $z$ is the density of $Z$ with respect to $\rho(s_2)$.
Equivalently, in the matrix case, $Z=z\rho$ with $\rho$ the diagonal
matrix with entries $\rho$. Then the Newton update becomes
\begin{equation}
z\rho\gets 2z\rho-z\rho\Delta z\rho
\end{equation}
which is equivalent to
\begin{equation}
z\gets 2z-z\rho\Delta z
\end{equation}
and now $\rho \Delta$ is samplable:
\begin{equation}
\rho\Delta=\E_{s\sim \rho,\,s'\sim P(s'|s)}
\1_s\transp{(\1_s-\gamma\1_{s'})}
\end{equation}
so we can apply the stochastic version of the Newton update. Explicitly
this is
\begin{equation}
z(s_1,s_2)\gets (1+\eta) z(s_1,s_2)-
\E_{s\sim \rho,\,s'\sim P(s'|s)}z(s_1,s)\left(z(s,s_2)-\gamma
z(s',s_2)\right)
\end{equation}

In the tabular case, we can directly implement the above for all values
of $s_1,s_2$ each time we observe a transition $s\to s'$. (We can also
initialize to $\Id$.)

In the non-tabular case, we parameterize $z(s_1,s_2)$ by a machine
learning model. Then, given a current value $z_0$ of $z$, we to TD-like
steps towards the Newton update. More precisely, we define
\begin{equation}
\ztar\deq 2z_0-z_0\rho\Delta z_0
\end{equation}
and perform any number of SGD steps on the parameters of $z$ to reduce
the error $\E_{s_1\sim \rho,\,s_2\sim \rho}
\,(z(s_1,s_2)-\ztar(s_1,s_2))^2$, namely, we draw two states $s_1$ and
$s_2$ according to $\rho$, a transition $s\to s'$ with $s\sim \rho$, and
updated the parameter $\theta$ of $z$
by
\begin{equation}
\theta\gets \theta+\eta \partial_\theta z(s_1,s_2)\left(
2z_0(s_1,s_2)-z_0(s_1,s)\left(z_0(s,s_2)-\gamma z_0(s',s_2)\right)-z(s_1,s_2)
\right)
\end{equation}
which is an unbiased estimate of the gradient of the error.

We can sample several values of $s_1$ and $s_2$ for each sampled
transition $s\to s'$.  The distribution used on $s_1$ and $s_2$ need not
be the same distribution used for $s\to s'$ (though in the
parameterization by $Z-\Id$ this will be needed).

We perform $\Nnewt$ gradient steps before resetting the Bellman--Newton
objective using the current value of $z$ (namely, $z_0\gets z$ every
$\Nnewt$ steps). If $\Nnewt\gg 1$ and the machine learning model
converges, and $Z_0$ is initialized to $\Id$ (namely $z_0=1/\rho(s_2)$ at
startup) then this is guaranteed to converge.

\paragraph{What to do with $Z$.} Since $V=\Delta^{-1}R$, approximating
$\Delta^{-1}$ provides the ingredients for policy evaluation and policy
gradient. With the representation of $\Delta^{-1}$ as
$Z_{s_1s_2}=\E_{s_2\sim \rho} z(s_1,s_2)$, one has
\begin{equation}
V(s_1)=\E_{s_2\sim \rho}\, z(s_1,s_2) R(s_2)
\end{equation}
The representation of $V$ as an expectation may not be convenient. So if
$V$ is explicitly needed, we may need an auxiliary neural network to
learn $V(s_1)$ via the above (supervised learning). However, for policy
gradient the auxiliary network is not needed. Indeed, by the policy
gradient formula, on-policy (with $\mu$ the stationary distribution) we find
\begin{align}
\partial \E R&=
\E_{s\sim \mu,s'\sim P(s'|s)} \partial \ln
\pi(a|s) \left(
r(s,a,s')+\gamma V(s')-V(s)
\right)
\\&=
\E_{s\sim \mu,s'\sim P(s'|s),\,s_2\sim \rho} \partial \ln
\pi(a|s) \left(
r(s,a,s')+R(s_2)\left(\gamma z(s',s_2)-z(s,s_2)\right)
\right)
\end{align}
To exploit this, two possibilities. Either work online on $s\sim \mu$,
and use many samples from $s_2$ with a model for $R(s_2)$. Or (better),
work online on $s_2$ with an on-policy replay buffer on $s$, and update
several values of $s$ each time a reward is observed at $s_2$.

\paragraph{All rewards at once.} The relationship between $R$ and $V$ is
linear (for a fied policy). The generic representation of $Z$ above
requires computing a model of $V$ for each reward we want to work with.
We can directly exploit the linearity by parameterizing $z$ as
\begin{equation}
z(s_1,s_2)=\langle A(s_1),B(s_2)\rangle
\end{equation}
with some parameterized functions $A$ and $B$ with values in some vector
space with dot product $\langle\cdot,\cdot\rangle$. Intuitively, $A(s_1)$
represents the future of $s_1$, and $B(s_2)$ the past of $s_2$.
Then we have
\begin{equation}
V(s_1)=\langle A(s_1),\E_{s_2\sim \rho} B(s_2)R(s_2)\rangle
\end{equation}
so we can abuse notation and compute
\begin{equation}
B(R)\deq \E_{s_2\sim \rho} B(s_2)R(s_2)
\end{equation}
as the representation of $R$.

\end{document}

### 2019-01-28 paper/thirdbell.tex

@@
 \newcommand{\Nnewt}{N_\mathrm{steps/Newt}}
+\newcommand{\noise}{\epsilon}
 
@@
 \end{equation}
-as the representation of $R$.
+as the representation of $R$. Then
+\begin{equation}
+V(s_1)=\langle A(s_1),B(R)\rangle
+\end{equation}
+
+I'm not too sure about this representation. The rank of $Z$ is limited to
+the dimension of the vector space, and this may cause problems. An
+infinite-dimensional representation can be used, by having $A$ and $B$
+take values in a space of $L^2$ functions over some probability space.
+Namely, the dot product is
+\begin{equation}
+z(s_1,s_2)=\E_{\noise} A(s_1,\eps)B(s_2,\eps)
+\end{equation}
+and we can use as many samples of $\noise$ as we want. Obviously this
+adds noise.
 

### 2019-01-29 paper/thirdbell.tex

@@
 
+This computes the value functions of all rewards for a given policy.
+Obviously when changing the policy, we may not change the policy in
+several ways at once. For this we have to introduce targets (options) in
+the policy.
+
+\paragraph{All targets at once.} Let $\omega$ be an input to the policy,
+which represents its current ``target''. We want to learn a policy with
+input $\omega$, together with a function $\Omega$ that maps each state
+$s$ to a target $\Omega(s)$, such that if we follow the policy with
+target $\Omega(s)$, then we move towards $s$.
+
+Concretely, we now have a function $z(s_1,s_2;\omega)$ which corresponds
+to the inverse Laplacian of the MDP using policy $\omega$. It is learned
+as above. Then we learn the parameters of $\pi(a|s,\omega)$ and of
+$\Omega$ so as to maximize the time spent in $s$ when using policy
+$\pi(\cdot|\Omega(s))$. Concretely, policy gradient is
+\begin{equation}
+\E_{s_1\sim \mu,\,a\sim \pi(a|s_1,\Omega(s)),\,s'\sim P(s'|s_1,a)}
+\partial \ln \pi(a|s_1,\Omega(s))(\1_{s_1=s}+\gamma
+z(s',s;\Omega(s))-z(s_1,s;\Omega(s)))
+\end{equation}
+(TODO: the $\1_{s_1=s}$ becomes a Dirac in the continuous case??)
+
+I think the policy gradient above covers both the parameters of the
+policy and the parameters of $\Omega$ (we can pretend the policy takes
+$s$ as its target and that computing $\Omega$ is part of the policy).
+But this conflicts with my previous intuition that we should train
+$\Omega(s)$ to maximize $z(s',s;\Omega(s))$, which is the locally optimal
+choice. I think we are missing some terms coming from the gradients of
+$z(s_1,s;\Omega(s))$ when updating $\Omega$. TODO!! (Might be correct in
+the end. Write that we want to maximize expected time spent in $s$ for
+policy $\Omega(s)$. Write policy gradient for the policy with target $\Omega(s)$,
+seeing $\Omega$ as part of the policy. Its policy gradient involves the
+value function for policy $\Omega(s)$, which is given by $z(;\Omega(s))$,
+but it does not involves the gradients of the value function.)
+
+Problem here: $\mu$ has to be the invariant distribution of the
+$\Omega(s)$-policy, so a bit complicated to apply online, we need to keep
+the policy fixed until convergence... this implies having separate runs
+with different targets. Or find an off-policy version...
+
+$\omega$ represents an option, but there is no risk of option collapse
+because each option has a different target.
+
+This still does not cover policies with arbitrary rewards $R$. We have to
+find a representation $\Omega(R)$. We could try to enforce linearity of
+$\Omega$, namely, train things such that the policy with parameter
+$\omega=\E_{s\sim \rho} \Omega(s)R(s)$ maximizes $R$, for every $R$ in some
+class. If the class includes Dirac, then 
+the policy with parameter
+$\Omega(s)$ indeed maximizes time spend in $s$. (This might help with the
+Dirac in the policy gradient above.)
+
 \end{document}

### 2019-01-29 paper/icml_drau/impl_details.tex
.

@@
 %! TEX root = supplementary.tex
-All the details specifying our implementation are hereby given. We first give precise pseudo code
-descriptions for both \emph{Continuous Deep Advantage Updating} (Alg.~\ref{alg:cdau}), as well as the variants of DDPG (Alg.~\ref{alg:ddpg})
-and DQN (Alg.~\ref{alg:dqn}) used.
+All the details specifying our implementation are given in this section. We
+first give precise pseudo code descriptions for both \emph{Continuous Deep
+Advantage Updating} (Alg.~\ref{alg:cdau}), as well as the variants of DDPG
+(Alg.~\ref{alg:ddpg}) and DQN (Alg.~\ref{alg:dqn}) used.
 
@@
 	\item {\bf Cheetah (C)}: State normalization is used. Discretization range: $[0.05, 0.02, 0.01, 0.005, 0.002]$
-	\item {\bf Bipedal Walker (C)}: State normalization is used, $\alpha_2 = 0.02$. Discretization range: $[0.01, 0.005, 0.002, 0.001]$.
+	\item {\bf Bipedal Walker (C)\footnote{
+				The reward for Bipedal Walker is modified not to scale with $\deltat$. This does not introduce any change for the default setup.
+		}}: State normalization is used, $\alpha_2 = 0.02$. Discretization range: $[0.01, 0.005, 0.002, 0.001]$.
 	\item {\bf Cartpole (D)}: $\alpha_2 = 0.02$, $\tau = 0$. Discretization range: $[0.01, 0.005, 0.002, 0.001, 0.0005]$.

### 2019-01-29 paper/icml_drau/pendulum_figs_suppl.tex

@@
 The lighter the pixel, the closer to $1$ the action, the darker, the closer to $-1$.}
-        \label{fig:pend}
+	\label{fig:pend1}
 \end{figure*}
@@
 The lighter the pixel, the closer to $1$ the action, the darker, the closer to $-1$.}
-        \label{fig:pend}
+	\label{fig:pend2}
 \end{figure*}
@@
 	\caption{Value functions obtained by DDPG (scaled version) and AU at different instants in physical time of training on the pendulum swing-up environment.
-Each image represents the policy learnt by the policy network, with $x$-axis representing angle, and $y$-axis angular velocity.
-The lighter the pixel, the closer to $1$ the action, the darker, the closer to $-1$.}
-        \label{fig:pend}
+Each image represents the value function learnt, with $x$-axis representing angle, and $y$-axis angular velocity.
+The lighter the pixel, the higher the value.}
+	\label{fig:pend3}
 \end{figure*}
+\begin{figure*}[h]
+	\centering
+	\includegraphics[width=\cw]{figs_data/fig_unscaled/full_results_unscaled.pdf}
+	\label{fig:full_results}
+	\caption{Learning curves for DAU and DDPG (scaled) on classic control benchmarks for  various time discretization $\deltat$: Scaled return as a function of the physical time spent in the environment.}
+\end{figure*}

### 2019-01-29 paper/icml_drau/proof_conv_adv.tex

@@
 %! TEX root = supplementary.tex
-To make things more precise, we denote $\tilde{s}_\deltat^t(s_0)$ the
+Denote $\tilde{s}_\deltat^t(s_0)$ the
 evaluation at instant $t$ of the solution of $d\tilde{s}_t/dt = F(\tilde{s}_t,
@@
 
-Bellman equation on $Q^\pi_\deltat$ yields
+The Bellman equation on $Q^\pi_\deltat$ yields
 \begin{align}
@@
 
-For all $s$, a first order taylor expansion yields
+For all $s$, a first-order Taylor expansion yields
 \begin{equation}
@@
 \end{equation}
-thus, by uniform boundedness of the hessian of $V^\pi_\deltat$,\footnote{
-	Without boundedness of the hessian, we cannot write the second order taylor expansion of
+where the constant in $\bigO()$ is uniformly bounded thanks to the
+assumptions on the Hessian.
+Thus, by uniform boundedness of the Hessian of $V^\pi_\deltat$,\footnote{
+	Without boundedness of the Hessian, we cannot write the second order Taylor expansion of
 	$V^\pi_\deltat(\tilde{s}_\deltat^\deltat(s))$ in term of $\deltat$.

### 2019-01-29 paper/icml_drau/proof_conv_params.tex

--- 2019-01-22 paper/icml_drau/proof_conv_params.tex
@@
 \end{align}
-Both $V$ and $A$ are uniformly lipschitz continuous in $\theta$ and $\psi$, thus
-$F$ is lipschitz continuous.
+From the bounded Hessians and Gradients hypothesis, $V$, $A$, $\partial_s V$,
+$\partial_\theta V$ and $\partial_\psi A$ are uniformly Lipschitz continuous in
+$\theta$ and $\psi$, thus $F$ is Lipschitz continuous.
 

### 2019-01-29 code/mylog.py
Code should be nearly ready for publication

@@ -1,4 +1,6 @@
-"""Log facilities"""
+"""Log facilities, strongly inspired by OpenAI baselines:
+https://github.com/openai/baselines/blob/master/baselines/logger.py
+"""
 from abc import ABC, abstractmethod
 from os import makedirs
 from os import remove

### 2019-01-29 code/nn.py

@@ -1,85 +1,6 @@
 """Some nn utilities."""
-from typing import Tuple, Optional
 import torch
-from torch import Tensor
-import torch.nn.functional as f
-from torch.distributions import Normal
-from abstract import ParametricFunction, Shape
-
-def gmm_loss(batch: Tensor,
-             mus: Tensor,
-             sigmas: Tensor,
-             logpi: Tensor,
-             correction_factor: Optional[float] = None,
-             reduce: bool = True) -> Tensor: # pylint: disable=too-many-arguments
-    """ Computes the gmm loss.
-    Compute minus the log probability of batch under the GMM model described
-    by mus, sigmas, pi. Precisely, with bs1, bs2, ... the sizes of the batch
-    dimensions (several batch dimension are useful when you have both a batch
-    axis and a time step axis), gs the number of mixtures and fs the number of
-    features.
-    :args batch: (bs1, bs2, *, fs) torch tensor
-    :args mus: (bs1, bs2, *, gs, fs) torch tensor
-    :args sigmas: (bs1, bs2, *, gs, fs) torch tensor
-    :args logpi: (bs1, bs2, *, gs) torch tensor
-    :args reduce: if not reduce, the mean in the following formula is ommited
-    :returns:
-    loss(batch) = - mean_{i1=0..bs1, i2=0..bs2, ...} log(
-        sum_{k=1..gs} pi[i1, i2, ..., k] * N(
-            batch[i1, i2, ..., :] | mus[i1, i2, ..., k, :], sigmas[i1, i2, ..., k, :]))
-    NOTE: The loss is not reduced along the feature dimension (i.e. it should scale ~linearily
-    with fs).
-    """
-    batch = batch.unsqueeze(-2)
-    normal_dist = Normal(mus, sigmas)
-    g_log_probs = normal_dist.log_prob(batch)
-    g_log_probs = logpi + torch.sum(g_log_probs, dim=-1)
-    max_log_probs = torch.max(g_log_probs, dim=-1, keepdim=True)[0]
-    g_log_probs = g_log_probs - max_log_probs
-
-    g_probs = torch.exp(g_log_probs)
-    if correction_factor is not None:
-        mask = (g_probs[..., 0] == 1).float()
-        g_probs = torch.stack([
-            mask * g_probs[..., 0] * correction_factor + (1 - mask) * g_probs[..., 0],
-            g_probs[..., 1]], dim=1)
[…]

### 2019-01-29 paper/thirdbell.tex

@@
 \newcommand{\ztar}{z^\mathrm{tar}}
+\newcommand{\zetatar}{\zeta^\mathrm{tar}}
 \newcommand{\Nnewt}{N_\mathrm{steps/Newt}}
@@
 transitions $s\to s'$ from some initial state $s$ distributed wrt some
-distribution $\rho$ (which we may not control). We then parameterize
+distribution $\rho$ (which we may not control). This means we can sample
+from $\rho \Delta$, more precisely,
+\begin{equation}
+\rho\Delta=\E_{s\sim \rho,\,s'\sim P(s'|s)}
+\1_s\transp{(\1_s-\gamma\1_{s'})}
+\end{equation}
+A solution is to parameterize
 \begin{equation}
@@
 \end{equation}
-and now $\rho \Delta$ is samplable:
-\begin{equation}
-\rho\Delta=\E_{s\sim \rho,\,s'\sim P(s'|s)}
-\1_s\transp{(\1_s-\gamma\1_{s'})}
-\end{equation}
+and now $\rho \Delta$ is samplable,
 so we can apply the stochastic version of the Newton update. Explicitly
@@
 
+\paragraph{Parameterizing $Z-\Id$ in the continuous case.} I think it is quite
+important to initialize at $Z=\Id$, but this is not easy in the
+continuous case (initializing a neural network $z(s_1,s_2)$ to
+$\delta_{s_1s_2}/\rho(s_2)$ is not convenient). Moreover, there is an
+analytical singularity: for a discrete-time Markov process in continuous
+space, the fundamental matrix $Z=\sum_t \gamma^t P^t$ does have two
+analytically distinct components: for every $t\geq 1$ the term $P^t$ is
+smooth, while for $t=0$ it is a Dirac component along the diagonal
+$s_1=s_2$ (corresponding to $\Id$). So it may not be a good idea to
+represent them together in the same network. Instead, we can choose a
+representation of the operator $Z$ as $Z=\Id+\E_\rho \zeta$ or more
+precisely
+\begin{equation}
+Z_{s_1s_2}=\delta_{s_1s_2}+\E_{s_2\sim \rho} \zeta(s_1,s_2)
+\end{equation}
+so that the operator $Z$ acting on a reward $R$ is
+\begin{equation}
+V(s_1)=(ZR)(s_1)=R(s_1)+\int_{s_2} \zeta(s_1,s_2)R(s_2)\d\rho(s_2)
+\end{equation}
+
+Let us work out the Newton step in this representation. Let us first go
+back to the discrete case and matrices. Decomposing $Z=\Id+\zeta\rho$,
+the variable $z$ is $z=Z\rho^{-1}=\rho^{-1}+\zeta$, and the objective on
+$z$ is 
+\begin{equation}
+z\gets 2z-z\rho\Delta z=\rho^{-1}-\zeta\rho\zeta+\gamma(
+P\rho^{-1}+\zeta\rho P \rho^{-1}+P\zeta+\zeta\rho P \zeta)
+\end{equation}
+so that the objective on $\zeta$ is
+\begin{equation}
+\zeta\gets -\zeta\rho\zeta+\gamma(
+P\rho^{-1}+\zeta\rho P \rho^{-1}+P\zeta+\zeta\rho P \zeta)
+\end{equation}
+The terms $-\zeta\rho \zeta$ and $\zeta\rho
+P \zeta$ are similar to what we had with $z$, but the other terms are not
+obviously samplable.
+
+The term in $P$ is a model of the transition probabilities. The term in
+$P\zeta$ is the left Bellman equation, the term in $\zeta P$ is the right
+Bellman equation (up to densities $\rho$ since $\zeta$ is defined as a
+density). I think this is how we get the right mix of the three Bellman
+equations. (Although we evaluate $P$, we do not really develop a world
+model, we only need to estimate $P(s'|s)$ given $s$ and $s'$, not build
+$s'$ from scratch as a sample from $P(s'|s)$.)
+
+The non-samplability problem solves itself when considering the gradient of the
+error. Let $\zetatar$ be the target above on $\zeta$. Define the loss
+$\E_{s_1\sim \rho,\,s_2\sim \rho} (\zeta(s_1,s_2)-\zetatar(s_1,s_2))^2$
+as above, whose gradient is
+\begin{equation}
+\E_{s_1\sim \rho,\,s_2\sim \rho} \,\partial \zeta(s_1,s_2)(\zeta(s_1,s_2)-\zetatar(s_1,s_2))
+\end{equation}
+and let us study the contributions from each term of $\zetatar$, starting
+with the non-samplable term $\gamma P \rho^{-1}$. Its contribution to the
+gradient is
+\begin{align}
+\E_{s_1\sim \rho,\,s_2\sim \rho} \,\partial \zeta(s_1,s_2)
+P_{s_1s_2}/\rho(s_2)&=\E_{s_1\sim \rho}\sum_{s_2} \partial
+\zeta(s_1,s_2)P_{s_1s_2}
+\\&=\E_{s_1\sim \rho,\,s_2\sim P(s_2|s_1)}\partial
+\zeta(s_1,s_2)
+\end{align}
+that is, the gradient of $\zeta$ along a transition $s_1\to s_2$.
+So each time we sample a transition $s\to s'$, the term $
+P\rho^{-1}$ contributes $\gamma \partial
+\zeta(s,s')$ to the gradient. (This differs from the updates above, in
+that above we sampled $s\to s'$ and then modified the value of $z$ at
+$s_1$ and $s_2$ for independently sampled states $s_1$ and $s_2$; here
+sampling $s\to s'$ contributes a change of $\zeta$ at $s$ and $s'$.)
+
+Likewise, the gradient contribution of the $P\zeta$ term is
+\begin{align}
+\E_{s_1\sim \rho,\,s_2\sim \rho} \,\partial \zeta(s_1,s_2)
+\,(P\zeta)(s_1,s_2)&=\E_{s_1\sim \rho,\,s_2\sim \rho}
+\partial \zeta(s_1,s_2)\E_{s_3\sim P(s_3|s_1)}\zeta(s_3,s_2)
+\end{align}
+so this time the transition is $s_1\to s_3$. That is, when we sample a
+transition $s\to s'$, we have a gradient term $\E_{s_2\sim \rho}
+\partial\zeta(s,s_2)\zeta(s',s_2)$.
+
+The last non-obvious term is $\zeta\rho P \rho^{-1}$. Its gradient
+contribution is
+\begin{equation}
+\E_{s_1\sim \rho,\,s_2\sim \rho} \,\partial \zeta(s_1,s_2)\sum_{s_3}
+\zeta(s_1,s_3)\rho(s_3)P_{s_3s_2}/\rho(s_2)
+\end{equation}
+because the $s_1s_2$ entry of $\zeta\rho P \rho^{-1}$ is $\sum_{s_3}
+\zeta(s_1,s_3)\rho(s_3)P_{s_3s_2}/\rho(s_2)$. This rewrites as
+\begin{equation}
+\E_{s_1\sim \rho}\sum_{s_2} \E_{s_3\sim\rho} P_{s_3s_2}
+\zeta(s_1,s_3)\partial\zeta(s_1,s_2)
+\end{equation}
+which in turn rewrites as
+\begin{equation}
+\E_{s_1\sim \rho}\E_{s_3\sim\rho} \E_{s_2\sim P(s_2|s_3)}
+\zeta(s_1,s_3)\partial\zeta(s_1,s_2)
+\end{equation}
+that is, when we sample a transition $s\to s'$, we have a contribution
+$\E_{s_1\sim \rho}\zeta(s_1,s)\partial \zeta(s_1,s')$.
+
+Thus, when sampling $s\to s'$, we have several gradient contributions:
+one is similar to what we have before and involves
+$\partial\zeta(s_1,s_2)$ for independent samples $s_1,s_2$; one involves
+$\partial\zeta(s,s')$, one involves $\partial\zeta(s,s_2)$, and the last
+involves $\partial\zeta(s_1,s')$. All are samplable.
+
+TODO rewrite gradient with current $\zeta$ and previous $\zeta_0$, with
+all correct signs and $\gamma$ factors...
+
+% Starting with a
+% previous value $Z_0=\Id+\E_{\rho}\zeta_0$, or
+% $z_0(s_1,s_2)=\delta_{s_1s_2}/\rho(s_2)+\zeta_0(s_1,s_2)$, the target is
+% $2Z_0-Z_0\Delta Z_0$. The loss is $\E_{s_1\sim \rho,\,s_2\sim \rho}
+% \,(z(s_1,s_2)-\ztar(s_1,s_2))^2$, where $z$ and $\ztar$ are the density
+% of $Z$ and of the target with respect to $\rho(s_2)$. On the variable
+% $z$, the target $\ztar$ is
+% $\ztar=2z_0-z_0\rho\Delta z_0$ which rewrites as
+% \begin{equation}
+% \ztar(s_1,s_2)=2\frac{\delta_{s_1s_2}}{\rho(s_2)}+2\zeta(s_1,s_2)-(\frac{\delta_{s_1s_2}}{\rho(s_2)}+\ztar(s_1,s_2))...
+% \end{equation}
+% % The expected
+% % gradient of the loss is $2\E_{s_1\sim \rho,\,s_2\sim \rho}\partial
+% % z(s_1,s_2)(z(s_1,s_2)-\ztar(s_1,s_2))$. We find
+% % \begin{align}
+% % z(s_1,s_2)-\ztar(s_1,s_2)=\frac{\delta_{s_1s_2}}{\rho(s_2)}-2\frac{\delta_{s_1s_2}}{\rho(s_2)}-2\zeta_0(s_1,s_2)+
+% %\end{align}
+
 \end{document}

### 2019-01-29 code/dataanalysis/plots.py
Plot to suitable file

@@ -1,10 +1,10 @@
 import matplotlib
-# matplotlib.use('Agg')
 import matplotlib.pyplot as plt
 import numpy as np
 from scipy.ndimage.filters import gaussian_filter1d
 from scipy.interpolate import interp1d
 from typing import List
+from os.path import join
 
 
 def plot_learning_curves(expdata, key_list: List[str], namefile: str,
@@ -116,4 +116,4 @@ def plot_learning_curves(expdata, key_list: List[str], namefile: str,
 
     plt.tight_layout()
     # plt.savefig(namefile+'.eps', format="eps")
-    plt.savefig(namefile+'.png', format='png', dpi=1000)
+    plt.savefig(join('plots', namefile+'.png'), format='png', dpi=1000)

### 2019-01-29 paper/thirdbell.tex

@@
 \end{equation}
-namely, $z$ is the density of $Z$ with respect to $\rho(s_2)$.
+namely, $z$ is the density of $Z$ with respect to $\rho(s_2)$ (and as
+matrices,
+$Z=z\rho$).
 Equivalently, in the matrix case, $Z=z\rho$ with $\rho$ the diagonal
@@
 and perform any number of SGD steps on the parameters of $z$ to reduce
-the error $\E_{s_1\sim \rho,\,s_2\sim \rho}
-\,(z(s_1,s_2)-\ztar(s_1,s_2))^2$, namely, we draw two states $s_1$ and
+the error 
+\begin{equation}
+\ell(z)\deq\tfrac12\, \E_{s_1\sim \rho,\,s_2\sim \rho}
+\,(z(s_1,s_2)-\ztar(s_1,s_2))^2
+\end{equation}
+namely, we draw two states $s_1$ and
 $s_2$ according to $\rho$, a transition $s\to s'$ with $s\sim \rho$, and
-updated the parameter $\theta$ of $z$
-by
+update the parameter $\theta$ of $z$
+by the gradient
 \begin{equation}
-\theta\gets \theta+\eta \partial_\theta z(s_1,s_2)\left(
+\partial\ell =-\partial_\theta z(s_1,s_2)\left(
 2z_0(s_1,s_2)-z_0(s_1,s)\left(z_0(s,s_2)-\gamma z_0(s',s_2)\right)-z(s_1,s_2)
@@
 \end{equation}
-which is an unbiased estimate of the gradient of the error.
+which is an unbiased estimate of the gradient $\partial \ell$ of the error.
 
@@
 represent them together in the same network. Instead, we can choose a
-representation of the operator $Z$ as $Z=\Id+\E_\rho \zeta$ or more
-precisely
+representation of the operator $Z$ as $Z=\Id+\zeta\rho$ or equivalently
 \begin{equation}
@@
 So each time we sample a transition $s\to s'$, the term $
-P\rho^{-1}$ contributes $\gamma \partial
+\gamma P\rho^{-1}$ contributes $\gamma \partial
 \zeta(s,s')$ to the gradient. (This differs from the updates above, in
@@
 transition $s\to s'$, we have a gradient term $\E_{s_2\sim \rho}
-\partial\zeta(s,s_2)\zeta(s',s_2)$.
+\partial\zeta(s,s_2)\zeta(s',s_2)$. This term is absolutely similar to
+standard TD (for which $s_2$ is fixed).
 
@@
 $\partial\zeta(s,s')$, one involves $\partial\zeta(s,s_2)$, and the last
-involves $\partial\zeta(s_1,s')$. All are samplable.
+involves $\partial\zeta(s_1,s')$. All are samplable. For each transition
+$s\to s'$, we can
+sample as many states $s_1,s_2$ as we want.
 
 TODO rewrite gradient with current $\zeta$ and previous $\zeta_0$, with
-all correct signs and $\gamma$ factors...
+all correct signs and $\gamma$ factors... and collect into a proposition
+about $\partial \ell$.
+
+TODO might be less easy to use in policy gradient: because of the
+expression for $V$ we have contributions $R(s')$.
 

### 2019-01-30 paper/icml_drau/advup.tex

@@
 \label{eq:Aparam}
-	A_\psi(s, a) \deq \bar{A}_\psi(s, a) - \bar{A}_\psi(s, \pi(s)).
+	A_\psi(s, a) \deq \bar{A}_\psi(s, a) - \bar{A}_\psi(s, \pi(s))
 \end{equation}
@@
 \end{align}
-as seen in \ref{subsec:reparam}. Then $V_{\theta}$
+as seen in \ref{subsec:reparam}. Thus $V_{\theta}$
 and $A_{\psi}$ are trained to approximately solve these equations.
 
-Maximization over actions in $\pi$ is implemented exactly for discrete actions,
-and approximated by a policy neural network $\pi_\phi(s)$, trained to maximize $A_\psi(s,
-\pi_\phi(s))$ for continuous actions, similarly to
+Maximization over actions for $\pi$ is implemented exactly for discrete actions,
+and, for continuous actions, approximated by a policy neural network $\pi_\phi(s)$ trained to maximize $A_\psi(s,
+\pi_\phi(s))$, similarly to
 \cite{ddpg}.
@@
 in~\ref{subsec:reparam}.  To approximately verify~\eqref{eq:bellman_A}, the
-corresponding squared residual is minimized by an approximate gradient descent.
+corresponding squared Bellman residual is minimized by an approximate gradient descent.
 The update equations when learning from a transition $(s, a, r, s')$, either from
@@
 Appropriate scalings for the learning rates $\eta^V_\deltat$ and
-$\eta^A_\deltat$ in term of $\deltat$ to obtain a well defined continuous
+$\eta^A_\deltat$ in terms of $\deltat$ to obtain a well defined continuous
 limit are derived next.
@@
 learning rates are of order $\deltat$, one would expect 
-the parameters $\theta$ and $\psi$ to move by $O(\deltat)$ in
+the parameters $\theta$ and $\psi$ to change by $O(\deltat)$ in
 each time interval $\deltat$, thus hopefully converging to smooth

### 2019-01-30 paper/icml_drau/ddpg.tex

%! TEX root = supplementary.tex

\begin{algorithmic}
	\STATE \textbf{Inputs:}
	\STATE $\psi$ and $\phi$, parameters of
	$Q_\psi$ and $\pi_\phi$.
	\STATE $\psi'$ and $\phi'$, parameters of target networks
	$Q_{\psi'}$ and $\pi_{\phi'}$.
	\STATE $\pi^{\text{explore}}$ and $\nu$ defining an exploration policy.
	\STATE \textbf{opt}$_Q$, \textbf{opt}$_\pi$, $\alpha^Q$ and $\alpha^\pi$, optimizers and learning rates.
	\STATE $\mathcal{D}$, buffer of transitions $(s, a, r, d, s')$, with $d$ the episode termination signal.
	\STATE $\gamma$ discount factor.
	\STATE $\tau$ target network update factor.
	\STATE \textbf{nb\_epochs} number of epochs.
	\STATE \textbf{nb\_steps}, number of steps per epoch.
	\STATE
	\STATE Observe initial state $s^0$
	\STATE $t \gets 0$
	\FOR {$e=0, \textbf{nb\_epochs}$}
	\FOR {$j=1, \textbf{nb\_steps}$}
	\STATE $a^k \leftarrow \pi^{\text{explore}}(s^k, \nu^k)$.
	\STATE Perform $a^k$ and observe $(r^{k+1}, d^{k+1}, s^{k+1})$.
	\STATE Store $(s^k, a^k, r^{k+1}, d^{k+1}, s^{k+1})$ in $\mathcal{D}$.
	\STATE $k \gets k + 1$
	\ENDFOR
	\FOR {$k=0, \text{nb\_learn}$}
	\STATE \text{Sample a batch of $N$ random transitions from $\mathcal{D}$}
	\STATE $\tilde{Q^i} \gets r^i + (1 - d^i) \gamma Q_{\psi'}(s'^i, \pi_{\phi'}(s'^i))$
	\STATE $\Delta \psi \gets \frac{1}{N}\sum\limits_{i=1}^N \left(Q^i - \tilde{Q^i}\right) \partial_\psi Q(s^i, a^i)$
	\STATE $\Delta \phi \gets \frac{1}{N} \sum\limits_{i=1}^N \partial_a Q_\psi(s^i, \pi_\phi(s^i)) \partial_\phi \pi_\phi(s^i)$
	\STATE Update $\psi$ with \textbf{opt}$_Q$, $\Delta \psi$ and learning rate $\alpha^Q$.
	\STATE Update $\phi$ with \textbf{opt}$_\pi$, $\Delta \phi$ and learning rate $\alpha^\pi$.
	\STATE $\psi' \gets \tau \psi' + (1 - \tau) \psi$
	\STATE $\phi' \gets \tau \phi' + (1 - \tau) \phi$
	\ENDFOR
	\ENDFOR
\end{algorithmic}

### 2019-01-30 paper/icml_drau/discussions.tex

@@
 that could explain this theoretical/practical discrepancy.
+
+Note that Alg.~\ref{alg:dau} requires
+knowledge of the timestep $\deltat$. In most environments, this is
+readily available, or even directly set by the practitioner: depending on
+the environment it is given by the frame rate, the maximum frequency of
+actuators or observation acquisition, the timestep of a physics
+simulator, etc.
 
@@
 
-\paragraph{Knowing $\deltat$.} Applying Alg.~\ref{alg:dau} requires
-knowledge of the timestep $\deltat$. In most environments, this is
-readily available, or even directly set by the practitioner: depending on
-the environment it is given by the frame rate, the maximum frequency of
-actuators or observation acquisition, etc.
-
 \paragraph{Near-continuous reinforcement learning and RMSProp.}

### 2019-01-30 paper/icml_drau/supplementary.tex
.

@@
 the continuous-time advantage function contains all the necessary
-information for policy values.
+information for policy improvement.
 
@@
 	%\label{eq:conv-value}
-	$V^\pi_\deltat(s) = V^\pi(s) + \smallo(\deltat)$
+	$V^\pi_\deltat(s) = V^\pi(s) + \smallo(1)$
 	%\end{equation}

### 2019-03-11 rebuttal/rebuttal.txt
Rebuttal

We thank the reviewers for their insightful comments. In what follows, we will
tried to extensively attend all of the mentionned concerns.

First off, and before going into answers to detailled comment, some of the
reviewers complained about the scale of the experiments:
Reviewer #8: "A few small scale experiments"
Reviewer #7: "the experiments are not ambitious in this sense"
Let us emphasize that the computational time required to run an experiment with
time discretization $\delta t$ is proportional to $\frac{1}{\delta t}$ (this is
the best you can expect, since you need to have the same amount of physical
time experience to learn the same behavior). With standard discretization, on
the mujoco environments, it takes approximately 1 hour with decent hardware to
obtain reasonable results. If we want to use a discretization 100 times finer
grained than the standard discretization, we can only expect to obtain
reasonable results after approximately 100 hours. Using radically finer grained
discretization, and typically going to the next order of magnitude would
require a full month of computation for a single experiment, which is
impossible to obtain with the ressources at our disposal.  At the moment, for
our smallest time discretizations, each experiment on Mujoco takes
approximately a day and a half. 
We propose adding the following mention as a footnote to the experimental section
in the text: ""

Reviewer #1:
Thank you for your encouraging review! We propose adding the following mention
at the end of the introduction, to make the contribution more precise.
"In what follows, theorem 1, 2 and 3 were informally stated in [Doya, Baird], but
not formally proven, while theorem 4 and 5 are novel".

Thank you for pointing out the theorem numbering issue.

Reviewer #2:
a. We believe there was a confusion between the BipedalWalker environment that
we are using (https://gym.openai.com/envs/BipedalWalker-v2/) whose best
obtainable result is around 300, and the walker2d environment which is used in
[Haarnoja 2019] whose return values are not comparable. For other environments,
(Ant and Cheetah) our baseline DDPG seem to behave comparably to their implementation,
and DAU provides better or comparable results on Ant, irrelevant of the time
discretization, and better results on Cheetah for the default time discretization. Do
note that, except for BipedalWalker, where our highest time discretiation is smaller
than the default one, the highest time discretization used is the standard one.

b.c. TODO

d. We have deffered the presentation of the discrete algorithm to the appendix, and
now present the continuous version in the main paper instead.

f. Thank you for pointing out the typos.

Reviewer #7:
- "Dueling DQN, as the closest relative to DAU should be a baseline in the
  experiment section.": We experimented with a DDPG variant of dueling DQN and
  found that it provided similar results to DDPG, we did not present the
  result, to prevent clouding the figures.
- In the $\delta t = 1$ case, we recover an algorithm which is close, but not exactly
equivalent to Dueling DQN, since
  1. We do not use a target network, while Dueling DQN does
  2. In their final version, Dueling DQN reparameterizes A as A(s, a) - mean_{a'} A(s, a'),
  while we remain closer to the theory and use A(s, a) - max_{a'} A(s, a') instead.
  We propose adding the following as a footnote in the text:
  "TODO"

- "The theoretical arguments are about the limit \delta t going to 0, however,
  the experiments are not ambitious in this sense, they barely cover two orders
  of magnitude." see comment above

- "you claim that performance “becomes more variable”, can you capture this
  diagnosis in a quantitative metric?". We are currently using the standard
  deviation across runs to quantify variability. For DDPG, the pendulum
  learning curves show increased standard deviation when $\delta t$ becomes
  smaller. We propose rephrasing "becomes more variable" by "the standard
  deviation accross runs increases".

- memory buffer: When experimenting, we found that, for all $\delta t$'s, using
  too small replay buffers had a very negative impact on performances. This is why
  we chose to use one, even though this does not exactly match our theory.

Thanks for pointing out the typos, we will make sure to fix them.

### 2019-03-11 rebuttal/rebuttal.txt
Rebuttal

@@
 Reviewer #7:
+- "on three domains, the baseline algorithm (DDPG) is in fact not sensitive to \delta t"
+  This is factually incorrect. On Pendulum, Cartpole, Ant and Cheetah, 
 - "Dueling DQN, as the closest relative to DAU should be a baseline in the
@@
 Thanks for pointing out the typos, we will make sure to fix them.
+
+Reviewer #8:

### 2019-03-11 rebuttal/rebuttal.txt
Rebuttal

--- 2019-03-11 rebuttal/rebuttal.txt
@@
 
-First off, and before going into answers to detailled comment, some of the
-reviewers complained about the scale of the experiments:
+First off, and before going into detailled answers to specific comments, some of the
+reviewers mentionned that experiments were relatively small scale:
 Reviewer #8: "A few small scale experiments"
-Reviewer #7: "the experiments are not ambitious in this sense"
-Let us emphasize that the computational time required to run an experiment with
-time discretization $\delta t$ is proportional to $\frac{1}{\delta t}$ (this is
-the best you can expect, since you need to have the same amount of physical
-time experience to learn the same behavior). With standard discretization, on
-the mujoco environments, it takes approximately 1 hour with decent hardware to
-obtain reasonable results. If we want to use a discretization 100 times finer
-grained than the standard discretization, we can only expect to obtain
-reasonable results after approximately 100 hours. Using radically finer grained
+Reviewer #7: "the experiments are not ambitious in this sense".
+The computational time required to run an experiment with time discretization
+$\delta t$ is proportional to $\frac{1}{\delta t}$ (this is the best you can
+expect, since you need to have the same amount of physical time experience to
+learn the same behavior). With standard discretization, on the mujoco
+environments, it takes approximately 1 hour with decent hardware to obtain
+reasonable results. If we want to use a discretization 100 times finer grained
+than the standard discretization, we can only expect to obtain reasonable
+results after approximately 100 hours. Using radically finer grained
 discretization, and typically going to the next order of magnitude would
@@
 our smallest time discretizations, each experiment on Mujoco takes
-approximately a day and a half. 
-We propose adding the following mention as a footnote to the experimental section
-in the text: ""
-
-Reviewer #1:
-Thank you for your encouraging review! We propose adding the following mention
-at the end of the introduction, to make the contribution more precise.
-"In what follows, theorem 1, 2 and 3 were informally stated in [Doya, Baird], but
-not formally proven, while theorem 4 and 5 are novel".
-
-Thank you for pointing out the theorem numbering issue.
-
-Reviewer #2:
-a. We believe there was a confusion between the BipedalWalker environment that
-we are using (https://gym.openai.com/envs/BipedalWalker-v2/) whose best
-obtainable result is around 300, and the walker2d environment which is used in
-[Haarnoja 2019] whose return values are not comparable. For other environments,
-(Ant and Cheetah) our baseline DDPG seem to behave comparably to their implementation,
-and DAU provides better or comparable results on Ant, irrelevant of the time
-discretization, and better results on Cheetah for the default time discretization. Do
-note that, except for BipedalWalker, where our highest time discretiation is smaller
-than the default one, the highest time discretization used is the standard one.
-
-b.c. TODO
-
-d. We have deffered the presentation of the discrete algorithm to the appendix, and
-now present the continuous version in the main paper instead.
-
-f. Thank you for pointing out the typos.
+approximately a day and a half.  We propose adding the following mention as a
+footnote to the experimental section in the text: ""
 
 Reviewer #7:
+- "on three domains, the baseline algorithm (DDPG) is in fact not sensitive to
+  \delta t"
+  We are surprised by this comment. On pendulum, both in the
+  qualitative and quantitative experiments, the performance of DDPG decreases
+  significantly with the discretization timestep: qualitatively, DDPG quickly
+  converges to the optimal V function for $\delta t = 0.01$, converges slower
+  for $\delta t = 0.005$ and fails to converge for smaller $\delta t$'s.
+  Quantitatively, for DDPG, on all benchmark environments, with very few exceptions,
+  lighter curves (small $\delta t$'s) display significantly decreased performances 
+  compared to darker curves (high $\delta t$'s). On the Bipedal Walker, this
+  ranking is less visible, but still present at the end of training (from physical
+  time 8 hours to the end of training). On Cheetah and Ant, the ranking of the two
+  highest $\delta t$'s is at times reversed, but this is explained by the fact that
+  not much learning is happening for these $\delta t$'s.
+- "However there, DAU also exhibits such sensitivity, just in the reverse order
+
+(things get worse with coarser discretisation)."
 - "Dueling DQN, as the closest relative to DAU should be a baseline in the
@@
 Thanks for pointing out the typos, we will make sure to fix them.
+
+Reviewer #1:
+Thank you for your encouraging review! We propose adding the following mention
+at the end of the introduction, to make the contribution more precise.
+"To the best of our knowledge, theorem 1, 2 and 3 were informally stated in
+[Doya, Baird], but not formally proven, while theorem 4 and 5 are novel".
+
+Thank you for pointing out the theorem numbering issue.
+
+Reviewer #2:
+a. We believe there was a confusion between the BipedalWalker environment that
+we are using (https://gym.openai.com/envs/BipedalWalker-v2/) whose best
+obtainable result is around 300, and the walker2d environment which is used in
+[Haarnoja 2019] whose return values are not comparable. For other environments,
+(Ant and Cheetah) our baseline DDPG (with no hyperparameter tuning) seems to
+behave comparably to their implementation on the environments we have in
+common. DAU provides better or comparable results on Ant, irrelevant of the
+time discretization, and better results on Cheetah for the default time
+discretization. Do note that, except for BipedalWalker, where our highest time
+discretiation is smaller than the default one, the highest time discretization
+used is the default time discretization.
+
+b.c. TODO
+
+d. We have deffered the presentation of the discrete algorithm to the appendix, and
+now present the continuous version in the main paper instead.
+
+f. Thank you for pointing out the typos.
+
+Reviewer #8:
